### 2022-11-29 reference_notes/meng_eloyan_2021.tex
Initial Overleaf Import

\section*{Principal manifold estimation via model complexity selection}
\subhead{Kun Meng and Ani Eloyan, 2021}

\subsection*{Helpful Definitions}
\begin{itemize}
    \item Seminorm: the norm of a vector space which does not need to be positive definite
    \item Sobolev space: a vector space of functions that are sufficiently differentiable for some application, and that has a norm that measures the size and regularity of a function
    \item Hilbert space: A vector space with an inner product that defines a distance function for which the space is a complete metric space
    \item Reproducing kernel Hilbert space: A Hilbert space of functions in which point evaluation is a continuous linear functional
\end{itemize}

\subsection*{Introduction}
Generally speaking, manifold learning is an approach to estimating a low-dimensional manifold underlying higher-dimensional data. The manifold learning process can be broken down into two interwoven steps:
\begin{enumerate}
    \item Parameterization: identifying the low-dimensional representation of the data.
    \item Embedding: identifying a function that relates the low-dimensional manifold to the high-dimensional space that the data is in.
\end{enumerate}

This paper proposes an approach that combines these two steps, using a partial parameterization to estimate an embedding function, then finding the full parameterization from this function. This approach yields differentiable manifolds, and is able to generalize to data of any intrinsic dimension. The proposed method incorporates a model complexity selection method, which helps avoid overfitting, eliminate the effects of outliers, and reduce computational intensity.

\subsection*{Notation}
\begin{itemize}
    \item Positive integers $d$ and $D$ represent the intrinsic dimension of the data and the dimension of the of the embedded space of the data, respectively, with $d < D$.
    \item $\|x\|_{\mathcal{R}^q} = \left(\sum_{k=1}^{q} x_k^2\right)^{1/2}$.
    \item Let $q_1, q_2 \in \{d, D\}$ and $k \in \{1, 2, \dots, \infty\}$, and $I$ be a subset of $\mathbb{R}^{q_1}$. Then $\mathcal{C}^k(I \to \mathbb{R}^{q_2}$ denotes the collection of $I \to \mathbb{R}^{q_2}$ maps whose components have up to the $k^{\text{th}}$ continuous derivative. To simplify, use $\mathcal{C}^k(I) = \mathcal{C}^k(I \to \mathbb{R}^1)$ and $\mathcal{C} = \mathcal{C^0}$.
\end{itemize}

\subsection*{Manifold Learning Background}

### 2022-11-29 paper/annotated_bibliography.tex
move all paper files to directory

\documentclass[11pt,reqno]{article}
%\usepackage{cases}

\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
%\RequirePackage[numbers]{natbib}
\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
\RequirePackage{graphicx}%% uncomment this for including figures

\usepackage{amsmath, amssymb, amscd, amsthm, amsfonts}

\begin{document}

\title{Annotated Bibliography}
\author{Rob Zielinski}

\maketitle

\nocite{*}
\bibliographystyle{plain-annote}
\bibliography{references}

\end{document}

### 2022-11-29 paper/main.tex
move all paper files to directory

%\documentclass[12pt]{amsart}
\documentclass[11pt,reqno]{article}
%\usepackage{cases}

\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
%\RequirePackage[numbers]{natbib}
\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
\RequirePackage{graphicx}%% uncomment this for including figures

\usepackage{amsmath, amssymb, amscd, amsthm, amsfonts}
\usepackage{graphicx}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}

\usepackage{algorithm2e}

\usepackage{natbib}
\usepackage{authblk}

\usepackage[english]{babel}
\bibliographystyle{abbrvnat}
\setcitestyle{authoryear,open={(},close={)}}

\renewcommand{\baselinestretch}{1.0}
\setlength{\oddsidemargin}{-0.5cm}
\renewcommand{\topmargin}{-2cm}
\renewcommand{\oddsidemargin}{0mm}
\renewcommand{\evensidemargin}{0mm}
\renewcommand{\textwidth}{180mm}
\renewcommand{\textheight}{240mm}

\begin{document}

\title{Longitudinal Manifold Estimation}
\author[1]{Robert Zielinski and Ani Eloyan}

\maketitle

\section{Introduction}

Manifold learning is an approach to modeling high-dimensional data that is assumed to lie along a low-dimensional manifold. These approaches have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical structures. In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space. For example, Yue et al. (2016) seeks to compute a parametric representation of the corpus callosum. A number of methods have been developed for manifold learning, including Isomap (Tenenbaum et al. 2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003). Meng and Eloyan (2021) proposed a principal manifold framework that resolved several limitations of previously developed extensions of the principal curves approach introduced by Hastie and Stuetzle (1989).

In many medical imaging situations, understanding how an anatomical structure changes over time could provide meaningful information into the function of that structure, or medical conditions that affect that function. In situations where the structure is better represented by a low-dimensional manifold, interest carries over to understanding how the underlying manifold changes longitudinally as well.

While the methods given above provide a variety of approaches to retrieving a low-dimensional manifold from high-dimensional data at a single time point, they are ill-equipped to reach meaningful estimates across multiple time points. This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension-reduction process. This treatment of the time dimension causes time to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time.

Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in Meng and Eloyan (2021) to estimate an appropriate principal manifold at each given time point. We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.

[DISCUSS LONGITUDINAL DIMENSION REDUCTION APPROACHES CURRENTLY AVAILABLE - NEED TO REVIEW ARTICLES]

The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold estimation method proposed in Meng and Eloyan (2021), discuss how this approach can be adapted for longitudinal settings, and show the algorithm being used to accomplish this. Section 3 demonstrates the performance of this approach on simulated data. In Section 4, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease. The article finishes with a discussion of the method in Section 5.

\section{Methods}

Let $\{x_{it}\}_{i=1, t=1}^{I, T}$ be data points in $D$-dimensional space.

Outline:
\begin{enumerate}
    \item Introduce principal surfaces / principal manifolds, PME algorithm, and the loss function it minimizes
    \item Show updated loss function for LPME algorithm
    \item LPME algorithm
\end{enumerate}

\section{Simulations}

\section{Application}

\section{Discussion}

\RestyleAlgo{ruled}
\LinesNumbered

\SetKwComment{Comment}{/* }{ */}

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:two}
\KwData{Data points $\{y_{i, t}\}_{i, t}^{I, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\{\lambda_k\}_{k=1}^K$, $\{\gamma_l\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $g^*: \mathbb{R}^{d+1} \to \mathbb{R}^{D+1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, ... T$}{
    Apply PME algorithm with input $(\{y_{i, t}\}_{i=1}^I, N_0, \alpha, \epsilon, \epsilon^*, \{\lambda_k\}_{k=1}^K)$ and obtain $\{f_t^*\}_{t=1}^T$ and $\{\lambda_t^*\}_{t=1}^T$\;
    Use $\{f_t^*\}_{t=1}^T$ to map a grid of $d$-dimensional parameters $\{r_j\}_{j=1}^N$ to $\{x_{j, t}\}_{j=1}^N$ in $\mathbb{R}^D$\;
}
Apply ISOMAP to parameterize the newly generated collection $\{x_{j, t}\}_{j, t}^{N, T}$ by the $d$-dimensional parameters $\{r_j\}_{j=1}^N$\;
Use parameters $\{r_j\}_{j=1}^N$ and time values $\{t_j\}_{j=1}^N$ to create new
$\pi_{g_{(1)}}(x_{j}) \gets r_j$ for $j=1, ..., N$\;
\For{$l = 1, 2, ..., L$}{
    $\gamma \gets \gamma_l$.\;
    Obtain $g_{(1)}$ by solving 
    $\left(\begin{array}{ccc} 
    2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
    2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
    \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
    s_l \\
    \alpha_l \\
    m_l 
    \end{array}\right) = \left(\begin{array}{c}
    2\mathbf{E}\mathbf{W}\mathbf{x} \\
    2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
    \mathbf{0} 
    \end{array}\right)$\;
    $\mathcal{E} \gets 2\epsilon^*$\;
    $n \gets 1$\;
    $\mathcal{D}_{\hat{Q}, N}(g_{(1)}) \gets \sum_{j=1}^N\|x_j - g_{(1)}\|_{\mathbb{R}^{D}}^2$\;
    \While{$\mathcal{E} \geq \epsilon^*$, $n < itr$}{
        Compute $g_{(n + 1)}$ by solving $\left(\begin{array}{ccc} 
        2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
        2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
        \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
        s_l \\
        \alpha_l \\
        m_l 
        \end{array}\right) = \left(\begin{array}{c}
        2\mathbf{E}\mathbf{W}\mathbf{x} \\
        2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
        \mathbf{0} 
        \end{array}\right)$\;
        $\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) \gets \sum_{j=1}^N\|x_j - g_{(n + 1)}(\pi_{g_{(n + 1)}}(x_j))\|_{\mathbb{R}^D}^2$\;
        $\mathcal{E} \gets |\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(g_{(n)})|/\mathcal{D}_{\hat{Q}_N}(g_{(n)})$\;
        $n \gets n + 1$\;
    }
    $\hat{g}_l \gets g_{(n)}$\;
}
$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbb{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \arg \min_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbb{R}^D}^2: l = 1, 2, \dots, L\}$\;
$g^*(r) = g^{l^*}(\kappa r)$, where the analytic formula of $g^*$ is from $g_{(n + 1), l}(r) = \sum_{j=1}^N \eta_{4-d}(r - \pi_{g_{(n)}}(x_j)) + \sum_{k=1}^{d + 1} \alpha_{k, l} \times p_k(r)$, and $\gamma^* \gets \gamma_{l^*}$
\end{algorithm}

	\begin{equation}\label{PMSEF}
	\mathcal{K}_{\lambda,\mathbb{P}}(f,g)=\mathbb{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbb{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbb{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbb{R})}^2, 
	\end{equation}
	\begin{equation}\nonumber
	\mathcal{K}_{\lambda,\mathbb{P}}(f)=\mathcal{K}_{\lambda,\mathbb{P}}(f,f),
	\end{equation}
	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbb{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbb{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$. A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbb{P}$) with the tuning parameter $\lambda$ if 
	\begin{align}\label{def: PM}
	f^*=\arg\min_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
	\end{align}

Questions:
\begin{enumerate}
    \item How can we ensure a consistent parameterization within the PME algorithm? If we aren't able to do that, can we modify parameter values after the fact to ensure that parameterization aligns at each time step? Do modifications need to differ based on the shape to be fit, or can we find universal function that generates consistency? I believe our ability to do this is key for being able to reach accurate longitudinal estimates.
    \item Should we worry about "trimming" the manifold estimates? If so, how should we do that?
    \item What is the best way to evaluate the performance of the estimates? We can use SSD for the embedded values, but how to assess the parameterized values?
    \item How can we summarize information about the estimated manifolds? How can we make this process easy to interpret from clinical perspective?
    \item Should there be updates to what is defined as the "optimal" model in the PME algorithm? Currently it is possible for a model with non-minimal MSD to be chosen.
    \item Algorithm seems to struggle more when fitting to data with greater translations between time points. Is this something we will need to worry about in practice, or could it be solved through preprocessing and registration?
    \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?
\end{enumerate}

\nocite{*}
\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2022-12-14 paper/reference_notes/meng_eloyan_2021.tex
add pme paper reference notes

\section*{Principal manifold estimation via model complexity selection}
\subhead{Kun Meng and Ani Eloyan, 2021}

\subsection*{Helpful Definitions}
\begin{itemize}
    \item Seminorm: the norm of a vector space which does not need to be positive definite
    \item Sobolev space: a vector space of functions that are sufficiently differentiable for some application, and that has a norm that measures the size and regularity of a function
    \item Hilbert space: A vector space with an inner product that defines a distance function for which the space is a complete metric space
    \item Reproducing kernel Hilbert space: A Hilbert space of functions in which point evaluation is a continuous linear functional
    \item Lebesgue space: A function space defined using a generalization of the p-norm for finite-dimensional vector spaces.
\end{itemize}

\subsection*{Introduction}
Generally speaking, manifold learning is an approach to estimating a low-dimensional manifold underlying higher-dimensional data. The manifold learning process can be broken down into two interwoven steps:
\begin{enumerate}
    \item Parameterization: identifying the low-dimensional representation of the data.
    \item Embedding: identifying a function that relates the low-dimensional manifold to the high-dimensional space that the data is in.
\end{enumerate}

This paper proposes an approach that combines these two steps, using a partial parameterization to estimate an embedding function, then finding the full parameterization from this function. This approach yields differentiable manifolds, and is able to generalize to data of any intrinsic dimension. The proposed method incorporates a model complexity selection method, which helps avoid overfitting, eliminate the effects of outliers, and reduce computational intensity.

Approaches commonly used for manifold estimation include ISOMAP, locally linear embedding, and Laplacian eigenmaps, each of which address the problem of developing parameterizations of high-dimensional data. Hastie and Stuetzle's principal curve framework is an approach to the embedding problem.

The principal curve framework is defined as follows: Let $I \in \mathbb{R}^{1}$ be a closed and possibly infinite interval. Suppose a map $f: I \to \mathbb{R}^{D}$ satisfies the following HS conditions:
\begin{enumerate}
  \label{enum:hs_conditions}

  \item $f \in \mathcal{C}^{\infty}(I \to \mathbb{R}^{D})$ ($f$ is infinitely differentiable)
  \item $\|f'(t)\|_{\mathbb{R}^{D}} = 1$ for all $t \in I$ ($f$ is arc-length parameterized)
  \item $f$ does not self intersect, meaning $t_1 \neq t_2$ implies $f(t_1) \neq f(t_2)$ (this means that we aren't able to fit closed surfaces because $f$ will not connect with itself)
  \item $\int_{\{t: f(t) \in B\}}dt < \infty$ for any finite ball $B$ in $\mathbb{R}^{D}$
\end{enumerate}

Then, $\pi_f: \mathbb{R}^{D} \to I$ is called the projection index with respect to $f$ and is defined as follows:
\[%
  \pi_f(x) = \text{sup}\left\{t \in I: \|x - f(t)\|_{\mathbb{R}^{D}} = \text{inf}_{t' \in I}\|x - f(t')\|_{\mathbb{R}^{D}}\right\}, \ \text{for all} \ x \in \mathbb{R}^{D}
.\]%

Essentially, this means that we denote the map from the high-dimensional space to the one-dimensional interval as $\pi_f$, and define it as the function that returns the least upper bound of the values of $t$ where the norm of $x - f(t)$ is equal to the greatest lower bound for $\|f - f(t')\|_{\mathbb{R}^{D}}$ such that $t' \in I$. Informally, we can think of this roughly meaning that $\pi_f$ gives the maximum value of $t$ that minimizes the $D$-dimensional difference between $x$ and $f(t)$.

Next, we suppose that $X$ is a continuous random $D$-vector with finite second moments. Principal curves of $X$ are all maps $f: I \to \mathbb{R}^{D}$ that satisfy the HS conditions above and the self-consistency condition, which is defined as:
\[%
  E(X | \pi_f(X) = t) = f(t)
.\]%

The HS conditions have the following restrictions:
\begin{enumerate}
  \label{enum:hs_restrictions}

  \item HS condition 2 requires that principal curves are arc-length parameterized, but this does not generalize to higher dimensions (essentially, they are principal curves, not surfaces).
  \item HS condition 3 eliminates the possibility of fitting a number of curves in practice because the curve does not self-intersect.
  \item It is difficult to verify whether HS condition 4 is met.
\end{enumerate}

While HS showed that principal curves are critical points of $\mathcal{D}_X(f)$ because they satisfy the self-consistency condition, Duchamp et al. (1996) showed that they may be saddle points, meaning that the principal curves are not necessarily the optimal curve under MSD. This is a flaw of both the HS framework and a related framework proposed by Tibshirani (1992), which also relied on self-consistency.

One approach to avoiding the saddle point issue is to avoid enforcing self-consistency and instead define principal curves as a curve minimizing MSD under a length constraint or regularity penalty. Using this approach, Smola et al. (2001) proposed a regularized principal manifold framework, where principal manifolds have the following form:

\[%
  \arg \min_{f \in \mathcal{F}}\left\{E\|X - f(\pi_f(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|Pf\|_{\mathcal{H}}^2\right\}
,\]%

where $\mathcal{F}$ is a collection of functions, and $P$ maps $f$ into inner product space $\mathcal{H}$.

This framework also has several key limitations:
\begin{enumerate}
  \label{enum:smola_limitations}

  \item It is still unknown how to select $\lambda$.
  \item There is not a defined projection operator $P$ that corresponds to the selection of the tuning parameter $\lambda$.
  \item Smola et al. (2001) does not discuss assumptions on the collection $\mathcal{F}$ that are necessary so that $\pi_f$ is well defined.
  \item The function spaces $\mathcal{H}$ and $\mathcal{F}$ compatible with $P$ and $\pi_f$ are not defined
\end{enumerate}

This paper addresses those limitations.

\subsection*{Notation}
\begin{itemize}
    \item Positive integers $d$ and $D$ represent the intrinsic dimension of the data and the dimension of the of the embedded space of the data, respectively, with $d < D$.
    \item $\|x\|_{\mathbb{R}^q} = \left(\sum_{k=1}^{q} x_k^2\right)^{1/2}$.
    \item Let $q_1, q_2 \in \{d, D\}$ and $k \in \{1, 2, \dots, \infty\}$, and $I$ be a subset of $\mathbb{R}^{q_1}$. Then $\mathcal{C}^{k}(I \to \mathbb{R}^{q_2})$ denotes the collection of $I \to \mathbb{R}^{q_2}$ maps whose components have up to the $k^{\text{th}}$ continuous derivative. To simplify, use $\mathcal{C}^{k}(I) = \mathcal{C}^{k}(I \to \mathbb{R}^{1})$ and $\mathcal{C} = \mathcal{C}^{0}$.
    \item $\delta_x$ represents the point mass at $x$.
    \item $L^{p}$ and $\|\cdot\|_{L^{p}}$, with $p \in [1, \infty]$ represent Lebesgue spaces and their norms.
    \item The mean squared distance (MSD) functional is expressed $\mathcal{D}_X(f) = E\|X - f(\pi_f(X))\|_{\mathbb{R}^{D}}^2$.
    \item $\nabla f$ denotes the gradient of $f$, $\left(\frac{\partial f}{\partial t_1}(t), \frac{\partial f}{\partial t_2}(t), \dots, \frac{\partial f}{\partial t_d}(t)\right)^{T}$.
    \item $\|\nabla  f\|_{\mathbb{R}^{d}}$ denotes the scalar-valued function $\left(\sum_{i=1}^{d}\left|\frac{\partial f}{\partial t_i}(t)\right|^2\right)^{\frac{1}{2}}$.
    \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$, $\left(\frac{\partial^{2}f}{\partial t_i \partial t_j}(t)\right)_{1 \leq i, j \leq d}$.
    \item $\|\nabla f\|_{\mathbb{R}^{d}} \in L^2(\mathbb{R}^{d})$ denotes the integral of the norm of the gradient of $f$ over the Lebesgue space, $\|\nabla f\|_{L^2(\mathbb{R}^{d})}^{2} = \int_{\mathbb{R}^{d}} \sum_{i=1}^d \left|\frac{\partial f}{\partial t_i}(t)\right|^2dt < \infty$.
    \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})$ denotes the integral of the norm of the Hessian matrix of $f$ over the Lebesgue space, $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(t)\right|^2dt < \infty$.
\end{itemize}

\subsection*{Manifold Learning Background}

The paper first introduces the concepts of manifolds and projection indices.

Let $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The image of $f$, $M_f^{d} = \{f(t): t \in \mathbb{R}^{d}\}$ is called a $d$-dimensional manifold determined by $f$, $f$ is called an embedding map, and $\mathbb{R}^{D}$ is called the parameter space. If $f$ has an existing inverse $f^{-1}: M_f^{d} \to \mathbb{R}^{d}$ and it is continuous, then $f$ is called a homeomorphism.

In practice, $\mathbb{R}^{d}$ is the space containing the latent parameterizations $\{t_i\}_{i=1}^{I}$ of observed data $\{x_i\}_{i=1}^{I} \subset \mathbb{R}^{D}$.

Next, the paper generalizes the concept of the projection matrix $\pi_f$ from the HS case, which works in cases where $d=1$, to all possible intrinsic dimensions $d$. Generally, we are looking for the parameter $t$ that minimizes $\|x - f(t)\|_{\mathbb{R}^{D}}$, but it is possible that there are multiple such values of $t$. With this in mind, the paper introduces the following function space:

\[%
  \mathcal{C}_{\infty}\left(\mathbb{R}^{d} \to \mathbb{R}^{D}\right) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|t\|_{\mathbb{R}^{d}} \to \infty} \|f(t)\|_{\mathbb{R}^{D}} = \infty \right\}
.\]%

Essentially, this function space is the space of embedding maps such that as the norm of parameter $t$ in the $d$-dimensional space approaches infinity, the norm of $f(t)$ in the $D$-dimensional space is also infinite.

Then, the paper proves that if $f \in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, then the $d$-dimensional set 
\[%
  \mathcal{A}_f(x) = \left\{t \in \mathbb{R}^{d}: \|x - f(t)\|_{\mathbb{R}^{D}} = dist(x, f)\right\}
\]% 
is nonempty and compact for all $x \in \mathbb{R}^{D}$. If $\mathcal{A}_f(x)$ contains more than one element (meaning that there are multiple values of $t$ where the norm of the difference of $x$ and $f(t)$ are equal to the distance between $x$ and $f$), then the point $x$ is called an ambiguity point. Then, we define the new projection index as follows:

\[%
  \begin{aligned}
    \pi_f(x) = (t_1^{*}, t_2^{*}, \dots, t_d^{*}), \ \text{where} \\
    t_1^{*} = \max \left\{t_1: (t_1, t_2, \dots, t_d) \in \mathcal{A}_f(x)\right\}, \quad t_j^{*} = \max \left\{t_j: (t_1^{*}, \dots, t_{j-1}^{*}, t_j, \dots, t_d) \in \mathcal{A}_f(x)\right\}, \ j = 2, 3, \dots, d
  \end{aligned}
.\]%

If $d=1$, then this definition of the projection index is identical to the projection index as defined by HS.

\section*{Principal Manifolds}

The principal manifold framework proposed in this article is conceptualized as a generalization of principal component analysis.

The first $d$ linear principal components of a random $D$-vector $X$ are the $d$-dimensional hyperplane given by $\arg \min_{L \in \mathcal{L}}E\|X - \Pi_L(X)\|_{\mathbb{R}^{D}}^2$, with $\mathcal{L}$ being the collection of all $d$-dimensional hyperplanes in $\mathbb{R}^{D}$ and $\Pi_L(X)$ is the projection of $X$ to the hyperplane $L$. In the generalized principal manifold framework, we replace $\mathcal{L}$ with a Sobolev space. In the principal manifold framework, all derivatives (except those used in the definition of $\mathcal{C}^{k}(\mathbb{R}^{d} \to \mathbb{R}^{D})$) are considered generalized derivatives defined on $\mathcal{D}'(\mathbb{R}^{d})$, which is the collection of generalized functions on $\mathbb{R}^{d}$ (definitions from functional analysis).

The following function spaces are introduced:
\begin{itemize}
  \label{item:sobolev_spaces}

  \item $\nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D'}(\mathbb{R}^{d}): \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$. This is the space of the integrals of the norms of the Hessian matrices of function $f$, where $f$ belongs to the collection of generalized functions on $\mathbb{R}^{d}$.

  \end{itemize}
Much of the notation regarding Sobolev spaces is quite difficult to work through, so it is skipped here.

One of the limitations of the model proposed by Kegl et al. (2000) is that the fitted embedding function $f$ is not differentiable exactly everywhere, largely because the regularization $\|f'\|_{L^2}^2$ may not place a large enough penalty on non-smoothness. Instead of using the first derivative, this paper instead uses the second derivative. This has several advantages, one of which being that the total squared curvature of $f$, $\int \|f''(t)\|_{\mathbb{R}^{D}}^2 dt$, is the penalty term used in cubic smoothing splines. Thin plate splines use the penalty term $\sum_{l=1}^{D}\int_{\mathbb{R}^{d}}\sum_{i, j=1}^{2}\left|\frac{\partial^2f_l}{\partial t_i \partial t_j}(t)\right|^2dt$ to measure the roughness of the surface $M_f^2$.

To generalize these penalty terms to larger intrinsic dimensions, we use
\[%
  \|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D}\|\nabla^{\bigotimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l = 1}^{D}\int_{\mathbb{R}^{d}}\sum_{i, j=1}^{d} \left|\frac{\partial^2f_l}{\partial t_i \partial t_j}(t)\right|^2dt
.\]%

We call again call this generalized term the total squared curvature. Manifolds that have high total squared curvature values tend to be more complex and less stable, so we seek to penalize this value.

With this in mind, we let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$. We define the following functionals below:
\[%
  \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E \|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
.\]%

Then, we define a principal manifold for $X$ with the tuning parameter $\lambda$ as a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ that satisfies
\[%
  f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
,\]%

where $\mathcal{F}(\mathbb{P}) = \left\{f \in C_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}$.

If we know $f_{\lambda}^{*}$, then $\pi_{f_{\lambda}^{*}}(X)$ is the projection index giving a $d$-dimensional parameterization of $X$. Using the theorems presented earlier in the paper, we have the following properties:
\begin{itemize}
  \label{item:principal_manifold_properties}

  \item $\pi_{f_{\lambda}^{*}}(X)$ is continuous under certain conditions on $f_{\lambda}^{*}$.
  \item The parameterizations $\left\{\pi_f(x): x \in \text{supp}(\mathbb{P})\right\}$ are restricted to the unit ball $\left\{t \in \mathbb{R}^{d}: \|t\|_{\mathbb{R}^{d}} \leq 1\right\}$ (i.e. the norm of $t$ must be less than or equal to 1).
  \item Regularity for the principal manifold holds, meaning $f_{\lambda}^{*} \in \mathcal{C}^{k}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ for $k < \max\left\{2 - \frac{d}{2}, 1\right\}$.
  \item The proposed framework is a special case of the regularized principal manifolds from Smola et al. (2001), where $P = \nabla^{\bigotimes 2}$, $\mathcal{H} = L^2(\mathbb{R}^{d})$, and $\mathcal{F} = \mathcal{F}(\mathbb{P})$.
\end{itemize}

To estimate $f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K_{\lambda, \mathbb{P}}}(f)$, we use an iterative procedure with 
\[%
  f_{(n)} = \arg \min_f\left\{\mathcal{K}_{\lambda, \mathbb{P}}(f, f_{(n-1)}): f \in C_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})\right\}
.\]%

If the above procedure stops when $n = n^{*}$, we let $f_{(n^{*})} \in \mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2$. Then we obtain an estimate of $f_{\lambda}^{*}$ as $\hat{f_{\lambda}^{*}} = f_{(n^{*})}(\kappa t)$, where $\kappa = \sup\left\{\|\pi_{f_{(n^{*})}}(x)\|_{\mathbb{R}^{d}}: x \in \text{supp}(\mathbb{P})\right\}$, essentially meaning that $\kappa$ takes the maximum value of the norms of the parameterizations of all possible $x$ values.

Usually, this estimation approach approximates local minima, meaning that the final outcome is sensitive to the chosen starting values.

In cases with extreme values of $\lambda$, the proposed principal manifold framework takes the form of linear PCA when $\lambda = \infty$ and the HS principal curve method when $\lambda = 0$.

In the case where $\lambda > 0$ but very close to 0, it is possible to encounter numerical problems due to the potentially large value of total squared curvature. This issue is called curvature singularity at $\lambda = 0$. If this problem is present, the fitted manifold may be very rough and potentially unstable for small values of $\lambda$.

\section*{Estimating Principal Manifolds}

The general principal manifold estimation algorithm has three main steps: reduction, fitting, and tuning. In the reduction step, the complete dataset $\left\{x_i\right\}_{i = 1}^{I}$ is reduced to points $\left\{\mu_j\right\}_{j=1}^{N}$ with smaller sample size.  Each point $\mu_j$ is associated with a weight $\theta_j$. We then use \[%
  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j=1}^{N}\theta_j\|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
\]%
to approximate $\mathcal{K}_{\lambda, \mathbb{P}} \approx \frac{1}{I}\sum_{i=1}^{I}\|x_i - f(\pi_f(x_i))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$. This step helps to alleviate computational burden while mitigating the effects of outliers.

In the second step, we estimate $\hat{f_{\lambda}} = \arg \min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)$ for a predefined set of tuning parameters $\lambda > 0$. In the third step, we choose the optimal value of the tuning parameter, $\lambda^{*}$.

\subsection*{Data Reduction}

The data reduction step uses the high-dimensional mixture density estimation (HDMDE) algorithm, a high-dimensional generalization motivated by Eloyan and Ghosh (2011) to approximate the distribution of the data as a convex combination of point masses, represented by $\hat{Q}_N = \sum_{j=1}^{N}\theta_j\delta_{\mu_j}$. The HDMDE algorithm allows us to estimate $\mu_j, \theta_j, \ \text{and} \ N$ using a mixture density estimation approach. We assume that the data were generated from the pdf $p(x) = \psi * \mathcal{Q}^{\star}(x) = \int \psi(x - t)\mathcal{Q}^{\star}(dt)$, where $\mathcal{Q}^{\star}$ is the probability measure responsible for generating the latent data, and $\psi(\cdot - t)$ is the pdf used in the data corruption stage. Then, we let $\psi_{\sigma}(x) = \frac{1}{\sigma^{D}}\psi\left(\frac{x}{\sigma}\right)$, $\left\{\mu_{j, N}\right\}_{j=1}^{N}$ be a collection of points in $\mathbb{R}^{D}$ depending on $N$, $\theta_N$ be a set of probabilities adding to 1 used as weights for the $\mu$ values, and $\sigma_N$ be a positive number. The mixture density is expressed as
\[%
  p_N(x | \theta_N) = \psi_{\sigma_N} * \hat{Q}_N(x) = \sum_{j=1}^{N}\theta_{j, N}\psi_{\sigma_N}(x - \mu_{j, N})
.\]%

The proposed approach estimates the parameters of interest as follows:
\begin{itemize}
  \label{item:hdmde_approach}
  \item To estimate $\mu_{j, N}$, we partition $\left\{x_i\right\}_{i=1}^{I}$ into $N$ clusters using k-means clustering, and let $\left\{\mu_{j, N}\right\}_{j=1}^{J}$ be the centers of the clusters.
  \item To estimate $\sigma_N$, we let $\left\{x_{j, l}\right\}_{l=1}^{L_j}$ be $x_i$ in the $j$th cluster, and estimate $\sigma_N$ by the scaled average of the average norms of the differences between the centers of each cluster and their included data points,
    \[%
    \hat{\sigma}_N = \left\{\frac{1}{D \times N}\sum_{j=1}^{N}\left(\frac{1}{L_j}\sum_{l=1}^{L_j}\|x_{j, l} - \mu_{j, N}\|_{\mathbb{R}^{D}}^2\right)\right\}^{1/2}
    .\]%
  \item To estimate $\theta_{j, N}$, we use a constrained EM-algorithm to achieve likelihood maximization.
    \[%
      \theta_{j, N}^{(k)} = \frac{\sum_{i=1}^{I}w_{ij}\theta_N^{(k-1)}}{\hat{\rho}_1 + \hat{\rho}_2^{T}\mu_{j, N}}
    ,\]%
    where
    \[%
      (\hat{\rho}_1, \hat{\rho}_2) = \arg \min_{\rho_1 \in \mathbb{R}, \rho_2 \in \mathbb{R}^{D}}\left\{\left|\sum_{j=1}^{N}\left(\frac{\sum_{i=1}^{I}w_{ij}\theta_N^{(k-1)}}{\rho_1 + \rho_2^{T}\mu_{j, N}}\right) - 1\right|^2 + \left\|\sum_{j=1}^{N}\left(\frac{\sum_{i=1}^{I}w_{ij}\theta_N^{(k-1)}}{\rho_1 + \rho_2^{T}\mu_{j, N}}\right)\mu_{j, N} - \overline{x}\right\|_{\mathbb{R}^{D}}^2\right\}
    \]%
    and
    \[%
      w_{ij}(\theta_N) = \frac{\theta_{j, N} \times \psi_{\hat{\sigma}_N}(x_i - \mu_{j, N})}{\sum_{j' = 1}^{N}\theta_{j', N} \times \psi_{\hat{\sigma}_N}(x_i - \mu_{j', N})}
    .\]%
    This process runs until $\sup\left\{\left|\theta_{j, N}^{(k^{*})} - \theta_{j, N}^{(k^{*} - 1)}\right|: j = 1, 2, \dots, N\right\}$ is smaller than some predefined threshold.
\end{itemize}

Estimating the appropriate value of $N$ is done by looking at the $L^{1}$-distance between $p_N(\cdot | \theta_N) = \psi * \hat{Q}_N$ and the PDF $p$ generating data $x_i$. Specifically, the hypotheses are as follows:
\[%
H_0: E_p\left\{p_{N+1}(X | \theta_{N + 1}) - p_N(X | \theta_N)\right\} = 0 \quad \text{versus} \quad H_a: E_p\left\{p_{N + 1}(X | \theta_{N + 1}) - p_N(X | \theta_N)\right\} \neq 0
,\]%
plugging in the estimates $\hat{\theta}_N$ for $\theta_N$.

The full HDMDE algorithm involves iteratively combining the steps described above.

\begin{algorithm}
  \caption{HDMDE}
  \label{alg:two}
  \KwData{(i) Data points $\left\{x_i\right\}_{i=1}^{I}$ in $\mathbb{R}^{D}$, (ii) a positive integer $N_0 < I - 1$, and (iii) $\epsilon, \alpha \in (0, 1)$.}
  \KwResult{$N, \left\{\mu_{j, N}\right\}_{j=1}^{N}, \hat{\theta}_N = (\hat{\theta}_{1, N}, \dots, \hat{\theta}_{N, N})^{T}, \ \text{and} \ \sigma_N$.}
  $N \leftarrow N_0$ and formally $Z_{I, N} \leftarrow 2 \times z_{1-\alpha / 2}$\;
  Estimate $\mu_{j, N}$ and $\sigma_N$ using k-means clustering\;
  Use constrained EM-algorithm to calculate $\hat{\theta}_N$\;
  Compute $p_N(x_i | \hat{\theta}_N) = \sum_{j=1}^{N}\hat{\theta}_{j, N} \times \psi_{\sigma_N}(x_i - \mu_{j, N})$ for all $i$\;
  \While{$\left|Z_{I, N}\right| \geq z_{1-\alpha / 2}$ and $N < I - 1$}{
    $N \leftarrow N + 1$, repeat the steps 2, 3, 4, and compute $Z_{I, N}$\;
  }
\end{algorithm}

As discussed briefly above, this algorithm has several desirable properties. First, by using the cluster centers instead of individual data points, the algorithm greatly reduces the computational burden of manifold estimation. Second, the algorithm reduces the influence of outliers through the weights $\theta_{j, N}$. Third, the algorithm performs well in approximating the data generating distribution. The paper shows that minimizing $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ is approximately equivalent to minimizing $\mathcal{K}_{\lambda, \hat{Q}_N}(f)$.

\subsection*{Fitting}

To fit $\hat{f}_{\lambda} = \arg \min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)$, the PME algorithm uses the iterative procedure above:
\[%
  f_{(n + 1)} = \arg \min_f\left\{\mathcal{K}_{\lambda, \hat{Q}_N}(f, f_{(n)}): f = (f_1, f_2, \dots, f_D)^{T} \in \mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})\right\}
,\]%
where 
\[%
  \mathcal{K}_{\lambda, \hat{Q}_N}(f, f_{(n)}) = \sum_{l=1}^{D}\left\{\sum_{j=1}^{N}\theta_{j, N}\left|\mu_{j, N, l} - f_{l}\left(\pi_{f_{(n)}}(\mu_{j, N})\right)\right|^2 + \lambda \left\|\nabla^{\bigotimes 2}f_l\right\|_{L^2(\mathbb{R}^{d})}^2\right\}
,\]%
with $\mu_{j, N, l}$ being the $l$th component of the $D$-vector $\mu_{j, N}$.

If $\nu$ is an even integer, we let $\eta_{\nu}(t) = \left\|t\right\|_{\mathbb{R}^{d}}^{\nu}\log\left(\left\|t\right\|_{\mathbb{R}^{d}}\right)$ when $\left\|t\right\|_{\mathbb{R}^{d}} \neq 0$ and $\eta_{\nu}(t) = 0$ when $\left\|t\right\|_{\mathbb{R}^{d}} = 0$. Otherwise, we let $\eta_{\nu}(t) = \left\|t\right\|_{\mathbb{R}^{d}}^{\nu}$. Additionally, we let $Poly_1[t]$ be the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and linear basis $\left\{p_k\right\}_{k=1}^{d+1}$.

We see that the function in $\mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2$ that minimizes $\mathcal{K}_{\lambda, \hat{Q}_N}(\cdot, f_{(n)})$ is of spline form. When the intrinsic dimension is equal to 1, the minimizer is a cubic smoothing spline, and when the intrinsic dimension is 2, the minimizer is a thin plate spline.

Then, for $l = 1, \dots, D$, with
\begin{itemize}
  \label{item:minimization_notation}

  \item $\mathbf{T}$ being an $N \times (d + 1)$ matrix with the $(i, j)$th element equal to $p_j(\pi_{f_{(n)}}(\mu_{i, N}))$
  \item $\mu_l = (\mu_{1, N, l}, \mu_{2, N, l}, \dots, \mu_{N, N, l})$
  \item $\alpha_l = (\alpha_{1, l}, \alpha_{2, l}, \dots, \alpha_{d+1, l})^{T}$
  \item $s_l = (s_{1, l}, s_{2, l}, \dots, s_{N, l})^{T}$
  \item $\mathbf{E}$ being an $N \times N$ matrix with the $(i, j)$th element equal to $\eta_{4-d}\left(\pi_{f_{(n)}}(\mu_{i, N}) - \pi_{f_{(n)}}(\mu_{j, N})\right)$
  \item $\mathbf{W} = diag(\theta_{1, N}, \theta_{2, N}, \dots, \theta_{N, N})$
\end{itemize}

solving the minimization problem is equivalent to finding
\[%
  \arg \min_{(s_l, \alpha_l)}\left\{\left\|\mathbf{W}^{\frac{1}{2}}\left(\mu_l - \mathbf{E}s_l - \mathbf{T}\alpha_l\right)\right\|_{\mathbb{R}^{N}}^2 + \lambda\left\|\mathbf{E}^{\frac{1}{2}}s_l\right\|_{\mathbb{R}^{N}}^2: s_l \in \mathbb{R}^{N}, \alpha_l \in \mathbb{R}^{d+1}, \ \text{and} \ \mathbf{T}^{T}s_l = 0\right\}
.\]%

We can use the Lagrange multiplier approach to find the minimizers by solving

\[%
  \left(
    \begin{array}{ccc}
      2\mathbf{EWE} + 2\lambda\mathbf{E} & 2\mathbf{EWT} & \mathbf{T} \\
      2\mathbf{T}^{T}\mathbf{WE} & 2\mathbf{T}^{T}\mathbf{WT} & \mathbf{0} \\
      \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      s_l \\
      \alpha_l \\
      m_l
    \end{array}
  \right) = \left(
    \begin{array}{c}
      2\mathbf{EW}\mu_l \\
      2\mathbf{T}^{T}\mathbf{W}\mu_l \\
      \mathbf{0}
    \end{array}
  \right), \quad l = 1, 2, \dots, D
,\]%

where $m_l$ represent the Lagrange multipliers.

\subsection*{Tuning}

The approximate embedding map $\hat{f}_{\lambda}$ is estimated for each value of $\lambda > 0$. From these options, the optimal element $\hat{f}_{\lambda^{*}}$ is chosen by using the observed data $\left\{x_i\right\}_{i=1}^{I}$ to calculate the MSD $\mathcal{D}(\hat{f}_\lambda)$ as follows:
\[%
  \lambda^{*} = \arg\min_{\lambda > 0}\left\{\mathcal{D}(\hat{f}_{\lambda})\right\}, \ \text{where} \ \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\left\|x_i - \hat{f}_{\lambda}\left(\pi_{\hat{f}_{\lambda}}(x_i)\right)\right\|_{\mathbb{R}^{D}}^2
.\]%

The full PME algorithm can be expressed as follows:

\begin{algorithm}
  \caption{PME Algorithm}
  \label{alg:pme}
  \KwData{(i) Data points $\left\{x_i\right\}_{i=1}^{I}$ in $\mathbb{R}^{D}$; (ii) a positive integer $N_0 < I - 1$; (iii) $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; (iv) candidate tuning parameters $\left\{\lambda_g\right\}_{g=1}^{G}$; (v) $itr \geq 1$, the maximum number of iterations allowed.}
  \KwResult{(i) Analytic formula of $f^{*}: \mathbb{R}^{d} \to \mathbb{R}^{D}$ determining the fitted manifold $M_{f^{*}}^{d}$; (ii) optimal tuning parameter $\lambda^{*}$.}
  Apply HDMDE with input $(\left\{x_i\right\}_{i=1}^{I}, N_0, \epsilon, \alpha)$ and obtain $N$, $\left\{\mu_{j, N}\right\}_{j=1}^{N}$, and $\left\{\theta_{j, N}\right\}_{j=1}^{N}$\;
  Parameterization: Apply ISOMAP to parameterize the reduced collection $\left\{\mu_{j, N}\right\}_{j=1}^{N}$ by the $d$-dimensional parameters $\left\{t_j\right\}_{j=1}^{N}$. Formally set $\pi_{f_{(0)}}(\mu_{j, N}) \leftarrow t_j$ for $j = 1, 2, \dots, N$\;
  \For{$g = 1, 2, \dots, G$}{
    $\lambda \leftarrow \lambda_g$ and obtain $f_{(1)}$ by solving the Lagrange multiplier equation\;
    $\mathcal{E} \leftarrow 2 \times \epsilon^{*}$, $n \leftarrow 1$, and $\mathcal{D}_{\hat{Q}_N}(f_{(1)}) \leftarrow \sum_{j=1}^{N}\theta_{j, N}\left\|\mu_{j, N} - f_{(1)}(\pi_{f_{(1)}}(\mu_{j, N}))\right\|_{\mathbb{R}^{D}}^2$\;
    \While{$\mathcal{E} \geq \epsilon^{*}$ and $n < itr$}{
      Compute $f_{(n + 1)}$ from $f_{(n)}$ by solving the Lagrange equation and compute SSD for the updated function estimate\;
      $\mathcal{E} \leftarrow \left|\left[\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(f_{(n)})\right] / \mathcal{D}_{\hat{Q}_N}(f_{(n)})\right|$ and $n \leftarrow n + 1$\;
    }
    $\hat{f}_g \leftarrow f_{(n)}$
  }
  $\kappa \leftarrow \max\left\{\left\|\pi_{\hat{f}_{g^{*}}}(x_i)\right\|_{\mathbb{R}^{d}}: i = 1, 2, \dots, I\right\}$, where $g^{*} = \arg\min_g\left\{\frac{1}{I}\sum_{i=1}^{I}\left\|x_i - \hat{f}_g\left(\pi_{\hat{f}_g}(x_i)\right)\right\|_{\mathbb{R}^{D}}^2: g = 1, 2, \dots, G\right\}$\;
  $f^{*}(t) = \hat{f}_{g^{*}}(\kappa t)$ and $\lambda^{*} \leftarrow \lambda_{g^{*}}$.
\end{algorithm}

\section*{Simulations}

The proposed PME algorithm is compared to other existing methods on data simulated from a variety of manifolds in three main dimensional cases: $(d = 1, D = 2)$, $(d = 1, D = 3)$, and $(d = 2, D = 3)$. In the first two dimensional cases, PME is compared to the HS principal curve algorithm and ISOMAP. In the $(d = 2, D = 3)$ case, PME is compared to the principal surface algorithm proposed in Yue et al. (2016) and ISOMAP. In several scenarios, the HS principal curve algorithm displays clear difficulties. PME and ISOMAP often show similar results. However, because ISOMAP uses the entire dataset, while PME first reduces the size of the data using HDMDE, PME runs more efficiently. Note that the PME algorithm can be made more efficient through relatively simple optimization steps, or potentially by recoding in C++.

\section*{Interior Identification}

The interior identification section is not clearly relevant to my work at this time.

### 2022-12-14 code/adni_mri_preprocess.R
extract surface points from adni hippocampus data

@@ -29,41 +29,45 @@ foreach (dir_idx = 1:length(sub_dirs)) %do% {
   nii_scans[[dir_idx]] <- list()
 
   for (img_idx in 1:length(img_dirs)) {
-    all_slices <- readDICOM(img_dirs[img_idx])
-    nii <- dicom2nifti(all_slices)
-
-    # Inhomogeneity correction
-    # Brain extraction / segmentation
-    # Image Registration
-    # Tissue Class Segmentation
-
-    nii_scans[[dir_idx]][[img_idx]] <- nii
-
-    # inhomogeneity correction
-    # nii_bc <- bias_correct(nii, correction = "N4")
-
-    # brain extraction
-    # nii_bc_be <- fslbet(nii_bc)
-    # if brain extraction shows signs of issues in some images,
-    # extrantsr::fslbet_robust() may be useful
-    # includes options for neck correction, etc.
-
-    # register image to MNI space
-    # not using coregistration to allow for future population-wide analysis
-    # nii_reg <- registration(
-    #   nii,
-    #   skull_strip = FALSE,
-    #   correct = FALSE
-    # )
-
     proc_dir <- paste0("data/ADNI_processed", gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx]))
+    if (file.exists(paste0(proc_dir, "/-L_Hipp_first.nii.gz"))) {
+      next
+    } else {
+      all_slices <- readDICOM(img_dirs[img_idx])
+      nii <- dicom2nifti(all_slices)
 
-    run_first_all(
-      # nii_reg$outfile,
-      nii,
-      oprefix = proc_dir,
-      brain_extracted = TRUE
-    )
+      # Inhomogeneity correction
+      # Brain extraction / segmentation
+      # Image Registration
+      # Tissue Class Segmentation
+
+      nii_scans[[dir_idx]][[img_idx]] <- nii
+
+      # inhomogeneity correction
+      # nii_bc <- bias_correct(nii, correction = "N4")
+
+      # brain extraction
+      # nii_bc_be <- fslbet(nii_bc)
+      # if brain extraction shows signs of issues in some images,
+      # extrantsr::fslbet_robust() may be useful
[…]

### 2022-12-14 paper/main.tex
update paper questions

@@
 \begin{enumerate}
-    \item How can we ensure a consistent parameterization within the PME algorithm?
-If we aren't able to do that, can we modify parameter values after the fact to ensure that parameterization aligns at each time step?
-Do modifications need to differ based on the shape to be fit, or can we find universal function that generates consistency?
-I believe our ability to do this is key for being able to reach accurate longitudinal estimates.
-    \item Should we worry about "trimming" the manifold estimates? If so, how should we do that?
   \item What is the best way to evaluate the performance of the estimates? We can use SSD for the embedded values, but how to assess the parameterized values?
-    \item How can we summarize information about the estimated manifolds? How can we make this process easy to interpret from clinical perspective?
-    \item Should there be updates to what is defined as the "optimal" model in the PME algorithm? Currently it is possible for a model with non-minimal MSD to be chosen.
-    \item Algorithm seems to struggle more when fitting to data with greater translations between time points.
-Is this something we will need to worry about in practice, or could it be solved through preprocessing and registration?
+  \item Should there be updates to what is defined as the "optimal" model in the PME algorithm?
+Currently it is possible for a model with non-minimal SSD to be chosen within the fitting of a single tuning value.
+  \item Why would we see the SSD increase substantially in late optimization steps?
+  \item Are there cases where the SSD and MSD differ substantially? Why would that be the case?
+  \item Are there any techniques that help with large amounts of noise in the data?
+Data reduction should help with outliers, but likely not consistently high noise levels.
+Early indications that possibly allowing for higher tuning parameters would help?
+  \item Why does PME algorithm struggle on Swiss Roll cross-section?
+It seems to work well up until spiral reaches second layer.
+The manifold should never intersect, but from the angle from the origin does overlap.
+  \item Generally, what are the limitations on the estimated manifolds intersecting? How close can it come to intersecting without showing strange behavior?
   \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?

### 2023-02-09 Examples_for_PME.R
change to pme function name

@@ -12,6 +12,8 @@ library("plot3D")
 ####################################################################
 par(mfrow=c(2,2))
 
+source("code/pme.R")
+
 ### Case I.
 
 # Simulated data
@@ -63,7 +65,7 @@ e2=rnorm(I,mean = 0, sd=sd.noise)
 data.points=X+cbind(e1,e2)
 
 ptm <- proc.time()
-result=PME(x.obs=data.points, d=1)
+result=pme(x.obs=data.points, d=1)
 proc.time() - ptm
 
 f=result$embedding.map
@@ -91,7 +93,7 @@ for(i in 1:I){ X[i,]=manifold(t[i]) }
 data.points=X+cbind(e1,e2)
 
 ptm <- proc.time()
-result=PME(x.obs=data.points, d=1)
+result=pme(x.obs=data.points, d=1)
 proc.time() - ptm
 
 f=result$embedding.map
@@ -124,7 +126,7 @@ e3=rnorm(I,mean=0,sd=noise)
 data.points=X+cbind(e1,e2,e3)
 
 ptm <- proc.time()
-result=PME(x.obs=data.points, d=1)
+result=pme(x.obs=data.points, d=1)
 proc.time() - ptm
 
 f=result$embedding.map
@@ -157,7 +159,7 @@ e3=rnorm(I,mean=0,sd=noise)
 data.points=X+cbind(e1,e2,e3)
 
 ptm <- proc.time()
-result=PME(x.obs=data.points, d=1)
+result=pme(x.obs=data.points, d=1)
 proc.time() - ptm
 
 f=result$embedding.map

### 2023-02-09 paper/main.tex
minor change to test compilation
updates to introduction

@@
 
-Manifold learning is an approach to modeling high-dimensional data that is assumed to lie along a low-dimensional manifold.
-These approaches have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical structures.
+Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
+These methods have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical shapes.
 In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space.
@@
 
-[DISCUSS LONGITUDINAL DIMENSION REDUCTION APPROACHES CURRENTLY AVAILABLE - NEED TO REVIEW ARTICLES]
+Some work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by Kinson et al.
+(2019).
+To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this paper is novel.
 
-The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold estimation method proposed in Meng and Eloyan (2021), discuss how this approach can be adapted for longitudinal settings, and show the algorithm being used to accomplish this.
-Section 3 demonstrates the performance of this approach on simulated data.
-In Section 4, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease.
-The article finishes with a discussion of the method in Section 5.
+The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold estimation method proposed in Meng and Eloyan (2021).
+Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this.
+Section 4 demonstrates the performance of this approach on simulated data.
+In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease.
+The article finishes with a discussion of the method in Section 6.
 
-\section{Methods}
+\section{Principal Manifolds and Estimation Algorithm}
+
+Because the algorithm proposed in this article relies on the same definitions and notation as those introduced in Meng and Eloyan (2021), we provide it here.
+To generalize the principal curve framework proposed in Hastie and Stuetzle (1989) to intrinsic dimensions $d > 1$, Meng and Eloyan introduce the function space
+\[%
+  C_{\infty}(\mathbb{R}^d \to \mathbb{R}^D) = \left\{f \in C()\right\}
+.\]%
+
+\section{Longitudinal Principal Manifold Estimation Algorithm}
 
@@
   \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?
+  \item How much error does the registration process introduce? Can the LPME algorithm successfully handle a certain number of bad registrations or extractions?
 \end{enumerate}

### 2023-02-14 paper/main.tex
begin introduction rewrite

@@
 
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al (2021), Poewe et al (2017)).
+Frequently, interest centers around changes in one or more substructures.
+For example, it is common to observe atrophy in the hippocampus of those with Alzheimer's disease or cognitive impairment from other causes.
+Image segmentation allows for the extraction and analysis of these subcortical structures.
+
+Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach, and is the standard.
+However, this approach is highly time and resource intensive.
+In studies analyzing a large number of images, this may be cost prohibitive.
+Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability.
+Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns.
+While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
+
+To demonstrate the potential extent of the inaccuracies introduced by segmentation, using FSL-FIRST to segment the hippocampus in images of one healthy individual over the course of four year in the ADNI dataset, the volume of the left and right hippocampi decreased by 3.6\% and 5.4\%, respectively, while each hippocampus was estimated to have increased in volume in subsequent images twice.
+In a larger comparison of the accuracy of FSL-FIRST and FreeSurfer, Mulder et al.
+(2014) found that 6.9\% of segmentations by FIRST failed visual inspection, as did 7.5\% of segmentations for FreeSurfer.
+After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
+If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with large numbers of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation.
+Mulder et al.
+(2014) also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches.
+
+Ultimately, high levels of variability are present at the subcortical level regardless of the segmentation approach used, but may be particularly influential when using automated segmentation methods.
+To address these elevated noise levels, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
+This approach, which extends the principal manifold estimation algorithm introduced by Meng and Eloyan (2021) for a longitudinal setting, includes a regularization parameter to limit the effects of between time-point variability.
+
+\section{Introduction - Old}
+
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
@@
 \[%
-  C_{\infty}(\mathbb{R}^d \to \mathbb{R}^D) = \left\{f \in C()\right\}
+  C_{\infty}(\mathbf{R}^d \to \mathbf{R}^D) = \left\{f \in C()\right\}
 .\]%
@@
 \KwData{Data points $\{y_{i, t}\}_{i, t}^{I, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\{\lambda_k\}_{k=1}^K$, $\{\gamma_l\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
-\KwResult{Analytic formula of $g^*: \mathbb{R}^{d+1} \to \mathbb{R}^{D+1}$, optimal tuning parameter $\gamma^*$.}
+\KwResult{Analytic formula of $g^*: \mathbf{R}^{d+1} \to \mathbf{R}^{D+1}$, optimal tuning parameter $\gamma^*$.}
 \For{$t = 1, 2, ... T$}{
     Apply PME algorithm with input $(\{y_{i, t}\}_{i=1}^I, N_0, \alpha, \epsilon, \epsilon^*, \{\lambda_k\}_{k=1}^K)$ and obtain $\{f_t^*\}_{t=1}^T$ and $\{\lambda_t^*\}_{t=1}^T$\;
-    Use $\{f_t^*\}_{t=1}^T$ to map a grid of $d$-dimensional parameters $\{r_j\}_{j=1}^N$ to $\{x_{j, t}\}_{j=1}^N$ in $\mathbb{R}^D$\;
+    Use $\{f_t^*\}_{t=1}^T$ to map a grid of $d$-dimensional parameters $\{r_j\}_{j=1}^N$ to $\{x_{j, t}\}_{j=1}^N$ in $\mathbf{R}^D$\;
 }
@@
     $n \gets 1$\;
-    $\mathcal{D}_{\hat{Q}, N}(g_{(1)}) \gets \sum_{j=1}^N\|x_j - g_{(1)}\|_{\mathbb{R}^{D}}^2$\;
+    $\mathcal{D}_{\hat{Q}, N}(g_{(1)}) \gets \sum_{j=1}^N\|x_j - g_{(1)}\|_{\mathbf{R}^{D}}^2$\;
     \While{$\mathcal{E} \geq \epsilon^*$, $n < itr$}{
@@
         \end{array}\right)$\;
-        $\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) \gets \sum_{j=1}^N\|x_j - g_{(n + 1)}(\pi_{g_{(n + 1)}}(x_j))\|_{\mathbb{R}^D}^2$\;
+        $\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) \gets \sum_{j=1}^N\|x_j - g_{(n + 1)}(\pi_{g_{(n + 1)}}(x_j))\|_{\mathbf{R}^D}^2$\;
         $\mathcal{E} \gets |\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(g_{(n)})|/\mathcal{D}_{\hat{Q}_N}(g_{(n)})$\;
@@
 }
-$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbb{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \arg \min_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbb{R}^D}^2: l = 1, 2, \dots, L\}$\;
+$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbf{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \arg \min_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbf{R}^D}^2: l = 1, 2, \dots, L\}$\;
 $g^*(r) = g^{l^*}(\kappa r)$, where the analytic formula of $g^*$ is from $g_{(n + 1), l}(r) = \sum_{j=1}^N \eta_{4-d}(r - \pi_{g_{(n)}}(x_j)) + \sum_{k=1}^{d + 1} \alpha_{k, l} \times p_k(r)$, and $\gamma^* \gets \gamma_{l^*}$
@@
 	\begin{equation}\label{PMSEF}
-	\mathcal{K}_{\lambda,\mathbb{P}}(f,g)=\mathbb{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbb{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbb{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbb{R})}^2,
+	\mathcal{K}_{\lambda,\mathbf{P}}(f,g)=\mathbf{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbf{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbf{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbf{R})}^2,
 	\end{equation}
 	\begin{equation}\nonumber
-	\mathcal{K}_{\lambda,\mathbb{P}}(f)=\mathcal{K}_{\lambda,\mathbb{P}}(f,f),
+	\mathcal{K}_{\lambda,\mathbf{P}}(f)=\mathcal{K}_{\lambda,\mathbf{P}}(f,f),
 	\end{equation}
-	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbb{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbb{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$.
-A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbb{P}$) with the tuning parameter $\lambda$ if
+	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbf{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbf{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$.
+A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbf{P}$) with the tuning parameter $\lambda$ if
 	\begin{align}\label{def: PM}
-	f^*=\arg\min_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
+	f^*=\arg\min_{f\in\mathscr{F}(\mathbf{P})}\mathcal{K}_{\lambda,\mathbf{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbf{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbf{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
 	\end{align}

### 2023-02-14 paper/main.tex
introduction draft

@@
 Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al (2021), Poewe et al (2017)).
-Frequently, interest centers around changes in one or more substructures.
+Frequently, interest centers around longitudinal changes in one or more substructures.
 For example, it is common to observe atrophy in the hippocampus of those with Alzheimer's disease or cognitive impairment from other causes.
@@
 
-Ultimately, high levels of variability are present at the subcortical level regardless of the segmentation approach used, but may be particularly influential when using automated segmentation methods.
-To address these elevated noise levels, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
-This approach, which extends the principal manifold estimation algorithm introduced by Meng and Eloyan (2021) for a longitudinal setting, includes a regularization parameter to limit the effects of between time-point variability.
-
-\section{Introduction - Old}
-
+Ultimately, high levels of variability are present at the subcortical level regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods.
+In this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
-These methods have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical shapes.
-In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space.
+In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive than in a higher-dimensional space.
 For example, Yue et al.
 (2016) seeks to compute a parametric representation of the corpus callosum.
-A number of methods have been developed for manifold learning, including Isomap (Tenenbaum et al.
-2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003).
-Meng and Eloyan (2021) proposed a principal manifold framework that resolved several limitations of previously developed extensions of the principal curves approach introduced by Hastie and Stuetzle (1989).
+To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by Meng and Eloyan (2021).
 
-In many medical imaging situations, understanding how an anatomical structure changes over time could provide meaningful information into the function of that structure, or medical conditions that affect that function.
-In situations where the structure is better represented by a low-dimensional manifold, interest carries over to understanding how the underlying manifold changes longitudinally as well.
-
-While the methods given above provide a variety of approaches to retrieving a low-dimensional manifold from high-dimensional data at a single time point, they are ill-equipped to reach meaningful estimates across multiple time points.
+Previously introduced methods for manifold learning, such as Isomap (Tenenbaum, et al.
+2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003), are ill-equipped to reach meaningful estimates across multiple time points.
 This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension-reduction process.
@@
 We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+This smoothing process conducted between longitudinal observations mitigates the effects of
 

### 2023-02-15 paper/main.tex
add section 2 outline

--- 2023-02-09 paper/main.tex
@@
 
-Because the algorithm proposed in this article relies on the same definitions and notation as those introduced in Meng and Eloyan (2021), we provide it here.
-To generalize the principal curve framework proposed in Hastie and Stuetzle (1989) to intrinsic dimensions $d > 1$, Meng and Eloyan introduce the function space
+The notation below is used to build to a definition of principal manifolds [still need to formally write out]:
+\begin{itemize}
+  \item Observed data $\left\{x_i\right\}_{i=1}^I \subset \mathbb{R}^D$
+  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$
+  \item $C_{\infty}\left(\mathbb{R}^d \to \mathbb{R}^D\right) = \left\{f \in C(\mathbb{R}^d \to \mathbb{R}^D): \lim_{\|t\|_{\mathbb{R}^d} \to \infty} \|f(t)\|_{\mathbb{R}^D} = \infty\right\}$ 
+  \item $\nabla f$ denotes the gradient of $f$
+  \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$
+  \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d} |\frac{\partial^2f}{\partial t_i \partial t_j}(t)|^2\right)^{\frac{1}{2}}$
+  \item $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l = 1}^{D}\|\nabla^{\bigotimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D}\int_{\mathbb{R}^d}\sum_{i,j = 1}^{d}|\frac{\partial^2f_l}{\partial t_i \partial t_j}(t)|^2dt$ represents the total squared curvature of manifold $M_f^{d}$
+  \item $\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|X - f(\pi_{g}(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$, with $\mathcal{K_{\lambda, \mathbb{P}}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)$
+  \item A manifold $M_{f^*}^{d}$ determined by $f^*$ is a principal manifold for random $D$-vector $X$ with compact support $\text{supp}(\mathbb{P})$ with the tuning parameter $\lambda$ if \[%
+    f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
+  ,\]%
+    where \[%
+      \mathcal{F}(\mathbb{P}) = \left\{f \in C_{\infty} \cap \nabla^{-biot 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
+    .\]%
+  \item $\pi_{f_\lambda^{*}}(X)$ gives a $d$-dimensional parameterization of $X$
+\end{itemize}
+
+Given this definition, we now summarize the estimation method provided.
+This algorithm has three main steps: data reduction, fitting, and tuning.
+In the reduction step, the HDMDE algorithm is used to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by points $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$.
+The fitting stage of the algorithm finds estimated functions
 \[%
-  C_{\infty}(\mathbb{R}^d \to \mathbb{R}^D) = \left\{f \in C()\right\}
+  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
+,\]%
+where 
+\[%
+  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
+\]% 
+approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
+\[%
+  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2
 .\]%
+
+Meng and Eloyan (2021) show that the optimal function will be of spline form.
+Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in Meng and Eloyan (2021).
 
@@
 \section{Discussion}
+
+Contributions:
+
+Limitations:
+\begin{enumerate}
+  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
+  \item Partitioning approach is somewhat ad hoc and may not work in every situation.
+In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent.
+There is not a clear approach to automatic partitioning.
+  \item The process of joining the partitions may introduce error.
+  \item Algorithm initialization is sensitive to magnitude of change between time points.
+  \item Computation time.
+\end{enumerate}
+
+Future Work:
 

### 2023-02-15 code/functions/plot_pme.R
delete redundant code

@@ -127,121 +127,3 @@ plot_pme <- function(time_points, x.obs, sol, tnew, I, d, lambda) {
     print(plt)
   }
 }
-
-if (print_plots == TRUE) {
-  time_vals <- seq(
-    min(time_points),
-    max(time_points),
-    0.1
-  )
-
-  r_vals <- seq(
-    from = -10,
-    to = 10,
-    by = 1
-  )
-  r_list <- lapply(numeric(d_new - 1), function(x) r_vals)
-  r_mat <- as.matrix(expand.grid(r_list))
-  r_length <- dim(r_mat)[1]
-  pred_grid <- expand_grid(r_vals, r_mat) %>%
-    as.matrix()
-
-  f_pred <- matrix(nrow = nrow(pred_grid), ncol = ncol(X_new))
-  for (i in 1:nrow(pred_grid)) {
-    f_pred[i, ] <- f_new(unlist(as.vector(pred_grid[i, ])))
-  }
-
-  idx_inrange <- matrix(nrow = dim(f_pred)[1], ncol = dim(f_pred)[2])
-  for (dim_idx in 1:dim(f_pred)[2]) {
-    idx_range <- max(X_new[, dim_idx]) - min(X_new[, dim_idx])
-    idx_min <- min(X_new[, dim_idx]) - (0.2 * idx_range)
-    idx_max <- max(X_new[, dim_idx]) + (0.2 * idx_range)
-    idx_inrange[, dim_idx] <- (f_pred[, dim_idx] > idx_min) &
-      (f_pred[, dim_idx] < idx_max)
-  }
-
-  r_inrange <- rowSums(idx_inrange) == dim(f_pred)[2]
-  if (sum(r_inrange) == 0) {
-    r_inrange <- rowSums(idx_inrange) > 0
-  }
-  if (sum(r_inrange) == 0) {
-    r_min <- -10
-    r_max <- 10
-  } else {
-    r_min <- min(pred_grid[r_inrange, -1])
-    r_min <- ifelse(is.na(r_min) | is.infinite(r_min), -10, r_min)
-    r_max <- max(pred_grid[r_inrange, -1])
-    r_max <- ifelse(is.na(r_max) | is.infinite(r_max), 10, r_max)
-  }
-
-  r_vals <- seq(
-    r_min,
-    r_max,
-    by = (r_max - r_min) / 40
-  )
-  r_list <- lapply(numeric(d_new - 1), function(x) r_vals)
[…]

### 2023-02-15 paper/main.tex
relocate algorithm, rearrange package calls

@@
 
-\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
 %\RequirePackage[numbers]{natbib}
@@
 
-\usepackage{amsmath, amssymb, amscd, amsthm, amsfonts}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amscd}
+\usepackage{amsthm}
+\usepackage{amsfonts}
+\usepackage{mathrsfs}
 \usepackage{graphicx}
@@
 \renewcommand{\textheight}{240mm}
+
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
   \item Observed data $\left\{x_i\right\}_{i=1}^I \subset \mathbb{R}^D$
-  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$
+  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$ [may change this notation to allow for time points]
   \item $C_{\infty}\left(\mathbb{R}^d \to \mathbb{R}^D\right) = \left\{f \in C(\mathbb{R}^d \to \mathbb{R}^D): \lim_{\|t\|_{\mathbb{R}^d} \to \infty} \|f(t)\|_{\mathbb{R}^D} = \infty\right\}$ 
@@
   \item A manifold $M_{f^*}^{d}$ determined by $f^*$ is a principal manifold for random $D$-vector $X$ with compact support $\text{supp}(\mathbb{P})$ with the tuning parameter $\lambda$ if \[%
-    f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
+    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
   ,\]%
@@
 \[%
-  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
+  \hat{f}_\lambda = \argmin_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
 ,\]%
@@
 \]% 
-approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
+approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \argmin_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
 \[%
@@
 
-Let $\{x_{it}\}_{i=1, t=1}^{I, T}$ be data points in $D$-dimensional space.
+We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as Meng and Eloyan (2021). Let
 
@@
 \begin{enumerate}
-    \item Introduce principal surfaces / principal manifolds, PME algorithm, and the loss function it minimizes
     \item Show updated loss function for LPME algorithm
@@
 \end{enumerate}
-
-\section{Simulations}
-
-\section{Application}
-
-\section{Discussion}
-
-Contributions:
-
-Limitations:
-\begin{enumerate}
-  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
-  \item Partitioning approach is somewhat ad hoc and may not work in every situation.
-In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent.
-There is not a clear approach to automatic partitioning.
-  \item The process of joining the partitions may introduce error.
-  \item Algorithm initialization is sensitive to magnitude of change between time points.
-  \item Computation time.
-\end{enumerate}
-
-Future Work:
-
-\RestyleAlgo{ruled}
-\LinesNumbered
-
-\SetKwComment{Comment}{/* }{ */}
 
@@
 }
-$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbb{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \arg \min_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbb{R}^D}^2: l = 1, 2, \dots, L\}$\;
+$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbb{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \argmin_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbb{R}^D}^2: l = 1, 2, \dots, L\}$\;
 $g^*(r) = g^{l^*}(\kappa r)$, where the analytic formula of $g^*$ is from $g_{(n + 1), l}(r) = \sum_{j=1}^N \eta_{4-d}(r - \pi_{g_{(n)}}(x_j)) + \sum_{k=1}^{d + 1} \alpha_{k, l} \times p_k(r)$, and $\gamma^* \gets \gamma_{l^*}$
 \end{algorithm}
+
+\section{Simulations}
+
+\section{Application}
+
+\section{Discussion}
+
+Contributions:
+
+Limitations:
+\begin{enumerate}
+  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
+  \item Partitioning approach is somewhat ad hoc and may not work in every situation.
+In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent.
+There is not a clear approach to automatic partitioning.
+  \item The process of joining the partitions may introduce error.
+  \item Algorithm initialization is sensitive to magnitude of change between time points.
+  \item Computation time.
+\end{enumerate}
+
+Future Work:
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
+\SetKwComment{Comment}{/* }{ */}
 
@@
 	\begin{align}\label{def: PM}
-	f^*=\arg\min_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
+	f^*=\argmin_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
 	\end{align}

### 2023-02-16 paper/main.tex
restore updated introduction

@@
 
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al.
+2021, Poewe et al.
+2017).
+Frequently, interest centers around longitudinal changes in one or more neurological substructures.
+For example, it is common to observe atrophy the hippocampus of those with Alzheimer's disease or cognitive impairment related to other causes.
+Image segmentation allows for the extraction of subcortical structures from images of the complete brain for further analysis.
+
+Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach and is the gold standard.
+However, this approach is highly time and resource intensive.
+In studies analyzing a large number of images, this approach may impose prohibitive costs.
+Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (Source).
+Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns [sources].
+While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
+
+To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
+In a larger comparison of the accuracy of FIRST and FreeSurfer, Mulder et al.
+(2014) found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
+If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation.
+Mulder et al.
+(2014) also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].
+
+Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods.
+Mitigating the extent of this variability is a priority from a statistical perspective.
+To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
-These methods have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical shapes.
-In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space.
+In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive than working with the structure in its original high-dimensional space.
 For example, Yue et al.
 (2016) seeks to compute a parametric representation of the corpus callosum.
-A number of methods have been developed for manifold learning, including Isomap (Tenenbaum et al.
-2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003).
-Meng and Eloyan (2021) proposed a principal manifold framework that resolved several limitations of previously developed extensions of the principal curves approach introduced by Hastie and Stuetzle (1989).
+To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by Meng and Eloyan (2021).
 
-In many medical imaging situations, understanding how an anatomical structure changes over time could provide meaningful information into the function of that structure, or medical conditions that affect that function.
-In situations where the structure is better represented by a low-dimensional manifold, interest carries over to understanding how the underlying manifold changes longitudinally as well.
-
-While the methods given above provide a variety of approaches to retrieving a low-dimensional manifold from high-dimensional data at a single time point, they are ill-equipped to reach meaningful estimates across multiple time points.
-This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension-reduction process.
-This treatment of the time dimension causes time to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
+Previously introduced methods for manifold learning, such as Isomap (Tenenbaum et al.
+2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003), are ill-equipped to reach meaningful estimates across multiple time points.
+This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension reduction process.
+This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
 Here, we propose an alternative approach to estimating manifolds over time.
@@
 Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension.
-Specifically, we use the method proposed in Meng and Eloyan (2021) to estimate an appropriate principal manifold at each given time point.
+Specifically, we use the method proposed in Meng and Eloyan (2021) to estimate the appropriate principal manifold at each given time point.
 We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+The regularization involved in this process mitigates the effects of variability between study visits.
 
-Some work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by Kinson et al.
+Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by Kinson et al.
 (2019).
-To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this paper is novel.
+To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this article is novel.
 
-The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold estimation method proposed in Meng and Eloyan (2021).
+The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold framework proposed in Meng and Eloyan (2021).
 Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this.
@@
 In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease.
-The article finishes with a discussion of the method in Section 6.
+The article concludes with a discussion of the method's contributions in Section 6.
 

### 2023-02-23 paper/main.tex
Updates from Overleaf
algorithm updates

@@
 
+\SetKwComment{Comment}{/* }{ */}
+
+	\begin{equation}\label{PMSEF}
+	\mathcal{K}_{\lambda,\mathbb{P}}(f,g)=\mathbb{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbb{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbb{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbb{R})}^2,
+	\end{equation}
+	\begin{equation}%\nonumber
+	\mathcal{K}_{\lambda,\mathbb{P}}(f)=\mathcal{K}_{\lambda,\mathbb{P}}(f,f),
+	\end{equation}
+	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbb{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbb{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$.
+A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbb{P}$) with the tuning parameter $\lambda$ if
+	\begin{align}\label{def: PM}
+	f^*=\argmin_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
+	\end{align}
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
 \begin{algorithm}
-\caption{Longitudinal Principal Manifold Estimation}\label{alg:two}
-\KwData{Data points $\{y_{i, t}\}_{i, t}^{I, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\{\lambda_k\}_{k=1}^K$, $\{\gamma_l\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
-\KwResult{Analytic formula of $g^*: \mathbb{R}^{d+1} \to \mathbb{R}^{D+1}$, optimal tuning parameter $\gamma^*$.}
-\For{$t = 1, 2, ... T$}{
-    Apply PME algorithm with input $(\{y_{i, t}\}_{i=1}^I, N_0, \alpha, \epsilon, \epsilon^*, \{\lambda_k\}_{k=1}^K)$ and obtain $\{f_t^*\}_{t=1}^T$ and $\{\lambda_t^*\}_{t=1}^T$\;
-    Use $\{f_t^*\}_{t=1}^T$ to map a grid of $d$-dimensional parameters $\{r_j\}_{j=1}^N$ to $\{x_{j, t}\}_{j=1}^N$ in $\mathbb{R}^D$\;
+\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
+\KwData{Data points $\left\{y_{i, t}\right\}_{i, t}^{I_t, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
+\KwResult{Analytic formula of $\hat{g}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D}$, optimal tuning parameter $\gamma^*$.}
+\For{$t = 1, 2, \dots, T$} {
+  Apply HDMDE algorithm with input $\left(\left\{y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, N_t, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, N_t, t}\right\}_{j = 1}^{N_t}$\;
 }
-Apply ISOMAP to parameterize the newly generated collection $\{x_{j, t}\}_{j, t}^{N, T}$ by the $d$-dimensional parameters $\{r_j\}_{j=1}^N$\;
-Use parameters $\{r_j\}_{j=1}^N$ and time values $\{t_j\}_{j=1}^N$ to create new
-$\pi_{g_{(1)}}(x_{j}) \gets r_j$ for $j=1, ..., N$\;
-\For{$l = 1, 2, ..., L$}{
-    $\gamma \gets \gamma_l$.\;
-    Obtain $g_{(1)}$ by solving 
-    $\left(\begin{array}{ccc} 
+
+Parameterization: Apply ISOMAP to parameterize $\left\{\mu_{j, N_t, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
+Formally set $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets r_j$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
+
+\For{t = 1, 2, \dots, T} {
+  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^N_t$ and obtain $f_t^*$ and $\lambda_t^*$\;
+  $\left\{r_{j, t}\right\} \gets \pi_{f_t^*}(\mu_{j, N_t, t})$ for $j = 1, \dots, N_t$ and $x_{j, t} \gets f_t^*(r_{j, t})$ for $j = 1, \dots, N_t$\;
+  Add column vector $\mathbf{t}$ as first column of $\mathbf{r}$\;
+}
+
+Let $N = \sum_{t=1}^{T}N_t$, $\mathbf{x}$ be a $N \times D$ matrix of stacked values of $\left\{x_{j, t}\right\}_{t = 1}^T$, and $\mathbf{r}$ be a $N \times d + 1$ matrix of stacked values of $\left\{r_{j, t}\right\}_{t = 1}^T$ \;
+
+Set $\pi_{g(0)}(x_i) \gets r_i$ for $i = 1, 2, \dots, N$\;
+
+\For{l = 1, 2, \dots, L}{
+  $\gamma \gets \gamma_l$\;
+  Obtain $g_{(1)}$ by solving $\left(\begin{array}{ccc}
       2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
-    2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
-    \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
-    s_l \\
-    \alpha_l \\
-    m_l 
-    \end{array}\right) = \left(\begin{array}{c}
-    2\mathbf{E}\mathbf{W}\mathbf{x} \\
-    2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
-    \mathbf{0} 
-    \end{array}\right)$\;
-    $\mathcal{E} \gets 2\epsilon^*$\;
-    $n \gets 1$\;
-    $\mathcal{D}_{\hat{Q}, N}(g_{(1)}) \gets \sum_{j=1}^N\|x_j - g_{(1)}\|_{\mathbb{R}^{D}}^2$\;
-    \While{$\mathcal{E} \geq \epsilon^*$, $n < itr$}{
-        Compute $g_{(n + 1)}$ by solving $\left(\begin{array}{ccc} 
-        2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
-        2\mathbf{T}^T\mathbf{W}\mathbf{E} & 2\mathbf{T}^T\mathbf{W}\mathbf{T} & \mathbf{0} \\
-        \mathbf{T}^T & \mathbf{0} & \mathbf{0} \end{array}\right)\left(\begin{array}{c} 
-        s_l \\
-        \alpha_l \\
-        m_l 
-        \end{array}\right) = \left(\begin{array}{c}
-        2\mathbf{E}\mathbf{W}\mathbf{x} \\
-        2\mathbf{T}^T\mathbf{W}\mathbf{x} \\
-        \mathbf{0} 
-        \end{array}\right)$\;
-        $\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) \gets \sum_{j=1}^N\|x_j - g_{(n + 1)}(\pi_{g_{(n + 1)}}(x_j))\|_{\mathbb{R}^D}^2$\;
-        $\mathcal{E} \gets |\mathcal{D}_{\hat{Q}_N}(g_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(g_{(n)})|/\mathcal{D}_{\hat{Q}_N}(g_{(n)})$\;
-        $n \gets n + 1$\;
+      2\mathbf{T}^{T}\mathbf{W}\mathbf{E} & 2\mathbf{T}^{T}\mathbf{W}\mathbf{T} & \mathbf{0} \\
+      \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
+  \end{array}\right) = $\;
 }
-    $\hat{g}_l \gets g_{(n)}$\;
-}
-$\kappa \gets \text{max}\{\|\pi_{\hat{g}_{l^*}}(x_j)\|_{\mathbb{R}^d}: j = 1, 2, \dots, I\}$, where $l^* = \argmin_l\{\frac{1}{N}\sum_{j=1}^N\|x_j - g_{(l)}(\pi_{\hat{g}_{(l)}}(x_j)\|_{\mathbb{R}^D}^2: l = 1, 2, \dots, L\}$\;
-$g^*(r) = g^{l^*}(\kappa r)$, where the analytic formula of $g^*$ is from $g_{(n + 1), l}(r) = \sum_{j=1}^N \eta_{4-d}(r - \pi_{g_{(n)}}(x_j)) + \sum_{k=1}^{d + 1} \alpha_{k, l} \times p_k(r)$, and $\gamma^* \gets \gamma_{l^*}$
 \end{algorithm}
@@
 Future Work:
-
-\RestyleAlgo{ruled}
-\LinesNumbered
-
-\SetKwComment{Comment}{/* }{ */}
-
-	\begin{equation}\label{PMSEF}
-	\mathcal{K}_{\lambda,\mathbb{P}}(f,g)=\mathbb{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbb{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbb{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbb{R})}^2,
-	\end{equation}
-	\begin{equation}\nonumber
-	\mathcal{K}_{\lambda,\mathbb{P}}(f)=\mathcal{K}_{\lambda,\mathbb{P}}(f,f),
-	\end{equation}
-	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbb{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbb{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$.
-A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbb{P}$) with the tuning parameter $\lambda$ if
-	\begin{align}\label{def: PM}
-	f^*=\argmin_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
-	\end{align}
 

### 2023-02-23 code/functions/plot_pme.R
add function for plotting intermediate results

@@ -1,0 +1,124 @@
+plot_pme <- function(x.obs, sol, tnew, I, d, lambda) {
+  D <- dim(x.obs)[2]
+  r_vals <- seq(
+    from = -10,
+    to = 10,
+    by = 1
+  )
+  r_list <- lapply(numeric(d - 1), function(x) r_vals)
+  r_mat <- as.matrix(expand.grid(r_list))
+  if (nrow(r_mat) > 0) {
+    pred_grid <- expand_grid(r_vals, r_mat) %>%
+      as.matrix()
+  } else {
+    pred_grid <- matrix(r_vals)
+  }
+
+  f_pred <- map(
+    1:nrow(pred_grid),
+    ~ fNew(
+      unlist(as.vector(pred_grid[.x, ])),
+      sol,
+      tnew,
+      I,
+      d,
+      lambda
+    ) %>%
+      as.vector()
+  ) %>%
+    unlist() %>%
+    matrix(nrow = nrow(pred_grid), byrow = TRUE)
+
+  idx_inrange <- matrix(nrow = dim(f_pred)[1], ncol = dim(f_pred)[2])
+  for (dim_idx in 1:dim(f_pred)[2]) {
+    idx_range <- max(x.obs[, dim_idx]) - min(x.obs[, dim_idx])
+    idx_min <- min(x.obs[, dim_idx]) - (0.2 * idx_range)
+    idx_max <- max(x.obs[, dim_idx]) + (0.2 * idx_range)
+    idx_inrange[, dim_idx] <- (f_pred[, dim_idx] > idx_min) &
+      (f_pred[, dim_idx] < idx_max)
+  }
+
+  r_inrange <- rowSums(idx_inrange) == dim(f_pred)[2]
+  if (sum(r_inrange) == 0) {
+    r_inrange <- rowSums(idx_inrange) > 0
+  }
+  if (sum(r_inrange) == 0) {
+    r_min <- -10
+    r_max <- 10
+  } else if (dim(pred_grid)[2] > 1) {
+    r_min <- min(pred_grid[r_inrange, -1])
+    r_min <- ifelse(is.na(r_min) | is.infinite(r_min), -10, r_min)
+    r_max <- max(pred_grid[r_inrange, -1])
+    r_max <- ifelse(is.na(r_max) | is.infinite(r_max), 10, r_max)
+  } else {
+    r_min <- min(pred_grid[r_inrange, ])
+    r_min <- ifelse(is.na(r_min) | is.infinite(r_min), -10, r_min)
[…]

### 2023-02-24 paper/main.tex

@@
 
+\usepackage{setspace}
 \usepackage{amsmath}
@@
 \maketitle
+
+\doublespacing
 
@@
   \item Observed data $\left\{x_i\right\}_{i=1}^I \subset \mathbb{R}^D$
-  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$ [may change this notation to allow for time points]
+  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$
   \item $C_{\infty}\left(\mathbb{R}^d \to \mathbb{R}^D\right) = \left\{f \in C(\mathbb{R}^d \to \mathbb{R}^D): \lim_{\|t\|_{\mathbb{R}^d} \to \infty} \|f(t)\|_{\mathbb{R}^D} = \infty\right\}$ 
@@
   \item A manifold $M_{f^*}^{d}$ determined by $f^*$ is a principal manifold for random $D$-vector $X$ with compact support $\text{supp}(\mathbb{P})$ with the tuning parameter $\lambda$ if \[%
-    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
+    f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
   ,\]%
@@
 \[%
-  \hat{f}_\lambda = \argmin_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
+  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
 ,\]%
@@
 \]% 
-approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \argmin_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
+approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
 \[%
@@
 \end{enumerate}
+
+\section{Simulations}
+
+\section{Application}
+
+\section{Discussion}
+
+Contributions:
+
+Limitations:
+\begin{enumerate}
+  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
+  \item Partitioning approach is somewhat ad hoc and may not work in every situation.
+In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent.
+There is not a clear approach to automatic partitioning.
+  \item The process of joining the partitions may introduce error.
+  \item Algorithm initialization is sensitive to magnitude of change between time points.
+  \item Computation time.
+\end{enumerate}
+
+Future Work:
+
+\RestyleAlgo{ruled}
+\LinesNumbered
 

### 2023-02-24 paper/main.tex
complete algorithm, update citation formatting

@@
 %\RequirePackage[numbers]{natbib}
-\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
+%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
 \RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
 \RequirePackage{graphicx}%% uncomment this for including figures
+
+\usepackage[
+backend=biber,
+style=authoryear
+]{biblatex}
+
+\addbibresource{references.bib}
 
@@
 
-\usepackage{natbib}
+%\usepackage{natbib}
+%\bibliographystyle{abbrvnat}
+%\setcitestyle{authoryear,open={(},close={)}}
 \usepackage{authblk}
@@
 \usepackage[english]{babel}
-\bibliographystyle{abbrvnat}
-\setcitestyle{authoryear,open={(},close={)}}
 
@@
 
-Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al.
-2021, Poewe et al.
-2017).
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}).
 Frequently, interest centers around longitudinal changes in one or more neurological substructures.
@@
 In studies analyzing a large number of images, this approach may impose prohibitive costs.
-Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (Source).
-Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns [sources].
+Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
+Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}).
 While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
@@
 To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
-In a larger comparison of the accuracy of FIRST and FreeSurfer, Mulder et al.
-(2014) found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+In a larger comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
 After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
 If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation.
-Mulder et al.
-(2014) also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].
+\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].
 
@@
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive than working with the structure in its original high-dimensional space.
-For example, Yue et al.
-(2016) seeks to compute a parametric representation of the corpus callosum.
-To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by Meng and Eloyan (2021).
+For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum.
+To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.
 
-Previously introduced methods for manifold learning, such as Isomap (Tenenbaum et al.
-2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003), are ill-equipped to reach meaningful estimates across multiple time points.
+Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003a}), are ill-equipped to reach meaningful estimates across multiple time points.
 This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension reduction process.
@@
 Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension.
-Specifically, we use the method proposed in Meng and Eloyan (2021) to estimate the appropriate principal manifold at each given time point.
+Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to estimate the appropriate principal manifold at each given time point.
 We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
@@
 
-Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by Kinson et al.
-(2019).
+Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
 To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this article is novel.
 
-The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold framework proposed in Meng and Eloyan (2021).
+The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}.
 Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this.
@@
 \section{Principal Manifolds and Estimation Algorithm}
+
+The framework for principal manifolds originated with the concept of principal curves and surfaces (\cite{hastiePrincipalCurves1989}), which are essentially curves or surfaces that pass through the middle of the data.
+Principal manifolds extend this concept to higher dimensions.
 
@@
 
-\section{Simulations}
-
-\section{Application}
-
-\section{Discussion}
-
-Contributions:
-
-Limitations:
-\begin{enumerate}
-  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
-  \item Partitioning approach is somewhat ad hoc and may not work in every situation.
-In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent.
-There is not a clear approach to automatic partitioning.
-  \item The process of joining the partitions may introduce error.
-  \item Algorithm initialization is sensitive to magnitude of change between time points.
-  \item Computation time.
-\end{enumerate}
-
-Future Work:
-
 \RestyleAlgo{ruled}
@@
       \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
-  \end{array}\right) = $\;
+    \end{array}\right)\left(\begin{array}{c}
+    s_l \\
+    \alpha_l \\
+    m_l
+    \end{array}\right) = \left(\begin{array}{c}
+    2\mathbf{E}\mathbf{W}\mu_l \\
+    2\mathbf{T}^{T}\mathbf{W}\mu_l \\
+    \mathbf{0}
+  \end{array}\right)$\;
+  $\mathcal{\epsilon} \gets 2 \times \epsilon^*$, $n \gets 1$, $\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) \gets \frac{1}{N}\sum_{j = 1}^{N}\|x_{j} - f_{(1)}(\pi_{f_{(1)}}(x_j))\|_{\mathbb{R}^{D}}^2$ \;
+  \While{$\epsilon > \epsilon^*$ and $n \leq itr$} {
+    Compute $f_{(n + 1)})$ from $f_{(n)}$ by solving (15) \;
+    $\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) \gets \frac{1}{N}\sum_{j=1}^{N}\|x_j - f_{(n + 1)}(\pi_{f_{(n + 1)}}(x_j))\|_{\mathbb{R}^D}^2$ \;
+    $\epsilon \gets |\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(f_{(n)})| / \mathcal{D}_{\hat{Q}_N}(f_{(n)})$ and $n \gets n + 1$\;
   }
+  $\hat{f}_l \gets f_{(n)}$\;
+}
+$l^* \gets \argmin_l\left\{\frac{1}{T}\sum_{t = 1}^{T}\frac{1}{I_t}\sum_{i = 1}^{I_t}\|y_{i, t} - \hat{f}_l(\pi_{f_l}(y_{i, t}))\|_{\mathbb{R}^D}^2: \ l = 1, 2, \dots, L \right\}$ \;
+$\kappa \gets \max\left\{\|\pi_{\hat{f}_{g^*}}(y_i, t)\|_{\mathbb{R}^{d}}: \ i = 1, 2, \dots, I_t; j = 1, 2, \dots, T \right\}$ \;
+$f^*(r) = \hat{f}_{l^*}(r)$, where the analytic formula of $f^*$ is from (14), and $\gamma^* \gets \gamma_l^*$.
 \end{algorithm}
@@
 \bibliographystyle{plain}
-\bibliography{references}
+%\bibliography{references}
+\printbibliography
 

### 2023-02-24 paper/main.tex
fix compilation errors

@@
 
+\usepackage{csquotes}
 \usepackage[english]{babel}
@@
 \For{t = 1, 2, \dots, T} {
-  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^N_t$ and obtain $f_t^*$ and $\lambda_t^*$\;
+  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t^*$ and $\lambda_t^*$\;
   $\left\{r_{j, t}\right\} \gets \pi_{f_t^*}(\mu_{j, N_t, t})$ for $j = 1, \dots, N_t$ and $x_{j, t} \gets f_t^*(r_{j, t})$ for $j = 1, \dots, N_t$\;
@@
 \nocite{*}
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 %\bibliography{references}

### 2023-03-02 code/lpme.R
update max.comp value

@@ -1,4 +1,4 @@
-lpme <- function(df, d, tuning.para.seq = exp(-25:5), alpha = 0.05, max.comp = 100, epsilon = 0.05, max.iter = 100, print.MSDs = TRUE, print_plots = TRUE, SSD_ratio_threshold = 100, init = "full") {
+lpme <- function(df, d, tuning.para.seq = exp(-25:5), alpha = 0.05, max.comp = 500, epsilon = 0.05, max.iter = 100, print.MSDs = TRUE, print_plots = TRUE, SSD_ratio_threshold = 100, init = "full") {
   # df is an N x (D + 1) matrix, with the first column corresponding
   # to the time point at which each observation was collected
   # this matrix should include the observations from all time points

### 2023-03-02 code/pme.R
allow adjustable SSD.ratio threshold

@@ -83,7 +83,7 @@ source("code/functions/plot_pme.R")
 ############ Section 3, Principal Manifold Estimation ######################
 ############################################################################
 
-pme <- function(x.obs, d, initialization = NULL, N0=20*D, tuning.para.seq=exp((-15:5)), alpha=0.05, max.comp=100, epsilon=0.05, max.iter=100, print.MSDs=TRUE, print_plots = FALSE) {
+pme <- function(x.obs, d, initialization = NULL, N0=20*D, tuning.para.seq=exp((-15:5)), alpha=0.05, max.comp=100, epsilon=0.05, max.iter=100, print.MSDs=TRUE, SSD_ratio_threshold = 10, print_plots = FALSE) {
 
   # "x.obs" is the data set of interest.
   #         There are n observations, and each observation is a D-dimensional point.
@@ -254,7 +254,7 @@ pme <- function(x.obs, d, initialization = NULL, N0=20*D, tuning.para.seq=exp((-
 
     while (
       (SSD.ratio > epsilon) &
-      (SSD.ratio <= 5) &
+      (SSD.ratio <= SSD_ratio_threshold) &
       (count <= (max.iter - 1))
     ) {
 

### 2023-03-02 paper/main.tex
update paper draft

@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
 
@@
 
-The framework for principal manifolds originated with the concept of principal curves and surfaces (\cite{hastiePrincipalCurves1989}), which are essentially curves or surfaces that pass through the middle of the data.
-Principal manifolds extend this concept to higher dimensions.
+The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data.
+\cite{hastiePrincipalCurves1989} defines principal curves as follows:
+
+\begin{definition}
+  \label{def:principal_curves}
+
+  (Part I) Let $I \in \mathbb{R}^{1}$ be a closed and possibly infinite interval. Suppose a map $f:I \to \mathbb{R}^{D}$ satisfies the conditions:
+  \begin{enumerate}
+    \label{enum:hs_conditions}
+
+    \item $f \in \mathcal{C}^{\infty}(I \to \mathbb{R}^{D})$
+    \item $\|f'(r)\|_{\mathbb{R}^{D}} = 1$ for all $r \in I$
+    \item $f$ does not self intersect, meaning $r_1 \neq r_2$ implies that $f(r_1) \neq f(r_2)$
+    \item $\int_{\left\{r: f(r) \in B\right\}}dt < \infty$ for any finite ball $B$ in $\mathbb{R}^{D}$
+  \end{enumerate}
+  Then the projection index with respect to $f$ is defined as \[%
+    \pi_f(x) = \sup\left\{r \in I: \|x - f(r)\|_{\mathbb{R}^{D}} = \inf_{r' \in I}\|x - f(r')\|_{\mathbb{R}^{D}} \right\}, \ \text{for all} \ x \in \mathbb{R}^{D}
+  .\]%
+
+  (Part II) Suppose $X$ is a continuous random $D$-vector with finite second moments.
+Principal curves of $X$ are all maps $f: I \to \mathbb{R}^{D}$ satisfying the conditions above and the Self-consistency defined as \[%
+    E(X | \pi_f(X) = t) = f(r)
+  .\]%
+\end{definition}
+
+Principal manifolds seek to extend this concept to higher dimensions.
+\cite{smolaRegularizedPrincipalManifolds2001} proposed a regularized principal manifold framework, where principal manifolds minimize \[%
+  \argmin_{f \in \mathcal{F}}\left\{E\|X - f\left(\pi_f(X)\right)\|_{\mathbb{R}^D}^2 + \lambda\|Pf\|_{\mathcal{H}}^2\right\}
+\]%
+with $\mathcal{F}$ being a collection of functions and $P$ being an operator that maps $f$ into inner product space $\mathcal{H}$.
+\cite{mengPrincipalManifoldEstimation2021} build on this framework to provide a more complete definition of principal manifolds, and proposes an algorithm to estimate these principal manifolds.
 
@@
   \item Why would we see the SSD increase substantially in late optimization steps?
-  \item Are there cases where the SSD and MSD differ substantially? Why would that be the case?
-  \item Are there any techniques that help with large amounts of noise in the data?
-Data reduction should help with outliers, but likely not consistently high noise levels.
-Early indications that possibly allowing for higher tuning parameters would help?
-  \item Why does PME algorithm struggle on Swiss Roll cross-section?
-It seems to work well up until spiral reaches second layer.
-The manifold should never intersect, but from the angle from the origin does overlap.
   \item Generally, what are the limitations on the estimated manifolds intersecting? How close can it come to intersecting without showing strange behavior?

### 2023-03-02 long_pme_simulations.R
update spherical simulation

@@ -1,5 +1,6 @@
 library(tidyverse)
 library(plotly)
+library(pracma)
 library(profvis)
 
 source("code/lpme.R")
@@ -1244,17 +1245,22 @@ time_vals <- 0:5
 set.seed(100)
 df_list <- lapply(
   time_vals,
-  sim_D4d2_case1,
+  sim_D5d2_case2,
   vertical_multiplier = 0.1,
   horizontal_multiplier = 0.1,
   depth_multiplier = 0.1,
   noise = 0.15,
   time_noise = 0.1,
-  shape_noise = 0.0001,
-  curvature = 1
+  shape_noise = 0.0001
+  # curvature = 1
 )
 
-data_points <- reduce(df_list, rbind)
+data_points <- reduce(df_list, rbind)[, -(5:6)]
+data_points_sph <- cart2sph(data_points[, 2:4])
+data_points <- cbind(data_points, data_points_sph)
+
+test_df <- data_points[data_points[, 1] == 0, -1]
+pme_result <- pme(test_df, 2, print_plots = TRUE)
 
 sim_result <- lpme(data_points, 2)
 

### 2023-03-15 paper/main.tex
add to pme section draft

@@
 with $\mathcal{F}$ being a collection of functions and $P$ being an operator that maps $f$ into inner product space $\mathcal{H}$.
+If $P = \frac{d}{dr}$, this framework minimizes the mean squared distance (MSD), $\mathcal{D}_X(f) = E\|X - f(\pi_f(X))\|_{\mathbb{R}^{D}}^2$, while penalizing the non-smoothness of the curve $f$.
 \cite{mengPrincipalManifoldEstimation2021} build on this framework to provide a more complete definition of principal manifolds, and proposes an algorithm to estimate these principal manifolds.
 
-The notation below is used to build to a definition of principal manifolds [still need to formally write out]:
+To ensure that the upcoming definitions are well behaved, we introduce the following function spaces:
 \begin{itemize}
-  \item Observed data $\left\{x_i\right\}_{i=1}^I \subset \mathbb{R}^D$
-  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$
-  \item $C_{\infty}\left(\mathbb{R}^d \to \mathbb{R}^D\right) = \left\{f \in C(\mathbb{R}^d \to \mathbb{R}^D): \lim_{\|t\|_{\mathbb{R}^d} \to \infty} \|f(t)\|_{\mathbb{R}^D} = \infty\right\}$ 
-  \item $\nabla f$ denotes the gradient of $f$
-  \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$
-  \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d} |\frac{\partial^2f}{\partial t_i \partial t_j}(t)|^2\right)^{\frac{1}{2}}$
-  \item $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l = 1}^{D}\|\nabla^{\bigotimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D}\int_{\mathbb{R}^d}\sum_{i,j = 1}^{d}|\frac{\partial^2f_l}{\partial t_i \partial t_j}(t)|^2dt$ represents the total squared curvature of manifold $M_f^{d}$
-  \item $\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|X - f(\pi_{g}(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$, with $\mathcal{K_{\lambda, \mathbb{P}}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)$
-  \item A manifold $M_{f^*}^{d}$ determined by $f^*$ is a principal manifold for random $D$-vector $X$ with compact support $\text{supp}(\mathbb{P})$ with the tuning parameter $\lambda$ if \[%
-    f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
-  ,\]%
-    where \[%
-      \mathcal{F}(\mathbb{P}) = \left\{f \in C_{\infty} \cap \nabla^{-biot 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
-    .\]%
-  \item $\pi_{f_\lambda^{*}}(X)$ gives a $d$-dimensional parameterization of $X$
+  \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$
+  \item $\nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
+  \item $H^2(\mathbb{R}^{d}) = \left\{f \in L^2(\mathbb{R}^{d}): \|\nabla f\|_{\mathbb{R}^{d}}, \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
+  \item $\nabla^{-\bigotimes 2}L^2(\Omega) = \left\{f|_{\Omega}: f \in \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d})\right\}$
+  \item $H^2(\Omega) = \left\{f|_{\Omega}: f \in H^2(\mathbb{R}^{d})\right\}$
 \end{itemize}
 
-Given this definition, we now summarize the estimation method provided.
+In these function spaces, we have the notation:
+\begin{itemize}
+  \item $\Omega$ is any open subset of $\mathbb{R}^{d}$ with a smooth boundary
+  \item $f|_{\Omega}$ denotes the restriction of $f$ to $\Omega$
+  \item $\nabla f$ denotes the gradient vector of $f$
+  \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$
+  \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\bigotimes 2}f(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(r)\right|^2\right)^{\frac{1}{2}}$
+  \item $\|\nabla f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i=1}^{d}\left|\frac{\partial f}{\partial r_i}(r)\right|^2dr $
+  \item $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial r_i \partial r_j}(r)\right|^2dr $
+\end{itemize}
+
+To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative.
+To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
+\[%
+  \|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr
+,\]%
+referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
+\begin{definition}
+  \label{def:principal_manifolds}
+  Let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments.
+Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
+  \[%
+    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
+  .\]%
+
+  Then a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $X$ with the tuning parameter $\lambda$ if 
+  \[%
+    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
+  .\]%
+\end{definition}
+
+In this setting, we can see that $\pi_{f_{\lambda}^{*}}(X)$ maps the $D$-vector $X$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
+
+To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm.
 This algorithm has three main steps: data reduction, fitting, and tuning.
-In the reduction step, the HDMDE algorithm is used to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by points $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$.
-The fitting stage of the algorithm finds estimated functions
+In the reduction step, the HDMDE algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$.
+The number of clusters is chosen using an iterative hypothesis testing approach.
+For each given value of $\lambda$, the fitting stage of the algorithm uses this reduced dataset to find estimated function
 \[%
@@
 \]% 
-approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
+approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses the optimal value of $\lambda$, $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
 \[%
   \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2
-.\]%
+\]%
+is an empirical estimate of the MSD using the full dataset.
 
-Meng and Eloyan (2021) show that the optimal function will be of spline form.
-Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in Meng and Eloyan (2021).
+\cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of spline form, meaning that when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$.
+Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in \cite{mengPrincipalManifoldEstimation2021}.
 
@@
 
-We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as Meng and Eloyan (2021). Let
+We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}. Let
 

### 2023-03-15 paper/main.tex
change otimes size

@@
   \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$
-  \item $\nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
-  \item $H^2(\mathbb{R}^{d}) = \left\{f \in L^2(\mathbb{R}^{d}): \|\nabla f\|_{\mathbb{R}^{d}}, \|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
-  \item $\nabla^{-\bigotimes 2}L^2(\Omega) = \left\{f|_{\Omega}: f \in \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d})\right\}$
+  \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
+  \item $H^2(\mathbb{R}^{d}) = \left\{f \in L^2(\mathbb{R}^{d}): \|\nabla f\|_{\mathbb{R}^{d}}, \|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
+  \item $\nabla^{-\otimes 2}L^2(\Omega) = \left\{f|_{\Omega}: f \in \nabla^{-\otimes 2}L^2(\mathbb{R}^{d})\right\}$
   \item $H^2(\Omega) = \left\{f|_{\Omega}: f \in H^2(\mathbb{R}^{d})\right\}$
@@
   \item $\nabla f$ denotes the gradient vector of $f$
-  \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$
-  \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\bigotimes 2}f(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(r)\right|^2\right)^{\frac{1}{2}}$
+  \item $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$
+  \item $\|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2}f(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(r)\right|^2\right)^{\frac{1}{2}}$
   \item $\|\nabla f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i=1}^{d}\left|\frac{\partial f}{\partial r_i}(r)\right|^2dr $
-  \item $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial r_i \partial r_j}(r)\right|^2dr $
+  \item $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial r_i \partial r_j}(r)\right|^2dr $
 \end{itemize}
@@
 \[%
-  \|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr
+  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr
 ,\]%
@@
   Let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments.
-Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
+Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
   \[%
-    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
+    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
   .\]%
@@
   \[%
-    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\bigotimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
+    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
   .\]%
@@
 \[%
-  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
+  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
 \]% 

### 2023-04-16 paper/main.tex
add lpme section intro draft

@@
 \title{Longitudinal Manifold Estimation}
-\author[1]{Robert Zielinski and Ani Eloyan}
+\author[1]{Robert Zielinski, Kun Meng, and Ani Eloyan}
 
@@
 
-We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}. Let
+We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}.
+Let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
+As in the setting described above, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
+However, in this updated case, the function determining the manifold may change over time.
+Note that we assume changes over time to be smooth.
+Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as
+\[%
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \sum_{t=1}^{T} \lambda_t \|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 + \gamma\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^d)}^2
+.\]%
 
-Outline:
-\begin{enumerate}
-    \item Show updated loss function for LPME algorithm
-    \item LPME algorithm
-\end{enumerate}
+To solve this problem, we propose an extension of the method developed in \cite{mengPrincipalManifoldEstimation2021}.
+The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space.
+In the LPME algorithm, we use the coefficients for spline functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time.
+This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set.
+Details of this approach are described below.
+
+\textit{Initialization}:
+
+\textit{Fitting}:
+
+\textit{Tuning}:
 
@@
 \SetKwComment{Comment}{/* }{ */}
-
-	\begin{equation}\label{PMSEF}
-	\mathcal{K}_{\lambda,\mathbb{P}}(f,g)=\mathbb{E}\left\Vert X_t - f\left(\pi_g(X_t)\right)\right\Vert^2_{\mathbb{R}^D} + \lambda \sum_{t = 1}^T \left\Vert\nabla^{\otimes 2} f(t)\right\Vert_{L^2(\mathbb{R}^d)}^2 + \gamma \sum_{X} \left\Vert\nabla^{\otimes 2} f(X_t)\right\Vert_{L^2(\mathbb{R})}^2,
-	\end{equation}
-	\begin{equation}%\nonumber
-	\mathcal{K}_{\lambda,\mathbb{P}}(f)=\mathcal{K}_{\lambda,\mathbb{P}}(f,f),
-	\end{equation}
-	where $\Vert\nabla^{\otimes 2}f \Vert_{L^2(\mathbb{R}^d)}^2 = \sum_{l=1}^D\int_{\mathbb{R}^d} \sum_{i,j=1}^d\vert\frac{\partial^2 f_l}{\partial t_i \partial t_j}(t)\vert^2dt$.
-A manifold $M_{f^*}^d$ determined by $f^*$ is called a principal manifold for $X$ (or $\mathbb{P}$) with the tuning parameter $\lambda$ if
-	\begin{align}\label{def: PM}
-	f^*=\argmin_{f\in\mathscr{F}(\mathbb{P})}\mathcal{K}_{\lambda,\mathbb{P}}(f),\ \ \mbox{ where }\mathscr{F}(\mathbb{P}):=\left\{f\in C_\infty\cap \nabla^{-\otimes 2}L^2( R^d \rightarrow R^D):\sup_{x\in supp(\mathbb{P})}\left\Vert\pi_f(x)\right\Vert_{R^d} = 1 \right\}.
-	\end{align}
 
@@
   $\left\{r_{j, t}\right\} \gets \pi_{f_t^*}(\mu_{j, N_t, t})$ for $j = 1, \dots, N_t$ and $x_{j, t} \gets f_t^*(r_{j, t})$ for $j = 1, \dots, N_t$\;
-  Add column vector $\mathbf{t}$ as first column of $\mathbf{r}$\;
 }
@@
 \begin{enumerate}
-  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
-  \item Partitioning approach is somewhat ad hoc and may not work in every situation.
-In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent.
-There is not a clear approach to automatic partitioning.
-  \item The process of joining the partitions may introduce error.
+  \item Algorithm assumes stability in shape over time.
   \item Algorithm initialization is sensitive to magnitude of change between time points.

### 2023-04-17 paper/main.tex
add lpme initialization explanation

@@
 The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space.
-In the LPME algorithm, we use the coefficients for spline functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time.
+In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time.
 This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set.
@@
 
-\textit{Initialization}:
+\textit{Initialization}: We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers from $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.
+The centers estimated in this step are then used to reach an initial low-dimensional parameterization.
+
+The PME algorithm uses ISOMAP (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
+However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e.
+low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
+Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
+Instead of using ISOMAP on the data for each time point separately, the LMPE approach removes the time dimension from the cluster centers of all time steps, then uses one ISOMAP run to reach a parameterization for the centers across all time points simultaneously.
+Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.
 

### 2023-04-19 code/functions/pme_functions.cpp
add function to calculate row-wise distances between matrices

@@ -39,6 +39,15 @@ double dist_euclideanF(arma::vec x, arma::vec y, Function f) {
 }
 
 // [[Rcpp::export]]
+arma::vec dist_euclideanC_vec(arma::mat x, arma::mat y) {
+  arma::vec distances(x.n_rows);
+  for (int i = 0; i < x.n_rows; i++) {
+    distances(i) = dist_euclideanC(x.row(i).t(), y.row(i).t());
+  }
+  return distances;
+}
+
+// [[Rcpp::export]]
 double eta_kernelC(arma::vec t, int lambda) {
 
   double lambda_num = lambda / 1.0;

### 2023-04-19 code/functions/sim_data.R
return true data pre-noise

@@ -128,7 +128,7 @@ sim_data <- function(time_val, case, noise, amp_noise, period_noise, N = 1000) {
     t[, 1] <- runif(I, min = -1, max = 1)
     t[, 2] <- runif(I, min = -1, max = 1)
   } else if (case == 8) {
-    t[, 1] <- runif(I, min = 0, max = 10)
+    t[, 1] <- runif(I, min = (1.1 * pi), max = (2.9 * pi))
     t[, 2] <- runif(I, min = -1, max = 1)
   } else if (case == 9) {
     t[, 1] <- runif(I, min = 0, max = pi)
@@ -149,6 +149,9 @@ sim_data <- function(time_val, case, noise, amp_noise, period_noise, N = 1000) {
   amp_noise <- rnorm(D, mean = amp_mean, sd = amp_noise)
   period_noise <- rnorm(D, mean = per_mean, sd = period_noise)
 
+  amp_noise2 <- rnorm(D, mean = amp_mean, sd = 0)
+  period_noise2 <- rnorm(D, mean = per_mean, sd = 0)
+
   X <- map(
     1:nrow(t),
     ~ manifold(
@@ -160,7 +163,19 @@ sim_data <- function(time_val, case, noise, amp_noise, period_noise, N = 1000) {
     unlist() %>%
     matrix(ncol = D, byrow = TRUE)
 
+  X2 <- map(
+    1:nrow(t),
+    ~ manifold(
+      t[.x, ],
+      amp_noise2,
+      period_noise2
+    )
+  ) %>%
+    unlist() %>%
+    matrix(ncol = D, byrow = TRUE)
+
   data.points <- X + noise_vals
   data.points <- cbind(time_val, data.points)
-  return(data.points)
+  X2 <- cbind(time_val, X2)
+  return(list(data.points, X2))
 }

### 2023-04-19 code/lpme.R
update error calculations

@@ -554,11 +554,11 @@ lpme <- function(df, d, tuning.para.seq = c(0, exp(-15:5)), alpha = 0.05, max.co
         temp_data_initial <- temp_data_initial[-which(errors), ]
       }
 
-      proj_error_cv <- map(
-        1:nrow(temp_data_initial),
-        ~ dist_euclideanC(temp_data_initial[.x, 1:D_new2], proj_points_cv[.x, ])
+      proj_error_cv <- dist_euclideanC_vec(
+        temp_data_initial[, 1:D_new2],
+        proj_points_cv
       ) %>%
-        reduce(c)
+        as.vector()
       cv_mse[time_idx] <- mean(proj_error_cv ^ 2)
     }
 

### 2023-04-19 code/pme.R
update error calculations

@@ -381,11 +381,12 @@ pme <- function(x.obs, d, initialization = NULL, N0=20*D, tuning.para.seq=exp((-
       ncol = d
     )
     proj.points <- t(apply(proj.para, 1, fnew))
-    diff.data.fit <- apply(
-      data.initial[, 1:D] - proj.points,
-      1,
-      norm_euclidean
-    )
+    diff.data.fit <- dist_euclideanC_vec(data.initial[, 1:D], proj.points)
+    # diff.data.fit <- apply(
+    #   data.initial[, 1:D] - proj.points,
+    #   1,
+    #   norm_euclidean
+    # )
     MSE <- mean(diff.data.fit ^ 2)
 
     MSE.seq[tuning.ind] <- MSE

### 2023-04-19 paper/main.tex
paper updates

@@
 \doublespacing
+
+\section*{Abstract}
 
@@
 
-\textit{Initialization}: We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+\subsection*{Initialization}
+We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
 Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers from $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
@@
 
-\textit{Fitting}:
+\subsection*{Fitting}
+Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point.
+Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
+This results in an estimated principal manifold for each time point.
+The remainder of the fitting stage involves estimating a thin plate spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
 
-\textit{Tuning}:
+At each time point, the estimated principal manifold has the functional form of a smoothing spline or thin plate spline, depending on the intrinsic dimension $d$.
+As detailed by \cite{greenSilverman1994}, smoothing splines, and thin plate splines as their higher dimensional analogues, may be considered to have knots at the points being used to fit them.
+This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fit must be identical.
+Hence, to obtain coefficients that can be compared across time points, a grid of values ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t^*$.
+This yields a set of outcomes corresponding to same parameter values at each time point.
+Following the discussion in Chapter 7 of \cite{greenSilverman1994}, comparable spline coefficients $\delta$ and $\mathbf{a}$ can then be calculated by minimizing
+\[%
+  S(g) = (\mathbf{X} - E\delta - T^{T}\mathbf{a})^{T}(\mathbf{Y} - E\delta - T^{T}\mathbf{a}) + \gamma\delta^{T}E\delta
+,\]%
+where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $T$ has $T_{ij} = \phi_i(\mathbf{r}_j)$. This minimization is conducted by solving
+\[%
+  \left[
+    \begin{array}{cc}
+      E + \gamma I & T^{T} \\
+      T^{T} & 0
+    \end{array}
+  \right]\left(
+    \begin{array}{c}
+      \mathbf{\delta} \\
+      \mathbf{a}
+    \end{array}
+  \right) = \left(
+    \begin{array}{c}
+      \mathbf{Y} \\
+      0
+    \end{array}
+  \right)
+.\]%
+
+\subsection*{Tuning}
 

### 2023-04-19 code/lpme_simulations.R
use real true values from sim_data

@@ -7,6 +7,7 @@ library(doParallel)
 library(doSNOW)
 library(furrr)
 library(progress)
+library(multimode)
 
 source("code/functions/sim_data.R")
 source("code/functions/calc_pme_est.R")
@@ -16,7 +17,7 @@ source("code/lpme.R")
 
 ### Simulation Case 1
 
-sim_error_case1 <- function(max_time, interval, noise, run = 1, print_plots = FALSE) {
+sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, n, run = 1, print_plots = FALSE) {
   require(tidyverse)
   source("code/functions/sim_data.R")
   source("code/pme.R")
@@ -24,14 +25,24 @@ sim_error_case1 <- function(max_time, interval, noise, run = 1, print_plots = FA
   source("code/functions/calc_pme_est.R")
   source("code/functions/calc_lpme_est.R")
   time_vals <- seq(0, max_time, interval)
-  sim_df <- lapply(
+  sim_list <- lapply(
     time_vals,
     sim_data,
     case = 1,
     noise = 0.15,
-    shape_noise = noise
-  ) %>%
-    reduce(rbind)
+    amp_noise = amp_noise,
+    period_noise = shape_noise,
+    N = n
+  )
+
+  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
+  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
+  for (i in 1:length(sim_list)) {
+    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
+    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
+  }
+  sim_df <- sim_df[-1, ]
+  true_vals <- true_vals[-1, ]
 
   lpme_result <- lpme(sim_df, 1, print_plots = print_plots, verbose = "MSD")
   lpme_vals <- calc_lpme_est(lpme_result, sim_df)
@@ -44,12 +55,6 @@ sim_error_case1 <- function(max_time, interval, noise, run = 1, print_plots = FA
     pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
   }
   pme_vals <- reduce(pme_vals, rbind)
-
-  tau <- sim_df[, 2]
-  true_vals <- matrix(nrow = nrow(sim_df), ncol = ncol(sim_df))
-  true_vals[, 1] <- sim_df[, 1]
[…]

### 2023-04-25 paper/main.tex
complete fitting section draft

@@
 We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers from $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\left\{\hat{\theta}_{j, t}\right\}$.
 HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.
@@
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
-Instead of using ISOMAP on the data for each time point separately, the LMPE approach removes the time dimension from the cluster centers of all time steps, then uses one ISOMAP run to reach a parameterization for the centers across all time points simultaneously.
+Instead of using ISOMAP on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one ISOMAP run to reach a parameterization for the centers across all time points simultaneously.
 Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.
@@
 Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
-This results in an estimated principal manifold for each time point.
+This results in an estimated principal manifold for each time point, $f_t$, and an optimal tuning parameter $\lambda_t^*$.
 The remainder of the fitting stage involves estimating a thin plate spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
@@
 This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fit must be identical.
-Hence, to obtain coefficients that can be compared across time points, a grid of values ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t^*$.
-This yields a set of outcomes corresponding to same parameter values at each time point.
-Following the discussion in Chapter 7 of \cite{greenSilverman1994}, comparable spline coefficients $\delta$ and $\mathbf{a}$ can then be calculated by minimizing
+Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\mathbf{r}_i\right\}_{i=1}^{N}$ ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t$.
+This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point.
+Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing
 \[%
-  S(g) = (\mathbf{X} - E\delta - T^{T}\mathbf{a})^{T}(\mathbf{Y} - E\delta - T^{T}\mathbf{a}) + \gamma\delta^{T}E\delta
+  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t
 ,\]%
-where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $T$ has $T_{ij} = \phi_i(\mathbf{r}_j)$. This minimization is conducted by solving
+where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $R$ has $R_{ij} = \phi_i(\mathbf{r}_j)$, where $\phi_i(\cdot)$ is a linear basis function.
+This minimization is conducted by solving
 \[%
@@
     \begin{array}{cc}
-      E + \gamma I & T^{T} \\
-      T^{T} & 0
+      E + \lambda_t^* I & R^{T} \\
+      R^{T} & 0
     \end{array}
@@
     \begin{array}{c}
-      \mathbf{\delta} \\
-      \mathbf{a}
+      \mathbf{s}_t \\
+      \mathbf{\alpha}_t
     \end{array}
@@
     \begin{array}{c}
-      \mathbf{Y} \\
+      \mathbf{X}_t \\
       0
+    \end{array}
+  \right)
+,\]%
+yielding function
+\[%
+  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k
+.\]%
+
+We denote the PME coefficients at each time point $t$ as $\omega_i = [\mathbf{s}_t, \mathbf{\alpha}_t]$.
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the PME coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$.
+With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function
+\[%
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\phi(t)_j
+,\]%
+where, with matrix $\mathbf{A}$ defined as $\mathbf{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\mathbf{T}$ defined as $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = diag(w_1, \dots, w_T)$,  $\mathbf{\delta}$ and $\mathbf{\nu}$ are the solutions to
+\[%
+  \left(
+  \begin{array}{ccc}
+    2\mathbf{A}\mathbf{W}\mathbf{A} + 2\gamma\mathbf{A} & 2\mathbf{A}\mathbf{W}\mathbf{T} & \mathbf{T} \\
+    2\mathbf{T}^{T}\mathbf{W}\mathbf{A} & 2\mathbf{T}^{T}\mathbf{W}\mathbf{T} & \mathbf{0} \\
+    \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
+  \end{array}
+  \right)\left(
+  \begin{array}{c}
+    \mathbf{\delta} \\
+    \mathbf{\nu} \\
+    \mathbf{m}
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    2\mathbf{A}\mathbf{W}\mathbf{\omega} \\
+    2\mathbf{T}^{T}\mathbf{W}\mathbf{\omega} \\
+    \mathbf{0}
   \end{array}

### 2023-04-25 paper/main.tex
add tuning section draft

@@
 
-We denote the PME coefficients at each time point $t$ as $\omega_i = [\mathbf{s}_t, \mathbf{\alpha}_t]$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the PME coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$.
+We denote the manifold coefficients at each time point $t$ as $\omega_i = [\mathbf{s}_t, \mathbf{\alpha}_t]$.
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$.
 With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function
@@
 
+With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $w_{\gamma}(t) = \left[\mathbf{s}_{\gamma}(t), \mathbf{\alpha}_{\gamma}(t)\right]$.
+Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\mathbf{r}$ may be written as
+\[%
+  f_{\gamma}(t, \mathbf{r}) = \sum_{j=1}^{N}\mathbf{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\mathbf{r} - \mathbf{r}_j^*\|\right) + \sum_{k=1}^{d+1}\mathbf{\alpha}_{\gamma}(t)_k \phi_k(\mathbf{r})
+.\]%
+
 \subsection*{Tuning}
+
+Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross-validation.
+In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
+We denote this function $g_{\gamma}^{(t)}$.
+We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
+\[%
+  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2
+.\]%
+
+The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
 
@@
 
-Contributions:
+\subsection*{Contributions}
 
-Limitations:
+\subsection*{Limitations}
+
 \begin{enumerate}

### 2023-04-25 paper/main.tex
initial algorithm updates

@@
 \[%
-  \left[
+  \left(
     \begin{array}{cc}
@@
     \end{array}
-  \right]\left(
+  \right)\left(
     \begin{array}{c}
@@
 
-Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross-validation.
+Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
 In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
@@
 .\]%
+The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
 
-The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
+Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
 
@@
 \caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
-\KwData{Data points $\left\{y_{i, t}\right\}_{i, t}^{I_t, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
-\KwResult{Analytic formula of $\hat{g}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D}$, optimal tuning parameter $\gamma^*$.}
+  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
+\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
 \For{$t = 1, 2, \dots, T$} {
-  Apply HDMDE algorithm with input $\left(\left\{y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, N_t, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, N_t, t}\right\}_{j = 1}^{N_t}$\;
+  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
 }
 
-Parameterization: Apply ISOMAP to parameterize $\left\{\mu_{j, N_t, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
-Formally set $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets r_j$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
+  Apply ISOMAP to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
+Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
 
 \For{t = 1, 2, \dots, T} {
-  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t^*$ and $\lambda_t^*$\;
-  $\left\{r_{j, t}\right\} \gets \pi_{f_t^*}(\mu_{j, N_t, t})$ for $j = 1, \dots, N_t$ and $x_{j, t} \gets f_t^*(r_{j, t})$ for $j = 1, \dots, N_t$\;
+  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$ and $\lambda_t^*$\;
+  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
 }
 
+  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$, and $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$ and $t = 1, \dots, T$\;
 Let $N = \sum_{t=1}^{T}N_t$, $\mathbf{x}$ be a $N \times D$ matrix of stacked values of $\left\{x_{j, t}\right\}_{t = 1}^T$, and $\mathbf{r}$ be a $N \times d + 1$ matrix of stacked values of $\left\{r_{j, t}\right\}_{t = 1}^T$ \;
@@
 
-Questions:
-\begin{enumerate}
-  \item What is the best way to evaluate the performance of the estimates? We can use SSD for the embedded values, but how to assess the parameterized values?
-  \item Should there be updates to what is defined as the "optimal" model in the PME algorithm?
-Currently it is possible for a model with non-minimal SSD to be chosen within the fitting of a single tuning value.
-  \item Why would we see the SSD increase substantially in late optimization steps?
-  \item Generally, what are the limitations on the estimated manifolds intersecting? How close can it come to intersecting without showing strange behavior?
-  \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?
-  \item How much error does the registration process introduce? Can the LPME algorithm successfully handle a certain number of bad registrations or extractions?
-\end{enumerate}
-
 \nocite{*}

### 2023-04-25 paper/main.tex
add equation numbers

@@
   \end{enumerate}
-  Then the projection index with respect to $f$ is defined as \[%
-    \pi_f(x) = \sup\left\{r \in I: \|x - f(r)\|_{\mathbb{R}^{D}} = \inf_{r' \in I}\|x - f(r')\|_{\mathbb{R}^{D}} \right\}, \ \text{for all} \ x \in \mathbb{R}^{D}
-  .\]%
+  Then the projection index with respect to $f$ is defined as 
+  \begin{equation}
+    \pi_f(x) = \sup\left\{r \in I: \|x - f(r)\|_{\mathbb{R}^{D}} = \inf_{r' \in I}\|x - f(r')\|_{\mathbb{R}^{D}} \right\}, \ \text{for all} \ x \in \mathbb{R}^{D}. \label{eq:1}
+  \end{equation}
 
   (Part II) Suppose $X$ is a continuous random $D$-vector with finite second moments.
-Principal curves of $X$ are all maps $f: I \to \mathbb{R}^{D}$ satisfying the conditions above and the Self-consistency defined as \[%
-    E(X | \pi_f(X) = t) = f(r)
-  .\]%
+Principal curves of $X$ are all maps $f: I \to \mathbb{R}^{D}$ satisfying the conditions above and the Self-consistency defined as
+    \begin{equation}
+      E(X | \pi_f(X) = t) = f(r). \label{eq:2}
+    \end{equation}
 \end{definition}
@@
 Principal manifolds seek to extend this concept to higher dimensions.
-\cite{smolaRegularizedPrincipalManifolds2001} proposed a regularized principal manifold framework, where principal manifolds minimize \[%
-  \argmin_{f \in \mathcal{F}}\left\{E\|X - f\left(\pi_f(X)\right)\|_{\mathbb{R}^D}^2 + \lambda\|Pf\|_{\mathcal{H}}^2\right\}
-\]%
+\cite{smolaRegularizedPrincipalManifolds2001} proposed a regularized principal manifold framework, where principal manifolds minimize
+\begin{equation}
+  \argmin_{f \in \mathcal{F}}\left\{E\|X - f\left(\pi_f(X)\right)\|_{\mathbb{R}^D}^2 + \lambda\|Pf\|_{\mathcal{H}}^2\right\} \label{eq:3}
+\end{equation}
 with $\mathcal{F}$ being a collection of functions and $P$ being an operator that maps $f$ into inner product space $\mathcal{H}$.
@@
 To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
-\[%
-  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr
-,\]%
+\begin{equation}
+  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
+\end{equation}
 referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
@@
 Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
-  \[%
-    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \quad \mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)
-  .\]%
+  \begin{align}
+    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) &= E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 \label{eq:5} \\
+    \mathcal{K}_{\lambda, \mathbb{P}}(f) &= \mathcal{K}_{\lambda, \mathbb{P}}(f, f) \label{eq:6}
+  \end{align}
 
   Then a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $X$ with the tuning parameter $\lambda$ if 
-  \[%
-    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
-  .\]%
+  \begin{equation}
+    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}.
+\label{eq:7}
+  \end{equation}
 \end{definition}
@@
 For each given value of $\lambda$, the fitting stage of the algorithm uses this reduced dataset to find estimated function
-\[%
-  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
-,\]%
+\begin{equation}
+  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f), \label{eq:8}
+\end{equation}
 where 
-\[%
-  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
-\]% 
+\begin{equation}
+  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 \label{eq:9}
+\end{equation}
 approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses the optimal value of $\lambda$, $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
-\[%
-  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2
-\]%
+\begin{equation}
+  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2 \label{eq:10}
+\end{equation}
 is an empirical estimate of the MSD using the full dataset.
@@
 Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as
-\[%
-  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \sum_{t=1}^{T} \lambda_t \|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 + \gamma\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^d)}^2
-.\]%
+\begin{equation}
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \sum_{t=1}^{T} \lambda_t \|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 + \gamma\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^d)}^2 \label{eq:11}
+\end{equation}
 
@@
 Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing
-\[%
-  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t
-,\]%
+\begin{equation}
+  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t, \label{eq:12}
+\end{equation}
 where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $R$ has $R_{ij} = \phi_i(\mathbf{r}_j)$, where $\phi_i(\cdot)$ is a linear basis function.
 This minimization is conducted by solving
-\[%
+\begin{equation}
   \left(
@@
     \end{array}
-  \right)
-,\]%
+  \right), \label{eq:13}
+\end{equation}
 yielding function
-\[%
-  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k
-.\]%
+\begin{equation}
+  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k. \label{eq:14}
+\end{equation}
 
@@
 With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function
-\[%
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\phi(t)_j
-,\]%
+\begin{equation}
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\phi(t)_j, \label{eq:15}
+\end{equation}
 where, with matrix $\mathbf{A}$ defined as $\mathbf{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\mathbf{T}$ defined as $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = diag(w_1, \dots, w_T)$,  $\mathbf{\delta}$ and $\mathbf{\nu}$ are the solutions to
-\[%
+\begin{equation}
   \left(
@@
   \end{array}
-  \right)
-.\]%
+  \right). \label{eq:16}
+\end{equation}
 
@@
 Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\mathbf{r}$ may be written as
-\[%
-  f_{\gamma}(t, \mathbf{r}) = \sum_{j=1}^{N}\mathbf{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\mathbf{r} - \mathbf{r}_j^*\|\right) + \sum_{k=1}^{d+1}\mathbf{\alpha}_{\gamma}(t)_k \phi_k(\mathbf{r})
-.\]%
+\begin{equation}
+  f_{\gamma}(t, \mathbf{r}) = \sum_{j=1}^{N}\mathbf{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\mathbf{r} - \mathbf{r}_j^*\|\right) + \sum_{k=1}^{d+1}\mathbf{\alpha}_{\gamma}(t)_k \phi_k(\mathbf{r}).
+\label{eq:17}
+\end{equation}
 
@@
 We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
-\[%
-  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2
-.\]%
+\begin{equation}
+  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2. \label{eq:18}
+\end{equation}
 The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
@@
   Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$, and $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$ and $t = 1, \dots, T$\;
-Let $N = \sum_{t=1}^{T}N_t$, $\mathbf{x}$ be a $N \times D$ matrix of stacked values of $\left\{x_{j, t}\right\}_{t = 1}^T$, and $\mathbf{r}$ be a $N \times d + 1$ matrix of stacked values of $\left\{r_{j, t}\right\}_{t = 1}^T$ \;
 

### 2023-04-25 paper/main.tex
update algorithm notation

@@
 \For{t = 1, 2, \dots, T} {
-  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$ and $\lambda_t^*$\;
+  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
   $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
@@
 
-  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$, and $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$ and $t = 1, \dots, T$\;
+  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$\;
 
-Set $\pi_{g(0)}(x_i) \gets r_i$ for $i = 1, 2, \dots, N$\;
+\For{t = 1, 2, \dots, T} {
+  Set $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
+  Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
+  Set $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
+}
 
-\For{l = 1, 2, \dots, L}{
-  $\gamma \gets \gamma_l$\;
-  Obtain $g_{(1)}$ by solving $\left(\begin{array}{ccc}
-      2\mathbf{E}\mathbf{W}\mathbf{E} + 2\gamma\mathbf{E} & 2\mathbf{E}\mathbf{W}\mathbf{T} & \mathbf{T} \\
-      2\mathbf{T}^{T}\mathbf{W}\mathbf{E} & 2\mathbf{T}^{T}\mathbf{W}\mathbf{T} & \mathbf{0} \\
-      \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
-    \end{array}\right)\left(\begin{array}{c}
-    s_l \\
-    \alpha_l \\
-    m_l
-    \end{array}\right) = \left(\begin{array}{c}
-    2\mathbf{E}\mathbf{W}\mu_l \\
-    2\mathbf{T}^{T}\mathbf{W}\mu_l \\
-    \mathbf{0}
-  \end{array}\right)$\;
-  $\mathcal{\epsilon} \gets 2 \times \epsilon^*$, $n \gets 1$, $\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) \gets \frac{1}{N}\sum_{j = 1}^{N}\|x_{j} - f_{(1)}(\pi_{f_{(1)}}(x_j))\|_{\mathbb{R}^{D}}^2$ \;
-  \While{$\epsilon > \epsilon^*$ and $n \leq itr$} {
-    Compute $f_{(n + 1)})$ from $f_{(n)}$ by solving (15) \;
-    $\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) \gets \frac{1}{N}\sum_{j=1}^{N}\|x_j - f_{(n + 1)}(\pi_{f_{(n + 1)}}(x_j))\|_{\mathbb{R}^D}^2$ \;
-    $\epsilon \gets |\mathcal{D}_{\hat{Q}_N}(f_{(n + 1)}) - \mathcal{D}_{\hat{Q}_N}(f_{(n)})| / \mathcal{D}_{\hat{Q}_N}(f_{(n)})$ and $n \gets n + 1$\;
+  Define $\mathbf{\omega}$ by setting $\mathbf{\omega}_t = \left[\mathbf{s}_t, \mathbf{\alpha}_t\right]$\;
+  \For{l = 1, 2, \dots, L} {
+    Compute $g_{\gamma_l}(t)$ by solving \eqref{eq:16}\;
+    \For{t = 1, 2, \dots, T} {
+      Compute $g_{\gamma_l}^{(t)}$ by solving \eqref{eq:16}\;
     }
-  $\hat{f}_l \gets f_{(n)}$\;
+    Estimate $MSD(\gamma_l)$ using \eqref{eq:18}\;
   }
-$l^* \gets \argmin_l\left\{\frac{1}{T}\sum_{t = 1}^{T}\frac{1}{I_t}\sum_{i = 1}^{I_t}\|y_{i, t} - \hat{f}_l(\pi_{f_l}(y_{i, t}))\|_{\mathbb{R}^D}^2: \ l = 1, 2, \dots, L \right\}$ \;
-$\kappa \gets \max\left\{\|\pi_{\hat{f}_{g^*}}(y_i, t)\|_{\mathbb{R}^{d}}: \ i = 1, 2, \dots, I_t; j = 1, 2, \dots, T \right\}$ \;
-$f^*(r) = \hat{f}_{l^*}(r)$, where the analytic formula of $f^*$ is from (14), and $\gamma^* \gets \gamma_l^*$.
+  $\gamma^* = \arg\min_{\gamma}MSD(\gamma)$\;
+  $f^*(t, \mathbf{r}) = f_{\gamma^*}(t, \mathbf{r})$, where the form of $f^*(t, \mathbf{r})$ is given in \eqref{eq:17}
 \end{algorithm}

### 2023-05-09 paper/main.tex
Add Kun comments

@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+\newcommand{\T}{\intercal}
+\newcommand{\commentout}[1]{}
+
+\newcommand{\kmedit}[1]{{\color{purple}  #1}}
+\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
+\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}
 
@@
 
-\title{Longitudinal Manifold Estimation}
+\title{Longitudinal Principal Manifold Estimation}
 \author[1]{Robert Zielinski, Kun Meng, and Ani Eloyan}
@@
 
-The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}.
+The remainder of this article is laid out as follows: In Section 2, we review the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}.
 Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this.
@@
 
-The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data.
+The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data.
+However, the PS framework in \cite{hastiePrincipalCurves1989} is flawed by the model complexity selection, which \cite{duchamp1996extremal} pointed out.
+Motivated by the penalization approaches in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework and algorithm.
+Since our LPME is based on the PME, we review the PME in this section for the audience's convenience.
+The following concepts and notations are necessary for both the PME and our proposed LPME, and they will be used throughout this paper
+\begin{itemize}
+\item $d$ and $D$ are two positive integers denoting dimensions, and $d<D$; \meng{Add some explanations of the dimensionalities $d$ and $D$.}
+\item for any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$;
+    \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$;
+    \item for every $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, the image of $f$ is referred to as the $d$-dimensional manifold $M_f^d$ determined by $f$, i.e., $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$;
+    \item For each $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and each point $x\in\mathbb{R}^D$, $\pi_f(x)$ is a point in $\mathbb{R}^d$ and denotes the ``projection index" of $x$; roughly, $\pi_f(x)$ is a point in $d$-dimensional space such that $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$; $\pi_f(x)$ is a generalization of the counterpart in \cite{hastiePrincipalCurves1989}, and its details can be found in Section 2 of \cite{mengPrincipalManifoldEstimation2021};
+    \item suppose $u(r_1,\ldots,r_d)$ is a scalar-valued function defined on $\mathbb{R}^d$; $\nabla^{\otimes 2}u=\left(\frac{\partial^2 u}{\partial r_i \partial r_j}\right)_{1\le i,j\le d}$ denotes the Hessian matrix of $u$, and $\|\nabla^{\otimes 2}u\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2} u(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2\right)^{\frac{1}{2}}$; in addition, $\|\nabla^{\otimes 2} u\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2dr $, and
+    \begin{align*}
+        \nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\},
+    \end{align*}
+    where $\mathcal{D}'$ denotes the collection of distributions (also known as generalized functions, see \cite{duchon1977splines} and Chapter 6 of \cite{rudin1991functional});
+    \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$ denotes the collection of vector-valued functions $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ whose components are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d})$; besides, $\mathcal{C}_\infty\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)=\mathcal{C}_\infty(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$; the following functional of $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ measures the curvature of the manifold $M_f^d$ and will be used throughout this paper
+    \begin{equation}
+  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \|\nabla^{\otimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2  = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
+\end{equation}
+\end{itemize}
+With the preparations above, the definition of principal manifolds proposed in \cite{mengPrincipalManifoldEstimation2021} is presented as follows
+\commentout{
 \cite{hastiePrincipalCurves1989} defines principal curves as follows:
@@
 \end{equation}
-referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
+referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:}
 \begin{definition}
   \label{def:principal_manifolds}
-  Let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments.
+  Let $\boldsymbol{X}$ be a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments.
 Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
-  \begin{align}
-    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) &= E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 \label{eq:5} \\
-    \mathcal{K}_{\lambda, \mathbb{P}}(f) &= \mathcal{K}_{\lambda, \mathbb{P}}(f, f) \label{eq:6}
+  \begin{align}\label{eq:5}
+    \begin{aligned}
+&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
+&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f).
+    \end{aligned}
   \end{align}
-
-  Then a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $X$ with the tuning parameter $\lambda$ if 
-  \begin{equation}
-    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}.
-\label{eq:7}
+A manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $\boldsymbol{X}$ with the tuning parameter $\lambda$ if 
+  \begin{equation}\label{eq:7}
+    \begin{aligned}
+        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \\ 
+    &\text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}.
+    \end{aligned}
   \end{equation}
 \end{definition}
-
-In this setting, we can see that $\pi_{f_{\lambda}^{*}}(X)$ maps the $D$-vector $X$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
+\meng{Add some explanations for the terms in Eq.~\eqref{eq:5}.}
+In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
 
 To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm.
-This algorithm has three main steps: data reduction, fitting, and tuning.
-In the reduction step, the HDMDE algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$.
-The number of clusters is chosen using an iterative hypothesis testing approach.
-For each given value of $\lambda$, the fitting stage of the algorithm uses this reduced dataset to find estimated function
+This algorithm has three main steps: data reduction, fitting, and tuning (see Figure 2 of \cite{mengPrincipalManifoldEstimation2021}).
+\meng{I would suggest presenting Figure 2 of \cite{mengPrincipalManifoldEstimation2021} in the present paper.}
+\begin{enumerate}
+    \item In the data reduction step, the high-dimensional mixture density estimation (HDMDE) algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$.
+The number $N$ of clusters is chosen using an iterative hypothesis testing approach, and $N$ tends to be much smaller than $I$ (see the simulation study in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
+    \item For each given value of $\lambda$, the fitting stage of the PME algorithm uses this reduced data set $\left\{\mu_j\right\}_{j=1}^{N}$ to find estimated function 
 \begin{equation}
@@
 \end{equation}
-where 
+where $\mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$ approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$.
+\item The tuning stage chooses the optimal value of the parameter $\lambda$ by $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
 \begin{equation}
-  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 \label{eq:9}
-\end{equation}
-approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses the optimal value of $\lambda$, $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
-\begin{equation}
-  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2 \label{eq:10}
+  \mathcal{D}\left(\hat{f}_\lambda\right) = \frac{1}{I}\sum_{i=1}^{I}\left\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\right\|_{\mathbb{R}^{D}}^2 \label{eq:10}
 \end{equation}
 is an empirical estimate of the MSD using the full dataset.
+\end{enumerate}
 
-\cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of spline form, meaning that when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$.
+Using results in \cite{duchon1977splines} and \cite{wahba1990}, \cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of a spline form.
+Specifically, when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$.
 Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in \cite{mengPrincipalManifoldEstimation2021}.
+
+\meng{I commented out many literature review discussions about \cite{hastiePrincipalCurves1989} and \cite{smolaRegularizedPrincipalManifolds2001} to make this section more concise.
+In addition, if the formulas above are not referred to in our discussions, they shouldn't be given labels.}
 
@@
 
-We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}.
+We now propose longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}.
+
 Let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
-As in the setting described above, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
+As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
 However, in this updated case, the function determining the manifold may change over time.
-Note that we assume changes over time to be smooth.
-Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as
+
+\subsection{Definition of LPM}
+
+Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively.
+In addition, we assume $F\in\mathcal{C}_\infty\cap\nabla^{-\otimes2}L^2(\mathbb{R}^d\times\mathbb{R}\rightarrow\mathbb{R}^D)$.
+For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold at the time point.
+For given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we plan to minimize the following functional
 \begin{equation}
-  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \sum_{t=1}^{T} \lambda_t \|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 + \gamma\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^d)}^2 \label{eq:11}
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) = \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt.
+\label{eq:11}
 \end{equation}
+\meng{Combine Eq.~\eqref{eq:5} and the functional above.}
 
-To solve this problem, we propose an extension of the method developed in \cite{mengPrincipalManifoldEstimation2021}.
+\meng{Add the interpretation of the terms above.
+This interpretation presents the key idea of this paper.
+I need to address it carefully.
+One thing I would strongly suggest is to explain why we chose to deal with space and time separately.}
+
+\meng{Mentioned the functional above in all the four steps of LPME algorithm --- to emphasize that we are going to minimize the functional above.}
+
+\commentout{Note that we assume changes over time to be smooth.
+Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as }
+
+\subsection{Alogirthm}
+
+To solve this problem, we propose an extension of the PME algorithm developed in \cite{mengPrincipalManifoldEstimation2021}.
 The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space.
@@
 This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set.
+Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning.
 Details of this approach are described below.
 
-\subsection*{Initialization}
-We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\left\{\hat{\theta}_{j, t}\right\}$.
+\meng{I would suggest making a flowchart similar to Figure 2 of \cite{mengPrincipalManifoldEstimation2021}.}
+
+\subsection*{Data Reduction}
+We begin by considering $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} (Section 5.3 therein) to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$.
 HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.
+
+This step reduces the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to the size $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
 The centers estimated in this step are then used to reach an initial low-dimensional parameterization.
 
-The PME algorithm uses ISOMAP (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
+\subsection*{Initialization}
+The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
 However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e.
-low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
+low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another) \meng{I would suggest writing more explanations for the ``inconsistent parameterization directions." Some visual illustrations are necessary.}.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
-Instead of using ISOMAP on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one ISOMAP run to reach a parameterization for the centers across all time points simultaneously.
+\meng{This paragraph should be devoted to the explanation for ``inconsistent parameterization directions." The comparison can be presented using visualizations.}
+
+Instead of using Isomap on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one Isomap run to reach a parameterization for the centers across all time points simultaneously.
 Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.
+\meng{I would suggest presenting simulation studies showing the ``consistency over time." Specifically, we may compare the difference between time-wise and over-time initializations via simulations.}
 
@@
 Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
-This results in an estimated principal manifold for each time point, $f_t$, and an optimal tuning parameter $\lambda_t^*$.
-The remainder of the fitting stage involves estimating a thin plate spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$, which deals with the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in Eq.~\eqref{eq:11}.
 
-At each time point, the estimated principal manifold has the functional form of a smoothing spline or thin plate spline, depending on the intrinsic dimension $d$.
-As detailed by \cite{greenSilverman1994}, smoothing splines, and thin plate splines as their higher dimensional analogues, may be considered to have knots at the points being used to fit them.
-This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fit must be identical.
-Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\mathbf{r}_i\right\}_{i=1}^{N}$ ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t$.
-This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point.
-Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing
-\begin{equation}
-  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t, \label{eq:12}
-\end{equation}
-where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $R$ has $R_{ij} = \phi_i(\mathbf{r}_j)$, where $\phi_i(\cdot)$ is a linear basis function.
+The remainder of the fitting stage deals with the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$, i.e., the curvature in the time dimension (see Eq.~\eqref{eq:11}).
+Specifically, the remainder of this stage involves estimating a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
+\meng{The term ``coefficient space" is clear to me, but it might be confusing to other readers who are unfair with the details of PME.}
+
+At each time point, the estimated principal manifold has a spline form.
+As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points being used to fit them.
+This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical.
+Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point  \meng{What is the $N$ herein?}.
+This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $X_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
+Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
+\begin{itemize}
+    \item $\mathcal{L}(f_t) = (\boldsymbol{X}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{X}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
+    \item \meng{specify the function $\eta_{4-d}$};
+    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$ \meng{$\eta_d$ or $\eta_{4-d}$?};
+    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
+    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
+\end{itemize}
 This minimization is conducted by solving
@@
     \begin{array}{cc}
-      E + \lambda_t^* I & R^{T} \\
-      R^{T} & 0
+      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
+      \boldsymbol{R} & \boldsymbol{0}
     \end{array}
@@
     \begin{array}{c}
-      \mathbf{s}_t \\
-      \mathbf{\alpha}_t
+      \boldsymbol{s}_t \\
+      \boldsymbol{\alpha}_t
     \end{array}
@@
     \begin{array}{c}
-      \mathbf{X}_t \\
-      0
+      \boldsymbol{X}_t \\
+      \boldsymbol{0}
     \end{array}
-  \right), \label{eq:13}
+  \right). \label{eq:13}
 \end{equation}
-yielding function
+The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
 \begin{equation}
-  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k. \label{eq:14}
+  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
 \end{equation}
 
-We denote the manifold coefficients at each time point $t$ as $\omega_i = [\mathbf{s}_t, \mathbf{\alpha}_t]$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$.
-With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function
+We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$.
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$ \meng{I would suggest writing the $\tau_t$ in a more precise way.}, the weights for each time are equal to
+$$w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
+
+With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function \meng{Explicitly say that we apply a weighted cubic spline regression using $\{t_i\}_{=1}^T$ as the predictors and $\{\boldsymbol{\omega}_t\}_t$ as responses.
+I mean, we need to clearly point out the predictors and responses of the spline regression.}
 \begin{equation}
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\phi(t)_j, \label{eq:15}
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
 \end{equation}
-where, with matrix $\mathbf{A}$ defined as $\mathbf{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\mathbf{T}$ defined as $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = diag(w_1, \dots, w_T)$,  $\mathbf{\delta}$ and $\mathbf{\nu}$ are the solutions to
+where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
 \begin{equation}
@@
   \begin{array}{ccc}
-    2\mathbf{A}\mathbf{W}\mathbf{A} + 2\gamma\mathbf{A} & 2\mathbf{A}\mathbf{W}\mathbf{T} & \mathbf{T} \\
-    2\mathbf{T}^{T}\mathbf{W}\mathbf{A} & 2\mathbf{T}^{T}\mathbf{W}\mathbf{T} & \mathbf{0} \\
-    \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
+    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{A} + 2\gamma\boldsymbol{A} & 2\boldsymbol{A}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{T} \\
+    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{A} & 2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{0} \\
+    \boldsymbol{T}^{T} & \boldsymbol{0} & \boldsymbol{0}
   \end{array}
@@
   \begin{array}{c}
-    \mathbf{\delta} \\
-    \mathbf{\nu} \\
-    \mathbf{m}
+    \boldsymbol{\delta} \\
+    \boldsymbol{\nu} \\
+    \boldsymbol{m}
   \end{array}
@@
   \begin{array}{c}
-    2\mathbf{A}\mathbf{W}\mathbf{\omega} \\
-    2\mathbf{T}^{T}\mathbf{W}\mathbf{\omega} \\
-    \mathbf{0}
+    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{\Omega} \\
+    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{\Omega} \\
+    \boldsymbol{0}
   \end{array}
-  \right). \label{eq:16}
+  \right), \label{eq:16}
 \end{equation}
+where $\boldsymbol{\Omega}=(\boldsymbol{\omega}_1,\ldots, \boldsymbol{\omega}_T)^\T$.
 
-With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $w_{\gamma}(t) = \left[\mathbf{s}_{\gamma}(t), \mathbf{\alpha}_{\gamma}(t)\right]$.
-Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\mathbf{r}$ may be written as
+\meng{Since the dimensionality of time $t$ is $d=1$, the spline $\eta_{4-d}$ is the cubic spline. We need to replace all $\eta_{4-d}(\Vert t-t_j\Vert)$ with $\Vert t-t_j\Vert^3$.}
+
+With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$.
+Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\boldsymbol{r}$ may be written as
 \begin{equation}
-  f_{\gamma}(t, \mathbf{r}) = \sum_{j=1}^{N}\mathbf{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\mathbf{r} - \mathbf{r}_j^*\|\right) + \sum_{k=1}^{d+1}\mathbf{\alpha}_{\gamma}(t)_k \phi_k(\mathbf{r}).
+  F_{\gamma}(t, \boldsymbol{r}) = \sum_{j=1}^{N}\boldsymbol{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\boldsymbol{r} - \boldsymbol{r}_j^*\|\right) + \sum_{k=1}^{d+1}\boldsymbol{\alpha}_{\gamma}(t)_k \phi_k(\boldsymbol{r}).
 \label{eq:17}
@@
 
+\meng{From the methodology viewpoint, the fitting section is the most important part of this paper.}
+
+\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction.
+The smoothness can be represented by the covariance kernel we use in the GPR.
+For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels.
+The taste of the GPR is similar to \cite{dunson2021inferring}, and we may cite \cite{dunson2021inferring}.}
+
 \subsection*{Tuning}
 
-Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
+Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
 In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
 We denote this function $g_{\gamma}^{(t)}$.
-We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
+We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows: \meng{What are the $Y_{i,t}$'s?}
 \begin{equation}
@@
 The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
+\meng{I would suggest some simulation studies showing the performance of the $\gamma^*$ something, like the Figure 4 of \cite{mengPrincipalManifoldEstimation2021}.}
 
 Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
+\meng{Do some simulation studies and compare the leave-one-out CV and k-fold CV.}
 
@@
 
+\meng{Make the notations in Algorithm \ref{alg:lpme} consistent with that in the description above.}
+
+\meng{FYI, to fit a circle-like manifold, you may need to refer to the following papers: \cite{fasy2014confidence}, \cite{dunson2021inferring}.}
+
 \begin{algorithm}
@@
 
-  Apply ISOMAP to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
+  Apply Isomap to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
 Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
@@
 
-  Define $\mathbf{\omega}$ by setting $\mathbf{\omega}_t = \left[\mathbf{s}_t, \mathbf{\alpha}_t\right]$\;
+  Define $\boldsymbol{\omega}$ by setting $\boldsymbol{\omega}_t = \left[\boldsymbol{s}_t, \boldsymbol{\alpha}_t\right]$\;
   \For{l = 1, 2, \dots, L} {
@@
   $\gamma^* = \arg\min_{\gamma}MSD(\gamma)$\;
-  $f^*(t, \mathbf{r}) = f_{\gamma^*}(t, \mathbf{r})$, where the form of $f^*(t, \mathbf{r})$ is given in \eqref{eq:17}
+  $f^*(t, \boldsymbol{r}) = f_{\gamma^*}(t, \boldsymbol{r})$, where the form of $f^*(t, \boldsymbol{r})$ is given in \eqref{eq:17}
 \end{algorithm}

### 2023-05-10 paper/main.tex
PME section updates

@@
 \begin{itemize}
-\item $d$ and $D$ are two positive integers denoting dimensions, and $d<D$; \meng{Add some explanations of the dimensionalities $d$ and $D$.}
+\item $d$ and $D$ are two positive integers denoting the dimensions of the low-dimensional manifold and high-dimensional space, respectively; 
 \item for any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$;
@@
 \end{definition}
-\meng{Add some explanations for the terms in Eq.~\eqref{eq:5}.}
 In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
+Thus, the first segment of the functional $\mathcal{K}_{\lambda, \mathbb{P}}$ provides a measure of the expected distance between the provided $D$-vector $\mathbf{X}$ and the value of $\mathbf{X}$ when projected onto the manifold.
+Meanwhile the second segment of $\mathcal{K}_{\lambda, \mathbb{P}}$ penalizes the roughness of the function $f$, with the magnitude of the penalty dependent on the provided value of $\lambda$.
 
@@
 For given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we plan to minimize the following functional
-\begin{equation}
-  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) = \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt.
-\label{eq:11}
-\end{equation}
+\begin{align}
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \label{eq:11} \\
+  &= \mathcal{K}_{\lambda, \mathbb{P}}(f_t) + \gamma \cdot \int{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \nonumber
+.\end{align}
 \meng{Combine Eq.~\eqref{eq:5} and the functional above.}

### 2023-06-16 paper/main.tex

@@
   \begin{align}\label{eq:5}
-    \begin{aligned}
 &\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
-&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f).
-    \end{aligned}
+&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
   \end{align}
@@
   \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \label{eq:11} \\
-  &= \mathcal{K}_{\lambda, \mathbb{P}}(f_t) + \gamma \cdot \int{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \nonumber
+  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \nonumber
 .\end{align}
-\meng{Combine Eq.~\eqref{eq:5} and the functional above.}
+
+By minimizing the functional above, we retain the goals of Eq.~\eqref{eq:5} while also penalizing the changes in $F$ over time.
+Thus, we seek to minimize the distance between the observations $\mathbf{X_t}$ and their projections to $f_t$ while separately discouraging roughness both over the manifolds at each time point and over the changes in manifolds in the time dimension.
+The ability to distinguish between smoothing at a fixed time point and smoothing over time differentiates this approach from a thin plate spline, which would smooth over both the manifold and time jointly.
+Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.
 
@@
 This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical.
-Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point  \meng{What is the $N$ herein?}.
-This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $X_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
+Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ being specified based on the maximum number of clusters identified in the data reduction step at any time point.
+This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
 Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
 \begin{itemize}
-    \item $\mathcal{L}(f_t) = (\boldsymbol{X}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{X}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
-    \item \meng{specify the function $\eta_{4-d}$};
-    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$ \meng{$\eta_d$ or $\eta_{4-d}$?};
+    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
+    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
+    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
     \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
@@
     \begin{array}{c}
-      \boldsymbol{X}_t \\
+      \boldsymbol{Y}_t \\
       \boldsymbol{0}
@@
 \begin{equation}
-  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
+  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
 \end{equation}
@@
 To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$ \meng{I would suggest writing the $\tau_t$ in a more precise way.}, the weights for each time are equal to
-$$w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
+$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
 
-With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function \meng{Explicitly say that we apply a weighted cubic spline regression using $\{t_i\}_{=1}^T$ as the predictors and $\{\boldsymbol{\omega}_t\}_t$ as responses.
-I mean, we need to clearly point out the predictors and responses of the spline regression.}
+With a given tuning value $\gamma$, we the fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$.
+This function takes the form
+
 \begin{equation}
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
 \end{equation}
-where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
+where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
 \begin{equation}
@@
 
-\meng{Since the dimensionality of time $t$ is $d=1$, the spline $\eta_{4-d}$ is the cubic spline. We need to replace all $\eta_{4-d}(\Vert t-t_j\Vert)$ with $\Vert t-t_j\Vert^3$.}
-
 With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$.
@@
 \begin{equation}
-  F_{\gamma}(t, \boldsymbol{r}) = \sum_{j=1}^{N}\boldsymbol{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\boldsymbol{r} - \boldsymbol{r}_j^*\|\right) + \sum_{k=1}^{d+1}\boldsymbol{\alpha}_{\gamma}(t)_k \phi_k(\boldsymbol{r}).
+  F_{\gamma}(t, \boldsymbol{r}) = \sum_{j=1}^{N}\boldsymbol{s}_{\gamma}(t)_j \eta_{d}\left(\|\boldsymbol{r} - \boldsymbol{r}_j^*\|\right) + \sum_{k=1}^{d+1}\boldsymbol{\alpha}_{\gamma}(t)_k \phi_k(\boldsymbol{r}).
 \label{eq:17}
@@
 \begin{equation}
-  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2. \label{eq:18}
+  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
 \end{equation}

### 2023-06-27 code/lpme_simulations.R
fix principal curve predictions

@@ -89,7 +89,7 @@ sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     }
     opt_curve <- which.min(curve_error)
     principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals <- curve_vals[[opt_curve]]
+    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
   }
   pme_vals <- reduce(pme_vals, rbind)
   principal_curve_vals <- reduce(principal_curve_vals, rbind)
@@ -265,7 +265,7 @@ sim_error_case2 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     }
     opt_curve <- which.min(curve_error)
     principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals <- curve_vals[[opt_curve]]
+    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
   }
   pme_vals <- reduce(pme_vals, rbind)
   principal_curve_vals <- reduce(principal_curve_vals, rbind)
@@ -445,7 +445,7 @@ sim_error_case3 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     }
     opt_curve <- which.min(curve_error)
     principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals <- curve_vals[[opt_curve]]
+    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
   }
   pme_vals <- reduce(pme_vals, rbind)
   principal_curve_vals <- reduce(principal_curve_vals, rbind)
@@ -639,7 +639,7 @@ sim_error_case4 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     }
     opt_curve <- which.min(curve_error)
     principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals <- curve_vals[[opt_curve]]
+    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
   }
   pme_vals <- reduce(pme_vals, rbind)
   principal_curve_vals <- reduce(principal_curve_vals, rbind)
@@ -817,7 +817,7 @@ sim_error_case5 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     }
     opt_curve <- which.min(curve_error)
     principal_curve_result[[t]] <- curves[[opt_curve]]
[…]

### 2023-06-27 paper/main.tex
add initial simulation study writeup

@@
 
+To assess the performance of the LPME algorithm, we compare its performance with the performance of two alternative approaches on simulated datasets.
+We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
+For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.
+
+In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run naively.
+The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}).
+In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
+Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$, when the spline smoothing method is used and $\lambda_g = g + 0.5$ for $g = 0, \dots, 25$ when the Gaussian process smoothing method is used.
+In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
+
+Notably, the manifolds estimated in the simulated datasets included several connected manifolds, which methods based on the principal curves, including the principal manifold framework used here, are typically unable to fit due to the self-consistency property discussed above.
+To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$.
+As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
+However, by adding a third dimension equal to the angle of each observation from the origin, we the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
+This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
+This process is illustrated graphically below [Figure still needs to be added].
+
 \section{Application}

### 2023-06-27 paper/main.tex
Updates from Overleaf

--- 2023-06-16 paper/main.tex
@@
   \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \label{eq:11} \\
-  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \nonumber
+  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \nonumber
 .\end{align}

### 2023-06-28 code/lpme_simulations.R
comment GP fitting trials

@@ -53,14 +53,14 @@ sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_cha
   )
   lpme_vals <- calc_lpme_est(lpme_result, sim_df)
 
-  lpme_result_gp <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "gp",
-    print_plots = print_plots,
-    verbose = TRUE
-  )
-  lpme_vals_gp <- calc_lpme_est(lpme_result_gp, sim_df)
+  # lpme_result_gp <- lpme(
+  #   sim_df,
+  #   1,
+  #   smoothing_method = "gp",
+  #   print_plots = print_plots,
+  #   verbose = TRUE
+  # )
+  # lpme_vals_gp <- calc_lpme_est(lpme_result_gp, sim_df)
 
   pme_result <- list()
   pme_vals <- list()
@@ -139,12 +139,12 @@ sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     unlist() %>%
     mean()
 
-  lpme_error_gp <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], lpme_vals_gp[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
+  # lpme_error_gp <- map(
+  #   1:nrow(true_vals),
+  #   ~ dist_euclidean(true_vals[.x, ], lpme_vals_gp[.x, ])^2
+  # ) %>%
+  #   unlist() %>%
+  #   mean()
 
   sim_case1 <- list(
     df = sim_df,
@@ -153,18 +153,18 @@ sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     period_noise = shape_noise,
     n = n,
     lpme_result = lpme_result,
-    lpme_gp_result = lpme_result_gp,
+    # lpme_gp_result = lpme_result_gp,
     pme_results = pme_result,
     principal_curve_results = principal_curve_result,
     lpme_error = lpme_error,
-    lpme_gp_error = lpme_error_gp,
+    # lpme_gp_error = lpme_error_gp,
     pme_error = pme_error,
     principal_curve_error = principal_curve_error,
     plot = p
   )
   sim_dir <- "simulations/case1/"
   if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir)
+    dir.create(sim_dir, recursive = TRUE)
   }
   filename <- paste0(
[…]

### 2023-07-03 paper/main.tex
add simulation table

@@
 \usepackage{tikz}
+\usepackage{geometry}
+\geometry{
+  left=10mm,
+  right=10mm
+  }
 
@@
 \setlength{\oddsidemargin}{-0.5cm}
+\setlength{\evensidemargin}{-0.5cm}
 \renewcommand{\topmargin}{-2cm}
@@
 
-To assess the performance of the LPME algorithm, we compare its performance with the performance of two alternative approaches on simulated datasets.
+To assess the performance of the LPME algorithm, we compare its performance with the that of two alternative approaches on simulated datasets.
 We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
 For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.
+The embedding functions used to define each manifold are given in Table \ref{table:1}.
+In these functions, the values of $\zeta$, $\alpha$, and $\beta$, as well as the functional form of $g(\cdot)$ are varied between trials.
+The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance varied between trials.
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c c c c c|}
+    \hline
+    Case & $d$ & $D$ & $f(\mathbf{r})$ & Domain \\
+    \hline
+    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
+    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
+    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{cos} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
+    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
+    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
+    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
+    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
+    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
+    9 & 2 & 3 & $\left(\theta\text{sin} \ (\beta_1r_1)\text{cos} \ (\beta_2r_2), \ \theta\text{sin} \ (\beta_1)\text{sin} \ (\beta_2r_2), \ \theta\text{cos} \ (\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
+    \hline
+  \end{tabular}
+  \caption{Embedding functions used for simulation trials.}
+  \label{table:1}
+\end{table}
 

### 2023-07-04 Examples_for_PME.R
delete unnecessary files

@@ -1,381 +1,0 @@
-####################################################################
-##########      A Demonstration of PME Using Examples    ###########
-##########               Kun (Michael) Meng              ###########
-##########        Department of Biostatistics            ###########
-##########               Brown University                ###########
-####################################################################
-
-# install.packages("plot3D")
-library("plot3D")
-
-################### Section 1, d=1 and D=2 #########################
-####################################################################
-par(mfrow=c(2,2))
-
-source("code/pme.R")
-
-### Case I.
-
-# Simulated data
-# I=1000                              # Sample size
-set.seed(100)
-I <- 1000
-t=rnorm(I,mean = 0,sd=1)
-sd.noise=0.15                       # Standard deviation of the noise
-e1=rnorm(I,mean = 0,sd=sd.noise)    # noise
-e2=rnorm(I,mean = 0,sd=sd.noise)
-X=matrix(NA,ncol = 2,nrow = I)
-manifold=function(tau){ return(c(tau,sin(tau+pi/2))) }
-for(i in 1:I){ X[i,]=manifold(t[i]) }
-data.points=X+cbind(e1,e2)
-
-# Apply PME()
-ptm <- proc.time()
-result=pme(x.obs=data.points, d=1)
-proc.time() - ptm
-
-# Plot the fitted manifold
-f=result2$embedding.map
-t.test=seq(from=-100,to=100,by=0.05)
-t.length=length(t.test)
-x.test=matrix(NA,ncol=2,nrow = t.length)
-for(i in 1:t.length){ x.test[i,]=f(t.test[i]) }
-plot(data.points[,1], data.points[,2],
-     pch=20, col="grey",
-     main="Principal Manifold Estimation",
-     xlab=" ", ylab=" ")
-lines(x.test[,1],x.test[,2],col="red",type = "l",lwd=3)
-
-ggplot() +
-  geom_point(aes(x = data.points[, 1], y = data.points[, 2])) +
[…]

### 2023-07-04 Principal_Manifold_Estimation.R

@@ -1,500 +1,0 @@
-#########################################################################
-##########    Principal Manifold Estimation (PME) Algorithm   ###########
-##########                  Kun (Michael) Meng                ###########
-##########    Department of Biostatistics, Brown University   ###########
-#########################################################################
-
-
-########### Introduction ###########################################
-####################################################################
-
-# We propose an R-software function for d-dimensional estimating principle manifolds, 
-# where d = 1, 2, 3, and manifolds embedded into D-dimensional Euclidean space with D 
-# strictly larger than d.
-
-# The proposed function is "PME()." To construct this function, we need some functions 
-# for "high dimensional mixture density estimation (HDMDE)."
-
-# This manuscript - whatever its name is - is organized as follows.
-#   (i)   In Section 1, we define some basic functions.
-#   (ii)  In Section 2, we construct a function for HDMDE. In this section, we first construct 
-#         the "weight.seq" function. Then we apply this function to construct "hdmde" function.
-#   (iii) Based on this "hdmde" function, "PME" estimation function is finally given in Section 3.
-# The theoretical foundation of this R-software function is the following paper.
-#
-# K. Meng and A. Eloyan, Principal Manifolds: A framework Using Sobolev Spaces and Model
-#                        Complexity Selection Using Mixture Densities
-
-########### Section 0, Packages ####################################
-####################################################################
-
-#install.packages("MASS")
-library("MASS")
-#install.packages("Matrix")
-library("Matrix")
-#install.packages("vegan")
-library("vegan")
-#install.packages("plot3D")
-library("plot3D")
-
-
-
-########### Section 1, Some Basic Functions ########################
-####################################################################
-
-## Subsection 1.1, Functions for Euclidean metrics
-
-# Norm function in an Euclidean space of any dimension
[…]

### 2023-07-04 process_adnitau.R

@@ -1,61 +1,0 @@
-library(oro.dicom)
-library(oro.nifti)
-library(tidyverse)
-library(tools)
-library(svMisc)
-
-adni_tau_path <- "data/ADNITau/"
-
-id_dirs <- list.files(adni_tau_path, full.names = TRUE)
-
-for (dir_idx in 1:length(id_dirs)) {
-  id_dir <- id_dirs[dir_idx]
-  subdirs <- list.dirs(id_dir)
-  num_nested <- grep("//", subdirs)
-  deepest <- subdirs[num_nested == max(num_nested)]
-  for (subdir in deepest) {
-    file_names <- list.files(subdir, full.names = TRUE)
-    if (sum(file_ext(file_names) == "dcm") == length(file_names)) {
-      all_slices <- readDICOM(subdir)
-      nii <- tryCatch(
-        dicom2nifti(all_slices),
-        error = function(e) {
-          print(paste0("ERROR in directory: ", id_dir))
-          print("")
-        }
-      )
-      if (class(nii) != "character") {
-        fname <- subdir %>% 
-          gsub(
-            pattern = "/",
-            replacement = "_"
-          ) %>% 
-          gsub(
-            pattern = "data_",
-            replacement = ""
-          ) %>% 
-          gsub(
-            pattern = "ADNITau",
-            replacement = "ADNI"
-          ) %>% 
-          gsub(
-            pattern = "__",
-            replacement = "_"
-          ) %>% 
-          gsub(
-            pattern = "\\.",
-            replacement = "_"
-          )
-        
-        writeNIfTI(
-          nim = nii, 
-          filename = paste(subdir, fname, sep = "/")
-        )
-        
-        file.remove(file_names)
-      }
-    }
-  }
-  progress(dir_idx, max.value = length(id_dirs))
-}
-

### 2023-07-04 paper/main.tex
add data augmentation figure

@@
 \usepackage{graphicx}
+\graphicspath{{figures/}}
 \usepackage[perpage,symbol*]{footmisc}
@@
 \usepackage{tikz}
-\usepackage{geometry}
-\geometry{
-  left=10mm,
-  right=10mm
-  }
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 The embedding functions used to define each manifold are given in Table \ref{table:1}.
-In these functions, the values of $\zeta$, $\alpha$, and $\beta$, as well as the functional form of $g(\cdot)$ are varied between trials.
-The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance varied between trials.
+In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials.
+The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
+Meanwhile, in cases where $d=1$, $\alpha$ and $\beta$ are scalars that are set externally.
+If $d>1$, then $\alpha$ and $\beta$ take the form of vectors of length $d$, with the values drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
 
@@
 In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
-Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$, when the spline smoothing method is used and $\lambda_g = g + 0.5$ for $g = 0, \dots, 25$ when the Gaussian process smoothing method is used.
+Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$.
 In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
@@
 This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
-This process is illustrated graphically below [Figure still needs to be added].
+This process is illustrated graphically in Figure \ref{fig:1}.
+
+\begin{figure}%
+  \centering
+  \subfloat[\centering $D = 2$]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
+  \hfill
+  \subfloat[\centering $D = 3$]{{\includegraphics[width=9cm]{unit_circle_D3}}}
+  \caption{Augmenting the Unit Circle}
+  \label{fig:1}
+\end{figure}
 

### 2023-07-05 code/lpme_simulations.R
update principal surface indexing

@@ -449,7 +449,7 @@ sim_error_case3 <- function(max_time, interval, amp_noise, shape_noise, time_cha
       curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
       curve_error[smoother_idx] <- map(
         1:nrow(true_vals[true_vals[, 1] == time_vals[t], ]),
-        ~ dist_euclidean(true_vals[true_vals[, 1] == time_vals[t], ][.x, ], curve_vals[[smoother_idx]][.x, ])^2
+        ~ dist_euclidean(true_vals[true_vals[, 1] == time_vals[t], ][.x, ], curve_vals[[smoother_idx]][.x, 1:3])^2
       ) %>%
         unlist() %>%
         mean()
@@ -1207,8 +1207,8 @@ sim_error_case7 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     ) %>%
       unlist()
     opt_surface <- which.min(surface_mse)
-    principal_curve_result[[t]] <- principal_surface[[opt_surface]]
-    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface]]$PS)
+    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
+    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
   }
   pme_vals <- reduce(pme_vals, rbind)
   principal_curve_vals <- reduce(principal_curve_vals, rbind)
@@ -1390,8 +1390,8 @@ sim_error_case8 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     ) %>%
       unlist()
     opt_surface <- which.min(surface_mse)
-    principal_curve_result[[t]] <- principal_surface[[opt_surface]]
-    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface]]$PS)
+    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
+    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
   }
   pme_vals <- reduce(pme_vals, rbind)
   principal_curve_vals <- reduce(principal_curve_vals, rbind)
@@ -1574,8 +1574,8 @@ sim_error_case9 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     ) %>%
       unlist()
[…]

### 2023-07-05 paper/main.tex
update simulation description

@@
 For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.
+
 The embedding functions used to define each manifold are given in Table \ref{table:1}.
-In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials.
-The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
-Meanwhile, in cases where $d=1$, $\alpha$ and $\beta$ are scalars that are set externally.
-If $d>1$, then $\alpha$ and $\beta$ take the form of vectors of length $d$, with the values drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
+In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent change in the underlying manifold over time.
+The function $g(\cdot)$ may be constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
+The values of $\iota$ represent noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
+Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
+These values describe random fluctuations in the manifold between time points.
 
@@
     \hline
-    Case & $d$ & $D$ & $f(\mathbf{r})$ & Domain \\
+    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
     \hline

### 2023-07-05 paper/main.tex
Updates from Overleaf

--- 2023-07-03 paper/main.tex
@@
 \usepackage{tikz}
-\usepackage{geometry}
-\geometry{
-  left=10mm,
-  right=10mm
-  }
 
@@
 \setlength{\oddsidemargin}{-0.5cm}
-\setlength{\evensidemargin}{-0.5cm}
 \renewcommand{\topmargin}{-2cm}
@@
 \title{Longitudinal Principal Manifold Estimation}
-\author[1]{Robert Zielinski, Kun Meng, and Ani Eloyan}
+\author[1]{Robert Zielinski}
+\author[2]{Kun Meng}
+\author[1]{Ani Eloyan}
+\affil[1]{Department of Biostatistics, Brown University}
+\affil[2]{Division of Applied Mathematics, Brown University}
 
@@
 
-Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}).
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}).
 Frequently, interest centers around longitudinal changes in one or more neurological substructures.
-For example, it is common to observe atrophy the hippocampus of those with Alzheimer's disease or cognitive impairment related to other causes.
-Image segmentation allows for the extraction of subcortical structures from images of the complete brain for further analysis.
+These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}).
+For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.
+Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for further analysis.
 
-Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach and is the gold standard.
+Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
 However, this approach is highly time and resource intensive.
@@
 
-To assess the performance of the LPME algorithm, we compare its performance with the that of two alternative approaches on simulated datasets.
+To assess the performance of the LPME algorithm, we compare its performance with the performance of two alternative approaches on simulated datasets.
 We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
 For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.
-The embedding functions used to define each manifold are given in Table \ref{table:1}.
-In these functions, the values of $\zeta$, $\alpha$, and $\beta$, as well as the functional form of $g(\cdot)$ are varied between trials.
-The values of $\iota$ are randomly sampled from a normal distribution with mean zero and variance varied between trials.
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c c c c c|}
-    \hline
-    Case & $d$ & $D$ & $f(\mathbf{r})$ & Domain \\
-    \hline
-    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
-    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
-    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{cos} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
-    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
-    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
-    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
-    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
-    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
-    9 & 2 & 3 & $\left(\theta\text{sin} \ (\beta_1r_1)\text{cos} \ (\beta_2r_2), \ \theta\text{sin} \ (\beta_1)\text{sin} \ (\beta_2r_2), \ \theta\text{cos} \ (\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
-    \hline
-  \end{tabular}
-  \caption{Embedding functions used for simulation trials.}
-  \label{table:1}
-\end{table}
 

### 2023-07-05 paper/main.tex
update simulation description

--- 2023-07-05 paper/main.tex
@@
 Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
-These values describe random fluctuations in the manifold between time points.
+These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.
 

### 2023-07-06 paper/main.tex
add simulation result table framework

@@
 \newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}
+
+\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}
 
@@
 
-\subsection{Alogirthm}
+\subsection{Algorithm}
 
@@
 For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels.
-The taste of the GPR is similar to \cite{dunson2021inferring}, and we may cite \cite{dunson2021inferring}.}
+The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
 
@@
 
-\meng{FYI, to fit a circle-like manifold, you may need to refer to the following papers: \cite{fasy2014confidence}, \cite{dunson2021inferring}.}
+\meng{FYI, to fit a circle-like manifold, you may need to refer to the following papers: \cite{fasy2014confidence}, \cite{dunsonInferringManifoldsNoisy2022}.}
 
@@
 
-To assess the performance of the LPME algorithm, we compare its performance with the that of two alternative approaches on simulated datasets.
+To assess the performance of the LPME algorithm, we compare its performance with that of two alternative approaches on simulated datasets.
 We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
@@
 
-The embedding functions used to define each manifold are given in Table \ref{table:1}.
-In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent change in the underlying manifold over time.
-The function $g(\cdot)$ may be constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
-The values of $\iota$ represent noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
+The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
+In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural change in the underlying manifold over time.
+The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
+The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
 Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
@@
   \caption{Embedding functions used for simulation trials.}
-  \label{table:1}
+  \label{table:simulation_embeddings}
 \end{table}
@@
 
-Notably, the manifolds estimated in the simulated datasets included several connected manifolds, which methods based on the principal curves, including the principal manifold framework used here, are typically unable to fit due to the self-consistency property discussed above.
+Notably, the manifolds estimated in the simulated datasets included several self-intersecting manifolds, which methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit due to the self-consistency property discussed above.
 To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$.
@@
 This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
-This process is illustrated graphically in Figure \ref{fig:1}.
+This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.
 
@@
   \caption{Augmenting the Unit Circle}
-  \label{fig:1}
+  \label{fig:unit_circle_augmentation}
+\end{figure}
+
+\zielinski{The simulation results shown below use partial results run on my computer.}
+
+Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case5} and \ref{fig:sim_case8}.
+Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
+Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.
+
+\zielinski{It may be necessary to put some of the simulation figures in an appendix.}
+
+A factorial design is used to run the simulation studies, with the factor levels set as follows:
+\begin{itemize}
+  \item $\alpha, \beta, \zeta \in \left\{0, 0.5, 1.0, 1.5, 2.0\right\}$
+  \item Study duration: $\left\{1, 2, 5, 10\right\}$
+  \item Interval between images: $\left\{0.1, 0.25, 0.5, 1\right\}$
+  \item Sample size: $\left\{1000, 10000\right\}$
+\end{itemize}
+
+Thus, each embedding map is simulated a total of 16,000 times. The results of these simulations are summarized in Table \ref{table:simulation_results}.
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c c c c c c c c c c|}
+    \hline
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
+    \hline
+    \hline
+  \end{tabular}
+  \caption{MSD Comparison.}
+  \label{table:simulation_results}
+\end{table}
+
+\begin{figure}%
+  \centering
+  \includegraphics[width=\textwidth]{sim_case1}
+  \caption{Simulation Case 1}
+  \label{fig:sim_case1}
 \end{figure}

### 2023-07-06 code/lpme_simulations.R
update factor levels to reduce number of runs

@@ -1869,14 +1869,14 @@ sim_error_case10 <- function(max_time, interval, amp_noise, shape_noise, time_ch
   return(0)
 }
 
-amp_noise_vals <- seq(0, 2, 0.25)
-shape_noise_vals <- seq(0, 2, 0.25)
-max_times <- c(1, 2, 5, 8, 10)
+amp_noise_vals <- seq(0, 2, 0.5)
+shape_noise_vals <- seq(0, 2, 0.5)
+max_times <- c(1, 2, 5, 10)
 intervals <- c(0.1, 0.25, 0.5, 1)
 n_vals <- 10^(3:4)
-replicates <- 1:4
+replicates <- 1
 case <- 1:10
-time_changes <- seq(0, 2, 0.25)
+time_changes <- seq(0, 2, 0.5)
 time_trends <- c("constant", "linear", "quadratic", "sinusoidal")
 
 param_grid <- expand.grid(

### 2023-07-06 code/generate_figures.R
add initial code to generate all relevant figures

@@ -1,0 +1,297 @@
+library(plot3D)
+library(pracma)
+library(tidyverse)
+
+set.seed(1202)
+theta <- runif(1000, min = 0, max = 2 * pi)
+x <- cos(theta)
+y <- sin(theta)
+
+unit_df <- data.frame(theta, x, y) %>%
+  arrange(theta)
+angle <- cart2pol(as.matrix(unit_df[, 2:3], ncol = 2))
+
+
+png("paper/figures/unit_circle_augmentation.png")
+par(mfrow = c(1, 2))
+plot(
+  x = unit_df$x,
+  y = unit_df$y,
+  type = "l",
+  asp = 1,
+  main = "Unit Circle, D = 2",
+  xlab = "x",
+  ylab = "y"
+)
+
+plt2 <- lines3D(
+  x = unit_df$x,
+  y = unit_df$y,
+  z = angle[, 1],
+  main = "Unit Circle, D = 3",
+  xlab = "x",
+  ylab = "y",
+  zlab = "theta"
+)
+dev.off()
+
+png("paper/figures/unit_circle_D2.png")
+plot(
+  x = unit_df$x,
+  y = unit_df$y,
+  type = "l",
+  asp = 1,
+  main = "Unit Circle, D = 2",
+  xlab = "x",
+  ylab = "y"
+)
+dev.off()
+
+png("paper/figures/unit_circle_D3.png")
+lines3D(
+  x = unit_df$x,
+  y = unit_df$y,
+  z = angle[, 1],
+  main = "Unit Circle, D = 3",
+  xlab = "x",
+  ylab = "y",
+  zlab = "theta"
+)
+dev.off()
+
+sim_example_case1 <- readRDS("simulations/case1/duration_05_interval_010_ampnoise_125_pernoise_000_n_1000_constant025_run_03.rds")
+sim_example_case5 <- readRDS("simulations/case5/duration_05_interval_100_ampnoise_175_pernoise_125_n_1000_quadratic000_run_04.rds")
+sim_example_case8 <- readRDS("simulations/case8/duration_02_interval_025_ampnoise_125_pernoise_025_n_1000_sinusoidal200_run_02.rds")
+
+df_labels <- c("Data", "True", "LPME", "PME", "Principal Curve")
+
+data_case1 <- list(
+  sim_example_case1$df,
+  sim_example_case1$true_vals,
+  sim_example_case1$lpme_vals,
+  sim_example_case1$pme_vals,
+  sim_example_case1$principal_curve_vals
+)
+
[…]

### 2023-07-06 paper/main.tex
add initial figures for simulation cases 4 and 7

@@
 
-Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case5} and \ref{fig:sim_case8}.
-Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
-Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.
-
-\zielinski{It may be necessary to put some of the simulation figures in an appendix.}
-
 A factorial design is used to run the simulation studies, with the factor levels set as follows:
@@
 
-Thus, each embedding map is simulated a total of 16,000 times. The results of these simulations are summarized in Table \ref{table:simulation_results}.
+Thus, each embedding map is simulated a total of 16,000 times.
+The results of these simulations are summarized in Table \ref{table:simulation_results}.
+All code required to reproduce the simulation results and the associated figures is provided at \texttt{https://github.com/rjzielinski/lpme-project}.
+
+Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case4} and \ref{fig:sim_case7}.
+Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
+Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.
+
+\zielinski{It may be necessary to put some of the simulation figures in an appendix.
+Additionally, the figures are still very much a work in progress and are serving more of a placeholder status at the moment.
+Any suggestions regarding how to improve the clarity and depth perception of the figures would be greatly appreciated!}
 
@@
     \hline
+    Data & & & & & & & & & \\
+    LPME & & & & & & & & & \\
+    PME & & & & & & & & & \\
+    PC/PS & & & & & & & & & \\
     \hline
@@
 \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{sim_case5}
+  \caption{Simulation Case 4}
+  \label{fig:sim_case4}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{sim_case8}
+  \caption{Simulation Case 7}
+  \label{fig:sim_case7}
+\end{figure}
+
+\zielinski{A brief discussion of the simulation results will follow when they become available.}
 

### 2023-07-06 code/generate_figures.R
add initial figures for simulation cases 4 and 7

@@ -1,5 +1,6 @@
 library(plot3D)
 library(pracma)
+library(scatterplot3d)
 library(tidyverse)
 
 set.seed(1202)
@@ -237,27 +238,28 @@ for (time_idx in 1:length(time_vals)) {
   # temp_pme <- sim_example_case1$pme_vals[temp_data[, 1] == time_vals[time_idx], ]
   # temp_princurve <- sim_example_case1$principal_curve_vals[temp_data[, 1] == time_vals[time_idx], ]
   # temp_true <- sim_example_case1$true_vals[temp_data[, 1] == time_vals[time_idx], ]
-  points3D(
+
+  plt <- scatterplot3d(
     x = temp_data$x,
     y = temp_data$y,
     z = temp_data$z,
     xlab = "x",
     ylab = "y",
     zlab = "z",
-    col = alpha("black", 0.4),
-    bg = alpha("black", 0.4),
     pch = 21,
+    color = alpha("black", 0.5),
+    bg = alpha("black", 0.5),
     main = paste0("Time = ", time_vals[time_idx])
   )
-  points3D(
+  plt$points3d(
     x = temp_true$x,
     y = temp_true$y,
     z = temp_true$z,
-    col = "red",
-    bg = "red",
+    col = alpha("red", 0.5),
+    bg = alpha("red", 0.5),
     pch = 21
   )
-  points3D(
+  plt$points3d(
     x = temp_lpme$x,
     y = temp_lpme$y,
     z = temp_lpme$z,
@@ -265,16 +267,145 @@ for (time_idx in 1:length(time_vals)) {
     bg = "blue",
     pch = 21
   )
-  points(
+  plt$points3d(
     x = temp_pme$x,
     y = temp_pme$y,
+    z = temp_pme$z,
     col = "green",
     bg = "green",
     pch = 21
   )
-  points(
+  plt$points3d(
     x = temp_princurve$x,
     y = temp_princurve$y,
+    z = temp_princurve$z,
+    col = "purple",
+    bg = "purple",
+    pch = 21
+  )
+}
+par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
+plot(0, 0, type = "l", bty = "n", xaxt = "n", yaxt = "n")
+legend(
+  "bottom",
+  c("Data", "True Manifold", "LPME", "PME", "Principal Curve"),
[…]

### 2023-07-07 paper/main.tex
include application data description

@@
 
-To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
+To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
 In a larger comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
@@
 
+As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images.
+This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates.
+To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).
+
+While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}).
+Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals.
+Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected.
+Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process.
+Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
+Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}.
+Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation.
+
 \section{Discussion}

### 2023-07-07 paper/main.tex
update application section

@@
 Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning.
-Details of this approach are described below.
+Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
 
@@
 Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
+
 Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}.
 Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation.
+Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
+The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.
+
+Provide a brief summary of the results of the model fitting here, along with the related figures.
 

### 2023-07-07 code/lpme_simulations.R
export functions and packages to parallel nodes

@@ -11,17 +11,18 @@ library(princurve)
 source("code/functions/sim_data.R")
 source("code/functions/calc_pme_est.R")
 source("code/functions/calc_lpme_est.R")
+source("code/prinSurf_v3.R")
 
 ### Simulation Case 1
 
 sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, run = 1, print_plots = FALSE) {
-  require(tidyverse)
-  require(pme)
-  require(princurve)
-  require(plotly)
-  source("code/functions/sim_data.R")
-  source("code/functions/calc_pme_est.R")
-  source("code/functions/calc_lpme_est.R")
+  # require(tidyverse)
+  # require(pme)
+  # require(princurve)
+  # require(plotly)
+  # source("code/functions/sim_data.R")
+  # source("code/functions/calc_pme_est.R")
+  # source("code/functions/calc_lpme_est.R")
   time_vals <- seq(0, max_time, interval)
   sim_list <- lapply(
     time_vals,
@@ -100,22 +101,33 @@ sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_cha
     z = lpme_vals[, 1],
     type = "scatter3d",
     mode = "markers",
-    marker = list(size = 3)
+    marker = list(size = 3),
+    name = "LPME"
   ) %>%
     add_markers(
       x = pme_vals[, 2],
       y = pme_vals[, 3],
-      z = pme_vals[, 1]
+      z = pme_vals[, 1],
+      name = "PME"
     ) %>%
     add_markers(
       x = principal_curve_vals[, 2],
       y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 1]
+      z = principal_curve_vals[, 1],
+      name = "Principal Curve"
     ) %>%
     add_markers(
       x = true_vals[, 2],
       y = true_vals[, 3],
-      z = true_vals[, 1]
+      z = true_vals[, 1],
+      name = "True Manifold"
+    ) %>%
+    add_markers(
+      x = sim_df[, 2],
+      y = sim_df[, 3],
+      z = sim_df[, 1],
+      name = "Data",
+      opacity = 0.2
     )
 
   pme_error <- map(
@@ -198,13 +210,13 @@ sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_cha
 ### Simulation Case 2
 
[…]

### 2023-07-12 paper/main.tex
update simulation description

@@
 \begin{itemize}
-  \item $\alpha, \beta, \zeta \in \left\{0, 0.5, 1.0, 1.5, 2.0\right\}$
+  \item $\alpha, \beta, \zeta \in \left\{0, 0.1, 0.25, 0.5, 1.0, 2.0\right\}$
   \item Study duration: $\left\{1, 2, 5, 10\right\}$
@@
   \item Sample size: $\left\{1000, 10000\right\}$
+  \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
 \end{itemize}
 
-Thus, each embedding map is simulated a total of 16,000 times.
-The results of these simulations are summarized in Table \ref{table:simulation_results}.
+Thus, each embedding map is simulated a total of 27,648 times.
 All code required to reproduce the simulation results and the associated figures is provided at \texttt{https://github.com/rjzielinski/lpme-project}.
 
-Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case4} and \ref{fig:sim_case7}.
+Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case4} and \ref{fig:sim_case6}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
     \hline
-    Data & & & & & & & & & \\
-    LPME & & & & & & & & & \\
-    PME & & & & & & & & & \\
-    PC/PS & & & & & & & & & \\
+    Data & 0.758 (1.28) & 0.884 (0.726) & N/A & 198 (466) & 1.91 (1.76) & 13.4 (51.4) & N/A & N/A & N/A \\
+    LPME & 4.55 (33.4) & 0.483 (0.471) & N/A & 198 (843) & 1.45 (1.34) & 8.99 (31.7) & N/A & N/A & N/A \\
+    PME & 1.01 (2.01) & 3.38 (12.1) & N/A & 194 (461) & 4.49 (6.26) & 12.4 (41.9) & N/A & N/A & N/A \\
+    PC/PS & 0.688 (1.23) & 0.544 (0.276) & N/A & 143 (379) & 1.37 (1.26) & 5.58 (19.6) & N/A & N/A & N/A \\
     \hline
   \end{tabular}
-  \caption{MSD Comparison.}
-  \label{table:simulation_results}
+  \caption{MSD Comparison, Mean (SD).}
+  \label{table:simulation_results_mean}
+\end{table}
+
+\zielinski{With results still incomplete, it seems like the worst LPME performance is concentrated in cases where the structure of the manifold changes drastically between time points.
+I have included these cases in the simulations to test where the method breaks down.
+For reporting purposes, would it be worth it to limit ourselves to more realistic cases without quite so much noise?
+Also, would it be worthwhile to limit the number of simulation cases we are considering for a more focused assessment?}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c c c c c c c c c c|}
+    \hline
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
+    \hline
+    Data & 0.244 (0.576) & 0.784 (0.585) & N/A & 14.3 (84.0) & 1.63 (1.85) & 0.870 (4.49) & N/A & N/A & N/A \\
+    LPME & 0.135 (0.265) & 0.406 (0.337) & N/A & 8.30 (35.6) & 1.21 (1.18) & 0.802 (2.85) & N/A & N/A & N/A \\
+    PME & 0.351 (0.814) & 0.980 (1.61) & N/A & 12.3 (65.9) & 2.50 (3.56) & 0.883 (4.17) & N/A & N/A & N/A \\
+    PC/PS & 0.201 (0.554) & 0.529 (0.215) & N/A & 7.66 (41.2) & 1.15 (1.13) & 0.867 (1.90) & N/A & N/A & N/A \\
+    \hline
+  \end{tabular}
+  \caption{MSD Comparison, Median (IQR).}
+  \label{table:simulation_results_median}
 \end{table}
@@
   \centering
-  \includegraphics[width=\textwidth]{sim_case8}
-  \caption{Simulation Case 7}
-  \label{fig:sim_case7}
+  \includegraphics[width=\textwidth]{sim_case7}
+  \caption{Simulation Case 6}
+  \label{fig:sim_case6}
 \end{figure}
 
-\zielinski{A brief discussion of the simulation results will follow when they become available.}
+The mean, median, and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
 

### 2023-07-12 code/generate_figures.R
update simulation case 1 figure

@@ -60,9 +60,9 @@ lines3D(
 )
 dev.off()
 
-sim_example_case1 <- readRDS("simulations/case1/duration_05_interval_010_ampnoise_125_pernoise_000_n_1000_constant025_run_03.rds")
-sim_example_case5 <- readRDS("simulations/case5/duration_05_interval_100_ampnoise_175_pernoise_125_n_1000_quadratic000_run_04.rds")
-sim_example_case8 <- readRDS("simulations/case8/duration_02_interval_025_ampnoise_125_pernoise_025_n_1000_sinusoidal200_run_02.rds")
+sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_050_pernoise_050_n_1000_constant000_run_01.rds")
+sim_example_case5 <- readRDS("simulations/case5/duration_01_interval_010_ampnoise_050_pernoise_000_n_10000_constant000_run_01.rds")
+sim_example_case7 <- readRDS("simulations/case7/duration_01_interval_010_ampnoise_050_pernoise_050_n_1000_constant000_run_01.rds")
 
 df_labels <- c("Data", "True", "LPME", "PME", "Principal Curve")
 
@@ -82,10 +82,10 @@ for (df_idx in 1:length(data_case1)) {
 }
 df_case1 <- reduce(data_case1, rbind)
 df_case1 <- df_case1 %>%
-  filter(mod(time, 1) == 0)
+  filter(mod(time * 10, 2) == 0)
 
 time_vals <- sim_example_case1$times
-time_vals <- time_vals[mod(time_vals, 1) == 0]
+time_vals <- time_vals[mod(time_vals * 10, 2) == 0]
 
 png("paper/figures/sim_case1.png")
 par(oma = c(4, 1, 1, 1), mfrow = c(2, 3), mar = c(2, 2, 1, 1))
@@ -132,6 +132,8 @@ for (time_idx in 1:length(time_vals)) {
     ylab = "y",
     col = alpha("black", 0.4),
     bg = alpha("black", 0.4),
+    xlim = c(-4, 4),
+    ylim = c(-3, 3),
     pch = 21,
     main = paste0("Time = ", time_vals[time_idx])
   )
@@ -195,14 +197,17 @@ for (df_idx in 1:length(data_case5)) {
 }
 df_case5 <- reduce(data_case5, rbind)
 df_case5 <- df_case5 %>%
-  filter(mod(time, 1) == 0)
[…]

### 2023-07-26 paper/main.tex
update new figures and wording

@@
     \hline
-    Data & 0.758 (1.28) & 0.884 (0.726) & N/A & 198 (466) & 1.91 (1.76) & 13.4 (51.4) & N/A & N/A & N/A \\
-    LPME & 4.55 (33.4) & 0.483 (0.471) & N/A & 198 (843) & 1.45 (1.34) & 8.99 (31.7) & N/A & N/A & N/A \\
-    PME & 1.01 (2.01) & 3.38 (12.1) & N/A & 194 (461) & 4.49 (6.26) & 12.4 (41.9) & N/A & N/A & N/A \\
-    PC/PS & 0.688 (1.23) & 0.544 (0.276) & N/A & 143 (379) & 1.37 (1.26) & 5.58 (19.6) & N/A & N/A & N/A \\
+    Data & 0.240 (0.346) & 0.489 (0.399) & N/A & 19.2 (48.7) & 0.906 (0.735) & 1.20 (3.16) & N/A & N/A & N/A \\
+    LPME & 0.166 (0.668) & 0.324 (0.327) & N/A & 11.8 (31.1) & 0.633 (0.692) & 1.11 (2.61) & N/A & N/A & N/A \\
+    PME & 0.265 (0.430) & 0.615 (0.850) & N/A & 19.2 (48.7) & 0.918 (0.831) & 1.26 (3.17) & N/A & N/A & N/A \\
+    PC/PS & 0.209 (0.342) & 0.557 (0.319) & N/A & 19.2 (48.7) & 0.863 (0.670) & 0.965 (1.17) & N/A & N/A & N/A \\
     \hline
   \end{tabular}
-  \caption{MSD Comparison, Mean (SD).}
+  \caption{MSD Comparison to True Values, Mean (SD).}
   \label{table:simulation_results_mean}
 \end{table}
-
-\zielinski{With results still incomplete, it seems like the worst LPME performance is concentrated in cases where the structure of the manifold changes drastically between time points.
-I have included these cases in the simulations to test where the method breaks down.
-For reporting purposes, would it be worth it to limit ourselves to more realistic cases without quite so much noise?
-Also, would it be worthwhile to limit the number of simulation cases we are considering for a more focused assessment?}
 
@@
     \hline
-    Data & 0.244 (0.576) & 0.784 (0.585) & N/A & 14.3 (84.0) & 1.63 (1.85) & 0.870 (4.49) & N/A & N/A & N/A \\
-    LPME & 0.135 (0.265) & 0.406 (0.337) & N/A & 8.30 (35.6) & 1.21 (1.18) & 0.802 (2.85) & N/A & N/A & N/A \\
-    PME & 0.351 (0.814) & 0.980 (1.61) & N/A & 12.3 (65.9) & 2.50 (3.56) & 0.883 (4.17) & N/A & N/A & N/A \\
-    PC/PS & 0.201 (0.554) & 0.529 (0.215) & N/A & 7.66 (41.2) & 1.15 (1.13) & 0.867 (1.90) & N/A & N/A & N/A \\
+    Data & 0.134 (0.232) & 0.382 (0.649) & N/A & 3.37 (15.7) & 0.729 (1.29) & 0.328 (0.742) & N/A & N/A & N/A \\
+    LPME & 0.0661 (0.119) & 0.236 (0.438) & N/A & 3.10 (9.54) & 0.370 (0.889) & 0.364 (0.671) & N/A & N/A & N/A \\
+    PME & 0.117 (0.281) & 0.463 (0.668) & N/A & 3.34 (15.6) & 0.675 (1.31) & 0.369 (0.680) & N/A & N/A & N/A \\
+    PC/PS & 0.110 (0.214) & 0.471 (0.467) & N/A & 3.33 (15.7) & 0.691 (1.18) & 0.607 (0.329) & N/A & N/A & N/A \\
     \hline
   \end{tabular}
-  \caption{MSD Comparison, Median (IQR).}
+  \caption{MSD Comparison to True Values, Median (IQR).}
   \label{table:simulation_results_median}
@@
 The mean, median, and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
+Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
+Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach should result in lower mean squared distance values.
+
+The result summaries indicate that in most cases, LPME provides a substantial improvement in performance over the PME and principal curve approaches when estimating the underlying manifold.
+As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates, shown in blue, remain relatively stable over time.
+This contrasts to the estimates found by the PME and principal curve approaches, which are highly sensitive to the data observed at each given time point, as expected.
+This ultimately results in the LPME estimates remaining closer to the true underlying manifold, shown in red.
+Similar results hold in most of the other simulation cases.
+
+\zielinski{It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on.
+This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold.
+In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point.
+
+This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold.
+When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.}
 

### 2023-07-26 code/generate_figures.R
udpate figure axes

@@ -60,9 +60,9 @@ lines3D(
 )
 dev.off()
 
-sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_050_pernoise_050_n_1000_constant000_run_01.rds")
-sim_example_case5 <- readRDS("simulations/case5/duration_01_interval_010_ampnoise_050_pernoise_000_n_10000_constant000_run_01.rds")
-sim_example_case7 <- readRDS("simulations/case7/duration_01_interval_010_ampnoise_050_pernoise_050_n_1000_constant000_run_01.rds")
+sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_010_pernoise_050_n_1000_constant000_run_01.rds")
+sim_example_case5 <- readRDS("simulations/case6/duration_01_interval_010_ampnoise_000_pernoise_005_n_1000_constant000_run_01.rds")
+sim_example_case7 <- readRDS("simulations/case7/duration_01_interval_010_ampnoise_010_pernoise_010_n_1000_constant000_run_01.rds")
 
 df_labels <- c("Data", "True", "LPME", "PME", "Principal Curve")
 
@@ -133,37 +133,42 @@ for (time_idx in 1:length(time_vals)) {
     col = alpha("black", 0.4),
     bg = alpha("black", 0.4),
     xlim = c(-4, 4),
-    ylim = c(-3, 3),
+    ylim = c(-2, 2),
     pch = 21,
-    main = paste0("Time = ", time_vals[time_idx])
+    main = paste0("Time = ", time_vals[time_idx]),
+    cex = 0.75
   )
   points(
     x = temp_true$x,
     y = temp_true$y,
     col = "red",
     bg = "red",
-    pch = 21
+    pch = 21,
+    cex = 0.5
   )
   points(
     x = temp_lpme$x,
     y = temp_lpme$y,
     col = "blue",
     bg = "blue",
-    pch = 21
+    pch = 21,
+    cex = 0.5
   )
   points(
     x = temp_pme$x,
     y = temp_pme$y,
     col = "green",
     bg = "green",
-    pch = 21
+    pch = 21,
+    cex = 0.5
   )
   points(
     x = temp_princurve$x,
     y = temp_princurve$y,
     col = "purple",
     bg = "purple",
[…]

### 2023-07-31 code/generate_figures.R
update figure legends

@@ -87,7 +87,7 @@ df_case1 <- df_case1 %>%
 time_vals <- sim_example_case1$times
 time_vals <- time_vals[mod(time_vals * 10, 2) == 0]
 
-png("paper/figures/sim_case1.png", res = 500, height = 2000, width = 2000)
+png("paper/figures/sim_case1.png", res = 500, height = 2000, width = 3000)
 par(oma = c(4, 1, 1, 1), mfrow = c(2, 3), mar = c(2, 2, 1, 1))
 for (time_idx in 1:length(time_vals)) {
   temp_data <- df_case1 %>%
@@ -179,14 +179,15 @@ for (time_idx in 1:length(time_vals)) {
     lwd = 1.25
   )
 }
-par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
+par(fig = c(0, 1, 0, 1), oma = c(0, 1, 0, 1), mar = c(0, 0, 0, 0), new = TRUE)
 plot(0, 0, type = "l", bty = "n", xaxt = "n", yaxt = "n")
 legend(
   "bottom",
   c("Data", "True Manifold", "LPME", "PME", "Principal Curve"),
   col = c("black", "red", "blue", "green", "purple"),
   # fill = c("black", "red", "blue", "green", "purple"),
-  lwd = 5,
+  lty = c("solid", "solid", "dashed", "dotted", "dotdash"),
+  lwd = 1.25,
   xpd = TRUE,
   horiz = TRUE,
   cex = 1,
@@ -215,7 +216,7 @@ df_case5 <- df_case5 %>%
 time_vals <- sim_example_case5$times
 time_vals <- time_vals[mod(time_vals * 10, 2) == 0]
 
-png("paper/figures/sim_case5.png", res = 500, height = 2000, width = 2000)
+png("paper/figures/sim_case5.png", res = 500, height = 2000, width = 3000)
 par(oma = c(4, 1, 1, 1), mfrow = c(2, 3), mar = c(2, 2, 1, 1))
 for (time_idx in 1:length(time_vals)) {
   temp_df <- df_case5 %>%
@@ -280,7 +281,8 @@ for (time_idx in 1:length(time_vals)) {
     col = alpha("red", 0.5),
     bg = alpha("red", 0.5),
     pch = 20,
-    cex = 0.33,
+    type = "l",
+    lwd = 1,
     add = TRUE
   )
   scatter3D(
[…]

### 2023-08-01 paper/main.tex
update simulation results
add adni hippocampus plots
update for new simulations

@@
 The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
-In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural change in the underlying manifold over time.
+In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural changes in the underlying manifold over time.
 The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
@@
 \begin{itemize}
-  \item $\alpha, \beta, \zeta \in \left\{0, 0.1, 0.25, 0.5, 1.0, 2.0\right\}$
-  \item Study duration: $\left\{1, 2, 5, 10\right\}$
-  \item Interval between images: $\left\{0.1, 0.25, 0.5, 1\right\}$
-  \item Sample size: $\left\{1000, 10000\right\}$
+  \item $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$
+  \item Study duration: $\left\{1, 2, 5\right\}$
+  \item Interval between images: $\left\{0.1, 0.25, 0.5\right\}$
   \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
@@
 
-Thus, each embedding map is simulated a total of 27,648 times.
+The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times.
 All code required to reproduce the simulation results and the associated figures is provided at \texttt{https://github.com/rjzielinski/lpme-project}.
@@
     \hline
-    Data & 0.240 (0.346) & 0.489 (0.399) & N/A & 19.2 (48.7) & 0.906 (0.735) & 1.20 (3.16) & N/A & N/A & N/A \\
-    LPME & 0.166 (0.668) & 0.324 (0.327) & N/A & 11.8 (31.1) & 0.633 (0.692) & 1.11 (2.61) & N/A & N/A & N/A \\
-    PME & 0.265 (0.430) & 0.615 (0.850) & N/A & 19.2 (48.7) & 0.918 (0.831) & 1.26 (3.17) & N/A & N/A & N/A \\
-    PC/PS & 0.209 (0.342) & 0.557 (0.319) & N/A & 19.2 (48.7) & 0.863 (0.670) & 0.965 (1.17) & N/A & N/A & N/A \\
+    Data & 0.223 (0.311) & 0.524 (0.462) & N/A & 25.5 (75.3) & 0.984 (0.754) & 2.00 (9.29) & N/A & N/A & N/A \\
+    LPME & 0.138 (0.174) & 0.363 (0.342) & N/A & 19.8 (69.1) & 0.751 (0.660) & 1.83 (9.19) & N/A & N/A & N/A \\
+    PME & 0.231 (0.412) & 0.871 (1.76) & N/A & 25.5 (75.3) & 1.11 (1.36) & 2.02 (9.29) & N/A & N/A & N/A \\
+    PC/PS & 0.189 (0.296) & 0.615 (0.357) & N/A & 25.5 (75.3) & 0.928 (0.702) & 1.25 (3.69) & N/A & N/A & N/A \\
     \hline
@@
     \hline
-    Data & 0.134 (0.232) & 0.382 (0.649) & N/A & 3.37 (15.7) & 0.729 (1.29) & 0.328 (0.742) & N/A & N/A & N/A \\
-    LPME & 0.0661 (0.119) & 0.236 (0.438) & N/A & 3.10 (9.54) & 0.370 (0.889) & 0.364 (0.671) & N/A & N/A & N/A \\
-    PME & 0.117 (0.281) & 0.463 (0.668) & N/A & 3.34 (15.6) & 0.675 (1.31) & 0.369 (0.680) & N/A & N/A & N/A \\
-    PC/PS & 0.110 (0.214) & 0.471 (0.467) & N/A & 3.33 (15.7) & 0.691 (1.18) & 0.607 (0.329) & N/A & N/A & N/A \\
+    Data & 0.124 (0.187) & 0.385 (0.640) & N/A & 4.11 (15.9) & 0.939 (1.26) & 0.272 (0.926) & N/A & N/A & N/A \\
+    LPME & 0.0707 (0.138) & 0.251 (0.511) & N/A & 3.23 (7.67) & 0.611 (1.05) & 0.290 (0.860) & N/A & N/A & N/A \\
+    PME & 0.105 (0.202) & 0.379 (0.708) & N/A & 4.06 (15.7) & 0.933 (1.32) & 0.295 (0.938) & N/A & N/A & N/A \\
+    PC/PS & 0.101 (0.177) & 0.504 (0.413) & N/A & 4.06 (15.9) & 0.830 (1.09) & 0.556 (0.317) & N/A & N/A & N/A \\
     \hline
@@
   \includegraphics[width=\textwidth]{sim_case5}
-  \caption{Simulation Case 4}
+  \caption{Simulation Case 5}
   \label{fig:sim_case4}
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{adni_lhipp}
+  \caption{Left Hippocampus, ADNI Participant}
+  \label{fig:adni_result}
+\end{figure}
+
 \section{Discussion}

### 2023-08-01 code/generate_figures.R

@@ -60,7 +60,7 @@ lines3D(
 )
 dev.off()
 
-sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_010_pernoise_050_n_1000_constant000_run_01.rds")
+sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_025_pernoise_050_n_1000_constant000_run_01.rds")
 sim_example_case5 <- readRDS("simulations/case6/duration_01_interval_010_ampnoise_000_pernoise_005_n_1000_constant000_run_01.rds")
 sim_example_case7 <- readRDS("simulations/case7/duration_01_interval_010_ampnoise_010_pernoise_010_n_1000_constant000_run_01.rds")
 

### 2023-08-04 paper/main.tex
update application section and add discussion

@@
     \hline
-    Data & 0.223 (0.311) & 0.524 (0.462) & N/A & 25.5 (75.3) & 0.984 (0.754) & 2.00 (9.29) & N/A & N/A & N/A \\
-    LPME & 0.138 (0.174) & 0.363 (0.342) & N/A & 19.8 (69.1) & 0.751 (0.660) & 1.83 (9.19) & N/A & N/A & N/A \\
-    PME & 0.231 (0.412) & 0.871 (1.76) & N/A & 25.5 (75.3) & 1.11 (1.36) & 2.02 (9.29) & N/A & N/A & N/A \\
-    PC/PS & 0.189 (0.296) & 0.615 (0.357) & N/A & 25.5 (75.3) & 0.928 (0.702) & 1.25 (3.69) & N/A & N/A & N/A \\
+    Data & 0.222 (0.261) & 0.513 (0.411) & N/A & 27.3 (74.9) & 0.986 (0.787) & 1.51 (6.04) & N/A & N/A & N/A \\
+    LPME & 0.129 (0.161) & 0.417 (0.960) & N/A & 19.0 (63.0) & 0.776 (0.777) & 1.27 (5.71) & N/A & N/A & N/A \\
+    PME & 0.247 (0.435) & 0.854 (1.62) & N/A & 27.3 (74.9) & 1.04 (1.07) & 1.55 (6.04) & N/A & N/A & N/A \\
+    PC/PS & 0.188 (0.247) & 0.600 (0.303) & N/A & 27.3 (74.9) & 0.939 (0.730) & 1.04 (2.20) & N/A & N/A & N/A \\
     \hline
@@
     \hline
-    Data & 0.124 (0.187) & 0.385 (0.640) & N/A & 4.11 (15.9) & 0.939 (1.26) & 0.272 (0.926) & N/A & N/A & N/A \\
-    LPME & 0.0707 (0.138) & 0.251 (0.511) & N/A & 3.23 (7.67) & 0.611 (1.05) & 0.290 (0.860) & N/A & N/A & N/A \\
-    PME & 0.105 (0.202) & 0.379 (0.708) & N/A & 4.06 (15.7) & 0.933 (1.32) & 0.295 (0.938) & N/A & N/A & N/A \\
-    PC/PS & 0.101 (0.177) & 0.504 (0.413) & N/A & 4.06 (15.9) & 0.830 (1.09) & 0.556 (0.317) & N/A & N/A & N/A \\
+    Data & 0.144 (0.222) & 0.466 (0.658) & N/A & 4.17 (19.4) & 0.890 (1.39) & 0.305 (1.07) & N/A & N/A & N/A \\
+    LPME & 0.0727 (0.129) & 0.250 (0.477) & N/A & 3.23 (11.3) & 0.575 (1.08) & 0.293 (0.907) & N/A & N/A & N/A \\
+    PME & 0.123 (0.249) & 0.516 (0.757) & N/A & 4.13 (19.2) & 0.897 (1.39) & 0.331 (1.04) & N/A & N/A & N/A \\
+    PC/PS & 0.112 (0.205) & 0.556 (0.427) & N/A & 4.12 (19.3) & 0.805 (1.26) & 0.567 (0.388) & N/A & N/A & N/A \\
     \hline
@@
 The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.
+Finally, the data were augmented with spherical coordinates in a manner similar to that described in the previous section to avoid the need to fit to a self-intersecting manifold.
 
-Provide a brief summary of the results of the model fitting here, along with the related figures.
+\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}
+
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}.
+Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data.
+There are two main sources of discrepancies between the data and the LPME estimate.
+First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation.
+The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.
+
+Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
+These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
+This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots.
+Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself.
+This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.
 
@@
   \includegraphics[width=\textwidth]{adni_lhipp}
-  \caption{Left Hippocampus, ADNI Participant}
+  \caption{Left Hippocampus, Cognitively Healthy ADNI Participant}
   \label{fig:adni_result}
+\end{figure}
+
+\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{adni_lhipp_cross_section}
+  \caption{Left Hippocampus Cross Section}
+  \label{fig:adni_cross_section}
 \end{figure}
@@
 
-\subsection*{Contributions}
+In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data.
+To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a longitudinal setting.
+In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point.
+This improvement was seen across several manifold types.
+We also suggest a data augmentation approach that may allow for the circumvention of the inability to fit principal curve-based methods to some self-intersecting manifolds.
 
-\subsection*{Limitations}
+While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations.
+First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly.
+Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that are too curvy.
+The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern.
+There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.
 
-\begin{enumerate}
-  \item Algorithm assumes stability in shape over time.
-  \item Algorithm initialization is sensitive to magnitude of change between time points.
-  \item Computation time.
-\end{enumerate}
+A second limitation is that the LPME method requires relative stability in the overall shape of the structure being fit over time.
+Thus, while this algorithm can be useful for separating meaningful changes in a manifold over time from inter-observation error, it will be unable to fit in situations where the error between observations is too substantial.
+How restrictive this requirement is depends largely on the area of application.
+When used to estimate subcortical surfaces as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.
 
-Future Work:
+Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset.
+It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus.
+This shortcoming could potentially be addressed in several ways.
+First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated.
+Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
+Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section 4.
+
+Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
+The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
+However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
+The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.
+While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
+
+\newpage
 

### 2023-08-04 code/adni_hipp_surface_test.R
update adni results

@@ -4,8 +4,9 @@ library(pracma)
 library(lubridate)
 library(scatterplot3d)
 library(tidyverse)
-source("code/pme.R")
-source("code/lpme_s3.R")
+library(pme)
+library(Rfast)
+
 source("code/functions/calc_pme_est.R")
 source("code/functions/calc_lpme_est.R")
 
@@ -42,6 +43,81 @@ create_scatter_mat <- function(mat, nrow = 1) {
   }
   scatter_grid <- plot_grid(plotlist = plots, nrow = nrow)
   return(scatter_grid)
+}
+
+create_cross_section_matrix <- function(mats, slice_idx, slice_widths, nrow = 1) {
+  time_vals <- unique(mats[[1]][, 1])
+  slices <- list()
+  for (time_idx in 1:length(time_vals)) {
+    slice_list <- list()
+    for (idx in 1:length(mats)) {
+      mat <- mats[[idx]]
+      temp_mat <- mat[mat[, 1] == time_vals[time_idx], ]
+      n_pixels <- length(unique(temp_mat[, slice_idx]))
+      slice_mean <- mean(temp_mat[, slice_idx])
+      slice_list[[idx]] <- temp_mat[(temp_mat[, slice_idx] >= slice_mean - slice_widths[idx]) &
+                                       (temp_mat[, slice_idx] <= slice_mean + slice_widths[idx]), ] %>%
+        cbind(idx)
+    }
+    slices[[time_idx]] <- reduce(slice_list, rbind)
+  }
+  mat_sliced <- reduce(slices, rbind)
+  # slice_pol <- cart2pol(mat_sliced[, 3:4])
+  # mat_sliced <- cbind(mat_sliced[, 1:4], slice_pol)
+
+  slice_df <- data.frame(mat_sliced)
+  names(slice_df) <- c("time", "x", "y", "z", "theta", "phi", "r", "Source")
+  slice_df <- slice_df %>%
+    mutate(
+      time = time * max_time,
+      time = round(time, 2),
+      time = as.factor(time),
+      Source = ifelse(
+        Source == 1,
+        "Data",
+        ifelse(
+          Source == 2,
+          "LPME",
+          "PME"
+        )
[…]

### 2023-08-04 code/lpme_simulation_analysis.R
add simulation study analysis code

@@ -1,0 +1,187 @@
+library(tidyverse)
+library(reticulate)
+library(pme)
+library(stringr)
+
+# files_case1 <- list.files("simulations/case1", full.names = TRUE)
+# files_case2 <- list.files("simulations/case2", full.names = TRUE)
+# files_case3 <- list.files("simulations/case3", full.names = TRUE)
+# files_case5 <- list.files("simulations/case5", full.names = TRUE)
+# files_case6 <- list.files("simulations/case6", full.names = TRUE)
+# files_case7 <- list.files("simulations/case7", full.names = TRUE)
+# files_case8 <- list.files("simulations/case8", full.names = TRUE)
+# files_case9 <- list.files("simulations/case9", full.names = TRUE)
+# files_case10 <- list.files("simulations/case10", full.names = TRUE)
+#
+# files_full <- c(
+#   files_case1,
+#   files_case2,
+#   files_case3,
+#   files_case5,
+#   files_case6,
+#   files_case7,
+#   files_case8,
+#   files_case9,
+#   files_case10
+# )
+
+files_full <- list.files("simulations", recursive = TRUE, full.names = TRUE)
+
+# files_full <- files_case1
+
+lpme_error <- vector()
+pme_error <- vector()
+# PME_error <- vector()
+pme_ssd <- vector()
+# PME_ssd <- vector()
+principal_curve_ssd <- vector()
+principal_curve_error <- vector()
+data_error <- vector()
+amp_noise <- vector()
+period_noise <- vector()
+n_visits <- vector()
+max_time <- vector()
+time_trend <- vector()
+sim_case <- vector()
+pme_ssd_raw <- vector()
+principal_curve_ssd_raw <- vector()
+for (i in 1:length(files_full)) {
+  error_list <- readRDS(files_full[i])
+  lpme_error[i] <- error_list$lpme_error
+  pme_error[i] <- error_list$pme_error
+  # PME_error[i] <- error_list$PME_error
+  principal_curve_error[i] <- error_list$principal_curve_error
+
+  # PME_ssd <- c(PME_ssd, error_list$PME_ssd)
+  principal_curve_ssd <- c(principal_curve_ssd, error_list$principal_curve_ssd)
+  data_error[i] <- error_list$data_error
+  amp_noise[i] <- error_list$amp_noise
[…]

### 2023-08-09 paper/main.tex
update literature review and augmentation explanation

@@
 Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
-To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this article is novel.
+\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to to incorporate longitudinal information when reducing neuroimaging information to lower dimensional spaces, but estimated parameterizations for baseline images and longitudinal image differences separately, where we seek to parameterize them simultaneously.
+\cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
+However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+To our knowledge, the approach to longitudinal smoothing in the general-purpose manifold learning setting taken in this article is novel.
 
@@
 As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
-However, by adding a third dimension equal to the angle of each observation from the origin, we the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
+However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
 This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
@@
 
+This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
+Reaching a combined estimate after fitting PME to each partition individually requires "gluing" the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
+This source of error is avoided using the data augmentation approach.
+Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold.
+This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances.
+However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.
+
 \begin{figure}%
   \centering
-  \subfloat[\centering $D = 2$]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
+  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
   \hfill
-  \subfloat[\centering $D = 3$]{{\includegraphics[width=9cm]{unit_circle_D3}}}
+  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
   \caption{Augmenting the Unit Circle}
@@
 \end{figure}
-
-\zielinski{The simulation results shown below use partial results run on my computer.}
 
@@
     \hline
-    Data & 0.222 (0.261) & 0.513 (0.411) & N/A & 27.3 (74.9) & 0.986 (0.787) & 1.51 (6.04) & N/A & N/A & N/A \\
-    LPME & 0.129 (0.161) & 0.417 (0.960) & N/A & 19.0 (63.0) & 0.776 (0.777) & 1.27 (5.71) & N/A & N/A & N/A \\
-    PME & 0.247 (0.435) & 0.854 (1.62) & N/A & 27.3 (74.9) & 1.04 (1.07) & 1.55 (6.04) & N/A & N/A & N/A \\
-    PC/PS & 0.188 (0.247) & 0.600 (0.303) & N/A & 27.3 (74.9) & 0.939 (0.730) & 1.04 (2.20) & N/A & N/A & N/A \\
+    Data & 0.222 (0.255) & 0.512 (0.409) & N/A & 30.5 (89.0) & 0.975 (0.770) & 1.43 (6.13) & N/A & N/A & N/A \\
+    LPME & 0.125 (0.161) & 0.384 (0.659) & N/A & 27.8 (266) & 0.788 (0.843) & 1.18 (5.71) & N/A & N/A & N/A \\
+    PME & 0.267 (0.862) & 0.839 (1.93) & N/A & 30.5 (89.2) & 1.03 (1.07) & 1.47 (6.14) & N/A & N/A & N/A \\
+    PC/PS & 0.188 (0.244) & 0.599 (0.297) & N/A & 30.4 (89.0) & 0.930 (0.713) & 1.00 (2.14) & N/A & N/A & N/A \\
     \hline
@@
     \hline
-    Data & 0.144 (0.222) & 0.466 (0.658) & N/A & 4.17 (19.4) & 0.890 (1.39) & 0.305 (1.07) & N/A & N/A & N/A \\
-    LPME & 0.0727 (0.129) & 0.250 (0.477) & N/A & 3.23 (11.3) & 0.575 (1.08) & 0.293 (0.907) & N/A & N/A & N/A \\
-    PME & 0.123 (0.249) & 0.516 (0.757) & N/A & 4.13 (19.2) & 0.897 (1.39) & 0.331 (1.04) & N/A & N/A & N/A \\
-    PC/PS & 0.112 (0.205) & 0.556 (0.427) & N/A & 4.12 (19.3) & 0.805 (1.26) & 0.567 (0.388) & N/A & N/A & N/A \\
+    Data & 0.145 (0.231) & 0.462 (0.661) & N/A & 4.21 (20.7) & 0.888 (1.33) & 0.279 (1.06) & N/A & N/A & N/A \\
+    LPME & 0.0733 (0.122) & 0.245 (0.460) & N/A & 3.27 (12.8) & 0.579 (1.08) & 0.269 (0.890) & N/A & N/A & N/A \\
+    PME & 0.129 (0.258) & 0.508 (0.747) & N/A & 4.19 (20.7) & 0.881 (1.36) & 0.313 (1.04) & N/A & N/A & N/A \\
+    PC/PS & 0.117 (0.205) & 0.560 (0.416) & N/A & 4.18 (20.7) & 0.817 (1.21) & 0.556 (0.386) & N/A & N/A & N/A \\
     \hline
@@
 First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly.
-Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that are too curvy.
+Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
 The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern.

### 2023-10-02 paper/main.tex
Updates from Overleaf
overleaf updates
fix table formatting
readd main.tex
Ani revisions

@@
 \author[1]{Ani Eloyan}
+\author[*]{for the Alzheimer’s Disease Neuroimaging Initiative}
 \affil[1]{Department of Biostatistics, Brown University}
@@
 For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.
-Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for further analysis.
-
+Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
 Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
@@
 
-To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
-In a larger comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
 After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
@@
 \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].
+In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
+To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
 
@@
 
-Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003a}), are ill-equipped to reach meaningful estimates across multiple time points.
-This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension reduction process.
+Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
+This is because these methods consider time as an additional dimension in the data, i.e.
+they include time in the dimension reduction process.
 This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
 Here, we propose an alternative approach to estimating manifolds over time.
-
 Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension.
-Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to estimate the appropriate principal manifold at each given time point.
-We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
-The regularization involved in this process mitigates the effects of variability between study visits.
-
-Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
-\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to to incorporate longitudinal information when reducing neuroimaging information to lower dimensional spaces, but estimated parameterizations for baseline images and longitudinal image differences separately, where we seek to parameterize them simultaneously.
+Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
+We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.
+Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
+\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously.
 \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
 However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
-To our knowledge, the approach to longitudinal smoothing in the general-purpose manifold learning setting taken in this article is novel.
+Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
 
-The remainder of this article is laid out as follows: In Section 2, we review the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}.
-Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this.
-Section 4 demonstrates the performance of this approach on simulated data.
-In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease.
-The article concludes with a discussion of the method's contributions in Section 6.
+The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework.
+In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework.
+Section \ref{s:simulations} demonstrates the performance of this approach on simulated data.
+In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus in individuals with Alzheimer's disease (AD).
+The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.
 
-\section{Principal Manifolds and Estimation Algorithm}
+\section{Principal Manifolds and Estimation Algorithm}\label{s:PME}
 
-The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data.
-However, the PS framework in \cite{hastiePrincipalCurves1989} is flawed by the model complexity selection, which \cite{duchamp1996extremal} pointed out.
-Motivated by the penalization approaches in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework and algorithm.
-Since our LPME is based on the PME, we review the PME in this section for the audience's convenience.
-The following concepts and notations are necessary for both the PME and our proposed LPME, and they will be used throughout this paper
+The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
+However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
+In addition, applying the methods in high-dimensional settings requires is problematic.
+Motivated by the penalization approaches presented by in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
+In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper.
+We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively.
+We assume that a data cloud is observed in the $D$ dimensional space.
+In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest.
+For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
+Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
+The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
+Most manifold estimation methods are based on defining a projection index, i.e.
+$\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
+The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
 \begin{itemize}
-\item $d$ and $D$ are two positive integers denoting the dimensions of the low-dimensional manifold and high-dimensional space, respectively; 
-\item for any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$;
-    \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$;
-    \item for every $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, the image of $f$ is referred to as the $d$-dimensional manifold $M_f^d$ determined by $f$, i.e., $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$;
-    \item For each $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and each point $x\in\mathbb{R}^D$, $\pi_f(x)$ is a point in $\mathbb{R}^d$ and denotes the ``projection index" of $x$; roughly, $\pi_f(x)$ is a point in $d$-dimensional space such that $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$; $\pi_f(x)$ is a generalization of the counterpart in \cite{hastiePrincipalCurves1989}, and its details can be found in Section 2 of \cite{mengPrincipalManifoldEstimation2021};
+%    \item  = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where  denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$;
+
     \item suppose $u(r_1,\ldots,r_d)$ is a scalar-valued function defined on $\mathbb{R}^d$; $\nabla^{\otimes 2}u=\left(\frac{\partial^2 u}{\partial r_i \partial r_j}\right)_{1\le i,j\le d}$ denotes the Hessian matrix of $u$, and $\|\nabla^{\otimes 2}u\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2} u(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2\right)^{\frac{1}{2}}$; in addition, $\|\nabla^{\otimes 2} u\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2dr $, and
@@
 With the preparations above, the definition of principal manifolds proposed in \cite{mengPrincipalManifoldEstimation2021} is presented as follows
-\commentout{
+
 \cite{hastiePrincipalCurves1989} defines principal curves as follows:
@@
 \end{equation}
-referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:}
+referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
+
 \begin{definition}
@@
 
-To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm.
-This algorithm has three main steps: data reduction, fitting, and tuning (see Figure 2 of \cite{mengPrincipalManifoldEstimation2021}).
-\meng{I would suggest presenting Figure 2 of \cite{mengPrincipalManifoldEstimation2021} in the present paper.}
-\begin{enumerate}
-    \item In the data reduction step, the high-dimensional mixture density estimation (HDMDE) algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$.
-The number $N$ of clusters is chosen using an iterative hypothesis testing approach, and $N$ tends to be much smaller than $I$ (see the simulation study in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
-    \item For each given value of $\lambda$, the fitting stage of the PME algorithm uses this reduced data set $\left\{\mu_j\right\}_{j=1}^{N}$ to find estimated function 
-\begin{equation}
-  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f), \label{eq:8}
-\end{equation}
-where $\mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$ approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$.
-\item The tuning stage chooses the optimal value of the parameter $\lambda$ by $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
-\begin{equation}
-  \mathcal{D}\left(\hat{f}_\lambda\right) = \frac{1}{I}\sum_{i=1}^{I}\left\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\right\|_{\mathbb{R}^{D}}^2 \label{eq:10}
-\end{equation}
-is an empirical estimate of the MSD using the full dataset.
-\end{enumerate}
-
-Using results in \cite{duchon1977splines} and \cite{wahba1990}, \cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of a spline form.
-Specifically, when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$.
-Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in \cite{mengPrincipalManifoldEstimation2021}.
-
-\meng{I commented out many literature review discussions about \cite{hastiePrincipalCurves1989} and \cite{smolaRegularizedPrincipalManifolds2001} to make this section more concise.
-In addition, if the formulas above are not referred to in our discussions, they shouldn't be given labels.}
-
-\section{Longitudinal Principal Manifold Estimation Algorithm}
+\section{Longitudinal Principal Manifold Estimation Algorithm}\label{s:LPME}
 
@@
 
-\section{Simulations}
+\section{Simulations}\label{s:simulations}
 
@@
 \begin{table}
+\small
   \centering
@@
 
-\begin{table}
+\begin{table}[h]
+\footnotesize
   \centering
@@
 
-\begin{figure}%
+\begin{figure}[h]
   \centering
@@
 
-\begin{figure}
+\begin{figure}[h]
   \centering
@@
 
-\section{Application}
+\section{Longitudinal Segmentation of MRI images in Alzheimer's Disease Studies}\label{s:application}
 
@@
 
-\section{Discussion}
+\section{Discussion}\label{s:discussion}
 

### 2023-10-04 paper/main.tex
update simulation results
update introduction

@@
 However, this approach is highly time and resource intensive.
-In studies analyzing a large number of images, this approach may impose prohibitive costs.
+When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
 Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
@@
 After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
-If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation.
-\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].
+If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement was much higher for FIRST and FreeSurfer than for manual segmentation.
+\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained.
 In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
@@
 Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods.
-Mitigating the extent of this variability is a priority from a statistical perspective.
+Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data.
 To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
-In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive than working with the structure in its original high-dimensional space.
+In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
 For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum.
@@
 Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
-This is because these methods consider time as an additional dimension in the data, i.e.
+These methods consider time as an additional dimension in the data, i.e.
 they include time in the dimension reduction process.
@@
     \hline
-    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
     \hline
-    Data & 0.222 (0.255) & 0.512 (0.409) & N/A & 30.5 (89.0) & 0.975 (0.770) & 1.43 (6.13) & N/A & N/A & N/A \\
-    LPME & 0.125 (0.161) & 0.384 (0.659) & N/A & 27.8 (266) & 0.788 (0.843) & 1.18 (5.71) & N/A & N/A & N/A \\
-    PME & 0.267 (0.862) & 0.839 (1.93) & N/A & 30.5 (89.2) & 1.03 (1.07) & 1.47 (6.14) & N/A & N/A & N/A \\
-    PC/PS & 0.188 (0.244) & 0.599 (0.297) & N/A & 30.4 (89.0) & 0.930 (0.713) & 1.00 (2.14) & N/A & N/A & N/A \\
+    Data & 0.223 (0.256) & 0.514 (0.408) & 0.454 (0.468) & 30.7 (88.1) & 0.980 (0.771) & 1.43 (6.04) & 0.571 (0.822) & 0.222 (0.268) \\
+    LPME & 0.125 (0.161) & 0.384 (0.648) & 0.414 (0.484) & 27.7 (260) & 0.791 (0.845) & 1.20 (5.66) & 3.87 (3.26) & 0.140 (0.177) \\
+    PME & 0.268 (0.850) & 0.843 (1.93) & 0.523 (0.648) & 30.7 (88.2) & 1.04 (1.07) & 1.47 (6.06) & 7.34 (1.14) & 0.238 (0.301) \\
+    PC/PS & 0.189 (0.245) & 0.600 (0.296) & 0.419 (0.444) & 30.6 (88.1) & 0.934 (0.713) & 1.01 (2.11) & 1.94 (0.786) & 0.271 (0.232) \\
     \hline
@@
     \hline
-    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
     \hline
-    Data & 0.145 (0.231) & 0.462 (0.661) & N/A & 4.21 (20.7) & 0.888 (1.33) & 0.279 (1.06) & N/A & N/A & N/A \\
-    LPME & 0.0733 (0.122) & 0.245 (0.460) & N/A & 3.27 (12.8) & 0.579 (1.08) & 0.269 (0.890) & N/A & N/A & N/A \\
-    PME & 0.129 (0.258) & 0.508 (0.747) & N/A & 4.19 (20.7) & 0.881 (1.36) & 0.313 (1.04) & N/A & N/A & N/A \\
-    PC/PS & 0.117 (0.205) & 0.560 (0.416) & N/A & 4.18 (20.7) & 0.817 (1.21) & 0.556 (0.386) & N/A & N/A & N/A \\
+    Data & 0.146 (0.233) & 0.467 (0.665) & 0.266 (0.607) & 4.26 (21.2) & 0.895 (1.33) & 0.284 (1.06) & 0.151 (0.539) & 0.113 (0.327) \\
+    LPME & 0.074 (0.122) & 0.248 (0.461) & 0.218 (0.562) & 3.29 (13.0) & 0.584 (1.08) & 0.273 (0.891) & 2.70 (4.20) & 0.077 (0.207) \\
+    PME & 0.131 (0.258) & 0.516 (0.750) & 0.311 (0.638) & 4.23 (21.1) & 0.894 (1.36) & 0.316 (1.04) & 6.93 (1.19) & 0.117 (0.336) \\
+    PC/PS & 0.118 (0.206) & 0.564 (0.419) & 0.242 (0.600) & 4.22 (21.2) & 0.821 (1.22) & 0.557 (0.387) & 1.57 (0.51) & 0.178 (0.237) \\
     \hline

### 2023-10-04 paper/main.tex
Ani updates and updated simulation results

--- 2023-10-02 paper/main.tex
@@
 The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
-\begin{itemize}
-%    \item  = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where  denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$;
-
-    \item suppose $u(r_1,\ldots,r_d)$ is a scalar-valued function defined on $\mathbb{R}^d$; $\nabla^{\otimes 2}u=\left(\frac{\partial^2 u}{\partial r_i \partial r_j}\right)_{1\le i,j\le d}$ denotes the Hessian matrix of $u$, and $\|\nabla^{\otimes 2}u\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2} u(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2\right)^{\frac{1}{2}}$; in addition, $\|\nabla^{\otimes 2} u\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2dr $, and
-    \begin{align*}
-        \nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\},
-    \end{align*}
-    where $\mathcal{D}'$ denotes the collection of distributions (also known as generalized functions, see \cite{duchon1977splines} and Chapter 6 of \cite{rudin1991functional});
-    \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$ denotes the collection of vector-valued functions $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ whose components are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d})$; besides, $\mathcal{C}_\infty\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)=\mathcal{C}_\infty(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$; the following functional of $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ measures the curvature of the manifold $M_f^d$ and will be used throughout this paper
-    \begin{equation}
-  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \|\nabla^{\otimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2  = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
-\end{equation}
-\end{itemize}
-With the preparations above, the definition of principal manifolds proposed in \cite{mengPrincipalManifoldEstimation2021} is presented as follows
-
-\cite{hastiePrincipalCurves1989} defines principal curves as follows:
-
-\begin{definition}
-  \label{def:principal_curves}
-
-  (Part I) Let $I \in \mathbb{R}^{1}$ be a closed and possibly infinite interval. Suppose a map $f:I \to \mathbb{R}^{D}$ satisfies the conditions:
-  \begin{enumerate}
-    \label{enum:hs_conditions}
-
-    \item $f \in \mathcal{C}^{\infty}(I \to \mathbb{R}^{D})$
-    \item $\|f'(r)\|_{\mathbb{R}^{D}} = 1$ for all $r \in I$
-    \item $f$ does not self intersect, meaning $r_1 \neq r_2$ implies that $f(r_1) \neq f(r_2)$
-    \item $\int_{\left\{r: f(r) \in B\right\}}dt < \infty$ for any finite ball $B$ in $\mathbb{R}^{D}$
-  \end{enumerate}
-  Then the projection index with respect to $f$ is defined as 
-  \begin{equation}
-    \pi_f(x) = \sup\left\{r \in I: \|x - f(r)\|_{\mathbb{R}^{D}} = \inf_{r' \in I}\|x - f(r')\|_{\mathbb{R}^{D}} \right\}, \ \text{for all} \ x \in \mathbb{R}^{D}. \label{eq:1}
-  \end{equation}
-
-  (Part II) Suppose $X$ is a continuous random $D$-vector with finite second moments.
-Principal curves of $X$ are all maps $f: I \to \mathbb{R}^{D}$ satisfying the conditions above and the Self-consistency defined as
-    \begin{equation}
-      E(X | \pi_f(X) = t) = f(r). \label{eq:2}
-    \end{equation}
-\end{definition}
-
-Principal manifolds seek to extend this concept to higher dimensions.
-\cite{smolaRegularizedPrincipalManifolds2001} proposed a regularized principal manifold framework, where principal manifolds minimize
-\begin{equation}
-  \argmin_{f \in \mathcal{F}}\left\{E\|X - f\left(\pi_f(X)\right)\|_{\mathbb{R}^D}^2 + \lambda\|Pf\|_{\mathcal{H}}^2\right\} \label{eq:3}
-\end{equation}
-with $\mathcal{F}$ being a collection of functions and $P$ being an operator that maps $f$ into inner product space $\mathcal{H}$.
-If $P = \frac{d}{dr}$, this framework minimizes the mean squared distance (MSD), $\mathcal{D}_X(f) = E\|X - f(\pi_f(X))\|_{\mathbb{R}^{D}}^2$, while penalizing the non-smoothness of the curve $f$.
-\cite{mengPrincipalManifoldEstimation2021} build on this framework to provide a more complete definition of principal manifolds, and proposes an algorithm to estimate these principal manifolds.
-
-To ensure that the upcoming definitions are well behaved, we introduce the following function spaces:
-\begin{itemize}
-  \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$
-  \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{f \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
-  \item $H^2(\mathbb{R}^{d}) = \left\{f \in L^2(\mathbb{R}^{d}): \|\nabla f\|_{\mathbb{R}^{d}}, \|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$
-  \item $\nabla^{-\otimes 2}L^2(\Omega) = \left\{f|_{\Omega}: f \in \nabla^{-\otimes 2}L^2(\mathbb{R}^{d})\right\}$
-  \item $H^2(\Omega) = \left\{f|_{\Omega}: f \in H^2(\mathbb{R}^{d})\right\}$
-\end{itemize}
-
-In these function spaces, we have the notation:
-\begin{itemize}
-  \item $\Omega$ is any open subset of $\mathbb{R}^{d}$ with a smooth boundary
-  \item $f|_{\Omega}$ denotes the restriction of $f$ to $\Omega$
-  \item $\nabla f$ denotes the gradient vector of $f$
-  \item $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$
-  \item $\|\nabla^{\otimes 2}f\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2}f(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial t_i \partial t_j}(r)\right|^2\right)^{\frac{1}{2}}$
-  \item $\|\nabla f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i=1}^{d}\left|\frac{\partial f}{\partial r_i}(r)\right|^2dr $
-  \item $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f}{\partial r_i \partial r_j}(r)\right|^2dr $
-\end{itemize}
-
-To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative.
-To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
-\begin{equation}
-  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
-\end{equation}
-referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
-
+The principal manifolds are then defined as follows.
 \begin{definition}
   \label{def:principal_manifolds}
-  Let $\boldsymbol{X}$ be a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments.
-Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
+  Given a random $D$-vector $\boldsymbol{X}$ associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments.
+Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty]$, and the following functionals:
   \begin{align}\label{eq:5}
@@
   \end{align}
-  A manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $\boldsymbol{X}$ with the tuning parameter $\lambda$ if 
-  \begin{equation}\label{eq:7}
+  The principal manifold of $\boldsymbol{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if 
+  \begin{equation}\nonumber
     \begin{aligned}
-        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \\ 
-    &\text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}.
+        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f),
     \end{aligned}
   \end{equation}
+where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
 \end{definition}
+The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements.
+First is the expected distance from each point in the cloud to the fitted manifold.
+In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on 3-dimensional brain region of interest.
+While the second element of the equation imposes the smoothness penalty on the estimated manifold.
+The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before.
+To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$.
 In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
-Thus, the first segment of the functional $\mathcal{K}_{\lambda, \mathbb{P}}$ provides a measure of the expected distance between the provided $D$-vector $\mathbf{X}$ and the value of $\mathbf{X}$ when projected onto the manifold.
-Meanwhile the second segment of $\mathcal{K}_{\lambda, \mathbb{P}}$ penalizes the roughness of the function $f$, with the magnitude of the penalty dependent on the provided value of $\lambda$.
+It is important to note, that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.
 
@@
 
-We now propose longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}.
+The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices.
+We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies.
+In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e.
+$D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest.
+Now consider that the data are observed longitudinally, during several visits by a study participant.
+Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.
+Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
+It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e.
+modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
+The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$ which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
+In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
+In order to address the limitations of the existing approaches for the setting described above, we now propose longitudinal principal manifold estimation (LPME) algorithm to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.
 
-Let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
+To give a more general representation of this problem, let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
 As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
@@
 
-\subsection{Definition of LPM}
+\subsection{The Longitudinal Principal Manifold Framework}
 
@@
     \hline
-    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
     \hline
-    Data & 0.222 (0.255) & 0.512 (0.409) & N/A & 30.5 (89.0) & 0.975 (0.770) & 1.43 (6.13) & N/A & N/A & N/A \\
-    LPME & 0.125 (0.161) & 0.384 (0.659) & N/A & 27.8 (266) & 0.788 (0.843) & 1.18 (5.71) & N/A & N/A & N/A \\
-    PME & 0.267 (0.862) & 0.839 (1.93) & N/A & 30.5 (89.2) & 1.03 (1.07) & 1.47 (6.14) & N/A & N/A & N/A \\
-    PC/PS & 0.188 (0.244) & 0.599 (0.297) & N/A & 30.4 (89.0) & 0.930 (0.713) & 1.00 (2.14) & N/A & N/A & N/A \\
+    Data & 0.223 (0.256) & 0.514 (0.408) & 0.446 (0.445) & 30.7 (88.1) & 0.980 (0.771) & 1.43 (6.04) & 0.580 (0.839) & 0.226 (0.275) \\
+    LPME & 0.125 (0.161) & 0.384 (0.648) & 0.401 (0.446) & 27.7 (260) & 0.791 (0.845) & 1.20 (5.66) & 4.07 (3.30) & 0.136 (0.169) \\
+    PME & 0.268 (0.850) & 0.843 (1.93) & 0.507 (0.594) & 30.7 (88.2) & 1.04 (1.07) & 1.47 (6.06) & 7.37 (1.14) & 0.242 (0.311) \\
+    PC/PS & 0.189 (0.245) & 0.600 (0.296) & 0.412 (0.423) & 30.6 (88.1) & 0.934 (0.713) & 1.01 (2.11) & 1.95 (0.800) & 0.274 (0.243) \\
     \hline
@@
     \hline
-    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
     \hline
-    Data & 0.145 (0.231) & 0.462 (0.661) & N/A & 4.21 (20.7) & 0.888 (1.33) & 0.279 (1.06) & N/A & N/A & N/A \\
-    LPME & 0.0733 (0.122) & 0.245 (0.460) & N/A & 3.27 (12.8) & 0.579 (1.08) & 0.269 (0.890) & N/A & N/A & N/A \\
-    PME & 0.129 (0.258) & 0.508 (0.747) & N/A & 4.19 (20.7) & 0.881 (1.36) & 0.313 (1.04) & N/A & N/A & N/A \\
-    PC/PS & 0.117 (0.205) & 0.560 (0.416) & N/A & 4.18 (20.7) & 0.817 (1.21) & 0.556 (0.386) & N/A & N/A & N/A \\
+    Data & 0.146 (0.233) & 0.467 (0.665) & 0.291 (0.601) & 4.26 (21.2) & 0.895 (1.33) & 0.284 (1.06) & 0.145 (0.552) & 0.110 (0.325) \\
+    LPME & 0.074 (0.122) & 0.248 (0.461) & 0.239 (0.564) & 3.29 (13.0) & 0.584 (1.08) & 0.273 (0.891) & 2.96 (4.65) & 0.074 (0.208) \\
+    PME & 0.131 (0.258) & 0.516 (0.750) & 0.317 (0.640) & 4.23 (21.1) & 0.894 (1.36) & 0.316 (1.04) & 6.92 (1.17) & 0.115 (0.331) \\
+    PC/PS & 0.118 (0.206) & 0.564 (0.419) & 0.264 (0.584) & 4.22 (21.2) & 0.821 (1.22) & 0.557 (0.387) & 1.58 (0.514) & 0.172 (0.234) \\
     \hline

### 2023-10-05 paper/main.tex
further updates

@@
 Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively.
-In addition, we assume $F\in\mathcal{C}_\infty\cap\nabla^{-\otimes2}L^2(\mathbb{R}^d\times\mathbb{R}\rightarrow\mathbb{R}^D)$.
-For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold at the time point.
-For given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we plan to minimize the following functional
+For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g.
+the surface of the 3D brain region, at time $t$.
+For given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the following functional, by extending the
 \begin{align}
@@
 .\end{align}
+where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
 
@@
 Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.
-
-\meng{Add the interpretation of the terms above.
-This interpretation presents the key idea of this paper.
-I need to address it carefully.
-One thing I would strongly suggest is to explain why we chose to deal with space and time separately.}
-
-\meng{Mentioned the functional above in all the four steps of LPME algorithm --- to emphasize that we are going to minimize the functional above.}
-
-\commentout{Note that we assume changes over time to be smooth.
-Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as }
 

### 2023-10-05 paper/main.tex

@@
 the surface of the 3D brain region, at time $t$.
-For given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the following functional, by extending the
+\begin{definition}
+  \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
 \begin{align}
@@
   &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \nonumber
-.\end{align}
+.\label{eq:newKappa}
+\end{align}
 where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
-
+Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+\end{definition}
 By minimizing the functional above, we retain the goals of Eq.~\eqref{eq:5} while also penalizing the changes in $F$ over time.

### 2023-10-06 paper/main.tex

@@
   \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
-\begin{align}
-  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \label{eq:11} \\
-  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \nonumber
-.\label{eq:newKappa}
+\begin{align}\label{eq:newKappa}
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
+  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt. \nonumber
 \end{align}
@@
 \end{definition}
-By minimizing the functional above, we retain the goals of Eq.~\eqref{eq:5} while also penalizing the changes in $F$ over time.
-Thus, we seek to minimize the distance between the observations $\mathbf{X_t}$ and their projections to $f_t$ while separately discouraging roughness both over the manifolds at each time point and over the changes in manifolds in the time dimension.
-The ability to distinguish between smoothing at a fixed time point and smoothing over time differentiates this approach from a thin plate spline, which would smooth over both the manifold and time jointly.
+By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, while imposing smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$.
+These goals correspond to minizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively.
+Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions.
+Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together.
 Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.
+Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered to be smooth over the other set of variables.
 
-\subsection{Algorithm}
+\subsection{The LPME Algorithm}
 
-To solve this problem, we propose an extension of the PME algorithm developed in \cite{mengPrincipalManifoldEstimation2021}.
 The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space.

### 2023-10-06 paper/enar_2024/enar_submission_2024.tex
formatting updates

%  template.tex for Biometrics papers
%
%  This file provides a template for Biometrics authors.  Use this
%  template as the starting point for creating your manuscript document.
%  See the file biomsample.tex for an example of a full-blown manuscript.

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS, usenatbib, referee]{biom}
%documentclass[useAMS]{biom}
%
%  If your system does not have the AMS fonts version 2.0 installed, then
%  remove the useAMS option.
%
%  useAMS allows you to obtain upright Greek characters.
%  e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
%  this guide for further information.
%
%  If you are using AMS 2.0 fonts, bold math letters/symbols are available
%  at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
%  preferably \bmath).
% 
%  Other options are described in the user guide. Here are a few:
% 
%  -  If you use Patrick Daly's natbib  to cross-reference your 
%     bibliography entries, use the usenatbib option
%
%  -  If you use \includegraphics (graphicx package) for importing graphics
%     into your figures, use the usegraphicx option
% 
%  If you wish to typeset the paper in Times font (if you do not have the
%  PostScript Type 1 Computer Modern fonts you will need to do this to get
%  smoother fonts in a PDF file) then uncomment the next line
%  \usepackage{Times}

%%%%% PLACE YOUR OWN MACROS HERE %%%%%

\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}

%\usepackage{algorithm2e}

\usepackage{authblk}

\usepackage{csquotes}
\usepackage[english]{babel}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\T}{\intercal}
\newcommand{\commentout}[1]{}

\newcommand{\kmedit}[1]{{\color{purple}  #1}}
\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}

\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}

%\theoremstyle{definition}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}

%  The rotating package allows you to have tables displayed in landscape
%  mode.  The rotating package is NOT included in this distribution, but
%  can be obtained from the CTAN archive.  USE OF LANDSCAPE TABLES IS
%  STRONGLY DISCOURAGED -- create landscape tables only as a last resort if
%  you see no other way to display the information.  If you do do this,
%  then you need the following command.

%\usepackage[figuresright]{rotating}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%  Here, place your title and author information.  Note that in 
%  use of the \author command, you create your own footnotes.  Follow
%  the examples below in creating your author and affiliation information.
%  Also consult a recent issue of the journal for examples of formatting.

\title[This is an Example of Recto Running Head]{This is an Example of
an Article Title}

%  Here are examples of different configurations of author/affiliation
%  displays.  According to the Biometrics style, in some instances,
%  the convention is to have superscript *, **, etc footnotes to indicate 
%  which of multiple email addresses belong to which author.  In this case,
%  use the \email{ } command to produce the emails in the display.

%  In other cases, such as a single author or two authors from 
%  different institutions, there should be no footnoting.  Here, use
%  the \emailx{ } command instead.

%  The examples below corrspond to almost every possible configuration
%  of authors and may be used as a guide.  For other configurations, consult
%  a recent issue of the the journal.

%  Single author -- USE \emailx{ } here so that no asterisk footnoting
%  for the email address will be produced.

%\author{John Author\emailx{email@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}

%  Two authors from the same institution, with both emails -- use
%  \email{ } here to produce the asterisk footnoting for each email address

%\author{John Author$^{*}$\email{author@address.edu} and
%Kathy Authoress$^{**}$\email{email2@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}

%  Exactly two authors from different institutions, with both emails  
%  USE \emailx{ } here so that no asterisk footnoting for the email address
%  is produced.

\author
{John Author\emailx{author@address.edu} \\
Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K. 
\and
Kathy Author\emailx{anotherauthor@address.edu} \\
Department of Biostatistics, University of North Carolina at Chapel Hill, 
Chapel Hill, North Carolina, U.S.A.}

%  Three or more authors from same institution with all emails displayed
%  and footnoted using asterisks -- use \email{ }

%\author{John Author$^*$\email{author@address.edu}, 
%Jane Author$^{**}$\email{jane@address.edu}, and 
%Dick Author$^{***}$\email{dick@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}

%  Three or more authors from same institution with one corresponding email
%  displayed

%\author{John Author$^*$\email{author@address.edu}, 
%Jane Author, and Dick Author \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}

%  Three or more authors, with at least two different institutions,
%  more than one email displayed

%\author{John Author$^{1,*}$\email{author@address.edu}, 
%Kathy Author$^{2,**}$\email{anotherauthor@address.edu}, and 
%Wilma Flinstone$^{3,***}$\email{wilma@bedrock.edu} \\
%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
%$^{2}$Department of Biostatistics, University of North Carolina at 
%Chapel Hill, Chapel Hill, North Carolina, U.S.A. \\
%$^{3}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}

%  Three or more authors with at least two different institutions and only
%  one email displayed

%\author{John Author$^{1,*}$\email{author@address.edu}, 
%Wilma Flinstone$^{2}$, and Barney Rubble$^{2}$ \\
%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
%$^{2}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}

\begin{document}

%  This will produce the submission and review information that appears
%  right after the reference section.  Of course, it will be unknown when
%  you submit your paper, so you can either leave this out or put in 
%  sample dates (these will have no effect on the fate of your paper in the
%  review process!)

\date{{\it Received October} 2007. {\it Revised February} 2008.  {\it
Accepted March} 2008.}

%  These options will count the number of pages and provide volume
%  and date information in the upper left hand corner of the top of the 
%  first page as in published papers.  The \pagerange command will only
%  work if you place the command \label{firstpage} near the beginning
%  of the document and \label{lastpage} at the end of the document, as we
%  have done in this template.

%  Again, putting a volume number and date is for your own amusement and
%  has no bearing on what actually happens to your paper!

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{64}
\pubyear{2008}
\artmonth{December}

%  The \doi command is where the DOI for your paper would be placed should it
%  be published.  Again, if you make one up and stick it here, it means 
%  nothing!

\doi{10.1111/j.1541-0420.2005.00454.x}

%  This label and the label ``lastpage'' are used by the \pagerange
%  command above to give the page range for the article.  You may have 
%  to process the document twice to get this to match up with what you 
%  expect.  When using the referee option, this will not count the pages
%  with tables and figures.

\label{firstpage}

%  put the summary for your paper here

\begin{abstract}
This is the summary for this paper.
\end{abstract}

%  Please place your key words in alphabetical order, separated
%  by semicolons, with the first letter of the first word capitalized,
%  and a period at the end of the list.
%

\begin{keywords}
A key word; But another key word; Still another key word; Yet another key word.
\end{keywords}

%  As usual, the \maketitle command creates the title and author/affiliations
%  display

\maketitle

%  If you are using the referee option, a new page, numbered page 1, will
%  start after the summary and keywords.  The page numbers thus count the
%  number of pages of your manuscript in the preferred submission style.
%  Remember, ``Normally, regular papers exceeding 25 pages and Reader Reaction 
%  papers exceeding 12 pages in (the preferred style) will be returned to 
%  the authors without review. The page limit includes acknowledgements, 
%  references, and appendices, but not tables and figures. The page count does 
%  not include the title page and abstract. A maximum of six (6) tables or 
%  figures combined is often required.''

%  You may now place the substance of your manuscript here.  Please use
%  the \section, \subsection, etc commands as described in the user guide.
%  Please use \label and \ref commands to cross-reference sections, equations,
%  tables, figures, etc.
%
%  Please DO NOT attempt to reformat the style of equation numbering!
%  For that matter, please do not attempt to redefine anything!

\section{Introduction}
\label{s:intro}

Here is the introduction. The file \texttt{biomsample.tex} gives an
example of using the \texttt{natbib} package to cross-reference
citations from the bibliography.  Here, we just do this manually to
demonstrate that the old-fashioned way also is acceptable.  See the
comment right before the bibliography section below for information on
using BiBTeX.

Please note that, although this document class produces a final
product that is very close to the format and appearance of a typeset
\textit{Biometrics} article, there may be some idiosyncracies that
cause the format to deviate slightly from that in the journal.  These
will be corrected at the typesetting phase should your paper be accepted
and forwarded for publication.  So do not worry if such things occur!

\section{Model}
\label{s:model}

\subsection{First Model Subsection}

The Cox model (Cox, 1972) is one of the most widely used statistical
models.  Hastie, Tibshirani, and Friedman (2001) is an example of a
citation to a work with three authors.  The first time you reference
one of these in the text, use all the authors names. However, in all
subsequent references, just use Hastie et al. (2001).  Works with four
or more authors are always referenced in the text using ``et al.''
All authors names should appear in the bibliography for all entries.

\subsection{Second Model Subsection}

Please use a recent issue of \textit{Biometrics} as a guide to the style
for citations and bibliography entries, and follow that style exactly!!

\section{Inference}
\label{s:inf}

Please see the file \texttt{biomsample.tex} for fancy examples of making
tables.  Here is a very simple one.  Use \texttt{table} for tables
that are narrow enough to fit in one column of the typeset journl; use
\texttt{table*} for tables that need to span two columns.  For
figures, use of \texttt{figure} and \texttt{figure*} is analogous.

\begin{table}
\caption{This is a simple table.}
\label{t:one}
\begin{center}
\begin{tabular}{lrrr}
\Hline
Estimator & \multicolumn{1}{c}{$\beta_1$} &  \multicolumn{1}{c}{$\beta_2$} & 
\multicolumn{1}{c}{$\beta_3$} \\ \hline
MLE & 10.18 & $-$3.26 & 0.13 \\
OLS & 9.92 & $-$3.19 & 0.11 \\
WLS & 9.88 & $-$3.33 & 0.12 \\
\hline
\end{tabular}
\end{center}
\end{table}

You can experiment with fancier tables than Table~\ref{t:one}.

We can get bold symbols using \verb+\bmath+, for example, $\bmath{\alpha}_i$.

\section{Discussion}
\label{s:discuss}

Put your final comments here. Here I test what happens when I cite \cite{mengPrincipalManifoldEstimation2021}.

%  The \backmatter command formats the subsequent headings so that they
%  are in the journal style.  Please keep this command in your document
%  in this position, right after the final section of the main part of 
%  the paper and right before the Acknowledgements, Supporting Information (Supplementary %  Materials),   and References sections.

\backmatter

%  This section is optional.  Here is where you will want to cite
%  grants, people who helped with the paper, etc.  But keep it short!

\section*{Acknowledgements}

The authors thank Professor A. Sen for some helpful suggestions,
Dr C. R. Rangarajan for a critical reading of the original version of the
paper, and an anonymous referee for very useful comments that improved
the presentation of the paper.\vspace*{-8pt}

%  Here, we create the bibliographic entries manually, following the
%  journal style.  If you use this method or use natbib, PLEASE PAY
%  CAREFUL ATTENTION TO THE BIBLIOGRAPHIC STYLE IN A RECENT ISSUE OF
%  THE JOURNAL AND FOLLOW IT!  Failure to follow stylistic conventions
%  just lengthens the time spend copyediting your paper and hence its
%  position in the publication queue should it be accepted.

%  We greatly prefer that you incorporate the references for your
%  article into the body of the article as we have done here 
%  (you can use natbib or not as you choose) than use BiBTeX,
%  so that your article is self-contained in one file.
%  If you do use BiBTeX, please use the .bst file that comes with 
%  the distribution.  In this case, replace the thebibliography
%  environment below by 
%
\bibliographystyle{biom} 
\bibliography{references}

%  If your paper refers to supporting web material, then you MUST
%  include this section!!  See Instructions for Authors at the journal
%  website http://www.biometrics.tibs.org

\section*{Supporting Information}

Web Appendix A, referenced in Section~\ref{s:model}, is available with
this paper at the Biometrics website on Wiley Online
Library.\vspace*{-8pt}

\appendix

%  To get the journal style of heading for an appendix, mimic the following.

\section{}
\subsection{Title of appendix}

Put your short appendix here.  Remember, longer appendices are
possible when presented as Supplementary Web Material.  Please 
review and follow the journal policy for this material, available
under Instructions for Authors at \texttt{http://www.biometrics.tibs.org}.

\label{lastpage}

\end{document}

### 2023-10-11 paper/enar_2024/enar_submission_2024.tex
add abstract

@@
 \usepackage{amscd}
-%\usepackage{amsthm}
+% \usepackage{amsthm}
 \usepackage{amsfonts}
@@
 
-%\usepackage{algorithm2e}
+\usepackage{algorithm2e}
 
@@
 \newcommand{\commentout}[1]{}
-
-\newcommand{\kmedit}[1]{{\color{purple}  #1}}
-\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
-\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}
-
-\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}
 
@@
 
-\title[This is an Example of Recto Running Head]{This is an Example of
-an Article Title}
+\title[Longitudinal Principal Manifold Estimation]{Longitudinal Principal Manifold Estimation}
 
@@
 %  is produced.
-
-\author
-{John Author\emailx{author@address.edu} \\
-Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K. 
-\and
-Kathy Author\emailx{anotherauthor@address.edu} \\
-Department of Biostatistics, University of North Carolina at Chapel Hill, 
-Chapel Hill, North Carolina, U.S.A.}
 
@@
 
-\date{{\it Received October} 2007. {\it Revised February} 2008.  {\it
-Accepted March} 2008.}
-
 %  These options will count the number of pages and provide volume
@@
 
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
-\volume{64}
-\pubyear{2008}
-\artmonth{December}
-
 %  The \doi command is where the DOI for your paper would be placed should it
@@
 %  nothing!
-
-\doi{10.1111/j.1541-0420.2005.00454.x}
 
@@
 \section{Introduction}
-\label{s:intro}
 
-Here is the introduction. The file \texttt{biomsample.tex} gives an
-example of using the \texttt{natbib} package to cross-reference
-citations from the bibliography.  Here, we just do this manually to
-demonstrate that the old-fashioned way also is acceptable.  See the
-comment right before the bibliography section below for information on
-using BiBTeX.
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}).
+Frequently, interest centers around longitudinal changes in one or more neurological substructures.
+These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}).
+For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.
+Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
+Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
+However, this approach is highly time and resource intensive.
+When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
+Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
+Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}).
+While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
 
-Please note that, although this document class produces a final
-product that is very close to the format and appearance of a typeset
-\textit{Biometrics} article, there may be some idiosyncracies that
-cause the format to deviate slightly from that in the journal.  These
-will be corrected at the typesetting phase should your paper be accepted
-and forwarded for publication.  So do not worry if such things occur!
+Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
+If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement was much higher for FIRST and FreeSurfer than for manual segmentation.
+\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained.
+In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
+To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
 
-\section{Model}
-\label{s:model}
+Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods.
+Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data.
+To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
+Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
+In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
+For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum.
+To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.
 
-\subsection{First Model Subsection}
+Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
+These methods consider time as an additional dimension in the data, i.e.
+they include time in the dimension reduction process.
+This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
+Here, we propose an alternative approach to estimating manifolds over time.
+Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension.
+Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
+We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.
+Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
+\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously.
+\cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
+However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
 
-The Cox model (Cox, 1972) is one of the most widely used statistical
-models.  Hastie, Tibshirani, and Friedman (2001) is an example of a
-citation to a work with three authors.  The first time you reference
-one of these in the text, use all the authors names. However, in all
-subsequent references, just use Hastie et al. (2001).  Works with four
-or more authors are always referenced in the text using ``et al.''
-All authors names should appear in the bibliography for all entries.
+The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework.
+In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework.
+Section \ref{s:simulations} demonstrates the performance of this approach on simulated data.
+In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus in individuals with Alzheimer's disease (AD).
+The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.
 
-\subsection{Second Model Subsection}
+\section{Principal Manifolds and Estimation Algorithm}\label{s:PME}
 
-Please use a recent issue of \textit{Biometrics} as a guide to the style
-for citations and bibliography entries, and follow that style exactly!!
+The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
+However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
+In addition, applying the methods in high-dimensional settings requires is problematic.
+Motivated by the penalization approaches presented by in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
+In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper.
+We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively.
+We assume that a data cloud is observed in the $D$ dimensional space.
+In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest.
+For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
+Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
+The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
+Most manifold estimation methods are based on defining a projection index, i.e.
+$\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
+The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
+The principal manifolds are then defined as follows.
+\begin{definition}
+  \label{def:principal_manifolds}
+  Given a random $D$-vector $\boldsymbol{X}$ associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments.
+Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty]$, and the following functionals:
+  \begin{align}\label{eq:5}
+&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
+&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
+  \end{align}
+  The principal manifold of $\boldsymbol{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if 
+  \begin{equation}\nonumber
+    \begin{aligned}
+        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f),
+    \end{aligned}
+  \end{equation}
+where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
+\end{definition}
+The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements.
+First is the expected distance from each point in the cloud to the fitted manifold.
+In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on 3-dimensional brain region of interest.
+While the second element of the equation imposes the smoothness penalty on the estimated manifold.
+The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before.
+To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$.
+In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
+It is important to note, that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.
 
-\section{Inference}
-\label{s:inf}
+\section{Longitudinal Principal Manifold Estimation Algorithm}\label{s:LPME}
 
-Please see the file \texttt{biomsample.tex} for fancy examples of making
-tables.  Here is a very simple one.  Use \texttt{table} for tables
-that are narrow enough to fit in one column of the typeset journl; use
-\texttt{table*} for tables that need to span two columns.  For
-figures, use of \texttt{figure} and \texttt{figure*} is analogous.
+The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices.
+We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies.
+In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e.
+$D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest.
+Now consider that the data are observed longitudinally, during several visits by a study participant.
+Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.
+Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
+It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e.
+modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
+The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$ which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
+In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
+In order to address the limitations of the existing approaches for the setting described above, we now propose longitudinal principal manifold estimation (LPME) algorithm to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.
+
+To give a more general representation of this problem, let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
+As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
+However, in this updated case, the function determining the manifold may change over time.
+
+\subsection{The Longitudinal Principal Manifold Framework}
+
+Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively.
+For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g.
+the surface of the 3D brain region, at time $t$.
+\begin{definition}
+  \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
+\begin{align}\label{eq:newKappa}
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
+  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt. \nonumber
+\end{align}
+where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
+Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+\end{definition}
+By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, while imposing smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$.
+These goals correspond to minizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively.
+Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions.
+Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together.
+Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.
+Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered to be smooth over the other set of variables.
+
+\subsection{The LPME Algorithm}
+
+The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space.
+In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time.
+This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set.
+Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning.
+Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
+
+\subsection*{Data Reduction}
+We begin by considering $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} (Section 5.3 therein) to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$.
+HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.
+
+This step reduces the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to the size $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
+The centers estimated in this step are then used to reach an initial low-dimensional parameterization.
+
+\subsection*{Initialization}
+The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
+However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e.
+low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
+Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
+
+Instead of using Isomap on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one Isomap run to reach a parameterization for the centers across all time points simultaneously.
+Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.
+
+\subsection*{Fitting}
+Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point.
+Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$, which deals with the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in Eq.~\eqref{eq:11}.
+
+The remainder of the fitting stage deals with the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$, i.e., the curvature in the time dimension (see Eq.~\eqref{eq:11}).
+Specifically, the remainder of this stage involves estimating a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
+
+At each time point, the estimated principal manifold has a spline form.
+As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points being used to fit them.
+This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical.
+Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ being specified based on the maximum number of clusters identified in the data reduction step at any time point.
+This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
+Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
+\begin{itemize}
+    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
+    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
+    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
+    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
+    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
+\end{itemize}
+This minimization is conducted by solving
+\begin{equation}
+  \left(
+    \begin{array}{cc}
+      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
+      \boldsymbol{R} & \boldsymbol{0}
+    \end{array}
+  \right)\left(
+    \begin{array}{c}
+      \boldsymbol{s}_t \\
+      \boldsymbol{\alpha}_t
+    \end{array}
+  \right) = \left(
+    \begin{array}{c}
+      \boldsymbol{Y}_t \\
+      \boldsymbol{0}
+    \end{array}
+  \right). \label{eq:13}
+\end{equation}
+The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
+\begin{equation}
+  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
+\end{equation}
+
+We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$.
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to
+$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
+
+With a given tuning value $\gamma$, we the fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$.
+This function takes the form
+
+\begin{equation}
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
+\end{equation}
+where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
+\begin{equation}
+  \left(
+  \begin{array}{ccc}
+    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{A} + 2\gamma\boldsymbol{A} & 2\boldsymbol{A}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{T} \\
+    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{A} & 2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{0} \\
+    \boldsymbol{T}^{T} & \boldsymbol{0} & \boldsymbol{0}
+  \end{array}
+  \right)\left(
+  \begin{array}{c}
+    \boldsymbol{\delta} \\
+    \boldsymbol{\nu} \\
+    \boldsymbol{m}
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{\Omega} \\
+    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{\Omega} \\
+    \boldsymbol{0}
+  \end{array}
+  \right), \label{eq:16}
+\end{equation}
+where $\boldsymbol{\Omega}=(\boldsymbol{\omega}_1,\ldots, \boldsymbol{\omega}_T)^\T$.
+
+With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$.
+Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\boldsymbol{r}$ may be written as
+\begin{equation}
+  F_{\gamma}(t, \boldsymbol{r}) = \sum_{j=1}^{N}\boldsymbol{s}_{\gamma}(t)_j \eta_{d}\left(\|\boldsymbol{r} - \boldsymbol{r}_j^*\|\right) + \sum_{k=1}^{d+1}\boldsymbol{\alpha}_{\gamma}(t)_k \phi_k(\boldsymbol{r}).
+\label{eq:17}
+\end{equation}
+
+\subsection*{Tuning}
+
+Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
+In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
+We denote this function $g_{\gamma}^{(t)}$.
+We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
+\begin{equation}
+  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
+\end{equation}
+The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
+
+Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
+\begin{algorithm}
+\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
+  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
+\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
+\For{$t = 1, 2, \dots, T$} {
+  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
+}
+
+  Apply Isomap to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
+Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
+
+\For{t = 1, 2, \dots, T} {
+  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
+  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
+}
+
+  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$\;
+
+\For{t = 1, 2, \dots, T} {
+  Set $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
+  Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
+  Set $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
+}
+
+  Define $\boldsymbol{\omega}$ by setting $\boldsymbol{\omega}_t = \left[\boldsymbol{s}_t, \boldsymbol{\alpha}_t\right]$\;
+  \For{l = 1, 2, \dots, L} {
+    Compute $g_{\gamma_l}(t)$ by solving \eqref{eq:16}\;
+    \For{t = 1, 2, \dots, T} {
+      Compute $g_{\gamma_l}^{(t)}$ by solving \eqref{eq:16}\;
+    }
+    Estimate $MSD(\gamma_l)$ using \eqref{eq:18}\;
+  }
+  $\gamma^* = \arg\min_{\gamma}MSD(\gamma)$\;
+  $f^*(t, \boldsymbol{r}) = f_{\gamma^*}(t, \boldsymbol{r})$, where the form of $f^*(t, \boldsymbol{r})$ is given in \eqref{eq:17}
+\end{algorithm}
+
+\section{Simulations}\label{s:simulations}
+
+To assess the performance of the LPME algorithm, we compare its performance with that of two alternative approaches on simulated datasets.
+We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
+For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.
+
+The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
+In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural changes in the underlying manifold over time.
+The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
+The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
+Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
+These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.
 
 \begin{table}
-\caption{This is a simple table.}
-\label{t:one}
-\begin{center}
-\begin{tabular}{lrrr}
-\Hline
-Estimator & \multicolumn{1}{c}{$\beta_1$} &  \multicolumn{1}{c}{$\beta_2$} & 
-\multicolumn{1}{c}{$\beta_3$} \\ \hline
-MLE & 10.18 & $-$3.26 & 0.13 \\
-OLS & 9.92 & $-$3.19 & 0.11 \\
-WLS & 9.88 & $-$3.33 & 0.12 \\
+\small
+  \centering
+  \begin{tabular}{|c c c c c|}
+    \hline
+    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
+    \hline
+    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
+    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
+    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{cos} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
+    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
+    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
+    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
+    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
+    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
+    9 & 2 & 3 & $\left(\theta\text{sin} \ (\beta_1r_1)\text{cos} \ (\beta_2r_2), \ \theta\text{sin} \ (\beta_1)\text{sin} \ (\beta_2r_2), \ \theta\text{cos} \ (\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
     \hline
   \end{tabular}
-\end{center}
+  \caption{Embedding functions used for simulation trials.}
+  \label{table:simulation_embeddings}
 \end{table}
 
-You can experiment with fancier tables than Table~\ref{t:one}.
+In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run naively.
+The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}).
+In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
+Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$.
+In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
 
-We can get bold symbols using \verb+\bmath+, for example, $\bmath{\alpha}_i$.
+Notably, the manifolds estimated in the simulated datasets included several self-intersecting manifolds, which methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit due to the self-consistency property discussed above.
+To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$.
+As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
+However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
+This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
+This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.
 
-\section{Discussion}
-\label{s:discuss}
+This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
+Reaching a combined estimate after fitting PME to each partition individually requires "gluing" the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
+This source of error is avoided using the data augmentation approach.
+Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold.
+This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances.
+However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.
 
-Put your final comments here. Here I test what happens when I cite \cite{mengPrincipalManifoldEstimation2021}.
+\begin{figure}%
+  \centering
+  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
+  \hfill
+  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
+  \caption{Augmenting the Unit Circle}
+  \label{fig:unit_circle_augmentation}
+\end{figure}
 
-%  The \backmatter command formats the subsequent headings so that they
-%  are in the journal style.  Please keep this command in your document
-%  in this position, right after the final section of the main part of 
-%  the paper and right before the Acknowledgements, Supporting Information (Supplementary %  Materials),   and References sections.
+A factorial design is used to run the simulation studies, with the factor levels set as follows:
+\begin{itemize}
+  \item $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$
+  \item Study duration: $\left\{1, 2, 5\right\}$
+  \item Interval between images: $\left\{0.1, 0.25, 0.5\right\}$
+  \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
+\end{itemize}
+
+The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times.
+All code required to reproduce the simulation results and the associated figures is provided at \texttt{https://github.com/rjzielinski/lpme-project}.
+
+Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case4} and \ref{fig:sim_case6}.
+Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
+Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.
+
+\begin{table}[h]
+\footnotesize
+  \centering
+  \begin{tabular}{|c c c c c c c c c c|}
+    \hline
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
+    \hline
+    Data & 0.223 (0.256) & 0.514 (0.408) & 0.446 (0.445) & 30.7 (88.1) & 0.980 (0.771) & 1.43 (6.04) & 0.580 (0.839) & 0.226 (0.275) \\
+    LPME & 0.125 (0.161) & 0.384 (0.648) & 0.401 (0.446) & 27.7 (260) & 0.791 (0.845) & 1.20 (5.66) & 4.07 (3.30) & 0.136 (0.169) \\
+    PME & 0.268 (0.850) & 0.843 (1.93) & 0.507 (0.594) & 30.7 (88.2) & 1.04 (1.07) & 1.47 (6.06) & 7.37 (1.14) & 0.242 (0.311) \\
+    PC/PS & 0.189 (0.245) & 0.600 (0.296) & 0.412 (0.423) & 30.6 (88.1) & 0.934 (0.713) & 1.01 (2.11) & 1.95 (0.800) & 0.274 (0.243) \\
+    \hline
+  \end{tabular}
+  \caption{MSD Comparison to True Values, Mean (SD).}
+  \label{table:simulation_results_mean}
+\end{table}
+
+\begin{table}[h]
+\footnotesize
+  \centering
+  \begin{tabular}{|c c c c c c c c c c|}
+    \hline
+    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
+    \hline
+    Data & 0.146 (0.233) & 0.467 (0.665) & 0.291 (0.601) & 4.26 (21.2) & 0.895 (1.33) & 0.284 (1.06) & 0.145 (0.552) & 0.110 (0.325) \\
+    LPME & 0.074 (0.122) & 0.248 (0.461) & 0.239 (0.564) & 3.29 (13.0) & 0.584 (1.08) & 0.273 (0.891) & 2.96 (4.65) & 0.074 (0.208) \\
+    PME & 0.131 (0.258) & 0.516 (0.750) & 0.317 (0.640) & 4.23 (21.1) & 0.894 (1.36) & 0.316 (1.04) & 6.92 (1.17) & 0.115 (0.331) \\
+    PC/PS & 0.118 (0.206) & 0.564 (0.419) & 0.264 (0.584) & 4.22 (21.2) & 0.821 (1.22) & 0.557 (0.387) & 1.58 (0.514) & 0.172 (0.234) \\
+    \hline
+  \end{tabular}
+  \caption{MSD Comparison to True Values, Median (IQR).}
+  \label{table:simulation_results_median}
+\end{table}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{sim_case1}
+  \caption{Simulation Case 1}
+  \label{fig:sim_case1}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{sim_case5}
+  \caption{Simulation Case 5}
+  \label{fig:sim_case4}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{sim_case7}
+  \caption{Simulation Case 6}
+  \label{fig:sim_case6}
+\end{figure}
+
+The mean, median, and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
+Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
+Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach should result in lower mean squared distance values.
+
+The result summaries indicate that in most cases, LPME provides a substantial improvement in performance over the PME and principal curve approaches when estimating the underlying manifold.
+As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates, shown in blue, remain relatively stable over time.
+This contrasts to the estimates found by the PME and principal curve approaches, which are highly sensitive to the data observed at each given time point, as expected.
+This ultimately results in the LPME estimates remaining closer to the true underlying manifold, shown in red.
+Similar results hold in most of the other simulation cases.
+
+\section{Longitudinal Segmentation of MRI images in Alzheimer's Disease Studies}\label{s:application}
+
+As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images.
+This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates.
+To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).
+
+While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}).
+Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals.
+Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected.
+Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process.
+Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
+
+Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}.
+Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation.
+Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
+The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.
+Finally, the data were augmented with spherical coordinates in a manner similar to that described in the previous section to avoid the need to fit to a self-intersecting manifold.
+
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}.
+Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data.
+There are two main sources of discrepancies between the data and the LPME estimate.
+First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation.
+The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.
+
+Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
+These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
+This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots.
+Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself.
+This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{adni_lhipp}
+  \caption{Left Hippocampus, Cognitively Healthy ADNI Participant}
+  \label{fig:adni_result}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{adni_lhipp_cross_section}
+  \caption{Left Hippocampus Cross Section}
+  \label{fig:adni_cross_section}
+\end{figure}
+
+\section{Discussion}\label{s:discussion}
+
+In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data.
+To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a longitudinal setting.
+In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point.
+This improvement was seen across several manifold types.
+We also suggest a data augmentation approach that may allow for the circumvention of the inability to fit principal curve-based methods to some self-intersecting manifolds.
+
+While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations.
+First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly.
+Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
+The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern.
+There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.
+
+A second limitation is that the LPME method requires relative stability in the overall shape of the structure being fit over time.
+Thus, while this algorithm can be useful for separating meaningful changes in a manifold over time from inter-observation error, it will be unable to fit in situations where the error between observations is too substantial.
+How restrictive this requirement is depends largely on the area of application.
+When used to estimate subcortical surfaces as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.
+
+Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset.
+It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus.
+This shortcoming could potentially be addressed in several ways.
+First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated.
+Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
+Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section 4.
+
+Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
+The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
+However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
+The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.
+While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
 
@@
 
-%  This section is optional.  Here is where you will want to cite
-%  grants, people who helped with the paper, etc.  But keep it short!
-
-\section*{Acknowledgements}
-
-The authors thank Professor A. Sen for some helpful suggestions,
-Dr C. R. Rangarajan for a critical reading of the original version of the
-paper, and an anonymous referee for very useful comments that improved
-the presentation of the paper.\vspace*{-8pt}
-
-%  Here, we create the bibliographic entries manually, following the
-%  journal style.  If you use this method or use natbib, PLEASE PAY
-%  CAREFUL ATTENTION TO THE BIBLIOGRAPHIC STYLE IN A RECENT ISSUE OF
-%  THE JOURNAL AND FOLLOW IT!  Failure to follow stylistic conventions
-%  just lengthens the time spend copyediting your paper and hence its
-%  position in the publication queue should it be accepted.
-
-%  We greatly prefer that you incorporate the references for your
-%  article into the body of the article as we have done here 
-%  (you can use natbib or not as you choose) than use BiBTeX,
-%  so that your article is self-contained in one file.
-%  If you do use BiBTeX, please use the .bst file that comes with 
-%  the distribution.  In this case, replace the thebibliography
-%  environment below by 
-%
 \bibliographystyle{biom} 
 \bibliography{references}
-
-%  If your paper refers to supporting web material, then you MUST
-%  include this section!!  See Instructions for Authors at the journal
-%  website http://www.biometrics.tibs.org
-
-\section*{Supporting Information}
-
-Web Appendix A, referenced in Section~\ref{s:model}, is available with
-this paper at the Biometrics website on Wiley Online
-Library.\vspace*{-8pt}
-
-\appendix
-
-%  To get the journal style of heading for an appendix, mimic the following.
-
-\section{}
-\subsection{Title of appendix}
-
-Put your short appendix here.  Remember, longer appendices are
-possible when presented as Supplementary Web Material.  Please 
-review and follow the journal policy for this material, available
-under Instructions for Authors at \texttt{http://www.biometrics.tibs.org}.
 

### 2023-10-11 paper/main.tex
add abstract

@@
 \section*{Abstract}
+
+We propose a longitudinal extension to principal manifold estimation, with the goal of recovering smooth, longitudinally meaningful manifold estimates over time.
+The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point.
+This smoothing mitigates the effects of random disturbances to the manifold between time points.
+A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds.
+We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
+These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
+We then apply the proposed method to estimating the surfaces of hippocampi of healthy individuals in the Alzheimer’s Disease Neuroimaging Initiative dataset.
 
@@
 
-The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space.
-In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time.
-This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set.
+Given the framework proposed in the previous Section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}).
+Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping.
+Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space.
+For example, the PME algorithm can be used to obtain this parameterization.
+In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, thus smoothing over the spatial spline coefficients with respect to time.
+This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set.
 Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning.
@@
 
-\subsection*{Data Reduction}
-We begin by considering $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} (Section 5.3 therein) to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$.
-HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.
+\subsection*{Sample Size Reduction}
 
-This step reduces the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to the size $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
+Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud.
+This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest.
+Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large.
+To reduce the sample size $I_t$, we use the HDMDE method (see\cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
+The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
+By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
 The centers estimated in this step are then used to reach an initial low-dimensional parameterization.
@@
 \subsection*{Initialization}
-The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
-However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e.
-low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another) \meng{I would suggest writing more explanations for the ``inconsistent parameterization directions." Some visual illustrations are necessary.}.
+
+The next step in developing the algorithm for estimation of the longitudinal principal manifold is initialization.
+Different algorithms use different approaches for this step.
+For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
+However, implementation of commonly used approaches for initialization in our setting can be problematic.
+For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e.
+the brain region of interest at each time point) is likely to result in inconsistent parameterization directions (i.e.
+low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
-\meng{This paragraph should be devoted to the explanation for ``inconsistent parameterization directions." The comparison can be presented using visualizations.}
 
-Instead of using Isomap on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one Isomap run to reach a parameterization for the centers across all time points simultaneously.
-Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.
-\meng{I would suggest presenting simulation studies showing the ``consistency over time." Specifically, we may compare the difference between time-wise and over-time initializations via simulations.}
+To obtain the same parameterization of the manifold over time, instead of using Isomap on the data for each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously.
+This approach is based on the assumption that the shape of the structure, i.e.
+the brain region of interest, remains relatively stable over time.
+In Section \ref{s:simulations}, we consider some deviations from this assumption and their effect on the results and we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.
 
 \subsection*{Fitting}
+
 Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point.
 Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
-This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$, which deals with the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in Eq.~\eqref{eq:11}.
-
-The remainder of the fitting stage deals with the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$, i.e., the curvature in the time dimension (see Eq.~\eqref{eq:11}).
-Specifically, the remainder of this stage involves estimating a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
-\meng{The term ``coefficient space" is clear to me, but it might be confusing to other readers who are unfair with the details of PME.}
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa}.
+At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+The principal manifold $\widehat{f_t}$ at time $t$ is characterized by spline coefficients.
+Hence, we estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
 
 At each time point, the estimated principal manifold has a spline form.
-As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points being used to fit them.
+As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points used to fit them.
 This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical.
-Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ being specified based on the maximum number of clusters identified in the data reduction step at any time point.
+Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point.
 This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.

### 2023-10-17 paper/main.tex
overleaf updates

@@
 Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large.
-To reduce the sample size $I_t$, we use the HDMDE method (see\cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
 Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
@@
 This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
-Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
-\begin{itemize}
-    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
-    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
-    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
-    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
-    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
-\end{itemize}
-This minimization is conducted by solving
+Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing the following function.
+\begin{equation}\label{eq:splineL}
+    \mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t
+\end{equation}    
+where $\boldsymbol{Y}_t$ is the $N \times N$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
+
+By computing the derivative of $\mathcal{L}(f_t)$ in (\ref{eq:splineL}), we conduct the minimization of $\mathcal{L}(f_t)$ by solving the following system of equations. 
 \begin{equation}
@@
 The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
-\begin{equation}
-  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
+\begin{equation}\nonumber
+  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). 
 \end{equation}
@@
 We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$ \meng{I would suggest writing the $\tau_t$ in a more precise way.}, the weights for each time are equal to
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where $\tau_t$ denote the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to
 $$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
 
-With a given tuning value $\gamma$, we the fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$.
-This function takes the form
-
-\begin{equation}
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
+With a given tuning parameter value $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$ resulting in a function taking the form
+\begin{equation}\nonumber
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j,
 \end{equation}
-where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
+where by defining the $T \times T$ matrix $\boldsymbol{A}$ as defined  $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\boldsymbol{T}$ as defined $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
 \begin{equation}
@@
 
-\meng{From the methodology viewpoint, the fitting section is the most important part of this paper.}
-
-\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction.
-The smoothness can be represented by the covariance kernel we use in the GPR.
-For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels.
-The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
-
 \subsection*{Tuning}
@@
 In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
-We denote this function $g_{\gamma}^{(t)}$.
-We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows: \meng{What are the $Y_{i,t}$'s?}
+We denote this function by $g_{\gamma}^{(t)}$.
+We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows.
 \begin{equation}
@@
 The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
-\meng{I would suggest some simulation studies showing the performance of the $\gamma^*$ something, like the Figure 4 of \cite{mengPrincipalManifoldEstimation2021}.}
+Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
 
-Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
-\meng{Do some simulation studies and compare the leave-one-out CV and k-fold CV.}
+%\meng{Do some simulation studies and compare the leave-one-out CV and k-fold CV.}
 
@@
 \LinesNumbered
-
-\meng{Make the notations in Algorithm \ref{alg:lpme} consistent with that in the description above.}
-
-\meng{FYI, to fit a circle-like manifold, you may need to refer to the following papers: \cite{fasy2014confidence}, \cite{dunsonInferringManifoldsNoisy2022}.}
 
@@
 
+\subsection{Self-intersecting Manifold Estimation}\label{ss:selfInt}
+
+Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds.
+Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency property.
+To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$.
+As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
+However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
+This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
+This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.
+
+This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
+Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
+This source of error is avoided using our proposed data augmentation approach.
+Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold.
+This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances.
+However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.
+
+\begin{figure}%
+  \centering
+  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
+  \hfill
+  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
+  \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
+The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
+  \label{fig:unit_circle_augmentation}
+\end{figure}
+
 \section{Simulations}\label{s:simulations}
@@
 We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
-For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.
-
+For each manifold, we consider differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time.
 The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
@@
 
-In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run naively.
+In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
 The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}).
@@
 In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
-
-Notably, the manifolds estimated in the simulated datasets included several self-intersecting manifolds, which methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit due to the self-consistency property discussed above.
-To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$.
-As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
-However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
-This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
-This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.
-
-This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
-Reaching a combined estimate after fitting PME to each partition individually requires "gluing" the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
-This source of error is avoided using the data augmentation approach.
-Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold.
-This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances.
-However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.
-
-\begin{figure}%
-  \centering
-  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
-  \hfill
-  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
-  \caption{Augmenting the Unit Circle}
-  \label{fig:unit_circle_augmentation}
-\end{figure}
 
@@
 \end{itemize}
-
 The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times.
-All code required to reproduce the simulation results and the associated figures is provided at \texttt{https://github.com/rjzielinski/lpme-project}.
+All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
 
@@
 
+\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction.
+The smoothness can be represented by the covariance kernel we use in the GPR.
+For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels.
+The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
 \newpage

### 2023-10-18 paper/main.tex
add lpme steps figures

@@
 Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.
+Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{sim_case1_plot2}
+  \caption{Longitudinal Simulated Data with $D = 2$}
+  \label{fig:sample_data}
+\end{figure}
+
 Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
@@
 Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning.
+A visualization of these steps is shown in Figure \ref{fig:lpme_steps}.
 Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
 
-\meng{I would suggest making a flowchart similar to Figure 2 of \cite{mengPrincipalManifoldEstimation2021}.}
+\begin{figure}%
+  \centering
+  \subfloat[\centering Data]{{\includegraphics[width=9cm]{sim_case1_plot2}}}%
+  \hfill
+  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=9cm]{lpme_data_reduction}}}
+  \vfill
+  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=9cm]{lpme_initialization}}}
+  \hfill
+  \caption{LPME Algorithm Steps}
+  \label{fig:lpme_steps}
+\end{figure}
 

### 2023-11-10 code/lpme_adni_application.R
add clean code file for adni application

@@ -1,0 +1,304 @@
+library(lubridate)
+library(plotly)
+library(plot3D)
+library(pme)
+library(pracma)
+library(RColorBrewer)
+library(Rfast)
+library(tidyverse)
+
+source("code/functions/calc_pme_est.R")
+source("code/functions/calc_lpme_est.R")
+
+# function for creating cross-section plots
+create_cross_section_matrix <- function(mats, slice_idx, slice_widths, nrow = 1, lhipp) {
+  time_vals <- unique(mats[[1]][, 1])
+  slices <- list()
+  for (time_idx in 1:length(time_vals)) {
+    slice_list <- list()
+    for (idx in 1:length(mats)) {
+      mat <- mats[[idx]]
+      temp_mat <- mat[mat[, 1] == time_vals[time_idx], ]
+      n_pixels <- length(unique(temp_mat[, slice_idx]))
+      slice_mean <- mean(temp_mat[, slice_idx])
+      # in the PME- and LPME-generated value matrices, voxel values are not all
+      # consistent as in the data
+      # to approximate a cross-section through the middle of the structure, we
+      # consider all voxel values within a certain range of the mean voxel value.
+      slice_list[[idx]] <- temp_mat[(temp_mat[, slice_idx] >= slice_mean - slice_widths[idx]) &
+                                      (temp_mat[, slice_idx] <= slice_mean + slice_widths[idx]), ] %>%
+        cbind(idx)
+    }
+    slices[[time_idx]] <- reduce(slice_list, rbind)
+  }
+  mat_sliced <- reduce(slices, rbind)
+
+  slice_df <- data.frame(mat_sliced)
+  names(slice_df) <- c("time", "x", "y", "z", "theta", "phi", "r", "Source")
+  slice_df <- slice_df %>%
+    mutate(
+      time = time * unique(lhipp$duration),
+      time = round(time, 2),
+      time = as.factor(time),
+      Source = ifelse(
+        Source == 1,
+        "Data",
+        ifelse(
+          Source == 2,
+          "LPME",
+          "PME"
+        )
[…]

### 2023-11-10 paper/main.tex
update figures

@@
   \hfill
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=9cm]{lpme_fitting}}}
   \caption{LPME Algorithm Steps}
@@
 low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
+This is demonstrated in Figure \ref{fig:inconsistent_parameterization}.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{inconsistent_parameterization}
+  \caption{Simulated Data with Inconsistent Parameterization}
+  \label{fig:inconsistent_parameterization}
+\end{figure}
 
@@
 
-\begin{figure}[h]
+\begin{figure}%
   \centering
-  \includegraphics[width=\textwidth]{adni_lhipp}
-  \caption{Left Hippocampus, Cognitively Healthy ADNI Participant}
+  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
+  \hfill
+  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}
+  \hfill
+  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_pme_plot}}}
+
+  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant}
   \label{fig:adni_result}
@@
   \centering
-  \includegraphics[width=\textwidth]{adni_lhipp_cross_section}
+  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
   \caption{Left Hippocampus Cross Section}

### 2023-11-10 paper/main.tex
Updates from Overleaf

--- 2023-10-18 paper/main.tex
@@
   \end{equation}
-where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
+where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$ (also known as ``distributions" in the mathematical community, see Chapter 6 of \cite{rudin1991functional} for details), and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
 \end{definition}

### 2023-11-10 paper/main.tex
update simulation table orientation

@@
     8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
-    9 & 2 & 3 & $\left(\theta\text{sin} \ (\beta_1r_1)\text{cos} \ (\beta_2r_2), \ \theta\text{sin} \ (\beta_1)\text{sin} \ (\beta_2r_2), \ \theta\text{cos} \ (\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
     \hline
@@
 
-\zielinski{It may be necessary to put some of the simulation figures in an appendix.
-Additionally, the figures are still very much a work in progress and are serving more of a placeholder status at the moment.
-Any suggestions regarding how to improve the clarity and depth perception of the figures would be greatly appreciated!}
-
 \begin{table}[h]
-\footnotesize
   \centering
-  \begin{tabular}{|c c c c c c c c c c|}
+  \begin{tabular}{|c c c c c|}
     \hline
-    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
+    Case & Data & LPME & PME & PC/PS \\
     \hline
-    Data & 0.223 (0.256) & 0.514 (0.408) & 0.446 (0.445) & 30.7 (88.1) & 0.980 (0.771) & 1.43 (6.04) & 0.580 (0.839) & 0.226 (0.275) \\
-    LPME & 0.125 (0.161) & 0.384 (0.648) & 0.401 (0.446) & 27.7 (260) & 0.791 (0.845) & 1.20 (5.66) & 4.07 (3.30) & 0.136 (0.169) \\
-    PME & 0.268 (0.850) & 0.843 (1.93) & 0.507 (0.594) & 30.7 (88.2) & 1.04 (1.07) & 1.47 (6.06) & 7.37 (1.14) & 0.242 (0.311) \\
-    PC/PS & 0.189 (0.245) & 0.600 (0.296) & 0.412 (0.423) & 30.6 (88.1) & 0.934 (0.713) & 1.01 (2.11) & 1.95 (0.800) & 0.274 (0.243) \\
+    1 & 0.223 (0.256) & 0.125 (0.161) & 0.268 (0.850) & 0.189 (0.245) \\
+    2 & 0.514 (0.408) & 0.384 (0.648) & 0.843 (1.93) & 0.600 (0.296) \\
+    3 & 0.446 (0.445) & 0.401 (0.446) & 0.507 (0.594) & 0.412 (0.423) \\
+    4 & 30.7 (88.1) & 27.7 (260) & 30.7 (88.2) & 30.6 (88.1) \\
+    5 & 0.980 (0.771) & 0.791 (0.845) & 1.04 (1.07) & 0.934 (0.713) \\
+    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & 1.01 (2.11) \\
+    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & 1.95 (0.800) \\
+    8 & 0.226 (0.275) & 0.136 (0.169) & 0.242 (0.311) & 0.274 (0.243) \\
     \hline
@@
 \begin{table}[h]
-\footnotesize
   \centering
-  \begin{tabular}{|c c c c c c c c c c|}
+  \begin{tabular}{|c c c c c|}
     \hline
-    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 \\
+    Case & Data & LPME & PME & PC/PS \\
     \hline
-    Data & 0.146 (0.233) & 0.467 (0.665) & 0.291 (0.601) & 4.26 (21.2) & 0.895 (1.33) & 0.284 (1.06) & 0.145 (0.552) & 0.110 (0.325) \\
-    LPME & 0.074 (0.122) & 0.248 (0.461) & 0.239 (0.564) & 3.29 (13.0) & 0.584 (1.08) & 0.273 (0.891) & 2.96 (4.65) & 0.074 (0.208) \\
-    PME & 0.131 (0.258) & 0.516 (0.750) & 0.317 (0.640) & 4.23 (21.1) & 0.894 (1.36) & 0.316 (1.04) & 6.92 (1.17) & 0.115 (0.331) \\
-    PC/PS & 0.118 (0.206) & 0.564 (0.419) & 0.264 (0.584) & 4.22 (21.2) & 0.821 (1.22) & 0.557 (0.387) & 1.58 (0.514) & 0.172 (0.234) \\
+    1 & 0.146 (0.233) & 0.074 (0.122) & 0.131 (0.258) & 0.118 (0.206) \\
+    2 & 0.467 (0.665) & 0.248 (0.516) & 0.516 (0.750) & 0.564 (0.419) \\
+    3 & 0.291 (0.601) & 0.239 (0.564) & 0.317 (0.640) & 0.264 (0.584) \\
+    4 & 4.26 (21.2) & 3.29 (13.0) & 4.23 (21.1) & 4.22 (21.2) \\
+    5 & 0.895 (1.33) & 0.584 (1.08) & 0.894 (1.36) & 0.821 (1.22) \\
+    6 & 0.284 (1.06) & 0.273 (0.891) & 0.316 (1.04) & 0.557 (0.387) \\
+    7 & 0.145 (0.552) & 2.96 (4.65) & 6.92 (1.17) & 1.58 (0.514) \\
+    8 & 0.110 (0.325) & 0.074 (0.208) & 0.115 (0.331) & 0.172 (0.234) \\
     \hline

### 2023-12-14 paper/main.tex

@@
 In addition, applying the methods in high-dimensional settings requires is problematic.
-Motivated by the penalization approaches presented by in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
+Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
 In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper.
@@
 Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.
+The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{sim_case1_plot2}
-  \caption{Longitudinal Simulated Data with $D = 2$}
+  \includegraphics[scale=0.4]{sim_case1_plot2}
+  \caption{Longitudinal Simulated Data with $D = 2$. Data from the function ADD FUNCTION are simulated at 5 time points. Random noise is added both in time and space as ADD FUNCTION.}
   \label{fig:sample_data}
@@
 
-\begin{figure}%
+\begin{figure}[ht]
   \centering
-  \subfloat[\centering Data]{{\includegraphics[width=9cm]{sim_case1_plot2}}}%
+  \subfloat[\centering Data]{{\includegraphics[width=6.5cm]{sim_case1_plot2}}}%
   \hfill
-  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=9cm]{lpme_data_reduction}}}
+  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=6.5cm]{lpme_data_reduction}}}
   \vfill
-  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=9cm]{lpme_initialization}}}
+  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=6.5cm]{lpme_initialization}}}
   \hfill
-  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=9cm]{lpme_fitting}}}
-  \caption{LPME Algorithm Steps}
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=6.5cm]{lpme_fitting}}}
+  \caption{LPME Algorithm Steps.
+In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.
+}
   \label{fig:lpme_steps}
@@
 To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
+Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_steps} (b), along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
 The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
@@
 low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
-This is demonstrated in Figure \ref{fig:inconsistent_parameterization}.
+This is demonstrated in Figure \ref{fig:inconsistent_parameterization} which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
@@
   \centering
-  \includegraphics[width=\textwidth]{inconsistent_parameterization}
+  \includegraphics[scale=0.6]{inconsistent_parameterization}
   \caption{Simulated Data with Inconsistent Parameterization}
@@
 
-To obtain the same parameterization of the manifold over time, instead of using Isomap on the data for each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously.
+To obtain the same parameterization of the manifold over time, instead of using Isomap at each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously.
 This approach is based on the assumption that the shape of the structure, i.e.
@@
 Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
-This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa}.
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_steps} (c).
 At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
@@
 
-Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
+Once $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
 In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
@@
 
-This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
-Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
-This source of error is avoided using our proposed data augmentation approach.
-Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold.
-This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances.
-However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.
-
-\begin{figure}%
+\begin{figure}[ht]
   \centering
@@
 \end{figure}
+
+This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
+Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
+This source of error is avoided using our proposed data augmentation approach.
+Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold.
+This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances.
+However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.
 
@@
 
-\begin{table}
+\begin{table}[ht]
 \small
@@
 
-A factorial design is used to run the simulation studies, with the factor levels set as follows:
-\begin{itemize}
-  \item $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$
-  \item Study duration: $\left\{1, 2, 5\right\}$
-  \item Interval between images: $\left\{0.1, 0.25, 0.5\right\}$
-  \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
-\end{itemize}
+A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$, 2) study duration: $\left\{1, 2, 5\right\}$, 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$, 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal.
 The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times.
@@
     \hline
-    1 & 0.223 (0.256) & 0.125 (0.161) & 0.268 (0.850) & 0.189 (0.245) \\
-    2 & 0.514 (0.408) & 0.384 (0.648) & 0.843 (1.93) & 0.600 (0.296) \\
-    3 & 0.446 (0.445) & 0.401 (0.446) & 0.507 (0.594) & 0.412 (0.423) \\
-    4 & 30.7 (88.1) & 27.7 (260) & 30.7 (88.2) & 30.6 (88.1) \\
-    5 & 0.980 (0.771) & 0.791 (0.845) & 1.04 (1.07) & 0.934 (0.713) \\
-    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & 1.01 (2.11) \\
-    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & 1.95 (0.800) \\
-    8 & 0.226 (0.275) & 0.136 (0.169) & 0.242 (0.311) & 0.274 (0.243) \\
+    1 & 0.223 (0.256) & {\bf 0.125 (0.161)} & 0.268 (0.850) & 0.189 (0.245) \\
+    2 & 0.514 (0.408) & {\bf 0.384 (0.648)} & 0.843 (1.93) & 0.600 (0.296) \\
+    3 & 0.446 (0.445) & {\bf 0.401 (0.446)} & 0.507 (0.594) & 0.412 (0.423) \\
+    4 & 30.7 (88.1) & {\bf 27.7 (260)} & 30.7 (88.2) & 30.6 (88.1) \\
+    5 & 0.980 (0.771) & {\bf 0.791 (0.845)} & 1.04 (1.07) & 0.934 (0.713) \\
+    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & {\bf 1.01 (2.11)} \\
+    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & {\bf 1.95 (0.800)} \\
+    8 & 0.226 (0.275) & {\bf 0.136 (0.169)} & 0.242 (0.311) & 0.274 (0.243) \\
     \hline
   \end{tabular}
-  \caption{MSD Comparison to True Values, Mean (SD).}
+  \caption{MSD comparison to true values, Mean (SD). For each case, the lowest algorithm-specific mean (SD) are highlighted in bold. }
   \label{table:simulation_results_mean}
@@
   \includegraphics[width=\textwidth]{sim_case1}
-  \caption{Simulation Case 1}
+  \includegraphics[width=\textwidth]{sim_case5}
+  \caption{Simulation Cases 1 and 5.}
   \label{fig:sim_case1}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth]{sim_case5}
-  \caption{Simulation Case 5}
-  \label{fig:sim_case4}
 \end{figure}
@@
 
-\zielinski{It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on.
-This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold.
-In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point.
-
-This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold.
-When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.}
-
 \section{Longitudinal Segmentation of MRI images in Alzheimer's Disease Studies}\label{s:application}
@@
 The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.
-Finally, the data were augmented with spherical coordinates in a manner similar to that described in the previous section to avoid the need to fit to a self-intersecting manifold.
+Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.
 
-\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}
+%\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}
 
@@
 
-\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}
+%\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}
 
@@
 We also suggest a data augmentation approach that may allow for the circumvention of the inability to fit principal curve-based methods to some self-intersecting manifolds.
+
+It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on.
+This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold.
+In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point.
+This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold.
+When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.
 
@@
 
-\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction.
+%\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction.
 The smoothness can be represented by the covariance kernel we use in the GPR.

### 2023-12-14 paper/main.tex
Updates from Overleaf

@@
 
-Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
 In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
 After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
-If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement was much higher for FIRST and FreeSurfer than for manual segmentation.
+If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
 \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained.
 In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
-To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
+To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
 
@@
 We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.
+
 Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
@@
 Section \ref{s:simulations} demonstrates the performance of this approach on simulated data.
-In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus in individuals with Alzheimer's disease (AD).
+In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus and thalamus in individuals with Alzheimer's disease.
 The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.
@@
 However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
-In addition, applying the methods in high-dimensional settings requires is problematic.
+In addition, applying the methods in high-dimensional settings is problematic.
 Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
@@
 First is the expected distance from each point in the cloud to the fitted manifold.
-In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on 3-dimensional brain region of interest.
-While the second element of the equation imposes the smoothness penalty on the estimated manifold.
+In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on the 3-dimensional brain region of interest.
+The second element of the equation imposes the smoothness penalty on the estimated manifold.
 The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before.
@@
 In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
-It is important to note, that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.
+It is important to note that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.
 
@@
 We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies.
-In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e.
+We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e.
 $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest.
-Now consider that the data are observed longitudinally, during several visits by a study participant.
+Now consider that the data are observed longitudinally during several visits by a study participant.
 Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.

### 2023-12-15 paper/main.tex
paper updates

@@
   \includegraphics[scale=0.4]{sim_case1_plot2}
-  \caption{Longitudinal Simulated Data with $D = 2$. Data from the function ADD FUNCTION are simulated at 5 time points. Random noise is added both in time and space as ADD FUNCTION.}
+  \caption{Longitudinal Simulated Data with $D = 2$.
+Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points.
+Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
   \label{fig:sample_data}
@@
 modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
-The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$ which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
+The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
 In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
@@
 Given the framework proposed in the previous Section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}).
-Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping.
+Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).
+At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping.
 Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space.
@@
 For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
-However, implementation of commonly used approaches for initialization in our setting can be problematic.
+However, implementation of commonly used approaches for initialization without considering the nuances posed by longitudinal data can be problematic.
 For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e.
-the brain region of interest at each time point) is likely to result in inconsistent parameterization directions (i.e.
+the brain region of interest at each time point) is likely to result in inconsistent parameterization directions across time points (i.e.
 low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
-This is demonstrated in Figure \ref{fig:inconsistent_parameterization} which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}.
+This is demonstrated in Figure \ref{fig:inconsistent_parameterization}, which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
@@
   \includegraphics[scale=0.6]{inconsistent_parameterization}
-  \caption{Simulated Data with Inconsistent Parameterization}
+  \caption{Simulated data with $D = 2$.
+The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
+The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates.
+Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps, as seen at two time points.}
   \label{fig:inconsistent_parameterization}
@@
 
-To obtain the same parameterization of the manifold over time, instead of using Isomap at each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously.
+To obtain the same parameterization of the manifold over time, instead of using Isomap at each time point separately, we propose using PME to develop a parameterization of the data taken from the first time point, then using the projection index associated with this estimate to find initial parameterizations of the centers for the remaining time points.
 This approach is based on the assumption that the shape of the structure, i.e.
 the brain region of interest, remains relatively stable over time.
-In Section \ref{s:simulations}, we consider some deviations from this assumption and their effect on the results and we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.
+In Section \ref{s:simulations}, we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure, and we consider some deviations from this assumption and their effect on the results.
 
@@
 
-Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point.
-Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run.
-This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_steps} (c).
+Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using the cluster centers, weights and parameterizations associated with each time point to initialize a PME run.
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_steps} (c).
+The fit of each $\hat{f}_t$ is measured using $\tau_t$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\hat{f}_t$.
 At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
@@
 At each time point, the estimated principal manifold has a spline form.
-As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points used to fit them.
+As detailed by \cite{greenSilverman1994}, smoothing splines can be considered to have knots at the points used to fit them.
 This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical.
@@
 We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where $\tau_t$ denote the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where the weights for each time point $t$ are equal to
 $$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
@@
 
-  Apply Isomap to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
-Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
+  Apply PME to parameterize $\left\{\mu_{j, 1}\right\}_{j = 1}^{N_1}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1}^{N_1}$ and use $\pi_{f_1(0)}$ to parameterize $\left\{\mu_{j, t}\right\}_{j=1, t = 1}^{N_t, T}$ by $\left\{r_{j, t}\right\}_{j=1, t=2}^{N_t, T}$.
+Formally set $\pi_{f_1(0)}(\mu_{j, 1}) \gets r_{j, 1}$ for $j = 1, 2, \dots, N_1$, and $\pi_{f_{t}(0)}(\mu_{j, t}) \gets \pi_{f_1(0)}(\mu_{j, t})$ for $j = 1, 2, \dots, N_t$, $t = 2, \dots, T$\;
 
@@
 Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds.
-Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency property.
+Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition.
 To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$.
@@
 
-Visualizations of the results from some of the simulation cases, truncated for concision to include only integer-valued time points, are shown in Figures \ref{fig:sim_case1}, \ref{fig:sim_case4} and \ref{fig:sim_case6}.
+Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
     \hline
-    1 & 0.146 (0.233) & 0.074 (0.122) & 0.131 (0.258) & 0.118 (0.206) \\
-    2 & 0.467 (0.665) & 0.248 (0.516) & 0.516 (0.750) & 0.564 (0.419) \\
-    3 & 0.291 (0.601) & 0.239 (0.564) & 0.317 (0.640) & 0.264 (0.584) \\
-    4 & 4.26 (21.2) & 3.29 (13.0) & 4.23 (21.1) & 4.22 (21.2) \\
-    5 & 0.895 (1.33) & 0.584 (1.08) & 0.894 (1.36) & 0.821 (1.22) \\
-    6 & 0.284 (1.06) & 0.273 (0.891) & 0.316 (1.04) & 0.557 (0.387) \\
-    7 & 0.145 (0.552) & 2.96 (4.65) & 6.92 (1.17) & 1.58 (0.514) \\
-    8 & 0.110 (0.325) & 0.074 (0.208) & 0.115 (0.331) & 0.172 (0.234) \\
+    1 & 0.146 (0.233) & {\bf 0.074 (0.122)} & 0.131 (0.258) & 0.118 (0.206) \\
+    2 & 0.467 (0.665) & {\bf 0.248 (0.516)} & 0.516 (0.750) & 0.564 (0.419) \\
+    3 & 0.291 (0.601) & {\bf 0.239 (0.564)} & 0.317 (0.640) & 0.264 (0.584) \\
+    4 & 4.26 (21.2) & {\bf 3.29 (13.0)} & 4.23 (21.1) & 4.22 (21.2) \\
+    5 & 0.895 (1.33) & {\bf 0.584 (1.08)} & 0.894 (1.36) & 0.821 (1.22) \\
+    6 & 0.284 (1.06) & {\bf 0.273 (0.891)} & 0.316 (1.04) & 0.557 (0.387) \\
+    7 & 0.145 (0.552) & 2.96 (4.65) & 6.92 (1.17) & {\bf 1.58 (0.514)} \\
+    8 & 0.110 (0.325) & {\bf 0.074 (0.208)} & 0.115 (0.331) & 0.172 (0.234) \\
     \hline
   \end{tabular}
-  \caption{MSD Comparison to True Values, Median (IQR).}
+  \caption{MSD comparison to true values, Median (IQR). The lowest algorithm-specific median (IQR) are highlighted in bold.}
   \label{table:simulation_results_median}
@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{}
+\end{figure}
+
 \section{Discussion}\label{s:discussion}
@@
 Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
-Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section 4.
+Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.
 

### 2023-12-15 paper/jsm_2024.tex
Updates from Overleaf

--- 2023-12-15 paper/main.tex
@@
 Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.
-Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.
+Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$.
+In this case, data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points.
+Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.
 The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.4]{sim_case1_plot2}
-  \caption{Longitudinal Simulated Data with $D = 2$.
-Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points.
-Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
-  \label{fig:sample_data}
-\end{figure}
 
@@
   \centering
-  \subfloat[\centering Data]{{\includegraphics[width=6.5cm]{sim_case1_plot2}}}%
+  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=6.5cm]{sim_case1_plot2}}%
   \hfill
-  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=6.5cm]{lpme_data_reduction}}}
+  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=6.5cm]{lpme_data_reduction}}
   \vfill
-  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=6.5cm]{lpme_initialization}}}
+  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=6.5cm]{lpme_initialization}}%
   \hfill
-  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=6.5cm]{lpme_fitting}}}
-  \caption{LPME Algorithm Steps.
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=6.5cm]{lpme_fitting}}
+  \caption{LPME Algorithm.
 In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.
@@
 To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_steps} (b), along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
+Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
 The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
@@
 low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
-This is demonstrated in Figure \ref{fig:inconsistent_parameterization}, which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}.
+This is demonstrated in Figure \ref{fig:inconsistent_parameterization}, which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:lpme_step1}.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
@@
   \centering
-  \includegraphics[scale=0.6]{inconsistent_parameterization}
+  \includegraphics[height=7cm]{inconsistent_parameterization}
   \caption{Simulated data with $D = 2$.
-The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
+The curves in this plot represent embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
 The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates.
@@
 Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using the cluster centers, weights and parameterizations associated with each time point to initialize a PME run.
-This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_steps} (c).
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_step3}.
 The fit of each $\hat{f}_t$ is measured using $\tau_t$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\hat{f}_t$.
@@
   \centering
-  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
+  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[height=9cm]{unit_circle_D2}}}%
   \hfill
-  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
+  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[height=9cm]{unit_circle_D3}}}
   \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
@@
 
-Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}.
+Visualizations of the results from two of the simulation cases, truncated for concision, are shown in Figure \ref{fig:sim_case1}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
 
-\begin{table}[h]
-  \centering
-  \begin{tabular}{|c c c c c|}
-    \hline
-    Case & Data & LPME & PME & PC/PS \\
-    \hline
-    1 & 0.146 (0.233) & {\bf 0.074 (0.122)} & 0.131 (0.258) & 0.118 (0.206) \\
-    2 & 0.467 (0.665) & {\bf 0.248 (0.516)} & 0.516 (0.750) & 0.564 (0.419) \\
-    3 & 0.291 (0.601) & {\bf 0.239 (0.564)} & 0.317 (0.640) & 0.264 (0.584) \\
-    4 & 4.26 (21.2) & {\bf 3.29 (13.0)} & 4.23 (21.1) & 4.22 (21.2) \\
-    5 & 0.895 (1.33) & {\bf 0.584 (1.08)} & 0.894 (1.36) & 0.821 (1.22) \\
-    6 & 0.284 (1.06) & {\bf 0.273 (0.891)} & 0.316 (1.04) & 0.557 (0.387) \\
-    7 & 0.145 (0.552) & 2.96 (4.65) & 6.92 (1.17) & {\bf 1.58 (0.514)} \\
-    8 & 0.110 (0.325) & {\bf 0.074 (0.208)} & 0.115 (0.331) & 0.172 (0.234) \\
-    \hline
-  \end{tabular}
-  \caption{MSD comparison to true values, Median (IQR). The lowest algorithm-specific median (IQR) are highlighted in bold.}
-  \label{table:simulation_results_median}
-\end{table}
-
 \begin{figure}[h]
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth]{sim_case7}
-  \caption{Simulation Case 6}
-  \label{fig:sim_case6}
-\end{figure}
-
-The mean, median, and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
+The mean and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_mean}.
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
@@
 
+The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data.
+To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously.
+Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.
+
+Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:adni_lthal_cross_section}.
+While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, Figure \ref{fig:adni_lthal_cross_section} indicates that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+
 \begin{figure}%
@@
   \centering
-  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
+  \subfloat[\centering Left Hippocampus Cross Section]{\includegraphics[height=10cm]{results/adni_plots/adni_cross_section}}%
+  \hfill
+  \subfloat[\centering Estimated Volumes]{\includegraphics[height=10cm]{}}
+  \includegraphics[height=10cm]{adni_plots/adni_cross_section}
   \caption{Left Hippocampus Cross Section}
@@
     \centering
-    \includegraphics[width=\textwidth]{}
+    \includegraphics[height=10cm]{adni_plots/adni_lthal_cross_section.png}
+    \caption{Left Thalamus Cross Section}
+    \label{fig:adni_lthal_cross_section}
+\end{figure}
+
+\begin{figure}%
+  \centering
+  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
+  \hfill
+  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}
+
+  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant}
+  \label{fig:adni_result}
 \end{figure}
@@
 
-\nocite{*}
+%\nocite{*}
 %\bibliographystyle{plain}

### 2023-12-16 paper/jsm_2024.tex

@@
 backend=biber,
-style=authoryear
+style=authoryear,
+url=false,
+isbn=false
 ]{biblatex}
@@
 \addbibresource{references.bib}
+\AtEveryBibitem{\clearfield{month}}
+\AtEveryBibitem{\clearfield{day}}
 
@@
 \begin{algorithm}
+\footnotesize
 \caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
@@
 This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
-This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.
-
-\begin{figure}[ht]
-  \centering
-  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[height=9cm]{unit_circle_D2}}}%
-  \hfill
-  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[height=9cm]{unit_circle_D3}}}
-  \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
-The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
-  \label{fig:unit_circle_augmentation}
-\end{figure}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{sim_case1}
-  \includegraphics[width=\textwidth]{sim_case5}
-  \caption{Simulation Cases 1 and 5.}
+  \includegraphics[height=11cm]{sim_case1}
+  \includegraphics[height=11cm]{sim_case5}
+  \caption{Simulation cases 1 and 5.
+The function estimates found to be closest to the true manifold function, shown in red, can be considered the best fitting.
+In each case, the LPME estimate in blue is closest to the true manifold over all time points.}
   \label{fig:sim_case1}
@@
 As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images.
-This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates.
+This section demonstrates how the LPME algorithm may be used to mitigate the effects of these variations on further analysis.
 To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).
@@
 
-Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}.
-Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data.
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}.
+Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to capture the primary patterns observed in the data.
 There are two main sources of discrepancies between the data and the LPME estimate.
@@
 
-Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
+Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
 This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots.
-Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself.
-This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.
+Notably, it appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with the highest level of curvature.
+
+To understand the impacts of these discrepancies on the estimated volume of the regions, the volume of the hippocampus was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
+The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
+Further details of this approach are beyond the scope of this article, but the function code is available at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
+The volume estimates can be found in Figure \ref{fig:lhipp_volumes}.
+Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
 
@@
 Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:adni_lthal_cross_section}.
-While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, Figure \ref{fig:adni_lthal_cross_section} indicates that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, Figure \ref{fig:lthal_cross_sections} indicates that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.
 
@@
 
-  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant}
+  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant.
+The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
+The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the structure at points with high levels of curvature.}
   \label{fig:adni_result}
@@
   \centering
-  \subfloat[\centering Left Hippocampus Cross Section]{\includegraphics[height=10cm]{results/adni_plots/adni_cross_section}}%
+  \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
   \hfill
-  \subfloat[\centering Estimated Volumes]{\includegraphics[height=10cm]{}}
-  \includegraphics[height=10cm]{adni_plots/adni_cross_section}
+  \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
   \caption{Left Hippocampus Cross Section}
@@
   \centering
-    \includegraphics[height=10cm]{adni_plots/adni_lthal_cross_section.png}
+  \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
+  \hfill
+  \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
   \caption{Left Thalamus Cross Section}
   \label{fig:adni_lthal_cross_section}
-\end{figure}
-
-\begin{figure}%
-  \centering
-  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
-  \hfill
-  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}
-
-  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant}
-  \label{fig:adni_result}
 \end{figure}
@@
 Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.
+The volume estimates obtained from the outputs of the PME and LPME algorithms reflected the insufficient fit of these algorithms to the hippocampus data.
+However, the volume estimation process includes ad hoc approaches to address missingness in the data, and the high level of variability in the estimated volumes of the thalamus despite visual indications of improved fit to this data suggests that this approach may itself be contributing to this variability.
+Further testing of this approach will be needed before the LPME algorithm will be viable for approximating the volume of brain regions.
 
-Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
+Additional work will also be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
 The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
-However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
+However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to the hippocampus or even more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
 The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.

### 2024-02-06 code/functions/estimate_volume.R
remove redundant print statements

@@ -173,8 +173,6 @@ estimate_volume <- function(data, voxel_volume) {
       missing_present <- TRUE
     }
 
-    print(paste("Number of Missing Values:", n_missing))
-    print(paste("Number of Missing Values (Previous):", n_missing_prev))
     if (!(n_missing < n_missing_prev)) {
       if (threshold_idx < length(missingness_thresholds)) {
         threshold_idx <- threshold_idx + 1
@@ -187,35 +185,21 @@ estimate_volume <- function(data, voxel_volume) {
     for (y_idx in 1:length(y_seq)) {
       for (z_idx in 1:length(z_seq)) {
         if (missing_mins[y_idx, z_idx] == TRUE) {
-          print(paste("y_idx", y_idx))
-          print(paste("z_idx", z_idx))
           x_min_vals <- get_surrounding_vals(x_mins, y_idx, z_idx)
-          print("Surrounding Minimum Values")
-          print(x_min_vals)
-          print(paste("Missingness Threshold:", missingness_thresholds[threshold_idx]))
           if (sum(is.na(x_min_vals)) / length(x_min_vals) <= missingness_thresholds[threshold_idx]) {
           # if (sum(!is.na(x_min_vals)) > 1) {
             # x_mins[y_idx, z_idx] <- min(x_min_vals, na.rm = TRUE)
             x_mins[y_idx, z_idx] <- round(mean(x_min_vals, na.rm = TRUE))
-            print(paste("Estimated minimum:", x_mins[y_idx, z_idx]))
             missing_mins[y_idx, z_idx] <- FALSE
             # n_missing <- n_missing - 1
-          } else {
-            print("No minimum value found")
           }
         }
         if (missing_maxs[y_idx, z_idx] == TRUE) {
-          print(paste("y_idx", y_idx))
-          print(paste("z_idx", z_idx))
           x_max_vals <- get_surrounding_vals(x_maxs, y_idx, z_idx)
-          print("Surrounding Maximum Values")
-          print(x_max_vals)
-          print(paste("Missingness Threshold:", missingness_thresholds[threshold_idx]))
           if (sum(!is.na(x_max_vals)) / length(x_max_vals) <= missingness_thresholds[threshold_idx]) {
           # if (sum(!is.na(x_max_vals)) > 1) {
             # x_maxs[y_idx, z_idx] <- max(x_max_vals, na.rm = TRUE)
[…]

### 2024-02-06 paper/main.tex
overleaf updates

@@
 
-We propose a longitudinal extension to principal manifold estimation, with the goal of recovering smooth, longitudinally meaningful manifold estimates over time.
+Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain.
+Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy.
+Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently.
+As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials.
+To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time.
 The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point.
@@
 To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
-Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold.
+Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{high-dimensional data tend to lie in the vicinity of a low-dimensional manifold} (\cite{fefferman2016testing}).
+Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}).
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
@@
 
-Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
+Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
 These methods consider time as an additional dimension in the data, i.e.
@@
 
+The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data.
+To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously.
+Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.
+
+Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:adni_lthal_cross_section}.
+While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:adni_lthal_cross_section} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+
 \begin{figure}%
@@
 
+\begin{figure}%
+  \centering
+  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_data_plot}}}%
+  \hfill
+  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_isomap_plot}}}
+  \hfill
+  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_pme_plot}}}
+
+  \caption{Left Thalamus, Cognitive Healthy ADNI Participant}
+  \label{fig:adni_lthal_result}
+\end{figure}
+
 \begin{figure}[h]
     \centering
-    \includegraphics[width=\textwidth]{}
+    \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section.png}
+    \caption{Left Thalamus Cross Section}
+    \label{fig:adni_lthal_cross_section}
 \end{figure}

### 2024-02-06 paper/jsm_2024.tex

@@
 \affil[2]{Division of Applied Mathematics, Brown University}
+\date{December 15, 2023}
 
@@
   \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
-  \caption{Left Hippocampus Cross Section}
+  \caption{Left Hippocampus Cross Section and Volume Estimates.
+In (a), we can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of surface in the observed data.
+In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it, corresponding to an unexpectedly high volume estimate seen in (b).
+Volume estimates derived from the observed data are consistently higher than those obtained from the embedding functions of PME and LPME estimates, respectively.}
   \label{fig:adni_cross_section}
@@
   \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
-  \caption{Left Thalamus Cross Section}
+  \caption{Left Thalamus Cross Section and Volume Estimates.
+In (a), the cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.
+The volume estimates in (b) show similar general trends to the hippocampus estimates, with the estimates reached using the observed data being higher than the PME- and LPME-based estimates.
+However, the differences between the estimates taken from the PME and LPME results are more comparable.}
   \label{fig:adni_lthal_cross_section}

### 2024-02-09 paper/main.tex
Updates from Overleaf

@@
 For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.
-Image segmentation allows for the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
+
+Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
 Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.

### 2024-02-09 paper/main.tex
Principal Manifold section updates -- explicitly define Hessian matrix

@@
 To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
+
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{high-dimensional data tend to lie in the vicinity of a low-dimensional manifold} (\cite{fefferman2016testing}).
@@
 For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
-Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
+Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
+Given a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $t \to \left( \frac{\partial ^2 f}{\partial t_i \partial t_j}(t) \right)_{1 \leq i, j \leq d}$.
 The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
-Most manifold estimation methods are based on defining a projection index, i.e.
-$\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
+Most manifold estimation methods are based on defining a projection index, denoted $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
 The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.

### 2024-02-09 paper/main.tex
LPME section updates - through tuning

@@
 \usepackage{subcaption}
+\usepackage{bm}
 
@@
 The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices.
+In this section, we introduce an approach to extending this framework to longitudinal clouds of vertices.
 We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies.
@@
 
-To give a more general representation of this problem, let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
+To give a more general representation of this problem, let $\left\{\bm{x}_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
 As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
@@
 \end{definition}
-By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, while imposing smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$.
-These goals correspond to minizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively.
+
+By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and impose smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$.
+These goals correspond to minimizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively.
 Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions.
@@
 
-Given the framework proposed in the previous Section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}).
+Given the framework proposed in the previous section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}).
 Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).
@@
 For example, the PME algorithm can be used to obtain this parameterization.
-In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, thus smoothing over the spatial spline coefficients with respect to time.
+In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, smoothing over the spatial spline coefficients with respect to time.
 This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set.
@@
 A visualization of these steps is shown in Figure \ref{fig:lpme_steps}.
-Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
+Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
 
@@
   \centering
-  \subfloat[\centering Data]{{\includegraphics[width=6.5cm]{sim_case1_plot2}}}%
+  \subfloat[\centering Data]{{\includegraphics[width=8cm]{sim_case1_plot2}}}%
   \hfill
-  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=6.5cm]{lpme_data_reduction}}}
+  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=8cm]{lpme_data_reduction}}}
   \vfill
-  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=6.5cm]{lpme_initialization}}}
+  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=8cm]{lpme_initialization}}}
   \hfill
-  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=6.5cm]{lpme_fitting}}}
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=8cm]{lpme_fitting}}}
   \caption{LPME Algorithm Steps.
@@
 
-Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud.
+In most practical manifold learning tasks data are often quite large, so we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud.
 This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest.
@@
 
-The next step in developing the algorithm for estimation of the longitudinal principal manifold is initialization.
+The next step in the algorithm for estimation of the longitudinal principal manifold is initialization.
 Different algorithms use different approaches for this step.
@@
 The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates.
-Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps, as seen at two time points.}
+Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps.
+For most functions, the $X$ values increase with increasing values for $R$.
+However, two functions have $X$ values that decrease with increasing $R$.
+The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
   \label{fig:inconsistent_parameterization}
@@
 To obtain the same parameterization of the manifold over time, instead of using Isomap at each time point separately, we propose using PME to develop a parameterization of the data taken from the first time point, then using the projection index associated with this estimate to find initial parameterizations of the centers for the remaining time points.
-This approach is based on the assumption that the shape of the structure, i.e.
-the brain region of interest, remains relatively stable over time.
+This approach is based on the assumption that the shape of the structure remains relatively stable over time.
 In Section \ref{s:simulations}, we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure, and we consider some deviations from this assumption and their effect on the results.
@@
 \end{equation}    
-where $\boldsymbol{Y}_t$ is the $N \times N$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
+where $\boldsymbol{Y}_t$ is the $N \times D$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
 
@@
 Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds.
-Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition.
+Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}.
 To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$.

### 2024-02-12 code/generate_figures.R
use theta symbol instead of word for plot label

@@ -34,8 +34,9 @@ plt2 <- lines3D(
   main = "Unit Circle, D = 3",
   xlab = "x",
   ylab = "y",
-  zlab = "theta"
+  zlab = ""
 )
+text3D(-2, -0.8, -2.25, labels = expression(theta), add = TRUE)
 dev.off()
 
 png("paper/figures/unit_circle_D2.png", res = 500, height = 2000, width = 2000)
@@ -58,8 +59,9 @@ lines3D(
   main = "Unit Circle, D = 3",
   xlab = "x",
   ylab = "y",
-  zlab = "theta"
+  zlab = ""
 )
+text3D(-2.25, -0.85, -2.25, labels = expression(theta), add = TRUE)
 dev.off()
 
 sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_025_pernoise_050_n_1000_constant000_run_01.rds")
@@ -480,9 +482,9 @@ dev.off()
 
 ### Following code follows from the sim_error_case1() function in lpme_simulations.R
 
-set.seed(500)
+set.seed(71927)
 source("code/functions/sim_data.R")
-max_time <- 2
+max_time <- 3
 interval <- 0.1
 amp_noise <- 0.5
 shape_noise <- 0.25
@@ -524,7 +526,7 @@ true_vals <- scale(
 )
 time_vals <- unique(sim_df[, 1])
 
-set.seed(10)
+set.seed(81752)
 pme_result <- list()
 pme_vals <- list()
 for (t in 1:length(time_vals)) {
@@ -600,12 +602,12 @@ fig <- plot_ly(
           z = 1
         )
       ),
-      xaxis = list(title = "X"),
+      xaxis = list(title = "X<sub>1</sub>"),
       yaxis = list(title = "R"),
-      zaxis = list(title = "Y")
+      zaxis = list(title = "X<sub>2</sub>")
     ),
     annotations = list(
-      x = 1.06,
+      x = 1.075,
       y = 1.02,
       text = "Time",
       xref = "paper",

### 2024-02-12 paper/main.tex
add algorithm udpates

@@
 \caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
-  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
+  \KwData{Data points $\left\{X_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
 \KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
 \For{$t = 1, 2, \dots, T$} {
-  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
+  Apply HDMDE algorithm with input $\left(\left\{X_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
 }
@@
   Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
-  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
+  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $Y_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
 }
@@
 \For{t = 1, 2, \dots, T} {
-  Set $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
+  Set $Y_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
   Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
-  Set $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
+  Set $w_t = \frac{1}{\tau_t \sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
 }
@@
 
-A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$, 2) study duration: $\left\{1, 2, 5\right\}$, 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$, 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal.
+A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal.
 The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times.
@@
 
-The mean, median, and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
+The mean, median, standard deviation and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.

### 2024-02-12 paper/main.tex
update application, discussion sections, figures

@@
 backend=biber,
-style=authoryear
+style=authoryear,
+url=false,
+isbn=false,
+date=year,
+doi=false,
+pmid=false,
+urldate=false
 ]{biblatex}
@@
   \centering
-  \subfloat[\centering Data]{{\includegraphics[width=8cm]{sim_case1_plot2}}}%
+  \subfloat[\centering Data]{{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}}%
   \hfill
-  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=8cm]{lpme_data_reduction}}}
+  \subfloat[\centering Step 1: Sample Size Reduction]{{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}}
   \vfill
-  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=8cm]{lpme_initialization}}}
+  \subfloat[\centering Step 2: Initialization]{{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}}
   \hfill
-  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=8cm]{lpme_fitting}}}
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}}
   \caption{LPME Algorithm Steps.
@@
   \centering
-  \includegraphics[scale=0.6]{inconsistent_parameterization}
+  \includegraphics[height=7cm]{inconsistent_parameterization}
   \caption{Simulated data with $D = 2$.
@@
 Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
-The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.
+The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in each dimension of Euclidean space and centered around the origin.
 Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.
@@
 
-Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}.
-Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data.
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}.
+Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data.
 There are two main sources of discrepancies between the data and the LPME estimate.
-First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation.
+First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
 The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.
 
-Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
+Figure \ref{fig:lhipp_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
@@
 This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.
+Notably, it also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with the highest levels of curvature.
+
+The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data.
+To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously.
+Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.
+
+To understand the impacts of these discrepancies on the estimated volume of the regions, the volume of the hippocampus was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
+The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
+The interior identification approach developed by \cite{mengPrincipalManifoldEstimation2021} is used to determine whether each voxel may be considered inside the surface of the structure or not.
+The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}.
+The volume estimates can be found in Figure \ref{fig:lhipp_volumes}.
+Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
 
@@
 While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:adni_lthal_cross_section} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
-  \caption{Left Hippocampus Cross Section}
+  \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
+  \hfill
+  \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
+  \caption{Left Hippocampus Cross Section and Volume Estimates.
+In (a), we can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of surface in the observed data.
+In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it, corresponding to an unexpectedly high volume estimate seen in (b).
+Volume estimates derived from the observed data are consistently higher than those obtained from the embedding functions of PME and LPME estimates, respectively.}
   \label{fig:adni_cross_section}
@@
   \centering
-    \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section.png}
-    \caption{Left Thalamus Cross Section}
+  \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
+  \hfill
+  \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
+  \caption{Left Thalamus Cross Section and Volume Estimates.
+In (a), the cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.
+The volume estimates in (b) show similar general trends to the hippocampus estimates, with the estimates reached using the observed data being higher than the PME- and LPME-based estimates.
+However, the differences between the estimates taken from the PME and LPME results are more comparable.}
   \label{fig:adni_lthal_cross_section}
@@
 In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data.
-To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a longitudinal setting.
+To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a general-purpose longitudinal setting.
+We also suggest a data augmentation approach that circumvents the inability to fit principal curve-based methods to self-intersecting manifolds in select settings.
 In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point.
 This improvement was seen across several manifold types.
-We also suggest a data augmentation approach that may allow for the circumvention of the inability to fit principal curve-based methods to some self-intersecting manifolds.
 
-It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on.
-This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold.
+It is key to note here that the MSD values shown in these tables compare the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on.
+This means that a model that performs well using this metric may not perform well when comparing the model's projections onto the manifold with the data.
 In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point.
@@
 First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly.
-Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
+Because the output of PME is dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
 The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern.
-There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.
+Opportunities also exist to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.
 
@@
 Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.
+The volume estimates obtained from the outputs of the PME and LPME algorithms reflected the insufficient fit of these algorithms to the hippocampus data.
+However, the volume estimation process includes ad hoc approaches to address missing data, and the high level of variability in the estimated volumes of the thalamus despite visual indications of improved fit to this data suggests that this approach may itself be contributing to this variability.
+Further testing of this approach will be needed before the LPME algorithm will be viable for approximating the volume of brain regions.
 
-Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
+Additional work will also be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
 The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
-However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
+However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to the hippocampus or more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
 The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.

### 2024-02-12 paper/main.tex
address latex warnings

@@
 date=year,
-doi=false,
-pmid=false,
-urldate=false
+doi=false
 ]{biblatex}
@@
   \centering
-  \subfloat[\centering Data]{{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}}%
+  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}%
   \hfill
-  \subfloat[\centering Step 1: Sample Size Reduction]{{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}}
+  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}
   \vfill
-  \subfloat[\centering Step 2: Initialization]{{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}}
+  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}
   \hfill
-  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}}
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}
   \caption{LPME Algorithm Steps.
@@
 
-Figure \ref{fig:lhipp_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
+Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point.
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.

### 2024-02-12 paper/main.tex
fitting section notation update

@@
 We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where the weights for each time point $t$ are equal to
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients $\boldsymbol{\omega}_t$, where the weights for each time point $t$ are equal to
 $$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$

### 2024-03-02 paper/main.tex

@@
 These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
-We then apply the proposed method to estimating the surfaces of hippocampi of healthy individuals in the Alzheimer’s Disease Neuroimaging Initiative dataset.
+We show that the proposed method improves the estimation of surfaces of hippocampi and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
 
@@
 
-Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{high-dimensional data tend to lie in the vicinity of a low-dimensional manifold} (\cite{fefferman2016testing}).
+Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
 Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}).
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
-For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum.
-To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.
+For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.
 
@@
 Here, we propose an alternative approach to estimating manifolds over time.
-Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension.
+Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension.
 Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
@@
 
-The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework.
+The remainder of this article is laid out as follows.
+In Section \ref{s:PME}, we review the principal manifold framework.
 In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework.
@@
 A visualization of these steps is shown in Figure \ref{fig:lpme_steps}.
-Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
+Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}.
 

### 2024-04-02 paper/main.tex

@@
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
-Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}).
+Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}).
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.

### 2024-04-03 paper/main.tex
add aggregate results for larger adni population

@@
 
-Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:adni_lthal_cross_section}.
-While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:adni_lthal_cross_section} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}.
+While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
 This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.
@@
 
-\begin{figure}[h]
-  \centering
-  \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
-  \hfill
-  \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
-  \caption{Left Hippocampus Cross Section and Volume Estimates.
+% \begin{figure}[h]
+%   \centering
+%   \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
+%   \hfill
+%   \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
+%   \caption{Left Hippocampus Cross Section and Volume Estimates.
 In (a), we can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of surface in the observed data.
@@
 Volume estimates derived from the observed data are consistently higher than those obtained from the embedding functions of PME and LPME estimates, respectively.}
-  \label{fig:adni_cross_section}
+%   \label{fig:adni_cross_section}
+% \end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
+  \caption{Left Hippocampus Cross Section.
+We can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the surface in the observed data.
+In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it.
+Gaps in the surface estimates frequently correspond with inaccurate volume estimates.}
+  \label{fig:lhipp_cross_sections}
 \end{figure}
@@
 
-\begin{figure}[h]
-  \centering
-  \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
-  \hfill
-  \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
-  \caption{Left Thalamus Cross Section and Volume Estimates.
+% \begin{figure}[h]
+%   \centering
+%   \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
+%   \hfill
+%   \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
+%   \caption{Left Thalamus Cross Section and Volume Estimates.
 In (a), the cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.
@@
 However, the differences between the estimates taken from the PME and LPME results are more comparable.}
-  \label{fig:adni_lthal_cross_section}
+%   \label{fig:adni_lthal_cross_section}
+% \end{figure}
+%
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section}
+  \caption{Left Thalamus Cross Section. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
+  \label{fig:lthal_cross_sections}
 \end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{adni_plots/adni_lhipp_volume_comp}
+  \caption{Left Hippocampus Volume Estimates for Six ADNI Participants.
+All volumes are estimated by summing the volumes of all voxels contained within the structure estimate as described by segmented data (red), LPME (blue), and PME (green).
+When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of volume estimates, gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data.
+These gaps may also cause inconsistency in the LPME volume estimates, resulting in higher aggregate variability than for the data or PME estimates.}
+  \label{fig:lhipp_volume_comparison}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{adni_plots/adni_lthal_volume_comp}
+  \caption{Left Thalamus Volume Estimates for Six ADNI Participants.
+Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by segmented data (red), LPME (blue), and PME (green).
+The PME-based volume estimates closely track the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data.
+The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participants 021\_S\_0626 and 053\_S\_0919.}
+  \label{fig:lthal_volume_comparison}
+\end{figure}
+
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{|c c c c|}
+    \hline
+    Structure & Data & LPME & PME  \\
+    \hline
+    Left Hippocampus & 218 & 225 & 191 \\
+    Right Hippocampus & 221 & 245 & 195 \\
+    Left Thalamus & 82.9 & 49.7 & 72.0 \\
+    Right Thalamus & 147 & 116 & 130 \\
+    \hline
+  \end{tabular}
+  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.}
+  \label{table:adni_volume_sds}
+\end{table}
 

### 2024-04-03 code/lpme_adni_results.R

@@ -175,33 +175,33 @@ hipp_info_ts %>%
 hipp_info_sd <- hipp_info %>%
   group_by(patno) %>%
   summarize(
-    data_sd = sd(lhipp_data_vol2),
-    lpme_sd = sd(lhipp_vol_lpme2),
-    pme_sd = sd(lhipp_vol_pme2),
-    data_iqr = quantile(lhipp_data_vol2, 0.75) - quantile(lhipp_data_vol2, 0.25),
-    lpme_iqr = quantile(lhipp_vol_lpme2, 0.75) - quantile(lhipp_vol_lpme2, 0.25),
-    pme_iqr = quantile(lhipp_vol_pme2, 0.75) - quantile(lhipp_vol_pme2, 0.25)
+    lhipp_data_sd = sd(lhipp_data_vol2),
+    lhipp_lpme_sd = sd(lhipp_vol_lpme2),
+    lhipp_pme_sd = sd(lhipp_vol_pme2),
+    rhipp_data_sd = sd(rhipp_data_vol2),
+    rhipp_lpme_sd = sd(rhipp_vol_lpme2),
+    rhipp_pme_sd = sd(rhipp_vol_pme2)
   ) %>%
   ungroup()
 
-hipp_info_sd %>%
+hipp_sd_mean <- hipp_info_sd %>%
   summarize(
-    data_sd_mean = mean(data_sd),
-    lpme_sd_mean = mean(lpme_sd),
-    pme_sd_mean = mean(pme_sd),
-    data_iqr_mean = mean(data_iqr),
-    lpme_iqr_mean = mean(lpme_iqr),
-    pme_iqr_mean = mean(pme_iqr)
+    lhipp_data_sd_mean = mean(lhipp_data_sd),
+    lhipp_lpme_sd_mean = mean(lhipp_lpme_sd),
+    lhipp_pme_sd_mean = mean(lhipp_pme_sd),
+    rhipp_data_sd_mean = mean(rhipp_data_sd),
+    rhipp_lpme_sd_mean = mean(rhipp_lpme_sd),
+    rhipp_pme_sd_mean = mean(rhipp_pme_sd)
   )
 
-hipp_info_sd %>%
+hipp_sd_med <- hipp_info_sd %>%
   summarize(
-    data_sd_med = median(data_sd),
-    lpme_sd_med = median(lpme_sd),
-    pme_sd_med = median(pme_sd),
-    data_iqr_med = median(data_iqr),
[…]

### 2024-04-03 paper/main.tex
update adni volume figure references

@@
 The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}.
-The volume estimates can be found in Figure \ref{fig:lhipp_volumes}.
+The volume estimates can be found in Figure \ref{fig:lhipp_volume_comparison}.
 Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
@@
 While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
-This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.
+This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volume_comparison}, though visible variations between time points remain.
 

### 2024-04-04 code/lpme_adni_results.R
remove lm lines from volume plots

@@ -388,34 +388,34 @@ hipp_info_ts %>%
   ggplot(aes(x = time_from_bl, group = patno)) +
   geom_point(aes(y = lhipp_data_vol2), color = colors[1]) +
   geom_line(aes(y = lhipp_data_vol2), color = colors[1]) +
-  geom_smooth(
-    aes(y = lhipp_data_vol2),
-    color = colors[1],
-    method = "lm",
-    se = FALSE,
-    linetype = "dashed",
-    linewidth = 1
-  ) +
+  # geom_smooth(
+  #   aes(y = lhipp_data_vol2),
+  #   color = colors[1],
+  #   method = "lm",
+  #   se = FALSE,
+  #   linetype = "dashed",
+  #   linewidth = 1
+  # ) +
   geom_point(aes(y = lhipp_vol_lpme2), color = colors[2]) +
   geom_line(aes(y = lhipp_vol_lpme2), color = colors[2]) +
-  geom_smooth(
-    aes(y = lhipp_vol_lpme2),
-    color = colors[2],
-    method = "lm",
-    se = FALSE,
-    linetype = "dashed",
-    linewidth = 1
-  ) +
+  # geom_smooth(
+  #   aes(y = lhipp_vol_lpme2),
+  #   color = colors[2],
+  #   method = "lm",
+  #   se = FALSE,
+  #   linetype = "dashed",
+  #   linewidth = 1
+  # ) +
   geom_point(aes(y = lhipp_vol_pme2), color = colors[3]) +
   geom_line(aes(y = lhipp_vol_pme2), color = colors[3]) +
-  geom_smooth(
-    aes(y = lhipp_vol_pme2),
-    color = colors[3],
-    method = "lm",
-    se = FALSE,
-    linetype = "dashed",
-    linewidth = 1
-  ) +
+  # geom_smooth(
+  #   aes(y = lhipp_vol_pme2),
+  #   color = colors[3],
+  #   method = "lm",
+  #   se = FALSE,
+  #   linetype = "dashed",
+  #   linewidth = 1
+  # ) +
   facet_wrap(~patno) +
   xlab("Time from Baseline Visit (Years)") +
   ylab("Estimated Left Hippocampus Volume")
@@ -426,34 +426,34 @@ thal_info_ts %>%
   ggplot(aes(x = time_from_bl, group = patno)) +
   geom_point(aes(y = lthal_data_vol2), color = colors[1]) +
[…]

### 2024-04-04 paper/main.tex
include regression-adjusted results and update caption

@@
   \centering
-  \begin{tabular}{|c c c c|}
+  \begin{tabular}{|c c c c c|}
     \hline
-    Structure & Data & LPME & PME  \\
+    Structure & Regression-Adjusted & Data & LPME & PME  \\
     \hline
-    Left Hippocampus & 218 & 225 & 191 \\
-    Right Hippocampus & 221 & 245 & 195 \\
-    Left Thalamus & 82.9 & 49.7 & 72.0 \\
-    Right Thalamus & 147 & 116 & 130 \\
+    Left Hippocampus & No & 218 & 225 & 191 \\
+    Right Hippocampus & No & 221 & 245 & 195 \\
+    Left Thalamus & No & 82.9 & 49.7 & 72.0 \\
+    Right Thalamus & No & 147 & 116 & 130 \\
+    Left Hippocampus & Yes & 184 & 197 & 160 \\
+    Right Hippocampus & Yes & 185 & 215 & 163 \\
+    Left Thalamus & Yes & 68.3 & 30.7 & 57.9 \\
+    Right Thalamus & Yes & 124 & 89.9 & 109 \\
     \hline
   \end{tabular}
-  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.}
+  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
+Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.}
   \label{table:adni_volume_sds}

### 2024-04-04 paper/main.tex
rearrange figures
update adni application section with new results

--- 2024-04-03 paper/main.tex
@@
 
-While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}).
-Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals.
-Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected.
-Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process.
+While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
+Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study.
+Follow-up duration for these participants ranged from approximately 6 months to 112 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study.
+To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants.
 Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
@@
 The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data.
-To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously.
-Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.
-
-To understand the impacts of these discrepancies on the estimated volume of the regions, the volume of the hippocampus was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
-The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
-The interior identification approach developed by \cite{mengPrincipalManifoldEstimation2021} is used to determine whether each voxel may be considered inside the surface of the structure or not.
-The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}.
-The volume estimates can be found in Figure \ref{fig:lhipp_volumes}.
-Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
-
-The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data.
-To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously.
+To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same set of individuals described previously.
 Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.
@@
 While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
-This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.
 
-\begin{figure}%
-  \centering
-  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
-  \hfill
-  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}
-  \hfill
-  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_pme_plot}}}
+To understand how the subcortical structure estimates obtained by the LPME and PME algorithms may impact the corresponding estimates of the volumes of these structures, the volume of each structure was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
+The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
+The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}.
+It is important to note that, instead of using the volume estimates produced by FSL's FIRST method during the segmentation process, the same volume estimation method used for the LPME- and PME-based estimates is used to reach estimates based on the raw data as well to align processing steps as closely as possible.
 
-  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant.
-The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
-The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the structure at points with high levels of curvature.}
-  \label{fig:adni_result}
-\end{figure}
+The volume estimates for randomly selected individuals can be found in Figures \ref{fig:lhipp_volume_comparison} and \ref{fig:lthal_volume_comparison}.
+Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
+In Figure \ref{fig:lhipp_volume_comparison}, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
+All three sources show substantial variability between study visits, particularly for subjects 137\_S\_0283 and 137\_S\_0686.
+In Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
+However, the LPME estimates demonstrate the ability to smooth over this variability, resulting in volume estimates that are not as sensitive to differences between time points.
+We also see that when basing volume estimates on the more reliable LPME fit to the thalamus data, LPME yields volume estimates similar in magnitude to those obtained from the raw data and PME estimates.
+
+A summary of the between-visit variability in the volume estimates of the subcortical structures considered here is shown in Table \ref{table:adni_volume_sds}.
+In cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e.
+raw data, LPME, and PME).
+We then took the mean of these values for each estimate source.
+When regression adjustments were used, instead of considering the raw volume estimates, we estimated the standard deviation of the residuals of a simple linear regression regressing volume on time.
+This additional step gives a better indication of smoothness when the volume of the structure cannot be assumed to be constant over time, as is the case for those with AD.
+The table shows that PME yields the smoothest volume estimates for both hippocampi.
+This conclusion aligns with the visual evidence of inconsistency in the LPME estimates caused by the irregular shape of the hippocampus.
+However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
+  %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
+  \includegraphics[height=8cm]{adni_plots/adni_cross_section}
   \caption{Left Hippocampus Cross Section.
@@
   \label{fig:lhipp_cross_sections}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  %\includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section}
+  \includegraphics[height=8cm]{adni_plots/adni_lthal_cross_section}
+  \caption{Left Thalamus Cross Section. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
+  \label{fig:lthal_cross_sections}
+\end{figure}
+
+\begin{figure}%
+  \centering
+  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
+  \hfill
+  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}
+  \hfill
+  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_pme_plot}}}
+
+  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant.
+The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
+The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the structure at points with high levels of curvature.}
+  \label{fig:adni_result}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section}
-  \caption{Left Thalamus Cross Section. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
-  \label{fig:lthal_cross_sections}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
   \includegraphics[width=\textwidth]{adni_plots/adni_lhipp_volume_comp}
@@
   \centering
-  \begin{tabular}{|c c c c|}
+  \begin{tabular}{|c c c c c|}
     \hline
-    Structure & Data & LPME & PME  \\
+    Structure & Regression-Adjusted & Data & LPME & PME  \\
     \hline
-    Left Hippocampus & 218 & 225 & 191 \\
-    Right Hippocampus & 221 & 245 & 195 \\
-    Left Thalamus & 82.9 & 49.7 & 72.0 \\
-    Right Thalamus & 147 & 116 & 130 \\
+    Left Hippocampus & No & 218 & 225 & \textbf{191} \\
+    Right Hippocampus & No & 221 & 245 & \textbf{195} \\
+    Left Hippocampus & Yes & 184 & 197 & \textbf{160} \\
+    Right Hippocampus & Yes & 185 & 215 & \textbf{163} \\
+    Left Thalamus & No & 82.9 & \textbf{49.7} & 72.0 \\
+    Right Thalamus & No & 147 & \textbf{116} & 130 \\
+    Left Thalamus & Yes & 68.3 & \textbf{30.7} & 57.9 \\
+    Right Thalamus & Yes & 124 & \textbf{89.9} & 109 \\
     \hline
   \end{tabular}
-  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.}
+  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
+Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.
+}
   \label{table:adni_volume_sds}

### 2024-04-04 paper/main.tex
update results for 2-year duration threshold

@@
     \hline
-    Left Hippocampus & No & 218 & 225 & \textbf{191} \\
-    Right Hippocampus & No & 221 & 245 & \textbf{195} \\
-    Left Hippocampus & Yes & 184 & 197 & \textbf{160} \\
-    Right Hippocampus & Yes & 185 & 215 & \textbf{163} \\
-    Left Thalamus & No & 82.9 & \textbf{49.7} & 72.0 \\
-    Right Thalamus & No & 147 & \textbf{116} & 130 \\
-    Left Thalamus & Yes & 68.3 & \textbf{30.7} & 57.9 \\
-    Right Thalamus & Yes & 124 & \textbf{89.9} & 109 \\
+    Left Hippocampus & No & 213 & 225 & \textbf{187} \\
+    Right Hippocampus & No & 210 & 236 & \textbf{186} \\
+    Left Hippocampus & Yes & 173 & 190 & \textbf{150} \\
+    Right Hippocampus & Yes & 174 & 206 & \textbf{154} \\
+    Left Thalamus & No & 78.3 & \textbf{49.0} & 67.8 \\
+    Right Thalamus & No & 139 & \textbf{112} & 122 \\
+    Left Thalamus & Yes & 62.8 & \textbf{29.0} & 53.0 \\
+    Right Thalamus & Yes & 115 & \textbf{85.3} & 99.9 \\
     \hline

### 2024-04-05 paper/main.tex
update discussion section

@@
 To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants.
+Of these remaining participants, 80 were cognitively normal, 126 had mild cognitive impairment, and 44 were diagnosed with AD at their baseline visits.
 Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
@@
 How restrictive this requirement is depends largely on the area of application.
-When used to estimate subcortical surfaces as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.
+When used to estimate subcortical surfaces in a single individual as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.
 
@@
 It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus.
-This shortcoming could potentially be addressed in several ways.
-First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated.
-Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
+There are two potential sources of this difficulty, each suggesting different steps that may be taken to address this issue.
+First, it is possible that this problem is caused by an inability of the PME and LPME algorithms to handle structures with as many sharp curves as the hippocampus.
+This may be addressed in several ways.
+As discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated.
+Also, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
 Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.
-The volume estimates obtained from the outputs of the PME and LPME algorithms reflected the insufficient fit of these algorithms to the hippocampus data.
-However, the volume estimation process includes ad hoc approaches to address missing data, and the high level of variability in the estimated volumes of the thalamus despite visual indications of improved fit to this data suggests that this approach may itself be contributing to this variability.
-Further testing of this approach will be needed before the LPME algorithm will be viable for approximating the volume of brain regions.
 
-Additional work will also be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities.
-The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
-However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to the hippocampus or more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin.
-The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.
-While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
+The second potential source of the algorithms' poor fit to the hippocampus data could be an insufficiency in the data augmentation approach that allows us to fit some self-intersecting manifolds.
+The simple calculations used here, augmenting with polar or spherical coordiantes, allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
+However, while this approach addressed a problem posed by well-behaved circular or spherical structures, such as the thalamus, its success may not extend to settings with more complex data.
+In the case of the hippocampus, we speculate that there may be intersections present in the spherical coordinates, which then result in the overlaps in certain regions of the estimated manifolds.
+Similar types of issues would arise in settings with other complex structures, like a figure 8 shape.
+The development of a procedure to handle these considerations would expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.
+
+Further attention should be paid to the process used to estimate volumes from the PME and LPME outputs if these volume estimates are to be used as outcomes of interest.
+The current procedure includes ad hoc adjustments to maintain the stability of estimates in the presence of gaps in the estimated surface.
+However, alternate approaches may yield less variable results.
+For example, \cite{mengPrincipalManifoldEstimation2021} introduced an interior identification approach that effectively makes use of the explicit representation of the embedding map obtained through the PME and LPME fitting process, and which may be adapted to settings with gaps in the estimated manifold.
+
+Finally, while the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
+Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.
 

### 2024-04-05 code/lpme_adni_results.R
summarize participant study arms

@@ -4,6 +4,21 @@ library(RColorBrewer)
 
 colors <- brewer.pal(3, "Set1")
 
+roster <- read_csv("data/adni_standard/ROSTER.csv")
+roster <- roster %>% 
+  select(
+    -USERDATE,
+    -USERDATE2,
+    -update_stamp
+  )
+adni_status <- read_csv("data/adni_standard/ADNIMERGE.csv")
+
+# adni_status <- full_join(
+#   adni_status,
+#   roster,
+#   by = c("Phase", "ID", "RID", "SITEID")
+# )
+
 sub_dirs <- list.dirs("results", recursive = FALSE)
 
 hipp_info <- data.frame(
@@ -62,7 +77,7 @@ thal_bl <- thal_info %>%
   )
 
 eligible_patnos <- hipp_bl %>%
-  filter(max_date - date_bl > 3) %>%
+  filter(max_date - date_bl > 2) %>%
   select(patno) %>%
   unlist() %>%
   unique()
@@ -250,7 +265,7 @@ hipp_sd_mean <- hipp_info_sd %>%
     rhipp_data_adj_sd_mean = mean(rhipp_data_adj_sd),
     rhipp_lpme_adj_sd_mean = mean(rhipp_lpme_adj_sd),
     rhipp_pme_adj_sd_mean = mean(rhipp_pme_adj_sd)
-  )
+)
 
 print(hipp_sd_mean, width = Inf)
 
@@ -458,3 +473,11 @@ thal_info_ts %>%
   xlab("Time from Baseline Visit (Years)") +
   ylab("Estimated Left Thalamus Volume")
 ggsave("paper/figures/adni_plots/adni_lthal_volume_comp.png", dpi = 1500)
+
+adni_status %>% 
+  filter(
+    PTID %in% eligible_patnos,
+    VISCODE == "bl"
+  ) %>% 
+  group_by(DX_bl) %>% 
+  tally()

### 2024-04-05 paper/main.tex
fix figure reference

@@
 
-Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}.
+Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}.
 While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

### 2024-04-05 paper/smi_2024.tex
add smi paper adaptation

--- 2024-04-05 paper/main.tex
@@
 backend=biber,
-style=authoryear,
+style=ext-authoryear,
 url=false,
@@
 date=year,
-doi=false
+doi=false,
+eprint=false,
+articlein=false
 ]{biblatex}
+
+%\renewbibmacro{in:}{}
+%\DeclareFieldFormat{pages}{#1}
 
@@
 \usepackage{bm}
+\usepackage[margin=0.7in]{geometry}
+\usepackage{titlesec}
+\titlespacing*{\section}{0pt}{0.1\baselineskip}{0.2\baselineskip}
+
+\doublespacing
 
@@
 \usepackage[english]{babel}
-
-\renewcommand{\baselinestretch}{1.0}
-\setlength{\oddsidemargin}{-0.5cm}
-\setlength{\evensidemargin}{-0.5cm}
-\renewcommand{\topmargin}{-2cm}
-\renewcommand{\oddsidemargin}{0mm}
-\renewcommand{\evensidemargin}{0mm}
-\renewcommand{\textwidth}{180mm}
-\renewcommand{\textheight}{240mm}
+%
+% \renewcommand{\baselinestretch}{1.0}
+% \setlength{\oddsidemargin}{-0.5cm}
+% \setlength{\evensidemargin}{-0.5cm}
+% \renewcommand{\topmargin}{-2cm}
+% \renewcommand{\oddsidemargin}{0mm}
+% \renewcommand{\evensidemargin}{0mm}
+% \renewcommand{\textwidth}{180mm}
+% \renewcommand{\textheight}{240mm}
 
@@
 These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
-We show that the proposed method improves the estimation of surfaces of hippocampi and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
+We apply the proposed method to estimate the volumes of hippocampi and thalamuses of participants in the Alzheimer's Disease Neuroimaging Initiative dataset.
 
@@
 These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}).
-For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.
+For example, it is common to observe atrophy in the hippocampus of those with AD or cognitive impairment related to other causes.
 
-Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
-Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
-However, this approach is highly time and resource intensive.
-When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
+Image segmentation is the process by which subcortical structures are extracted from MRI images of the brain, allowing the identification of differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
+Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation.
+However, this approach is highly time and resource intensive, and may impose prohibitive costs when analyzing large numbers of images.
 Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
@@
 
-Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+Failures of automatic segmentation algorithms may have substantial effects when analyzing large datasets of images where manual quality control may not be feasible.
 In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
 After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
-If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
-\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained.
 In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
-To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
 
-Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods.
+Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used.
 Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data.
@@
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
-Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}).
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
@@
 
-Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
-These methods consider time as an additional dimension in the data, i.e.
-they include time in the dimension reduction process.
+Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
+These methods consider time as an additional dimension in the data and include it in the dimension reduction process.
 This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
-Here, we propose an alternative approach to estimating manifolds over time.
-Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension.
+Here, we propose an alternative approach that estimates manifolds over time while maintaining the interpretability of the time dimension.
 Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
-We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+We then impose smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
 We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.
@@
 
-The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
-However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
-In addition, applying the methods in high-dimensional settings is problematic.
+The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
+However, this approach has several limitations, and applying the methods in high-dimensional settings is problematic (\cite{duchamp1996extremal}).
 Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
+
 In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper.
@@
   \end{equation}
-where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$ (also known as ``distributions" in the mathematical community, see Chapter 6 of \cite{rudin1991functional} for details), and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
+where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
 \end{definition}
@@
 Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.
-Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.
+Figure \ref{fig:sim_case1} displays a simplified example of this setting using simulated data where $D = 2$.
 The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.4]{sim_case1_plot2}
-  \caption{Longitudinal Simulated Data with $D = 2$.
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[scale=0.4]{sim_case1_plot2}
+%   \caption{Longitudinal Simulated Data with $D = 2$.
 Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points.
 Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
-  \label{fig:sample_data}
-\end{figure}
-
+%   \label{fig:sample_data}
+% \end{figure}
+%
 Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
@@
   \centering
-  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}%
+  \subfloat[\centering Data]{\label{fig:sim_case1} \includegraphics[width=8cm]{sim_case1_plot2}}%
   \hfill
@@
 low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
-This is demonstrated in Figure \ref{fig:inconsistent_parameterization}, which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[height=7cm]{inconsistent_parameterization}
-  \caption{Simulated data with $D = 2$.
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[height=7cm]{inconsistent_parameterization}
+%   \caption{Simulated data with $D = 2$.
 The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
@@
 The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
-  \label{fig:inconsistent_parameterization}
-\end{figure}
+%   \label{fig:inconsistent_parameterization}
+% \end{figure}
 
@@
 This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
-This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.
 
-\begin{figure}[ht]
-  \centering
-  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
-  \hfill
-  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
-  \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
+% \begin{figure}[ht]
+%   \centering
+%   \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=8.75cm]{unit_circle_D2}}}%
+%   \hfill
+%   \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=8.75cm]{unit_circle_D3}}}
+%   \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
 The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
-  \label{fig:unit_circle_augmentation}
-\end{figure}
+%   \label{fig:unit_circle_augmentation}
+% \end{figure}
 
@@
 
-Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}.
+Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figure \ref{fig:sim_case1}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
 
-\begin{table}[h]
-  \centering
-  \begin{tabular}{|c c c c c|}
-    \hline
-    Case & Data & LPME & PME & PC/PS \\
-    \hline
-    1 & 0.146 (0.233) & {\bf 0.074 (0.122)} & 0.131 (0.258) & 0.118 (0.206) \\
-    2 & 0.467 (0.665) & {\bf 0.248 (0.516)} & 0.516 (0.750) & 0.564 (0.419) \\
-    3 & 0.291 (0.601) & {\bf 0.239 (0.564)} & 0.317 (0.640) & 0.264 (0.584) \\
-    4 & 4.26 (21.2) & {\bf 3.29 (13.0)} & 4.23 (21.1) & 4.22 (21.2) \\
-    5 & 0.895 (1.33) & {\bf 0.584 (1.08)} & 0.894 (1.36) & 0.821 (1.22) \\
-    6 & 0.284 (1.06) & {\bf 0.273 (0.891)} & 0.316 (1.04) & 0.557 (0.387) \\
-    7 & 0.145 (0.552) & 2.96 (4.65) & 6.92 (1.17) & {\bf 1.58 (0.514)} \\
-    8 & 0.110 (0.325) & {\bf 0.074 (0.208)} & 0.115 (0.331) & 0.172 (0.234) \\
-    \hline
-  \end{tabular}
-  \caption{MSD comparison to true values, Median (IQR). The lowest algorithm-specific median (IQR) are highlighted in bold.}
-  \label{table:simulation_results_median}
-\end{table}
-
+% \begin{table}[h]
+%   \centering
+%   \begin{tabular}{|c c c c c|}
+%     \hline
+%     Case & Data & LPME & PME & PC/PS \\
+%     \hline
+%     1 & 0.146 (0.233) & {\bf 0.074 (0.122)} & 0.131 (0.258) & 0.118 (0.206) \\
+%     2 & 0.467 (0.665) & {\bf 0.248 (0.516)} & 0.516 (0.750) & 0.564 (0.419) \\
+%     3 & 0.291 (0.601) & {\bf 0.239 (0.564)} & 0.317 (0.640) & 0.264 (0.584) \\
+%     4 & 4.26 (21.2) & {\bf 3.29 (13.0)} & 4.23 (21.1) & 4.22 (21.2) \\
+%     5 & 0.895 (1.33) & {\bf 0.584 (1.08)} & 0.894 (1.36) & 0.821 (1.22) \\
+%     6 & 0.284 (1.06) & {\bf 0.273 (0.891)} & 0.316 (1.04) & 0.557 (0.387) \\
+%     7 & 0.145 (0.552) & 2.96 (4.65) & 6.92 (1.17) & {\bf 1.58 (0.514)} \\
+%     8 & 0.110 (0.325) & {\bf 0.074 (0.208)} & 0.115 (0.331) & 0.172 (0.234) \\
+%     \hline
+%   \end{tabular}
+%   \caption{MSD comparison to true values, Median (IQR). The lowest algorithm-specific median (IQR) are highlighted in bold.}
+%   \label{table:simulation_results_median}
+% \end{table}
+%
 \begin{figure}[h]
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth]{sim_case7}
-  \caption{Simulation Case 6}
-  \label{fig:sim_case6}
-\end{figure}
-
-The mean, median, standard deviation and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=\textwidth]{sim_case7}
+%   \caption{Simulation Case 6}
+%   \label{fig:sim_case6}
+% \end{figure}
+%
+The mean and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_mean}.
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
@@
 
-While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
-Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study.
+While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal (CN), 150 displayed symptoms of mild cognitive impairment (MCI), and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
+Of the 85 CN participants, 28 were subsequently observed with MCI or dementia during the course of the study.
 Follow-up duration for these participants ranged from approximately 6 months to 112 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study.
-To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants.
-Of these remaining participants, 80 were cognitively normal, 126 had mild cognitive impairment, and 44 were diagnosed with AD at their baseline visits.
+To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, resulting in a final sample of 250 participants (80 CN, 126 MCI, 44 AD).
 Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
@@
 There are two main sources of discrepancies between the data and the LPME estimate.
-First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
+First, at time points where the observed surface changes orientation, as is visible between the first two time points, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
 The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.
@@
 
-Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}.
-While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}.
+While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
 
-To understand how the subcortical structure estimates obtained by the LPME and PME algorithms may impact the corresponding estimates of the volumes of these structures, the volume of each structure was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
+To understand how the subcortical structure estimates obtained by the LPME and PME algorithms may impact the corresponding estimates of the volumes of these structures, the volume of each structure was approximated from the raw data and the PME and LPME estimates at each time point.
 The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
@@
 The volume estimates for randomly selected individuals can be found in Figures \ref{fig:lhipp_volume_comparison} and \ref{fig:lthal_volume_comparison}.
-Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
+The hippocampus estimates indicate that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
 In Figure \ref{fig:lhipp_volume_comparison}, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
-All three sources show substantial variability between study visits, particularly for subjects 137\_S\_0283 and 137\_S\_0686.
-In Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
-However, the LPME estimates demonstrate the ability to smooth over this variability, resulting in volume estimates that are not as sensitive to differences between time points.
+All three sources show substantial variability between study visits.
+Figure \ref{fig:lthal_volume_comparison} again shows that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
+However, the LPME estimates demonstrate the ability to smooth over this variability, yielding results that are not as sensitive to differences between time points.
 We also see that when basing volume estimates on the more reliable LPME fit to the thalamus data, LPME yields volume estimates similar in magnitude to those obtained from the raw data and PME estimates.
@@
 A summary of the between-visit variability in the volume estimates of the subcortical structures considered here is shown in Table \ref{table:adni_volume_sds}.
-In cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e.
+We estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e.
 raw data, LPME, and PME).
 We then took the mean of these values for each estimate source.
-When regression adjustments were used, instead of considering the raw volume estimates, we estimated the standard deviation of the residuals of a simple linear regression regressing volume on time.
-This additional step gives a better indication of smoothness when the volume of the structure cannot be assumed to be constant over time, as is the case for those with AD.
 The table shows that PME yields the smoothest volume estimates for both hippocampi.
@@
 However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.
+Analogous results were found when conducting this analysis for the residuals of simple linear regression models using time as the independent variable and volume as the outcome.
 
@@
 
-\begin{figure}%
-  \centering
-  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_data_plot}}}%
-  \hfill
-  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_isomap_plot}}}
-  \hfill
-  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_pme_plot}}}
-
-  \caption{Left Thalamus, Cognitive Healthy ADNI Participant}
-  \label{fig:adni_lthal_result}
-\end{figure}
-
+% \begin{figure}%
+%   \centering
+%   \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_data_plot}}}%
+%   \hfill
+%   \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_isomap_plot}}}
+%   \hfill
+%   \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_pme_plot}}}
+%
+%   \caption{Left Thalamus, Cognitive Healthy ADNI Participant}
+%   \label{fig:adni_lthal_result}
+% \end{figure}
+%
 % \begin{figure}[h]
@@
 
+% \begin{table}[ht]
+%   \centering
+%   \begin{tabular}{|c c c c c|}
+%     \hline
+%     Structure & Regression-Adjusted & Data & LPME & PME  \\
+%     \hline
+%     Left Hippocampus & No & 213 & 225 & \textbf{187} \\
+%     Right Hippocampus & No & 210 & 236 & \textbf{186} \\
+%     %Left Hippocampus & Yes & 173 & 190 & \textbf{150} \\
+%     %Right Hippocampus & Yes & 174 & 206 & \textbf{154} \\
+%     Left Thalamus & No & 78.3 & \textbf{49.0} & 67.8 \\
+%     Right Thalamus & No & 139 & \textbf{112} & 122 \\
+%     %Left Thalamus & Yes & 62.8 & \textbf{29.0} & 53.0 \\
+%     %Right Thalamus & Yes & 115 & \textbf{85.3} & 99.9 \\
+%     \hline
+%   \end{tabular}
+%   \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
+Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.
+}
+%   \label{table:adni_volume_sds}
+% \end{table}
+%
 \begin{table}[ht]
   \centering
-  \begin{tabular}{|c c c c c|}
+  \begin{tabular}{|c c c c|}
     \hline
-    Structure & Regression-Adjusted & Data & LPME & PME  \\
+    Structure & Data & LPME & PME  \\
     \hline
-    Left Hippocampus & No & 213 & 225 & \textbf{187} \\
-    Right Hippocampus & No & 210 & 236 & \textbf{186} \\
-    Left Hippocampus & Yes & 173 & 190 & \textbf{150} \\
-    Right Hippocampus & Yes & 174 & 206 & \textbf{154} \\
-    Left Thalamus & No & 78.3 & \textbf{49.0} & 67.8 \\
-    Right Thalamus & No & 139 & \textbf{112} & 122 \\
-    Left Thalamus & Yes & 62.8 & \textbf{29.0} & 53.0 \\
-    Right Thalamus & Yes & 115 & \textbf{85.3} & 99.9 \\
+    Left Hippocampus & 213 & 225 & \textbf{187} \\
+    Right Hippocampus & 210 & 236 & \textbf{186} \\
+    %Left Hippocampus & Yes & 173 & 190 & \textbf{150} \\
+    %Right Hippocampus & Yes & 174 & 206 & \textbf{154} \\
+    Left Thalamus & 78.3 & \textbf{49.0} & 67.8 \\
+    Right Thalamus & 139 & \textbf{112} & 122 \\
+    %Left Thalamus & Yes & 62.8 & \textbf{29.0} & 53.0 \\
+    %Right Thalamus & Yes & 115 & \textbf{85.3} & 99.9 \\
     \hline
@@
   \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
-Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.
-}
+When applied to the irregularly shaped hippocampus data, PME estimates display the lowest variability.
+However, LPME estimates of volume show the lowest variability when applied to the thalamus.}
   \label{table:adni_volume_sds}

### 2024-04-23 code/03_adni_mri_preprocess.qmd
convert preprocessing script to notebook

---
title: "Preprocessing ADNI Structural MRI Data for LPME"
author: "Robert Zielinski"
date: "April 23, 2024"
output: html_document
---

This notebook contains the code used to preprocess structural MRI images from the ADNI dataset and extract the hippocampuses and thalamuses from each image for further analysis with the LPME algorithm. To run properly, the code assumes that the ADNI data is contained within the `data/adni` directory in this project. To obtain this data, please follow the steps outlined in the `README.md` document, seen [here](https://github.com/rjzielinski/lpme-project).

The packages required for preprocessing are loaded below:

```{r, message=FALSE}
library(foreach)      # used for parallelization
library(fslr)         # includes functions for registration and segmentation
library(oro.dicom)    # dicom file operations
library(oro.nifti)    # nifti file operations
library(neurobase)    # added functions for nifti operations
library(stringr)      # string and regex operations
library(tidyverse)    # used for dataframe construction
```

The following code chunk obtains the directory names for each subject incldued in the ADNI data, loads the MNI space mask, and creates two empty lists. The first will contain the original scans converted into nifti format, while the second will store the scan outputs after the preprocessing steps have been applied.

```{r}
sub_dirs <- list.dirs("data/adni", recursive = FALSE)

mni_nifti <- readNIfTI(
  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm.nii.gz")
)

nii_scans <- list()

processed_scans <- list()
```

The preprocessing steps that follow are computationally intensive. To relieve some of this burden, the code below enables parallelization.

```{r}
ncores <- parallel::detectCores() / 2
cl <- parallel::makeCluster(ncores, type = "FORK")
doParallel::registerDoParallel(cl = cl)
```

The following code preprocesses every image in our dataset. The data is originally formatted as DICOM files, so we begin processing by converting these to NIFTI format. Details of FSL's FIRST algorithm are given in the User Guide provided [here](https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide). The `fslr` package's `run_first_all()` function calls FSL's `run_first_all` function, which in turn has two main steps:

1. Affine transformation to MNI space using `first_flirt`. This is designed to be run on whole-head t1-weighted images.
2. Segment structures by running `run_first`. The `run_first_all` function repeats this function call for all available structures.

Following the guidance of this documentation, we use the `run_first_all` function on the T1-weighted images that have been loaded from the original DICOM files and converted to NIFTI format. No other preprocessing steps are used.

```{r, eval=FALSE}
foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
  # for each subject...

  # retrieve the filenames of all images
  img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
    str_remove(pattern = "([^/]+$)") %>%  # remove all text after final slash
    unique()

  nii_scans[[dir_idx]] <- list() # list to store all scans for given subject

  for (img_idx in 1:length(img_dirs)) {
    # for each of the subject's images

    # save all processed data to data/adni_processed_fsl
    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/adni", replacement = "", img_dirs[img_idx]))

    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
      # avoid repeat processing if output file is already present
      next
    } else {
      # read all dicom images in directory
      all_slices <- readDICOM(img_dirs[img_idx])

      # convert to nifti format
      nii <- dicom2nifti(all_slices)

      # run segmentation function
      run_first_all(
        nii,
        oprefix = proc_dir,
        verbose = FALSE,
        opts = "-d"
      )
    }
  }
}

parallel::stopCluster(cl = cl) # stop parallelized cluster
```

The segmentation process above saves aggregated volumetric estimates in the `_all_fast_origsegs.nii.gz` files within each output directory. Note that these results have been backtransformed to the original image space, while the results given in `_all_fast_firstsegs.nii.gz` are still registered to the MNI space. As this analysis is primarily concerned with structural differences between images, we prefer to operate in the original image space.

Next, we must extract the results specific to the right and left hippocampuses and the left and right thalamuses.

```{r}
patnos <- list.dirs("data/adni_processed_fsl", recursive = FALSE, full.names = FALSE)
processed_dirs <- list.dirs("data/adni_processed_fsl", recursive = FALSE)

# create matrices to hold voxel information
lhipp <- matrix(ncol = 6)
rhipp <- matrix(ncol = 6)

lthal <- matrix(ncol = 6)
rthal <- matrix(ncol = 6)

# vectors to hold FIRST-estimated volume information
lhipp_vol <- vector()
rhipp_vol <- vector()

lthal_vol <- vector()
rthal_vol <- vector()

# image collection information
img_id <- vector()
img_date <- vector()

# matrices for surface voxel information
lhipp_surface <- matrix(ncol = 6)
rhipp_surface <- matrix(ncol = 6)

lthal_surface <- matrix(ncol = 6)
rthal_surface <- matrix(ncol = 6)
```

The following code contains all information necessary to extract voxel-level location estimates for the relevant structures in all images. The indices for the structures of interest are based on the structure order in the "Labels" section at the end of the [user guide](https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide).

The essential operation being performed is the conversion of structure information from a 170 x 256 x 256 tensor of image voxel intensities to a $N \times 6$ matrix with two columns of image metadata, three columns of positional values and one column representing voxel intensity. In this case, $N$ represents the number of voxels with nonzero intensity. This allows us to represent a masked T1-weighted image as a point cloud with $N$ observations.

This process outputs a similar point cloud matrix that only includes voxels found to be on the surface of the structure. To approximate the structure surface while considering the irregularity of the structures being considered, any voxel having an adjacent voxel with zero intensity was included in the surface estimate. Despite the potential for error in estimating the surface, in practice, this procedure provides surface estimates that appear to reasonably reflect the expected shapes upon visual inspection.

```{r, eval=FALSE}
for (dir_idx in 1:length(processed_dirs)) {
  # for each subject

  # identify image recording dates
  scan_dates <- list.dirs(
    paste0(processed_dirs[dir_idx], "/MPRAGE"),
    full.names = FALSE,
    recursive = FALSE
  )
  scan_dirs <- list.files(
    paste0(processed_dirs[dir_idx], "/MPRAGE"),
    recursive = TRUE,
    full.names = TRUE
  ) %>%
    gsub(pattern = "([^/]+$)", replacement = "") %>%
    unique()

  for (scan_idx in 1:length(scan_dirs)) {
    # for each image

    # if results file exists
    if (file.exists(paste0(scan_dirs[scan_idx], "_all_fast_origsegs.nii.gz"))) {
      img_id <- c(img_id, patnos[dir_idx])
      img_date <- c(img_date, scan_dates[scan_idx])

      # read segmentation estimates
      temp_segs <- readnii(paste0(scan_dirs[scan_idx], "_all_fast_origsegs.nii.gz"))

      # estimate volumes for all structures
      seg_vols <- fslstats(
        paste0(scan_dirs[scan_idx], "_all_fast_origsegs.nii.gz"),
        opts = "-V",
        ts = TRUE
      )

      # left hippocampus corresponds to index 6
      temp_lhipp <- temp_segs[, , , 6]
      # volume estimates contain two values: (1) number of voxels; (2) volume
      # retrieve second value
      temp_lhipp_vol <- seg_vols[6] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()
      lhipp_vol <- c(lhipp_vol, temp_lhipp_vol)

      # find left hippocampus voxels with intensity > 0
      # arr.ind saves index values in all three dimensions
      # this gives us positional locations of active voxels
      lhipp_idx <- which(temp_lhipp > 0, arr.ind = TRUE)

      # logical vector indicating whether voxel is on the structure's surface
      surface <- rep(FALSE, nrow(lhipp_idx))

      ### Surface Identification

      # loop through all voxels
      for (dim in 1:3) {
        for (dim2 in 1:3) {
          if (dim != dim2) {
            # ensure dim, dim2, dim3 are all different to select different values
            dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
            # for each of the first two specified dimensions,
            # get all available values in third dimension
            unique_vals <- unique(lhipp_idx[, c(dim, dim2)])
            for (i in 1:nrow(unique_vals)) {
              # get all third dimension values given first two dimensions
              vals <- lhipp_idx[lhipp_idx[, dim] == unique_vals[i, 1] & lhipp_idx[, dim2] == unique_vals[i, 2], dim3]
              min_vals <- vector()
              max_vals <- vector()

              for (v in vals) {
                # for each third dimension value
                # if position immediately before or after given value
                # does not have positive intensity, then selected value must
                # be on structure surface
                if (!((v-1) %in% vals)) {
                  min_vals <- c(min_vals, v)
                } else if (!((v+1) %in% vals)) {
                  max_vals <- c(max_vals, v)
                }
              }
              surface_vals <- c(min_vals, max_vals)
              for (v in surface_vals) {
                surface[which(lhipp_idx[, dim] == unique_vals[i, 1] & lhipp_idx[, dim2] == unique_vals[i, 2] & lhipp_idx[, dim3] == v, arr.ind = TRUE)] <- TRUE
              }
            }
          } else {
            next
          }
        }
      }
      lhipp_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        # ensure proper structure of positional indices
        lhipp_idx %*% diag(pixdim(as.nifti(temp_lhipp))[2:4]),
        # append intensity values
        temp_lhipp[lhipp_idx]
      )
      # repeat process, but only for voxels where surface == TRUE
      lhipp_surface_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        lhipp_idx[surface, ] %*% diag(pixdim(as.nifti(temp_lhipp))[2:4]),
        temp_lhipp[lhipp_idx[surface, ]]
      )
      lhipp <- rbind(lhipp, lhipp_temp)
      lhipp_surface <- rbind(lhipp_surface, lhipp_surface_temp)

      # repeat for right hippocampus, index 13
      temp_rhipp <- temp_segs[, , , 13]

      temp_rhipp_vol <- seg_vols[13] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()
      rhipp_vol <- c(rhipp_vol, temp_rhipp_vol)
      rhipp_idx <- which(temp_rhipp > 0, arr.ind = TRUE)
      surface <- rep(FALSE, nrow(rhipp_idx))
      for (dim in 1:3) {
        for (dim2 in 1:3) {
          if (dim != dim2) {
            dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
            unique_vals <- unique(rhipp_idx[, c(dim, dim2)])
            for (i in 1:nrow(unique_vals)) {
              vals <- rhipp_idx[rhipp_idx[, dim] == unique_vals[i, 1] & rhipp_idx[, dim2] == unique_vals[i, 2], dim3]
              min_vals <- vector()
              max_vals <- vector()
              for (v in vals) {
                if (!((v-1) %in% vals)) {
                  min_vals <- c(min_vals, v)
                } else if (!((v+1) %in% vals)) {
                  max_vals <- c(max_vals, v)
                }
              }
              surface_vals <- c(min_vals, max_vals)
              for (v in surface_vals) {
                surface[which(rhipp_idx[, dim] == unique_vals[i, 1] & rhipp_idx[, dim2] == unique_vals[i, 2] & rhipp_idx[, dim3] == v, arr.ind = TRUE)] <- TRUE
              }
            }
          } else {
            next
          }
        }
      }
      rhipp_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        rhipp_idx %*% diag(pixdim(as.nifti(temp_rhipp))[2:4]),
        temp_rhipp[rhipp_idx]
      )
      rhipp_surface_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        rhipp_idx[surface, ] %*% diag(pixdim(as.nifti(temp_rhipp))[2:4]),
        temp_rhipp[rhipp_idx[surface, ]]
      )
      rhipp <- rbind(rhipp, rhipp_temp)
      rhipp_surface <- rbind(rhipp_surface, rhipp_surface_temp)

      # repeat for left thalamus, index 1
      temp_lthal <- temp_segs[, , , 1]
      temp_lthal_vol <- seg_vols[1] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()
      lthal_vol <- c(lthal_vol, temp_lthal_vol)
      lthal_idx <- which(temp_lthal > 0, arr.ind = TRUE)
      surface <- rep(FALSE, nrow(lthal_idx))
      for (dim in 1:3) {
        for (dim2 in 1:3) {
          if (dim != dim2) {
            dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
            unique_vals <- unique(lthal_idx[, c(dim, dim2)])
            for (i in 1:nrow(unique_vals)) {
              vals <- lthal_idx[lthal_idx[, dim] == unique_vals[i, 1] & lthal_idx[, dim2] == unique_vals[i, 2], dim3]
              min_vals <- vector()
              max_vals <- vector()
              for (v in vals) {
                if (!((v-1) %in% vals)) {
                  min_vals <- c(min_vals, v)
                } else if (!((v+1) %in% vals)) {
                  max_vals <- c(max_vals, v)
                }
              }
              surface_vals <- c(min_vals, max_vals)
              for (v in surface_vals) {
                surface[which(lthal_idx[, dim] == unique_vals[i, 1] & lthal_idx[, dim2] == unique_vals[i, 2] & lthal_idx[, dim3] == v, arr.ind = TRUE)] <- TRUE
              }
            }
          } else {
            next
          }
        }
      }
      lthal_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        lthal_idx %*% diag(pixdim(as.nifti(temp_lthal))[2:4]),
        temp_lthal[lthal_idx]
      )
      lthal_surface_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        lthal_idx[surface, ] %*% diag(pixdim(as.nifti(temp_lthal))[2:4]),
        temp_lthal[lthal_idx[surface, ]]
      )
      lthal <- rbind(lthal, lthal_temp)
      lthal_surface <- rbind(lthal_surface, lthal_surface_temp)

      # repeat for right thalamus, index 9
      temp_rthal <- temp_segs[, , , 9]
      temp_rthal_vol <- seg_vols[9] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()
      rthal_vol <- c(rthal_vol, temp_rthal_vol)
      rthal_idx <- which(temp_rthal > 0, arr.ind = TRUE)
      surface <- rep(FALSE, nrow(rthal_idx))
      for (dim in 1:3) {
        for (dim2 in 1:3) {
          if (dim != dim2) {
            dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
            unique_vals <- unique(rthal_idx[, c(dim, dim2)])
            for (i in 1:nrow(unique_vals)) {
              vals <- rthal_idx[rthal_idx[, dim] == unique_vals[i, 1] & rthal_idx[, dim2] == unique_vals[i, 2], dim3]
              min_vals <- vector()
              max_vals <- vector()
              for (v in vals) {
                if (!((v-1) %in% vals)) {
                  min_vals <- c(min_vals, v)
                } else if (!((v+1) %in% vals)) {
                  max_vals <- c(max_vals, v)
                }
              }
              surface_vals <- c(min_vals, max_vals)
              for (v in surface_vals) {
                surface[which(rthal_idx[, dim] == unique_vals[i, 1] & rthal_idx[, dim2] == unique_vals[i, 2] & rthal_idx[, dim3] == v, arr.ind = TRUE)] <- TRUE
              }
            }
          } else {
            next
          }
        }
      }
      rthal_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        rthal_idx %*% diag(pixdim(as.nifti(temp_rthal))[2:4]),
        temp_rthal[rthal_idx]
      )
      rthal_surface_temp <- cbind(
        patnos[dir_idx],
        scan_dates[scan_idx],
        rthal_idx[surface, ] %*% diag(pixdim(as.nifti(temp_rthal))[2:4]),
        temp_rthal[rthal_idx[surface, ]]
      )
      rthal <- rbind(rthal, rthal_temp)
      rthal_surface <- rbind(rthal_surface, rthal_surface_temp)
    }
  }
}
```

The following code takes the matrix-formatted results from the previous code chunk and reformats them as `tibble` objects, providing column names and ensuring use of the appropriate data classes.

```{r, eval=FALSE}
lhipp <- lhipp[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

lhipp_surface <- lhipp_surface[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

rhipp <- rhipp[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

rhipp_surface <- rhipp_surface[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

lthal <- lthal[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

lthal_surface <- lthal_surface[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

rthal <- rthal[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )

rthal_surface <- rthal_surface[-1, ] %>%
  as_tibble() %>%
  rename(
    patno = V1,
    scan_date = V2,
    x = V3,
    y = V4,
    z = V5,
    intensity = V6
  ) %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    z = as.numeric(z),
    intensity = as.numeric(intensity)
  )
```

Finally, the `hipp_info` and `thal_info` data frames are created to store volume information for the hippocampuses and thalamuses, respectively. All results are then saved as CSV files to be used in the analysis step.

```{r, eval=FALSE}
hipp_info <- data.frame(img_id, img_date, lhipp_vol, rhipp_vol)
names(hipp_info) <- c("patno", "date", "lhipp_vol", "rhipp_vol")

thal_info <- data.frame(img_id, img_date, lthal_vol, rthal_vol)
names(thal_info) <- c("patno", "date", "lthal_vol", "rthal_vol")

write_csv(hipp_info, "data/adni_fsl_hipp_info.csv")
write_csv(thal_info, "data/adni_fsl_thal_info.csv")

write_csv(lhipp, "data/adni_fsl_lhipp.csv")
write_csv(rhipp, "data/adni_fsl_rhipp.csv")

write_csv(lthal, "data/adni_fsl_lthal.csv")
write_csv(rthal, "data/adni_fsl_rthal.csv")

write_csv(lhipp_surface, "data/adni_fsl_lhipp_surface.csv")
write_csv(rhipp_surface, "data/adni_fsl_rhipp_surface.csv")

write_csv(lthal_surface, "data/adni_fsl_lthal_surface.csv")
write_csv(rthal_surface, "data/adni_fsl_rthal_surface.csv")
```

### 2024-05-09 paper/main.tex
fix typo in function definition

@@
 The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point.
-This smoothing mitigates the effects of random disturbances to the manifold between time points.
-A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds.
+This approach mitigates the effects of random disturbances to the manifold between time points.
+We propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds.
 We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
@@
 These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}).
-For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.
-
-Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
+For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes.Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
 Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
@@
 Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
-In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
-After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
+In a study on comparison of the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
 If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
-\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained.
+\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain image of a cognitively normal person which compose the majority of the data used for training the segmentation models.
 In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
@@
 
-Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods.
+Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods.
 Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data.
@@
 Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
-Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}).
+Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}).
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
 For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.
+Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.
 
 Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
-These methods consider time as an additional dimension in the data, i.e.
-they include time in the dimension reduction process.
-This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
+These methods consider time as an additional dimension in the data, by including time in the dimension reduction process.
+Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
 Here, we propose an alternative approach to estimating manifolds over time.
@@
 
-Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}.
-\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously.
+Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
+\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
 \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
@@
 
+Our work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time.
+One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this manuscript.
+The data are publicly available and are hosted at \texttt{adni.loni.usc.edu}.
+The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.
+
 The remainder of this article is laid out as follows.
 In Section \ref{s:PME}, we review the principal manifold framework.
-In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework.
+In Section \ref{s:LPME}, we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show our proposed novel  algorithm for dimension reduction using the longitudinal principal manifold framework.
 Section \ref{s:simulations} demonstrates the performance of this approach on simulated data.
-In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus and thalamus in individuals with Alzheimer's disease.
+In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in individuals with AD and cognitively normal adults.
 The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.
@@
 
-The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
-However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
+The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
+However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
 In addition, applying the methods in high-dimensional settings is problematic.
@@
 The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
-Most manifold estimation methods are based on defining a projection index, denoted $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
-The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
+Most manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
+The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for two- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
 The principal manifolds are then defined as follows.
@@
 The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements.
-First is the expected distance from each point in the cloud to the fitted manifold.
-In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on the 3-dimensional brain region of interest.
+First is the expected distance from each point in the cloud to the fitted manifold, e.g.
+the distance between the fitted surface and the coordinates of the vertices on the 3-dimensional brain region of interest which we intend to minimize.
 The second element of the equation imposes the smoothness penalty on the estimated manifold.
-The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before.
+The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting.
 To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$.
-In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
+In this setting,  $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
 It is important to note that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.
@@
 Now consider that the data are observed longitudinally during several visits by a study participant.
-Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit.
+Hence, the observed data are the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of $j$th study visit.
 Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.
@@
   \includegraphics[scale=0.4]{sim_case1_plot2}
-  \caption{Longitudinal Simulated Data with $D = 2$.
-Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points.
+  \caption{Longitudinal simulated data with $D = 2$.
+Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are generated at 5 time points.
 Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
@@
 
-Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
+Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
 It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e.
@@
 In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
-In order to address the limitations of the existing approaches for the setting described above, we now propose longitudinal principal manifold estimation (LPME) algorithm to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.
+In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.
 
-To give a more general representation of this problem, let $\left\{\bm{x}_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
+To give a more general representation of this problem, let $\left\{\bm{x}_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
 As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
@@
 \begin{definition}
-  \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
+  \label{def:lpme} Given $\boldsymbol{X} = \{ \boldsymbol{X}_t \}_{t=1}^T$, denoting the collection of data clouds observed at all time points, where $\boldsymbol{X}_t$ is the random  $D$-vector observed at time $t$, the manifold $M_F^{d+1}$, which is representative of $\boldsymbol{X}$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
 \begin{align}\label{eq:newKappa}
   \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
-  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt. \nonumber
+  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt, \nonumber
 \end{align}
@@
 Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.
-Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered to be smooth over the other set of variables.
+Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered smooth over the other set of variables.
 
@@
 
-Given the framework proposed in the previous section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}).
+Given the framework proposed in the previous section, estimation of the longitudinal principal manifold entails minimizing the functional in (\ref{eq:newKappa}).
 Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).
@@
 
-In most practical manifold learning tasks data are often quite large, so we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud.
+In most practical manifold learning tasks data are often quite large, so we begin by implementing an approach for selecting a small number of points in $D$-dimensional space that represent the structure of the data cloud.
 This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest.
-Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large.
+Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is large.
 To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data.
+Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that  characterize the shape of the original data.
 The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
@@
   \caption{Simulated data with $D = 2$.
-The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
+The curves shown in this plot represent embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
 The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates.
@@
 At each time point, the estimated principal manifold has a spline form.
-As detailed by \cite{greenSilverman1994}, smoothing splines can be considered to have knots at the points used to fit them.
-This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical.
-Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point.
+As detailed by \cite{greenSilverman1994}, smoothing splines are fitted at a collection of knots.
+This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the knot values used as inputs when the functions are fitted must be identical.
+Hence, to obtain coefficients that can be compared across time points, a grid of knots $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the knots in each of the $d$ dimensions is selected and used as input for the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point.
 This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
@@
 \begin{equation}\label{eq:splineL}
-    \mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t
+    \mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t,
 \end{equation}
@@
 \end{equation}
-The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
+The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields the function
 \begin{equation}\nonumber
@@
 
-With a given tuning parameter value $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$ resulting in a function taking the form
+With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$ resulting in a function taking the form
 \begin{equation}\nonumber
@@
 \end{equation}
-where by defining the $T \times T$ matrix $\boldsymbol{A}$ as defined  $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\boldsymbol{T}$ as defined $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to
+where by defining the $T \times T$ matrix $\boldsymbol{A}$ as  $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\boldsymbol{T}$ as  $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$, we obtain the coefficients $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ as the solutions to
 \begin{equation}
@@
 
-With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$.
-Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\boldsymbol{r}$ may be written as
+For a given $\gamma$, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$.
+Hence, given $\gamma$, the estimated embedding function at time $t$ and $d$-dimensional parameterization $\boldsymbol{r}$ is
 \begin{equation}
@@
 \end{equation}
-The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.
+The optimal tuning value $\gamma^*$ is chosen as the value of $\gamma$ that minimizes $MSD(\gamma)$.
 Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
@@
 
-Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds.
-Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}.
-To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$.
+Notably, most manifold learning methods are developed under the assumption that the underlying manifold is not self-intersecting.
+This is not the case for in our motivating problem, where we are interested in modeling the surface of a brain region that, by definition, is a closed 3D shape.
+This may often be the case in other applications where manifold learning is implemented.
+One such simple example is modeling digits in 2D space where the number 8 for instance is a self-intersecting manifold.
+Methods based on the principal curve framework, including the PME framework, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}.
+A common approach for dealing with this problem is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space.
+To avoid this issue, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
 As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
-However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections.
+However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be viewed as given in the space with $d = 1$ with the data cloud of $D = 3$, while avoiding any self-intersections.
 This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.
@@
 
-To assess the performance of the LPME algorithm, we compare its performance with that of two alternative approaches on simulated datasets.
-We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
+To assess the performance of the LPME algorithm, we use simulation studies to compare how closely (in terms of reduction of MSE) the LPME estimated underlying functions approximate the true data generating functions in comparison with that of three alternative approaches.
+We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, as well as $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
 For each manifold, we consider differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time.
 The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
+In order to put these results in context of previous published work in this area, we use true underlying functions that are similar to those used in previous literature and build on these existing results.
+Based on our observations of systematic noise sources in imaging data, we add various types of terms to these functions at each visit.
 In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural changes in the underlying manifold over time.
 The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
-The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials.
-Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials.
+The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between visits.
+Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean 1 and variance that is varied between visits.
 These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.
@@
   \end{tabular}
-  \caption{Embedding functions used for simulation trials.}
+  \caption{Embedding functions used for simulation studies.}
   \label{table:simulation_embeddings}
@@
 A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal.
-The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times.
-All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
+The sample size for each time point is set to 1,000, so each embedding map is simulated a total of 7,776 times.
+All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
 
-Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}.
+Visualizations of the results from some example simulated cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
   \includegraphics[width=\textwidth]{sim_case5}
-  \caption{Simulation Cases 1 and 5.}
+  \caption{Scatterplots of two example simulated data clouds (gray points) with $d=1$ overlaid true embedding manifold (red), the estimated LMPE function (blue, dashed), PME (green, dotted), and principal curve or surface (purple, dot-dashed).}
   \label{fig:sim_case1}
@@
   \includegraphics[width=\textwidth]{sim_case7}
-  \caption{Simulation Case 6}
+  \caption{Scatterplots of two example simulated data clouds (gray points) with $d=2$ overlaid true embedding manifold (red), the estimated LMPE function (blue, dashed), PME (green, dotted), and principal curve or surface (purple, dot-dashed).}
   \label{fig:sim_case6}
@@
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
-Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach should result in lower mean squared distance values.
+Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach results in lower mean squared distance values.
 
@@
 As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates, shown in blue, remain relatively stable over time.
-This contrasts to the estimates found by the PME and principal curve approaches, which are highly sensitive to the data observed at each given time point, as expected.
+This contrasts to the estimates found by the PME and principal curve approaches, which are highly sensitive to the systematic added noise in the data observed at each given time point, as expected.
 This ultimately results in the LPME estimates remaining closer to the true underlying manifold, shown in red.
@@
 
-As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images.
+As discussed previously, meaningful between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images.
 This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates.
-To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).
+To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of AD.
 
-While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
+While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
 Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study.
-Follow-up duration for these participants ranged from approximately 6 months to 112 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study.
-To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants.
+Follow-up duration for these participants ranged from approximately 6 months to 12 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study.
+To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow-up were considered.
+Our final analysis set included 250 participants.
 Of these remaining participants, 80 were cognitively normal, 126 had mild cognitive impairment, and 44 were diagnosed with AD at their baseline visits.
-Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.
+Due to the changes experienced by those with AD, we focused on two brain regions of interest starting from modeling the hippocampi of the selected individuals then applying the proposed methods to model the thalamuses.
 
@@
 
-%\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}
-
 Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}.
@@
 First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
-The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.
+The second difference between the LPME estimates and the observed data is seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of curves in the observed data.
 
@@
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
-This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots.
+This Figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each plot.
 Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself.
@@
 Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}.
-While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+While the segmentation estimates of the thalamus do not show deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
 
@@
 We then took the mean of these values for each estimate source.
-When regression adjustments were used, instead of considering the raw volume estimates, we estimated the standard deviation of the residuals of a simple linear regression regressing volume on time.
+When regression adjustments were used, instead of considering the raw volume estimates, we estimated the standard deviation of the residuals of a simple linear regression of volume on time.
 This additional step gives a better indication of smoothness when the volume of the structure cannot be assumed to be constant over time, as is the case for those with AD.
-The table shows that PME yields the smoothest volume estimates for both hippocampi.
+Table \ref{table:adni_volume_sds} shows that PME yields the smoothest volume estimates for both hippocampi.
 This conclusion aligns with the visual evidence of inconsistency in the LPME estimates caused by the irregular shape of the hippocampus.
@@
   \caption{Left Hippocampus Cross Section.
-We can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the surface in the observed data.
-In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it.
-Gaps in the surface estimates frequently correspond with inaccurate volume estimates.}
+LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the surface in the observed data.
+In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap.
+Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.}
   \label{fig:lhipp_cross_sections}
@@
 
-In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data.
+In this article, we propose the LPME framework for modeling longitudinal changes in a low-dimensional manifold underlying high-dimensional data.
 To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a general-purpose longitudinal setting.
 We also suggest a data augmentation approach that circumvents the inability to fit principal curve-based methods to self-intersecting manifolds in select settings.
-In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point.
-This improvement was seen across several manifold types.
+In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance, in terms of minimizing the MSD, in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve/surface algorithms at each time point.
+This improvement is observed across several manifold types.
 
-It is key to note here that the MSD values shown in these tables compare the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on.
-This means that a model that performs well using this metric may not perform well when comparing the model's projections onto the manifold with the data.
-In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point.
-This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold.
-When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.
+Similar to other applications of statistical smoothing techniques, it is difficult to compare performance of models in real data experiments.
+Specifically, the MSD provides a summary of the distance of the data points and their projections onto the estimated manifolds for each model.
+This means that a model that performs well using this metric may not actually be the best fit for the underlying manifold, e.g.
+in the case of overfitting.
+This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are estimating a curve or surface using the data available at a given time point, which may be affected by systematic noise that increases the distance to the true manifold.
+When comparing the fit of the PME and principal surface and principal curve methods in the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.
 
-While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations.
-First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly.
-Because the output of PME is dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
-The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern.
-Opportunities also exist to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.
-
-A second limitation is that the LPME method requires relative stability in the overall shape of the structure being fit over time.
-Thus, while this algorithm can be useful for separating meaningful changes in a manifold over time from inter-observation error, it will be unable to fit in situations where the error between observations is too substantial.
+While the LPME algorithm demonstrated strong performance in a number of settings, it is important to note the assumptions under which the method performs well.
+First, because Isomap and PME are used for initialization and time-specific fitting, it is necessary that the assumptions required by these algorithms hold for the LPME algorithm to perform well.
+In practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
+In such cases, we recommend the use of alternative methods, such as Locally Linear Embedding (CITATION) or Laplacian Eigenmaps (CITATION), to generate the initial parameterization to alleviate this concern.
+Opportunities also exist to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise.
+A second assumption is that the LPME method requires relative stability in the overall shape of the underlying manifold structure.
+Thus, while this algorithm is useful for separating meaningful changes in a manifold over time from inter-observation error, it may not perform well in situations where the error between observations is too substantial, similar to any smoothing method.
 How restrictive this requirement is depends largely on the area of application.
@@
 
-Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset.
+An interesting future research direction is extension of the LPME method for modeling  shapes such as the hippocampus surfaces obtained from the ADNI dataset.
 It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus.
 There are two potential sources of this difficulty, each suggesting different steps that may be taken to address this issue.
-First, it is possible that this problem is caused by an inability of the PME and LPME algorithms to handle structures with as many sharp curves as the hippocampus.
-This may be addressed in several ways.
-As discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated.
+The manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better estimate the sharp curvature of the hippocampus.
 Also, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
-Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.
 
-The second potential source of the algorithms' poor fit to the hippocampus data could be an insufficiency in the data augmentation approach that allows us to fit some self-intersecting manifolds.
-The simple calculations used here, augmenting with polar or spherical coordiantes, allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
+The second potential interesting direction of research in the future is more in-depth investigation of the data augmentation approach  described in Section \ref{s:LPME}, that allows us to fit some self-intersecting manifolds.
+The proposed augmentation with polar or spherical coordinates, allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
 However, while this approach addressed a problem posed by well-behaved circular or spherical structures, such as the thalamus, its success may not extend to settings with more complex data.
-In the case of the hippocampus, we speculate that there may be intersections present in the spherical coordinates, which then result in the overlaps in certain regions of the estimated manifolds.
-Similar types of issues would arise in settings with other complex structures, like a figure 8 shape.
+In the case of the hippocampus, potential intersections in the spherical coordinates may result in the overlaps in certain regions of the estimated manifolds.
+Similar types of issues may arise in settings with other complex structures, like a figure 8 shape.
 The development of a procedure to handle these considerations would expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.
 
-Further attention should be paid to the process used to estimate volumes from the PME and LPME outputs if these volume estimates are to be used as outcomes of interest.
-The current procedure includes ad hoc adjustments to maintain the stability of estimates in the presence of gaps in the estimated surface.
-However, alternate approaches may yield less variable results.
-For example, \cite{mengPrincipalManifoldEstimation2021} introduced an interior identification approach that effectively makes use of the explicit representation of the embedding map obtained through the PME and LPME fitting process, and which may be adapted to settings with gaps in the estimated manifold.
+Estimation of volumes of the resulting manifolds is essential, as the volumes are often used as outcomes in modeling disease processes over time in clinical trials and observational studies.
+In this manuscript, the volumes were obtained by simply computing the number of within region voxels resulting from each fitting procedure.
+If the resulting manifold has gaps, however, this procedure may not be possible to implement.
+\cite{mengPrincipalManifoldEstimation2021} introduced an interior identification approach that effectively makes use of the explicit representation of the embedding map obtained through the PME and LPME fitting processes.
+Future work may adapt this procedure to settings with gaps in the estimated manifold.
 
-Finally, while the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
+While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, extension of this method to manifold estimation for groups of similar structures will be a valuable contribution to the literature allowing population-level comparisons of time-dependent trajectories of shapes.
 Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.
+In population-level modeling of the data, it will be important to also consider registration of the participant-level MRI data to a common template space, raising important considerations for alignment of the shapes without loosing important information.
 
-%\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction.
-The smoothness can be represented by the covariance kernel we use in the GPR.
-For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels.
-The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
+Finally, Gaussian process regression (GPR) may be applied to smooth the time direction.
+The smoothness can be represented by the covariance kernel used in the GPR, e.g.
+by applying the Gaussian or Matérn kernels (\cite{dunsonInferringManifoldsNoisy2022}).
+Future work may consider implementation of a similar approach in the framework we proposed in this article.
+
 \newpage

### 2024-05-09 recode_attempt1/Principal_Manifold_Estimation_Recode.R
remove redundant code

@@ -1,423 +1,0 @@
-library(MASS)
-library(Matrix)
-library(Rfast)
-library(vegan)
-library(plot3D)
-
-
-PME <- function(x.obs, d, N0=20*D, tuning.para.seq=exp((-15:5)), alpha=0.05, max.comp=100, epsilon=0.05, max.iter=100, print.MSDs=TRUE){
-  
-  source("norm_euclidean.R")
-  source("dist_euclidean.R")
-  source("hdmde.R")
-  source("projection.R")
-  source("eta_kernel.R")
-  source("ker.R")
-  source("weight_seq.R")
-
-  
-  # "x.obs" is the data set of interest. 
-  #         There are n observations, and each observation is a D-dimensional point.
-  #         x.obs is a n-by-D matrix.
-  # "d" is the intrinsic dimension of the underlying manifold
-  # "N0" is a predetermined lower bound for N - the number of density components, default value is 20*D
-  # "tuning.para.seq" is a vector of tuning parameter candidates, its default value is exp((-15:5)).
-  #                   If you would like to fit a manifold for a specific lambda, set tuning.prar.seq=c(lambda).
-  # "alpha" is the pre-determined confidence level, which determines the number of the components in a mixture density.
-  # "max.comp" is the upper bound of the number of components in the desired mixture density.
-  # "epsilon" is the tolerance for distance between manifolds f.old and f.new.
-  # "max.iter" is the upper bound of the number of steps in this iteration.
-  
-  # Remark: The larger N0 is, the less time consuming the function is.
-  
-  dimension.size <- dim(x.obs)
-  D <- dimension.size[2] # "D" is the dimension of the input space.
-  n <- dimension.size[1] # "n" is the number of observed D-dimensional data points.
-  lambda <- 4-d # "lambda" determines the form of reproducing kernels
-  
-  if (N0 == 0) { 
-    N0 <- 20 * D 
-  }
-  
-  est <- hdmde(x.obs, N0, alpha, max.comp)
-  theta.hat <- est$theta.hat  
-  centers <- est$mu
-  sigma <- est$sigma
-  W <- diag(theta.hat) # The matrix W
-  X <- est$mu
-  I <- length(theta.hat)
-  
-  # The (i,j)th element of this matrix is the Euclidean 
-  dissimilarity.matrix <- as.matrix(dist(X)) 
[…]

### 2024-05-09 recode_attempt1/dist_euclidean.R

@@ -1,4 +1,0 @@
-dist.euclidean <- function(x, y) {
-  # Distance function in a Euclidean space of any dimension
-  return(norm.euclidean(x - y))
-}

### 2024-05-09 recode_attempt1/eta_kernel.R

@@ -1,13 +1,0 @@
-eta.kernel <- function(t, lambda) {
-  norm_val <- norm.euclidean(t)
-  if (lambda %% 2 == 0) {
-    if (norm_val == 0) {
-      y <- 0
-    } else {
-      y <- (norm_val ^ lambda) * log(norm_val)
-    }
-  } else {
-    y <- norm_val ^ lambda
-  }
-  return(y)
-}

### 2024-05-09 recode_attempt1/hdmde.R

@@ -1,94 +1,0 @@
-hdmde2 <- function(x.obs, N0, alpha, max.comp, p.old = NULL) {
-  
-  zalpha <- qnorm(1 - alpha / 2)
-  n.D <- dim(x.obs)
-  n <- n.D[1]
-  D <- n.D[2]
-  N <- N0
-  
-  print(N)
-  
-  km <- kmeans(
-    x.obs,
-    N,
-    nstart = 100
-  )  
-  mu <- km$centers
-  
-  # calc_sigma <- function(km, j) {
-  #   index.temp <- which(km$cluster == j)
-  #   xi.j <- x.obs[index.temp, ]
-  #   sig.prepare <- function(x) {
-  #     return((dist.euclidean(x, mu[j, ]))^2)
-  #   }
-  #   s <- apply(xi.j, 1, sig.prepare)
-  #   return(mean(s))
-  # }
-  
-  # sigma.vec <- sapply(1:N, calc_sigma, km = km)
-  
-  sigma.vec <- rep(NA, N)
-  # for (idx in 1:length(sort(unique(km$cluster)))) {
-  #   sigma.vec[idx] <- calc_sigma(km, sort(unique(km$cluster))[idx])
-  # }
-  for (j in 1:N) {
-    index.temp <- which(km$cluster == j)
-    xi.j <- x.obs[index.temp, ]
-    sig.prepare <- function(x) {
-      return((dist.euclidean(x, mu[j, ]))^2)
-    }
-    s <- apply(xi.j, 1, sig.prepare)
-    sigma.vec[j] <- mean(s)
-  }
-  
-  sig <- sqrt(mean(sigma.vec) / dim(x.obs)[2])
-  
-  theta.hat <- weight.seq(x.obs, mu, sig)
-  
-  f.test <- function(x) {
-    comp.vec <- ker(x, mu, sig)
-    return(sum(theta.hat * comp.vec))
-  }
-  
-  p.old <- apply(x.obs, 1, f.test)
-  
-  if (is.null(p.old)) {
-    p.new <- apply(x.obs, 1, f.test)
-    test.rejection <- 1
-  } else {
-    p.new <- apply(x.obs, 1, f.test)
-    delta.hat <- p.new - p.old
-    sigma.hat.sq <- mean((delta.hat - mean(delta.hat))^2)
[…]

### 2024-05-09 recode_attempt1/ker.R

@@ -1,7 +1,0 @@
-ker <- function(x, mu, sigma) {
-  # Smoothing kernel for density estimation
-  # We applied Gaussian kernel
-  require(Rfast)
-  yseq <- dnorm((x - t(mu)) / sigma)
-  return((sigma ^ {-length(x)}) * colprods(yseq))
-}

### 2024-05-09 recode_attempt1/norm_euclidean.R

@@ -1,5 +1,0 @@
-norm.euclidean <- function(x) {
-  # Norm function in a Euclidean space of any dimension
-  require(Matrix)
-  return(norm(matrix(x, ncol = 1), type = "F"))
-}

### 2024-05-09 recode_attempt1/projection.R

@@ -1,7 +1,0 @@
-projection <- function(x, f, initial.guess) {
-  DD <- function(t) {
-    return(dist.euclidean(x, f(t)))
-  }
-  est <- nlm(DD, p = initial.guess)
-  return(est$estimate)
-}

### 2024-05-09 recode_attempt1/weight_seq.R

@@ -1,65 +1,0 @@
-weight.seq <- function(x.obs, mu, sigma, epsilon=0.001, max.iter = 1000) {
-  
-  # x.obs is the dataset of interest.
-  #       There are n observations, with each observation being a D-dimensional point.
-  #       x.obs is an n x D matrix
-  # mu is a vector of the knots in a mixture density estimation
-  # sigma is the bandwidth of this density estimation
-  # epsilon is a predetermined tolerance of the Euclidean distance between thetas in two consecutive steps
-  # max.iter is a predetermined upper bound of the number of steps in this iteration
-  
-  require(Rfast)
-  
-  n.D <- dim(x.obs)
-  n <- n.D[1]
-  D <- n.D[2]
-  N <- dim(mu)[1]
-  
-  A <- matrix(NA, ncol = N, nrow = n)
-  
-  for (i in 1:n) {
-    A[i, ] <- ker(x.obs[i, ], mu = mu, sigma = sigma)
-  }
-  
-  theta.old <- rep(1 / N, N) # The initial guess of the weights, theta_j
-  abs.diff <- 10 * epsilon # Absolute value of the difference between "theta.new" and "theta.old"
-  count <- 0 # Counting the number of steps in this iteration
-  lambda.hat.old <- c(n, rep(-1, D)) # The initial guess of the Lagrangian multipliers
-  
-  while ((abs.diff > epsilon) & (count <= max.iter)) { # The iteration for computing desired thetas
-    W <- t(t(A) * theta.old)
-    W <- W / rowsums(W)
-    w <- colsums(W)
-    
-    flambda <- function(lambda) {
-      denom.temp <- rowsums(t(t(cbind(1, mu)) * lambda))
-      num.temp <- mu * w
-      
-      f1 <- sum(w / denom.temp)
-      f2 <- colsums(num.temp * as.vector(1 / denom.temp))
-      f <- dist.euclidean(f1, 1) + dist.euclidean(f2, colmeans(x.obs))
-      return(f)
-    }
-    
-    lambda.hat <- nlm(flambda, lambda.hat.old, iterlim = 1000)$estimate
-    
-    # We set the Lagrangian multipliers in the previous step
-    # as the initial guess in this step
-    
-    theta.new <- w / rowsums(t(t(cbind(1, mu)) * lambda.hat))
-    abs.diff <- dist.euclidean(theta.new, theta.old)
[…]

### 2024-06-05 paper/main.tex

@@
 Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently.
-As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials.
+As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials.
 To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time.
@@
 These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
-We show that the proposed method improves the estimation of surfaces of hippocampi and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
+We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
 
@@
 These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}).
-For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes.Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
+For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes.
+Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
 Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
@@
 When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
-Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
+Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
 Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}).
@@
 Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
-In a study on comparison of the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
 After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
@@
 \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain image of a cognitively normal person which compose the majority of the data used for training the segmentation models.
-In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
+In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
 To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
@@
 Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
-These methods consider time as an additional dimension in the data, by including time in the dimension reduction process.
+These methods consider time as an additional dimension in the data, including time in the dimension reduction process.
 Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
@@
 Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
-In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper.
+In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper.
 We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively.
@@
 Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
-Given a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $t \to \left( \frac{\partial ^2 f}{\partial t_i \partial t_j}(t) \right)_{1 \leq i, j \leq d}$.
+Given a function $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $r \to \left( \frac{\partial ^2 f}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$.
 The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
@@
   \label{def:principal_manifolds}
-  Given a random $D$-vector $\boldsymbol{X}$ associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments.
-Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty]$, and the following functionals:
+  Let $\boldsymbol{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments.
+Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty)$, and the following functionals:
   \begin{align}\label{eq:5}
@@
 We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e.
-$D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest.
+$D=3$ to $(r_1, r_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest.
 Now consider that the data are observed longitudinally during several visits by a study participant.
 Hence, the observed data are the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of $j$th study visit.
-Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.
+Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$.
 The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.4]{sim_case1_plot2}
-  \caption{Longitudinal simulated data with $D = 2$.
-Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are generated at 5 time points.
-Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
-  \label{fig:sample_data}
-\end{figure}
-
 Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
-It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e.
+It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g.
 modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
-The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
+The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
 In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
@@
 Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together.
-Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.
-Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered smooth over the other set of variables.
+Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of spatial roughness shows minimal changes over time, or vice versa.
+Generally, our proposed framework can be used in applications where a high-dimensional subset of variables of interest set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered smooth over the other set of variables.
 
@@
   \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}
-  \caption{LPME Algorithm Steps.
+  \caption{LPME algorithm steps, demonstrated using longitudinal simulated data with $D=2$.
+Data from the functions $x_1 = =r$ and $x_2 = \alpha \sin \left(\beta r + \frac{\pi}{2}\right)$ are generated at 5 time points.
+Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually.
 In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.
@@
 To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that  characterize the shape of the original data.
+Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
 The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
@@
 For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e.
-the brain region of interest at each time point) is likely to result in inconsistent parameterization directions across time points (i.e.
-low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another).
-This is demonstrated in Figure \ref{fig:inconsistent_parameterization}, which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}.
+the brain region of interest at each time point) is likely to result in inconsistent parameterization directions across time points.
+To provide a concrete example using the data shown in Figure \ref{fig:lpme_step1}, ISOMAP could estimate parameters according to the functions $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$.
+However, parameter estimates according to $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$ would be equally appropriate.
+In practice, ISOMAP varies in which solution it reaches, yielding inconsistent parameterizations between time points.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[height=7cm]{inconsistent_parameterization}
-  \caption{Simulated data with $D = 2$.
-The curves shown in this plot represent embedding maps estimated by PME for each time point, with time indicated by the color of the curve.
-The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates.
-Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps.
-For most functions, the $X$ values increase with increasing values for $R$.
-However, two functions have $X$ values that decrease with increasing $R$.
-The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
-  \label{fig:inconsistent_parameterization}
-\end{figure}
 
@@
 This approach is based on the assumption that the shape of the structure remains relatively stable over time.
-In Section \ref{s:simulations}, we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure, and we consider some deviations from this assumption and their effect on the results.
+The results given in Section \ref{s:simulations} indicate that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.
 
@@
 
-With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$ resulting in a function taking the form
+With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_{t=1}^T$ as the response values, with weights $\left\{w_t\right\}_{t=1}^T$ resulting in a function taking the form
 \begin{equation}\nonumber
@@
 \end{equation}
+
+We note that, once the spline knots have been harmonized across time points, this function takes the form of a case in the varying-coefficient model framework discussed in \cite{}.
 
@@
     2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
-    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{cos} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
+    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
     4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\

### 2024-06-05 paper/main.tex

@@
 
-Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain.
-Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy.
+Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain.
+Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy.
 Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently.
@@
 This approach mitigates the effects of random disturbances to the manifold between time points.
-We propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds.
+In addition, we propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds.
 We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
@@
 
+\subsection{A Motivating Question from the Medical Imaging}
+
 Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
@@
 
-Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
-Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}).
+\subsection{A Brief Overview of Existing Approaches and Our Contributions}
+
+Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
+Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}), single-cell biology (\cite{ding2023learning}), and robotics (\cite{gao2023k}; \cite{gao2024bi}).
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
@@
 Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
+Unlike most commonly used manifold learning methods, this approach results in an explicit functional representation of the manifold estimate.
 We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
@@
 However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+
 Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
-
-Our work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time.
+This work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time.
 One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this manuscript.
-The data are publicly available and are hosted at \texttt{adni.loni.usc.edu}.
+The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}.
 The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.
@@
 The remainder of this article is laid out as follows.
-In Section \ref{s:PME}, we review the principal manifold framework.
+In Section \ref{s:PME}, we review the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}.
 In Section \ref{s:LPME}, we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show our proposed novel  algorithm for dimension reduction using the longitudinal principal manifold framework.
@@
 
-\section{Principal Manifolds and Estimation Algorithm}\label{s:PME}
+\section{Principal Manifold Estimation Algorithm}\label{s:PME}
 
 The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
-However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection.
-In addition, applying the methods in high-dimensional settings is problematic.
+However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings.
 Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
@@
 We assume that a data cloud is observed in the $D$ dimensional space.
-In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest.
+In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud (i.e., $D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest (i.e., $d=2$).
 For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
@@
 The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
-Most manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
+Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
 The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for two- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
@@
   \label{def:principal_manifolds}
-  Let $\boldsymbol{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments.
+  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\mathbf{X}$ has compact support and finite second moments.
 Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty)$, and the following functionals:
-  \begin{align}\label{eq:5}
-&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
+  \begin{align}\label{eq:pme_kappa}
+&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\mathbf{X} - f(\pi_g(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
 &\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
   \end{align}
-  The principal manifold of $\boldsymbol{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
+  The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
   \begin{equation}\nonumber
@@
   \end{equation}
-where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$ (also known as ``distributions" in the mathematical community, see Chapter 6 of \cite{rudin1991functional} for details), and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
+where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
 \end{definition}
-The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements.
-First is the expected distance from each point in the cloud to the fitted manifold, e.g.
-the distance between the fitted surface and the coordinates of the vertices on the 3-dimensional brain region of interest which we intend to minimize.
-The second element of the equation imposes the smoothness penalty on the estimated manifold.
-The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting.
-To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$.
-In this setting,  $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
-It is important to note that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.
+The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements.
+The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$.
+The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold.
+The coefficient $\lambda$ is a tuning parameter.
+In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
+Thus, finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ takes the form of a penalized regression problem.
+\cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form $$f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ l = 1, 2, \dots, D,$$
+under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud $\bf{X}$.
+This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$.
+These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.
 
@@
 
-The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices.
-In this section, we introduce an approach to extending this framework to longitudinal clouds of vertices.
+The PME framework described in Section \ref{s:PME} is an excellent approach for fitting a 2-dimensional surface that passes through the ``middle'' of a 3-dimensional cloud of vertices.
+In this section, we introduce an approach to extending the PME framework to longitudinal clouds of vertices.
 We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies.
-We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e.
-$D=3$ to $(r_1, r_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest.
-Now consider that the data are observed longitudinally during several visits by a study participant.
-Hence, the observed data are the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of $j$th study visit.
+We consider that 3-dimensional data clouds are observed longitudinally during several visits by a study participant.
+Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit.
 Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$.
@@
 Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
-It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g.
-modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
+It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g., modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
 The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
-In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
+In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
 In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.
 
-To give a more general representation of this problem, let $\left\{\bm{x}_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
+This problem can be posed for positive integer values of $D$ and $d$ with $d < D$.
+Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
 As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
@@
 
-\subsection{The Longitudinal Principal Manifold Framework}
+\subsection{The Longitudinal Principal Manifold Framework}\label{section: The Longitudinal Principal Manifold Framework}
 
@@
 \begin{definition}
-  \label{def:lpme} Given $\boldsymbol{X} = \{ \boldsymbol{X}_t \}_{t=1}^T$, denoting the collection of data clouds observed at all time points, where $\boldsymbol{X}_t$ is the random  $D$-vector observed at time $t$, the manifold $M_F^{d+1}$, which is representative of $\boldsymbol{X}$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
+  \label{def:lpme} Given $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$, denoting the collection of data clouds observed at all time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, the manifold $M_F^{d+1}$, which is representative of $\mathbf{X}$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
 \begin{align}\label{eq:newKappa}
-  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
+  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
   &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt, \nonumber
 \end{align}
-where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
-Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D}\times\mathbb{R})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$ and the coordinates of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
+Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
 \end{definition}
 
-By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and impose smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$.
+By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and impose smoothness of the function $F$ over time regularized by the tuning parameter $\gamma$.
 These goals correspond to minimizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively.
@@
 
-Given the framework proposed in the previous section, estimation of the longitudinal principal manifold entails minimizing the functional in (\ref{eq:newKappa}).
+Given the framework proposed in Section \ref{section: The Longitudinal Principal Manifold Framework}, estimation of the longitudinal principal manifold entails minimizing the functional in (\ref{eq:newKappa}).
 Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).
@@
   \caption{LPME algorithm steps, demonstrated using longitudinal simulated data with $D=2$.
-Data from the functions $x_1 = =r$ and $x_2 = \alpha \sin \left(\beta r + \frac{\pi}{2}\right)$ are generated at 5 time points.
-Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually.
+Data from the functions $x_1 = r$ and $x_2 = \alpha \sin \left(\beta r + \frac{\pi}{2}\right)$ are generated at 5 time points.
+Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually.
 In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.
@@
 
-\subsection*{Sample Size Reduction}
+\subsubsection{Sample Size Reduction}
 
@@
 Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is large.
-To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
 Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
@@
 
-\subsection*{Initialization}
+\subsubsection{Initialization}
 
@@
 However, implementation of commonly used approaches for initialization without considering the nuances posed by longitudinal data can be problematic.
-For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e.
-the brain region of interest at each time point) is likely to result in inconsistent parameterization directions across time points.
-To provide a concrete example using the data shown in Figure \ref{fig:lpme_step1}, ISOMAP could estimate parameters according to the functions $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$.
+For instance, using Isomap to find parameterizations of similarly shaped structures (i.e., the brain region of interest at each time point) is likely to result in inconsistent parameterizations across time points.
+To provide a concrete example using the data shown in Figure \ref{fig:lpme_step1}, Isomap could estimate parameters according to the functions $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$.
 However, parameter estimates according to $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$ would be equally appropriate.
-In practice, ISOMAP varies in which solution it reaches, yielding inconsistent parameterizations between time points.
+In practice, since Isomap is applied at each time point separately, it varies in which solution it reaches, yielding inconsistent parameterizations between time points.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
@@
 
-\subsection*{Fitting}
+\subsubsection{Fitting}
 
@@
 This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_step3}.
-The fit of each $\hat{f}_t$ is measured using $\tau_t$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\hat{f}_t$.
-At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
-The principal manifold $\widehat{f_t}$ at time $t$ is characterized by spline coefficients.
-Hence, we estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
+The fit of each $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I}\sum_{i=1}^{I}\| \mathbf{X}_t - \widehat{f_t}(\pi_{\widehat{f_t}}(\mathbf{X}_t))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\widehat{f_t}$.
+Further smoothing of the time-specific manifold estimates with respect to time is needed to consider curvature in the time dimension, corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
 
-At each time point, the estimated principal manifold has a spline form.
-As detailed by \cite{greenSilverman1994}, smoothing splines are fitted at a collection of knots.
-This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the knot values used as inputs when the functions are fitted must be identical.
-Hence, to obtain coefficients that can be compared across time points, a grid of knots $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the knots in each of the $d$ dimensions is selected and used as input for the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point.
-This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$.
-Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing the following function.
+As described in Section \ref{s:PME}, the principal manifold $\widehat{f_t}$ at time $t$ has spline form and is characterized by spline coefficients $\{s_{j, l}\}_{j=1, l=1}^{N, D}$ and $\{\alpha_{k, l}\}_{k=1, l=1}^{d+1, D}$.
+To smooth over the functions $\widehat{f_t}$ for each time $t$, we seek to estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
+As detailed by \cite{greenSilverman1994}, smoothing splines are fitted at a collection of knots corresponding to the data points.
+Because time-specific principal manifold estimates were fit with different cluster centers for each time point, the coefficients for each function are not comparable.
+Hence, to obtain coefficients that can be compared across time points, we design a grid of knots $\left\{\mathbf{r}_i\right\}_{i=1}^{N} \subset \mathbb{R}^d$ ranging from the minimum to the maximum of the approximated parameters $\pi_{\widehat{f_t}}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $t = 1, \dots, T$.
+This grid is used as input for the estimated embedding functions $\widehat{f_t}$ at all time points, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point.
+This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\mathbf{r}_i)$ for all $i$ and $t$.
+Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing the following function.
 \begin{equation}\label{eq:splineL}
-    \mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t,
+    \mathcal{L}(f_t) = (\mathbf{Y}_t - \mathbf{E}\mathbf{s}_t - \mathbf{R}^\T\mathbf{\alpha}_t)^\T(\mathbf{Y}_t - \mathbf{E}\mathbf{s}_t - \mathbf{R}^\T\mathbf{\alpha}_t) + \lambda_t^*\mathbf{s}_t^\T \mathbf{E}\mathbf{s}_t,
 \end{equation}
-where $\boldsymbol{Y}_t$ is the $N \times D$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
+where $\mathbf{Y}_t$ is the $N \times D$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\mathbf{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\mathbf{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\mathbf{r}_j)$.
 
@@
     \begin{array}{cc}
-      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
-      \boldsymbol{R} & \boldsymbol{0}
+      \mathbf{E} + \lambda_t^* \mathbf{I} & \mathbf{R}^\T \\
+      \mathbf{R} & \mathbf{0}
     \end{array}
@@
     \begin{array}{c}
-      \boldsymbol{s}_t \\
-      \boldsymbol{\alpha}_t
+      \mathbf{s}_t \\
+      \mathbf{\alpha}_t
     \end{array}
@@
     \begin{array}{c}
-      \boldsymbol{Y}_t \\
-      \boldsymbol{0}
+      \mathbf{Y}_t \\
+      \mathbf{0}
     \end{array}
@@
 \end{equation}
-The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields the function
+The solution to Eq.~\eqref{eq:13} , denoted by $\mathbf{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\mathbf{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields the function
 \begin{equation}\nonumber
@@
 
-We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients $\boldsymbol{\omega}_t$, where the weights for each time point $t$ are equal to
-$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$
+We denote the manifold coefficients at each time point $t$ as $\mathbf{b}_t = (\mathbf{s}_t^{*\T}, \mathbf{\alpha}_t^{*\T})^\T$.
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients $\mathbf{b}_t$, where the weights for each time point $t$ are equal to
+$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$ Thus, the weights correspond to the normalized inverse errors of the PME estimates at each time point.
 
-With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_{t=1}^T$ as the response values, with weights $\left\{w_t\right\}_{t=1}^T$ resulting in a function taking the form
+With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{b}_t\right\}_{t=1}^T$ as the response values, with weights $\left\{w_t\right\}_{t=1}^T$ resulting in a function taking the form
 \begin{equation}\nonumber
@@
 \end{equation}
-where by defining the $T \times T$ matrix $\boldsymbol{A}$ as  $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\boldsymbol{T}$ as  $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$, we obtain the coefficients $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ as the solutions to
+where by defining the $T \times T$ matrix $\mathbf{A}$ as  $\mathbf{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\mathbf{T}$ as  $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = \operatorname{diag}(w_1, \dots, w_T)$, we obtain the coefficients $\mathbf{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\mathbf{\nu}=(\nu_1,\nu_2)^\T$ as the solutions to
 \begin{equation}
@@
   \begin{array}{ccc}
-    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{A} + 2\gamma\boldsymbol{A} & 2\boldsymbol{A}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{T} \\
-    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{A} & 2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{0} \\
-    \boldsymbol{T}^{T} & \boldsymbol{0} & \boldsymbol{0}
+    2\mathbf{A}\mathbf{W}\mathbf{A} + 2\gamma\mathbf{A} & 2\mathbf{A}\mathbf{W}\mathbf{T} & \mathbf{T} \\
+    2\mathbf{T}^\T\mathbf{W}\mathbf{A} & 2\mathbf{T}^\T\mathbf{W}\mathbf{T} & \mathbf{0} \\
+    \mathbf{T}^\T & \mathbf{0} & \mathbf{0}
   \end{array}
@@
   \begin{array}{c}
-    \boldsymbol{\delta} \\
-    \boldsymbol{\nu} \\
-    \boldsymbol{m}
+    \mathbf{\delta} \\
+    \mathbf{\nu} \\
+    \mathbf{m}
   \end{array}
@@
   \begin{array}{c}
-    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{\Omega} \\
-    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{\Omega} \\
-    \boldsymbol{0}
+    2\mathbf{A}\mathbf{W}\mathbf{B} \\
+    2\mathbf{T}^\T\mathbf{W}\mathbf{B} \\
+    \mathbf{0}
   \end{array}
@@
 \end{equation}
-where $\boldsymbol{\Omega}=(\boldsymbol{\omega}_1,\ldots, \boldsymbol{\omega}_T)^\T$.
+where $\mathbf{B}=(\mathbf{b}_1,\ldots, \mathbf{B}_T)^\T$.
 
-For a given $\gamma$, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\boldsymbol{\Omega}_{\gamma}(t) = \left(\boldsymbol{s}_{\gamma}(t)^\T, \boldsymbol{\alpha}_{\gamma}(t)^\T\right)^\T$.
-Hence, given $\gamma$, the estimated embedding function at time $t$ and $d$-dimensional parameterization $\boldsymbol{r}$ is
+For a given $\gamma$, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\mathbf{B}_{\gamma}(t) = \left(\mathbf{s}_{\gamma}(t)^\T, \mathbf{\alpha}_{\gamma}(t)^\T\right)^\T$.
+Hence, given $\gamma$, the estimated embedding function at time $t$ and $d$-dimensional parameterization $\mathbf{r}$ is
 \begin{equation}
-  F_{\gamma}(t, \boldsymbol{r}) = \sum_{j=1}^{N}\boldsymbol{s}_{\gamma}(t)_j \eta_{d}\left(\|\boldsymbol{r} - \boldsymbol{r}_j^*\|\right) + \sum_{k=1}^{d+1}\boldsymbol{\alpha}_{\gamma}(t)_k \phi_k(\boldsymbol{r}).
+  F_{\gamma}(t, \mathbf{r}) = \sum_{j=1}^{N}\mathbf{s}_{\gamma}(t)_j \eta_{d}\left(\|\mathbf{r} - \mathbf{r}_j^*\|\right) + \sum_{k=1}^{d+1}\mathbf{\alpha}_{\gamma}(t)_k \phi_k(\mathbf{r}).
 \label{eq:17}
@@
 
-We note that, once the spline knots have been harmonized across time points, this function takes the form of a case in the varying-coefficient model framework discussed in \cite{}.
+We note that, once the spline knots correspond across time points, this function takes the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}.
+However, instead of attempting to estimate each coefficient function individually, we consider all coefficients together as the output of a single smoothing spline function.
+The initial use of smoothing splines to estimate principal manifolds complicates the direct use of estimation approaches developed for this model class in our use case.
 
-\subsection*{Tuning}
+\subsubsection{Tuning}
 
-Once $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
+Once $F_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
 In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
@@
 \begin{equation}
-  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
+  \operatorname{MSD}(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - F_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
 \end{equation}
@@
 Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
+
+Further work must be done to demonstrate the convergence properties of the proposed algorithm.
+However, \cite{mengPrincipalManifoldEstimation2021} demonstrated that the function that minimizes Equation \ref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimate in a penalized regression setting (\cite{wahba1990}).
+These results motivate our expectation that the estimate obtained by the LPME algorithm will minimize Equation \ref{eq:newKappa}.
 
@@
 \begin{algorithm}
-\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
+\caption{Longitudinal Principal Manifold Estimation (LPME)}\label{alg:lpme}
   \KwData{Data points $\left\{X_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
@@
 
-  Define $\boldsymbol{\omega}$ by setting $\boldsymbol{\omega}_t = \left[\boldsymbol{s}_t, \boldsymbol{\alpha}_t\right]$\;
+  Define $\mathbf{\omega}$ by setting $\mathbf{\omega}_t = \left[\mathbf{s}_t, \mathbf{\alpha}_t\right]$\;
   \For{l = 1, 2, \dots, L} {
@@
   $\gamma^* = \arg\min_{\gamma}MSD(\gamma)$\;
-  $f^*(t, \boldsymbol{r}) = f_{\gamma^*}(t, \boldsymbol{r})$, where the form of $f^*(t, \boldsymbol{r})$ is given in \eqref{eq:17}
+  $f^*(t, \mathbf{r}) = f_{\gamma^*}(t, \mathbf{r})$, where the form of $f^*(t, \mathbf{r})$ is given in \eqref{eq:17}
 \end{algorithm}
@@
 
-Notably, most manifold learning methods are developed under the assumption that the underlying manifold is not self-intersecting.
-This is not the case for in our motivating problem, where we are interested in modeling the surface of a brain region that, by definition, is a closed 3D shape.
+Notably, many manifold learning methods are developed under the assumption that the underlying manifold is not self-intersecting.
+This is not the case for in our motivating problem, where we are interested in modeling the surface of a brain region that is a closed 2-dimensional surface embedded in 3-dimensional space.
 This may often be the case in other applications where manifold learning is implemented.
-One such simple example is modeling digits in 2D space where the number 8 for instance is a self-intersecting manifold.
+One such simple example is modeling digits where the number 8 for instance is a self-intersecting manifold.
 Methods based on the principal curve framework, including the PME framework, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}.
-A common approach for dealing with this problem is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space.
+A common approach for dealing with this problem is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space (e.g., \cite{mengPrincipalManifoldEstimation2021}; \cite{dunsonInferringManifoldsNoisy2022}).
 To avoid this issue, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
+This approach is motivated by the concept of ``lift" in algebraic topology (\cite{hatcher2002algebraic}, Section 1.1).
 As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
@@
 
+Because the added dimensions duplicate information contained within the unaugmented observations, these dimensions can be discarded to reach estimates in the original $D$-dimensional space.
 This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
@@
 
-To assess the performance of the LPME algorithm, we use simulation studies to compare how closely (in terms of reduction of MSE) the LPME estimated underlying functions approximate the true data generating functions in comparison with that of three alternative approaches.
+To assess the performance of the LPME algorithm (Algorithm \ref{alg:lpme}), we use simulation studies to compare how closely (in terms of reduction of MSE) the LPME-estimated underlying functions approximate the true data generating functions in comparison with that of three alternative approaches.
 We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, as well as $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting.
@@
 Based on our observations of systematic noise sources in imaging data, we add various types of terms to these functions at each visit.
-In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural changes in the underlying manifold over time.
+In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between visits and represent structural changes in the underlying manifold over time.
 The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
@@
   \end{tabular}
-  \caption{Embedding functions used for simulation studies.}
+  \caption{Embedding functions used for simulation studies.
+Function $\bf{g}(t)$ denotes structural change in the underlying manifold over time, while $\zeta$ represents the scale of this change.
+Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}
   \label{table:simulation_embeddings}
@@
 In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
-The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}).
+The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package (\cite{hastie2022package}), developed using \texttt{R} (\cite{rSoftware2023}).
 In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
@@
 A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal.
-The sample size for each time point is set to 1,000, so each embedding map is simulated a total of 7,776 times.
+The sample size for each time point is set to 1,000 observations.
+Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 7,776 times.
 All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
@@
 Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.
-
-\begin{table}[h]
-  \centering
-  \begin{tabular}{|c c c c c|}
-    \hline
-    Case & Data & LPME & PME & PC/PS \\
-    \hline
-    1 & 0.223 (0.256) & {\bf 0.125 (0.161)} & 0.268 (0.850) & 0.189 (0.245) \\
-    2 & 0.514 (0.408) & {\bf 0.384 (0.648)} & 0.843 (1.93) & 0.600 (0.296) \\
-    3 & 0.446 (0.445) & {\bf 0.401 (0.446)} & 0.507 (0.594) & 0.412 (0.423) \\
-    4 & 30.7 (88.1) & {\bf 27.7 (260)} & 30.7 (88.2) & 30.6 (88.1) \\
-    5 & 0.980 (0.771) & {\bf 0.791 (0.845)} & 1.04 (1.07) & 0.934 (0.713) \\
-    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & {\bf 1.01 (2.11)} \\
-    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & {\bf 1.95 (0.800)} \\
-    8 & 0.226 (0.275) & {\bf 0.136 (0.169)} & 0.242 (0.311) & 0.274 (0.243) \\
-    \hline
-  \end{tabular}
-  \caption{MSD comparison to true values, Mean (SD). For each case, the lowest algorithm-specific mean (SD) are highlighted in bold. }
-  \label{table:simulation_results_mean}
-\end{table}
 
@@
   \end{tabular}
-  \caption{MSD comparison to true values, Median (IQR). The lowest algorithm-specific median (IQR) are highlighted in bold.}
+  \caption{Mean Squared Distance comparison of simulated data, longitudinal principal manifold estimation (LPME)-, principal manifold estimation (PME)-, and principal curve / surface-based estimates to true values, Median (IQR).
+The lowest algorithm-specific median (IQR) are highlighted in bold.}
   \label{table:simulation_results_median}
@@
 
-The mean, median, standard deviation and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}.
+The median and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_median}.
+Analogous mean and standard deviation summaries of the results are given in the Appendix.
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
@@
 
-While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
-Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study.
-Follow-up duration for these participants ranged from approximately 6 months to 12 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study.
+While the original ADNI 1 cohort included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 463 participants, of whom 130 were cognitively normal, 223 displayed symptoms of mild cognitive impairment, and 101 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
+Diagnostic information was unavailable for 9 of the 463 participants for whom imaging data was available.
+Of the 130 cognitively normal participants, 34 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study.
+Follow-up duration for these participants ranged from 0 months to approximately 52 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study.
 To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow-up were considered.
-Our final analysis set included 250 participants.
-Of these remaining participants, 80 were cognitively normal, 126 had mild cognitive impairment, and 44 were diagnosed with AD at their baseline visits.
-Due to the changes experienced by those with AD, we focused on two brain regions of interest starting from modeling the hippocampi of the selected individuals then applying the proposed methods to model the thalamuses.
+Our final analysis set included 236 participants.
+Of these remaining participants, 88 were cognitively normal, 107 had mild cognitive impairment, and 41 were diagnosed with AD at their baseline visits.
+We focused on two brain regions of interest: the hippocampus and the thalamus.
+The hippocampus is of interest due to the changes experienced by those with AD, while the shape of the thalamus more closely aligns with the spherical structure used in simulation case 8.
 
-Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}.
-Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation.
-Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
-The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in each dimension of Euclidean space and centered around the origin.
+Images were processed using FSL via the \texttt{fslr}package in \texttt{R}, with FSL's FLIRT method linearly registering the images to MNI space, and the FIRST method being used for image segmentation.
+Following image segmentation, the surfaces of each region were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
+The estimated surface positions were then standardized to a maximum distance of one from the origin in each dimension of Euclidean space and centered around the origin.
 Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.
@@
 Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}.
+The previously described data augmentation approach was used to enable the fitting of closed surfaces.
 Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data.
 There are two main sources of discrepancies between the data and the LPME estimate.
-First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
+First, at time points where the observed surface changes orientation, as is visible between the first two time points, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
 The second difference between the LPME estimates and the observed data is seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of curves in the observed data.
@@
 The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
-The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}.
-It is important to note that, instead of using the volume estimates produced by FSL's FIRST method during the segmentation process, the same volume estimation method used for the LPME- and PME-based estimates is used to reach estimates based on the raw data as well to align processing steps as closely as possible.
+The code used to implement this approach is available at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
+To align processing steps as closely as possible, the volume estimates shwon for the raw data use the same volume estimation method used to find LPME- and PME-based approximations, rather than the volume estimates produced by FSL's FIRST method during segmentation.
 
-The volume estimates for randomly selected individuals can be found in Figures \ref{fig:lhipp_volume_comparison} and \ref{fig:lthal_volume_comparison}.
+The volume estimates for randomly selected individuals can be found in Figures \ref{fig:lhipp_volume_comparison} and \ref{fig:lthal_volume_comparison}, corresponding to estimates for the left hippocampus and left thalamus, respectively.
 Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
@@
 All three sources show substantial variability between study visits, particularly for subjects 137\_S\_0283 and 137\_S\_0686.
-In Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
+Considering the thalamus estimates in Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
 However, the LPME estimates demonstrate the ability to smooth over this variability, resulting in volume estimates that are not as sensitive to differences between time points.
@@
 A summary of the between-visit variability in the volume estimates of the subcortical structures considered here is shown in Table \ref{table:adni_volume_sds}.
-In cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e.
+In addition to reporting the standard deviations of the volume estimates, we also report regression-adjusted estimates of variability to account for cases in which the structure cannot be assumed to be constant over time, as is the case for those with AD.
+The regression-adjusted summaries estimate the standard deviation of the residuals of a simple linear regression of volume on time.
+Meanwhile, in cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e.
 raw data, LPME, and PME).
 We then took the mean of these values for each estimate source.
-When regression adjustments were used, instead of considering the raw volume estimates, we estimated the standard deviation of the residuals of a simple linear regression of volume on time.
-This additional step gives a better indication of smoothness when the volume of the structure cannot be assumed to be constant over time, as is the case for those with AD.
-Table \ref{table:adni_volume_sds} shows that PME yields the smoothest volume estimates for both hippocampi.
+Table \ref{table:adni_volume_sds} shows that PME yields the smoothest volume estimates for both hippocampuses.
 This conclusion aligns with the visual evidence of inconsistency in the LPME estimates caused by the irregular shape of the hippocampus.
 However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.
-
-%\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}
-
-% \begin{figure}[h]
-%   \centering
-%   \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
-%   \hfill
-%   \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
-%   \caption{Left Hippocampus Cross Section and Volume Estimates.
-In (a), we can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of surface in the observed data.
-In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it, corresponding to an unexpectedly high volume estimate seen in (b).
-Volume estimates derived from the observed data are consistently higher than those obtained from the embedding functions of PME and LPME estimates, respectively.}
-%   \label{fig:adni_cross_section}
-% \end{figure}
 
@@
   \hfill
-  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_isomap_plot}}}
+  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_Isomap_plot}}}
   \hfill
@@
     \hline
-    Left Hippocampus & No & 213 & 225 & \textbf{187} \\
-    Right Hippocampus & No & 210 & 236 & \textbf{186} \\
-    Left Hippocampus & Yes & 173 & 190 & \textbf{150} \\
-    Right Hippocampus & Yes & 174 & 206 & \textbf{154} \\
-    Left Thalamus & No & 78.3 & \textbf{49.0} & 67.8 \\
-    Right Thalamus & No & 139 & \textbf{112} & 122 \\
-    Left Thalamus & Yes & 62.8 & \textbf{29.0} & 53.0 \\
-    Right Thalamus & Yes & 115 & \textbf{85.3} & 99.9 \\
+    Left Hippocampus & No & 129 & 132 & \textbf{106} \\
+    Right Hippocampus & No & 137 & 136 & \textbf{110} \\
+    Left Hippocampus & Yes & 100 & 108 & \textbf{87.1} \\
+    Right Hippocampus & Yes & 108 & 121 & \textbf{85.9} \\
+    Left Thalamus & No & 57.5 & \textbf{31.3} & 42.0 \\
+    Right Thalamus & No & 108 & \textbf{66.4} & 85.3 \\
+    Left Thalamus & Yes & 45.8 & \textbf{20.6} & 35.4 \\
+    Right Thalamus & Yes & 89.3 & \textbf{50.1} & 73.1 \\
     \hline
@@
 In practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
-In such cases, we recommend the use of alternative methods, such as Locally Linear Embedding (CITATION) or Laplacian Eigenmaps (CITATION), to generate the initial parameterization to alleviate this concern.
+In such cases, we recommend the use of alternative methods, such as Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}) or Laplacian Eigenmaps (\cite{belkin2003laplacian}), to generate the initial parameterization to alleviate this concern.
 Opportunities also exist to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise.
@@
 The proposed augmentation with polar or spherical coordinates, allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
-However, while this approach addressed a problem posed by well-behaved circular or spherical structures, such as the thalamus, its success may not extend to settings with more complex data.
+However, while this approach addressed a problem posed by structures with roughly circular or spherical shapes, such as the thalamus, its success may not extend to settings with more complex data.
 In the case of the hippocampus, potential intersections in the spherical coordinates may result in the overlaps in certain regions of the estimated manifolds.
@@
 Estimation of volumes of the resulting manifolds is essential, as the volumes are often used as outcomes in modeling disease processes over time in clinical trials and observational studies.
-In this manuscript, the volumes were obtained by simply computing the number of within region voxels resulting from each fitting procedure.
+In this manuscript, the volumes were obtained by simply computing the number of within-region voxels resulting from each fitting procedure.
 If the resulting manifold has gaps, however, this procedure may not be possible to implement.
@@
 Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.
-In population-level modeling of the data, it will be important to also consider registration of the participant-level MRI data to a common template space, raising important considerations for alignment of the shapes without loosing important information.
+In population-level modeling of the data, it will be important to also consider registration of the participant-level MRI data to a common template space, raising important considerations for alignment of the shapes without losing important information.
 
@@
 
+\newpage
+
+\appendix
+
+\section*{Appendix}
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{|c c c c c|}
+    \hline
+    Case & Data & LPME & PME & PC/PS \\
+    \hline
+    1 & 0.223 (0.256) & {\bf 0.125 (0.161)} & 0.268 (0.850) & 0.189 (0.245) \\
+    2 & 0.514 (0.408) & {\bf 0.384 (0.648)} & 0.843 (1.93) & 0.600 (0.296) \\
+    3 & 0.446 (0.445) & {\bf 0.401 (0.446)} & 0.507 (0.594) & 0.412 (0.423) \\
+    4 & 30.7 (88.1) & {\bf 27.7 (260)} & 30.7 (88.2) & 30.6 (88.1) \\
+    5 & 0.980 (0.771) & {\bf 0.791 (0.845)} & 1.04 (1.07) & 0.934 (0.713) \\
+    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & {\bf 1.01 (2.11)} \\
+    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & {\bf 1.95 (0.800)} \\
+    8 & 0.226 (0.275) & {\bf 0.136 (0.169)} & 0.242 (0.311) & 0.274 (0.243) \\
+    \hline
+  \end{tabular}
+  \caption{MSD comparison to true values, Mean (SD). For each case, the lowest algorithm-specific mean (SD) are highlighted in bold. }
+  \label{table:simulation_results_mean}
+\end{table}
+
 \end{document}

### 2024-06-05 paper/main.tex
Updates from Overleaf

@@
   \includegraphics[width=\textwidth]{sim_case1}
-  \includegraphics[width=\textwidth]{sim_case5}
-  \caption{Scatterplots of two example simulated data clouds (gray points) with $d=1$ overlaid true embedding manifold (red), the estimated LMPE function (blue, dashed), PME (green, dotted), and principal curve or surface (purple, dot-dashed).}
-  \label{fig:sim_case1}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
   \includegraphics[width=\textwidth]{sim_case7}
-  \caption{Scatterplots of two example simulated data clouds (gray points) with $d=2$ overlaid true embedding manifold (red), the estimated LMPE function (blue, dashed), PME (green, dotted), and principal curve or surface (purple, dot-dashed).}
-  \label{fig:sim_case6}
+  \caption{Scatterplots of two example simulated data clouds (gray points) with $d=1$ and $d=2$ overlaid with points on the true embedding manifold (red), the estimated LMPE function (blue), PME (green), and principal curve or surface (purple).}
+  \label{fig:sim_results}
 \end{figure}

### 2024-06-07 paper/main.tex

@@
 
-Visualizations of the results from some example simulated cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}.
+Visualizations of the results from some example simulated cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case7}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
   \centering
-  \includegraphics[width=\textwidth]{sim_case1}
-  \includegraphics[width=\textwidth]{sim_case7}
+  \subfloat[\centering Simulation Case 1]{{\includegraphics[height=8cm]{sim_case1}} \label{fig:sim_case1}}%
+  \vfill
+  \subfloat[\centering Simulation Case 7]{{\includegraphics[height=8cm]{sim_case7}} \label{fig:sim_case7}}
+
   \caption{Scatterplots of two example simulated data clouds (gray points) with $d=1$ and $d=2$ overlaid with points on the true embedding manifold (red), the estimated LMPE function (blue), PME (green), and principal curve or surface (purple).}
@@
   %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
-  \includegraphics[height=8cm]{adni_plots/adni_cross_section}
-  \caption{Left Hippocampus Cross Section.
-LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the surface in the observed data.
-In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap.
-Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.}
-  \label{fig:lhipp_cross_sections}
-\end{figure}
+  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[height=8cm]{adni_plots/adni_cross_section}} \label{fig:lhipp_cross_sections}}%
+  \vfill
+  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[height=8cm]{adni_plots/adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}
 
-\begin{figure}[h]
-  \centering
-  %\includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section}
-  \includegraphics[height=8cm]{adni_plots/adni_lthal_cross_section}
-  \caption{Left Thalamus Cross Section. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
-  \label{fig:lthal_cross_sections}
+  \caption{Left Hippocampus and Left Thalamus Cross Sections.
+LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface.
+In the fifth hippocampus observation with $t=3.04$, the LPME-estimated surface has a gap.
+Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.
+The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
+  \label{fig:adni_cross_sections}
 \end{figure}

### 2024-06-10 code/lpme_adni_results.R
update volume comparison plots

@@ -397,11 +397,15 @@ thal_sd_med <- thal_info_sd %>%
 
 print(thal_sd_med, width = Inf)
 
-set.seed(6819)
+set.seed(8310)
+sampled_patnos <- hipp_info_ts %>% 
+  sample_n_keys(3) %>% 
+  .$patno %>% 
+  unique()
 hipp_info_ts %>%
-  sample_n_keys(6) %>%
+  filter(patno %in% sampled_patnos) %>%
   ggplot(aes(x = time_from_bl, group = patno)) +
-  geom_point(aes(y = lhipp_data_vol2), color = colors[1]) +
+  geom_point(aes(y = lhipp_data_vol2), shape = 16, color = colors[1]) +
   geom_line(aes(y = lhipp_data_vol2), color = colors[1]) +
   # geom_smooth(
   #   aes(y = lhipp_data_vol2),
@@ -411,7 +415,7 @@ hipp_info_ts %>%
   #   linetype = "dashed",
   #   linewidth = 1
   # ) +
-  geom_point(aes(y = lhipp_vol_lpme2), color = colors[2]) +
+  geom_point(aes(y = lhipp_vol_lpme2), shape = 15, color = colors[2]) +
   geom_line(aes(y = lhipp_vol_lpme2), color = colors[2]) +
   # geom_smooth(
   #   aes(y = lhipp_vol_lpme2),
@@ -421,7 +425,7 @@ hipp_info_ts %>%
   #   linetype = "dashed",
   #   linewidth = 1
   # ) +
-  geom_point(aes(y = lhipp_vol_pme2), color = colors[3]) +
+  geom_point(aes(y = lhipp_vol_pme2), shape = 17, color = colors[3]) +
   geom_line(aes(y = lhipp_vol_pme2), color = colors[3]) +
   # geom_smooth(
   #   aes(y = lhipp_vol_pme2),
@@ -437,10 +441,10 @@ hipp_info_ts %>%
 ggsave("paper/figures/adni_plots/adni_lhipp_volume_comp.png", dpi = 1500)
 
 thal_info_ts %>%
-  sample_n_keys(6) %>%
+  filter(patno %in% sampled_patnos) %>% 
   ggplot(aes(x = time_from_bl, group = patno)) +
-  geom_point(aes(y = lthal_data_vol2), color = colors[1]) +
-  geom_line(aes(y = lthal_data_vol2), color = colors[1]) +
+  geom_point(aes(y = lthal_data_vol2), shape = 16, color = colors[1]) +
[…]

### 2024-06-10 paper/main.tex

@@
   \centering
-  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
+  \subfloat{
+      \subfloat[\centering Data]{{\includegraphics[height=12cm]{adni_plots/adni_lhipp_data_plot}}}%
       \hfill
-  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}
+      \subfloat[\centering LPME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lhipp_lpme_isomap_plot}}}
       \hfill
-  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_pme_plot}}}
+      \subfloat[\centering PME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lhipp_pme_plot}}}
+      \label{fig:adni_result}
+  }
+  \hfill
+  \subfloat{
+      \subfloat[\centering Data]{{\includegraphics[height=12cm]{adni_plots/adni_lthal_data_plot}}}%
+      \hfill
+      \subfloat[\centering LPME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lthal_lpme_Isomap_plot}}}
+      \hfill
+      \subfloat[\centering PME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lthal_pme_plot}}}
+      \label{fig:adni_lthal_result}
+  }
+  \caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant.
+The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
+The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
 
-  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant.
-The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
-The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the structure at points with high levels of curvature.}
-  \label{fig:adni_result}
-\end{figure}
-
-\begin{figure}%
-  \centering
-  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_data_plot}}}%
-  \hfill
-  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_lpme_Isomap_plot}}}
-  \hfill
-  \subfloat[\centering PME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lthal_pme_plot}}}
-
-  \caption{Left Thalamus, Cognitive Healthy ADNI Participant}
-  \label{fig:adni_lthal_result}
 \end{figure}

### 2024-06-14 code/03_adni_mri_preprocess.qmd
use here package for managing working directory in rmarkdown file

@@
 library(fslr)         # includes functions for registration and segmentation
+library(here)
 library(oro.dicom)    # dicom file operations
@@
 parallel::stopCluster(cl = cl) # stop parallelized cluster
+
 ```
@@
 ```{r}
-patnos <- list.dirs("data/adni_processed_fsl", recursive = FALSE, full.names = FALSE)
-processed_dirs <- list.dirs("data/adni_processed_fsl", recursive = FALSE)
+patnos <- list.dirs(here("data/adni_processed_fsl"), recursive = FALSE, full.names = FALSE)
+processed_dirs <- list.dirs(here("data/adni_processed_fsl"), recursive = FALSE)
 
@@
 ```{r, eval=FALSE}
+
 for (dir_idx in 1:length(processed_dirs)) {
@@
 }
+
 ```
@@
 
-write_csv(hipp_info, "data/adni_fsl_hipp_info.csv")
-write_csv(thal_info, "data/adni_fsl_thal_info.csv")
+write_csv(hipp_info, here("data/adni_fsl_hipp_info.csv"))
+write_csv(thal_info, here("data/adni_fsl_thal_info.csv"))
 
-write_csv(lhipp, "data/adni_fsl_lhipp.csv")
-write_csv(rhipp, "data/adni_fsl_rhipp.csv")
+write_csv(lhipp, here("data/adni_fsl_lhipp.csv"))
+write_csv(rhipp, here("data/adni_fsl_rhipp.csv"))
 
-write_csv(lthal, "data/adni_fsl_lthal.csv")
-write_csv(rthal, "data/adni_fsl_rthal.csv")
+write_csv(lthal, here("data/adni_fsl_lthal.csv"))
+write_csv(rthal, here("data/adni_fsl_rthal.csv"))
 
-write_csv(lhipp_surface, "data/adni_fsl_lhipp_surface.csv")
-write_csv(rhipp_surface, "data/adni_fsl_rhipp_surface.csv")
+write_csv(lhipp_surface, here("data/adni_fsl_lhipp_surface.csv"))
+write_csv(rhipp_surface, here("data/adni_fsl_rhipp_surface.csv"))
 
-write_csv(lthal_surface, "data/adni_fsl_lthal_surface.csv")
-write_csv(rthal_surface, "data/adni_fsl_rthal_surface.csv")
+write_csv(lthal_surface, here("data/adni_fsl_lthal_surface.csv"))
+write_csv(rthal_surface, here("data/adni_fsl_rthal_surface.csv"))
 ```

### 2024-06-14 code/01_lpme_simulations.qmd
add notebook to run lpme simulations

---
title: "LPME Simulation Studies"
author: "Robert Zielinski"
date: "April 23, 2024"
output: html_document
---

This notebook includes all code necessary to replicate the results of the simulations comparing performance between the LPME, PME, and PC/PS algorithms on data generated from a common underlying manifold, but with varying degrees of noise added between time points. The LPME algorithm is designed to obtain *smooth* estimates of the manifolds underlying noisy data. Consequently, if the algorithm achieves its stated goal, it will not show the best goodness of fit to the observed data in these conditions. Simulated data is useful in this setting because we can generate the data from a known underlying manifold, allowing us to measure the algorithm's success at recovering this known manifold.

```{r, message=FALSE}
library(tidyverse)
library(plotly)
library(pracma)
library(foreach)
library(doParallel)
library(furrr)
library(progress)
library(pme)
library(princurve)

# note that file paths are relative to the data/ directory where this notebook resides
source("functions/sim_data.R")
source("functions/calc_pme_est.R")
source("functions/calc_lpme_est.R")
source("prinSurf_v3.R")
```

Manifold learning algorithms traditionally have tradeoffs when compared to each other, meaning that an algorithm that displays strong performance for one manifold may be outperformed by another algorithm when considering a different manifold. Because of this, manifold learning algorithms are often evaluated with several manifolds meant to test an algorithm's performance on data with various attributes. We consider eight different manifolds:

1. For $r \in [-3, 3]$, $f(r) = (r, \sin(r + \frac{\pi}{2}))$.

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 1,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 1],
  z = sim_df[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 1],
    z = true_vals[, 3],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 1")
```

2. For $r \in [-3\pi, 3\pi]$, $f(r) = (r, \sin(r))$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 2,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 1],
  z = sim_df[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 1],
    z = true_vals[, 3],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 2")
```

3. For $r \in [-\frac{4}{5}\pi, \frac{\pi}{2}]$, $f(r) = (\cos(r), \sin(r))$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 3,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 1],
  z = sim_df[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 1],
    z = true_vals[, 3],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 3")
```

4. For $r \in [-1, 1]$, $f(r) = (r, r^2, r^3)$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 4,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 4")
```

5. For $r \in [0, 3\pi]$, $f(r) = (r, \cos(r), \sin(r))$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 5,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 5")
```

6. For $r_1 \in [-1, 1]$, $r_2 \in [-1, 1]$, $f(r_1, r_2) = (r_1, r_2, \|\mathbf{r}\|^2)$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 6,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 6")
```

7. For $r_1 \in [0, 3\pi]$, $r_2 \in [-1, 1]$, $f(r_1, r_2) = (r_1\cos(r_1), r_1\sin(r_1), r_2)$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 7,
  noise = 0.15,
  amp_noise = 0.25 / 25, # noise reduced to maintain integrity of shape
  period_noise = 0.25 / 25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 7")
```

8. For $r_1 \in [0, \pi]$, $r_2 \in [0, 2\pi]$, $f(r_1, r_2) = (\sin(r_1)\cos(r_2), \sin(r_1)\sin(r_2), \cos(r_1))$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 8,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25 / 5,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 8")
```

Simulation runs were conducted using functions for each case, with each function taking the following arguments:

- `max_time`: The maximum time from baseline of a set of observations, or the study duration.
- `interval`: The interval of study visits. Intervals between study visits were assumed to be constant in these simulations, but the LPME algorithm can handle settings with varying intervals between visits.
- `amp_noise`: Specifies the magnitude of between-time point noise impacting the amplitude of the functions used to generate the data.
- `shape_noise`: Specifies the magnitude of between-time point noise impacting the shape, or period, of the functions used to generate the data.
- `time_change`: Specifies the magnitude of structural change in the underlying manifold over time.
- `time_trend`: Describes the type of longitudinal change in the underlying manifold. May take one of four values - constant, linear, quadratic, or sinusoidal.
- `n`: The number of observations generated at each time point.
- `initialization_alg`: The manifold learning algorithm used to initialize the PME algorithm. May currently take one of three values - isomap, diffusion_maps, or laplacian_eigenmaps.
- `run`: Used to save multiple simulation results with the same parameter inputs.
- `print_plots`: Specify whether plots of preliminary results should be viewed. This is useful for debugging but is computationally expensive.

The functions are given in the code chunks below:

### Simulation Case 1

```{r}
sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 1,
    noise = 0.15,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    1,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal curve estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  # test all smoothing options for principal curve algorithm
  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 1,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers",
      verbose = FALSE
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    curves <- list()
    curve_vals <- list()
    curve_error <- vector()
    for (smoother_idx in 1:length(smoothing_options)) {
      curves[[smoother_idx]] <- principal_curve(
        temp_data,
        smoother = smoothing_options[smoother_idx]
      )
      # save projections of observations onto estimated principal curve
      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
    }
    opt_curve <- which.min(curve_error)
    # save principal curve results for smoothing approach with the lowest error
    principal_curve_result[[t]] <- curves[[opt_curve]]
    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 3-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal curve results,
  # projections and errors for all models
  sim_case1 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case1/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case1,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

### Simulation Case 2

```{r}
sim_error_case2 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 2,
    noise = 0.15,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    1,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected onto lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal curve estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  # test all smoothing options for principal curve algorithm
  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 1,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers",
      verbose = FALSE
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    curves <- list()
    curve_vals <- list()
    curve_error <- vector()
    for (smoother_idx in 1:length(smoothing_options)) {
      curves[[smoother_idx]] <- principal_curve(
        temp_data,
        smoother = smoothing_options[smoother_idx]
      )
      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
    }
    opt_curve <- which.min(curve_error)
    # save principal curve results for smoothing approach with lowest error
    principal_curve_result[[t]] <- curves[[opt_curve]]
    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 3-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal curve results,
  # projections and errors for all models
  sim_case2 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case2/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case2,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

### Simulation Case 3

```{r}
sim_error_case3 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 3,
    noise = 0.15,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # compute polar coordinates for the simulated observations
  # and points on the true manifold
  pol <- sim_df[, -1] %>%
    cart2pol()
  true_pol <- true_vals[, -1] %>%
    cart2pol()
  # augment data with polar coordinates, excluding the radius for this case
  sim_df <- cbind(sim_df, pol)[, -5]
  true_vals <- cbind(true_vals, true_pol)[, -5]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    1,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal curve estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  # test all smoothing options for principal curve algorithm
  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 1,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers"
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    curves <- list()
    curve_vals <- list()
    curve_error <- vector()
    for (smoother_idx in 1:length(smoothing_options)) {
      curves[[smoother_idx]] <- principal_curve(
        temp_data,
        smoother = smoothing_options[smoother_idx]
      )
      # save projections of observations onto estimated principal curve
      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
    }
    opt_curve <- which.min(curve_error)
    # save principal curve results for smoothing approach with the lowest error
    principal_curve_result[[t]] <- curves[[opt_curve]]
    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 3-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:3], pme_vals[.x, 1:3])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:3], lpme_vals[.x, 1:3])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:3], sim_df[.x, 1:3])^2
  ) %>%
    unlist() %>%
    mean()

  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:3], principal_curve_vals[.x, 1:3])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal curve results,
  # projections and errors for all models
  sim_case3 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    # lpme_gp_result = lpme_result_gp,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    # lpme_gp_error = lpme_error_gp,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case3/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case3,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

### Simulation Case 4

```{r}
sim_error_case4 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 4,
    noise = 0.15,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    1,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal curve estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  # test all smoothing options for principal curve algorithm
  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 1,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers"
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    curves <- list()
    curve_vals <- list()
    curve_error <- vector()
    for (smoother_idx in 1:length(smoothing_options)) {
      curves[[smoother_idx]] <- principal_curve(
        temp_data,
        smoother = smoothing_options[smoother_idx]
      )
      # save projections of observations onto estimated principal curve
      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
    }
    opt_curve <- which.min(curve_error)
    # save principal curve results for smoothing approach with the lowest error
    principal_curve_result[[t]] <- curves[[opt_curve]]
    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 4-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 4],
    frame = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 4],
      frame = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 4],
      frame = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 4],
      frame = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 4],
      frame = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal curve results,
  # projections and errors for all models
  sim_case4 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case4/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case4,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

## Simulation Case 5

```{r}
sim_error_case5 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 5,
    noise = 0.15,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    1,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal curve estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  # test all smoothing options for principal curve algorithm
  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 1,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers"
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    curves <- list()
    curve_vals <- list()
    curve_error <- vector()
    for (smoother_idx in 1:length(smoothing_options)) {
      curves[[smoother_idx]] <- principal_curve(
        temp_data,
        smoother = smoothing_options[smoother_idx]
      )
      # save projections of observations onto estimated principal curve
      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
    }
    opt_curve <- which.min(curve_error)
    # save principal curve results for smoothing approach with the lowest error
    principal_curve_result[[t]] <- curves[[opt_curve]]
    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 4-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 4],
    frame = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 4],
      frame = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 4],
      frame = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 4],
      frame = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 4],
      frame = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal curve results,
  # projections and errors for all models
  sim_case5 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    # lpme_gp_result = lpme_result_gp,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    # lpme_gp_error = lpme_error_gp,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )

  sim_dir <- "simulations/case5/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )

  saveRDS(
    sim_case5,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

### Simulation Case 6

```{r}
sim_error_case6 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 6,
    noise = 0.05,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme_algorithm on observations
  lpme_result <- lpme(
    sim_df,
    2,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal surface estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 2,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers"
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    principal_surface <- prinSurf(temp_data)
    # obtain principal surface error estimates
    surface_mse <- map(
      1:length(principal_surface),
      ~ principal_surface[[.x]]$MSE
    ) %>%
      unlist()
    opt_surface <- which.min(surface_mse)
    # save principal surface results
    # using opt_surface + 2 as list index because first two list entries are NULL
    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 4-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 4],
    frame = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 4],
      frame = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 4],
      frame = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 4],
      frame = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 4],
      frame = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  # keep principal_curve_error as variable name for continuity
  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal surface results,
  # projections and errors for all models
  sim_case6 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case6/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case6,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

### Simulation Case 7

```{r}
sim_error_case7 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 7,
    noise = 0.15,
    amp_noise = amp_noise / 25,
    period_noise = shape_noise / 25,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # compute polar coordinates for the simulated observations
  # and points on the true manifold
  # removing 4th column from consideration removes the depth from the Swiss roll
  pol <- sim_df[, -c(1, 4)] %>%
    cart2pol()
  true_pol <- true_vals[, -c(1, 4)] %>%
    cart2pol()
  # augment data with polar coordinates, excluding the radius
  sim_df <- cbind(sim_df, pol)
  sim_df <- sim_df[, -5]
  true_vals <- cbind(true_vals, true_pol)
  true_vals <- true_vals[, -5]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    2,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal surface estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 2,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers"
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    principal_surface <- prinSurf(temp_data)
    # obtain principal surface error estimates
    surface_mse <- map(
      1:length(principal_surface),
      ~ principal_surface[[.x]]$MSE
    ) %>%
      unlist()
    opt_surface <- which.min(surface_mse)
    # save principal surface results
    # using opt_surface + 2 as list index because first two list entries are NULL
    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 4-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 4],
    frame = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 4],
      frame = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 4],
      frame = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 4],
      frame = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 4],
      frame = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], pme_vals[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], lpme_vals[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], sim_df[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  # keep principal_curve_error as variable name for continuity
  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], principal_curve_vals[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal surface results,
  # projections and errors for all models
  sim_case7 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    # lpme_gp_result = lpme_result_gp,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    # lpme_gp_error = lpme_error_gp,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case7/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case7,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

### Simulation Case 8

```{r}
sim_error_case8 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 9,
    noise = 0.05,
    amp_noise = amp_noise,
    period_noise = shape_noise / 5,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }

  # compute spherical coordinates for the simulated observations
  # and points on the true manifold
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]
  sph <- sim_df[, -1] %>%
    cart2sph()
  true_sph <- true_vals[, -1] %>%
    cart2sph()
  # augment data with spherical coordinates
  sim_df <- cbind(sim_df, sph)
  true_vals <- cbind(true_vals, true_sph)

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    # results improved in this case by
    # excluding spherical coordinates from the standardization
    if (col_idx < 4) {
      col_max <- max(abs(sim_df[, col_idx]))
      sim_df[, col_idx] <- sim_df[, col_idx] / col_max
      true_vals[, col_idx] <- true_vals[, col_idx] / col_max
    }
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    2,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal surface estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 2,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers"
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    principal_surface <- prinSurf(temp_data)
    surface_mse <- map(
      1:length(principal_surface),
      ~ principal_surface[[.x]]$MSE
    ) %>%
      unlist()
    opt_surface <- which.min(surface_mse)
    # save principal surface results
    # using opt_surface + 2 as list index because first two list entries are NULL
    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 4-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 4],
    frame = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 4],
      frame = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 4],
      frame = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 4],
      frame = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 4],
      frame = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], pme_vals[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], lpme_vals[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], sim_df[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  # keep principal_curve_error as variable name for continuity
  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, 1:4], principal_curve_vals[.x, 1:4])^2
  ) %>%
    unlist() %>%
    mean()

  # save list of parameter values, LPME, PME, and principal surface results,
  # projections and errors for all models
  sim_case8 <- list(
    df = sim_df,
    times = time_vals,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    n = n,
    lpme_result = lpme_result,
    pme_results = pme_result,
    principal_curve_results = principal_curve_result,
    lpme_error = lpme_error,
    pme_error = pme_error,
    data_error = data_error,
    principal_curve_error = principal_curve_error,
    true_vals = true_vals,
    lpme_vals = lpme_vals,
    pme_vals = pme_vals,
    principal_curve_vals = principal_curve_vals,
    plot = p,
    initialization_algorithm = initialization_alg
  )
  sim_dir <- "simulations/case8/"
  if (!dir.exists(sim_dir)) {
    dir.create(sim_dir, recursive = TRUE)
  }
  filename <- paste0(
    initialization_alg,
    "_duration_",
    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
    "_interval_",
    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
    "_ampnoise_",
    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
    "_pernoise_",
    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
    "_n_",
    str_pad(as.character(n), 4, side = "left", pad = "0"),
    "_",
    time_trend,
    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
    "_run_",
    str_pad(as.character(run), 2, side = "left", pad = "0"),
    ".rds"
  )
  saveRDS(
    sim_case8,
    paste0(sim_dir, filename)
  )
  return(0)
}
```

Finally, the last code chunks contains the code needed to set the parameters for all simulation runs and run those simulations. We consider amplitude and shape noise values of $\{0, 0.05, 0.1, 0.25, 0.5, 1\}$, study durations of $\{1, 2, 5\}$, and intervals between study visits of $\{0.1, 0.25, 0.5\}$. Each simulation run has 1,000 observations and every combination of parameters is run one time. Finally, we consider three types of change to the underlying manifold (linear, quadratic, and sinusoidal), and the possibility of the manifold remaining constant over time. In cases where the manifold changes over time, the following magnitudes of longitudinal change were used: $\{0, 0.05, 0.1, 0.25, 0.5, 1\}$. Finally, in addition to using ISOMAP to initialize the PME algorithm, we also considered using Laplacian eigenmaps and diffusion maps. All possible combinations of parameter values were considered.

```{r, eval=FALSE}
# set parameter values
amp_noise_vals <- c(0, 0.05, 0.1, 0.25, 0.5, 1)
shape_noise_vals <- c(0, 0.05, 0.1, 0.25, 0.5, 1)
max_times <- c(1, 2, 5)
intervals <- c(0.1, 0.25, 0.5)
n_vals <- 1000
replicates <- 1
case <- 1:8
time_changes <- c(0, 0.05, 0.1, 0.25, 0.5, 1)
time_trends <- c("constant", "linear", "quadratic", "sinusoidal")
initialization_algorithms = c(
  "diffusion_maps",
  "laplacian_eigenmaps",
  "isomap"
)

# create data frame where each row is a combination of parameter values
param_grid <- expand_grid(
  case,
  amp_noise_vals,
  shape_noise_vals,
  n_vals,
  intervals,
  max_times,
  replicates,
  time_changes,
  time_trends,
  initialization_algorithms
) %>%
  data.frame()
```

The final chunk below includes the code required to run the simulations. This code loops through the rows of the data frame created in the chunk above, running the appropriate simulation function with the parameter values given in the row in question.

```{r, eval=FALSE}
# run code in parallel using the foreach package
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

set.seed(10972)
errors <- foreach(
  case = param_grid[, 1],
  amp_noise = param_grid[, 2],
  shape_noise = param_grid[, 3],
  n = param_grid[, 4],
  interval = param_grid[, 5],
  duration = param_grid[, 6],
  initialization_algorithm = param_grid[, 10],
  run = param_grid[, 7],
  time_change = param_grid[, 8],
  time_trend = param_grid[, 9],
  .inorder = FALSE,
  # the .export and .packages arguments are used to make functions and packages
  # available for the additional workers initialized by the parallel cluster.
  # in my experience, this may take some tinkering with to get everything
  # running properly
  .export = c(
    "sim_data",
    "calc_pme_est",
    "calc_lpme_est",
    "prinSurf",
    "cart2sph",
    "cart2pol"
  ),
  .packages = c(
    "tidyverse",
    "pme",
    "princurve",
    "plotly",
    "doParallel"
  )
) %dopar% {
    # control statements are used to run the appropriate function for each case value
    if (case == 1) {
      try({
        simlist <- sim_error_case1(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      })
    } else if (case == 2) {
      try(
        sim_error_case2(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    } else if (case == 3) {
      try(
        sim_error_case3(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    } else if (case == 4) {
      try(
        sim_error_case4(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    } else if (case == 5) {
      try(
        sim_error_case5(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    } else if (case == 6) {
      try(
        sim_error_case6(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    } else if (case == 7) {
      try(
        sim_error_case7(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    } else if (case == 8) {
      try(
        sim_error_case8(
          duration,
          interval,
          amp_noise,
          shape_noise,
          time_change,
          time_trend,
          n,
          initialization_algorithm,
          run,
          print_plots = FALSE
        )
      )
    }
  }

stopCluster(cl)
```

### 2024-06-14 paper/main.tex

@@
 These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
-We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
+We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
 
@@
 
-\subsection{A Motivating Question from the Medical Imaging}
+\subsection{A Motivating Question from Medical Imaging}
 
@@
 If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
-\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain image of a cognitively normal person which compose the majority of the data used for training the segmentation models.
+\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models.
 In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
 To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
+A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.
 
@@
 
-Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
+Nonlinear manifold learning methods have been extensively studied in the literature.
+However, many of them are not applicable in longitudinal studies, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}).
+That is, these methods include the time dimension in the dimension reduction process, as a result of which the time dimension is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of the time dimension.
+We use the oldest linear manifold learning method, principal component analysis (PCA, \cite{Pearson1901on}), as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.
+
+Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
+As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space.
+However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
+In addition, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
+However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+
+In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time.
+Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension.
+Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
+Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation~\eqref{eq: spline representation of the PME}, in Section \ref{s:PME}).
+Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit.
+We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+We demonstrate that the regularization involved in this process mitigates the effects of variability between time points.
+Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
+
+%Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
 These methods consider time as an additional dimension in the data, including time in the dimension reduction process.
@@
 
-Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
+%Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis (PCA) method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
 \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
@@
 
-Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
-This work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time.
-One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this manuscript.
+%Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
+
+This work was motivated by longitudinal imaging studies, where data of participants are collected during several study visits to observe changes in brain structure (among other variables of interest) over time.
+One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this article.
 The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}.
-The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.
+The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources, including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.
 
@@
 In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper.
-We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively.
+We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively.
 We assume that a data cloud is observed in the $D$ dimensional space.
-In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud (i.e., $D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest (i.e., $d=2$).
+In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure \ref{fig:adni_result}).
 For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
 Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
-Given a function $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $r \to \left( \frac{\partial ^2 f}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$.
+Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$.
 The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
 Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
-The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for two- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
+The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
 The principal manifolds are then defined as follows.
@@
   \label{def:principal_manifolds}
-  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\mathbf{X}$ has compact support and finite second moments.
-Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty)$, and the following functionals:
+  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability distribution $\mathbb{P}$ such that $\mathbf{X}$ has compact support, $\text{supp}(\mathbb{P})$, and finite second moments.
+Under some assumptions on limiting behavior and smoothness on function $f$, given a $\lambda \in [0, \infty)$, define the following functional:
   \begin{align}\label{eq:pme_kappa}
-&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\mathbf{X} - f(\pi_g(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
-&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber
+\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathbb{E}\|\mathbf{X} - f(\pi_f(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2. 
   \end{align}
@@
   \end{equation}
-where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
+where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
 \end{definition}
@@
 Thus, finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ takes the form of a penalized regression problem.
-\cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form $$f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ l = 1, 2, \dots, D,$$
-under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud $\bf{X}$.
+\cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form
+\begin{align}\label{eq: spline representation of the PME}
+    f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ \ \text{ for } \ l = 1, 2, \dots, D,
+\end{align}
+under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud.
+Importantly, $N$ is generally much smaller than the number of points in the data cloud (see Figure 3 of \cite{mengPrincipalManifoldEstimation2021}), which indicates that the representation in equation~\eqref{eq: spline representation of the PME} is computationally efficient.
 This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$.
@@
 
-The PME framework described in Section \ref{s:PME} is an excellent approach for fitting a 2-dimensional surface that passes through the ``middle'' of a 3-dimensional cloud of vertices.
-In this section, we introduce an approach to extending the PME framework to longitudinal clouds of vertices.
-We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies.
+In this section, we introduce an approach to extending the PME framework described in Section \ref{s:PME} to longitudinal point clouds.
+We first heuristically explain the goal of our proposed model and its importance in the segmentation of brain regions for longitudinal imaging studies.
+Our goal is to estimate a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
+The estimated surfaces across different time points enable further inference, e.g., modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
 We consider that 3-dimensional data clouds are observed longitudinally during several visits by a study participant.
-Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit.
-Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$.
-The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.
+Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit, and $(x_1, x_2, x_3)$ denotes a spatial point on the surface of the segmented brain region of interest.
 
-Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth.
-It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g., modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc.
-The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
+Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data, in which 2-dimensional point clouds are observed longitudinally at several time points.
+The goal in this simplified example is to fit a surface in 3-dimensional spacetime---comprising two spatial dimensions and one time dimension---that fits the longitudinal data $(x_1, x_2, t_j)$ in spacetime and is not affected by the random noise added at each time point and in space (see Figure \ref{fig:lpme_step4}).
+The PME framework does not directly apply due to the following:
+\begin{enumerate}
+    \item Suppose we apply the PME approach to the data $(x_1, x_2, t_j)$ in spacetime and obtain an estimated surface $(r_1, r_2) \mapsto F(r_1, r_2)\in\mathbb{R}^3$ in spacetime.
+It is likely that neither coordinate $r_1$ nor $r_2$ coincides with the time dimension---they are (potentially nonlinear) combinations of both time and spatial coordinates, which prevents meaningful interpretation of the time dimension.
+
+    \item If the manifold is estimated separately at each time point, the results may not be smooth over time.
+\end{enumerate}
+We expect that our proposed dimension reduction method will yield a smooth surface $(r,t)\mapsto F(r,t)$ in spacetime while ensuring that one coordinate, $t$, coincides with the time dimension.
+This can be achieved by the longitudinal principal manifold estimation (LPME) framework proposed in this section.
+
+%The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
 In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
@@
 
-This problem can be posed for positive integer values of $D$ and $d$ with $d < D$.
-Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points.
-As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
-However, in this updated case, the function determining the manifold may change over time.
+%Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$.
+Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points.
+As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
+In this setting, the manifold to estimate is represented by $(\mathbf{r},t)\mapsto F(\mathbf{r},t)\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
 
 \subsection{The Longitudinal Principal Manifold Framework}\label{section: The Longitudinal Principal Manifold Framework}
+
+Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$.
+Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points.
+As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
+In this setting, the manifold to estimate is represented by $(\mathbf{r},t)\mapsto F(\mathbf{r},t)=:f_t(\mathbf{r})\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
 
@@
 
-\begin{figure}%
+\begin{figure}[h]
 \centering
 \subfloat{
-      \subfloat[\centering Data]{{\includegraphics[height=12cm]{adni_plots/adni_lhipp_data_plot}}}%
+\label{fig:adni_lhipp_result}
+\subfloat[\centering Data]{
+\includegraphics[height=10cm]{adni_plots/adni_lhipp_data_plot}
+}
 \hfill
-      \subfloat[\centering LPME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lhipp_lpme_isomap_plot}}}
+\subfloat[\centering LPME]{
+\includegraphics[height=10cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
+}
 \hfill
-      \subfloat[\centering PME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lhipp_pme_plot}}}
-      \label{fig:adni_result}
+\subfloat[\centering PME]{
+\includegraphics[height=10cm]{adni_plots/adni_lhipp_pme_plot}
+}
 }
@@
 \subfloat{
-      \subfloat[\centering Data]{{\includegraphics[height=12cm]{adni_plots/adni_lthal_data_plot}}}%
+\label{fig:adni_lthal_result}
+\subfloat[\centering Data]{
+\includegraphics[height=10cm]{adni_plots/adni_lthal_data_plot}
+}
 \hfill
-      \subfloat[\centering LPME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lthal_lpme_Isomap_plot}}}
+\subfloat[\centering LPME]{
+\includegraphics[height=10cm]{adni_plots/adni_lthal_lpme_isomap_plot}
+}
 \hfill
-      \subfloat[\centering PME Estimate]{{\includegraphics[height=12cm]{adni_plots/adni_lthal_pme_plot}}}
-      \label{fig:adni_lthal_result}
+\subfloat[\centering PME]{
+\includegraphics[height=10cm]{adni_plots/adni_lthal_pme_plot}
+}
 }
@@
 The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
-
+\label{fig:adni_result}
 \end{figure}
 
-% \begin{figure}[h]
-%   \centering
-%   \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
+%\begin{figure}[h]
+%\centering
+%\subfloat{
+%\label{fig:adni_lhipp_result}
+%\begin{subfigure}
+%\includegraphics{adni_plots/adni_lhipp_data_plot}
+%\caption{Data}
+%\label{fig:adni_lhipp_result_data}
+%\end{subfigure}
+%}
+
 % \hfill
-%   \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
-%   \caption{Left Thalamus Cross Section and Volume Estimates.
-In (a), the cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.
-The volume estimates in (b) show similar general trends to the hippocampus estimates, with the estimates reached using the observed data being higher than the PME- and LPME-based estimates.
-However, the differences between the estimates taken from the PME and LPME results are more comparable.}
-%   \label{fig:adni_lthal_cross_section}
-% \end{figure}
-%
+%\subfloat{
+%\label{fig:adni_lthal_result}
+%\begin{subfigure}
+%\includegraphics{adni_plots/adni_lthal_data_plot}
+%\caption{Data}
+%\label{fig:adni_lthal_result_data}
+%\end{subfigure}
+%}
+
+%\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant.
+The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
+The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
+%\label{fig:adni_result}
+
+%\end{figure}
 
@@
     \centering
-  \includegraphics[width=\textwidth]{adni_plots/adni_lhipp_volume_comp}
-  \caption{Left Hippocampus Volume Estimates for Six ADNI Participants.
-All volumes are estimated by summing the volumes of all voxels contained within the structure estimate as described by segmented data (red), LPME (blue), and PME (green).
-When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of volume estimates, gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data.
-These gaps may also cause inconsistency in the LPME volume estimates, resulting in higher aggregate variability than for the data or PME estimates.}
+    \subfloat[\centering Left Hippocampus Volume Estimates]{
+      \includegraphics[height=8cm]{adni_plots/adni_lhipp_volume_comp}
       \label{fig:lhipp_volume_comparison}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth]{adni_plots/adni_lthal_volume_comp}
-  \caption{Left Thalamus Volume Estimates for Six ADNI Participants.
-Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by segmented data (red), LPME (blue), and PME (green).
-The PME-based volume estimates closely track the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data.
-The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participants 021\_S\_0626 and 053\_S\_0919.}
+    }
+    \vfill
+    \subfloat[\centering Left Thalamus Volume Estimates]{
+      \includegraphics[height=8cm]{adni_plots/adni_lthal_volume_comp}
       \label{fig:lthal_volume_comparison}
+    }
+    \caption{Left hippocampus and left thalamus volume estimates for three ADNI participants.
+Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (red, circles), LPME (blue, squares), and PME (green, triangles).
+When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of the volume estimates.
+Gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data.
+These gaps may also cause inconsistency in the LPME volume estimates, resulting in higher aggregate variability than for the data or PME estimates.
+When applied to the thalamus, the PME-based volume estimates demonstrate similar time point-to-time point changes in the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data.
+The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 011\_S\_0003.}
 \end{figure}

### 2024-06-25 paper/main.tex

@@
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url}
+
 %\documentclass[12pt]{amsart}
-\documentclass[11pt,reqno]{article}
+%\documentclass[11pt,reqno]{article}
 %\usepackage{cases}
@@
 %\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
-\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
-\RequirePackage{graphicx}%% uncomment this for including figures
+%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
+%\RequirePackage{graphicx}%% uncomment this for including figures
 
-\usepackage[
-backend=biber,
-style=authoryear,
-url=false,
-isbn=false,
-date=year,
-doi=false
-]{biblatex}
+%\usepackage[
+%backend=biber,
+%style=authoryear,
+%url=false,
+%isbn=false,
+%date=year,
+%doi=false
+%]{biblatex}
 
-\addbibresource{references.bib}
+%\addbibresource{references.bib}
 
@@
 
-Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain.
+Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease.
+Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies.
+To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time.
+The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point.
+This mitigates the effects of random disturbances to the manifold between time points.
+Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds.
+Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
+The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
+An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
+\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
+
+% Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain.
 Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy.
@@
 These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
-We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
+We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
+We present a comprehensive analysis of magnetic resonance imaging data from 263 participants demonstrating high levels of variability of regional volumes when using existing methods for segmentation of hippocampuses and thalamnuses and the advantages gained by using our proposed methods that leverage longitudinal regional trends for segmentation.
+\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
 
@@
 
-Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
 In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
@@
 However, many of them are not applicable in longitudinal studies, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}).
-That is, these methods include the time dimension in the dimension reduction process, as a result of which the time dimension is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of the time dimension.
+That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of the time dimension.
 We use the oldest linear manifold learning method, principal component analysis (PCA, \cite{Pearson1901on}), as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.
@@
 The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources, including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.
+We implement the proposed methods to estimate the surfaces of two regions of interest in the brain - the hippocampus and the thalamus.
+The thalamus is selected as a ``easy-to-fit" region of interest to evaluate the performance of our proposed methods when the underlying structure is somewhat close to a spherical shape.
+The hippocampus is more of a pear-shaped region and is selected as a region that may exhibit atrophy for people with AD.
 
@@
 Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
-The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection.
+The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection.
 The principal manifolds are then defined as follows.
@@
 \end{definition}
+
+It is important to note that, unlike manifold learning methods that use an eigendecomposition to directly estimate coordinates in the $d$-dimensional manifold space (e.g.
+Isomap, locally linear embedding, and others) with only an implicit representation of the embedding function $f$, principal manifolds may be estimated via an explicit approximation of the embedding function $f$, from which the projection index $\pi_f(x)$ and the desired coordinates on the manifold can be found.
+In this way, regression approaches become relevant to the manifold estimation problem.
+
 The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements.
@@
 In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
-Thus, finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ takes the form of a penalized regression problem.
-\cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form
+Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem making this a supervised learning problem.
+\cite{mengPrincipalManifoldEstimation2021} also show that this function takes spline form
 \begin{align}\label{eq: spline representation of the PME}
@@
 under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud.
-Importantly, $N$ is generally much smaller than the number of points in the data cloud (see Figure 3 of \cite{mengPrincipalManifoldEstimation2021}), which indicates that the representation in equation~\eqref{eq: spline representation of the PME} is computationally efficient.
+Importantly, $N$ tends to be much smaller than the number of points in the data cloud (see Figure 3 of \cite{mengPrincipalManifoldEstimation2021}), which indicates that the representation in equation~\eqref{eq: spline representation of the PME} is computationally efficient.
 This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$.
@@
 We consider that 3-dimensional data clouds are observed longitudinally during several visits by a study participant.
-Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit, and $(x_1, x_2, x_3)$ denotes a spatial point on the surface of the segmented brain region of interest.
+Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of the $j$th study visit, and $(x_1, x_2, x_3)$ denotes a spatial point on the surface of the segmented brain region of interest.
 
@@
 \begin{enumerate}
-    \item Suppose we apply the PME approach to the data $(x_1, x_2, t_j)$ in spacetime and obtain an estimated surface $(r_1, r_2) \mapsto F(r_1, r_2)\in\mathbb{R}^3$ in spacetime.
+    \item Suppose we apply the PME approach to the data $(x_1, x_2, t_j)$ in spacetime and obtain an estimated surface $(r_1, r_2) \mapsto F(r_1, r_2)\in\mathbb{R}^2\times\mathbb{R}$ in spacetime.
 It is likely that neither coordinate $r_1$ nor $r_2$ coincides with the time dimension---they are (potentially nonlinear) combinations of both time and spatial coordinates, which prevents meaningful interpretation of the time dimension.
@@
 \end{enumerate}
-We expect that our proposed dimension reduction method will yield a smooth surface $(r,t)\mapsto F(r,t)$ in spacetime while ensuring that one coordinate, $t$, coincides with the time dimension.
+We expect that our proposed dimension reduction method will yield a smooth surface $(t,r)\mapsto F(t,r)$ in spacetime, $\mathbb{R}^2\times\mathbb{R}$, while ensuring that one coordinate, $t$, coincides with the time dimension.
 This can be achieved by the longitudinal principal manifold estimation (LPME) framework proposed in this section.
@@
 As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
-In this setting, the manifold to estimate is represented by $(\mathbf{r},t)\mapsto F(\mathbf{r},t)\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
+In this setting, the manifold to estimate is represented by $(t,\mathbf{r})\mapsto F(t,\mathbf{r})\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
 
@@
 
-Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$.
-Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points.
+Hereafter, we utilize the following notation to describe the LPME framework for generic low and high dimensions, $d$ and $D$.
+Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each image $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points.
 As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
-In this setting, the manifold to estimate is represented by $(\mathbf{r},t)\mapsto F(\mathbf{r},t)=:f_t(\mathbf{r})\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
-
-Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively.
-For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g.
-the surface of the 3D brain region, at time $t$.
+In this setting, the manifold to estimate is represented by
+\begin{align}\label{eq: F form}
+    \begin{aligned}
+        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R}, \\
+    & (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
+    \end{aligned}
+\end{align}
+where $t$ coincides with the time dimension. We define a longitudinal principal manifold as follows.
 \begin{definition}
-  \label{def:lpme} Given $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$, denoting the collection of data clouds observed at all time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, the manifold $M_F^{d+1}$, which is representative of $\mathbf{X}$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
+  \label{def:lpme} Given a collection $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$ of data clouds observed at a series of time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ as follows
 \begin{align}\label{eq:newKappa}
-  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
-  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt, \nonumber
+  \begin{aligned}
+      \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
+  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt,
+  \end{aligned}
 \end{align}
-where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D}\times\mathbb{R})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$ and the coordinates of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
-Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation~\eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
+Then, the manifold $M_{F^{*}}^{d+1}:=\{F^*(t,\mathbf{r}): (t,\mathbf{r})\in \mathbb{R}^d\times\mathbb{R}\}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
 \end{definition}
 
-By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and impose smoothness of the function $F$ over time regularized by the tuning parameter $\gamma$.
+By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to (i) minimize the distance between each data point $\mathbf{X}_t$ and its projection $f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)$ on the fitted manifold at each time point $t$, (ii) penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and (iii) impose smoothness of the function $F$ over time regularized by the tuning parameter $\gamma$.
 These goals correspond to minimizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively.
-Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions.
-Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together.
 Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of spatial roughness shows minimal changes over time, or vice versa.
-Generally, our proposed framework can be used in applications where a high-dimensional subset of variables of interest set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered smooth over the other set of variables.
 
@@
 Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).
-At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping.
+Specifically, at each individual time point, we apply the PME algorithm (\cite{mengPrincipalManifoldEstimation2021}, Algorithm 2) to fit the data cloud at the time point and represent the fitted manifold using the spline form in equation~\eqref{eq: spline representation of the PME}.
+Hence, we have a collection of spline coefficients, i.e., the $s_{j,l}$ and $\alpha_{k,l}$ in equation~\eqref{eq: spline representation of the PME}, associated with each time point.
+Then, we smooth the time-dependent spline coefficients with respect to time.
+Details of this procedure are encapsulated in our proposed LPME algorithm, which consists of four steps: data reduction, initialization, fitting, and tuning.
+A visualization of these steps is shown in Figure \ref{fig:lpme_steps}.
+Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}.
+
+%At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping.
 Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space.
@@
 
-\subsubsection{Sample Size Reduction}
+\subsubsection{Data Reduction}
 
-In most practical manifold learning tasks data are often quite large, so we begin by implementing an approach for selecting a small number of points in $D$-dimensional space that represent the structure of the data cloud.
+In most practical manifold learning tasks, data are often quite large.
+We begin by deriving a small number of feature points in $D$-dimensional space that capture the intrinsic structure of the data cloud.
 This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest.
 Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is large.
-To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method (see \cite{mengPrincipalManifoldEstimation2021}, Algorithm 1) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
 Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
-The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
-By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
-The centers estimated in this step are then used to reach an initial low-dimensional parameterization.
+The number $N_t$ of clusters is estimated for each time point $t$ and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
+By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
+The centers $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$ estimated in this step are then used to reach an initial low-dimensional parameterization.
 
@@
 
-The next step in the algorithm for estimation of the longitudinal principal manifold is initialization.
-Different algorithms use different approaches for this step.
-For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve.
-However, implementation of commonly used approaches for initialization without considering the nuances posed by longitudinal data can be problematic.
-For instance, using Isomap to find parameterizations of similarly shaped structures (i.e., the brain region of interest at each time point) is likely to result in inconsistent parameterizations across time points.
-To provide a concrete example using the data shown in Figure \ref{fig:lpme_step1}, Isomap could estimate parameters according to the functions $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$.
-However, parameter estimates according to $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$ would be equally appropriate.
-In practice, since Isomap is applied at each time point separately, it varies in which solution it reaches, yielding inconsistent parameterizations between time points.
-Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.
+The next step in the LPME algorithm for the estimation of the longitudinal principal manifold is initialization.
+The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve.
+Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points.
+Here, consistency refers to the similarity of the parameterizations of similarly shaped point clouds.
+However, using Isomap to find parameterizations of similarly shaped manifolds (i.e., the surfaces of brain regions observed at consecutive time points) is likely to result in inconsistent parameterizations.
+For example, the following two drastically different expressions parameterize the same curve: $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$ and $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$.
+If Isomap is applied at each time point separately, it tends to yield inconsistent parameterizations between time points.
+Therefore, a different approach to generating an initial parameterization is needed for the LPME algorithm.
 
-To obtain the same parameterization of the manifold over time, instead of using Isomap at each time point separately, we propose using PME to develop a parameterization of the data taken from the first time point, then using the projection index associated with this estimate to find initial parameterizations of the centers for the remaining time points.
-This approach is based on the assumption that the shape of the structure remains relatively stable over time.
-The results given in Section \ref{s:simulations} indicate that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.
+To obtain consistent parameterizations of the obtained cluster centers $\left\{\mu_{j, t}\right\}_{j=1}^{N_t}$ over time $t$, we propose using the PME algorithm to develop a parameterization of the centers $\left\{\mu_{j, t_1}\right\}_{j=1}^{N_{t_1}}$ taken from the first time point $t_1$, then using the projection index associated with this estimate to find initial parameterizations of the centers for the remaining time points.
+This approach is based on the assumption that the underlying manifolds of interest remains relatively stable over time.
+%The results given in Section \ref{s:simulations} indicate that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.
 
@@
 
-Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using the cluster centers, weights and parameterizations associated with each time point to initialize a PME run.
-This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_step3}.
-The fit of each $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I}\sum_{i=1}^{I}\| \mathbf{X}_t - \widehat{f_t}(\pi_{\widehat{f_t}}(\mathbf{X}_t))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\widehat{f_t}$.
-Further smoothing of the time-specific manifold estimates with respect to time is needed to consider curvature in the time dimension, corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+Once the initial parameterization for each cluster center $\mu_{j, t}$ is obtained, the model fitting process begins by using the cluster centers $\mu_{j, t}$, weights $\hat{\theta}_{j, t}$ and the parameterizations associated with each time point to initialize a PME run.
+Specifically, the cluster centers $\mu_{j, t}$ and weights $\hat{\theta}_{j, t}$ replace those obtained using the HDMDE algorithm, and the Isomap-derived parameters obtained in Steps 1 and 2 of Algorithm 2 in \cite{mengPrincipalManifoldEstimation2021}, respectively.
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using multicolored lines in Figure \ref{fig:lpme_step3}.
+The performance of each fit $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I_t}\sum_{i=1}^{I_t}\| x_{it} - \widehat{f_t}(\pi_{\widehat{f_t}}(x_{it}))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\widehat{f_t}$.
 
-As described in Section \ref{s:PME}, the principal manifold $\widehat{f_t}$ at time $t$ has spline form and is characterized by spline coefficients $\{s_{j, l}\}_{j=1, l=1}^{N, D}$ and $\{\alpha_{k, l}\}_{k=1, l=1}^{d+1, D}$.
-To smooth over the functions $\widehat{f_t}$ for each time $t$, we seek to estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.
-As detailed by \cite{greenSilverman1994}, smoothing splines are fitted at a collection of knots corresponding to the data points.
-Because time-specific principal manifold estimates were fit with different cluster centers for each time point, the coefficients for each function are not comparable.
-Hence, to obtain coefficients that can be compared across time points, we design a grid of knots $\left\{\mathbf{r}_i\right\}_{i=1}^{N} \subset \mathbb{R}^d$ ranging from the minimum to the maximum of the approximated parameters $\pi_{\widehat{f_t}}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $t = 1, \dots, T$.
-This grid is used as input for the estimated embedding functions $\widehat{f_t}$ at all time points, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point.
-This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\mathbf{r}_i)$ for all $i$ and $t$.
+Further smoothing of the time-specific manifold estimates $\widehat{f_t}$ with respect to time is needed to consider curvature in the time dimension, corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
+As described in Section \ref{s:PME}, the principal manifold $\widehat{f_t}$ at time $t$ has spline form (see equation~\eqref{eq: spline representation of the PME}) and is characterized by spline coefficients $\{s_{t, j, l}\}_{j=1, l=1}^{N, D}$ and $\{\alpha_{t, k, l}\}_{k=1, l=1}^{d+1, D}$.
+To smooth the functions $\widehat{f_t}$ over time $t$, we seek to estimate a spline function mapping from the time dimension to the coefficient space given a tuning parameter $\gamma$ (see the third term in equation~\eqref{eq:newKappa}).
+As detailed by \cite{greenSilverman1994}, smoothing splines are fitted at a collection of knots corresponding to the design points (in this case, the cluster center parameterizations $\{\pi_{\widehat{f_t}}(\mu_{j, t})\}_{j=1, t=1}^{N_t, T}$).
+Because time-specific principal manifold estimates were fit with different cluster centers $\mu_{j, t}$ for each time point $t$, the coefficients of fitted functions at different time points are not comparable.
+Hence, to obtain coefficients that can be compared across time points, we design a grid of knots $\left\{\mathbf{r}_i\right\}_{i=1}^{N} \subset \mathbb{R}^d$ ranging from the minimum to the maximum of the approximated parameters $\{\pi_{\widehat{f_t}}(\mu_{j, t})\}_{j=1}^{N_t}$ in each dimension of $\mathbb{R}^d$.
+Define
+\begin{align*}
+    N:=\max_{t=1,2,\ldots,T} N_t,\ \ \ \text{ and }\ \ \ Y_{i,t}=\widehat{f_t}(\mathbf{r}_i)\ \text{ for all }i=1,2,\ldots,N \text{ and }t=1,2,\ldots,T.
+\end{align*}
 Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing the following function.
 \begin{equation}\label{eq:splineL}
-    \mathcal{L}(f_t) = (\mathbf{Y}_t - \mathbf{E}\mathbf{s}_t - \mathbf{R}^\T\mathbf{\alpha}_t)^\T(\mathbf{Y}_t - \mathbf{E}\mathbf{s}_t - \mathbf{R}^\T\mathbf{\alpha}_t) + \lambda_t^*\mathbf{s}_t^\T \mathbf{E}\mathbf{s}_t,
+    \mathcal{L}(\widehat{f_t}) = (\mathbf{Y}_t - \mathbf{E}\mathbf{s}_t - \mathbf{R}^\T\mathbf{\alpha}_t)^\T(\mathbf{Y}_t - \mathbf{E}\mathbf{s}_t - \mathbf{R}^\T\mathbf{\alpha}_t) + \lambda_t^*\mathbf{s}_t^\T \mathbf{E}\mathbf{s}_t,
 \end{equation}
 where $\mathbf{Y}_t$ is the $N \times D$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\mathbf{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\mathbf{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\mathbf{r}_j)$.
-
-By computing the derivative of $\mathcal{L}(f_t)$ in (\ref{eq:splineL}), we conduct the minimization of $\mathcal{L}(f_t)$ by solving the following system of equations.
+By computing the derivative of $\mathcal{L}(\widehat{f_t})$ in (\ref{eq:splineL}) with respect to the coefficients, we conduct the minimization of $\mathcal{L}(\widehat{f_t})$ by solving the following system of equations.
 \begin{equation}
@@
 \end{equation}
-The solution to Eq.~\eqref{eq:13} , denoted by $\mathbf{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\mathbf{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields the function
+The solution to equation~\eqref{eq:13} , denoted by $\mathbf{\alpha}^*_t=(\alpha^*_{t,1},\ldots,\alpha^*_{t,d+1})^\T$ and $\mathbf{s}^*_t=(s^*_{t,1},\ldots,s^*_{t,N})^\T$, yields the function
 \begin{equation}\nonumber
-  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r).
+  f_{t, l}^*(r) = \sum_{j=1}^{N}s_{t, j, l}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{t, k, l}^* \cdot \phi_k(r).
 \end{equation}
@@
 $$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$ Thus, the weights correspond to the normalized inverse errors of the PME estimates at each time point.
-
 With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{b}_t\right\}_{t=1}^T$ as the response values, with weights $\left\{w_t\right\}_{t=1}^T$ resulting in a function taking the form
@@
 \end{equation}
-where $\mathbf{B}=(\mathbf{b}_1,\ldots, \mathbf{B}_T)^\T$.
+where $\mathbf{B}=(\mathbf{b}_1,\ldots, \mathbf{b}_T)^\T$.
 
@@
 
-We note that, once the spline knots correspond across time points, this function takes the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}.
+Depending on the locations of spline knots  across time points, this function may take the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}.
 However, instead of attempting to estimate each coefficient function individually, we consider all coefficients together as the output of a single smoothing spline function.
@@
 
-Once $F_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
+The optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation.
 In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$.
@@
 The optimal tuning value $\gamma^*$ is chosen as the value of $\gamma$ that minimizes $MSD(\gamma)$.
-Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
+Leave-one-out cross validation was chosen for this application due to the relatively limited number of time points available for individuals in the ADNI dataset.
+In other circumstances with greater data availability, alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.
 
-Further work must be done to demonstrate the convergence properties of the proposed algorithm.
-However, \cite{mengPrincipalManifoldEstimation2021} demonstrated that the function that minimizes Equation \ref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimate in a penalized regression setting (\cite{wahba1990}).
-These results motivate our expectation that the estimate obtained by the LPME algorithm will minimize Equation \ref{eq:newKappa}.
-
-%\meng{Do some simulation studies and compare the leave-one-out CV and k-fold CV.}
+In applying the proposed algorithms in extensive simulations, we did not encounter any issues with convergence of the algorithms.
+Further work must be done to demonstrate that the proposed estimator does indeed minimize the objective function, however, \cite{mengPrincipalManifoldEstimation2021} demonstrated that the function that minimizes equation \eqref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimator in a penalized regression setting (\cite{wahba1990}).
+These results motivate our expectation that the estimator from the LPME algorithm will minimize equation \eqref{eq:newKappa}.
 
@@
 Notably, many manifold learning methods are developed under the assumption that the underlying manifold is not self-intersecting.
-This is not the case for in our motivating problem, where we are interested in modeling the surface of a brain region that is a closed 2-dimensional surface embedded in 3-dimensional space.
+This is not the case in our motivating problem, where we are interested in modeling the surface of a brain region that is a closed 2-dimensional surface embedded in 3-dimensional space, e.g.
+see the images of the hippocampus and thalamus in Figure \ref{fig:adni_result}.
 This may often be the case in other applications where manifold learning is implemented.
 One such simple example is modeling digits where the number 8 for instance is a self-intersecting manifold.
-Methods based on the principal curve framework, including the PME framework, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}.
-A common approach for dealing with this problem is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space (e.g., \cite{mengPrincipalManifoldEstimation2021}; \cite{dunsonInferringManifoldsNoisy2022}).
-To avoid this issue, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
+
+The self-consistency condition used to define principal curves in \cite{hastiePrincipalCurves1989} theoretically precludes the estimation of curves that are self-intersecting.
+However, Hastie and Stuetzle demonstrated that this can be done empirically through the use of periodic smoothers.
+\cite{banfieldIceFloeIdentification1992} generalized the algorithm proposed in \cite{hastiePrincipalCurves1989}, averaging over projection residuals rather than data points, to fit closed curves.
+These approaches do not adapt easily to the higher dimensional penalized regression setting we consider here.
+
+A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space (e.g., \cite{mengPrincipalManifoldEstimation2021}).
+To avoid this process, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
 This approach is motivated by the concept of ``lift" in algebraic topology (\cite{hatcher2002algebraic}, Section 1.1).
@@
   \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
-The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
+The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the polar coordinates results in a manifold that is not self-intersecting.}
   \label{fig:unit_circle_augmentation}
@@
 In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
-The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package (\cite{hastie2022package}), developed using \texttt{R} (\cite{rSoftware2023}).
+The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package (\cite{Cannoodt2018princurve}), developed using \texttt{R} (\cite{rSoftware2023}).
 In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
@@
 
-Images were processed using FSL via the \texttt{fslr}package in \texttt{R}, with FSL's FLIRT method linearly registering the images to MNI space, and the FIRST method being used for image segmentation.
+Images were processed using FSL (\cite{jenkinsonFSL2012}) via the \texttt{fslr} package (\cite{muschelliFslrConnectingFSL2015}) in \texttt{R}, with FSL's FLIRT method linearly registering the images to MNI space, and the FIRST method being used for image segmentation.
 Following image segmentation, the surfaces of each region were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
@@
 
-Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}.
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure \ref{fig:adni_result}.
 The previously described data augmentation approach was used to enable the fitting of closed surfaces.
-Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data.
+Visual inspection of the hippocampus estimates indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data.
 There are two main sources of discrepancies between the data and the LPME estimate.
@@
 
-Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}.
-While the segmentation estimates of the thalamus do not show deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, cross sectional results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}.
+While the segmentation estimates of the thalamus in Figure \ref{fig:adni_result} do not show deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
 
@@
 The code used to implement this approach is available at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
-To align processing steps as closely as possible, the volume estimates shwon for the raw data use the same volume estimation method used to find LPME- and PME-based approximations, rather than the volume estimates produced by FSL's FIRST method during segmentation.
+To align processing steps as closely as possible, the volume estimates shown for the raw data use the same volume estimation method used to find LPME- and PME-based approximations, rather than the volume estimates produced by FSL's FIRST method during segmentation.
 
@@
 In Figure \ref{fig:lhipp_volume_comparison}, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
-All three sources show substantial variability between study visits, particularly for subjects 137\_S\_0283 and 137\_S\_0686.
+All three sources show substantial variability between study visits, particularly for subjects 011\_S\_0003 and 036\_S\_1240.
 Considering the thalamus estimates in Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
@@
   %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
-  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[height=8cm]{adni_plots/adni_cross_section}} \label{fig:lhipp_cross_sections}}%
+  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[height=8cm]{adni_plots/adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
   \vfill
@@
 LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface.
-In the fifth hippocampus observation with $t=3.04$, the LPME-estimated surface has a gap.
+In the fifth hippocampus observation with $t=6.55$, the LPME-estimated surface has a gap.
 Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.
@@
 To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a general-purpose longitudinal setting.
-We also suggest a data augmentation approach that circumvents the inability to fit principal curve-based methods to self-intersecting manifolds in select settings.
+We also suggest a data augmentation approach that circumvents challenges in fitting principal manifold-based methods to self-intersecting manifolds in select settings.
 In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance, in terms of minimizing the MSD, in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve/surface algorithms at each time point.
@@
 The smoothness can be represented by the covariance kernel used in the GPR, e.g.
-by applying the Gaussian or Matérn kernels (\cite{dunsonInferringManifoldsNoisy2022}).
+by applying the Gaussian or Matérn kernels (\cite{li2023inference}).
 Future work may consider implementation of a similar approach in the framework we proposed in this article.
+
+\section*{Acknowledgements}
+
+Research presented in this manuscript was funded by National Institute On Aging of the National Institutes of Health (NIH) under Award Number R01AG075511.
+The ideas presented in this paper are the responsibility of the authors and may not necessarily represent the official views of the NIH.
 
@@
 
-\nocite{*}
+%\nocite{*}
 %\bibliographystyle{plain}
 %\bibliography{references}
-\printbibliography
+%\printbibliography
 
@@
 
+\bibliographystyle{plainnat}
+\bibliography{references}
+
 \end{document}

### 2024-06-25 paper/main.tex

@@
 Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
-As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space.
+As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space.
 However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{agsm}
 \bibliography{references}

### 2024-07-10 paper/main.tex
overleaf updates

@@
 \documentclass[12pt]{article}
+\usepackage{algorithm2e}
+\usepackage{amscd}
+\usepackage{amsfonts}
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage{authblk}
+\usepackage[english]{babel}
+\usepackage{bm}
+\usepackage{caption}
+\usepackage{color}
+%\usepackage{csquotes}
+\usepackage{float}
+\usepackage[perpage, symbol*]{footmisc}
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usepackage{enumerate}
+\usepackage{mathrsfs}
 \usepackage{natbib}
+\usepackage{setspace}
+\usepackage{subcaption}
+\usepackage{tikz}
 \usepackage{url}
 
-%\documentclass[12pt]{amsart}
-%\documentclass[11pt,reqno]{article}
-%\usepackage{cases}
+\newcommand{\blind}{1}
+\graphicspath{{figures/}}
 
-%\RequirePackage[numbers]{natbib}
-%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
-%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
-%\RequirePackage{graphicx}%% uncomment this for including figures
-
-%\usepackage[
-%backend=biber,
-%style=authoryear,
-%url=false,
-%isbn=false,
-%date=year,
-%doi=false
-%]{biblatex}
-
-%\addbibresource{references.bib}
-
-\usepackage{setspace}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amscd}
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\usepackage{mathrsfs}
-\usepackage{graphicx}
-\graphicspath{{figures/}}
-\usepackage[perpage,symbol*]{footmisc}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{color}
-\usepackage{tikz}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{bm}
-
-\usepackage{algorithm2e}
-
-%\usepackage{natbib}
-%\bibliographystyle{abbrvnat}
-%\setcitestyle{authoryear,open={(},close={)}}
-\usepackage{authblk}
-
-\usepackage{csquotes}
-\usepackage[english]{babel}
-
-\renewcommand{\baselinestretch}{1.0}
-\setlength{\oddsidemargin}{-0.5cm}
-\setlength{\evensidemargin}{-0.5cm}
-\renewcommand{\topmargin}{-2cm}
-\renewcommand{\oddsidemargin}{0mm}
-\renewcommand{\evensidemargin}{0mm}
-\renewcommand{\textwidth}{180mm}
-\renewcommand{\textheight}{240mm}
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-1in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.7in}%
+\addtolength{\topmargin}{-1in}%
 
@@
 \newcommand{\T}{\intercal}
-\newcommand{\commentout}[1]{}
-
-\newcommand{\kmedit}[1]{{\color{purple}  #1}}
-\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
-\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}
-
-\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}
 
@@
 
+\setcitestyle{aysep={}}
+
 \begin{document}
 
-\title{Longitudinal Principal Manifold Estimation}
-\author[1]{Robert Zielinski}
-\author[2]{Kun Meng}
-\author[1]{Ani Eloyan}
-\author[*]{for the Alzheimer’s Disease Neuroimaging Initiative}
-\affil[1]{Department of Biostatistics, Brown University}
-\affil[2]{Division of Applied Mathematics, Brown University}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if1\blind
+{
+  \title{\bf Longitudinal Principal Manifold Estimation}
+  \author{
+    Robert Zielinski\hspace{.2cm}\\
+    Department of Biostatistics, Brown University\\
+    Kun Meng \\
+    Division of Applied Mathematics, Brown University\\
+    and \\
+    Ani Eloyan \\
+    Department of Biostatistics, Brown University\\
+    for the Alzheimer's Disease Neuroimaging Initiative}
   \maketitle
+} \fi
 
-\doublespacing
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Longitudinal Principal Manifold Estimation}
+\end{center}
+  \medskip
+} \fi
 
-\section*{Abstract}
-
+\bigskip
+\begin{abstract}
 Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease.
@@
 An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
-\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
+%\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
+\end{abstract}
 
-% Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain.
-Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy.
-Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently.
-As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials.
-To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time.
-The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point.
-This approach mitigates the effects of random disturbances to the manifold between time points.
-In addition, we propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds.
-We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
-These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points.
-We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.
-We present a comprehensive analysis of magnetic resonance imaging data from 263 participants demonstrating high levels of variability of regional volumes when using existing methods for segmentation of hippocampuses and thalamnuses and the advantages gained by using our proposed methods that leverage longitudinal regional trends for segmentation.
-\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
+\noindent%
+{\it Keywords:}  Alzheimer's disease, smoothing spline, magnetic resonance imaging
+\vfill
 
+\newpage
+\spacingset{1.9}
 \section{Introduction}
 
-Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}).
+%Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}.
 Frequently, interest centers around longitudinal changes in one or more neurological substructures.
-These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}).
+These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}.
 For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes.
@@
 When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
-Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}).
-Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}).
+Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability \citep{boccardiSurveyProtocolsManual2011}.
+Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}.
 While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
 
-\subsection{A Motivating Question from Medical Imaging}
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}.
+Frequently, interest centers around longitudinal changes in one or more neurological substructures.
+These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}.
+For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes.
+Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
+Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
+However, this approach is highly time and resource intensive.
+When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
+Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}.
+While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
 
-Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+%\subsection{A Motivating Question from Medical Imaging}
+
+%Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
 In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
@@
 If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
-\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models.
-In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).
-To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
+%\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models.
+%In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging \citep{scahill2003longitudinal}.
+%To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
 A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.
 
+Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
+In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
+If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
 Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods.
@@
 
-\subsection{A Brief Overview of Existing Approaches and Our Contributions}
+%\subsection{Background and Our Contributions}
 
-Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}).
-Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}), single-cell biology (\cite{ding2023learning}), and robotics (\cite{gao2023k}; \cite{gao2024bi}).
+%Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}.
+Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing \citep{shen2022robust}, single-cell biology \citep{ding2023learning}, and robotics \citep{gao2024bi}.
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
@@
 
+Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}.
+Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}.
+In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
+For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.
+Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.
 Nonlinear manifold learning methods have been extensively studied in the literature.
-However, many of them are not applicable in longitudinal studies, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}).
-That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of the time dimension.
-We use the oldest linear manifold learning method, principal component analysis (PCA, \cite{Pearson1901on}), as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.
+However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies.
+That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.
+%We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.
 
@@
 However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
-In addition, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
+Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
 However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+
+%In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time.
+Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension.
+Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
+Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}).
+Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit.
+We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+We demonstrate that the regularization involved in this process mitigates the effects of variability between time points.
+Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
 
@@
 Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
-Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation~\eqref{eq: spline representation of the PME}, in Section \ref{s:PME}).
-Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit.
-We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
+Unlike most popular manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}).
+We build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
 We demonstrate that the regularization involved in this process mitigates the effects of variability between time points.
@@
 This work was motivated by longitudinal imaging studies, where data of participants are collected during several study visits to observe changes in brain structure (among other variables of interest) over time.
-One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this article.
+One such study is the ADNI, where we obtained data that we analyzed to reach the results presented in this article.
 The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}.
@@
 The remainder of this article is laid out as follows.
-In Section \ref{s:PME}, we review the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}.
-In Section \ref{s:LPME}, we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show our proposed novel  algorithm for dimension reduction using the longitudinal principal manifold framework.
+Section \ref{s:PME} reviews the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}.
+In Section \ref{s:LPME}, we adapt the principal manifold framework for longitudinal settings, and show our proposed novel algorithm for dimension reduction using the longitudinal principal manifold framework.
 Section \ref{s:simulations} demonstrates the performance of this approach on simulated data.
-In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in individuals with AD and cognitively normal adults.
+In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in ADNI participants.
 The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.
@@
 
-The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud.
+%The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud.
 However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings.
@@
 The principal manifolds are then defined as follows.
+
+The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud.
+Motivated by the penalization approach presented by \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions.
+In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper.
+We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively.
+We assume that a data cloud is observed in the $D$ dimensional space.
+In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure 1 in the online supplement.
+For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
+Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
+Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$.
+The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
+Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
+The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection.
+The principal manifolds are defined as follows.
 \begin{definition}
@@
   The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
-  \begin{equation}\nonumber
-    \begin{aligned}
-        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f),
-    \end{aligned}
-  \end{equation}
+  $f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)$,
 where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
@@
 
-The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements.
+The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation \eqref{eq:pme_kappa} consists of two elements.
 The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$.
@@
 In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
-Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem making this a supervised learning problem.
+Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem, making this a supervised learning problem.
 \cite{mengPrincipalManifoldEstimation2021} also show that this function takes spline form
@@
 under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud.
-Importantly, $N$ tends to be much smaller than the number of points in the data cloud (see Figure 3 of \cite{mengPrincipalManifoldEstimation2021}), which indicates that the representation in equation~\eqref{eq: spline representation of the PME} is computationally efficient.
+Importantly, $N$ tends to be much smaller than the number of points in the data cloud \citep[fig.
+3]{mengPrincipalManifoldEstimation2021}, which indicates that the representation in equation \eqref{eq: spline representation of the PME} is computationally efficient.
 This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$.
@@
 
-\section{Longitudinal Principal Manifold Estimation Algorithm}\label{s:LPME}
+\section{Longitudinal Principal Manifold Estimation}\label{s:LPME}
 
@@
 The goal in this simplified example is to fit a surface in 3-dimensional spacetime---comprising two spatial dimensions and one time dimension---that fits the longitudinal data $(x_1, x_2, t_j)$ in spacetime and is not affected by the random noise added at each time point and in space (see Figure \ref{fig:lpme_step4}).
-The PME framework does not directly apply due to the following:
-\begin{enumerate}
-    \item Suppose we apply the PME approach to the data $(x_1, x_2, t_j)$ in spacetime and obtain an estimated surface $(r_1, r_2) \mapsto F(r_1, r_2)\in\mathbb{R}^2\times\mathbb{R}$ in spacetime.
+The PME framework does not directly apply due to the following: (1) Suppose we apply the PME approach to the data $(x_1, x_2, t_j)$ in spacetime and obtain an estimated surface $(r_1, r_2) \mapsto F(r_1, r_2)\in\mathbb{R}^2\times\mathbb{R}$ in spacetime.
 It is likely that neither coordinate $r_1$ nor $r_2$ coincides with the time dimension---they are (potentially nonlinear) combinations of both time and spatial coordinates, which prevents meaningful interpretation of the time dimension.
-
-    \item If the manifold is estimated separately at each time point, the results may not be smooth over time.
-\end{enumerate}
+(2) If the manifold is estimated separately at each time point, the results may not be smooth over time.
 We expect that our proposed dimension reduction method will yield a smooth surface $(t,r)\mapsto F(t,r)$ in spacetime, $\mathbb{R}^2\times\mathbb{R}$, while ensuring that one coordinate, $t$, coincides with the time dimension.
@@
 As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
-In this setting, the manifold to estimate is represented by
+In this setting, the manifold to estimate is represented by the following, where $t$ coincides with the time dimension
 \begin{align}\label{eq: F form}
     \begin{aligned}
-        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R}, \\
-    & (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
+        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R},\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
     \end{aligned}
 \end{align}
-where $t$ coincides with the time dimension. We define a longitudinal principal manifold as follows.
+We define a longitudinal principal manifold as follows.
 \begin{definition}
@@
 \begin{align}\label{eq:newKappa}
-  \begin{aligned}
-      \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
+      \notag\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
   &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt,
-  \end{aligned}
 \end{align}
-where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation~\eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
+\sloppy where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation \eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u: \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$.
 Then, the manifold $M_{F^{*}}^{d+1}:=\{F^*(t,\mathbf{r}): (t,\mathbf{r})\in \mathbb{R}^d\times\mathbb{R}\}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
@@
 Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing).
-Specifically, at each individual time point, we apply the PME algorithm (\cite{mengPrincipalManifoldEstimation2021}, Algorithm 2) to fit the data cloud at the time point and represent the fitted manifold using the spline form in equation~\eqref{eq: spline representation of the PME}.
+Specifically, at each individual time point, we apply the PME algorithm \citep[alg.
+2]{mengPrincipalManifoldEstimation2021} to fit the data cloud at the time point and represent the fitted manifold using the spline form in equation~\eqref{eq: spline representation of the PME}.
 Hence, we have a collection of spline coefficients, i.e., the $s_{j,l}$ and $\alpha_{k,l}$ in equation~\eqref{eq: spline representation of the PME}, associated with each time point.
@@
 A visualization of these steps is shown in Figure \ref{fig:lpme_steps}.
-Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}.
+Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}\if1\blind{, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}}\fi.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}
   \centering
-  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}%
+  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=7cm]{sim_case1_plot2}}%
   \hfill
-  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}
+  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=7cm]{lpme_data_reduction}}
   \vfill
-  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}
+  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=7cm]{lpme_initialization}}
   \hfill
-  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}
-  \caption{LPME algorithm steps, demonstrated using longitudinal simulated data with $D=2$.
+  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=7cm]{lpme_fitting}}
+  \caption{{\footnotesize LPME algorithm steps, demonstrated using longitudinal simulated data with $D=2$.
 Data from the functions $x_1 = r$ and $x_2 = \alpha \sin \left(\beta r + \frac{\pi}{2}\right)$ are generated at 5 time points.
 Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually.
-In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.
-}
+In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.} }
   \label{fig:lpme_steps}
@@
 Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is large.
-To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method (see \cite{mengPrincipalManifoldEstimation2021}, Algorithm 1) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method \citep[see][alg.
+1]{mengPrincipalManifoldEstimation2021} to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
 Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
 The number $N_t$ of clusters is estimated for each time point $t$ and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
-By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}).
+By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ \citep[see the simulation study presented in][fig.
+3(c)]{mengPrincipalManifoldEstimation2021}.
 The centers $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$ estimated in this step are then used to reach an initial low-dimensional parameterization.
@@
 The next step in the LPME algorithm for the estimation of the longitudinal principal manifold is initialization.
-The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve.
+The PME algorithm uses Isomap \citep{tenenbaumGlobalGeometricFramework2000} to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points.
@@
 For example, the following two drastically different expressions parameterize the same curve: $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$ and $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$.
-If Isomap is applied at each time point separately, it tends to yield inconsistent parameterizations between time points.
-Therefore, a different approach to generating an initial parameterization is needed for the LPME algorithm.
+Because Isomap tends to yield inconsistent parameterizations when applied to each time point separately, a different approach is needed for the LPME algorithm.
 
@@
 This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using multicolored lines in Figure \ref{fig:lpme_step3}.
-The performance of each fit $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I_t}\sum_{i=1}^{I_t}\| x_{it} - \widehat{f_t}(\pi_{\widehat{f_t}}(x_{it}))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the manifold defined by $\widehat{f_t}$.
+The performance of each fit $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I_t}\sum_{i=1}^{I_t}\| x_{it} - \widehat{f_t}(\pi_{\widehat{f_t}}(x_{it}))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the fitted manifold $\widehat{f_t}$.
 
 Further smoothing of the time-specific manifold estimates $\widehat{f_t}$ with respect to time is needed to consider curvature in the time dimension, corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
-As described in Section \ref{s:PME}, the principal manifold $\widehat{f_t}$ at time $t$ has spline form (see equation~\eqref{eq: spline representation of the PME}) and is characterized by spline coefficients $\{s_{t, j, l}\}_{j=1, l=1}^{N, D}$ and $\{\alpha_{t, k, l}\}_{k=1, l=1}^{d+1, D}$.
-To smooth the functions $\widehat{f_t}$ over time $t$, we seek to estimate a spline function mapping from the time dimension to the coefficient space given a tuning parameter $\gamma$ (see the third term in equation~\eqref{eq:newKappa}).
+As described in Section \ref{s:PME}, the principal manifold $\widehat{f_t}$ at time $t$ has spline form (equation~\eqref{eq: spline representation of the PME}) and is characterized by spline coefficients $\{s_{t, j, l}\}_{j=1, l=1}^{N, D}$ and $\{\alpha_{t, k, l}\}_{k=1, l=1}^{d+1, D}$.
+To smooth the functions $\widehat{f_t}$ over time $t$, we seek to estimate a spline function mapping from the time dimension to the coefficient space given a tuning parameter $\gamma$.
 As detailed by \cite{greenSilverman1994}, smoothing splines are fitted at a collection of knots corresponding to the design points (in this case, the cluster center parameterizations $\{\pi_{\widehat{f_t}}(\mu_{j, t})\}_{j=1, t=1}^{N_t, T}$).
@@
 Hence, to obtain coefficients that can be compared across time points, we design a grid of knots $\left\{\mathbf{r}_i\right\}_{i=1}^{N} \subset \mathbb{R}^d$ ranging from the minimum to the maximum of the approximated parameters $\{\pi_{\widehat{f_t}}(\mu_{j, t})\}_{j=1}^{N_t}$ in each dimension of $\mathbb{R}^d$.
-Define
-\begin{align*}
-    N:=\max_{t=1,2,\ldots,T} N_t,\ \ \ \text{ and }\ \ \ Y_{i,t}=\widehat{f_t}(\mathbf{r}_i)\ \text{ for all }i=1,2,\ldots,N \text{ and }t=1,2,\ldots,T.
-\end{align*}
+Define $N:=\max_{t=1,2,\ldots,T} N_t$ and $Y_{i,t}=\widehat{f_t}(\mathbf{r}_i)$ for all $i=1,2,\ldots,N$ and $t=1,2,\ldots,T$.
 Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing the following function.
@@
 \end{equation}
-where $\mathbf{Y}_t$ is the $N \times D$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\mathbf{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\mathbf{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\mathbf{r}_j)$.
+where $\mathbf{Y}_t$ is the $N \times D$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$; the matrix $\mathbf{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$; $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even and $\eta_d(r) = r^{4-d}$ if $d$ is odd; the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\mathbf{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\mathbf{r}_j)$.
 By computing the derivative of $\mathcal{L}(\widehat{f_t})$ in (\ref{eq:splineL}) with respect to the coefficients, we conduct the minimization of $\mathcal{L}(\widehat{f_t})$ by solving the following system of equations.
@@
 The solution to equation~\eqref{eq:13} , denoted by $\mathbf{\alpha}^*_t=(\alpha^*_{t,1},\ldots,\alpha^*_{t,d+1})^\T$ and $\mathbf{s}^*_t=(s^*_{t,1},\ldots,s^*_{t,N})^\T$, yields the function
-\begin{equation}\nonumber
-  f_{t, l}^*(r) = \sum_{j=1}^{N}s_{t, j, l}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{t, k, l}^* \cdot \phi_k(r).
-\end{equation}
+$f_{t, l}^*(r) = \sum_{j=1}^{N}s_{t, j, l}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{t, k, l}^* \cdot \phi_k(r).$
 
 We denote the manifold coefficients at each time point $t$ as $\mathbf{b}_t = (\mathbf{s}_t^{*\T}, \mathbf{\alpha}_t^{*\T})^\T$.
-To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients $\mathbf{b}_t$, where the weights for each time point $t$ are equal to
-$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$ Thus, the weights correspond to the normalized inverse errors of the PME estimates at each time point.
+To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients $\mathbf{b}_t$, where the weights for each time point $t$ are equal to $w_t = 1 / \left(\tau_t\sum_{i=1}^T(1/\tau_i)\right)$.
+Thus, the weights correspond to the normalized inverse errors of the PME estimates at each time point.
 With a given tuning parameter $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{b}_t\right\}_{t=1}^T$ as the response values, with weights $\left\{w_t\right\}_{t=1}^T$ resulting in a function taking the form
-\begin{equation}\nonumber
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j,
-\end{equation}
+$g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j$,
 where by defining the $T \times T$ matrix $\mathbf{A}$ as  $\mathbf{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\mathbf{T}$ as  $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = \operatorname{diag}(w_1, \dots, w_T)$, we obtain the coefficients $\mathbf{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\mathbf{\nu}=(\nu_1,\nu_2)^\T$ as the solutions to
@@
 where $\mathbf{B}=(\mathbf{b}_1,\ldots, \mathbf{b}_T)^\T$.
-
 For a given $\gamma$, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $\mathbf{B}_{\gamma}(t) = \left(\mathbf{s}_{\gamma}(t)^\T, \mathbf{\alpha}_{\gamma}(t)^\T\right)^\T$.
@@
 
-Depending on the locations of spline knots  across time points, this function may take the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}.
+%Depending on the locations of spline knots  across time points, this function may take the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}.
 However, instead of attempting to estimate each coefficient function individually, we consider all coefficients together as the output of a single smoothing spline function.
@@
 In applying the proposed algorithms in extensive simulations, we did not encounter any issues with convergence of the algorithms.
-Further work must be done to demonstrate that the proposed estimator does indeed minimize the objective function, however, \cite{mengPrincipalManifoldEstimation2021} demonstrated that the function that minimizes equation \eqref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimator in a penalized regression setting (\cite{wahba1990}).
+Further work must be done to demonstrate that the proposed estimator does indeed minimize the objective function, however \cite{mengPrincipalManifoldEstimation2021} demonstrated that the function that minimizes equation \eqref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimator in a penalized regression setting \citep{wahba1990}.
 These results motivate our expectation that the estimator from the LPME algorithm will minimize equation \eqref{eq:newKappa}.
@@
 Notably, many manifold learning methods are developed under the assumption that the underlying manifold is not self-intersecting.
-This is not the case in our motivating problem, where we are interested in modeling the surface of a brain region that is a closed 2-dimensional surface embedded in 3-dimensional space, e.g.
-see the images of the hippocampus and thalamus in Figure \ref{fig:adni_result}.
-This may often be the case in other applications where manifold learning is implemented.
-One such simple example is modeling digits where the number 8 for instance is a self-intersecting manifold.
+This is not the case in our motivating problem, where we are interested in modeling the surface of a brain region that is a closed 2-dimensional surface embedded in 3-dimensional space.
+Self-intersecting manifolds may appear in other applications where manifold learning is implemented.
+One such simple example is modeling digits, where the number 8, for instance, is a self-intersecting manifold.
 
@@
 
-A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space (e.g., \cite{mengPrincipalManifoldEstimation2021}).
+A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{mengPrincipalManifoldEstimation2021}.
 To avoid this process, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
-This approach is motivated by the concept of ``lift" in algebraic topology (\cite{hatcher2002algebraic}, Section 1.1).
+This approach is motivated by the concept of ``lift" in algebraic topology \citep[sec.
+1.1]{hatcher2002algebraic}.
 As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions.
@@
 
-\begin{figure}[ht]
+\begin{figure}
   \centering
-  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
+  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=7cm]{unit_circle_D2}}}%
   \hfill
-  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
-  \caption{Illustration of our proposed data augmentation approach using the Unit Circle.
-The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the polar coordinates results in a manifold that is not self-intersecting.}
+  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=7cm]{unit_circle_D3}}}
+  \caption{{\footnotesize Illustration of our proposed data augmentation approach using the Unit Circle.
+The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the polar coordinates results in a manifold that is not self-intersecting.}}
   \label{fig:unit_circle_augmentation}
@@
 Because the added dimensions duplicate information contained within the unaugmented observations, these dimensions can be discarded to reach estimates in the original $D$-dimensional space.
-This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition.
+This approach to data augmentation offers an alternative to manual partitioning used by \cite{mengPrincipalManifoldEstimation2021}.
 Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
@@
 The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
-In order to put these results in context of previous published work in this area, we use true underlying functions that are similar to those used in previous literature and build on these existing results.
 Based on our observations of systematic noise sources in imaging data, we add various types of terms to these functions at each visit.
 In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between visits and represent structural changes in the underlying manifold over time.
-The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
+The function $g(\cdot)$ may show change with respect to time that is constant (no change), linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
 The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between visits.
@@
 \begin{table}[ht]
-\small
   \centering
-  \begin{tabular}{|c c c c c|}
+  \resizebox{\textwidth}{!}{\begin{tabular}{|c c c c c|}
     \hline
@@
     \hline
-  \end{tabular}
-  \caption{Embedding functions used for simulation studies.
+  \end{tabular}}
+  \caption{{\footnotesize Embedding functions used for simulation studies.
 Function $\bf{g}(t)$ denotes structural change in the underlying manifold over time, while $\zeta$ represents the scale of this change.
-Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}
+Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}}
   \label{table:simulation_embeddings}
@@
 In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
-The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package (\cite{Cannoodt2018princurve}), developed using \texttt{R} (\cite{rSoftware2023}).
+The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package \citep{Cannoodt2018princurve}, developed using \texttt{R} \citep{rSoftware2023}.
 In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
@@
 Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 7,776 times.
-All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
+\if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.}\fi
 
-Visualizations of the results from some example simulated cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case7}.
+Visualizations of the results from one example simulated case, truncated for concision, are shown in Figure \ref{fig:sim_case1}.
 Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
@@
   \end{tabular}
-  \caption{Mean Squared Distance comparison of simulated data, longitudinal principal manifold estimation (LPME)-, principal manifold estimation (PME)-, and principal curve / surface-based estimates to true values, Median (IQR).
-The lowest algorithm-specific median (IQR) are highlighted in bold.}
+  \caption{{\footnotesize Mean Squared Distance comparison of simulated data, longitudinal principal manifold estimation (LPME)-, principal manifold estimation (PME)-, and principal curve / surface-based estimates to true values, Median (IQR).
+The lowest algorithm-specific median (IQR) are highlighted in bold.}}
   \label{table:simulation_results_median}
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
-  \subfloat[\centering Simulation Case 1]{{\includegraphics[height=8cm]{sim_case1}} \label{fig:sim_case1}}%
-  \vfill
-  \subfloat[\centering Simulation Case 7]{{\includegraphics[height=8cm]{sim_case7}} \label{fig:sim_case7}}
-
-  \caption{Scatterplots of two example simulated data clouds (gray points) with $d=1$ and $d=2$ overlaid with points on the true embedding manifold (red), the estimated LMPE function (blue), PME (green), and principal curve or surface (purple).}
-  \label{fig:sim_results}
+  \includegraphics[height=9cm]{sim_case1} 
+  \caption{{\footnotesize Simulation Case 1.
+Scatterplots of one example simulated data cloud (gray points) with $d=1$ overlaid with points on the true embedding manifold (red), the estimated LMPE function (blue), PME (green), and principal curve or surface (purple).}}
+  \label{fig:sim_case1}
 \end{figure}
@@
 The median and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_median}.
-Analogous mean and standard deviation summaries of the results are given in the Appendix.
+Analogous mean and standard deviation summaries of the results are given in the online supplement.
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
@@
 
-\section{Longitudinal Segmentation of MRI images in Alzheimer's Disease Studies}\label{s:application}
+\section{Longitudinal Segmentation of MRI Images}\label{s:application}
 
@@
 
-While the original ADNI 1 cohort included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 463 participants, of whom 130 were cognitively normal, 223 displayed symptoms of mild cognitive impairment, and 101 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}).
+While the original ADNI 1 cohort included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 463 participants, of whom 130 were cognitively normal, 223 displayed symptoms of mild cognitive impairment, and 101 were diagnosed with AD at the baseline study visit.
 Diagnostic information was unavailable for 9 of the 463 participants for whom imaging data was available.
@@
 
-Images were processed using FSL (\cite{jenkinsonFSL2012}) via the \texttt{fslr} package (\cite{muschelliFslrConnectingFSL2015}) in \texttt{R}, with FSL's FLIRT method linearly registering the images to MNI space, and the FIRST method being used for image segmentation.
+Images were processed using FSL via the \texttt{fslr} package \citep{muschelliFslrConnectingFSL2015} in \texttt{R}, with FSL's FLIRT method linearly registering the images to MNI space, and the FIRST method being used for image segmentation.
 Following image segmentation, the surfaces of each region were identified by finding the extreme voxels in each dimension with nonzero intensity readings.
 The estimated surface positions were then standardized to a maximum distance of one from the origin in each dimension of Euclidean space and centered around the origin.
-Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.
+Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{ss:selfInt} to avoid the need to fit to a self-intersecting manifold.
 
-Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure \ref{fig:adni_result}.
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure 1 in the online supplement.
 The previously described data augmentation approach was used to enable the fitting of closed surfaces.
@@
 There are two main sources of discrepancies between the data and the LPME estimate.
-First, at time points where the observed surface changes orientation, as is visible between the first two time points, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
+First, at time points where the observed surface changes orientation, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points.
 The second difference between the LPME estimates and the observed data is seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of curves in the observed data.
@@
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
-This Figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each plot.
-Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself.
+This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curve in the structure at the upper-left corner of each plot.
+Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface.
 This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.
-Notably, it also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with the highest levels of curvature.
+It also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with high levels of curvature.
+
+\begin{figure}
+  \centering
+  %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
+  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_plots/adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
+  \hfill
+  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_plots/adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}
+
+  \caption{{\footnotesize Left Hippocampus and Left Thalamus Cross Sections.
+LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface.
+In the fifth hippocampus observation with $t=6.55$, the LPME-estimated surface has a gap.
+Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.
+The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}}
+  \label{fig:adni_cross_sections}
+\end{figure}
 
@@
 Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.
+Considering the same individual shown in Figure \ref{fig:lhipp_cross_sections}, cross sectional results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}.
+The cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
 
-Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, cross sectional results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}.
-While the segmentation estimates of the thalamus in Figure \ref{fig:adni_result} do not show deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+To understand how the subcortical structure estimates obtained by the LPME and PME algorithms impact the corresponding volume estimates for these structures, the volume of each structure was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
+The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
+\if1\blind{The code used to implement this approach is available at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.}\fi To align processing steps as closely as possible, the volume estimates shown for the raw data use the same volume estimation method used to find LPME- and PME-based approximations, rather than the volume estimates produced by FSL's FIRST method during segmentation.
 
-To understand how the subcortical structure estimates obtained by the LPME and PME algorithms may impact the corresponding estimates of the volumes of these structures, the volume of each structure was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point.
-The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.
-The code used to implement this approach is available at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.
-To align processing steps as closely as possible, the volume estimates shown for the raw data use the same volume estimation method used to find LPME- and PME-based approximations, rather than the volume estimates produced by FSL's FIRST method during segmentation.
-
-The volume estimates for randomly selected individuals can be found in Figures \ref{fig:lhipp_volume_comparison} and \ref{fig:lthal_volume_comparison}, corresponding to estimates for the left hippocampus and left thalamus, respectively.
+The volume estimates for randomly selected individuals can be found in Figures 2a and 2b in the online supplement, corresponding to estimates for the left hippocampus and left thalamus, respectively.
 Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
-In Figure \ref{fig:lhipp_volume_comparison}, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
-All three sources show substantial variability between study visits, particularly for subjects 011\_S\_0003 and 036\_S\_1240.
-Considering the thalamus estimates in Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
+In the supplement's Figure 2a, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
+All three sources show substantial variability between study visits, particularly for subjects \texttt{011\_S\_0003} and \texttt{036\_S\_1240}.
+Considering the thalamus estimates in the supplement's Figure 2b, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
 However, the LPME estimates demonstrate the ability to smooth over this variability, resulting in volume estimates that are not as sensitive to differences between time points.
@@
 However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.
-
-\begin{figure}[h]
-  \centering
-  %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
-  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[height=8cm]{adni_plots/adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
-  \vfill
-  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[height=8cm]{adni_plots/adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}
-
-  \caption{Left Hippocampus and Left Thalamus Cross Sections.
-LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface.
-In the fifth hippocampus observation with $t=6.55$, the LPME-estimated surface has a gap.
-Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.
-The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
-  \label{fig:adni_cross_sections}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\subfloat{
-\label{fig:adni_lhipp_result}
-\subfloat[\centering Data]{
-\includegraphics[height=10cm]{adni_plots/adni_lhipp_data_plot}
-}
-\hfill
-\subfloat[\centering LPME]{
-\includegraphics[height=10cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
-}
-\hfill
-\subfloat[\centering PME]{
-\includegraphics[height=10cm]{adni_plots/adni_lhipp_pme_plot}
-}
-}
-\hfill
-\subfloat{
-\label{fig:adni_lthal_result}
-\subfloat[\centering Data]{
-\includegraphics[height=10cm]{adni_plots/adni_lthal_data_plot}
-}
-\hfill
-\subfloat[\centering LPME]{
-\includegraphics[height=10cm]{adni_plots/adni_lthal_lpme_isomap_plot}
-}
-\hfill
-\subfloat[\centering PME]{
-\includegraphics[height=10cm]{adni_plots/adni_lthal_pme_plot}
-}
-}
-\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant.
-The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
-The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
-\label{fig:adni_result}
-\end{figure}
 
@@
 
-\begin{figure}[h]
-    \centering
-    \subfloat[\centering Left Hippocampus Volume Estimates]{
-      \includegraphics[height=8cm]{adni_plots/adni_lhipp_volume_comp}
-      \label{fig:lhipp_volume_comparison}
-    }
-    \vfill
-    \subfloat[\centering Left Thalamus Volume Estimates]{
-      \includegraphics[height=8cm]{adni_plots/adni_lthal_volume_comp}
-      \label{fig:lthal_volume_comparison}
-    }
-    \caption{Left hippocampus and left thalamus volume estimates for three ADNI participants.
-Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (red, circles), LPME (blue, squares), and PME (green, triangles).
-When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of the volume estimates.
-Gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data.
-These gaps may also cause inconsistency in the LPME volume estimates, resulting in higher aggregate variability than for the data or PME estimates.
-When applied to the thalamus, the PME-based volume estimates demonstrate similar time point-to-time point changes in the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data.
-The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 011\_S\_0003.}
-\end{figure}
-
 \begin{table}[ht]
@@
   \end{tabular}
-  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
-Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.
-}
+  \caption{{\footnotesize Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
+Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.} }
   \label{table:adni_volume_sds}
@@
 We also suggest a data augmentation approach that circumvents challenges in fitting principal manifold-based methods to self-intersecting manifolds in select settings.
-In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance, in terms of minimizing the MSD, in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve/surface algorithms at each time point.
-This improvement is observed across several manifold types.
+In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve/surface algorithms at each time point.
 
-Similar to other applications of statistical smoothing techniques, it is difficult to compare performance of models in real data experiments.
+%Similar to other applications of statistical smoothing techniques, it is difficult to compare performance of models in real data experiments.
 Specifically, the MSD provides a summary of the distance of the data points and their projections onto the estimated manifolds for each model.
@@
 In practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
-In such cases, we recommend the use of alternative methods, such as Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}) or Laplacian Eigenmaps (\cite{belkin2003laplacian}), to generate the initial parameterization to alleviate this concern.
-Opportunities also exist to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise.
+In such cases, the use of alternative methods, such as Locally Linear Embedding \citep{roweisNonlinearDimensionalityReduction2000} or Laplacian Eigenmaps \citep{belkin2003laplacian}, to generate the initial parameterization may alleviate this concern.
 A second assumption is that the LPME method requires relative stability in the overall shape of the underlying manifold structure.
@@
 There are two potential sources of this difficulty, each suggesting different steps that may be taken to address this issue.
-The manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better estimate the sharp curvature of the hippocampus.
-Also, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated.
+Replacing the simple grid search that is used to select the optimal value of the smoothing parameters in the PME and LPME algorithms with an improved search method could yield results that are more sensitive to the nuances of the structure being estimated, such as the areas of sharp curvature in the hippocampus.
 
-The second potential interesting direction of research in the future is more in-depth investigation of the data augmentation approach  described in Section \ref{s:LPME}, that allows us to fit some self-intersecting manifolds.
-The proposed augmentation with polar or spherical coordinates, allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
+Second, further investigation of the  data augmentation approach described in Section \ref{ss:selfInt} may enable fitting the PME or LPME algorithms to a more extensive set of structures.
+The proposed augmentation with polar or spherical coordinates allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods.
 However, while this approach addressed a problem posed by structures with roughly circular or spherical shapes, such as the thalamus, its success may not extend to settings with more complex data.
@@
 Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.
-In population-level modeling of the data, it will be important to also consider registration of the participant-level MRI data to a common template space, raising important considerations for alignment of the shapes without losing important information.
 
@@
 The smoothness can be represented by the covariance kernel used in the GPR, e.g.
-by applying the Gaussian or Matérn kernels (\cite{li2023inference}).
+by applying the Gaussian or Matérn kernels \citep{li2023inference}.
 Future work may consider implementation of a similar approach in the framework we proposed in this article.
-
-\section*{Acknowledgements}
-
-Research presented in this manuscript was funded by National Institute On Aging of the National Institutes of Health (NIH) under Award Number R01AG075511.
-The ideas presented in this paper are the responsibility of the authors and may not necessarily represent the official views of the NIH.
-
-\newpage
 
@@
 
-\newpage
+%\if1\blind
+%{
+%\section*{Acknowledgment}
 
-\appendix
+%\section*{Funding}
+%Research presented in this manuscript was funded by National Institute On Aging of the National Institutes of Health (NIH) under Award Number R01AG075511.
+The ideas presented in this paper are the responsibility of the authors and may not necessarily represent the official views of the NIH.
+%} \fi
 
-\section*{Appendix}
-
-\begin{table}[h]
-  \centering
-  \begin{tabular}{|c c c c c|}
-    \hline
-    Case & Data & LPME & PME & PC/PS \\
-    \hline
-    1 & 0.223 (0.256) & {\bf 0.125 (0.161)} & 0.268 (0.850) & 0.189 (0.245) \\
-    2 & 0.514 (0.408) & {\bf 0.384 (0.648)} & 0.843 (1.93) & 0.600 (0.296) \\
-    3 & 0.446 (0.445) & {\bf 0.401 (0.446)} & 0.507 (0.594) & 0.412 (0.423) \\
-    4 & 30.7 (88.1) & {\bf 27.7 (260)} & 30.7 (88.2) & 30.6 (88.1) \\
-    5 & 0.980 (0.771) & {\bf 0.791 (0.845)} & 1.04 (1.07) & 0.934 (0.713) \\
-    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & {\bf 1.01 (2.11)} \\
-    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & {\bf 1.95 (0.800)} \\
-    8 & 0.226 (0.275) & {\bf 0.136 (0.169)} & 0.242 (0.311) & 0.274 (0.243) \\
-    \hline
-  \end{tabular}
-  \caption{MSD comparison to true values, Mean (SD). For each case, the lowest algorithm-specific mean (SD) are highlighted in bold. }
-  \label{table:simulation_results_mean}
-\end{table}
-
-\bibliographystyle{agsm}
+%\newpage
+\bibliographystyle{jasa}
 \bibliography{references}

### 2024-07-10 paper/supplement.tex
overleaf updates

\documentclass[12pt]{article}

\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
%\usepackage{csquotes}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

\graphicspath{{figures/}}

\begin{document}

\begin{table}[h]
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Case & Data & LPME & PME & PC/PS \\
    \hline
    1 & 0.223 (0.256) & {\bf 0.125 (0.161)} & 0.268 (0.850) & 0.189 (0.245) \\
    2 & 0.514 (0.408) & {\bf 0.384 (0.648)} & 0.843 (1.93) & 0.600 (0.296) \\
    3 & 0.446 (0.445) & {\bf 0.401 (0.446)} & 0.507 (0.594) & 0.412 (0.423) \\
    4 & 30.7 (88.1) & {\bf 27.7 (260)} & 30.7 (88.2) & 30.6 (88.1) \\
    5 & 0.980 (0.771) & {\bf 0.791 (0.845)} & 1.04 (1.07) & 0.934 (0.713) \\
    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & {\bf 1.01 (2.11)} \\
    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & {\bf 1.95 (0.800)} \\
    8 & 0.226 (0.275) & {\bf 0.136 (0.169)} & 0.242 (0.311) & 0.274 (0.243) \\
    \hline
  \end{tabular}
  \caption{MSD comparison to true values, Mean (SD). For each case, the lowest algorithm-specific mean (SD) are highlighted in bold. }
  \label{table:simulation_results_mean}
\end{table}

\begin{figure}
\centering
\subfloat{
\label{fig:adni_lhipp_result}
\subfloat[\centering Data]{
\includegraphics[height=10cm]{adni_plots/adni_lhipp_data_plot}
}
\hfill
\subfloat[\centering LPME]{
\includegraphics[height=10cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
}
\hfill
\subfloat[\centering PME]{
\includegraphics[height=10cm]{adni_plots/adni_lhipp_pme_plot}
}
}
\hfill
\subfloat{
\label{fig:adni_lthal_result}
\subfloat[\centering Data]{
\includegraphics[height=10cm]{adni_plots/adni_lthal_data_plot}
}
\hfill
\subfloat[\centering LPME]{
\includegraphics[height=10cm]{adni_plots/adni_lthal_lpme_isomap_plot}
}
\hfill
\subfloat[\centering PME]{
\includegraphics[height=10cm]{adni_plots/adni_lthal_pme_plot}
}
}
\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant. The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue. The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
\label{fig:adni_result}
\end{figure}

\begin{figure}
    \centering
    \subfloat[\centering Left Hippocampus Volume Estimates]{
      \includegraphics[height=8cm]{adni_plots/adni_lhipp_volume_comp}
      \label{fig:lhipp_volume_comparison}
    }
    \vfill
    \subfloat[\centering Left Thalamus Volume Estimates]{
      \includegraphics[height=8cm]{adni_plots/adni_lthal_volume_comp}
      \label{fig:lthal_volume_comparison}
    }
    \caption{Left hippocampus and left thalamus volume estimates for three ADNI participants. Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (red, circles), LPME (blue, squares), and PME (green, triangles). When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of the volume estimates. Gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data. These gaps may also cause inconsistency in the LPME volume estimates, resulting in higher aggregate variability than for the data or PME estimates. When applied to the thalamus, the PME-based volume estimates demonstrate similar time point-to-time point changes in the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data. The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 011\_S\_0003.}
\end{figure}

\end{document}

### 2024-07-10 code/04_lpme_adni_application.qmd
complete lpme adni application code in notebooks

---
title: "Application of LPME Algorithm to ADNI Data"
author: "Robert Zielinski"
date: "June 14, 2024"
output: html_document

execute:
  eval: false
---

This notebook contains the code necessary to run the PME and LPME algorithms on the processed MRI data from the ADNI study. The code assumes that the data for the left hippocampuses, left thalamuses, right hippocampuses, and right thalamuses of all particpants of interest are stored in the following files

- `data/adni_fsl_lhipp_surface.csv`*Path to your R executable*
- `data/adni_fsl_lthal_surface.csv`
- `data/adni_fsl_rhipp_surface.csv`
- `data/adni_fsl_rthal_surface.csv`

The previous notebook, `code/03_adni_mri_preprocess.qmd`, includes the code used to process the MRI data into the proper format.

```{r}
knitr::opts_chunk$set(eval = FALSE)

# load all required packages and custom functions
library(foreach)
library(here)
library(lubridate)
library(plotly)
library(plot3D)
library(pme)
library(pracma)
library(RColorBrewer)
library(Rfast)
library(tidyverse)

source(here("code/functions/calc_pme_est.R"))
source(here("code/functions/calc_lpme_est_params.R"))
source(here("code/functions/estimate_volume.R"))
source(here("code/functions/interior_identification.R"))
source(here("code/functions/create_cross_section_matrix.R"))
```

```{r}
# read surface voxel data for the structures of interest
lhipp_surface <- read_csv(here("data/adni_fsl_lhipp_surface.csv"))
lthal_surface <- read_csv(here("data/adni_fsl_lthal_surface.csv"))
rhipp_surface <- read_csv(here("data/adni_fsl_rhipp_surface.csv"))
rthal_surface <- read_csv(here("data/adni_fsl_rthal_surface.csv"))

# read FSL-generated volume data for hippocampus and thalamus structures
hipp_info <- read_csv(here("data/adni_fsl_hipp_info.csv"))
thal_info <- read_csv(here("data/adni_fsl_thal_info.csv"))
```

```{r}
# create data frames to store volume data from hippocampus and thalamus estimates
est_hipp_info <- data.frame(
  patno = character(),
  date = numeric(),
  lhipp_data_vol2 = numeric(),
  lhipp_vol_lpme1 = numeric(),
  lhipp_vol_lpme2 = numeric(),
  lhipp_vol_pme1 = numeric(),
  lhipp_vol_pme2 = numeric(),
  rhipp_data_vol2 = numeric(),
  rhipp_vol_lpme1 = numeric(),
  rhipp_vol_lpme2 = numeric(),
  rhipp_vol_pme1 = numeric(),
  rhipp_vol_pme2 = numeric()
)

est_thal_info <- data.frame(
  patno = character(),
  date = numeric(),
  lthal_data_vol2 = numeric(),
  lthal_vol_lpme1 = numeric(),
  lthal_vol_lpme2 = numeric(),
  lthal_vol_pme1 = numeric(),
  lthal_vol_pme2 = numeric(),
  rthal_data_vol2 = numeric(),
  rthal_vol_lpme1 = numeric(),
  rthal_vol_lpme2 = numeric(),
  rthal_vol_pme1 = numeric(),
  rthal_vol_pme2 = numeric()
)
```

```{r}
# process scan dates
lhipp_surface <- lhipp_surface %>%
  mutate(
    scan_date = gsub("\\.0", "", scan_date),
    scan_date = ymd_hms(scan_date),
    scan_date = decimal_date(scan_date)
  )
lthal_surface <- lthal_surface %>%
  mutate(
    scan_date = gsub("\\.0", "", scan_date),
    scan_date = ymd_hms(scan_date),
    scan_date = decimal_date(scan_date)
  )
rhipp_surface <- rhipp_surface %>%
  mutate(
    scan_date = gsub("\\.0", "", scan_date),
    scan_date = ymd_hms(scan_date),
    scan_date = decimal_date(scan_date)
  )
rthal_surface <- rthal_surface %>%
  mutate(
    scan_date = gsub("\\.0", "", scan_date),
    scan_date = ymd_hms(scan_date),
    scan_date = decimal_date(scan_date)
  )

hipp_info <- hipp_info %>%
  mutate(
    date = gsub("\\.0", "", date),
    date = ymd_hms(date),
    date = decimal_date(date)
  )
thal_info <- thal_info %>%
  mutate(
    date = gsub("\\.0", "", date),
    date = ymd_hms(date),
    date = decimal_date(date)
  )

# calculate mean, max, and sd of voxel coordinates
lhipp_surface_centers <- lhipp_surface %>%
  group_by(patno, scan_date) %>%
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    mean_z = mean(z),
    max_x = max(abs(x)),
    max_y = max(abs(y)),
    max_z = max(abs(z)),
    sd_x = sd(x),
    sd_y = sd(y),
    sd_z = sd(z)
  )

lthal_surface_centers <- lthal_surface %>%
  group_by(patno, scan_date) %>%
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    mean_z = mean(z),
    max_x = max(abs(x)),
    max_y = max(abs(y)),
    max_z = max(abs(z)),
    sd_x = sd(x),
    sd_y = sd(y),
    sd_z = sd(z)
  )

rhipp_surface_centers <- rhipp_surface %>%
  group_by(patno, scan_date) %>%
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    mean_z = mean(z),
    max_x = max(abs(x)),
    max_y = max(abs(y)),
    max_z = max(abs(z)),
    sd_x = sd(x),
    sd_y = sd(y),
    sd_z = sd(z)
  )

rthal_surface_centers <- rthal_surface %>%
  group_by(patno, scan_date) %>%
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    mean_z = mean(z),
    max_x = max(abs(x)),
    max_y = max(abs(y)),
    max_z = max(abs(z)),
    sd_x = sd(x),
    sd_y = sd(y),
    sd_z = sd(z)
  )

```

```{r}
# for each participant, calculate the date of their baseline visit and
# the date of their final visit
lhipp_bl <- lhipp_surface %>%
  group_by(patno) %>%
  arrange(scan_date) %>%
  summarize(
    time_bl = first(scan_date),
    max_time = max(scan_date)
  ) %>%
  # study duration for this participant is the difference between the dates
  # of their first and last visits
  mutate(duration = max_time - time_bl)

lthal_bl <- lthal_surface %>%
  group_by(patno) %>%
  arrange(scan_date) %>%
  summarize(
    time_bl = first(scan_date),
    max_time = max(scan_date)
  ) %>%
  mutate(duration = max_time - time_bl)

rhipp_bl <- rhipp_surface %>%
  group_by(patno) %>%
  arrange(scan_date) %>%
  summarize(
    time_bl = first(scan_date),
    max_time = max(scan_date)
  ) %>%
  mutate(duration = max_time - time_bl)

rthal_bl <- rthal_surface %>%
  group_by(patno) %>%
  arrange(scan_date) %>%
  summarize(
    time_bl = first(scan_date),
    max_time = max(scan_date)
  ) %>%
  mutate(duration = max_time - time_bl)
```

```{r}
# center and standardize data
lhipp_surface <- lhipp_surface %>%
  full_join(lhipp_surface_centers, by = c("patno", "scan_date")) %>%
  full_join(lhipp_bl, by = "patno") %>%
  mutate(
    x = (x - mean_x) / max_x,
    y = (y - mean_y) / max_y,
    z = (z - mean_z) / max_z,
    time_from_bl = scan_date - time_bl
  )

lthal_surface <- lthal_surface %>%
  full_join(lthal_surface_centers, by = c("patno", "scan_date")) %>%
  full_join(lthal_bl, by = "patno") %>%
  mutate(
    x = (x - mean_x) / max_x,
    y = (y - mean_y) / max_y,
    z = (z - mean_z) / max_z,
    time_from_bl = scan_date - time_bl
  )

rhipp_surface <- rhipp_surface %>%
  full_join(rhipp_surface_centers, by = c("patno", "scan_date")) %>%
  full_join(rhipp_bl, by = "patno") %>%
  mutate(
    x = (x - mean_x) / max_x,
    y = (y - mean_y) / max_y,
    z = (z - mean_z) / max_z,
    time_from_bl = scan_date - time_bl
  )

rthal_surface <- rthal_surface %>%
  full_join(rthal_surface_centers, by = c("patno", "scan_date")) %>%
  full_join(rthal_bl, by = "patno") %>%
  mutate(
    x = (x - mean_x) / max_x,
    y = (y - mean_y) / max_y,
    z = (z - mean_z) / max_z,
    time_from_bl = scan_date - time_bl
  )
```

```{r}
# exclude participants with duration shorter than 2 years
lhipp_patnos <- lhipp_surface %>%
  filter(duration > 2) %>%
  .$patno %>%
  unique()

lhipp_surface <- lhipp_surface %>%
  filter(patno %in% lhipp_patnos)

lthal_surface <- lthal_surface %>%
  filter(patno %in% lhipp_patnos)

rhipp_surface <- rhipp_surface %>%
  filter(patno %in% lhipp_patnos)

rthal_surface <- rthal_surface %>%
  filter(patno %in% lhipp_patnos)
```

```{r}
# augment data with spherical coordinates
lhipp_surface_spherical <- lhipp_surface %>%
  dplyr::select(x, y, z) %>%
  as.matrix() %>%
  cart2sph() %>%
  as_tibble()

lthal_surface_spherical <- lthal_surface %>%
  dplyr::select(x, y, z) %>%
  as.matrix() %>%
  cart2sph() %>%
  as_tibble()

rhipp_surface_spherical <- rhipp_surface %>%
  dplyr::select(x, y, z) %>%
  as.matrix() %>%
  cart2sph() %>%
  as_tibble()

rthal_surface_spherical <- rthal_surface %>%
  dplyr::select(x, y, z) %>%
  as.matrix() %>%
  cart2sph() %>%
  as_tibble()

lhipp_surface <- bind_cols(lhipp_surface, lhipp_surface_spherical)
lthal_surface <- bind_cols(lthal_surface, lthal_surface_spherical)
rhipp_surface <- bind_cols(rhipp_surface, rhipp_surface_spherical)
rthal_surface <- bind_cols(rthal_surface, rthal_surface_spherical)
```

```{r}
write_csv(lhipp_surface, here("data/adni_lhipp_surface_processed.csv"))
write_csv(lthal_surface, here("data/adni_lthal_surface_processed.csv"))
write_csv(rhipp_surface, here("data/adni_rhipp_surface_processed.csv"))
write_csv(rthal_surface, here("data/adni_rthal_surface_processed.csv"))
```

```{r}
patnos <- lhipp_surface$patno %>%
  unique()

ncores <- parallel::detectCores()
cl <- parallel::makeCluster(ncores / 2, type = "FORK")
doParallel::registerDoParallel(cl)
```

```{r}
set.seed(10283)
foreach(
  patno_val = patnos,
  .packages = c(
    "pme",
    "plotly",
    "plot3D",
    "pracma",
    "RColorBrewer",
    "Rfast",
    "tidyverse"
  ),
  .export = c(
    "calc_pme_est",
    "calc_lpme_est_params",
    "estimate_volume",
    "interior_identification",
    "create_cross_section_matrix"
  )
) %dopar% {
# for (patno_val in patnos) {
  print(patno_val)
  lhipp <- lhipp_surface %>%
    filter(patno == patno_val)
  lthal <- lthal_surface %>%
    filter(patno == patno_val)
  rhipp <- rhipp_surface %>%
    filter(patno == patno_val)
  rthal <- rthal_surface %>%
    filter(patno == patno_val)

  lhipp_mat <- lhipp %>%
    dplyr::select(
      time_from_bl,
      x,
      y,
      z,
      theta,
      phi,
      r
    ) %>%
    as.matrix()

  lthal_mat <- lthal %>%
    filter(patno == patno_val) %>%
    dplyr::select(
      time_from_bl,
      x,
      y,
      z,
      theta,
      phi,
      r
    ) %>%
    as.matrix()

  rhipp_mat <- rhipp %>%
    filter(patno == patno_val) %>%
    dplyr::select(
      time_from_bl,
      x,
      y,
      z,
      theta,
      phi,
      r
    ) %>%
    as.matrix()
  rthal_mat <- rthal %>%
    filter(patno == patno_val) %>%
    dplyr::select(
      time_from_bl,
      x,
      y,
      z,
      theta,
      phi,
      r
    ) %>%
    as.matrix()

  time_vals <- unique(lhipp_mat[, 1])

  print("Fitting LHIPP Models")

  lhipp_lpme_result_isomap <- lpme(
    lhipp_mat,
    d = 2,
    gamma = exp(-15:5),
    lambda = exp(-12:5),
    initialization_algorithm = "isomap",
    init_type = "centers",
    verbose = FALSE,
    print_plots = FALSE
  )
  lhipp_lpme_isomap_est <- calc_lpme_est_params(lhipp_lpme_result_isomap, lhipp_mat)
  lhipp_lpme_isomap_vals <- lhipp_lpme_isomap_est$results
  lhipp_lpme_isomap_params <- lhipp_lpme_isomap_est$params

  lhipp_pme_result <- list()
  lhipp_pme_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- lhipp_mat[lhipp_mat[, 1] == time_vals[t], -1]
    lhipp_pme_result[[t]] <- pme(temp_data, d = 2, lambda = exp(-12:5), verbose = FALSE)
    lhipp_pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(lhipp_pme_result[[t]], temp_data))
  }
  lhipp_pme_vals <- purrr::reduce(lhipp_pme_vals, rbind)

  patno_path <- here(paste0("results/", patno_val))
  if (!dir.exists(patno_path)) {
    dir.create(patno_path)
  }
  saveRDS(lhipp_lpme_result_isomap, paste0(patno_path, "/adni_lhipp_lpme_isomap.rds"))
  saveRDS(lhipp_pme_result, paste0(patno_path, "/adni_lhipp_pme.rds"))
  write.csv(lhipp_mat, paste0(patno_path, "/adni_lhipp_mat.csv"))
  write.csv(lhipp_lpme_isomap_vals, paste0(patno_path, "/adni_lhipp_lpme_isomap_vals.csv"))
  write.csv(lhipp_pme_vals, paste0(patno_path, "/adni_lhipp_pme_vals.csv"))

  print("Files saved")

  ### VOLUME ESTIMATION

  print("LHIPP Volume Estimation")

  lhipp_lpme_vals <- lhipp_lpme_isomap_vals

  lhipp_data_rescaled <- list()
  lhipp_lpme_rescaled <- list()
  lhipp_pme_rescaled <- list()

  for (time_idx in 1:length(time_vals)) {
    temp_data <- lhipp_mat[lhipp_mat[, 1] == time_vals[time_idx], ]
    temp_lpme <- lhipp_lpme_vals[lhipp_lpme_vals[, 1] == time_vals[time_idx], ]
    temp_pme <- lhipp_pme_vals[lhipp_pme_vals[, 1] == time_vals[time_idx], ]

    temp_max_x <- unique(lhipp[lhipp$time_from_bl == time_vals[time_idx], ]$max_x)
    temp_max_y <- unique(lhipp[lhipp$time_from_bl == time_vals[time_idx], ]$max_y)
    temp_max_z <- unique(lhipp[lhipp$time_from_bl == time_vals[time_idx], ]$max_z)

    temp_data_rescaled <- temp_data[, -(5:7)]
    temp_lpme_rescaled <- temp_lpme[, -(5:7)]
    temp_pme_rescaled <- temp_pme[, -(5:7)]

    temp_data_rescaled[, 2] <- round((temp_data_rescaled[, 2] * temp_max_x) / 1.2)
    temp_data_rescaled[, 3] <- round((temp_data_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_data_rescaled[, 4] <- round((temp_data_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_lpme_rescaled[, 2] <- round((temp_lpme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_lpme_rescaled[, 3] <- round((temp_lpme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_lpme_rescaled[, 4] <- round((temp_lpme_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_pme_rescaled[, 2] <- round((temp_pme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_pme_rescaled[, 3] <- round((temp_pme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_pme_rescaled[, 4] <- round((temp_pme_rescaled[, 4] * temp_max_z) / 0.9375)

    lhipp_data_rescaled[[time_idx]] <- temp_data_rescaled
    lhipp_lpme_rescaled[[time_idx]] <- temp_lpme_rescaled
    lhipp_pme_rescaled[[time_idx]] <- temp_pme_rescaled
  }

  lhipp_data_rescaled_full <- reduce(lhipp_data_rescaled, rbind)
  lhipp_lpme_rescaled_full <- reduce(lhipp_lpme_rescaled, rbind)
  lhipp_pme_rescaled_full <- reduce(lhipp_pme_rescaled, rbind)

  voxel_vol <- 1.2 * 0.9375 * 0.9375

  lhipp_data_volumes <- map(
    lhipp_data_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  lhipp_data_volume_points <- map(
    lhipp_data_volumes,
    ~ .x[[2]]
  )

  lhipp_data_volumes <- map(
    lhipp_data_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  lhipp_lpme_volumes <- map(
    lhipp_lpme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  lhipp_lpme_volume_points <- map(
    lhipp_lpme_volumes,
    ~ .x[[2]]
  )

  lhipp_lpme_volumes <- map(
    lhipp_lpme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  lhipp_pme_volumes <- map(
    lhipp_pme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  lhipp_pme_volume_points <- map(
    lhipp_pme_volumes,
    ~ .x[[2]]
  )

  lhipp_pme_volumes <- map(
    lhipp_pme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  lhipp_data_volume_points_full <- map(
    1:length(lhipp_data_volume_points),
    ~ cbind(time_vals[.x], lhipp_data_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  lhipp_lpme_volume_points_full <- map(
    1:length(lhipp_lpme_volume_points),
    ~ cbind(time_vals[.x], lhipp_lpme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  lhipp_pme_volume_points_full <- map(
    1:length(lhipp_pme_volume_points),
    ~ cbind(time_vals[.x], lhipp_pme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  print("Fitting LTHAL Models")

  lthal_lpme_result_isomap <- lpme(
    lthal_mat,
    d = 2,
    gamma = exp(-15:5),
    lambda = exp(-12:5),
    initialization_algorithm = "isomap",
    init_type = "centers",
    verbose = FALSE,
    print_plots = FALSE
  )
  lthal_lpme_isomap_est <- calc_lpme_est_params(lthal_lpme_result_isomap, lthal_mat)
  lthal_lpme_isomap_vals <- lthal_lpme_isomap_est$results
  lthal_lpme_isomap_params <- lthal_lpme_isomap_est$params

  lthal_pme_result <- list()
  lthal_pme_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- lthal_mat[lthal_mat[, 1] == time_vals[t], -1]
    lthal_pme_result[[t]] <- pme(temp_data, d = 2, lambda = exp(-12:5), verbose = FALSE)
    lthal_pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(lthal_pme_result[[t]], temp_data))
  }
  lthal_pme_vals <- reduce(lthal_pme_vals, rbind)

  saveRDS(lthal_lpme_result_isomap, paste0(patno_path, "/adni_lthal_lpme_isomap.rds"))
  saveRDS(lthal_pme_result, paste0(patno_path, "/adni_lthal_pme.rds"))
  write.csv(lthal_mat, paste0(patno_path, "/adni_lthal_mat.csv"))
  write.csv(lthal_lpme_isomap_vals, paste0(patno_path, "/adni_lthal_lpme_isomap_vals.csv"))
  write.csv(lthal_pme_vals, paste0(patno_path, "/adni_lthal_pme_vals.csv"))

  print("LTHAL Files Saved")

  ### VOLUME ESTIMATION

  print("LTHAL Volume Estimation")

  lthal_lpme_vals <- lthal_lpme_isomap_vals

  lthal_data_rescaled <- list()
  lthal_lpme_rescaled <- list()
  lthal_pme_rescaled <- list()

  for (time_idx in 1:length(time_vals)) {
    temp_data <- lthal_mat[lthal_mat[, 1] == time_vals[time_idx], ]
    temp_lpme <- lthal_lpme_vals[lthal_lpme_vals[, 1] == time_vals[time_idx], ]
    temp_pme <- lthal_pme_vals[lthal_pme_vals[, 1] == time_vals[time_idx], ]

    temp_max_x <- unique(lthal[lthal$time_from_bl == time_vals[time_idx], ]$max_x)
    temp_max_y <- unique(lthal[lthal$time_from_bl == time_vals[time_idx], ]$max_y)
    temp_max_z <- unique(lthal[lthal$time_from_bl == time_vals[time_idx], ]$max_z)

    temp_data_rescaled <- temp_data[, -(5:7)]
    temp_lpme_rescaled <- temp_lpme[, -(5:7)]
    temp_pme_rescaled <- temp_pme[, -(5:7)]

    temp_data_rescaled[, 2] <- round((temp_data_rescaled[, 2] * temp_max_x) / 1.2)
    temp_data_rescaled[, 3] <- round((temp_data_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_data_rescaled[, 4] <- round((temp_data_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_lpme_rescaled[, 2] <- round((temp_lpme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_lpme_rescaled[, 3] <- round((temp_lpme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_lpme_rescaled[, 4] <- round((temp_lpme_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_pme_rescaled[, 2] <- round((temp_pme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_pme_rescaled[, 3] <- round((temp_pme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_pme_rescaled[, 4] <- round((temp_pme_rescaled[, 4] * temp_max_z) / 0.9375)

    lthal_data_rescaled[[time_idx]] <- temp_data_rescaled
    lthal_lpme_rescaled[[time_idx]] <- temp_lpme_rescaled
    lthal_pme_rescaled[[time_idx]] <- temp_pme_rescaled
  }

  lthal_data_rescaled_full <- reduce(lthal_data_rescaled, rbind)
  lthal_lpme_rescaled_full <- reduce(lthal_lpme_rescaled, rbind)
  lthal_pme_rescaled_full <- reduce(lthal_pme_rescaled, rbind)

  voxel_vol <- 1.2 * 0.9375 * 0.9375

  lthal_data_volumes <- map(
    lthal_data_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  lthal_data_volume_points <- map(
    lthal_data_volumes,
    ~ .x[[2]]
  )

  lthal_data_volumes <- map(
    lthal_data_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  lthal_lpme_volumes <- map(
    lthal_lpme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  lthal_lpme_volume_points <- map(
    lthal_lpme_volumes,
    ~ .x[[2]]
  )

  lthal_lpme_volumes <- map(
    lthal_lpme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  lthal_pme_volumes <- map(
    lthal_pme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  lthal_pme_volume_points <- map(
    lthal_pme_volumes,
    ~ .x[[2]]
  )

  lthal_pme_volumes <- map(
    lthal_pme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  lthal_data_volume_points_full <- map(
    1:length(lthal_data_volume_points),
    ~ cbind(time_vals[.x], lthal_data_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  lthal_lpme_volume_points_full <- map(
    1:length(lthal_lpme_volume_points),
    ~ cbind(time_vals[.x], lthal_lpme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  lthal_pme_volume_points_full <- map(
    1:length(lthal_pme_volume_points),
    ~ cbind(time_vals[.x], lthal_pme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  print("Fitting RHIPP Models")

  rhipp_lpme_result_isomap <- lpme(
    rhipp_mat,
    d = 2,
    gamma = exp(-15:5),
    lambda = exp(-12:5),
    initialization_algorithm = "isomap",
    init_type = "centers",
    verbose = FALSE,
    print_plots = FALSE
  )
  rhipp_lpme_isomap_est <- calc_lpme_est_params(rhipp_lpme_result_isomap, rhipp_mat)
  rhipp_lpme_isomap_vals <- rhipp_lpme_isomap_est$results
  rhipp_lpme_isomap_params <- rhipp_lpme_isomap_est$params

  rhipp_pme_result <- list()
  rhipp_pme_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- rhipp_mat[rhipp_mat[, 1] == time_vals[t], -1]
    rhipp_pme_result[[t]] <- pme(temp_data, d = 2, lambda = exp(-12:5), verbose = FALSE)
    rhipp_pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(rhipp_pme_result[[t]], temp_data))
  }
  rhipp_pme_vals <- reduce(rhipp_pme_vals, rbind)

  saveRDS(rhipp_lpme_result_isomap, paste0(patno_path, "/adni_rhipp_lpme_isomap.rds"))
  saveRDS(rhipp_pme_result, paste0(patno_path, "/adni_rhipp_pme.rds"))
  write.csv(rhipp_mat, paste0(patno_path, "/adni_rhipp_mat.csv"))
  write.csv(rhipp_lpme_isomap_vals, paste0(patno_path, "/adni_rhipp_lpme_isomap_vals.csv"))
  write.csv(rhipp_pme_vals, paste0(patno_path, "/adni_rhipp_pme_vals.csv"))

  print("RHIPP Files Saved")

  #### VOLUME ESTIMATION

  # Voxel dimension: 1.2mm x 0.9375mm x 0.9375mm
  # Voxel volume: 1.054688 mm^3
  print("RHIPP Volume Estimation")
  ### VOLUME ESTIMATION

  rhipp_lpme_vals <- rhipp_lpme_isomap_vals

  rhipp_data_rescaled <- list()
  rhipp_lpme_rescaled <- list()
  rhipp_pme_rescaled <- list()

  for (time_idx in 1:length(time_vals)) {
    temp_data <- rhipp_mat[rhipp_mat[, 1] == time_vals[time_idx], ]
    temp_lpme <- rhipp_lpme_vals[rhipp_lpme_vals[, 1] == time_vals[time_idx], ]
    temp_pme <- rhipp_pme_vals[rhipp_pme_vals[, 1] == time_vals[time_idx], ]

    temp_max_x <- unique(rhipp[rhipp$time_from_bl == time_vals[time_idx], ]$max_x)
    temp_max_y <- unique(rhipp[rhipp$time_from_bl == time_vals[time_idx], ]$max_y)
    temp_max_z <- unique(rhipp[rhipp$time_from_bl == time_vals[time_idx], ]$max_z)

    temp_data_rescaled <- temp_data[, -(5:7)]
    temp_lpme_rescaled <- temp_lpme[, -(5:7)]
    temp_pme_rescaled <- temp_pme[, -(5:7)]

    temp_data_rescaled[, 2] <- round((temp_data_rescaled[, 2] * temp_max_x) / 1.2)
    temp_data_rescaled[, 3] <- round((temp_data_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_data_rescaled[, 4] <- round((temp_data_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_lpme_rescaled[, 2] <- round((temp_lpme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_lpme_rescaled[, 3] <- round((temp_lpme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_lpme_rescaled[, 4] <- round((temp_lpme_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_pme_rescaled[, 2] <- round((temp_pme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_pme_rescaled[, 3] <- round((temp_pme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_pme_rescaled[, 4] <- round((temp_pme_rescaled[, 4] * temp_max_z) / 0.9375)

    rhipp_data_rescaled[[time_idx]] <- temp_data_rescaled
    rhipp_lpme_rescaled[[time_idx]] <- temp_lpme_rescaled
    rhipp_pme_rescaled[[time_idx]] <- temp_pme_rescaled
  }

  rhipp_data_rescaled_full <- reduce(rhipp_data_rescaled, rbind)
  rhipp_lpme_rescaled_full <- reduce(rhipp_lpme_rescaled, rbind)
  rhipp_pme_rescaled_full <- reduce(rhipp_pme_rescaled, rbind)

  voxel_vol <- 1.2 * 0.9375 * 0.9375

  rhipp_data_volumes <- map(
    rhipp_data_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  rhipp_data_volume_points <- map(
    rhipp_data_volumes,
    ~ .x[[2]]
  )

  rhipp_data_volumes <- map(
    rhipp_data_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  rhipp_lpme_volumes <- map(
    rhipp_lpme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  rhipp_lpme_volume_points <- map(
    rhipp_lpme_volumes,
    ~ .x[[2]]
  )

  rhipp_lpme_volumes <- map(
    rhipp_lpme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  rhipp_pme_volumes <- map(
    rhipp_pme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  rhipp_pme_volume_points <- map(
    rhipp_pme_volumes,
    ~ .x[[2]]
  )

  rhipp_pme_volumes <- map(
    rhipp_pme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  rhipp_data_volume_points_full <- map(
    1:length(rhipp_data_volume_points),
    ~ cbind(time_vals[.x], rhipp_data_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  rhipp_lpme_volume_points_full <- map(
    1:length(rhipp_lpme_volume_points),
    ~ cbind(time_vals[.x], rhipp_lpme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  rhipp_pme_volume_points_full <- map(
    1:length(rhipp_pme_volume_points),
    ~ cbind(time_vals[.x], rhipp_pme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  print("Fitting RTHAL Models")
  rthal_lpme_result_isomap <- lpme(
    rthal_mat,
    d = 2,
    gamma = exp(-15:5),
    lambda = exp(-12:5),
    initialization_algorithm = "isomap",
    init_type = "centers",
    verbose = FALSE,
    print_plots = FALSE
  )
  rthal_lpme_isomap_est <- calc_lpme_est_params(rthal_lpme_result_isomap, rthal_mat)
  rthal_lpme_isomap_vals <- rthal_lpme_isomap_est$results
  rthal_lpme_isomap_params <- rthal_lpme_isomap_est$params

  rthal_pme_result <- list()
  rthal_pme_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- rthal_mat[rthal_mat[, 1] == time_vals[t], -1]
    rthal_pme_result[[t]] <- pme(temp_data, d = 2, lambda = exp(-12:5), verbose = FALSE)
    rthal_pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(rthal_pme_result[[t]], temp_data))
  }
  rthal_pme_vals <- reduce(rthal_pme_vals, rbind)

  saveRDS(rthal_lpme_result_isomap, paste0(patno_path, "/adni_rthal_lpme_isomap.rds"))
  saveRDS(rthal_pme_result, paste0(patno_path, "/adni_rthal_pme.rds"))
  write.csv(rthal_mat, paste0(patno_path, "/adni_rthal_mat.csv"))
  write.csv(rthal_lpme_isomap_vals, paste0(patno_path, "/adni_rthal_lpme_isomap_vals.csv"))
  write.csv(rthal_pme_vals, paste0(patno_path, "/adni_rthal_pme_vals.csv"))

  print("RTHAL Files Saved")

  ### VOLUME ESTIMATION

  print("RTHAL Volume Estimation")
  rthal_lpme_vals <- rthal_lpme_isomap_vals

  rthal_data_rescaled <- list()
  rthal_lpme_rescaled <- list()
  rthal_pme_rescaled <- list()

  for (time_idx in 1:length(time_vals)) {
    temp_data <- rthal_mat[rthal_mat[, 1] == time_vals[time_idx], ]
    temp_lpme <- rthal_lpme_vals[rthal_lpme_vals[, 1] == time_vals[time_idx], ]
    temp_pme <- rthal_pme_vals[rthal_pme_vals[, 1] == time_vals[time_idx], ]

    temp_max_x <- unique(rthal[rthal$time_from_bl == time_vals[time_idx], ]$max_x)
    temp_max_y <- unique(rthal[rthal$time_from_bl == time_vals[time_idx], ]$max_y)
    temp_max_z <- unique(rthal[rthal$time_from_bl == time_vals[time_idx], ]$max_z)

    temp_data_rescaled <- temp_data[, -(5:7)]
    temp_lpme_rescaled <- temp_lpme[, -(5:7)]
    temp_pme_rescaled <- temp_pme[, -(5:7)]

    temp_data_rescaled[, 2] <- round((temp_data_rescaled[, 2] * temp_max_x) / 1.2)
    temp_data_rescaled[, 3] <- round((temp_data_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_data_rescaled[, 4] <- round((temp_data_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_lpme_rescaled[, 2] <- round((temp_lpme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_lpme_rescaled[, 3] <- round((temp_lpme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_lpme_rescaled[, 4] <- round((temp_lpme_rescaled[, 4] * temp_max_z) / 0.9375)

    temp_pme_rescaled[, 2] <- round((temp_pme_rescaled[, 2] * temp_max_x) / 1.2)
    temp_pme_rescaled[, 3] <- round((temp_pme_rescaled[, 3] * temp_max_y) / 0.9375)
    temp_pme_rescaled[, 4] <- round((temp_pme_rescaled[, 4] * temp_max_z) / 0.9375)

    rthal_data_rescaled[[time_idx]] <- temp_data_rescaled
    rthal_lpme_rescaled[[time_idx]] <- temp_lpme_rescaled
    rthal_pme_rescaled[[time_idx]] <- temp_pme_rescaled
  }

  rthal_data_rescaled_full <- reduce(rthal_data_rescaled, rbind)
  rthal_lpme_rescaled_full <- reduce(rthal_lpme_rescaled, rbind)
  rthal_pme_rescaled_full <- reduce(rthal_pme_rescaled, rbind)

  voxel_vol <- 1.2 * 0.9375 * 0.9375

  rthal_data_volumes <- map(
    rthal_data_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  rthal_data_volume_points <- map(
    rthal_data_volumes,
    ~ .x[[2]]
  )

  rthal_data_volumes <- map(
    rthal_data_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  rthal_lpme_volumes <- map(
    rthal_lpme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  rthal_lpme_volume_points <- map(
    rthal_lpme_volumes,
    ~ .x[[2]]
  )

  rthal_lpme_volumes <- map(
    rthal_lpme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  rthal_pme_volumes <- map(
    rthal_pme_rescaled,
    ~ estimate_volume(.x[, -1], voxel_vol)
  )

  rthal_pme_volume_points <- map(
    rthal_pme_volumes,
    ~ .x[[2]]
  )

  rthal_pme_volumes <- map(
    rthal_pme_volumes,
    ~ .x[[1]]
  ) %>%
    reduce(c)

  rthal_data_volume_points_full <- map(
    1:length(rthal_data_volume_points),
    ~ cbind(time_vals[.x], rthal_data_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  rthal_lpme_volume_points_full <- map(
    1:length(rthal_lpme_volume_points),
    ~ cbind(time_vals[.x], rthal_lpme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  rthal_pme_volume_points_full <- map(
    1:length(rthal_pme_volume_points),
    ~ cbind(time_vals[.x], rthal_pme_volume_points[[.x]])
  ) %>%
    reduce(rbind)

  dates <- unique(lhipp$time_bl) + time_vals

  print("Creating Temporary Files...")

  temp_hipp_info <- tibble(
    patno = patno_val,
    date = dates,
    lhipp_data_vol = lhipp_data_volumes,
    lhipp_vol_lpme = lhipp_lpme_volumes,
    lhipp_vol_pme = lhipp_pme_volumes,
    rhipp_data_vol = rhipp_data_volumes,
    rhipp_vol_lpme = rhipp_lpme_volumes,
    rhipp_vol_pme = rhipp_pme_volumes
  )

  temp_thal_info <- tibble(
    patno = patno_val,
    date = dates,
    lthal_data_vol = lthal_data_volumes,
    lthal_vol_lpme = lthal_lpme_volumes,
    lthal_vol_pme = lthal_pme_volumes,
    rthal_data_vol = rthal_data_volumes,
    rthal_vol_lpme = rthal_lpme_volumes,
    rthal_vol_pme = rthal_pme_volumes
  )

  est_hipp_info <- bind_rows(est_hipp_info, temp_hipp_info)
  est_thal_info <- bind_rows(est_thal_info, temp_thal_info)

  write.csv(temp_hipp_info, file = paste0(patno_path, "/hipp_info.csv"))
  write.csv(temp_thal_info, file = paste0(patno_path, "/thal_info.csv"))

  print("Temporary Files Saved")
}

parallel::stopCluster(cl)
```

```{r}
write.csv(est_hipp_info, here("results/est_hipp_info.csv"))
write.csv(est_thal_info, here("results/est_thal_info.csv"))
```

### 2024-07-10 code/05_lpme_adni_results.qmd
complete lpme adni application code in notebooks

---
title: "ADNI Application Results"
author: "Robert Zielinski"
date: "June 14, 2024"
output: html_document

execute:
  eval: false
---

This notebook includes the code used to generate plots and other visualizations from the results of applying the LPME and PME algorithms to ADNI data, as well as the code needed to summarize the results as shown in the paper.

```{r}
library(bazar)
library(brolgar)
library(here)
library(numbers)
library(plot3D)
library(plotly)
library(RColorBrewer)
library(scatterplot3d)
library(tidyverse)

source(here("code/functions/create_cross_section_matrix.R"))
```

```{r}
# set color palette to maintain consistency between plots
colors <- brewer.pal(3, "Set1")
```

```{r}
result_dirs <- list.dirs(here("results"), recursive = FALSE)
```

```{r}
lhipp_full <- read_csv(here("data/adni_lhipp_surface_processed.csv"))
rhipp_full <- read_csv(here("data/adni_rhipp_surface_processed.csv"))
lthal_full <- read_csv(here("data/adni_lthal_surface_processed.csv"))
rthal_full <- read_csv(here("data/adni_rthal_surface_processed.csv"))

patnos <- unique(lhipp_full$patno)
```

```{r}
for (dir_idx in 1:length(result_dirs)) {
  dir_path <- result_dirs[dir_idx]

  # load files with structure surface data
  lhipp_mat <- read_csv(paste0(dir_path, "/adni_lhipp_mat.csv"))[, -1]
  rhipp_mat <- read_csv(paste0(dir_path, "/adni_rhipp_mat.csv"))[, -1]
  lthal_mat <- read_csv(paste0(dir_path, "/adni_lthal_mat.csv"))[, -1]
  rthal_mat <- read_csv(paste0(dir_path, "/adni_rthal_mat.csv"))[, -1]

  lhipp <- lhipp_full %>% 
    filter(patno == patnos[dir_idx])
  rhipp <- rhipp_full %>% 
    filter(patno == patnos[dir_idx])
  lthal <- lthal_full %>%
    filter(patno == patnos[dir_idx])
  rthal <- rthal_full %>%
    filter(patno == patnos[dir_idx])

  lhipp_mat <- lhipp %>% 
    select(time_from_bl, x, y, z, theta, phi, r) %>% 
    as.matrix()
  rhipp_mat <- rhipp %>% 
    select(time_from_bl, x, y, z, theta, phi, r) %>% 
    as.matrix()
  lthal_mat <- lthal %>% 
    select(time_from_bl, x, y, z, theta, phi, r) %>% 
    as.matrix()
  rthal_mat <- rthal %>% 
    select(time_from_bl, x, y, z, theta, phi, r) %>% 
    as.matrix()

  # load lpme and pme estimates of all structures
  lhipp_lpme_isomap_vals <- read_csv(paste0(dir_path, "/adni_lhipp_lpme_isomap_vals.csv"))[, -1] %>% 
    as.matrix()
  lhipp_pme_vals <- read_csv(paste0(dir_path, "/adni_lhipp_pme_vals.csv"))[, -1] %>% 
    as.matrix()
  rhipp_lpme_isomap_vals <- read_csv(paste0(dir_path, "/adni_rhipp_lpme_isomap_vals.csv"))[, -1] %>% 
    as.matrix()
  rhipp_pme_vals <- read_csv(paste0(dir_path, "/adni_rhipp_pme_vals.csv"))[, -1] %>% 
    as.matrix()
  lthal_lpme_isomap_vals <- read_csv(paste0(dir_path, "/adni_lthal_lpme_isomap_vals.csv"))[, -1] %>% 
    as.matrix()
  lthal_pme_vals <- read_csv(paste0(dir_path, "/adni_lthal_pme_vals.csv"))[, -1] %>% 
    as.matrix()
  rthal_lpme_isomap_vals <- read_csv(paste0(dir_path, "/adni_rthal_lpme_isomap_vals.csv"))[, -1] %>% 
    as.matrix()
  rthal_pme_vals <- read_csv(paste0(dir_path, "/adni_rthal_pme_vals.csv"))[, -1] %>% 
    as.matrix()

  time_vals <- unlist(lhipp_mat[, 1]) %>% 
    unique()

  nplots <- floor(length(time_vals) / 2)

  lhipp_range <- colMinsMaxs(lhipp_mat[, 2:4]) %>% 
    abs() %>% 
    max()
  rhipp_range <- colMinsMaxs(rhipp_mat[, 2:4]) %>% 
    abs() %>% 
    max()
  lthal_range <- colMinsMaxs(lthal_mat[, 2:4]) %>% 
    abs() %>% 
    max()
  rthal_range <- colMinsMaxs(rthal_mat[, 2:4]) %>% 
    abs() %>% 
    max()

  lhipp_positions <- data.frame(
    c(0, round(-1.35 * lhipp_range, 2), round(1.35 * lhipp_range, 2)),
    c(round(1.45 * lhipp_range, 2), 0, round(1.35 * lhipp_range, 2)),
    c(round(-1.2 * lhipp_range, 2), round(-1.3 * lhipp_range, 2), 0)
  ) %>% 
    as.matrix(byrow = TRUE)

  rhipp_positions <- data.frame(
    c(0, round(-1.35 * rhipp_range, 2), round(1.35 * rhipp_range, 2)),
    c(round(1.45 * rhipp_range, 2), 0, round(1.35 * rhipp_range, 2)),
    c(round(-1.2 * rhipp_range, 2), round(-1.3 * rhipp_range, 2), 0)
  ) %>% 
    as.matrix(byrow = TRUE)

  lthal_positions <- data.frame(
    c(0, round(-1.35 * lthal_range, 2), round(1.35 * lthal_range, 2)),
    c(round(1.45 * lthal_range, 2), 0, round(1.35 * lthal_range, 2)),
    c(round(-1.2 * lthal_range, 2), round(-1.3 * lthal_range, 2), 0)
  ) %>% 
    as.matrix(byrow = TRUE)

  rthal_positions <- data.frame(
    c(0, round(-1.35 * rthal_range, 2), round(1.35 * rthal_range, 2)),
    c(round(1.45 * rthal_range, 2), 0, round(1.35 * rthal_range, 2)),
    c(round(-1.2 * rthal_range, 2), round(-1.3 * rthal_range, 2), 0)
  ) %>% 
    as.matrix(byrow = TRUE)

  plot_labels <- c("x", "y", "z")

  p <- plot_ly(
    x = lhipp_mat[, 2],
    y = lhipp_mat[, 3],
    z = lhipp_mat[, 4],
    frame = lhipp_mat[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "Data"
  ) %>%
    add_markers(
      x = lhipp_lpme_isomap_vals[, 2],
      y = lhipp_lpme_isomap_vals[, 3],
      z = lhipp_lpme_isomap_vals[, 4],
      frame = lhipp_lpme_isomap_vals[, 1],
      name = "LPME"
    ) %>%
    add_markers(
      x = lhipp_pme_vals[, 2],
      y = lhipp_pme_vals[, 3],
      z = lhipp_pme_vals[, 4],
      frame = lhipp_pme_vals[, 1],
      name = "PME"
    )

  lhipp_lpme_plots <- list()
  lhipp_pme_plots <- list()
  lhipp_data_plots <- list()

  lims <- c(-lhipp_range, lhipp_range)
  png(
    paste0(dir_path, "/adni_lhipp_data_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_data <- lhipp_mat[almost.equal(lhipp_mat[, 1], time_vals[t]), ]

      scatter3D(
        x = temp_data[, 2],
        y = temp_data[, 3],
        z = temp_data[, 4],
        pch = 20,
        col = alpha(colors[1], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = lhipp_positions[, 1],
        y = lhipp_positions[, 2],
        z = lhipp_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_lhipp_lpme_isomap_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_lpme <- lhipp_lpme_isomap_vals[almost.equal(lhipp_lpme_isomap_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_lpme[, 2],
        y = temp_lpme[, 3],
        z = temp_lpme[, 4],
        pch = 15,
        col = alpha(colors[2], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = lhipp_positions[, 1],
        y = lhipp_positions[, 2],
        z = lhipp_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_lhipp_pme_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_pme <- lhipp_pme_vals[almost.equal(lhipp_pme_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_pme[, 2],
        y = temp_pme[, 3],
        z = temp_pme[, 4],
        pch = 17,
        col = alpha(colors[3], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = lhipp_positions[, 1],
        y = lhipp_positions[, 2],
        z = lhipp_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  lhipp_cross_section <- create_cross_section_matrix(
    list(lhipp_mat, lhipp_lpme_isomap_vals, lhipp_pme_vals),
    4,
    c(0.005, 0.005, 0.005),
    nrow = 2,
    lhipp = lhipp
  )
  ggsave(
    paste0(dir_path, "/adni_lhipp_cross_section.png"),
    plot = lhipp_cross_section,
    height = 5000,
    width = 7000,
    units = "px",
    dpi = 1250
  )

  lims <- c(-rhipp_range, rhipp_range)
  png(
    paste0(dir_path, "/adni_rhipp_data_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_data <- rhipp_mat[almost.equal(rhipp_mat[, 1], time_vals[t]), ]

      scatter3D(
        x = temp_data[, 2],
        y = temp_data[, 3],
        z = temp_data[, 4],
        pch = 20,
        col = alpha(colors[1], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = rhipp_positions[, 1],
        y = rhipp_positions[, 2],
        z = rhipp_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_rhipp_lpme_isomap_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_lpme <- rhipp_lpme_isomap_vals[almost.equal(rhipp_lpme_isomap_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_lpme[, 2],
        y = temp_lpme[, 3],
        z = temp_lpme[, 4],
        pch = 15,
        col = alpha(colors[2], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = rhipp_positions[, 1],
        y = rhipp_positions[, 2],
        z = rhipp_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_rhipp_pme_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_pme <- rhipp_pme_vals[almost.equal(rhipp_pme_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_pme[, 2],
        y = temp_pme[, 3],
        z = temp_pme[, 4],
        pch = 17,
        col = alpha(colors[3], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = rhipp_positions[, 1],
        y = rhipp_positions[, 2],
        z = rhipp_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  rhipp_cross_section <- create_cross_section_matrix(
    list(rhipp_mat, rhipp_lpme_isomap_vals, rhipp_pme_vals),
    4,
    c(0.005, 0.005, 0.005),
    nrow = 2,
    lhipp = rhipp
  )
  ggsave(
    paste0(dir_path, "/adni_rhipp_cross_section.png"),
    plot = rhipp_cross_section,
    height = 5000,
    width = 7000,
    units = "px",
    dpi = 1250
  )

  lims <- c(-lthal_range, lthal_range)
  png(
    paste0(dir_path, "/adni_lthal_data_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_data <- lthal_mat[almost.equal(lthal_mat[, 1], time_vals[t]), ]

      scatter3D(
        x = temp_data[, 2],
        y = temp_data[, 3],
        z = temp_data[, 4],
        pch = 20,
        col = alpha(colors[1], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = lthal_positions[, 1],
        y = lthal_positions[, 2],
        z = lthal_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_lthal_lpme_isomap_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_lpme <- lthal_lpme_isomap_vals[almost.equal(lthal_lpme_isomap_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_lpme[, 2],
        y = temp_lpme[, 3],
        z = temp_lpme[, 4],
        pch = 15,
        col = alpha(colors[2], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = lthal_positions[, 1],
        y = lthal_positions[, 2],
        z = lthal_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_lthal_pme_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_pme <- lthal_pme_vals[almost.equal(lthal_pme_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_pme[, 2],
        y = temp_pme[, 3],
        z = temp_pme[, 4],
        pch = 17,
        col = alpha(colors[3], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = lthal_positions[, 1],
        y = lthal_positions[, 2],
        z = lthal_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  lthal_cross_section <- create_cross_section_matrix(
    list(lthal_mat, lthal_lpme_isomap_vals, lthal_pme_vals),
    4,
    c(0.005, 0.005, 0.005),
    nrow = 2,
    lhipp = lthal
  )
  ggsave(
    paste0(dir_path, "/adni_lthal_cross_section.png"),
    plot = lthal_cross_section,
    height = 5000,
    width = 7000,
    units = "px",
    dpi = 1250
  )

  lims <- c(-rthal_range, rthal_range)
  png(
    paste0(dir_path, "/adni_rthal_data_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_data <- rthal_mat[almost.equal(rthal_mat[, 1], time_vals[t]), ]

      scatter3D(
        x = temp_data[, 2],
        y = temp_data[, 3],
        z = temp_data[, 4],
        pch = 20,
        col = alpha(colors[1], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = rthal_positions[, 1],
        y = rthal_positions[, 2],
        z = rthal_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_rthal_lpme_isomap_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_lpme <- rthal_lpme_isomap_vals[almost.equal(rthal_lpme_isomap_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_lpme[, 2],
        y = temp_lpme[, 3],
        z = temp_lpme[, 4],
        pch = 15,
        col = alpha(colors[2], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = rthal_positions[, 1],
        y = rthal_positions[, 2],
        z = rthal_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  png(
    paste0(dir_path, "/adni_rthal_pme_plot.png"),
    res = 1000,
    height = 3500,
    width = 5000 * nplots
  )
  par(
    oma = c(0, 0, 1, 0), 
    mfrow = c(1, floor(length(time_vals) / 2)), 
    mar = c(0, 0, 1, 0), 
    mgp = c(5, 1, 0)
  )

  for (t in 1:length(time_vals)) {
    if (mod(t, 2) > 1e-10) {
      next
    } else {
      temp_pme <- rthal_pme_vals[almost.equal(rthal_pme_vals[, 1], time_vals[t]), ]
      scatter3D(
        x = temp_pme[, 2],
        y = temp_pme[, 3],
        z = temp_pme[, 4],
        pch = 17,
        col = alpha(colors[3], 0.3),
        theta = 220,
        ticktype = "detailed",
        nticks = 1,
        main = paste0("Time = ", round((time_vals)[t], 2)),
        xlim = lims,
        ylim = lims,
        zlim = lims,
        xlab = "",
        ylab = "",
        zlab = "",
        cex.axis = 0.67,
        cex = 0.75
      )
      text3D(
        x = rthal_positions[, 1],
        y = rthal_positions[, 2],
        z = rthal_positions[, 3],
        labels = plot_labels,
        add = TRUE
      )
    }
  }
  dev.off()

  rthal_cross_section <- create_cross_section_matrix(
    list(rthal_mat, rthal_lpme_isomap_vals, rthal_pme_vals),
    4,
    c(0.005, 0.005, 0.005),
    nrow = 2,
    lhipp = rthal
  )
  ggsave(
    paste0(dir_path, "/adni_rthal_cross_section.png"),
    plot = rthal_cross_section,
    height = 5000,
    width = 7000,
    units = "px",
    dpi = 1250
  )
}
```
```{r}
colors <- brewer.pal(3, "Set1")

roster <- read_csv(here("data/adni_standard/ROSTER.csv"))
roster <- roster %>% 
  select(
    -USERDATE,
    -USERDATE2,
    -update_stamp
  )
adni_status <- read_csv(here("data/adni_standard/ADNIMERGE.csv"))

# adni_status <- full_join(
#   adni_status,
#   roster,
#   by = c("Phase", "ID", "RID", "SITEID")
# )

sub_dirs <- result_dirs

hipp_info <- data.frame(
  patno = character(),
  date = numeric(),
  lhipp_data_vol = numeric(),
  lhipp_vol_lpme = numeric(),
  lhipp_vol_pme = numeric(),
  rhipp_data_vol = numeric(),
  rhipp_vol_lpme = numeric(),
  rhipp_vol_pme = numeric()
)

thal_info <- data.frame(
  patno = character(),
  date = numeric(),
  lthal_data_vol = numeric(),
  lthal_vol_lpme = numeric(),
  lthal_vol_pme = numeric(),
  rthal_data_vol = numeric(),
  rthal_vol_lpme = numeric(),
  rthal_vol_pme = numeric()
)

for (dir in sub_dirs) {
  file_list <- list.files(dir, full.names = FALSE)
  if ("hipp_info.csv" %in% file_list) {
    temp_hipp_info <- read_csv(paste0(dir, "/hipp_info.csv"))[, -1]
    hipp_info <- bind_rows(hipp_info, temp_hipp_info)
  }
  if ("thal_info.csv" %in% file_list) {
    temp_thal_info <- read_csv(paste0(dir, "/thal_info.csv"))[, -1]
    thal_info <- bind_rows(thal_info, temp_thal_info)
  }
}

hipp_bl <- hipp_info %>%
  group_by(patno) %>%
  summarize(
    date_bl = min(date),
    max_date = max(date)
  )
thal_bl <- thal_info %>%
  group_by(patno) %>%
  summarize(
    date_bl = min(date),
    max_date = max(date)
  )

eligible_patnos <- hipp_bl %>%
  filter(max_date - date_bl > 2) %>%
  select(patno) %>%
  unlist() %>%
  unique()

hipp_info <- full_join(hipp_info, hipp_bl, by = "patno") %>%
  mutate(time_from_bl = date - date_bl) %>%
  filter(patno %in% eligible_patnos)

thal_info <- full_join(thal_info, thal_bl, by = "patno") %>%
  mutate(time_from_bl = date - date_bl) %>%
  filter(patno %in% eligible_patnos)

hipp_summary <- hipp_info %>%
  group_by(patno) %>%
  summarize(
    lhipp_data_vol_sd = sd(lhipp_data_vol),
    lhipp_vol_lpme_sd = sd(lhipp_vol_lpme),
    lhipp_vol_pme_sd = sd(lhipp_vol_pme),
    rhipp_data_vol_sd = sd(rhipp_data_vol),
    rhipp_vol_lpme_sd = sd(rhipp_vol_lpme),
    rhipp_vol_pme_sd = sd(rhipp_vol_pme)
  )

hipp_long <- hipp_info %>%
  pivot_longer(
    lhipp_data_vol:lhipp_vol_pme,
    names_to = "lhipp_method",
    values_to = "lhipp_volume"
  ) %>%
  pivot_longer(
    rhipp_data_vol:rhipp_vol_pme,
    names_to = "rhipp_method",
    values_to = "rhipp_volume"
  ) %>%
  mutate(
    lhipp_method = gsub("lhipp_", "", lhipp_method),
    rhipp_method = gsub("rhipp_", "", rhipp_method)
  ) %>%
  filter(lhipp_method == rhipp_method) %>%
  select(-rhipp_method) %>%
  separate(lhipp_method, c("source", "method"), sep = "_") %>%
  mutate(
    source_mod = ifelse(source == "data", source, method),
    method_mod = ifelse(source == "vol", source, method)
  ) %>%
  select(-source, -method) %>%
  rename(source = source_mod, method = method_mod)

thal_long <- thal_info %>%
  pivot_longer(
    lthal_data_vol:lthal_vol_pme,
    names_to = "lthal_method",
    values_to = "lthal_volume"
  ) %>%
  pivot_longer(
    rthal_data_vol:rthal_vol_pme,
    names_to = "rthal_method",
    values_to = "rthal_volume"
  ) %>%
  mutate(
    lthal_method = gsub("lthal_", "", lthal_method),
    rthal_method = gsub("rthal_", "", rthal_method)
  ) %>%
  filter(lthal_method == rthal_method) %>%
  select(-rthal_method) %>%
  separate(lthal_method, c("source", "method"), sep = "_") %>%
  mutate(
    source_mod = ifelse(source == "data", source, method),
    method_mod = ifelse(source == "vol", source, method)
  ) %>%
  select(-source, -method) %>%
  rename(source = source_mod, method = method_mod)

thal_info_ts <- as_tsibble(
  thal_info,
  key = patno,
  index = time_from_bl,
  regular = FALSE
)

hipp_info_ts <- as_tsibble(
  hipp_info,
  key = patno,
  index = time_from_bl,
  regular = FALSE
)

hipp_info_ts %>%
  sample_n_keys(3) %>%
  ggplot(aes(x = time_from_bl, group = patno, color = patno)) +
  geom_line(aes(y = lhipp_data_vol), linetype = "solid") +
  geom_line(aes(y = lhipp_vol_lpme), linetype = "dashed") +
  geom_line(aes(y = lhipp_vol_pme), linetype = "dotted") +
  xlab("Time from Baseline Visit (Years)") +
  ylab("Estimated Left Hippocampus Volume")

lhipp_data_vol_pred <- vector()
lhipp_lpme_vol_pred <- vector()
lhipp_pme_vol_pred <- vector()
rhipp_data_vol_pred <- vector()
rhipp_lpme_vol_pred <- vector()
rhipp_pme_vol_pred <- vector()

for (patno_val in unique(hipp_info$patno)) {
  patno_data <- hipp_info %>%
    filter(patno == patno_val)
  lhipp_data_lm <- lm(lhipp_data_vol ~ date, data = patno_data)
  lhipp_data_pred <- predict(lhipp_data_lm, newdata = patno_data)
  lhipp_data_vol_pred <- c(lhipp_data_vol_pred, lhipp_data_pred)

  lhipp_lpme_lm <- lm(lhipp_vol_lpme ~ date, data = patno_data)
  lhipp_lpme_pred <- predict(lhipp_lpme_lm, newdata = patno_data)
  lhipp_lpme_vol_pred <- c(lhipp_lpme_vol_pred, lhipp_lpme_pred)

  lhipp_pme_lm <- lm(lhipp_vol_pme ~ date, data = patno_data)
  lhipp_pme_pred <- predict(lhipp_pme_lm, newdata = patno_data)
  lhipp_pme_vol_pred <- c(lhipp_pme_vol_pred, lhipp_pme_pred)

  rhipp_data_lm <- lm(rhipp_data_vol ~ date, data = patno_data)
  rhipp_data_pred <- predict(rhipp_data_lm, newdata = patno_data)
  rhipp_data_vol_pred <- c(rhipp_data_vol_pred, rhipp_data_pred)

  rhipp_lpme_lm <- lm(rhipp_vol_lpme ~ date, data = patno_data)
  rhipp_lpme_pred <- predict(rhipp_lpme_lm, newdata = patno_data)
  rhipp_lpme_vol_pred <- c(rhipp_lpme_vol_pred, rhipp_lpme_pred)

  rhipp_pme_lm <- lm(rhipp_vol_pme ~ date, data = patno_data)
  rhipp_pme_pred <- predict(rhipp_pme_lm, newdata = patno_data)
  rhipp_pme_vol_pred <- c(rhipp_pme_vol_pred, rhipp_pme_pred)
}

hipp_lm_pred <- tibble(
  lhipp_data_vol_pred,
  lhipp_lpme_vol_pred,
  lhipp_pme_vol_pred,
  rhipp_data_vol_pred,
  rhipp_lpme_vol_pred,
  rhipp_pme_vol_pred
)

hipp_info <- bind_cols(hipp_info, hipp_lm_pred)

hipp_info_sd <- hipp_info %>%
  group_by(patno) %>%
  summarize(
    lhipp_data_sd = sd(lhipp_data_vol),
    lhipp_lpme_sd = sd(lhipp_vol_lpme),
    lhipp_pme_sd = sd(lhipp_vol_pme),
    rhipp_data_sd = sd(rhipp_data_vol),
    rhipp_lpme_sd = sd(rhipp_vol_lpme),
    rhipp_pme_sd = sd(rhipp_vol_pme),
    lhipp_data_adj_sd = sd(lhipp_data_vol - lhipp_data_vol_pred),
    lhipp_lpme_adj_sd = sd(lhipp_vol_lpme - lhipp_lpme_vol_pred),
    lhipp_pme_adj_sd = sd(lhipp_vol_pme - lhipp_pme_vol_pred),
    rhipp_data_adj_sd = sd(rhipp_data_vol - rhipp_data_vol_pred),
    rhipp_lpme_adj_sd = sd(rhipp_vol_lpme - rhipp_lpme_vol_pred),
    rhipp_pme_adj_sd = sd(rhipp_vol_pme - rhipp_pme_vol_pred)
  ) %>%
  ungroup()

hipp_sd_mean <- hipp_info_sd %>%
  summarize(
    lhipp_data_sd_mean = mean(lhipp_data_sd),
    lhipp_lpme_sd_mean = mean(lhipp_lpme_sd),
    lhipp_pme_sd_mean = mean(lhipp_pme_sd),
    rhipp_data_sd_mean = mean(rhipp_data_sd),
    rhipp_lpme_sd_mean = mean(rhipp_lpme_sd),
    rhipp_pme_sd_mean = mean(rhipp_pme_sd),
    lhipp_data_adj_sd_mean = mean(lhipp_data_adj_sd),
    lhipp_lpme_adj_sd_mean = mean(lhipp_lpme_adj_sd),
    lhipp_pme_adj_sd_mean = mean(lhipp_pme_adj_sd),
    rhipp_data_adj_sd_mean = mean(rhipp_data_adj_sd),
    rhipp_lpme_adj_sd_mean = mean(rhipp_lpme_adj_sd),
    rhipp_pme_adj_sd_mean = mean(rhipp_pme_adj_sd)
)

# print(hipp_sd_mean, width = Inf)

hipp_sd_med <- hipp_info_sd %>%
  summarize(
    lhipp_data_sd_med = median(lhipp_data_sd),
    lhipp_lpme_sd_med = median(lhipp_lpme_sd),
    lhipp_pme_sd_med = median(lhipp_pme_sd),
    rhipp_data_sd_med = median(rhipp_data_sd),
    rhipp_lpme_sd_med = median(rhipp_lpme_sd),
    rhipp_pme_sd_med = median(rhipp_pme_sd),
    lhipp_data_adj_sd_med = median(lhipp_data_adj_sd),
    lhipp_lpme_adj_sd_med = median(lhipp_lpme_adj_sd),
    lhipp_pme_adj_sd_med = median(lhipp_pme_adj_sd),
    rhipp_data_adj_sd_med = median(rhipp_data_adj_sd),
    rhipp_lpme_adj_sd_med = median(rhipp_lpme_adj_sd),
    rhipp_pme_adj_sd_med = median(rhipp_pme_adj_sd)
  )

print(hipp_sd_med, width = Inf)

thal_info_ts %>%
  sample_n_keys(3) %>%
  ggplot(aes(x = time_from_bl, group = patno, color = patno)) +
  geom_line(aes(y = lthal_data_vol), linetype = "solid") +
  geom_line(aes(y = lthal_vol_lpme), linetype = "dashed") +
  geom_line(aes(y = lthal_vol_pme), linetype = "dotted") +
  xlab("Time from Baseline Visit (Years)") +
  ylab("Estimated Left Thalamus Volume")

lthal_data_vol_pred <- vector()
lthal_lpme_vol_pred <- vector()
lthal_pme_vol_pred <- vector()
rthal_data_vol_pred <- vector()
rthal_lpme_vol_pred <- vector()
rthal_pme_vol_pred <- vector()

for (patno_val in unique(thal_info$patno)) {
  patno_data <- thal_info %>%
    filter(patno == patno_val)
  lthal_data_lm <- lm(lthal_data_vol ~ date, data = patno_data)
  lthal_data_pred <- predict(lthal_data_lm, newdata = patno_data)
  lthal_data_vol_pred <- c(lthal_data_vol_pred, lthal_data_pred)

  lthal_lpme_lm <- lm(lthal_vol_lpme ~ date, data = patno_data)
  lthal_lpme_pred <- predict(lthal_lpme_lm, newdata = patno_data)
  lthal_lpme_vol_pred <- c(lthal_lpme_vol_pred, lthal_lpme_pred)

  lthal_pme_lm <- lm(lthal_vol_pme ~ date, data = patno_data)
  lthal_pme_pred <- predict(lthal_pme_lm, newdata = patno_data)
  lthal_pme_vol_pred <- c(lthal_pme_vol_pred, lthal_pme_pred)

  rthal_data_lm <- lm(rthal_data_vol ~ date, data = patno_data)
  rthal_data_pred <- predict(rthal_data_lm, newdata = patno_data)
  rthal_data_vol_pred <- c(rthal_data_vol_pred, rthal_data_pred)

  rthal_lpme_lm <- lm(rthal_vol_lpme ~ date, data = patno_data)
  rthal_lpme_pred <- predict(rthal_lpme_lm, newdata = patno_data)
  rthal_lpme_vol_pred <- c(rthal_lpme_vol_pred, rthal_lpme_pred)

  rthal_pme_lm <- lm(rthal_vol_pme ~ date, data = patno_data)
  rthal_pme_pred <- predict(rthal_pme_lm, newdata = patno_data)
  rthal_pme_vol_pred <- c(rthal_pme_vol_pred, rthal_pme_pred)
}

thal_lm_pred <- tibble(
  lthal_data_vol_pred,
  lthal_lpme_vol_pred,
  lthal_pme_vol_pred,
  rthal_data_vol_pred,
  rthal_lpme_vol_pred,
  rthal_pme_vol_pred
)

thal_info <- bind_cols(thal_info, thal_lm_pred)

thal_info_sd <- thal_info %>%
  group_by(patno) %>%
  summarize(
    lthal_data_sd = sd(lthal_data_vol),
    lthal_lpme_sd = sd(lthal_vol_lpme),
    lthal_pme_sd = sd(lthal_vol_pme),
    rthal_data_sd = sd(rthal_data_vol),
    rthal_lpme_sd = sd(rthal_vol_lpme),
    rthal_pme_sd = sd(rthal_vol_pme),
    lthal_data_adj_sd = sd(lthal_data_vol - lthal_data_vol_pred),
    lthal_lpme_adj_sd = sd(lthal_vol_lpme - lthal_lpme_vol_pred),
    lthal_pme_adj_sd = sd(lthal_vol_pme - lthal_pme_vol_pred),
    rthal_data_adj_sd = sd(rthal_data_vol - rthal_data_vol_pred),
    rthal_lpme_adj_sd = sd(rthal_vol_lpme - rthal_lpme_vol_pred),
    rthal_pme_adj_sd = sd(rthal_vol_pme - rthal_pme_vol_pred)
  ) %>%
  ungroup()

thal_sd_mean <- thal_info_sd %>%
  summarize(
    lthal_data_sd_mean = mean(lthal_data_sd),
    lthal_lpme_sd_mean = mean(lthal_lpme_sd),
    lthal_pme_sd_mean = mean(lthal_pme_sd),
    rthal_data_sd_mean = mean(rthal_data_sd),
    rthal_lpme_sd_mean = mean(rthal_lpme_sd),
    rthal_pme_sd_mean = mean(rthal_pme_sd),
    lthal_data_adj_sd_mean = mean(lthal_data_adj_sd),
    lthal_lpme_adj_sd_mean = mean(lthal_lpme_adj_sd),
    lthal_pme_adj_sd_mean = mean(lthal_pme_adj_sd),
    rthal_data_adj_sd_mean = mean(rthal_data_adj_sd),
    rthal_lpme_adj_sd_mean = mean(rthal_lpme_adj_sd),
    rthal_pme_adj_sd_mean = mean(rthal_pme_adj_sd)
  )

# print(thal_sd_mean, width = Inf)

thal_sd_med <- thal_info_sd %>%
  summarize(
    lthal_data_sd_med = median(lthal_data_sd),
    lthal_lpme_sd_med = median(lthal_lpme_sd),
    lthal_pme_sd_med = median(lthal_pme_sd),
    rthal_data_sd_med = median(rthal_data_sd),
    rthal_lpme_sd_med = median(rthal_lpme_sd),
    rthal_pme_sd_med = median(rthal_pme_sd),
    lthal_data_adj_sd_med = median(lthal_data_adj_sd),
    lthal_lpme_adj_sd_med = median(lthal_lpme_adj_sd),
    lthal_pme_adj_sd_med = median(lthal_pme_adj_sd),
    rthal_data_adj_sd_med = median(rthal_data_adj_sd),
    rthal_lpme_adj_sd_med = median(rthal_lpme_adj_sd),
    rthal_pme_adj_sd_med = median(rthal_pme_adj_sd)
  )

print(thal_sd_med, width = Inf)

set.seed(8310)
sampled_patnos <- hipp_info_ts %>% 
  sample_n_keys(3) %>% 
  .$patno %>% 
  unique()

scale_factor <- 3
hipp_info_ts %>%
  filter(patno %in% sampled_patnos) %>%
  ggplot(aes(x = time_from_bl, group = patno)) +
  geom_point(aes(y = lhipp_data_vol), shape = 16, color = colors[1]) +
  geom_line(aes(y = lhipp_data_vol), linetype = "solid", color = colors[1]) +
  # geom_smooth(
  #   aes(y = lhipp_data_vol2),
  #   color = colors[1],
  #   method = "lm",
  #   se = FALSE,
  #   linetype = "dashed",
  #   linewidth = 1
  # ) +
  geom_point(aes(y = lhipp_vol_lpme), shape = 15, color = colors[2]) +
  geom_line(aes(y = lhipp_vol_lpme), linetype = "dashed", color = colors[2]) +
  # geom_smooth(
  #   aes(y = lhipp_vol_lpme2),
  #   color = colors[2],
  #   method = "lm",
  #   se = FALSE,
  #   linetype = "dashed",
  #   linewidth = 1
  # ) +
  geom_point(aes(y = lhipp_vol_pme), shape = 17, color = colors[3]) +
  geom_line(aes(y = lhipp_vol_pme), linetype = "dotted", color = colors[3]) +
  # geom_smooth(
  #   aes(y = lhipp_vol_pme2),
  #   color = colors[3],
  #   method = "lm",
  #   se = FALSE,
  #   linetype = "dashed",
  #   linewidth = 1
  # ) +
  facet_wrap(~patno) +
  xlab("Time from Baseline Visit (Years)") +
  ylab("Estimated Left Hippocampus Volume") + 
  theme(
    axis.title.x = element_text(size = scale_factor * 5),
    axis.title.y = element_text(size = scale_factor * 5),
    axis.text.x = element_text(size = scale_factor * 5),
    axis.text.y = element_text(size = scale_factor * 5),
    strip.text.x = element_text(size = scale_factor * 5)
  )
ggsave(here("paper/figures/adni_plots/adni_lhipp_volume_comp.png"), dpi = 1500)

thal_info_ts %>%
  filter(patno %in% sampled_patnos) %>% 
  ggplot(aes(x = time_from_bl, group = patno)) +
  geom_point(aes(y = lthal_data_vol), shape = 16, color = colors[1]) +
  geom_line(aes(y = lthal_data_vol), linetype = "solid", color = colors[1]) +
  # geom_smooth(
  #   aes(y = lthal_data_vol2),
  #   color = colors[1],
  #   method = "lm",
  #   se = FALSE,
  #   linetype = "dashed",
  #   linewidth = 1
  # ) +
  geom_point(aes(y = lthal_vol_lpme), shape = 15, color = colors[2]) +
  geom_line(aes(y = lthal_vol_lpme), linetype = "dashed", color = colors[2]) +
  # geom_smooth(
  #   aes(y = lthal_vol_lpme2),
  #   color = colors[2],
  #   method = "lm",
  #   se = FALSE,
  #   linetype = "dashed",
  #   linewidth = 1
  # ) +
  geom_point(aes(y = lthal_vol_pme), shape = 17, color = colors[3]) +
  geom_line(aes(y = lthal_vol_pme), linetype = "dotted", color = colors[3]) +
  # geom_smooth(
  #   aes(y = lthal_vol_pme2),
  #   color = colors[3],
  #   method = "lm",
  #   se = FALSE,
  #   linetype = "dashed",
  #   linewidth = 1
  # ) +
  facet_wrap(~patno) +
  xlab("Time from Baseline Visit (Years)") +
  ylab("Estimated Left Thalamus Volume") + 
  guides(
    color = guide_legend(title = "Participant ID"),
    shape = guide_legend(title = "Estimate Source")
  ) +
  theme(
    axis.title.x = element_text(size = scale_factor * 5),
    axis.title.y = element_text(size = scale_factor * 5),
    axis.text.x = element_text(size = scale_factor * 5),
    axis.text.y = element_text(size = scale_factor * 5),
    strip.text.x = element_text(size = scale_factor * 5)
  )
ggsave(here("paper/figures/adni_plots/adni_lthal_volume_comp.png"), dpi = 1500)

adni_status %>% 
  filter(
    PTID %in% eligible_patnos,
    VISCODE == "bl"
  ) %>% 
  group_by(DX_bl) %>% 
  tally()
```

### 2024-07-10 code/functions/sim_data.R
update case numbers

@@ -25,22 +25,15 @@ sim_data <- function(time_val, case, noise, amp_noise, period_noise, time_change
         amp_noise[1] * cos(period_noise[1] * tau),
         amp_noise[2] * sin(period_noise[2] * tau)
       ) +
-      time_change_d2
-    },
-    function(tau, amp_noise, period_noise) {
-      c(
-        amp_noise[1] * cos(period_noise[1] * tau),
-        amp_noise[2] * sin(period_noise[2] * tau)
-      ) +
-      time_change_d2
+        time_change_d2
     },
     function(tau, amp_noise, period_noise) {
       c(
         tau,
-        (tau * amp_noise[1] + period_noise[1]) ^ 2,
-        (tau * amp_noise[2] + period_noise[2]) ^ 3
+        (tau * amp_noise[1] + period_noise[1])^2,
+        (tau * amp_noise[2] + period_noise[2])^3
       ) +
-      time_change_d3
+        time_change_d3
     },
     function(tau, amp_noise, period_noise) {
       c(
@@ -48,15 +41,15 @@ sim_data <- function(time_val, case, noise, amp_noise, period_noise, time_change
         amp_noise[1] * cos(period_noise[1] * tau),
         amp_noise[2] * sin(period_noise[2] * tau)
       ) +
-      time_change_d3
+        time_change_d3
     },
     function(tau, amp_noise, period_noise) {
       c(
         (period_noise[1] / 2) * tau[1],
         (period_noise[2] / 2) * tau[2],
-        (amp_noise[1]) * ((amp_noise[2]) * norm_euclidean(period_noise[1:d] * tau) ^ 2)
+        (amp_noise[1]) * ((amp_noise[2]) * norm_euclidean(period_noise[1:d] * tau)^2)
       ) +
-      time_change_d3
+        time_change_d3
     },
     function(tau, amp_noise, period_noise) {
       c(
@@ -64,22 +57,13 @@ sim_data <- function(time_val, case, noise, amp_noise, period_noise, time_change
         (period_noise[2] * amp_noise[2] * tau[1]) * sin(amp_noise[2] * tau[1]),
         tau[2]
       ) +
-      time_change_d3
+        time_change_d3
     },
[…]

### 2024-07-10 paper/main.tex
Updates from Overleaf

@@
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
-This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curve in the structure at the upper-left corner of each plot.
+This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curves in the structure near the top and bottom corners of each plot.
 Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface.
@@
 Considering the same individual shown in Figure \ref{fig:lhipp_cross_sections}, cross sectional results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}.
-The cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
+The cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus, though the LPME estimate still falls inside the observed surface.
 
@@
 raw data, LPME, and PME).
-We then took the mean of these values for each estimate source.
-Table \ref{table:adni_volume_sds} shows that PME yields the smoothest volume estimates for both hippocampuses.
-This conclusion aligns with the visual evidence of inconsistency in the LPME estimates caused by the irregular shape of the hippocampus.
-However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.
+We then took the median of these values for each estimate source.
+Table \ref{table:adni_volume_sds} shows that, despite the visual evidence of inconsistency in the LPME estimates of the hippocampus caused by its irregular shape, LPME still yields the smoothest volume estimates for both hippocampuses.
+LPME also results in the smoothest volume estimates when applied to the thalamus, showing even greater improvements over the alternative volume estimates.
 
@@
     \hline
-    Left Hippocampus & No & 129 & 132 & \textbf{106} \\
-    Right Hippocampus & No & 137 & 136 & \textbf{110} \\
-    Left Hippocampus & Yes & 100 & 108 & \textbf{87.1} \\
-    Right Hippocampus & Yes & 108 & 121 & \textbf{85.9} \\
-    Left Thalamus & No & 57.5 & \textbf{31.3} & 42.0 \\
-    Right Thalamus & No & 108 & \textbf{66.4} & 85.3 \\
-    Left Thalamus & Yes & 45.8 & \textbf{20.6} & 35.4 \\
-    Right Thalamus & Yes & 89.3 & \textbf{50.1} & 73.1 \\
+    Left Hippocampus & No & 129 & \textbf{119} & 124 \\
+    Right Hippocampus & No & 133 & \textbf{124} & 179 \\
+    Left Hippocampus & Yes & 104 & \textbf{93.2} & 98.4 \\
+    Right Hippocampus & Yes & 107 & \textbf{100} & 149 \\
+    Left Thalamus & No & 55.7 & \textbf{28.5} & 42.5 \\
+    Right Thalamus & No & 108 & \textbf{61.7} & 93.3 \\
+    Left Thalamus & Yes & 45.2 & \textbf{16.0} & 35.3 \\
+    Right Thalamus & Yes & 86.9 & \textbf{42.7} & 76.6 \\
     \hline
   \end{tabular}
-  \caption{{\footnotesize Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.
+  \caption{{\footnotesize Median of the empirical standard deviation of volume estimates computed for each ADNI participant.
 Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.} }

### 2024-07-10 paper/supplement.tex
Updates from Overleaf

@@
 
+\renewcommand\thefigure{S\arabic{figure}}
+\renewcommand\thetable{S\arabic{table}}
+
+\setcounter{figure}{0}
+\setcounter{table}{0}
+
 \begin{table}[h]
@@
 \begin{figure}
+
+  \begin{subfigure}{\textwidth}
+    \label{fig:adni_lhipp_result}
+    \begin{subfigure}{\textwidth}
       \centering
-\subfloat{
-\label{fig:adni_lhipp_result}
-\subfloat[\centering Data]{
-\includegraphics[height=10cm]{adni_plots/adni_lhipp_data_plot}
-}
-\hfill
-\subfloat[\centering LPME]{
-\includegraphics[height=10cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
-}
-\hfill
-\subfloat[\centering PME]{
-\includegraphics[height=10cm]{adni_plots/adni_lhipp_pme_plot}
-}
-}
-\hfill
-\subfloat{
+      \includegraphics[height=3cm]{adni_plots/adni_lhipp_data_plot}
+      \caption{Data}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[height=3cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
+      \caption{LPME}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[height=3cm]{adni_plots/adni_lhipp_pme_plot}
+      \caption{PME}
+    \end{subfigure}   
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
     \label{fig:adni_lthal_result}
-\subfloat[\centering Data]{
-\includegraphics[height=10cm]{adni_plots/adni_lthal_data_plot}
-}
-\hfill
-\subfloat[\centering LPME]{
-\includegraphics[height=10cm]{adni_plots/adni_lthal_lpme_isomap_plot}
-}
-\hfill
-\subfloat[\centering PME]{
-\includegraphics[height=10cm]{adni_plots/adni_lthal_pme_plot}
-}
-}
-\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant.
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[height=3cm]{adni_plots/adni_lthal_data_plot}
+      \caption{Data}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[height=3cm]{adni_plots/adni_lthal_lpme_isomap_plot}
+      \caption{LPME}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[height=3cm]{adni_plots/adni_lthal_pme_plot}
+      \caption{PME}
+    \end{subfigure}   
+  \end{subfigure}
+
+  \caption{Left Hippocampus (panels a-c), and Left Thalamus (panels d-f), Cognitive Healthy ADNI Participant.
 The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.

### 2024-07-13 paper/main.tex
Updates from Overleaf

@@
     Department of Biostatistics, Brown University\\
-    for the Alzheimer's Disease Neuroimaging Initiative}
+    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
+As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
+A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
   \maketitle
@@
 However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies.
-That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.
+These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.
 %We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.
 
-Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
+Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional PCA method proposed by \cite{greven2011longitudinal}.
 As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space.
@@
 However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.
 
@@
 We assume that a data cloud is observed in the $D$ dimensional space.
-In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure 1 in the online supplement.
+In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure S1 in the online supplement).
 For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
@@
 
-It is important to note that, unlike manifold learning methods that use an eigendecomposition to directly estimate coordinates in the $d$-dimensional manifold space (e.g.
-Isomap, locally linear embedding, and others) with only an implicit representation of the embedding function $f$, principal manifolds may be estimated via an explicit approximation of the embedding function $f$, from which the projection index $\pi_f(x)$ and the desired coordinates on the manifold can be found.
+It is important to note that, unlike manifold learning methods that use an eigendecomposition to directly estimate coordinates in the $d$-dimensional manifold space with only an implicit representation of the embedding function $f$ (e.g.
+Isomap, locally linear embedding, and others), principal manifolds may be estimated via an explicit approximation of the embedding function $f$, from which the projection index $\pi_f(x)$ and the desired coordinates on the manifold can be found.
 In this way, regression approaches become relevant to the manifold estimation problem.
@@
 Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually.
-In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.} }
+In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the diamond-shaped points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the surface in (d) shows the estimated smooth surface.} }
   \label{fig:lpme_steps}
@@
 1]{mengPrincipalManifoldEstimation2021} to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
-Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
+Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using diamond-shaped points in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
 The number $N_t$ of clusters is estimated for each time point $t$ and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure.
@@
 Specifically, the cluster centers $\mu_{j, t}$ and weights $\hat{\theta}_{j, t}$ replace those obtained using the HDMDE algorithm, and the Isomap-derived parameters obtained in Steps 1 and 2 of Algorithm 2 in \cite{mengPrincipalManifoldEstimation2021}, respectively.
-This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using multicolored lines in Figure \ref{fig:lpme_step3}.
+This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using lines in Figure \ref{fig:lpme_step3}.
 The performance of each fit $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I_t}\sum_{i=1}^{I_t}\| x_{it} - \widehat{f_t}(\pi_{\widehat{f_t}}(x_{it}))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the fitted manifold $\widehat{f_t}$.
@@
 
-A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{mengPrincipalManifoldEstimation2021}.
+A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results, preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{mengPrincipalManifoldEstimation2021}.
 To avoid this process, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
@@
 
-In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
+In the situations where $d = 1$, LPME is compared to the PME method naïvely run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
 The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package \citep{Cannoodt2018princurve}, developed using \texttt{R} \citep{rSoftware2023}.
 In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
-Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$.
-In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
+Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$.
+The minimum number of cluster centers is set to $10 \times d$.
+In cases where $d = 2$, LPME is again compared to the naïve PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
 
@@
 Visualizations of the results from one example simulated case, truncated for concision, are shown in Figure \ref{fig:sim_case1}.
-Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time.
+Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold over time.
 Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.
@@
   \caption{{\footnotesize Simulation Case 1.
-Scatterplots of one example simulated data cloud (gray points) with $d=1$ overlaid with points on the true embedding manifold (red), the estimated LMPE function (blue), PME (green), and principal curve or surface (purple).}}
+Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).}}
   \label{fig:sim_case1}
@@
 The median and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_median}.
-Analogous mean and standard deviation summaries of the results are given in the online supplement.
+Analogous mean and standard deviation summaries of the results are given in Table S1 of the online supplement.
 Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data.
@@
 The result summaries indicate that in most cases, LPME provides a substantial improvement in performance over the PME and principal curve approaches when estimating the underlying manifold.
-As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates, shown in blue, remain relatively stable over time.
-This contrasts to the estimates found by the PME and principal curve approaches, which are highly sensitive to the systematic added noise in the data observed at each given time point, as expected.
-This ultimately results in the LPME estimates remaining closer to the true underlying manifold, shown in red.
+As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates remain relatively stable over time.
+This contrasts to the estimates found by the PME and principal curve approaches, which, as expected, are highly sensitive to the added systematic noise in the data observed at each given time point.
+This ultimately results in the LPME estimates remaining closer to the true underlying manifold.
 Similar results hold in most of the other simulation cases.
@@
 
-Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure 1 in the online supplement.
+Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure S1 in the online supplement.
 The previously described data augmentation approach was used to enable the fitting of closed surfaces.
@@
 These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points.
-This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curves in the structure near the top and bottom corners of each plot.
+This Figure reiterates that both algorithms struggle to capture the distinctive sharp curves in the structure near the top and bottom corners of each plot.
 Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface.
@@
 
-The volume estimates for randomly selected individuals can be found in Figures 2a and 2b in the online supplement, corresponding to estimates for the left hippocampus and left thalamus, respectively.
+The volume estimates for randomly selected individuals can be found in panels (a) and (b) of Figure S2 in the online supplement, corresponding to estimates for the left hippocampus and left thalamus, respectively.
 Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.
-In the supplement's Figure 2a, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
-All three sources show substantial variability between study visits, particularly for subjects \texttt{011\_S\_0003} and \texttt{036\_S\_1240}.
-Considering the thalamus estimates in the supplement's Figure 2b, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
+In panel (a) of Figure S2, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates.
+All three sources show substantial variability between study visits, particularly for subjects \texttt{033\_S\_0514} and \texttt{099\_S\_0291}.
+Considering the thalamus estimates in panel (b), we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points.
 However, the LPME estimates demonstrate the ability to smooth over this variability, resulting in volume estimates that are not as sensitive to differences between time points.
 We also see that when basing volume estimates on the more reliable LPME fit to the thalamus data, LPME yields volume estimates similar in magnitude to those obtained from the raw data and PME estimates.
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c c c c c|}
+    \hline
+    Structure & Regression-Adjusted & Data & LPME & PME  \\
+    \hline
+    Left Hippocampus & No & 129 & \textbf{119} & 124 \\
+    Right Hippocampus & No & 133 & \textbf{124} & 179 \\
+    Left Hippocampus & Yes & 104 & \textbf{93.2} & 98.4 \\
+    Right Hippocampus & Yes & 107 & \textbf{100} & 149 \\
+    Left Thalamus & No & 55.7 & \textbf{28.5} & 42.5 \\
+    Right Thalamus & No & 108 & \textbf{61.7} & 93.3 \\
+    Left Thalamus & Yes & 45.2 & \textbf{16.0} & 35.3 \\
+    Right Thalamus & Yes & 86.9 & \textbf{42.7} & 76.6 \\
+    \hline
+  \end{tabular}
+  \caption{{\footnotesize Median of the empirical standard deviation of volume estimates computed for each ADNI participant.
+Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.} }
+  \label{table:adni_volume_sds}
+\end{table}
 
@@
 
-\begin{table}[ht]
-  \centering
-  \begin{tabular}{|c c c c c|}
-    \hline
-    Structure & Regression-Adjusted & Data & LPME & PME  \\
-    \hline
-    Left Hippocampus & No & 129 & \textbf{119} & 124 \\
-    Right Hippocampus & No & 133 & \textbf{124} & 179 \\
-    Left Hippocampus & Yes & 104 & \textbf{93.2} & 98.4 \\
-    Right Hippocampus & Yes & 107 & \textbf{100} & 149 \\
-    Left Thalamus & No & 55.7 & \textbf{28.5} & 42.5 \\
-    Right Thalamus & No & 108 & \textbf{61.7} & 93.3 \\
-    Left Thalamus & Yes & 45.2 & \textbf{16.0} & 35.3 \\
-    Right Thalamus & Yes & 86.9 & \textbf{42.7} & 76.6 \\
-    \hline
-  \end{tabular}
-  \caption{{\footnotesize Median of the empirical standard deviation of volume estimates computed for each ADNI participant.
-Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.} }
-  \label{table:adni_volume_sds}
-\end{table}
-
 \section{Discussion}\label{s:discussion}
@@
 In this article, we propose the LPME framework for modeling longitudinal changes in a low-dimensional manifold underlying high-dimensional data.
-To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a general-purpose longitudinal setting.
+This work to adapt a nonlinear manifold learning method to a general-purpose longitudinal setting expands the situations in which such methods may be applicable.
 We also suggest a data augmentation approach that circumvents challenges in fitting principal manifold-based methods to self-intersecting manifolds in select settings.
-In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve/surface algorithms at each time point.
+In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance in recovering the underlying manifold when compared to a naïve approach of applying either the PME or principal curve/surface algorithms at each time point.
 
@@
 While the LPME algorithm demonstrated strong performance in a number of settings, it is important to note the assumptions under which the method performs well.
-First, because Isomap and PME are used for initialization and time-specific fitting, it is necessary that the assumptions required by these algorithms hold for the LPME algorithm to perform well.
+First, because Isomap and PME are used for initialization and time-specific fitting, the assumptions required by these algorithms are also needed for the LPME algorithm to perform well.
 In practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
-In such cases, the use of alternative methods, such as Locally Linear Embedding \citep{roweisNonlinearDimensionalityReduction2000} or Laplacian Eigenmaps \citep{belkin2003laplacian}, to generate the initial parameterization may alleviate this concern.
+The use of alternative methods, such as Locally Linear Embedding \citep{roweisNonlinearDimensionalityReduction2000} or Laplacian Eigenmaps \citep{belkin2003laplacian}, to generate the initial parameterization may alleviate this concern.
 A second assumption is that the LPME method requires relative stability in the overall shape of the underlying manifold structure.
-Thus, while this algorithm is useful for separating meaningful changes in a manifold over time from inter-observation error, it may not perform well in situations where the error between observations is too substantial, similar to any smoothing method.
+Thus, while this algorithm is useful for separating meaningful changes in a manifold over time from inter-observation error, it may not perform well when the error between observations is too substantial, similar to any smoothing method.
 How restrictive this requirement is depends largely on the area of application.
@@
 While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, extension of this method to manifold estimation for groups of similar structures will be a valuable contribution to the literature allowing population-level comparisons of time-dependent trajectories of shapes.
-Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.
+Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which could prove beneficial in a number of clinical trial settings.
 
@@
 %{
-%\section*{Acknowledgment}
+%\section*{Acknowledgments}
 
-%\section*{Funding}
 %Research presented in this manuscript was funded by National Institute On Aging of the National Institutes of Health (NIH) under Award Number R01AG075511.
 The ideas presented in this paper are the responsibility of the authors and may not necessarily represent the official views of the NIH.
+
+%Data collection and sharing for the Alzheimer's Disease Neuroimaging Initiative (ADNI) is funded by the National Institute on Aging (National Institutes of Health Grant U19 AG024904).
+The grantee organization is the Northern California Institute for Research and Education.
+In the past, ADNI has also received funding from the National Institute of Biomedical Imaging and Bioengineering, the Canadian Institutes of Health Research, and private sector contributions through the Foundation for the National Institutes of Health (FNIH) including generous contributions from the following: AbbVie, Alzheimer’s Association; Alzheimer’s Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and Company; EuroImmun; F.
+Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research & Development, LLC.; Johnson & Johnson Pharmaceutical Research &Development LLC.; Lumosity; Lundbeck; Merck & Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
 %} \fi

### 2024-07-26 jasa/main.tex
add jasa submission directory

--- 2024-07-13 paper/main.tex
@@
 \usepackage{color}
-%\usepackage{csquotes}
 \usepackage{float}
@@
 \addtolength{\topmargin}{-1in}%
+
+\pdfminorversion=4
 
@@
 An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
-%\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
 \end{abstract}
@@
 \section{Introduction}
-
-%Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}.
-Frequently, interest centers around longitudinal changes in one or more neurological substructures.
-These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}.
-For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes.
-Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time.
-Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard.
-However, this approach is highly time and resource intensive.
-When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
-Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability \citep{boccardiSurveyProtocolsManual2011}.
-Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}.
-While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
 
@@
 
-%\subsection{A Motivating Question from Medical Imaging}
-
-%Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
-In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
-After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
-If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
-%\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models.
-%In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging \citep{scahill2003longitudinal}.
-%To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.
-A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.
-
 Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
@@
 To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.
-
-%\subsection{Background and Our Contributions}
-
-%Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}.
-Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing \citep{shen2022robust}, single-cell biology \citep{ding2023learning}, and robotics \citep{gao2024bi}.
-In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
-For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.
-Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.
 
@@
 These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.
-%We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.
 
@@
 
-%In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time.
-Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension.
-Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
-Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}).
-Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit.
-We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
-We demonstrate that the regularization involved in this process mitigates the effects of variability between time points.
-Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
-
 In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time.
@@
 Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
-
-%Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points.
-These methods consider time as an additional dimension in the data, including time in the dimension reduction process.
-Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension.
-Here, we propose an alternative approach to estimating manifolds over time.
-Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension.
-Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
-Unlike most commonly used manifold learning methods, this approach results in an explicit functional representation of the manifold estimate.
-We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.
-We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.
-
-%Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis (PCA) method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}.
-\cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
-\cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
-However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
-
-%Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.
 
@@
 \section{Principal Manifold Estimation Algorithm}\label{s:PME}
-
-%The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud.
-However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings.
-Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.
-In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper.
-We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively.
-We assume that a data cloud is observed in the $D$ dimensional space.
-In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure \ref{fig:adni_result}).
-For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$.
-Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$.
-Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$.
-The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$.
-Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
-The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection.
-The principal manifolds are then defined as follows.
 
@@
 
-%The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time.
-In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$.
-In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.
-
-%Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$.
-Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points.
-As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise.
-In this setting, the manifold to estimate is represented by $(t,\mathbf{r})\mapsto F(t,\mathbf{r})\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
-
 \subsection{The Longitudinal Principal Manifold Framework}\label{section: The Longitudinal Principal Manifold Framework}
@@
 Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}\if1\blind{, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}}\fi.
-
-%At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping.
-Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space.
-For example, the PME algorithm can be used to obtain this parameterization.
-In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, smoothing over the spatial spline coefficients with respect to time.
-This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set.
-Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning.
-A visualization of these steps is shown in Figure \ref{fig:lpme_steps}.
-Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}.
 
@@
 This approach is based on the assumption that the underlying manifolds of interest remains relatively stable over time.
-%The results given in Section \ref{s:simulations} indicate that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.
 
@@
 \end{equation}
-
-%Depending on the locations of spline knots  across time points, this function may take the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}.
-However, instead of attempting to estimate each coefficient function individually, we consider all coefficients together as the output of a single smoothing spline function.
-The initial use of smoothing splines to estimate principal manifolds complicates the direct use of estimation approaches developed for this model class in our use case.
 
@@
   \centering
-  %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
-  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_plots/adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
+  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
   \hfill
-  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_plots/adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}
+  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}
 
   \caption{{\footnotesize Left Hippocampus and Left Thalamus Cross Sections.
-LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface.
-In the fifth hippocampus observation with $t=6.55$, the LPME-estimated surface has a gap.
+LPME-approximated embedding function routinely falls inside the boundaries of the observed hippocampus surface.
+In the fifth hippocampus observation with $t=2.11$, the LPME-estimated surface has a gap.
 Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.
@@
 
-%\begin{figure}[h]
-%\centering
-%\subfloat{
-%\label{fig:adni_lhipp_result}
-%\begin{subfigure}
-%\includegraphics{adni_plots/adni_lhipp_data_plot}
-%\caption{Data}
-%\label{fig:adni_lhipp_result_data}
-%\end{subfigure}
-%}
-
-% \hfill
-%\subfloat{
-%\label{fig:adni_lthal_result}
-%\begin{subfigure}
-%\includegraphics{adni_plots/adni_lthal_data_plot}
-%\caption{Data}
-%\label{fig:adni_lthal_result_data}
-%\end{subfigure}
-%}
-
-%\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant.
-The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue.
-The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
-%\label{fig:adni_result}
-
-%\end{figure}
-
 \section{Discussion}\label{s:discussion}
@@
 In simulated datasets where spurious changes in the structure to be estimated are introduced between time points, the LPME algorithm shows improved performance in recovering the underlying manifold when compared to a naïve approach of applying either the PME or principal curve/surface algorithms at each time point.
-
-%Similar to other applications of statistical smoothing techniques, it is difficult to compare performance of models in real data experiments.
-Specifically, the MSD provides a summary of the distance of the data points and their projections onto the estimated manifolds for each model.
-This means that a model that performs well using this metric may not actually be the best fit for the underlying manifold, e.g.
-in the case of overfitting.
-This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are estimating a curve or surface using the data available at a given time point, which may be affected by systematic noise that increases the distance to the true manifold.
-When comparing the fit of the PME and principal surface and principal curve methods in the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.
 
@@
 
-%\nocite{*}
-%\bibliographystyle{plain}
-%\bibliography{references}
-%\printbibliography
-
 %\if1\blind
@@
 In the past, ADNI has also received funding from the National Institute of Biomedical Imaging and Bioengineering, the Canadian Institutes of Health Research, and private sector contributions through the Foundation for the National Institutes of Health (FNIH) including generous contributions from the following: AbbVie, Alzheimer’s Association; Alzheimer’s Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and Company; EuroImmun; F.
-Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research & Development, LLC.; Johnson & Johnson Pharmaceutical Research &Development LLC.; Lumosity; Lundbeck; Merck & Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
+Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research \& Development, LLC.; Johnson \& Johnson Pharmaceutical Research \& Development LLC.; Lumosity; Lundbeck; Merck \& Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
 %} \fi

### 2024-07-26 paper/main.tex

@@
   \caption{{\footnotesize Left Hippocampus and Left Thalamus Cross Sections.
-LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface.
-In the fifth hippocampus observation with $t=6.55$, the LPME-estimated surface has a gap.
+LPME-approximated embedding function routinely falls inside the boundaries of the observed hippocampus surface.
+In the fifth hippocampus observation with $t=2.11$, the LPME-estimated surface has a gap.
 Such gaps in the surface estimates frequently correspond with inaccurate volume estimates.

### 2025-01-08 code/05_lpme_adni_results.qmd
load Rfast library

@@
 library(RColorBrewer)
+library(Rfast)
 library(scatterplot3d)

### 2025-02-10 paper/main.tex
Updates from Overleaf

@@
 
+\newcommand{\revision}[1]{{\color{blue}}}
+
 \theoremstyle{definition}
@@
 {
-  \title{\bf Longitudinal Principal Manifold Estimation}
+  \title{\bf Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}
   \author{
@@
     and \\
-    Ani Eloyan \\
-    Department of Biostatistics, Brown University\\
+    Ani Eloyan \\    Department of Biostatistics, Brown University\\
     for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
@@
   \begin{center}
-    {\LARGE\bf Longitudinal Principal Manifold Estimation}
+    {\LARGE\bf Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}
 \end{center}
@@
 \begin{abstract}
-Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease.
+Longitudinal structural magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease.
 Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies.
@@
 Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
-The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
+The proposed method improves the estimation of surfaces of hippocampi and thalami using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
 An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
@@
 If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
+Similarly, \cite{akudjedu2018comparative} found that FIRST and FreeSurfer showed strong correlation with manual segmentation for the caudate, but relatively poor absolute agreement and moderate relative consistency when applied to the hippocampus.
+When comparing estimated subcortical volumes obtained using FIRST and FreeSurfer in two consecutive yearly study visits, \cite{quilis-sancho2020comparative} found that considerable variability was present in the relative differences in subcortical volume estimates, even after removing outliers from analysis.
+Furthermore, this variability was much greater for FIRST than FreeSurfer, with standard deviations for relative volumetric difference in the left and right hippocampus reaching 12.2\% and 11.3\%, respectively \citep{quilis-sancho2020comparative}.
 Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods.
@@
 Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
-However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
+However, this approach assumes that the progression trajectory for each subject must be parallel to a straight reference trajectory in some latent space, which may be overly restrictive in the case of heterogeneous conditions like AD.
 Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.
@@
     \begin{aligned}
-        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R},\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
+        F:\ \ & \mathbb{R}\times\mathbb{R}^d \rightarrow \mathbb{R}\times\mathbb{R}^D,\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
     \end{aligned}
@@
 \end{figure}
-
 Because the added dimensions duplicate information contained within the unaugmented observations, these dimensions can be discarded to reach estimates in the original $D$-dimensional space.
@@
 The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}.
-Based on our observations of systematic noise sources in imaging data, we add various types of terms to these functions at each visit.
-In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between visits and represent structural changes in the underlying manifold over time.
-The function $g(\cdot)$ may show change with respect to time that is constant (no change), linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier.
-The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between visits.
-Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean 1 and variance that is varied between visits.
-These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.
+We incorporate several sources of systematic variability to reflect the patterns of noise commonly observed in imaging data.
+In these functions, we specify $\alpha = 1 - \zeta\mathbf{g}(t) + \upsilon_{\text{a}}$ and $\beta = 1 + \upsilon_{\text{p}}$, where $\zeta$ and the functional form of $g(\cdot)$ are varied between visits to represent structural changes in the underlying manifold over time.
+The function $g(\cdot)$ may show change with respect to time that is constant ($g(t) = 0$), linear ($g(t) = t$), or quadratic ($g(t) = t^2$) with $\zeta$ serving as a scalar multiplier.
+The values $\upsilon_{\text{a}}$ and $\upsilon_{\text{p}}$, denoting random fluctuations in the manifold amplitude and period between visits, are randomly sampled from a normal distribution with mean zero and variance that is varied between simulation runs.
+Finally, the $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between simulation runs.
+
+Of particular interest is simulation case 8, in which the simulated data takes the form of a sphere with $D = 3$.
+Because $t \geq 0$ for all simulated visits, the linear and quadratic forms of $g(\cdot)$ result in a sphere with a decreasing radius, approximating atrophy in a subcortical structure.
 
@@
     \hline
-    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
-    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
-    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
-    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
-    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
-    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
-    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
-    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
+    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \iota$ & $-3 \leq r \leq 3$ \\
+    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
+    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
+    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \iota$ & $-1 \leq r \leq 1$ \\
+    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \iota$ & $0 \leq r \leq 3\pi$ \\
+    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
+    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
+    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
     \hline
@@
 
+%\begin{table}[ht]
+%  \centering
+%  \resizebox{\textwidth}{!}{\begin{tabular}{|c c c c c|}
+%    \hline
+%    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
+%    \hline
+%    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
+%    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
+%    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
+%    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
+%    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
+%    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
+%    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
+%    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
+%    \hline
+%  \end{tabular}}
+%  \caption{{\footnotesize Embedding functions used for simulation studies.
+Function $\bf{g}(t)$ denotes structural change in the underlying manifold over time, while $\zeta$ represents the scale of this change.
+Parameters $\alpha$ and $\beta$ are normally distributed with mean 1, and represent random changes in the manifold between time points, while $\iota$ represents within-image noise.}}
+%  \label{table:simulation_embeddings}
+%\end{table}
+
 In the situations where $d = 1$, LPME is compared to the PME method naïvely run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
@@
 
-A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal.
+A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\iota, \upsilon_{\text{a}}, \upsilon_{\text{p}} \in \left\{0.05, 0.1, 0.25, 0.5\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic; 5) $\zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5\right\}$.
 The sample size for each time point is set to 1,000 observations.
-Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 7,776 times.
-\if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.}\fi
+Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 5,760 times.
+\if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme}{\texttt{https://github.com/rjzielinski/lpme}}.}\fi Note that while discrete time points were used in these simulations, the spline model is also able to accommodate continuous time values, as seen in the ADNI application given in Section \ref{s:application}.
 
@@
   \caption{{\footnotesize Simulation Case 1.
-Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).}}
+Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).
+Noise introduced in the data generation process causes the observed data to deviate from the true manifold.
+The PME and principal curve estimates tend to closely follow the data at each time point, while the LPME estimates remain closer to the true manifold.}}
   \label{fig:sim_case1}
@@
 Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach results in lower mean squared distance values.
+
+INCLUDE VOLUME ESTIMATES FROM CASE 8 TO INDICATE PERFORMANCE IN ATROPHY SITUATION
 
@@
 
-As discussed previously, meaningful between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images.
+As discussed previously, substantial between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images.
 This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates.

### 2025-02-10 manuscript/paper/JASA_template_2021/JASA-TEMPLATE.tex
restructure paper files

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Title}
  \author{Author 1\thanks{
    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
    Department of YYY, University of XXX\\
    and \\
    Author 2 \\
    Department of ZZZ, University of WWW}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract. 200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  3 to 6 keywords, that do not appear in the title
\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Body of paper.  Margins in this document are roughly 0.75 inches all
around, letter size paper.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig1.pdf}
\end{center}
\caption{Consistency comparison in fitting surrogate model in the tidal
power example. \label{fig:first}}
\end{figure}

\begin{table}
\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
\begin{center}
\begin{tabular}{rrrrr}
one & two & three & four & five\\\hline
1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
\end{tabular}
\end{center}
\end{table}

\begin{itemize}
\item Note that figures and tables (such as Figure~\ref{fig:first} and
Table~\ref{tab:tabone}) should appear in the paper, not at the end or
in separate files.
\item In the latex source, near the top of the file the command
\verb+\newcommand{\blind}{1}+ can be used to hide the authors and
acknowledgements, producing the required blinded version.
\item Remember that in the blind version, you should not identify authors
indirectly in the text.  That is, don't say ``In Smith et. al.  (2009) we
showed that ...''.  Instead, say ``Smith et. al. (2009) showed that ...''.
\item These points are only intended to remind you of some requirements.
Please refer to the instructions for authors
at \url{http://amstat.tandfonline.com/action/authorSubmission?journalCode=uasa20&page=instructions#.VFkk7fnF_0c}
\item For more about ASA\ style, please see \url{http://journals.taylorandfrancis.com/amstat/asa-style-guide/}
\item If you have supplementary material (e.g., software, data, technical
proofs), identify them in the section below.  In early stages of the
submission process, you may be unsure what to include as supplementary
material.  Don't worry---this is something that can be worked out at later stages.
\end{itemize}

\section{Methods}
\label{sec:meth}
Don't take any of these section titles seriously.  They're just for
illustration.

\section{Verifications}
\label{sec:verify}
This section will be just long enough to illustrate what a full page of
text looks like, for margins and spacing.

\addtolength{\textheight}{.5in}%

\cite{Campbell02}, \cite{Schubert13}, \cite{Chi81}

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
{\bf With this spacing we have 26 lines per page.}
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

%new margin nere

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\addtolength{\textheight}{-.3in}%

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\section{Conclusion}
\label{sec:conc}

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Brief description. (file type)

\item[R-package for  MYNEW routine:] R-package ÒMYNEWÓ containing code to perform the diagnostic methods described in the article. The package also contains all datasets used as examples in the article. (GNU zipped tar file)

\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)

\end{description}

\section{BibTeX}

We hope you've chosen to use BibTeX!\ If you have, please feel free to use the package natbib with any bibliography style you're comfortable with. The .bst file agsm has been included here for your convenience.

\bibliographystyle{Chicago}

\bibliography{Bibliography-MM-MC}
\end{document}

### 2025-02-10 manuscript/paper/supplement.tex
restructure paper files

--- 2024-07-10 paper/supplement.tex
@@
 \begin{document}
+
+\title{\bf{Supplementary Material for ``Longitudinal Principal Manifold Estimation"}}
+\author{}
+
+\maketitle
 
@@
     \caption{Left hippocampus and left thalamus volume estimates for three ADNI participants.
-Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (red, circles), LPME (blue, squares), and PME (green, triangles).
+Volume estimates are obtained by summing the volumes of voxels contained within the structure as described by the segmented data (circles, solid line), LPME (squares, dashed line), and PME (triangles, dotted line).
 When PME and LPME are applied to the irregular shape of the hippocampus, there is a clear ordering of the volume estimates.
 Gaps in the estimated surface induce underestimates of the volume compared to the volume values estimated from the data.
-These gaps may also cause inconsistency in the LPME volume estimates, resulting in higher aggregate variability than for the data or PME estimates.
 When applied to the thalamus, the PME-based volume estimates demonstrate similar time point-to-time point changes in the volumes estimated from the segmented data, reflecting the close fit of the PME algorithm to the regularly-shaped thalamus data.
-The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 011\_S\_0003.}
+The LPME estimates appear to successfully smooth over regions with large variations in subsequent volume measures from the data and PME estimates, as seen for participant 033\_S\_0514.}
 \end{figure}

### 2025-02-10 manuscript/paper/main.tex
reset to JASA submission version, add title/abstract revisions

--- 2024-07-26 jasa/main.tex
@@
 \usepackage{tikz}
+\usepackage{ulem}
 \usepackage{url}
@@
 
+\newcommand{\revision}[1]{\textcolor{blue}{#1}}
+\newcommand{\outdated}[1]{\sout{#1}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 {
-  \title{\bf Longitudinal Principal Manifold Estimation}
+  \title{\bf \outdated{Longitudinal Principal Manifold Estimation} \revision{Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}}
   \author{
@@
 \begin{abstract}
-Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease.
+Longitudinal \revision{structural} magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease.
 Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies.
@@
 Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
-The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
+The proposed method improves the estimation of surfaces of \outdated{hippocampuses} \revision{hippocampi} and \outdated{thalamuses} \revision{thalami} using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
 An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.

### 2025-02-10 manuscript/paper/supplement.tex
reset to JASA submission version, add title/abstract revisions

@@
 \usepackage{url}
+
+\pdfminorversion=4
 
@@
       \centering
-      \includegraphics[height=3cm]{adni_plots/adni_lhipp_data_plot}
+      \includegraphics[height=3cm]{adni_lhipp_data_plot}
       \caption{Data}
@@
       \centering
-      \includegraphics[height=3cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
+      \includegraphics[height=3cm]{adni_lhipp_lpme_isomap_plot}
       \caption{LPME}
@@
       \centering
-      \includegraphics[height=3cm]{adni_plots/adni_lhipp_pme_plot}
+      \includegraphics[height=3cm]{adni_lhipp_pme_plot}
       \caption{PME}
@@
       \centering
-      \includegraphics[height=3cm]{adni_plots/adni_lthal_data_plot}
+      \includegraphics[height=3cm]{adni_lthal_data_plot}
       \caption{Data}
@@
       \centering
-      \includegraphics[height=3cm]{adni_plots/adni_lthal_lpme_isomap_plot}
+      \includegraphics[height=3cm]{adni_lthal_lpme_isomap_plot}
       \caption{LPME}
@@
       \centering
-      \includegraphics[height=3cm]{adni_plots/adni_lthal_pme_plot}
+      \includegraphics[height=3cm]{adni_lthal_pme_plot}
       \caption{PME}
@@
     \subfloat[\centering Left Hippocampus Volume Estimates]{
-      \includegraphics[height=8cm]{adni_plots/adni_lhipp_volume_comp}
+      \includegraphics[height=8cm]{adni_lhipp_volume_comp}
       \label{fig:lhipp_volume_comparison}
@@
     \subfloat[\centering Left Thalamus Volume Estimates]{
-      \includegraphics[height=8cm]{adni_plots/adni_lthal_volume_comp}
+      \includegraphics[height=8cm]{adni_lthal_volume_comp}
       \label{fig:lthal_volume_comparison}

### 2025-02-10 manuscript/paper/main.tex
complete textual revisions

@@
 \section{Introduction}
-
-Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}.
+Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopman2021Alzheimer}.
 Frequently, interest centers around longitudinal changes in one or more neurological substructures.
@@
 When analyzing data from studies with a large number of images, this approach may impose prohibitive costs.
-Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}.
+Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaude2011Bayesian}.
 While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.
@@
 Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study.
-In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
+In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulder2014Hippocampal} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer.
 After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively.
 If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation.
-Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods.
+\revision{Similarly, \cite{akudjedu2018comparative} found that FIRST and FreeSurfer showed strong correlation with manual segmentation for the caudate, but relatively poor absolute agreement and moderate relative consistency when applied to the hippocampus.
+When comparing estimated subcortical volumes obtained using FIRST and FreeSurfer in two consecutive yearly study visits, \cite{quilis-sancho2020comparative} found that considerable variability was present in the relative differences in subcortical volume estimates, even after removing outliers from analysis.
+Furthermore, this variability was much greater for FIRST than FreeSurfer, with standard deviations for relative volumetric difference in the left and right hippocampus reaching 12.2\% and 11.3\%, respectively \citep{quilis-sancho2020comparative}.} Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods.
 Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data.
@@
 In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space.
-For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.
+For example, \cite{yue2016Parameterization} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.
 Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.
 Nonlinear manifold learning methods have been extensively studied in the literature.
-However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies.
+However, many popular approaches, such as Isomap \citep{tenenbaum2000Global}, locally linear embedding \citep{roweis2000Nonlinear}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies.
 These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.
 
-Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional PCA method proposed by \cite{greven2011longitudinal}.
-As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space.
+Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional\outdated{PCA} \revision{principal component analysis} method proposed by \cite{greven2011longitudinal}.
+As for nonlinear methods, \cite{wolz2010Manifold} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space.
 However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously.
-Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
-However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.
-Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.
+Additionally, \cite{louis2019Riemannian} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling.
+\outdated{However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.}\revision{However, this approach assumes that the progression trajectory for each subject must be parallel to a straight reference trajectory in some latent space, which may be overly restrictive in the case of heterogeneous conditions like AD.} Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.
 
@@
 Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension.
-Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point.
+Specifically, we use the method proposed by \cite{meng2021Principal} to model the appropriate principal manifold at each given time point.
 Unlike most popular manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}).
@@
 The remainder of this article is laid out as follows.
-Section \ref{s:PME} reviews the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}.
+Section \ref{s:PME} reviews the principal manifold framework proposed by \cite{meng2021Principal}.
 In Section \ref{s:LPME}, we adapt the principal manifold framework for longitudinal settings, and show our proposed novel algorithm for dimension reduction using the longitudinal principal manifold framework.
@@
 
-The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud.
-Motivated by the penalization approach presented by \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions.
+The framework for principal manifolds originated with the concept of principal curves \citep{hastie1989Principal}, which are essentially curves that pass through the middle of a data cloud.
+Motivated by the penalization approach presented by \cite{smola2001Regularized}, \cite{meng2021Principal} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions.
 In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper.
@@
 Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$.
-The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection.
+The definitions of $\pi_f(x)$ given by \cite{hastie1989Principal} and \cite{meng2021Principal} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection.
 The principal manifolds are defined as follows.
@@
 In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.
-Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem, making this a supervised learning problem.
-\cite{mengPrincipalManifoldEstimation2021} also show that this function takes spline form
+Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{meng2021Principal} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem, making this a supervised learning problem.
+\cite{meng2021Principal} also show that this function takes spline form
 \begin{align}\label{eq: spline representation of the PME}
@@
 Importantly, $N$ tends to be much smaller than the number of points in the data cloud \citep[fig.
-3]{mengPrincipalManifoldEstimation2021}, which indicates that the representation in equation \eqref{eq: spline representation of the PME} is computationally efficient.
+3]{meng2021Principal}, which indicates that the representation in equation \eqref{eq: spline representation of the PME} is computationally efficient.
 This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$.
@@
 In this setting, the manifold to estimate is represented by the following, where $t$ coincides with the time dimension
+
+\revision{Note revision in equation from $\mathbb{R}^d \times \mathbb{R} \rightarrow \mathbb{R}^D \times \mathbb{R}$ to $\mathbb{R} \times \mathbb{R}^d \rightarrow \mathbb{R} \times \mathbb{R}^D$ to match function domain.}
 \begin{align}\label{eq: F form}
     \begin{aligned}
-        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R},\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
+        F:\ \ & \mathbb{R}\times\mathbb{R}^d \rightarrow \mathbb{R}\times\mathbb{R}^D,\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
     \end{aligned}
@@
 Specifically, at each individual time point, we apply the PME algorithm \citep[alg.
-2]{mengPrincipalManifoldEstimation2021} to fit the data cloud at the time point and represent the fitted manifold using the spline form in equation~\eqref{eq: spline representation of the PME}.
+2]{meng2021Principal} to fit the data cloud at the time point and represent the fitted manifold using the spline form in equation~\eqref{eq: spline representation of the PME}.
 Hence, we have a collection of spline coefficients, i.e., the $s_{j,l}$ and $\alpha_{k,l}$ in equation~\eqref{eq: spline representation of the PME}, associated with each time point.
@@
 To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method \citep[see][alg.
-1]{mengPrincipalManifoldEstimation2021} to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
+1]{meng2021Principal} to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$.
 Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using diamond-shaped points in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data.
@@
 By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ \citep[see the simulation study presented in][fig.
-3(c)]{mengPrincipalManifoldEstimation2021}.
+3(c)]{meng2021Principal}.
 The centers $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$ estimated in this step are then used to reach an initial low-dimensional parameterization.
@@
 The next step in the LPME algorithm for the estimation of the longitudinal principal manifold is initialization.
-The PME algorithm uses Isomap \citep{tenenbaumGlobalGeometricFramework2000} to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve.
+The PME algorithm uses Isomap \citep{tenenbaum2000Global} to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve.
 Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points.
@@
 Once the initial parameterization for each cluster center $\mu_{j, t}$ is obtained, the model fitting process begins by using the cluster centers $\mu_{j, t}$, weights $\hat{\theta}_{j, t}$ and the parameterizations associated with each time point to initialize a PME run.
-Specifically, the cluster centers $\mu_{j, t}$ and weights $\hat{\theta}_{j, t}$ replace those obtained using the HDMDE algorithm, and the Isomap-derived parameters obtained in Steps 1 and 2 of Algorithm 2 in \cite{mengPrincipalManifoldEstimation2021}, respectively.
+Specifically, the cluster centers $\mu_{j, t}$ and weights $\hat{\theta}_{j, t}$ replace those obtained using the HDMDE algorithm, and the Isomap-derived parameters obtained in Steps 1 and 2 of Algorithm 2 in \cite{meng2021Principal}, respectively.
 This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using lines in Figure \ref{fig:lpme_step3}.
@@
 In applying the proposed algorithms in extensive simulations, we did not encounter any issues with convergence of the algorithms.
-Further work must be done to demonstrate that the proposed estimator does indeed minimize the objective function, however \cite{mengPrincipalManifoldEstimation2021} demonstrated that the function that minimizes equation \eqref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimator in a penalized regression setting \citep{wahba1990}.
+Further work must be done to demonstrate that the proposed estimator does indeed minimize the objective function, however \cite{meng2021Principal} demonstrated that the function that minimizes equation \eqref{eq:pme_kappa} is of smoothing spline form, while smoothing splines can be derived as a Bayes estimator in a penalized regression setting \citep{wahba1990}.
 These results motivate our expectation that the estimator from the LPME algorithm will minimize equation \eqref{eq:newKappa}.
@@
 
-The self-consistency condition used to define principal curves in \cite{hastiePrincipalCurves1989} theoretically precludes the estimation of curves that are self-intersecting.
+The self-consistency condition used to define principal curves in \cite{hastie1989Principal} theoretically precludes the estimation of curves that are self-intersecting.
 However, Hastie and Stuetzle demonstrated that this can be done empirically through the use of periodic smoothers.
-\cite{banfieldIceFloeIdentification1992} generalized the algorithm proposed in \cite{hastiePrincipalCurves1989}, averaging over projection residuals rather than data points, to fit closed curves.
+\cite{banfieldIceFloeIdentification1992} generalized the algorithm proposed in \cite{hastie1989Principal}, averaging over projection residuals rather than data points, to fit closed curves.
 These approaches do not adapt easily to the higher dimensional penalized regression setting we consider here.
 
-A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results, preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{mengPrincipalManifoldEstimation2021}.
+A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results, preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{meng2021Principal}.
 To avoid this process, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$.
@@
 Because the added dimensions duplicate information contained within the unaugmented observations, these dimensions can be discarded to reach estimates in the original $D$-dimensional space.
-This approach to data augmentation offers an alternative to manual partitioning used by \cite{mengPrincipalManifoldEstimation2021}.
+This approach to data augmentation offers an alternative to manual partitioning used by \cite{meng2021Principal}.
 Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate.
@@
 
-In the situations where $d = 1$, LPME is compared to the PME method naïvely run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point.
+In the situations where $d = 1$, LPME is compared to the PME method naïvely run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastie1989Principal}, also run independently at each time point.
 The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package \citep{Cannoodt2018princurve}, developed using \texttt{R} \citep{rSoftware2023}.
 In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen.
-Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$.
+Following \cite{meng2021Principal}, the inputs for the PME algorithm are set to $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$.
 The minimum number of cluster centers is set to $10 \times d$.
-In cases where $d = 2$, LPME is again compared to the naïve PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.
+In cases where $d = 2$, LPME is again compared to the naïve PME approach described above, and the principal surface estimation algorithm developed by \cite{yue2016Parameterization}.
 
@@
 Each combination of factor levels is run once, resulting in each embedding map being simulated a total of 7,776 times.
-\if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.}\fi
+\if1\blind{All code required to reproduce the simulation results and associated Figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.}\fi \revision{Note that while discrete time points were used in these simulations, the spline model is also able to accommodate continuous time values, as seen in the ADNI application given in Section \ref{s:application}.}
 
@@
   \caption{{\footnotesize Simulation Case 1.
-Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).}}
+Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).
+\revision{Noise introduced in the data generation process causes the observed data to deviate from the true manifold.
+The PME and principal curve estimates tend to closely follow the data at each time point, while the LPME estimates remain closer to the true manifold.}}}
   \label{fig:sim_case1}
@@
 
-As discussed previously, meaningful between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images.
+As discussed previously, \outdated{meaningful} \revision{substantial} between-image errors in estimates of subcortical structures are introduced during the process of segmenting MRI images.
 This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates.
@@
 In practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature.
-The use of alternative methods, such as Locally Linear Embedding \citep{roweisNonlinearDimensionalityReduction2000} or Laplacian Eigenmaps \citep{belkin2003laplacian}, to generate the initial parameterization may alleviate this concern.
+The use of alternative methods, such as Locally Linear Embedding \citep{roweis2000Nonlinear} or Laplacian Eigenmaps \citep{belkin2003laplacian}, to generate the initial parameterization may alleviate this concern.
 A second assumption is that the LPME method requires relative stability in the overall shape of the underlying manifold structure.
@@
 If the resulting manifold has gaps, however, this procedure may not be possible to implement.
-\cite{mengPrincipalManifoldEstimation2021} introduced an interior identification approach that effectively makes use of the explicit representation of the embedding map obtained through the PME and LPME fitting processes.
+\cite{meng2021Principal} introduced an interior identification approach that effectively makes use of the explicit representation of the embedding map obtained through the PME and LPME fitting processes.
 Future work may adapt this procedure to settings with gaps in the estimated manifold.
@@
 
-%\newpage
+\newpage
 \bibliographystyle{jasa}

### 2025-02-10 code/01_lpme_simulations.qmd
align with reproducibility template

@@
 author: "Robert Zielinski"
-date: "April 23, 2024"
-output: html_document
+date: "January 31, 2025"
+format: html
 ---
 
-This notebook includes all code necessary to replicate the results of the simulations comparing performance between the LPME, PME, and PC/PS algorithms on data generated from a common underlying manifold, but with varying degrees of noise added between time points.
+This notebook includes all code needed to replicate the results of the simulations comparing performance between the Principal Manifold Estimation (PME), Longitudinal Principal Manifold Estimation (LPME), and Principal Curve/Surface (PC/PS) algorithms on data generated from a common underlying manifold, but with varying degrees of noise added independently between time points.
 The LPME algorithm is designed to obtain *smooth* estimates of the manifolds underlying noisy data.
@@
 
-```{r, message=FALSE}
+```{r, echo=FALSE}
+#| label: chunk_setup
+knitr::opts_chunk$set(message = FALSE, warning = FALSE)
+```
+
+```{r, echo=FALSE}
+#| label: load_packages
 library(tidyverse)
@@
 library(princurve)
-
-# note that file paths are relative to the data/ directory where this notebook resides
-source("functions/sim_data.R")
-source("functions/calc_pme_est.R")
-source("functions/calc_lpme_est.R")
+library(Rfast)
+
 source("prinSurf_v3.R")
-```
-
-Manifold learning algorithms traditionally have tradeoffs when compared to each other, meaning that an algorithm that displays strong performance for one manifold may be outperformed by another algorithm when considering a different manifold.
-Because of this, manifold learning algorithms are often evaluated with several manifolds meant to test an algorithm's performance on data with various attributes.
-We consider eight different manifolds:
+source("functions/simulations/simulate_data.R")
+source("functions/simulations/preprocess_data.R")
+source("functions/simulations/fit_models.R")
+source("functions/simulations/evaluate_models.R")
+source("functions/simulations/display_results.R")
+source("functions/calculate_lpme_reconstructions.R")
+source("functions/calculate_pme_reconstructions.R")
+source("functions/estimate_volume.R")
+source("functions/interior_identification.R")
+```
+## Simulated Manifolds
+
+Machine learning algorithms traditionally have tradeoffs when compared to each other, meaning that an algorithm that displays strong performance for one manifold may be outperformed by another algorithm when considering a different manifold.
+Because of this, it is advantageous to evaluate manifold learning algorithms with several manifolds meant to test an algorithm's performance on data with a variety of attributes.
+The function given in the code chunk below will be used to generate data lying along the manifolds of interest.
+In the function above, we consider eight different manifolds:
 
 1. For $r \in [-3, 3]$, $f(r) = (r, \sin(r + \frac{\pi}{2}))$.
-
-```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+```{r, echo=FALSE}
+#| label: plot_case1
+set.seed(100)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 1,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 1],
-  z = sim_df[, 3],
+  data = sim_df,
+  x = ~X1,
+  y = ~time,
+  z = ~X2,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 1],
-    z = true_vals[, 3],
+    x = ~X_true1,
+    y = ~time,
+    z = ~X_true2,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 1")
-```
-
-2. For $r \in [-3\pi, 3\pi]$, $f(r) = (r, \sin(r))$
-
-```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+  )
+```
+
+2. For $r \in [-3\pi, 3\pi]$, $f(r) = (r, \sin(r))$.
+```{r, echo=FALSE}
+#| label: plot_case2
+set.seed(200)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 2,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 1],
-  z = sim_df[, 3],
+  data = sim_df,
+  x = ~X1,
+  y = ~time,
+  z = ~X2,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 1],
-    z = true_vals[, 3],
+    x = ~X_true1,
+    y = ~time,
+    z = ~X_true2,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 2")
-```
-
-3. For $r \in [-\frac{4}{5}\pi, \frac{\pi}{2}]$, $f(r) = (\cos(r), \sin(r))$
-
-```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+  )
+```
+
+3. For $r \in [-\frac{4}{5}\pi, \frac{\pi}{2}]$, $f(r) = (\cos(r), \sin(r))$.
+```{r, echo=FALSE}
+#| label: plot_case3
+set.seed(300)
+sim_df <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 3,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.15,
+  amplitude_noise = 0.25,
+  period_noise = 0.10,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 1],
-  z = sim_df[, 3],
+  data = sim_df,
+  x = ~X1,
+  y = ~time,
+  z = ~X2,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 1],
-    z = true_vals[, 3],
+    x = ~X_true1,
+    y = ~time,
+    z = ~X_true2,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 3")
+  )
 ```
@@
 4. For $r \in [-1, 1]$, $f(r) = (r, r^2, r^3)$
-
-```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+```{r, echo=FALSE}
+#| label: plot_case4
+set.seed(400)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 4,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.15,
+  amplitude_noise = 0.25,
+  period_noise = 0.10,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 3],
-  z = sim_df[, 4],
-  frame = sim_df[, 1],
+  data = sim_df,
+  x = ~X1,
+  y = ~X2,
+  z = ~X3,
+  frame = ~time,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 3],
-    z = true_vals[, 4],
-    frame = true_vals[, 1],
+    x = ~X_true1,
+    y = ~X_true2,
+    z = ~X_true3,
+    frame = ~time,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 4")
+  )
 ```
@@
 ```{r, echo=FALSE}
+#| label: plot_case5
 set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 5,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.15,
+  amplitude_noise = 0.25,
+  period_noise = 0.10,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 3],
-  z = sim_df[, 4],
-  frame = sim_df[, 1],
+  data = sim_df,
+  x = ~X1,
+  y = ~X2,
+  z = ~X3,
+  frame = ~time,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 3],
-    z = true_vals[, 4],
-    frame = true_vals[, 1],
+    x = ~X_true1,
+    y = ~X_true2,
+    z = ~X_true3,
+    frame = ~time,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 5")
+  )
+
 ```
@@
 ```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+#| label: plot_case6
+set.seed(600)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 6,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.15,
+  amplitude_noise = 0.25,
+  period_noise = 0.10,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 3],
-  z = sim_df[, 4],
-  frame = sim_df[, 1],
+  data = sim_df,
+  x = ~X1,
+  y = ~X2,
+  z = ~X3,
+  frame = ~time,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 3],
-    z = true_vals[, 4],
-    frame = true_vals[, 1],
+    x = ~X_true1,
+    y = ~X_true2,
+    z = ~X_true3,
+    frame = ~time,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 6")
+  )
+
 ```
@@
 ```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+#| label: plot_case7
+set.seed(700)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 7,
-  noise = 0.15,
-  amp_noise = 0.25 / 25, # noise reduced to maintain integrity of shape
-  period_noise = 0.25 / 25,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.15,
+  amplitude_noise = 0.05,
+  period_noise = 0.01,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 3],
-  z = sim_df[, 4],
-  frame = sim_df[, 1],
+  data = sim_df,
+  x = ~X1,
+  y = ~X2,
+  z = ~X3,
+  frame = ~time,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 3],
-    z = true_vals[, 4],
-    frame = true_vals[, 1],
+    x = ~X_true1,
+    y = ~X_true2,
+    z = ~X_true3,
+    frame = ~time,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 7")
+  )
+
 ```
@@
 ```{r, echo=FALSE}
-set.seed(500)
-time_vals <- seq(0, 1, 0.2)
-sim_list <- lapply(
-  time_vals,
-  sim_data,
+#| label: plot_case8
+set.seed(800)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
   case = 8,
-  noise = 0.15,
-  amp_noise = 0.25,
-  period_noise = 0.25 / 5,
-  N = 1000,
-  time_change = 0,
-  time_trend = "constant"
-)
-
-sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-for (i in 1:length(sim_list)) {
-  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-}
-sim_df <- sim_df[-1, ]
-true_vals <- true_vals[-1, ]
+  obs_noise = 0.15,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
 
 plot_ly(
-  x = sim_df[, 2],
-  y = sim_df[, 3],
-  z = sim_df[, 4],
-  frame = sim_df[, 1],
+  data = sim_df,
+  x = ~X1,
+  y = ~X2,
+  z = ~X3,
+  frame = ~time,
   type = "scatter3d",
@@
   name = "Data"
-) %>%
+) |>
   add_markers(
-    x = true_vals[, 2],
-    y = true_vals[, 3],
-    z = true_vals[, 4],
-    frame = true_vals[, 1],
+    x = ~X_true1,
+    y = ~X_true2,
+    z = ~X_true3,
+    frame = ~time,
     marker = list(size = 3, color = "red"),
     name = "True Manifold"
-  ) %>%
-  layout(title = "Simulation Case 8")
-```
-
-Simulation runs were conducted using functions for each case, with each function taking the following arguments:
-
-- `max_time`: The maximum time from baseline of a set of observations, or the study duration.
-- `interval`: The interval of study visits.
-Intervals between study visits were assumed to be constant in these simulations, but the LPME algorithm can handle settings with varying intervals between visits.
-- `amp_noise`: Specifies the magnitude of between-time point noise impacting the amplitude of the functions used to generate the data.
-- `shape_noise`: Specifies the magnitude of between-time point noise impacting the shape, or period, of the functions used to generate the data.
-- `time_change`: Specifies the magnitude of structural change in the underlying manifold over time.
-- `time_trend`: Describes the type of longitudinal change in the underlying manifold. May take one of four values - constant, linear, quadratic, or sinusoidal.
-- `n`: The number of observations generated at each time point.
-- `initialization_alg`: The manifold learning algorithm used to initialize the PME algorithm. May currently take one of three values - isomap, diffusion_maps, or laplacian_eigenmaps.
-- `run`: Used to save multiple simulation results with the same parameter inputs.
-- `print_plots`: Specify whether plots of preliminary results should be viewed. This is useful for debugging but is computationally expensive.
-
-The functions are given in the code chunks below:
-
-### Simulation Case 1
+  )
+
+```
+
+Note that case 8, in which observations drawn from a 2-dimensional manifold take the form of a sphere in 3 dimensions, is of particular interest due to our application to the surfaces of subcortical structures in the brain.
+
+Simulation runs were conducted by varying the following arguments:
+
+- `duration`: The simulated study's follow-up duration. Values tested are 1, 2, and 5, representing study durations of 1, 2, and 5 years, respectively.
+- `interval`: The scheduled interval of study visits. Values tested are 0.25 and 0.5, representing study visits every 3 months and every 6 months, respectively.
+- `case`: Specifies the manifold used to generate the data. 
+- `obs_noise`: The standard deviation of the independent Gaussian observation-level noise. Values tested include {0.05, 0.1, 0.25, 0.5}.
+- `amplitude_noise`: The standard deviation of Gaussian noise applied to the amplitude of the manifold at each time point. Values tested include {0.05, 0.1, 0.25, 0.5}.
+- `period_noise`: The standard deviation of Gaussian noise applied to the period of the manifold at each time point. Values tested include {0.05, 0.1, 0.25, 0.5}.
+- `time_trend`: Specifies systematic longitudinal changes in the manifold amplitude.
+Possible values include "constant", indicating that the manifold does not change over time, "linear", representing a constant rate of change in manifold amplitude, and "quadratic", indicating a quadratic rate of change in the manifold amplitude.
+To approximate trends observed when the brain atrophies, all changes in amplitude are downward.
+- `time_change`: Scales the longitudinal amplitude changes. Values tested include {0, 0.05, 0.1, 0.25, 0.5}.
+
+The sample size at each time point is set to 1,000. A factorial design was used to specify combinations of parameter values, with each simulation case being run a total of 5,760 times.
+
+## Simulation Analysis Pipeline
+
+For our purposes, analysis of the simulated data consists of the following steps:
+
+- Preprocessing: Standardize observations to fall between -1 and 1 and augment with polar or spherical coordinates if needed.
+- Manifold learning: Fit the LPME, PME and Principal Curve/Surface algorithms to the observed data.
+- Model evaluation: For each model, obtain reconstructed estimates of the original observations.
+Compute the mean squared distance between these estimates and the D-dimensional observations generated from the true manifold.
+
+Functions used to obtain the algorithm-specific reconstructions are provided in the code chunk below:
+
+Functions specifying each step of the analysis process can be found at `code/functions/simulations/preprocess_data.R`, `code/functions/simulations/fit_models.R`, and `code/functions/simulations/evaluate_models.R`.
+An additional function used to generate plots to display the resulting model estimates can be found at `code/functions/simulations/display_results.R`.
+
+Demonstrations of this analysis pipeline for several simulation cases can be found below.
+
+### Simulation Demonstrations
+
+```{r, eval=FALSE}
+#| label: demo_case1
+set.seed(100)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
+  case = 1,
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
+
+sim_processed <- preprocess_data(sim_df, case = 1, d = 1, D = 2)
+sim_models <- fit_models(sim_processed, case = 1, d = 1, D = 2)
+model_errors <- evaluate_models(
+  sim_processed,
+  sim_models,
+  case = 1,
+  d = 1,
+  D = 2
+)
+model_errors
+result_plot <- display_results(
+  sim_processed,
+  sim_models,
+  case = 1,
+  d = 1,
+  D = 2
+)
+result_plot
+```
+
+```{r, eval=FALSE}
+#| label: demo_case3
+set.seed(300)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
+  case = 3,
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
+
+sim_processed <- preprocess_data(sim_df, case = 3, d = 1, D = 2)
+sim_models <- fit_models(sim_processed, case = 3, d = 1, D = 2)
+model_errors <- evaluate_models(
+  sim_processed,
+  sim_models,
+  case = 3,
+  d = 1,
+  D = 2
+)
+model_errors
+
+result_plot <- display_results(
+  sim_processed,
+  sim_models,
+  case = 3,
+  d = 1,
+  D = 2
+)
+result_plot
+```
+
+```{r, eval=FALSE}
+#| label: demo_case7
+set.seed(700)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
+  case = 7,
+  obs_noise = 0.1,
+  amplitude_noise = 0.01,
+  period_noise = 0.01,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
+
+sim_processed <- preprocess_data(sim_df, case = 7, d = 2, D = 3)
+sim_models <- fit_models(sim_processed, case = 7, d = 2, D = 3)
+model_errors <- evaluate_models(
+  sim_processed,
+  sim_models,
+  case = 7,
+  d = 2,
+  D = 3
+)
+model_errors
+
+result_plot <- display_results(
+  sim_processed,
+  sim_models,
+  case = 7,
+  d = 2,
+  D = 3
+)
+result_plot
+```
+
+```{r, eval=FALSE}
+#| label: demo_case8
+set.seed(800)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
+  case = 8,
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
+
+sim_processed <- preprocess_data(sim_df, case = 8, d = 2, D = 3)
+sim_models <- fit_models(sim_processed, case = 8, d = 2, D = 3)
+model_errors <- evaluate_models(
+  sim_processed,
+  sim_models,
+  case = 8,
+  d = 2,
+  D = 3
+)
+model_errors
+
+result_plot <- display_results(
+  sim_processed,
+  sim_models,
+  case = 8,
+  d = 2,
+  D = 3
+)
+result_plot
+```
+
+## Running Simulations
+
+The code chunks below specify the combinations of simulation parameters and runs the simulations in parallel.
+Note that due to computational demands, these simulations are not run in the process of compiling this notebook.
+This was handled by the `code/run_simulations.R` script.
+However, the code chunk below will yield identical results, and those same results are discussed later in the notebook.
+
+```{r, eval=FALSE}
+#| label: simulation_parameters
+
+# set possible values of simulation parameters as discussed above
+durations <- c(1, 2, 5)
+intervals <- c(0.25, 0.5)
+cases <- 1:8
+obs_noises <- c(0.05, 0.1, 0.25, 0.5)
+amplitude_noises <- c(0.05, 0.1, 0.25, 0.5)
+period_noises <- c(0.05, 0.1, 0.25, 0.5)
+time_trends <- c("constant", "linear", "quadratic")
+time_changes <- c(0, 0.05, 0.1, 0.25, 0.5)
+
+# create dataframe of all combinations of parameter values
+parameter_df <- expand_grid(
+  duration = durations,
+  interval = intervals,
+  case = cases,
+  obs_noise = obs_noises,
+  amplitude_noise = amplitude_noises,
+  period_noise = period_noises,
+  time_trend = time_trends,
+  time_change = time_changes
+)
+
+# create empty dataframe to store error estimates
+error_df <- tibble(
+  lpme_error = vector(mode = "numeric", length = nrow(parameter_df)),
+  pme_error = vector(mode = "numeric", length = nrow(parameter_df)),
+  pc_error = vector(mode = "numeric", length = nrow(parameter_df)),
+  data_error = vector(mode = "numeric", length = nrow(parameter_df))
+)
+
+# save the seed state that was set when each simulation was run
+seed_states <- list()
+```
+
+```{r, eval=FALSE}
+#| label: run_simulations
+print("Starting Simulations...")
+
+cl <- makeCluster(parallel::detectCores() - 1)
+# unclear why, but registerDoSNOW worked better when running with slurm
+registerDoSNOW(cl)
+
+# the registerDoRNG function allows reproducibility
+# of the simulations when run in parallel
+registerDoRNG(611492)
+
+pb  <- txtProgressBar(max = nrow(parameter_df), style = 3)
+progress <- function(n) setTxtProgressBar(pb, n)
+opts <- list(progress = progress)
+error_df <- foreach(
+  row_idx = seq_len(nrow(parameter_df)),
+  .export = c(
+    "simulate_data",
+    "preprocess_data",
+    "fit_models",
+    "evaluate_models",
+    "display_results"
+  ),
+  .packages = c("tidyverse", "pme"),
+  .combine = rbind,
+  .options.snow = opts
+) %dopar% {
+  seed_val <- .Random.seed
+  seed_states[[row_idx]] <- seed_val
+
+  duration_value <- parameter_df$duration[row_idx]
+  interval_value <- parameter_df$interval[row_idx]
+  case_value <- parameter_df$case[row_idx]
+  obs_noise_value <- parameter_df$obs_noise[row_idx]
+  amplitude_noise_value <- parameter_df$amplitude_noise[row_idx]
+  period_noise_value <- parameter_df$period_noise[row_idx]
+  trend_value <- parameter_df$time_trend[row_idx]
+  change_value <- parameter_df$time_change[row_idx]
+
+  # additional stability is needed for Swiss roll and spherical cases
+  # to maintain structural integrity
+  if (case_value == 7) {
+    amplitude_noise_value <- amplitude_noise_value / 10
+    period_noise_value <- period_noise_value / 10
+  } else if (case_value == 8) {
+    period_noise_value <- period_noise_value / 10
+  }
+
+  D_value <- case_when(
+    case_value == 1 ~ 2,
+    case_value == 2 ~ 2,
+    case_value == 3 ~ 2,
+    case_value == 4 ~ 3,
+    case_value == 5 ~ 3,
+    case_value == 6 ~ 3,
+    case_value == 7 ~ 3,
+    case_value == 8 ~ 3
+  )
+  d_value <- case_when(
+    case_value == 1 ~ 1,
+    case_value == 2 ~ 1,
+    case_value == 3 ~ 1,
+    case_value == 4 ~ 1,
+    case_value == 5 ~ 1,
+    case_value == 6 ~ 2,
+    case_value == 7 ~ 2,
+    case_value == 8 ~ 2
+  )
+
+  # following the same analysis pipeline as described above
+  sim_data <- simulate_data(
+    duration = duration_value,
+    interval = interval_value,
+    case = case_value,
+    obs_noise = obs_noise_value,
+    amplitude_noise = amplitude_noise_value,
+    period_noise = period_noise_value,
+    time_trend = trend_value,
+    time_change = change_value,
+    N = 1000
+  )
+  sim_df <- sim_data$df
+
+  sim_processed <- preprocess_data(
+    sim_df,
+    case = case_value,
+    d = d_value,
+    D = D_value
+  )
+
+  sim_models <- fit_models(
+    sim_processed,
+    case = case_value,
+    d = d_value,
+    D = D_value
+  )
+
+  model_errors <- evaluate_models(
+    sim_processed,
+    sim_models,
+    case = case_value,
+    d = d_value,
+    D = D_value
+  )
+
+  result_plot <- display_results(
+    sim_processed,
+    sim_models,
+    case = case_value,
+    d = d_value,
+    D = D_value
+  )
+
+  simulation_output <- list(
+    data = sim_data,
+    processed_data = sim_processed,
+    models = sim_models,
+    errors = model_errors,
+    plot = result_plot,
+    seed = seed_val
+  )
+
+  result_dir <- paste0("../output/simulations/case", case_value)
+  if (!dir.exists(result_dir)) {
+    dir.create(result_dir, recursive = TRUE)
+  }
+
+  simulation_file <- paste0(
+    "duration_",
+    duration_value,
+    "_interval_",
+    str_pad(as.character(interval_value * 100), 3, pad = "0"),
+    "_obs_noise_",
+    str_pad(as.character(obs_noise_value * 100), 3, pad = "0"),
+    "_amplitude_noise_",
+    str_pad(as.character(amplitude_noise_value * 100), 3, pad = "0"),
+    "_period_noise_",
+    str_pad(as.character(period_noise_value * 100), 3, pad = "0"),
+    "_time_trend_",
+    trend_value,
+    "_time_change_",
+    str_pad(as.character(change_value * 100), 3, pad = "0"),
+    ".RDS"
+  )
+
+  saveRDS(simulation_output, file.path(result_dir, simulation_file))
+
+  unlist(model_errors)
+}
+
+stopCluster(cl)
+```
+
+```{r, eval=FALSE}
+#| label: save_simulation_results
+error_df <- error_df %>%
+  as_tibble()
+print(error_df)
+
+print("Simulations Complete")
+
+simulation_results <- bind_cols(parameter_df, error_df)
+write_csv(simulation_results, "../output/simulation_results.csv")
+```
+## Simulation Results
 
 ```{r}
-sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 1,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
+#| label: sample_result
+result_case1 <- readRDS(
+  file.path(
+    "../output/simulations/case1",
+    "duration_1_interval_025_obs_noise_005_amplitude_noise_010_period_noise_010_time_trend_linear_time_change_010.RDS"
+    # "duration_1_interval_050_obs_noise_010_amplitude_noise_025_period_noise_010_time_trend_linear_time_change_010.RDS"
+    # "duration_1_interval_025_obs_noise_005_amplitude_noise_005_period_noise_005_time_trend_constant_time_change_025.RDS"
+  )
+)
+
+case1_observed <- result_case1$processed_data$df_observed
+case1_true <- result_case1$processed_data$df_true
+case1_sim <- result_case1$data$df
+
+lpme_data <- result_case1$models$lpme$reconstructions
+pme_data <- result_case1$models$pme$reconstructions
+pc_data <- result_case1$models$pc$reconstructions
+
+case1_names <- c("time", "X1", "X2")
+names(case1_observed) <- case1_names
+names(case1_true) <- case1_names
+case1_observed$source <- "Data"
+case1_observed$param <- case1_sim$r1
+case1_true$source <- "True Manifold"
+case1_true$param <- case1_sim$r1
+
+lpme_df <- as_tibble(lpme_data, ..name_repair = NULL)
+names(lpme_df) <- case1_names
+lpme_df$source <- "LPME"
+lpme_df$param <- case1_sim$r1
+
+pme_df <- as_tibble(pme_data, ..name_repair = NULL)
+names(pme_df) <- case1_names
+pme_df$source <- "PME"
+pme_df$param <- case1_sim$r1
+
+pc_df <- as_tibble(pc_data, ..name_repair = NULL)
+names(pc_df) <- case1_names
+pc_df$source <- "PC"
+pc_df$param <- case1_sim$r1
+
+result_case1$plot
+```
+
+As an example simulation result, we consider the run using the first manifold case with a study duration of 1 year, visits every 3 months, observation noise = 0.025, amplitude noise = 0.005, period noise = 0.005, and no change in the underlying over time.
+The model estimates are shown in 3 dimensions in the interactive plot above, or in 2 dimensions below.
+Visual ispection indicates that the PME estimate, shown in green, tends to follow the observed data at each time point, while the LPME estimate appears to fall between the PME estimate and the observed manifold at most time points.
+When `time = 1`, the LPME estimate is much further from the true manifold, as there are no further time points to offset the error in the observations collected at that visit.
+
+```{r, echo=FALSE}
+#| label: result_plot
+case1_data <- bind_rows(
+  case1_observed,
+  case1_true,
+  lpme_df,
+  pme_df,
+  pc_df
+) |>
+  arrange(param)
+
+color_values <- c("black", "blue", "purple", "green", "red")
+
+set.seed(8195)
+# subsample data to improve clarity of visualization
+data_sampled <- case1_data |>
+  group_by(time, source) |>
+  sample_n(100)
+```
+
+```{r, echo=FALSE}
+#| label: result_facet_plot
+ggplot(case1_data, aes(x = X1, y = X2)) +
+  facet_wrap(~time, nrow = 3) +
+  geom_point(
+    data = subset(case1_data, source == "Data"),
+    color = "black",
+    alpha = 0.2,
+    size = 0.25
+  ) +
+  geom_point(
+    data = subset(case1_data, source != "Data"),
+    aes(color = source, shape = source),
+    size = 0.2
+  ) +
+  scale_color_manual(values = color_values[-1])
+```
+
+```{r}
+result_case1$errors
+```
+
+Strikingly, the principal curve algorithm does not appear to fit the data well at many points, visibly losing the structure of the underlying manifold at many points.
+
+## Atrophy Estimation
+
+The proposed use case for the LPME algorithm is to obtain smooth estimates of the manifold underlying the shape of the hippocampus over time, which can in turn be used to estimate the volume of the hippocampus, giving insight into atrophy that occurs as Alzheimer's disease progresses.
+To understand how well this approach performs, we consider simulation case 8, in which data from a 2-dimensional manifold is embedded in a spherical form.
+
+Analogous to the proposed application, the LPME estimate can be used to approximate the volume of the simulated sphere.
+Unlike the application, the volume of the true sphere is known and provides a ground truth, allowing us to evaluate the performance of the smoothing in the presence of data generating patterns commonly found in neuroimaging data.
+
+```{r}
+dir_case8 <- "../output/simulations/case8"
+files_case8 <- list.files(dir_case8)
+
+file_name <- files_case8[1]
+file_stripped <- str_replace(file_name, ".RDS", "")
+file_parsed <- str_split(file_stripped, "_")
+
+duration_val <- as.numeric(file_parsed[[1]][2])
+interval_val <- as.numeric(file_parsed[[1]][4]) / 100
+time_trend_val <- file_parsed[[1]][16]
+time_change_val <- as.numeric(file_parsed[[1]][19]) / 100
+
+sim <- readRDS(file.path(dir_case8, file_name))
+
+sim_data <- sim$processed_data$df
+time_points <- sim_data |>
+  select(time) |>
+  unlist() |>
+  unique()
+
+sim_mat <- sim$processed_data$df_observed |>
+  select(time, contains("X")) |>
+  as.matrix()
+
+radii <- map(
+  seq_along(sim$data$amplitude_values),
+  ~ sim$data$amplitude_values[[.x]][1]
+) |>
+  reduce(c)
+
+observed_volumes <- (4 / 3) * pi * radii^3
+
+time_adjustments <- case_when(
+  time_trend_val == "constant" ~ 0 * time_points,
+  time_trend_val == "linear" ~ time_points,
+  time_trend_val == "quadratic" ~ time_points^2
+)
+
+# scale time adjustments to be proportion of 1
+if (max(time_adjustments) == 0) {
+  time_adjustments_scaled <- time_adjustments * time_change_val * 0
+} else {
+  time_adjustments_scaled <- (
+    time_adjustments / max(time_adjustments)
+  ) * time_change_val
 }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
+# subtract from initial amplitude to replicate effects of atrophy
+true_radii <- 1 - time_adjustments_scaled
+true_volumes <- (4 / 3) * pi * true_radii^3
+```
+
+```{r}
+lpme_interior_points <- matrix(ncol = ncol(sim_mat))
+lpme_volume <- vector(mode = "numeric", length = length(time_points))
+pme_interior_points <- matrix(ncol = ncol(sim_mat))
+pme_volumes <- vector(mode = "numeric", length = length(time_points))
+
+df_observed <- sim$processed_data$df_observed |>
+  select(time, contains("X")) |>
+  as.matrix()
+df_true <- sim$processed_data$df_true |>
+  select(time, contains("X")) |>
+  as.matrix()
+df_lpme <- sim$models$lpme$reconstructions[, 1:4]
+df_pme <- sim$models$pme$reconstructions[, 1:4]
+
+for (time_idx in seq_along(time_points)) {
+  temp_observed <- df_observed[df_observed[, 1] == time_points[time_idx], -1]
+  temp_true <- df_true[df_true[, 1] == time_points[time_idx], -1]
+  temp_lpme <- df_lpme[df_lpme[, 1] == time_points[time_idx], -1]
+  temp_pme <- df_pme[df_pme[, 1] == time_points[time_idx], -1]
+
+  scaling_factor <- 0.01
+  temp_observed_scaled <- round(temp_observed / scaling_factor)
+  temp_true_scaled <- round(temp_true / scaling_factor)
+  temp_lpme_scaled <- round(temp_lpme / scaling_factor)
+  temp_pme_scaled <- round(temp_pme / scaling_factor)
 }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal curve estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  # test all smoothing options for principal curve algorithm
-  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 1,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers",
-      verbose = FALSE
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    curves <- list()
-    curve_vals <- list()
-    curve_error <- vector()
-    for (smoother_idx in 1:length(smoothing_options)) {
-      curves[[smoother_idx]] <- principal_curve(
-        temp_data,
-        smoother = smoothing_options[smoother_idx]
-      )
-      # save projections of observations onto estimated principal curve
-      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
-      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
+```
+
+```{r, eval=FALSE}
+lpme_interior_points <- matrix(ncol = ncol(sim_mat))
+lpme_volumes <- vector(mode = "numeric", length = length(time_points))
+
+pme_interior_points <- matrix(ncol = ncol(sim_mat))
+pme_volumes <- vector(mode = "numeric", length = length(time_points))
+
+for (time_idx in seq_along(time_points)) {
+  temp_data <- sim_mat[sim_mat[, 1] == time_points[time_idx], -1]
+
+  data_limits <- colMinsMaxs(temp_data)
+
+  limit_scaler <- 0.1
+  x_seq <- seq(
+    data_limits[1, 1] - abs(data_limits[1, 1] * limit_scaler),
+    data_limits[2, 1] + abs(data_limits[2, 1] * limit_scaler),
+    by = 0.1
+  )
+  y_seq <- seq(
+    data_limits[1, 2] - abs(data_limits[1, 2] * limit_scaler),
+    data_limits[2, 2] + abs(data_limits[2, 2] * limit_scaler),
+    by = 0.1
+  )
+  z_seq <- seq(
+    data_limits[1, 3] - abs(data_limits[1, 3] * limit_scaler),
+    data_limits[2, 3] + abs(data_limits[2, 3] * limit_scaler),
+    by = 0.1
+  )
+
+  point_volume <- 0.1^3
+
+  lpme_candidates  <- expand.grid(x = x_seq, y = y_seq, z = z_seq) |>
+    as.matrix()
+  pme_candidates <- expand.grid(x = x_seq, y = y_seq, z = z_seq) |>
+    as.matrix()
+
+  temp_pme <- sim$models$pme$pme[[time_idx]]
+  temp_pme_embedding <- temp_pme$embedding_map
+  temp_pme_coefs <- temp_pme$coefs[[which.min(temp_pme$MSD)]]
+  temp_pme_params <- temp_pme$parameterization[[which.min(temp_pme$MSD)]]
+
+  lpme_est <- sim$models$lpme$lpme
+  temp_lpme_coefs <- lpme_est$sol_coef_functions[[
+    which.min(lpme_est$msd)
+  ]](time_points[time_idx])
+  temp_lpme_params <- lpme_est$parameterization_list[[which.min(lpme_est$msd)]]
+  temp_lpme_params <- temp_lpme_params[
+    temp_lpme_params[, 1] == time_points[time_idx],
+    -1
+  ]
+  lpme_n_knots <- nrow(temp_lpme_params)
+  d <- ncol(temp_lpme_params)
+  coef_mat <- matrix(
+    temp_lpme_coefs,
+    lpme_n_knots + d + 1,
+    byrow = TRUE
+  )
+
+  temp_lpme_embedding <- function(r) {
+    t(coef_mat[1:lpme_n_knots, ]) %*% pme::etaFunc(r, temp_lpme_params, 4 - d) +
+      t(coef_mat[(lpme_n_knots + 1):(lpme_n_knots + d + 1), ]) %*%
+        matrix(c(1, r), ncol = 1)
   }
-    opt_curve <- which.min(curve_error)
-    # save principal curve results for smoothing approach with the lowest error
-    principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
+
+  lpme_interior_voxels <- interior_identification(
+    temp_lpme_embedding,
+    coef_mat,
+    temp_lpme_params,
+    lpme_candidates,
+    c(0, 0, 0)
+  )
+  pme_interior_voxels <- interior_identification(
+    temp_pme_embedding,
+    temp_pme_coefs,
+    temp_pme_params,
+    pme_candidates,
+    c(0, 0, 0)
+  )
+
+  lpme_interior <- cbind(
+    time_points[time_idx],
+    lpme_candidates[lpme_interior_voxels, ]
+  )
+  lpme_interior_points <- rbind(lpme_interior_points, lpme_interior)
+  lpme_volumes[time_idx] <- nrow(lpme_interior) * point_volume
+
+  pme_interior <- cbind(
+    time_points[time_idx],
+    pme_candidates[pme_interior_voxels, ]
+  )
+  pme_interior_points <- rbind(pme_interior_points, pme_interior)
+  pme_volumes[time_idx] <- nrow(pme_interior) * point_volume
 }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 3-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal curve results,
-  # projections and errors for all models
-  sim_case1 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case1/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case1,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-### Simulation Case 2
-
-```{r}
-sim_error_case2 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 2,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected onto lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal curve estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  # test all smoothing options for principal curve algorithm
-  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 1,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers",
-      verbose = FALSE
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    curves <- list()
-    curve_vals <- list()
-    curve_error <- vector()
-    for (smoother_idx in 1:length(smoothing_options)) {
-      curves[[smoother_idx]] <- principal_curve(
-        temp_data,
-        smoother = smoothing_options[smoother_idx]
-      )
-      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
-      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
-    }
-    opt_curve <- which.min(curve_error)
-    # save principal curve results for smoothing approach with lowest error
-    principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 3-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal curve results,
-  # projections and errors for all models
-  sim_case2 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case2/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case2,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-### Simulation Case 3
-
-```{r}
-sim_error_case3 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 3,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  # compute polar coordinates for the simulated observations
-  # and points on the true manifold
-  pol <- sim_df[, -1] %>%
-    cart2pol()
-  true_pol <- true_vals[, -1] %>%
-    cart2pol()
-  # augment data with polar coordinates, excluding the radius for this case
-  sim_df <- cbind(sim_df, pol)[, -5]
-  true_vals <- cbind(true_vals, true_pol)[, -5]
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal curve estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  # test all smoothing options for principal curve algorithm
-  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 1,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers"
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    curves <- list()
-    curve_vals <- list()
-    curve_error <- vector()
-    for (smoother_idx in 1:length(smoothing_options)) {
-      curves[[smoother_idx]] <- principal_curve(
-        temp_data,
-        smoother = smoothing_options[smoother_idx]
-      )
-      # save projections of observations onto estimated principal curve
-      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
-      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
-    }
-    opt_curve <- which.min(curve_error)
-    # save principal curve results for smoothing approach with the lowest error
-    principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 3-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:3], pme_vals[.x, 1:3])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:3], lpme_vals[.x, 1:3])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:3], sim_df[.x, 1:3])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:3], principal_curve_vals[.x, 1:3])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal curve results,
-  # projections and errors for all models
-  sim_case3 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    # lpme_gp_result = lpme_result_gp,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    # lpme_gp_error = lpme_error_gp,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case3/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case3,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-### Simulation Case 4
-
-```{r}
-sim_error_case4 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 4,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal curve estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  # test all smoothing options for principal curve algorithm
-  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 1,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers"
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    curves <- list()
-    curve_vals <- list()
-    curve_error <- vector()
-    for (smoother_idx in 1:length(smoothing_options)) {
-      curves[[smoother_idx]] <- principal_curve(
-        temp_data,
-        smoother = smoothing_options[smoother_idx]
-      )
-      # save projections of observations onto estimated principal curve
-      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
-      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
-    }
-    opt_curve <- which.min(curve_error)
-    # save principal curve results for smoothing approach with the lowest error
-    principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 4-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 4],
-    frame = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 4],
-      frame = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 4],
-      frame = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 4],
-      frame = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 4],
-      frame = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal curve results,
-  # projections and errors for all models
-  sim_case4 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case4/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case4,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-## Simulation Case 5
-
-```{r}
-sim_error_case5 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 5,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal curve estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  # test all smoothing options for principal curve algorithm
-  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 1,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers"
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    curves <- list()
-    curve_vals <- list()
-    curve_error <- vector()
-    for (smoother_idx in 1:length(smoothing_options)) {
-      curves[[smoother_idx]] <- principal_curve(
-        temp_data,
-        smoother = smoothing_options[smoother_idx]
-      )
-      # save projections of observations onto estimated principal curve
-      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
-      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
-    }
-    opt_curve <- which.min(curve_error)
-    # save principal curve results for smoothing approach with the lowest error
-    principal_curve_result[[t]] <- curves[[opt_curve]]
-    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 4-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 4],
-    frame = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 4],
-      frame = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 4],
-      frame = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 4],
-      frame = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 4],
-      frame = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal curve results,
-  # projections and errors for all models
-  sim_case5 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    # lpme_gp_result = lpme_result_gp,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    # lpme_gp_error = lpme_error_gp,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-
-  sim_dir <- "simulations/case5/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-
-  saveRDS(
-    sim_case5,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-### Simulation Case 6
-
-```{r}
-sim_error_case6 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 6,
-    noise = 0.05,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme_algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    2,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal surface estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 2,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers"
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    principal_surface <- prinSurf(temp_data)
-    # obtain principal surface error estimates
-    surface_mse <- map(
-      1:length(principal_surface),
-      ~ principal_surface[[.x]]$MSE
-    ) %>%
-      unlist()
-    opt_surface <- which.min(surface_mse)
-    # save principal surface results
-    # using opt_surface + 2 as list index because first two list entries are NULL
-    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
-    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 4-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 4],
-    frame = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 4],
-      frame = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 4],
-      frame = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 4],
-      frame = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 4],
-      frame = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # keep principal_curve_error as variable name for continuity
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal surface results,
-  # projections and errors for all models
-  sim_case6 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case6/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case6,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-### Simulation Case 7
-
-```{r}
-sim_error_case7 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 7,
-    noise = 0.15,
-    amp_noise = amp_noise / 25,
-    period_noise = shape_noise / 25,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  # compute polar coordinates for the simulated observations
-  # and points on the true manifold
-  # removing 4th column from consideration removes the depth from the Swiss roll
-  pol <- sim_df[, -c(1, 4)] %>%
-    cart2pol()
-  true_pol <- true_vals[, -c(1, 4)] %>%
-    cart2pol()
-  # augment data with polar coordinates, excluding the radius
-  sim_df <- cbind(sim_df, pol)
-  sim_df <- sim_df[, -5]
-  true_vals <- cbind(true_vals, true_pol)
-  true_vals <- true_vals[, -5]
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    2,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal surface estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 2,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers"
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    principal_surface <- prinSurf(temp_data)
-    # obtain principal surface error estimates
-    surface_mse <- map(
-      1:length(principal_surface),
-      ~ principal_surface[[.x]]$MSE
-    ) %>%
-      unlist()
-    opt_surface <- which.min(surface_mse)
-    # save principal surface results
-    # using opt_surface + 2 as list index because first two list entries are NULL
-    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
-    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 4-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 4],
-    frame = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 4],
-      frame = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 4],
-      frame = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 4],
-      frame = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 4],
-      frame = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], pme_vals[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], lpme_vals[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], sim_df[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # keep principal_curve_error as variable name for continuity
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], principal_curve_vals[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal surface results,
-  # projections and errors for all models
-  sim_case7 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    # lpme_gp_result = lpme_result_gp,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    # lpme_gp_error = lpme_error_gp,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case7/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case7,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-### Simulation Case 8
-
-```{r}
-sim_error_case8 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  # at each time point, simulate a dataset
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 9,
-    noise = 0.05,
-    amp_noise = amp_noise,
-    period_noise = shape_noise / 5,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  # combine the simulated observations into one matrix
-  # do the same with points from the true manifold
-  # these true points correspond to the simulated observations without added noise
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-
-  # compute spherical coordinates for the simulated observations
-  # and points on the true manifold
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-  sph <- sim_df[, -1] %>%
-    cart2sph()
-  true_sph <- true_vals[, -1] %>%
-    cart2sph()
-  # augment data with spherical coordinates
-  sim_df <- cbind(sim_df, sph)
-  true_vals <- cbind(true_vals, true_sph)
-
-  # standardize observations to fall between -1 and 1 in each dimension
-  for (col_idx in 1:ncol(sim_df)) {
-    # results improved in this case by
-    # excluding spherical coordinates from the standardization
-    if (col_idx < 4) {
-      col_max <- max(abs(sim_df[, col_idx]))
-      sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-      true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-    }
-  }
-
-  # update time_vals to use standardized values
-  time_vals <- unique(sim_df[, 1])
-
-  # run lpme algorithm on observations
-  lpme_result <- lpme(
-    sim_df,
-    2,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  # compute values of observations projected to lpme estimated manifold
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # fit pme and principal surface estimates independently for each time point
-  pme_result <- list()
-  pme_vals <- list()
-  principal_curve_result <- list()
-  principal_curve_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(
-      temp_data,
-      d = 2,
-      initialization_algorithm = initialization_alg,
-      initialization_type = "centers"
-    )
-    # save projections of observations onto PME manifold
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-    principal_surface <- prinSurf(temp_data)
-    surface_mse <- map(
-      1:length(principal_surface),
-      ~ principal_surface[[.x]]$MSE
-    ) %>%
-      unlist()
-    opt_surface <- which.min(surface_mse)
-    # save principal surface results
-    # using opt_surface + 2 as list index because first two list entries are NULL
-    principal_curve_result[[t]] <- principal_surface[[opt_surface + 2]]
-    principal_curve_vals[[t]] <- cbind(time_vals[t], principal_surface[[opt_surface + 2]]$PS)
-  }
-  pme_vals <- reduce(pme_vals, rbind)
-  principal_curve_vals <- reduce(principal_curve_vals, rbind)
-
-  # create 4-dimensional plot of data, estimated manifolds, and true manifold
-  p <- plot_ly(
-    x = lpme_vals[, 2],
-    y = lpme_vals[, 3],
-    z = lpme_vals[, 4],
-    frame = lpme_vals[, 1],
-    type = "scatter3d",
-    mode = "markers",
-    marker = list(size = 3),
-    name = "LPME"
-  ) %>%
-    add_markers(
-      x = pme_vals[, 2],
-      y = pme_vals[, 3],
-      z = pme_vals[, 4],
-      frame = pme_vals[, 1],
-      name = "PME"
-    ) %>%
-    add_markers(
-      x = principal_curve_vals[, 2],
-      y = principal_curve_vals[, 3],
-      z = principal_curve_vals[, 4],
-      frame = principal_curve_vals[, 1],
-      name = "Principal Curve"
-    ) %>%
-    add_markers(
-      x = true_vals[, 2],
-      y = true_vals[, 3],
-      z = true_vals[, 4],
-      frame = true_vals[, 1],
-      name = "True Manifold"
-    ) %>%
-    add_markers(
-      x = sim_df[, 2],
-      y = sim_df[, 3],
-      z = sim_df[, 4],
-      frame = sim_df[, 1],
-      name = "Data",
-      opacity = 0.2
-    )
-
-  # for each estimate, compute mean squared distance between
-  # points on the true manifold and the same points projected to
-  # the relevant manifold estimate
-  pme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], pme_vals[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  lpme_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], lpme_vals[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  data_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], sim_df[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # keep principal_curve_error as variable name for continuity
-  principal_curve_error <- map(
-    1:nrow(true_vals),
-    ~ dist_euclidean(true_vals[.x, 1:4], principal_curve_vals[.x, 1:4])^2
-  ) %>%
-    unlist() %>%
-    mean()
-
-  # save list of parameter values, LPME, PME, and principal surface results,
-  # projections and errors for all models
-  sim_case8 <- list(
-    df = sim_df,
-    times = time_vals,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    n = n,
-    lpme_result = lpme_result,
-    pme_results = pme_result,
-    principal_curve_results = principal_curve_result,
-    lpme_error = lpme_error,
-    pme_error = pme_error,
-    data_error = data_error,
-    principal_curve_error = principal_curve_error,
-    true_vals = true_vals,
-    lpme_vals = lpme_vals,
-    pme_vals = pme_vals,
-    principal_curve_vals = principal_curve_vals,
-    plot = p,
-    initialization_algorithm = initialization_alg
-  )
-  sim_dir <- "simulations/case8/"
-  if (!dir.exists(sim_dir)) {
-    dir.create(sim_dir, recursive = TRUE)
-  }
-  filename <- paste0(
-    initialization_alg,
-    "_duration_",
-    str_pad(as.character(max_time), 2, side = "left", pad = "0"),
-    "_interval_",
-    str_pad(as.character(100 * interval), 3, side = "left", pad = "0"),
-    "_ampnoise_",
-    str_pad(as.character(100 * amp_noise), 3, side = "left", pad = "0"),
-    "_pernoise_",
-    str_pad(as.character(100 * shape_noise), 3, side = "left", pad = "0"),
-    "_n_",
-    str_pad(as.character(n), 4, side = "left", pad = "0"),
-    "_",
-    time_trend,
-    str_pad(as.character(100 * time_change), 3, side = "left", pad = "0"),
-    "_run_",
-    str_pad(as.character(run), 2, side = "left", pad = "0"),
-    ".rds"
-  )
-  saveRDS(
-    sim_case8,
-    paste0(sim_dir, filename)
-  )
-  return(0)
-}
-```
-
-Finally, the last code chunks contains the code needed to set the parameters for all simulation runs and run those simulations.
-We consider amplitude and shape noise values of $\{0, 0.05, 0.1, 0.25, 0.5, 1\}$, study durations of $\{1, 2, 5\}$, and intervals between study visits of $\{0.1, 0.25, 0.5\}$.
-Each simulation run has 1,000 observations and every combination of parameters is run one time.
-Finally, we consider three types of change to the underlying manifold (linear, quadratic, and sinusoidal), and the possibility of the manifold remaining constant over time.
-In cases where the manifold changes over time, the following magnitudes of longitudinal change were used: $\{0, 0.05, 0.1, 0.25, 0.5, 1\}$.
-Finally, in addition to using ISOMAP to initialize the PME algorithm, we also considered using Laplacian eigenmaps and diffusion maps.
-All possible combinations of parameter values were considered.
-
-```{r, eval=FALSE}
-# set parameter values
-amp_noise_vals <- c(0, 0.05, 0.1, 0.25, 0.5, 1)
-shape_noise_vals <- c(0, 0.05, 0.1, 0.25, 0.5, 1)
-max_times <- c(1, 2, 5)
-intervals <- c(0.1, 0.25, 0.5)
-n_vals <- 1000
-replicates <- 1
-case <- 1:8
-time_changes <- c(0, 0.05, 0.1, 0.25, 0.5, 1)
-time_trends <- c("constant", "linear", "quadratic", "sinusoidal")
-initialization_algorithms = c(
-  "diffusion_maps",
-  "laplacian_eigenmaps",
-  "isomap"
-)
-
-# create data frame where each row is a combination of parameter values
-param_grid <- expand_grid(
-  case,
-  amp_noise_vals,
-  shape_noise_vals,
-  n_vals,
-  intervals,
-  max_times,
-  replicates,
-  time_changes,
-  time_trends,
-  initialization_algorithms
-) %>%
-  data.frame()
-```
-
-The final chunk below includes the code required to run the simulations.
-This code loops through the rows of the data frame created in the chunk above, running the appropriate simulation function with the parameter values given in the row in question.
-
-```{r, eval=FALSE}
-# run code in parallel using the foreach package
-cl <- makeCluster(detectCores() - 1)
-registerDoParallel(cl)
-
-set.seed(10972)
-errors <- foreach(
-  case = param_grid[, 1],
-  amp_noise = param_grid[, 2],
-  shape_noise = param_grid[, 3],
-  n = param_grid[, 4],
-  interval = param_grid[, 5],
-  duration = param_grid[, 6],
-  initialization_algorithm = param_grid[, 10],
-  run = param_grid[, 7],
-  time_change = param_grid[, 8],
-  time_trend = param_grid[, 9],
-  .inorder = FALSE,
-  # the .export and .packages arguments are used to make functions and packages
-  # available for the additional workers initialized by the parallel cluster.
-  # in my experience, this may take some tinkering with to get everything
-  # running properly
-  .export = c(
-    "sim_data",
-    "calc_pme_est",
-    "calc_lpme_est",
-    "prinSurf",
-    "cart2sph",
-    "cart2pol"
-  ),
-  .packages = c(
-    "tidyverse",
-    "pme",
-    "princurve",
-    "plotly",
-    "doParallel"
-  )
-) %dopar% {
-    # control statements are used to run the appropriate function for each case value
-    if (case == 1) {
-      try({
-        simlist <- sim_error_case1(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      })
-    } else if (case == 2) {
-      try(
-        sim_error_case2(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    } else if (case == 3) {
-      try(
-        sim_error_case3(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    } else if (case == 4) {
-      try(
-        sim_error_case4(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    } else if (case == 5) {
-      try(
-        sim_error_case5(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    } else if (case == 6) {
-      try(
-        sim_error_case6(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    } else if (case == 7) {
-      try(
-        sim_error_case7(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    } else if (case == 8) {
-      try(
-        sim_error_case8(
-          duration,
-          interval,
-          amp_noise,
-          shape_noise,
-          time_change,
-          time_trend,
-          n,
-          initialization_algorithm,
-          run,
-          print_plots = FALSE
-        )
-      )
-    }
-  }
-
-stopCluster(cl)
-```
+```

### 2025-02-10 code-old/adni_hipp_surface_test.R

@@ -1,0 +1,501 @@
+library(cowplot)
+library(gridGraphics)
+library(pracma)
+library(lubridate)
+library(scatterplot3d)
+library(tidyverse)
+library(pme)
+library(Rfast)
+
+source("code/functions/calc_pme_est.R")
+source("code/functions/calc_lpme_est.R")
+
+create_scatter_mat <- function(mat, nrow = 1) {
+  time_points <- unique(mat[, 1])
+  xmin <- min(mat[, 2])
+  xmax <- max(mat[, 2])
+  ymin <- min(mat[, 3])
+  ymax <- max(mat[, 3])
+  zmin <- min(mat[, 4])
+  zmax <- max(mat[, 4])
+
+  # graphics::layout(matrix(c(1:length(time_points)), 1))
+
+  plots <- list()
+  for (time_idx in 1:length(time_points)) {
+    time <- time_points[time_idx]
+    scatterplot3d(
+      x = mat[mat[, 1] == time, 2],
+      y = mat[mat[, 1] == time, 3],
+      z = mat[mat[, 1] == time, 4],
+      main = paste0("Time = ", as.character(round(time, 3))),
+      highlight.3d = TRUE,
+      pch = 20,
+      xlim = c(xmin, xmax),
+      ylim = c(ymin, ymax),
+      zlim = c(zmin, zmax),
+      xlab = "x",
+      ylab = "y",
+      zlab = "z",
+      box = FALSE
+    )
+    plots[[time_idx]] <- recordPlot()
+  }
+  scatter_grid <- plot_grid(plotlist = plots, nrow = nrow)
+  return(scatter_grid)
+}
+
+create_cross_section_matrix <- function(mats, slice_idx, slice_widths, nrow = 1) {
+  time_vals <- unique(mats[[1]][, 1])
+  slices <- list()
+  for (time_idx in 1:length(time_vals)) {
+    slice_list <- list()
+    for (idx in 1:length(mats)) {
+      mat <- mats[[idx]]
+      temp_mat <- mat[mat[, 1] == time_vals[time_idx], ]
+      n_pixels <- length(unique(temp_mat[, slice_idx]))
+      slice_mean <- mean(temp_mat[, slice_idx])
[…]

### 2025-02-10 code-old/adni_mri_preprocess.R

@@ -1,0 +1,527 @@
+library(oro.dicom)
+library(oro.nifti)
+library(neurobase)
+library(fslr)
+# library(extrantsr)
+library(stringr)
+library(foreach)
+library(tidyverse)
+library(ANTsRCore)
+
+sub_dirs <- list.dirs("data/adni", recursive = FALSE)
+
+nii_scans <- list()
+
+processed_scans <- list()
+
+hipp_mask <- readNIfTI(
+  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm_Hipp_mask_dil8.nii.gz")
+)
+
+mni_nifti <- readNIfTI(
+  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm.nii.gz")
+)
+
+ncores <- parallel::detectCores() / 2
+cl <- parallel::makeCluster(ncores, type = "FORK")
+doParallel::registerDoParallel(cl = cl)
+
+foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
+# foreach(dir_idx = 1:8) %dopar% {
+  img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
+    gsub(pattern = "([^/]+$)", replacement = "") %>%
+    unique()
+
+  nii_scans[[dir_idx]] <- list()
+
+  for (img_idx in 1:length(img_dirs)) {
+    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/adni", replacement = "", img_dirs[img_idx]))
+    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
+      next
+    } else {
+      all_slices <- readDICOM(img_dirs[img_idx])
+      nii <- dicom2nifti(all_slices)
+
+      # Inhomogeneity correction
+      # Brain extraction / segmentation
+      # Image Registration
+      # Tissue Class Segmentation
+
+      # nii_scans[[dir_idx]][[img_idx]] <- nii
+
+      nii <- fslreorient2std(nii, verbose = FALSE) # orient nii to mni space
+      # inhomogeneity correction
+      # nii <- bias_correct(
+      #   nii,
+      #   correction = "N4",
+      #   verbose = FALSE
+      # ) # N4 bias correction using ANTs
+      # nii_bc <- fast(nii, bias_correct = TRUE)
+
+      # nii <- fast(nii, bias_correct = TRUE)
[…]

### 2025-02-10 code-old/functions/calc_lpme_est.R

@@ -1,0 +1,40 @@
+calc_lpme_est <- function(result, df) {
+  opt_idx <- which.min(result$msd)
+  r_full <- expand_grid(result$times, result$initial_parameterization) %>%
+    as.matrix()
+  # x_vals <- map(
+  #   1:nrow(r_full),
+  #   ~ embed(result, r_full[.x, ])
+  # ) %>%
+  #   reduce(rbind)
+  x_vals <- map(
+    1:nrow(r_full),
+    ~ result$embedding_map(r_full[.x, ])
+  ) %>%
+    reduce(rbind)
+  tnew_init <- result$optimal_parameterization
+
+  nearest_x <- map(
+    1:nrow(df),
+    ~ which.min(apply(x_vals[x_vals[, 1] == df[.x, 1], ], 1, dist_euclidean, y = df[.x, ]))
+  ) %>%
+    reduce(c)
+
+  init_param <- map(
+    1:nrow(df),
+    ~ tnew_init[tnew_init[, 1] == df[.x, 1], ][nearest_x[.x], ]
+  ) %>%
+    reduce(rbind)
+
+  tnew <- map(
+    1:nrow(df),
+    ~ projection_lpme(df[.x, ], function(x) result$embedding_map(x), init_param[.x, ])
+  ) %>%
+    reduce(rbind)
+  results <- map(
+    1:nrow(df),
+    ~ result$embedding_map(tnew[.x, ])
+  ) %>%
+    reduce(rbind)
+  return(results)
+}

### 2025-02-10 code-old/functions/calc_pme_est.R

@@ -1,0 +1,26 @@
+calc_pme_est <- function(pme, df) {
+  center_order <- order(pme$knots$centers[, 1])
+  centers <- pme$knots$centers[center_order, ]
+
+  nearest_clusters <- map(
+    1:nrow(df),
+    ~ which.min(as.vector(apply(centers, 1, dist_euclidean, y = df[.x, ])))
+  ) %>%
+    reduce(c)
+
+  opt_run <- which.min(pme$MSD)
+  tnew_init <- pme$parameterization[[opt_run]][nearest_clusters, ] %>%
+    matrix(nrow = nrow(df), byrow = TRUE)
+
+  tnew <- map(
+    1:nrow(df),
+    ~ projection_pme(df[.x, ], pme$embedding_map, tnew_init[.x, ])
+  ) %>%
+    reduce(rbind)
+  results <- map(
+    1:nrow(df),
+    ~ pme$embedding_map(tnew[.x, ])
+  ) %>%
+    reduce(rbind)
+  return(results)
+}

### 2025-02-10 code-old/functions/create_cross_section_matrix.R

@@ -1,0 +1,81 @@
+# function for creating cross-section plots
+create_cross_section_matrix <- function(mats, slice_idx, slice_widths, nrow = 1, lhipp) {
+  time_vals <- unlist(mats[[1]][, 1]) %>% 
+    unique()
+  slices <- list()
+  for (time_idx in 1:length(time_vals)) {
+    slice_list <- list()
+    for (idx in 1:length(mats)) {
+      mat <- mats[[idx]]
+      temp_mat <- mat[almost.equal(mat[, 1], time_vals[time_idx]), ] %>% 
+        as.matrix()
+      n_pixels <- length(unique(temp_mat[, slice_idx]))
+      slice_mean <- mean(temp_mat[, slice_idx])
+      # in the PME- and LPME-generated value matrices, voxel values are not all
+      # consistent as in the data
+      # to approximate a cross-section through the middle of the structure, we
+      # consider all voxel values within a certain range of the mean voxel value.
+      slice_list[[idx]] <- temp_mat[(temp_mat[, slice_idx] >= slice_mean - slice_widths[idx]) &
+                                      (temp_mat[, slice_idx] <= slice_mean + slice_widths[idx]), ] %>%
+        cbind(idx)
+    }
+    slices[[time_idx]] <- reduce(slice_list, rbind)
+  }
+  mat_sliced <- reduce(slices, rbind)
+
+  slice_df <- data.frame(mat_sliced)
+  names(slice_df) <- c("time", "x", "y", "z", "theta", "phi", "r", "Source")
+  slice_df <- slice_df %>%
+    mutate(
+      # time = time * unique(lhipp$duration),
+      time = round(time, 2),
+      time = as.factor(time),
+      Source = ifelse(
+        Source == 1,
+        "Data",
+        ifelse(
+          Source == 2,
+          "LPME",
+          "PME"
+        )
+      ),
+      # code to change opacity by data source if needed
+      opacity = case_when(
+        Source == "Data" ~ 1,
+        Source == "LPME" ~ 1,
+        Source == "PME" ~ 1
+      )
+    )
+
+  slice_df <- arrange(slice_df, Source)
+  slice_df <- slice_df %>%
+    mutate(
+      Source = reorder(
+        Source,
+        X = c(
+          rep(1, nrow(mat_sliced[mat_sliced[, 8] == 1, ])),
[…]

### 2025-02-10 code-old/functions/dist_euclidean.R

@@ -1,0 +1,4 @@
+dist_euclidean <- function(x, y) {
+  # Distance function in a Euclidean space of any dimension
+  return(norm_euclidean(x - y))
+}

### 2025-02-10 code-old/functions/estimate_volume.R

@@ -1,0 +1,298 @@
+estimate_volume <- function(data, voxel_volume) {
+  indices <- matrix(ncol = 3)
+  data_limits <- colMinsMaxs(data)
+
+  x_seq <- seq(data_limits[1, 1], data_limits[2, 1])
+  y_seq <- seq(data_limits[1, 2], data_limits[2, 2])
+  z_seq <- seq(data_limits[1, 3], data_limits[2, 3])
+
+  # fill_candidates <- expand.grid(x_seq, y_seq, z_seq)
+  fill_candidates <- matrix(ncol = 3)
+  x_mins <- matrix(nrow = length(y_seq), ncol = length(z_seq))
+  x_maxs <- matrix(nrow = length(y_seq), ncol = length(z_seq))
+
+  for (y_idx in 1:length(y_seq)) {
+    for (z_idx in 1:length(z_seq)) {
+      y_val <- y_seq[y_idx]
+      z_val <- z_seq[z_idx]
+      data_red <- data[data[, 2] == y_val & data[, 3] == z_val, ] %>%
+        matrix(ncol = ncol(data))
+
+      if (nrow(data_red) == 0) {
+        x_mins[y_idx, z_idx] <- NA
+        x_maxs[y_idx, z_idx] <- NA
+      } else if (nrow(data_red) > 1) {
+        min_val <- min(unique(data_red[, 1]))
+        max_val <- max(unique(data_red[, 1]))
+        if (max_val - min_val > 1) {
+          x_mins[y_idx, z_idx] <- min(unique(data_red[, 1]))
+          x_maxs[y_idx, z_idx] <- max(unique(data_red[, 1]))
+        }
+      } else {
+        x_mins[y_idx, z_idx] <- NA
+        x_maxs[y_idx, z_idx] <- NA
+      }
+    }
+  }
+
+  for (y_idx in 1:length(y_seq)) {
+    for (z_idx in 1:length(z_seq)) {
+      y_val <- y_seq[y_idx]
+      z_val <- z_seq[z_idx]
+      data_red <- data[data[, 2] == y_val & data[, 3] == z_val, ] %>%
+        matrix(ncol = ncol(data))
+
+      if (nrow(data_red) == 0) {
+        next
+      } else if (nrow(data_red) > 1) {
+        next
+      } else {
+        x_min_vals <- get_surrounding_vals(x_mins, y_idx, z_idx)
[…]

### 2025-02-10 code-old/functions/hdmde.R

@@ -1,0 +1,120 @@
+hdmde <- function(x.obs, N0, alpha, max.comp) {
+
+  # "x.obs" is the data set of interest.
+  # There are n observations, and each observation is a D-dimensional point.
+  # x.obs is a n-by-D matrix.
+  # "N0" is a predetermined lower bound for N - the number of density components
+  # "alpha" is the predetermined confidence level.
+  # "max.comp" is the upper bound of the number of components in the desired mixture density.
+
+  zalpha <- qnorm(1 - alpha / 2)
+  n.D <- dim(x.obs) # "x.obs" is a n by D matrix. Each row of it denotes a data point in R^D.
+  n <- n.D[1]
+  D <- n.D[2]
+  N <- N0
+
+  km <- kmeans(x.obs, N, iter.max = 10000, nstart = 100) # Use k-means clustering to get N clusters of x.obs.
+  mu <- km$centers # Set the centers of the N clusters as the means of the density components.
+
+  sigma.vec <- rep(NA, N) # The following block estimates \sigma_N.
+  for(j in 1:N) {
+    index.temp <- which(km$cluster == j)
+    xi.j <- matrix(x.obs[index.temp, ], nrow = length(index.temp))
+    sig.prepare <- function(x) {
+      return((dist_euclidean(x, mu[j, ])) ^ 2)
+    }
+    s <- apply(xi.j, 1, sig.prepare)
+    sigma.vec[j] <- mean(s)
+  }
+  sig <- sqrt(mean(sigma.vec) / dim(x.obs)[2])
+
+  # It gives an estimation of theta_j's with weight.seq().
+  theta.hat <- weight_seq(x.obs, mu, sig)
+
+  # The following block gives an approximation to the underlying density function of interest.
+  # This estimation is of the form of weights times scaled kernels.
+  f.test <- function(x) {
+    fun.prepare <- function(t) {
+      return(ker(x, t, sig))
+    }
+    comp.vec <- apply(mu, 1, fun.prepare)
+    return(sum(theta.hat * comp.vec))
+  }
+
+  p.old <- apply(x.obs,1,f.test) # The second/negative term p_N in Delta.hat.
+
+  test.rejection <- 1
+  while ((test.rejection == 1) & (N <= min(n, max.comp))) {
+
+    if (max.comp > N0) {
+      N <- N + 1
+    }
+
+    ##################################################
[…]

### 2025-02-10 code-old/functions/interior_identification.R

@@ -1,0 +1,35 @@
+interior_identification <- function(embedding_map, coefs, params, x, reference) {
+  orientation_c <- get_orientation(embedding_map, coefs, params, reference)[1]
+  orientation_x <- map(
+    1:nrow(x), 
+    ~ get_orientation(embedding_map, coefs, params, unlist(x[.x, ]))
+  ) %>% 
+    unlist()
+  return(orientation_x == orientation_c)
+}
+
+  get_orientation <- function(embedding_map, coefs, params, x) {
+  d <- ncol(params)
+  n <- nrow(params)
+
+  augment_x <- c(x, cart2sph(x))
+  x_params <- projection_pme(augment_x, embedding_map, params[1, ])
+  x_proj <- embedding_map(x_params)
+
+  
+  param_diff <- -1 * sweep(params, 2, x_params)
+  param_diff_norms <- wordspace::rowNorms(param_diff)
+
+  jacobian <- t(coefs[1:n, ]) %*% (param_diff * ((2 * log(param_diff_norms)) + 1)) +
+    t(coefs[(n + 1):(n + d + 1), ][-1, ])
+
+  normal_vector <- c(
+    (jacobian[2, 1] * jacobian[3, 2]) - (jacobian[3, 1] * jacobian[2, 2]),
+    (jacobian[3, 1] * jacobian[1, 2]) - (jacobian[1, 1] * jacobian[3, 2]),
+    (jacobian[1, 1] * jacobian[2, 2]) - (jacobian[2, 1] * jacobian[1, 2])
+  ) %>%
+    matrix(ncol = 1)
+
+  orientation <- sign(t(matrix(x_proj - augment_x, nrow = 1)[, 1:3]) %*% normal_vector)
+  return(orientation)
+}

### 2025-02-10 code-old/functions/lpme_init.R

@@ -1,0 +1,125 @@
+lpme_init <- function(df, d, alpha = 0.05, max.comp = 100, epsilon = 0.05, max.iter = 100, init = "first") {
+
+  # init argument can take three values:
+  # "first" indicates that an isomap initialization run on the first time point is used
+  # "full" indicates that an isomap initialization run on the full dataset (all time points) is used
+  # "separate" indicates that each time point will use its own isomap initialization
+
+  source("code/pme.R")
+  require(plotly)
+
+  time_points <- df[, 1] %>%
+    unique()
+
+  pme_results <- list()
+  funcs <- list()
+  x_test <- list()
+  r <- list()
+
+  msd <- list()
+
+  if (init %in% c("first", "full")) {
+    if (init == "first") {
+      init_df <- df[df[, 1] == time_points[1], -1]
+    } else if (init == "full") {
+      init_df <- df[, -1]
+    }
+
+    init_dimension.size <- dim(init_df)
+    init_D <- init_dimension.size[2]
+    init_n <- init_dimension.size[1]
+    lambda <- 4 - d
+
+    init_N0 <- 20 * init_D
+    init_est <- hdmde(init_df, init_N0, alpha, max.comp)
+    init_order <- order(init_est$mu[, 1])
+    init_theta.hat <- init_est$theta.hat[init_order]
+    init_centers <- init_est$mu[init_order, ]
+    init_sigma <- init_est$sigma
+    init_W <- diag(init_theta.hat)
+    init_X <- init_est$mu[init_order, ]
+    init_I <- length(init_theta.hat)
+
+    init_dissimilarity.matrix <- as.matrix(dist(init_X))
+    init_isomap <- isomap(init_dissimilarity.matrix, ndim = d, k = 10)
+  }
+
+  for (idx in 1:length(time_points)) {
+    df_temp <- df[df[, 1] == time_points[idx], ]
+    if (init %in% c("first", "full")) {
+      pme_results[[idx]] <- pme(
+        x.obs = df_temp[, -1],
+        d = d,
+        initialization = list(init_est, init_isomap)
+      )
+    } else {
+      pme_results[[idx]] <- pme(
[…]

### 2025-02-10 code-old/functions/plot_lpme.R

@@ -1,0 +1,102 @@
+plot_lpme <- function(x.obs, f, tnew, d_new, D_new, time_points) {
+  time_vals <- seq(
+    min(time_points),
+    max(time_points),
+    0.05
+  )
+  # pred_grid <- calc_tnew(centers, tnew, sol, I, d_new, gamma)
+  # r_bounds <- colMinsMaxs(pred_grid)
+  # r_list <- list()
+  # for (idx in 1:dim(r_bounds)[2]) {
+  #   r_list[[idx]] <- seq(
+  #     r_bounds[1, idx],
+  #     r_bounds[2, idx],
+  #     length.out = sqrt(nrow(centers))
+  #   )
+  # }
+  # r_mat <- as.matrix(expand.grid(r_list))
+  # pred_grid <- r_mat
+  new_pred <- matrix(ncol = ncol(tnew) + 1)
+  for (time in time_vals) {
+    new_pred <- rbind(new_pred, cbind(time, tnew))
+  }
+  pred_grid <- new_pred[-1, ]
+  f_pred <- map(
+    1:nrow(pred_grid),
+    ~ f(pred_grid[.x, ])
+  ) %>%
+    reduce(rbind)
+
+  f_pred_full <- cbind(pred_grid, f_pred)
+
+  # if (D_new == 3) {
+  if (D_new == 2) {
+    plt <- plot_ly(
+      x = f_pred_full[, d_new + 3],
+      y = f_pred_full[, d_new + 4],
+      # x = f_pred_full[, d_new + 1],
+      # y = f_pred_full[, d_new + 2],
+      z = f_pred_full[, 1],
+      type = "scatter3d",
+      mode = "markers",
+      marker = list(
+        size = 3
+      )
+    ) %>%
+      add_markers(
+        x = x.obs[, 2],
+        y = x.obs[, 3],
+        z = x.obs[, 1],
+        opacity = 0.15
+      )
+    print(plt)
+  } else {
+    if (D_new >= 3) {
+    # if (D_new >= 4) {
+    plt <- plot_ly(
+      x = f_pred_full[, d_new + 3],
+      # x = f_pred_full[, d_new + 1],
+      y = f_pred_full[, d_new + 4],
+      # y = f_pred_full[, d_new + 2],
+      z = f_pred_full[, d_new + 5],
+      # z = f_pred_full[, d_new + 3],
+      frame = f_pred_full[, d_new + 2],
+      # frame = f_pred_full[, d_new],
[…]

### 2025-02-10 code-old/functions/plot_lpme_complete.R

@@ -1,0 +1,9 @@
+plot_lpme_complete <- function(df, result) {
+  opt_result <- which.min(result$MSD)
+  # t_new <- result$TNEW[[opt_result]]
+  coefs <- result$coefs[[opt_result]]
+  d_new <- ncol(coefs)
+  D_new <- ncol(df) - 1
+  time_points <- unique(df[,  1])
+  plot_lpme(df, result$embedding_map, coefs, d_new, D_new, time_points)
+}

### 2025-02-10 code-old/functions/plot_pme.R

@@ -1,0 +1,66 @@
+plot_pme <- function(x.obs, centers, sol, tnew, I, d, lambda) {
+  pred_grid <- calc_tnew(centers, tnew, sol, I, d, lambda)
+  r_bounds <- colMinsMaxs(pred_grid)
+  r_list <- list()
+  for (idx in 1:dim(r_bounds)[2]) {
+    r_list[[idx]] <- seq(
+      r_bounds[1, idx],
+      r_bounds[2, idx],
+      length.out = nrow(centers)
+    )
+  }
+  r_mat <- as.matrix(expand.grid(r_list))
+
+  pred_grid <- r_mat
+  f_pred <- map(
+    1:nrow(pred_grid),
+    ~ fNew(
+      unlist(as.vector(pred_grid[.x, ])),
+      sol,
+      tnew,
+      I,
+      d,
+      lambda
+    ) %>%
+      as.vector()
+  ) %>%
+    unlist() %>%
+    matrix(nrow = nrow(pred_grid), byrow = TRUE)
+
+  f_pred_full <- cbind(pred_grid, f_pred)
+
+  if (dim(x.obs)[2] == 2) {
+    plt <- ggplot() +
+      geom_point(
+        aes(
+          x =x.obs[, 1],
+          y = x.obs[, 2]
+        ),
+        alpha = 0.5
+      ) +
+      geom_point(
+        aes(
+          x = f_pred_full[, d + 1],
+          y = f_pred_full[, d + 2]
+        ),
+        color = "red"
+      )
+    print(plt)
+  } else if (dim(x.obs)[2] >= 3) {
+    plt <- plot_ly(
+      x = f_pred_full[, d + 1],
+      y = f_pred_full[, d + 2],
+      z = f_pred_full[, d + 3],
+      type = "scatter3d",
+      mode = "markers",
+      opacity = 0.5
+    ) %>%
+      add_markers(
+        x = x.obs[, 1],
+        y = x.obs[, 2],
+        z = x.obs[, 3],
+        opacity = 0.15
+      )
+    print(plt)
+  }
+}

### 2025-02-10 code-old/functions/pme_functions.cpp

@@ -1,0 +1,195 @@
+#include <RcppCommon.h>
+#include <dlib/optimization/optimization.h>
+
+#include <RcppArmadillo.h>
+// [[Rcpp::depends(RcppArmadillo)]]
+using namespace Rcpp;
+
+// [[Rcpp::depends(dlib)]]
+//#include <dlib/optimization.h>
+
+
+using namespace std;
+using namespace dlib;
+
+
+#include <numeric>
+#include <vector>
+#include <cmath>
+#include <complex>
+
+// [[Rcpp::export]]
+double norm_euclideanC(arma::vec x) {
+  double sq_sum = 0;
+  for (int i = 0; i < x.size(); i++) {
+    sq_sum += pow(x[i], 2);
+  }
+  return sqrt(sq_sum);
+}
+
+// [[Rcpp::export]]
+double dist_euclideanC(arma::vec x, arma::vec y) {
+  return norm_euclideanC(x - y);
+}
+
+// [[Rcpp::export]]
+double dist_euclideanF(arma::vec x, arma::vec y, Function f) {
+  NumericVector mod_x = f(x);
+  return dist_euclideanC(mod_x, y);
+}
+
+// [[Rcpp::export]]
+arma::vec dist_euclideanC_vec(arma::mat x, arma::mat y) {
+  arma::vec distances(x.n_rows);
+  for (int i = 0; i < x.n_rows; i++) {
+    distances(i) = dist_euclideanC(x.row(i).t(), y.row(i).t());
+  }
+  return distances;
+}
+
+// [[Rcpp::export]]
+double eta_kernelC(arma::vec t, int lambda) {
+
+  double lambda_num = lambda / 1.0;
+  double norm_val = norm_euclideanC(t);
+  double y;
+  if (lambda % 2 == 0) {
+    if (norm_val == 0) {
+      y = 0;
+    } else {
+      y = pow(norm_val, lambda_num) * log(norm_val);
+    }
+  } else {
+    y = pow(norm_val, lambda_num);
+  }
+  return y;
+}
+
+// [[Rcpp::export]]
+arma::mat calcE(arma::mat x, int lambda) {
+  int nrow = x.n_rows;
+  int ncol = x.n_cols;
+  arma::mat E(nrow, nrow);
+
+  for (int i = 0; i < nrow; i++) {
+    for (int j = 0; j < nrow; j++) {
[…]

### 2025-02-10 code-old/functions/projection.R

@@ -1,0 +1,10 @@
+projection <- function(x, f, initial.guess) {
+  est <- nlm(function(t) dist_euclideanC(x = x, f(t)), p = initial.guess)
+  # opts <- list("algorithm" = "NLOPT_LN_COBYLA", "xtol_rel" = 1e-04)
+  # est <- nloptr(
+  #   x0 = initial.guess,
+  #   function(t) dist_euclideanC(x = x, f(t)),
+  #   opts = opts
+  # )
+  return(est$estimate)
+}

### 2025-02-10 code-old/functions/projection_lpme.R

@@ -1,0 +1,25 @@
+projection_lpme <- function(x, f, initial.guess) {
+  nlm_est <- try(
+    nlm(
+      function(t) dist_euclideanC(x = x, f(matrix(c(initial.guess[1], t), nrow = 1))),
+      p = initial.guess[-1]
+    )
+  )
+  if (class(nlm_est) == "try-error") {
+    opts <- list("algorithm" = "NLOPT_LN_COBYLA", "xtol_rel" = 1e-07)
+    nlopt_est <- try(
+      nloptr(
+        x0 = initial.guess[-1],
+        function(t) dist_euclideanC(x = x, f(c(initial.guess[1], t))),
+        opts = opts
+      )
+    )
+    if (class(nlopt_est) == "try-error") {
+      return(NULL)
+    } else {
+      return(c(initial.guess[1], nlopt_est$solution))
+    }
+  } else {
+    return(c(initial.guess[1], nlm_est$estimate))
+  }
+}

### 2025-02-10 code-old/functions/sim_data.R

@@ -1,0 +1,162 @@
+sim_data <- function(time_val, case, noise, amp_noise, period_noise, time_change, time_trend, N = 1000) {
+  require(pme)
+  if (time_trend == "constant") {
+    time_trend_val <- 0
+  } else if (time_trend == "linear") {
+    time_trend_val <- time_val
+  } else if (time_trend == "quadratic") {
+    time_trend_val <- time_val^2
+  } else if (time_trend == "sinusoidal") {
+    time_trend_val <- sin(time_val)
+  }
+  time_change_d2 <- rep(time_change * time_trend_val, 2)
+  time_change_d3 <- rep(time_change * time_trend_val, 3)
+  manifolds <- list(
+    function(tau, amp_noise, period_noise) {
+      c(tau, amp_noise[1] * sin(period_noise[1] * tau + pi / 2)) +
+        time_change_d2
+    },
+    function(tau, amp_noise, period_noise) {
+      c(tau, amp_noise[1] * sin(period_noise[1] * tau)) +
+        time_change_d2
+    },
+    function(tau, amp_noise, period_noise) {
+      c(
+        amp_noise[1] * cos(period_noise[1] * tau),
+        amp_noise[2] * sin(period_noise[2] * tau)
+      ) +
+        time_change_d2
+    },
+    function(tau, amp_noise, period_noise) {
+      c(
+        tau,
+        (tau * amp_noise[1] + period_noise[1])^2,
+        (tau * amp_noise[2] + period_noise[2])^3
+      ) +
+        time_change_d3
+    },
+    function(tau, amp_noise, period_noise) {
+      c(
+        tau,
+        amp_noise[1] * cos(period_noise[1] * tau),
+        amp_noise[2] * sin(period_noise[2] * tau)
+      ) +
+        time_change_d3
+    },
+    function(tau, amp_noise, period_noise) {
+      c(
+        (period_noise[1] / 2) * tau[1],
+        (period_noise[2] / 2) * tau[2],
+        (amp_noise[1]) * ((amp_noise[2]) * norm_euclidean(period_noise[1:d] * tau)^2)
+      ) +
+        time_change_d3
+    },
[…]

### 2025-02-10 code-old/functions/solve_eq.R

@@ -1,0 +1,28 @@
+solve_eq <- function(E, W, t_val, X, w, d, D) {
+  M1 <- cbind(
+    2 * E %*% W %*% E + 2 * w * E,
+    2 * E %*% W %*% t_val,
+    t_val
+  )
+  M2 <- cbind(
+    2 * t(t_val) %*% W %*% E,
+    2 * t(t_val) %*% W %*% t_val,
+    matrix(0, ncol = d + 1, nrow = d + 1)
+  )
+  M3 <- cbind(
+    t(t_val),
+    matrix(0, ncol = d + 1, nrow = d + 1),
+    matrix(0, ncol = d + 1, nrow = d + 1)
+  )
+  M <- rbind(M1, M2, M3) # The coefficient matrix of the linear equations
+
+  # The nonhomogeneous term of the linear equations
+  b <- rbind(
+    2 * E %*% W %*% X,
+    2 * t(t_val) %*% W %*% X,
+    matrix(0, nrow = d + 1, ncol = D)
+  )
+  sol <- ginv(M) %*% b # Solve the linear equations
+  # sol <- solve(M, tol = 1e-100) %*% b
+  return(sol)
+}

### 2025-02-10 code-old/functions/solve_eq2.R

@@ -1,0 +1,8 @@
+solve_eq2 <- function(E, T_val, Y, I, w, d, D) {
+  M1 <- cbind(E + (w * diag(rep(1, I))), T_val)
+  M2 <- cbind(t(T_val), matrix(0, ncol = d + 1, nrow = d + 1))
+  M <- rbind(M1, M2)
+  b <- rbind(Y, matrix(0, nrow = d + 1, ncol = D))
+  sol <- ginv(M) %*% b
+  return(sol)
+}

### 2025-02-10 code-old/functions/weight_seq.R

@@ -1,0 +1,64 @@
+weight_seq <- function(x.obs, mu, sigma, epsilon = 0.001, max.iter = 1000) {
+  # x.obs is the dataset of interest
+  # there are n observations, and each observation is a D-dimensional point
+  # x.obs is a n-by-D matrix
+
+  # mu is a vector of the knots in a mixture density estimation
+  # sigma is the bandwidth of the density estimation
+  # epsilon is a predetermined tolerance of the Euclidean distance between thetas in two consecutive steps
+  # max.iter is a predetermined upper bound on the number of steps
+
+  n.D <- dim(x.obs)
+  n <- n.D[1]
+  D <- n.D[2]
+  N <- dim(mu)[1]
+
+  A <- matrix(NA, nrow = n, ncol = N)
+  for (j in 1:N) {
+    A.prepare <- function(x) {
+      return(ker(x, mu[j, ], sigma))
+    }
+    A[, j] <- apply(x.obs, 1, A.prepare)
+  }
+
+  theta.old <- rep(1 / N, N)
+  abs.diff <- 10 * epsilon
+  count <- 0
+  lambda.hat.old <- c(n, rep(-1, D))
+
+  while((abs.diff > epsilon) & (count <= max.iter)) {
+    W <- t(t(A) * theta.old)
+    W <- W / rowsums(W)
+    w <- colsums(W)
+
+    flambda <- function(lambda) {
+
+      denom.temp <- rowsums(t(t(cbind(rep(1, dim(mu)[1]), mu)) * lambda))
+      num.temp <- mu * w
+
+      f1 <- sum(w / denom.temp)
+      f2 <- colsums(num.temp * (as.vector(1 / denom.temp)))
+      f <- dist_euclidean(f1, 1) + dist_euclidean(f2, apply(x.obs, 2, mean))
+      return(f)
+    }
+
+    lambda.hat <- nlm(flambda, lambda.hat.old, iterlim = 1000)$estimate
+
+    theta.new <- w / rowsums(t(t(cbind(rep(1, dim(mu)[1]), mu)) * lambda.hat))
+
+    abs.diff <- dist_euclidean(theta.new, theta.old)
+    if (is.na(abs.diff)) {
+      abs.diff <- 0
+      theta.new <- theta.old
+    }
+
+    theta.old <- pmax(theta.new, 0)
+    theta.old <- pmin(theta.old, 1)
+    count <- count + 1
[…]

### 2025-02-10 code-old/gen_lpme_figures.R

@@ -1,0 +1,406 @@
+library(tidyverse)
+library(plotly)
+library(pracma)
+library(foreach)
+library(doParallel)
+library(furrr)
+library(progress)
+library(princurve)
+library(devtools)
+library(plot3D)
+library(latex2exp)
+
+load_all()
+
+source("~/Documents/brown/research/lpme-project/code/functions/sim_data.R")
+source("~/Documents/brown/research/lpme-project/code/functions/calc_pme_est.R")
+source("~/Documents/brown/research/lpme-project/code/functions/calc_lpme_est.R")
+source("~/Documents/brown/research/lpme-project/code/prinSurf_v3.R")
+
+lambda <- exp(-15:5)
+
+max_time <- 1
+interval <- 0.25
+amp_noise <- 0.5
+shape_noise <- 0.25
+n <- 1000
+time_change <- 0
+time_trend <- "constant"
+
+set.seed(1000)
+
+time_vals <- seq(0, max_time, interval)
+sim_list <- lapply(
+  time_vals,
+  sim_data,
+  case = 1,
+  noise = 0.15,
+  amp_noise = amp_noise,
+  period_noise = shape_noise,
+  N = n,
+  time_change = time_change,
+  time_trend = time_trend
+)
+
+sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
+true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
+for (i in 1:length(sim_list)) {
+  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
+  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
+}
+sim_df <- sim_df[-1, ]
+true_vals <- true_vals[-1, ]
+
+png(
+  "~/Documents/brown/research/lpme-project/paper/figures/sim_case1_plot2.png",
+  width = 2500,
+  height = 2500,
+  res = 500
+)
+scatter3D(
+  x = sim_df[, 1],
+  y = sim_df[, 2],
+  z = sim_df[, 3],
+  pch = 20,
+  col = alpha("black", 0.15),
+  colkey = FALSE,
+  # xlab = "",
+  # ylab = "",
+  # zlab = ""
+  xlab = "Time",
+  ylab = "",
+  zlab = "",
+  cex = 0.5
+)
+text3D(
+  1.17, 
+  -2.5, 
+  0, 
+  labels = expression(X[1]), 
+  add = TRUE
+)
+text3D(
+  -1.25, 
+  -1.25, 
[…]

### 2025-02-10 code-old/generate_figures.R

@@ -1,0 +1,629 @@
+library(plot3D)
+library(plotly)
+library(pracma)
+library(RColorBrewer)
+library(scatterplot3d)
+library(tidyverse)
+
+set.seed(1202)
+theta <- runif(1000, min = 0, max = 2 * pi)
+x <- cos(theta)
+y <- sin(theta)
+
+unit_df <- data.frame(theta, x, y) %>%
+  arrange(theta)
+angle <- cart2pol(as.matrix(unit_df[, 2:3], ncol = 2))
+
+
+png("paper/figures/unit_circle_augmentation.png", res = 500, height = 2000, width = 2000)
+par(mfrow = c(1, 2))
+plot(
+  x = unit_df$x,
+  y = unit_df$y,
+  type = "l",
+  asp = 1,
+  main = "Unit Circle, D = 2",
+  xlab = "x",
+  ylab = "y"
+)
+
+plt2 <- lines3D(
+  x = unit_df$x,
+  y = unit_df$y,
+  z = angle[, 1],
+  main = "Unit Circle, D = 3",
+  xlab = "x",
+  ylab = "y",
+  zlab = ""
+)
+text3D(-2, -0.8, -2.25, labels = expression(theta), add = TRUE)
+dev.off()
+
+png("paper/figures/unit_circle_D2.png", res = 500, height = 2000, width = 2000)
+plot(
+  x = unit_df$x,
+  y = unit_df$y,
+  type = "l",
+  asp = 1,
+  main = "Unit Circle, D = 2",
+  xlab = "x",
+  ylab = "y"
+)
+dev.off()
+
+png("paper/figures/unit_circle_D3.png", res = 500, height = 2000, width = 2000)
+lines3D(
+  x = unit_df$x,
+  y = unit_df$y,
+  z = angle[, 1],
+  main = "Unit Circle, D = 3",
+  xlab = "x",
+  ylab = "y",
+  zlab = ""
+)
+text3D(-2.25, -0.85, -2.25, labels = expression(theta), add = TRUE)
+dev.off()
+
+sim_example_case1 <- readRDS("simulations/case1/duration_01_interval_010_ampnoise_050_pernoise_050_n_1000_constant025_run_01.rds")
+sim_example_case5 <- readRDS("simulations/case6/duration_01_interval_010_ampnoise_005_pernoise_000_n_1000_constant010_run_01.rds")
[…]

### 2025-02-10 code-old/lpme.R

@@ -1,0 +1,706 @@
+lpme <- function(df, d, tuning.para.seq = c(0, exp(-15:10)), alpha = 0.05, max.comp = 500, epsilon = 0.05, max.iter = 100, verbose = "all", print_plots = TRUE, SSD_ratio_threshold = 100, increase_threshold = 1.05, init = "full") {
+  # df is an N x (D + 1) matrix, with the first column corresponding
+  # to the time point at which each observation was collected
+  # this matrix should include the observations from all time points
+
+  source("code/pme.R")
+  source("code/functions/plot_lpme.R")
+  source("code/functions/projection_lpme.R")
+  source("code/functions/solve_eq2.R")
+  require(plotly)
+  require(svMisc)
+  require(Rcpp)
+  require(nloptr)
+  sourceCpp("code/functions/pme_functions.cpp")
+  # source("Principal_Manifold_Estimation.R")
+
+  if (verbose == "none") {
+    print.MSDs <- FALSE
+    print.SSDs <- FALSE
+  } else if (verbose == "MSD") {
+    print.MSDs <- TRUE
+    print.SSDs <- FALSE
+  } else if (verbose == "SSD") {
+    print.MSDs <- FALSE
+    print.SSDs <- TRUE
+  } else if (verbose == "all") {
+    print.MSDs <- TRUE
+    print.SSDs <- TRUE
+  }
+
+  time_points <- df[, 1] %>%
+    unique()
+
+  pme_results <- list()
+  pme_coefs <- list()
+  pme_coefs2 <- list()
+  funcs <- list()
+  clusters <- list()
+  centers <- list()
+  x_test <- list()
+  r <- list()
+  r2 <- list()
+
+  init_timevals <- list()
+  init_theta_hat <- list()
+  init_centers <- list()
+  init_sigma <- list()
+  init_clusters <- list()
+
+  if (init %in% c("first", "full")) {
+    if (init == "first") {
+      init_df <- df[df[, 1] == time_points[1], -1]
+    } else if (init == "full") {
+      for (idx in 1:length(time_points)) {
+        init_dimension_size <- dim(df[, -1])
+        init_D <- init_dimension_size[2]
[…]

### 2025-02-10 code-old/lpme_adni_results.R

@@ -1,0 +1,507 @@
+library(tidyverse)
+library(brolgar)
+library(RColorBrewer)
+
+colors <- brewer.pal(3, "Set1")
+
+roster <- read_csv("data/adni_standard/ROSTER.csv")
+roster <- roster %>% 
+  select(
+    -USERDATE,
+    -USERDATE2,
+    -update_stamp
+  )
+adni_status <- read_csv("data/adni_standard/ADNIMERGE.csv")
+
+# adni_status <- full_join(
+#   adni_status,
+#   roster,
+#   by = c("Phase", "ID", "RID", "SITEID")
+# )
+
+sub_dirs <- list.dirs("results", recursive = FALSE)
+
+hipp_info <- data.frame(
+  patno = character(),
+  date = numeric(),
+  lhipp_data_vol2 = numeric(),
+  lhipp_vol_lpme1 = numeric(),
+  lhipp_vol_lpme2 = numeric(),
+  lhipp_vol_pme1 = numeric(),
+  lhipp_vol_pme2 = numeric(),
+  rhipp_data_vol2 = numeric(),
+  rhipp_vol_lpme1 = numeric(),
+  rhipp_vol_lpme2 = numeric(),
+  rhipp_vol_pme1 = numeric(),
+  rhipp_vol_pme2 = numeric()
+)
+
+thal_info <- data.frame(
+  patno = character(),
+  date = numeric(),
+  lthal_data_vol2 = numeric(),
+  lthal_vol_lpme1 = numeric(),
+  lthal_vol_lpme2 = numeric(),
+  lthal_vol_pme1 = numeric(),
+  lthal_vol_pme2 = numeric(),
+  rthal_data_vol2 = numeric(),
+  rthal_vol_lpme1 = numeric(),
+  rthal_vol_lpme2 = numeric(),
+  rthal_vol_pme1 = numeric(),
+  rthal_vol_pme2 = numeric()
+)
+
+for (dir in sub_dirs) {
+  file_list <- list.files(dir, full.names = FALSE)
+  if ("hipp_info.csv" %in% file_list) {
+    temp_hipp_info <- read_csv(paste0(dir, "/hipp_info.csv"))[, -1]
+    hipp_info <- bind_rows(hipp_info, temp_hipp_info)
+  }
+  if ("thal_info.csv" %in% file_list) {
[…]

### 2025-02-10 code-old/lpme_s3.R

@@ -1,0 +1,783 @@
+library(Rcpp)
+library(vctrs)
+
+source("code/pme.R")
+source("code/functions/plot_lpme.R")
+source("code/functions/projection_lpme.R")
+source("code/functions/solve_eq2.R")
+require(plotly)
+require(svMisc)
+require(Rcpp)
+require(nloptr)
+sourceCpp("code/functions/pme_functions.cpp")
+
+new_lpme <- function(
+    msd,
+    sol_coef,
+    times,
+    r_init,
+    r_fit,
+    d,
+    D,
+    n_knots,
+    lambda,
+    gamma,
+    sol_coef_list,
+    r_fit_list
+) {
+  lpme_list <- list(
+    d = d,
+    D = D,
+    lambda = lambda,
+    gamma = gamma,
+    times = times,
+    n_knots = n_knots,
+    r_init = r_init,
+    r_fit = r_fit,
+    sol_coef = sol_coef,
+    msd = msd,
+    sol_coef_list = sol_coef_list,
+    r_fit_list = r_fit_list
+  )
+  new_vctr(lpme_list, class = "lpme")
+}
+
+is_lpme <- function(x) {
+  inherits(x, "lpme")
+}
+
+lpme <- function(df, d, tuning.para.seq = c(0, exp(-15:10)), alpha = 0.05, max.comp = 500, epsilon = 0.05, max.iter = 100, verbose = "all", print_plots = TRUE, SSD_ratio_threshold = 100, increase_threshold = 1.05, init = "full") {
+  # df is an N x (D + 1) matrix, with the first column corresponding
+  # to the time point at which each observation was collected
+  # this matrix should include the observations from all time points
+
+  source("code/pme.R")
+  source("code/functions/plot_lpme.R")
+  source("code/functions/projection_lpme.R")
+  source("code/functions/solve_eq2.R")
+  require(plotly)
+  require(svMisc)
+  require(Rcpp)
+  require(nloptr)
+  sourceCpp("code/functions/pme_functions.cpp")
+  # source("Principal_Manifold_Estimation.R")
+
+  if (verbose == "none") {
+    print.MSDs <- FALSE
+    print.SSDs <- FALSE
+  } else if (verbose == "MSD") {
+    print.MSDs <- TRUE
+    print.SSDs <- FALSE
[…]

### 2025-02-10 code-old/lpme_simulation_analysis.R

@@ -1,0 +1,197 @@
+library(tidyverse)
+library(reticulate)
+library(pme)
+library(stringr)
+
+# files_case1 <- list.files("simulations/case1", full.names = TRUE)
+# files_case2 <- list.files("simulations/case2", full.names = TRUE)
+# files_case3 <- list.files("simulations/case3", full.names = TRUE)
+# files_case5 <- list.files("simulations/case5", full.names = TRUE)
+# files_case6 <- list.files("simulations/case6", full.names = TRUE)
+# files_case7 <- list.files("simulations/case7", full.names = TRUE)
+# files_case8 <- list.files("simulations/case8", full.names = TRUE)
+# files_case9 <- list.files("simulations/case9", full.names = TRUE)
+# files_case10 <- list.files("simulations/case10", full.names = TRUE)
+#
+# files_full <- c(
+#   files_case1,
+#   files_case2,
+#   files_case3,
+#   files_case5,
+#   files_case6,
+#   files_case7,
+#   files_case8,
+#   files_case9,
+#   files_case10
+# )
+
+files_full <- list.files("simulations", recursive = TRUE, full.names = TRUE)
+
+# files_full <- files_case1
+
+initialization_algorithm <- vector()
+lpme_error <- vector()
+pme_error <- vector()
+# PME_error <- vector()
+pme_ssd <- vector()
+# PME_ssd <- vector()
+principal_curve_ssd <- vector()
+principal_curve_error <- vector()
+data_error <- vector()
+amp_noise <- vector()
+period_noise <- vector()
+n_visits <- vector()
+max_time <- vector()
+time_trend <- vector()
+sim_case <- vector()
+pme_ssd_raw <- vector()
+principal_curve_ssd_raw <- vector()
+for (i in 1:length(files_full)) {
+  error_list <- readRDS(files_full[i])
+  initialization_algorithm[i] <- error_list$initialization_algorithm
+  lpme_error[i] <- error_list$lpme_error
+  pme_error[i] <- error_list$pme_error
+  # PME_error[i] <- error_list$PME_error
+  principal_curve_error[i] <- error_list$principal_curve_error
+
+  # PME_ssd <- c(PME_ssd, error_list$PME_ssd)
+  principal_curve_ssd <- c(principal_curve_ssd, error_list$principal_curve_ssd)
[…]

### 2025-02-10 code-old/lpme_simulations.R

@@ -1,0 +1,2295 @@
+library(tidyverse)
+library(plotly)
+library(pracma)
+library(foreach)
+library(doParallel)
+library(furrr)
+library(progress)
+library(pme)
+library(princurve)
+
+source("code/functions/sim_data.R")
+source("code/functions/calc_pme_est.R")
+source("code/functions/calc_lpme_est.R")
+source("code/prinSurf_v3.R")
+
+### Simulation Case 1
+
+sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
+  time_vals <- seq(0, max_time, interval)
+  sim_list <- lapply(
+    time_vals,
+    sim_data,
+    case = 1,
+    noise = 0.15,
+    amp_noise = amp_noise,
+    period_noise = shape_noise,
+    N = n,
+    time_change = time_change,
+    time_trend = time_trend
+  )
+
+  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
+  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
+  for (i in 1:length(sim_list)) {
+    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
+    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
+  }
+  sim_df <- sim_df[-1, ]
+  true_vals <- true_vals[-1, ]
+
+  for (col_idx in 1:ncol(sim_df)) {
+    col_max <- max(abs(sim_df[, col_idx]))
+    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
+    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
+  }
+
+  time_vals <- unique(sim_df[, 1])
+
+  lpme_result <- lpme(
+    sim_df,
+    1,
+    smoothing_method = "spline",
+    print_plots = print_plots,
+    verbose = TRUE,
+    init = "first",
+    initialization_algorithm = initialization_alg
+  )
+  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
+
+  # lpme_result_gp <- lpme(
+  #   sim_df,
+  #   1,
+  #   smoothing_method = "gp",
+  #   print_plots = print_plots,
+  #   verbose = TRUE
+  # )
+  # lpme_vals_gp <- calc_lpme_est(lpme_result_gp, sim_df)
+
+  pme_result <- list()
[…]

### 2025-02-10 code-old/pme.R

@@ -1,0 +1,507 @@
+#########################################################################
+##########    Principal Manifold Estimation (PME) Algorithm   ###########
+##########                  Kun (Michael) Meng                ###########
+##########    Department of Biostatistics, Brown University   ###########
+#########################################################################
+
+
+########### Introduction ###########################################
+####################################################################
+
+# We propose an R-software function for d-dimensional estimating principle manifolds,
+# where d = 1, 2, 3, and manifolds embedded into D-dimensional Euclidean space with D
+# strictly larger than d.
+
+# The proposed function is "PME()." To construct this function, we need some functions
+# for "high dimensional mixture density estimation (HDMDE)."
+
+# This manuscript - whatever its name is - is organized as follows.
+#   (i)   In Section 1, we define some basic functions.
+#   (ii)  In Section 2, we construct a function for HDMDE. In this section, we first construct
+#         the "weight.seq" function. Then we apply this function to construct "hdmde" function.
+#   (iii) Based on this "hdmde" function, "PME" estimation function is finally given in Section 3.
+# The theoretical foundation of this R-software function is the following paper.
+#
+# K. Meng and A. Eloyan, Principal Manifolds: A framework Using Sobolev Spaces and Model
+#                        Complexity Selection Using Mixture Densities
+
+########### Section 0, Packages ####################################
+####################################################################
+
+#install.packages("MASS")
+library("MASS")
+#install.packages("Matrix")
+library("Matrix")
+#install.packages("vegan")
+library("vegan")
+#install.packages("plot3D")
+library("plot3D")
+library(Rfast)
+library(tidyverse)
+library(plotly)
+library(Rcpp)
+
+
+########### Section 1, Some Basic Functions ########################
+####################################################################
+
+## Subsection 1.1, Functions for Euclidean metrics
+source("code/functions/norm_euclidean.R")
+
[…]

### 2025-02-10 code-old/simulation_test.R

@@ -1,0 +1,1448 @@
+library(tidyverse)
+library(plotly)
+library(pracma)
+library(profvis)
+library(foreach)
+library(doParallel)
+library(doSNOW)
+library(furrr)
+library(progress)
+# library(multimode)
+
+source("code/functions/sim_data.R")
+source("code/functions/calc_pme_est.R")
+source("code/functions/calc_lpme_est.R")
+source("code/pme.R")
+# source("code/lpme.R")
+source("code/lpme_s3.R")
+
+### Simulation Case 1
+
+sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, n, run = 1, print_plots = FALSE) {
+  require(tidyverse)
+  source("code/functions/sim_data.R")
+  source("code/pme.R")
+  # source("code/lpme.R")
+  source("code/lpme_s3.R")
+  source("code/functions/calc_pme_est.R")
+  source("code/functions/calc_lpme_est.R")
+  time_vals <- seq(0, max_time, interval)
+  sim_list <- lapply(
+    time_vals,
+    sim_data,
+    case = 1,
+    noise = 0.15,
+    amp_noise = amp_noise,
+    period_noise = shape_noise,
+    N = n
+  )
+
+  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
+  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
+  for (i in 1:length(sim_list)) {
+    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
+    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
+  }
+  sim_df <- sim_df[-1, ]
+  true_vals <- true_vals[-1, ]
+
+  lpme_result <- lpme(sim_df, 1, print_plots = print_plots, verbose = "MSD")
+  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
+  # lpme_vals[, 1] <- sim_df[, 1]
+  pme_result <- list()
+  pme_vals <- list()
+  for (t in 1:length(time_vals)) {
+    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
+    pme_result[[t]] <- pme(temp_data, d = 1, verbose = "none")
+    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
+  }
[…]

### 2025-02-10 code-old/volume_estimation_simulations.R

@@ -1,0 +1,384 @@
+library(tidyverse)
+library(plotly)
+library(pracma)
+library(foreach)
+library(doParallel)
+library(furrr)
+library(progress)
+library(pme)
+library(princurve)
+library(Rfast)
+
+source("code/functions/sim_data.R")
+source("code/functions/calc_pme_est.R")
+source("code/functions/calc_lpme_est.R")
+source("code/prinSurf_v3.R")
+source("code/functions/estimate_volume.R")
+
+est_sphere_vol <- function(max_time, interval, amp_noise, r, pct_missingness, run) {
+
+  time_vals <- seq(0, max_time, interval)
+  sim_list <- lapply(
+    time_vals,
+    sim_data,
+    case = 9,
+    noise = 0.05,
+    amp_noise = amp_noise,
+    period_noise = 0,
+    N = 1000,
+    time_change = 0,
+    time_trend = "constant"
+  )
+
+  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
+  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
+  for (i in 1:length(sim_list)) {
+    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
+    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
+  }
+  sim_df <- sim_df[-1, ]
+  true_vals <- true_vals[-1, ]
+
+  for (col_idx in 2:ncol(sim_df)) {
+    sim_df[, col_idx] <- r * sim_df[, col_idx]
+    true_vals[, col_idx] <- r * true_vals[, col_idx]
+  }
+
+  true_sph <- cart2sph(true_vals[, -1])
+
+  true_r <- vector()
+  true_volume <- vector()
+  for (time_idx in 1:length(unique(true_vals[, 1]))) {
+    time_val <- unique(true_vals[, 1])[time_idx]
+    true_r[time_idx] <- mean(true_sph[true_vals[, 1] == time_val, 3])
+    true_volume[time_idx] <- (4 / 3) * pi * true_r[time_idx]^3
+  }
+
+  col_maxs <- vector()
+  for (col_idx in 1:ncol(sim_df[, -(5:7)])) {
+    col_max <- max(abs(sim_df[, col_idx]))
+    col_maxs[col_idx] <- col_max
+    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
[…]

### 2025-02-10 code/01_lpme_simulations.qmd.otter.R

@@ -1,0 +1,1021 @@
+
+
+
+
+
+
+
+
+
+
+#| label: chunk_setup
+knitr::opts_chunk$set(message = FALSE, warning = FALSE)
+
+
+
+#| label: load_packages
+library(tidyverse)
+library(plotly)
+library(pracma)
+library(foreach)
+library(doParallel)
+library(furrr)
+library(progress)
+library(pme)
+library(princurve)
+library(Rfast)
+
+source("prinSurf_v3.R")
+source("functions/simulations/simulate_data.R")
+source("functions/simulations/preprocess_data.R")
+source("functions/simulations/fit_models.R")
+source("functions/simulations/evaluate_models.R")
+source("functions/simulations/display_results.R")
+source("functions/calculate_lpme_reconstructions.R")
+source("functions/calculate_pme_reconstructions.R")
+source("functions/estimate_volume.R")
+source("functions/interior_identification.R")
+
+
+
+
+
+
+
+
+#| label: plot_case1
+set.seed(100)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
+  case = 1,
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
+
+plot_ly(
+  data = sim_df,
+  x = ~X1,
+  y = ~time,
+  z = ~X2,
+  type = "scatter3d",
+  mode = "markers",
+  marker = list(size = 3, color = "black"),
+  name = "Data"
+) |>
+  add_markers(
+    x = ~X_true1,
+    y = ~time,
+    z = ~X_true2,
+    marker = list(size = 3, color = "red"),
+    name = "True Manifold"
+  )
+
+
+
+
+#| label: plot_case2
+set.seed(200)
+sim_data <- simulate_data(
+  duration = 1,
+  interval = 0.25,
+  case = 2,
+  obs_noise = 0.1,
+  amplitude_noise = 0.25,
+  period_noise = 0.05,
+  time_trend = "constant",
+  time_change = 0,
+  N = 1000
+)
+sim_df <- sim_data$df
+
+plot_ly(
+  data = sim_df,
+  x = ~X1,
+  y = ~time,
+  z = ~X2,
+  type = "scatter3d",
+  mode = "markers",
[…]

### 2025-02-10 code/adni_mri_preprocess.R

@@ -1,527 +1,0 @@
-library(oro.dicom)
-library(oro.nifti)
-library(neurobase)
-library(fslr)
-# library(extrantsr)
-library(stringr)
-library(foreach)
-library(tidyverse)
-library(ANTsRCore)
-
-sub_dirs <- list.dirs("data/adni", recursive = FALSE)
-
-nii_scans <- list()
-
-processed_scans <- list()
-
-hipp_mask <- readNIfTI(
-  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm_Hipp_mask_dil8.nii.gz")
-)
-
-mni_nifti <- readNIfTI(
-  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm.nii.gz")
-)
-
-ncores <- parallel::detectCores() / 2
-cl <- parallel::makeCluster(ncores, type = "FORK")
-doParallel::registerDoParallel(cl = cl)
-
-foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
-# foreach(dir_idx = 1:8) %dopar% {
-  img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
-    gsub(pattern = "([^/]+$)", replacement = "") %>%
-    unique()
-
-  nii_scans[[dir_idx]] <- list()
-
-  for (img_idx in 1:length(img_dirs)) {
-    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/adni", replacement = "", img_dirs[img_idx]))
-    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
-      next
-    } else {
-      all_slices <- readDICOM(img_dirs[img_idx])
-      nii <- dicom2nifti(all_slices)
-
-      # Inhomogeneity correction
-      # Brain extraction / segmentation
-      # Image Registration
-      # Tissue Class Segmentation
-
-      # nii_scans[[dir_idx]][[img_idx]] <- nii
-
-      nii <- fslreorient2std(nii, verbose = FALSE) # orient nii to mni space
-      # inhomogeneity correction
-      # nii <- bias_correct(
-      #   nii,
-      #   correction = "N4",
-      #   verbose = FALSE
-      # ) # N4 bias correction using ANTs
-      # nii_bc <- fast(nii, bias_correct = TRUE)
-
-      # nii <- fast(nii, bias_correct = TRUE)
[…]

### 2025-02-10 code/functions/calc_lpme_est.R

@@ -1,40 +1,0 @@
-calc_lpme_est <- function(result, df) {
-  opt_idx <- which.min(result$msd)
-  r_full <- expand_grid(result$times, result$initial_parameterization) %>%
-    as.matrix()
-  # x_vals <- map(
-  #   1:nrow(r_full),
-  #   ~ embed(result, r_full[.x, ])
-  # ) %>%
-  #   reduce(rbind)
-  x_vals <- map(
-    1:nrow(r_full),
-    ~ result$embedding_map(r_full[.x, ])
-  ) %>%
-    reduce(rbind)
-  tnew_init <- result$optimal_parameterization
-
-  nearest_x <- map(
-    1:nrow(df),
-    ~ which.min(apply(x_vals[x_vals[, 1] == df[.x, 1], ], 1, dist_euclidean, y = df[.x, ]))
-  ) %>%
-    reduce(c)
-
-  init_param <- map(
-    1:nrow(df),
-    ~ tnew_init[tnew_init[, 1] == df[.x, 1], ][nearest_x[.x], ]
-  ) %>%
-    reduce(rbind)
-
-  tnew <- map(
-    1:nrow(df),
-    ~ projection_lpme(df[.x, ], function(x) result$embedding_map(x), init_param[.x, ])
-  ) %>%
-    reduce(rbind)
-  results <- map(
-    1:nrow(df),
-    ~ result$embedding_map(tnew[.x, ])
-  ) %>%
-    reduce(rbind)
-  return(results)
-}

### 2025-02-10 code/functions/calc_lpme_est_params.R

@@ -1,40 +1,0 @@
-calc_lpme_est_params <- function(result, df) {
-  opt_idx <- which.min(result$msd)
-  r_full <- expand_grid(result$times, result$initial_parameterization) %>%
-    as.matrix()
-  # x_vals <- map(
-  #   1:nrow(r_full),
-  #   ~ embed(result, r_full[.x, ])
-  # ) %>%
-  #   reduce(rbind)
-  x_vals <- map(
-    1:nrow(r_full),
-    ~ result$embedding_map(r_full[.x, ])
-  ) %>%
-    reduce(rbind)
-  tnew_init <- result$optimal_parameterization
-
-  nearest_x <- map(
-    1:nrow(df),
-    ~ which.min(apply(x_vals[x_vals[, 1] == df[.x, 1], ], 1, dist_euclidean, y = df[.x, ]))
-  ) %>%
-    reduce(c)
-
-  init_param <- map(
-    1:nrow(df),
-    ~ tnew_init[tnew_init[, 1] == df[.x, 1], ][nearest_x[.x], ]
-  ) %>%
-    reduce(rbind)
-
-  tnew <- map(
-    1:nrow(df),
-    ~ projection_lpme(df[.x, ], function(x) result$embedding_map(x), init_param[.x, ])
-  ) %>%
-    reduce(rbind)
-  results <- map(
-    1:nrow(df),
-    ~ result$embedding_map(tnew[.x, ])
-  ) %>%
-    reduce(rbind)
-  return(list(results = results, params = tnew))
-}

### 2025-02-10 code/functions/calc_pme_est.R

@@ -1,26 +1,0 @@
-calc_pme_est <- function(pme, df) {
-  center_order <- order(pme$knots$centers[, 1])
-  centers <- pme$knots$centers[center_order, ]
-
-  nearest_clusters <- map(
-    1:nrow(df),
-    ~ which.min(as.vector(apply(centers, 1, dist_euclidean, y = df[.x, ])))
-  ) %>%
-    reduce(c)
-
-  opt_run <- which.min(pme$MSD)
-  tnew_init <- pme$parameterization[[opt_run]][nearest_clusters, ] %>%
-    matrix(nrow = nrow(df), byrow = TRUE)
-
-  tnew <- map(
-    1:nrow(df),
-    ~ projection_pme(df[.x, ], pme$embedding_map, tnew_init[.x, ])
-  ) %>%
-    reduce(rbind)
-  results <- map(
-    1:nrow(df),
-    ~ pme$embedding_map(tnew[.x, ])
-  ) %>%
-    reduce(rbind)
-  return(results)
-}

### 2025-02-10 code/functions/calculate_lpme_reconstructions.R

@@ -1,0 +1,71 @@
+calculate_lpme_reconstructions <- function(lpme_result, data) {
+  #' Calculate LPME Reconstructions
+  #'
+  #' Uses a fitted LPME model to estimate reconstructed observations in D-dimensional space
+  #'
+  #' @param lpme_result The fitted LPME model
+  #' @param data A numeric matrix including a column of time values and d-dimensional parameters
+
+  # obtain lpme run with optimal tuning value and
+  # parameter values for the associated cluster centers
+  lpme_opt <- which.min(lpme_result$msd)
+  params <- lpme_result$parameterization_list[[lpme_opt]]
+
+  # convert the cluster center parameters to D dimensions
+  centers <- map(
+    1:nrow(params),
+    ~ lpme_result$embedding_map(params[.x, ])
+  ) |>
+    reduce(rbind)
+
+  # within the clusters calculated for each time point, find the nearest
+  # cluster center for each observation.
+  # adjustments are needed to compare to indices among full set of clusters.
+  nearest_center <- map(
+    1:nrow(data),
+    ~ which.min(
+      apply(
+        centers[centers[, 1] == data[.x, 1], ],
+        1,
+        dist_euclidean,
+        y = data[.x, ]
+      )
+    )
+  ) |>
+    reduce(c)
+  center_idx_adjustment <- map(
+    1:length(nearest_center),
+    ~ nrow(
+      centers[centers[, 1] < data[.x, 1], ]
+    )
+  ) |>
+    reduce(c)
+  nearest_center <- nearest_center + center_idx_adjustment
+
+  # obtain the parameters associated with each nearest_cluster
+  nearest_params <- map(
+    1:length(nearest_center),
+    ~ params[nearest_center[.x], ]
+  ) |>
+    reduce(rbind)
+
+  # find d-dimensional projections to the estimated manifold for each observation
+  lpme_projections <- map(
+    1:nrow(data),
+    ~ projection_lpme(
+      data[.x, ],
+      function(x) lpme_result$embedding_map(x),
+      nearest_params[.x, ]
+    )
+  ) |>
+    reduce(rbind)
+
+  # embed projections in D-dimensional space to calculate reconstructions
+  lpme_reconstructions <- map(
+    1:nrow(data),
+    ~ lpme_result$embedding_map(lpme_projections[.x, ])
+  ) |>
+    reduce(rbind)
+
+  return(lpme_reconstructions)
+}

### 2025-02-10 code/functions/calculate_pme_reconstructions.R

@@ -1,0 +1,62 @@
+calculate_pme_reconstructions <- function(pme_result, data) {
+  #' Calculate PME Reconstructions
+  #'
+  #' Uses a fitted PME model to estimate reconstructed observations in D-dimensional space
+  #'
+  #' @param lpme_result The fitted PME model
+  #' @param data A numeric matrix of d-dimensional parameters, with time values excluded
+
+  require(purrr)
+
+  # obtain pme run with optimal tuning value and
+  # parameter values for the associated cluster centers
+  pme_opt <- which.min(pme_result$MSD)
+  params <- pme_result$parameterization[[pme_opt]]
+
+  centers <- map(
+    seq_len(nrow(params)),
+    ~ pme_result$embedding_map(params[.x, ])
+  ) |>
+    reduce(rbind)
+
+  # calculate the cluster centers nearest to each observation
+  nearest_center <- map(
+    seq_len(nrow(data)),
+    ~ which.min(
+      apply(
+        centers,
+        1,
+        dist_euclidean,
+        y = data[.x, ]
+      )
+    )
+  ) |>
+    reduce(c)
+
+  # obtain the parameters associated with each nearest_cluster
+  nearest_params <- map(
+    seq_along(nearest_center),
+    ~ params[nearest_center[.x], ]
+  ) |>
+    reduce(rbind)
+
+  # find d-dimensional projections to the estimated manifold for each observation
+  pme_projections <- map(
+    seq_len(nrow(data)),
+    ~ projection_pme(
+      data[.x, ],
+      function(x) pme_result$embedding_map(x),
+      nearest_params[.x, ]
+    )
+  ) |>
+    reduce(rbind)
+
+  # embed projections in D-dimensional space to calculate reconstructions
+  pme_reconstructions <- map(
+    seq_len(nrow(data)),
+    ~ pme_result$embedding_map(pme_projections[.x, ])
+  ) |>
+    reduce(rbind)
+
+  return(pme_reconstructions)
+}

### 2025-02-10 code/functions/estimate_volume.R

@@ -1,4 +1,6 @@
 estimate_volume <- function(data, voxel_volume) {
+  require(Rfast, quietly = TRUE, warn.conflicts = FALSE)
+
   indices <- matrix(ncol = 3)
   data_limits <- colMinsMaxs(data)
 
@@ -11,11 +13,11 @@ estimate_volume <- function(data, voxel_volume) {
   x_mins <- matrix(nrow = length(y_seq), ncol = length(z_seq))
   x_maxs <- matrix(nrow = length(y_seq), ncol = length(z_seq))
 
-  for (y_idx in 1:length(y_seq)) {
-    for (z_idx in 1:length(z_seq)) {
+  for (y_idx in seq_along(y_seq)) {
+    for (z_idx in seq_along(z_seq)) {
       y_val <- y_seq[y_idx]
       z_val <- z_seq[z_idx]
-      data_red <- data[data[, 2] == y_val & data[, 3] == z_val, ] %>%
+      data_red <- data[data[, 2] == y_val & data[, 3] == z_val, ] |>
         matrix(ncol = ncol(data))
 
       if (nrow(data_red) == 0) {
@@ -35,11 +37,11 @@ estimate_volume <- function(data, voxel_volume) {
     }
   }
 
-  for (y_idx in 1:length(y_seq)) {
-    for (z_idx in 1:length(z_seq)) {
+  for (y_idx in seq_along(y_seq)) {
+    for (z_idx in seq_along(z_seq)) {
       y_val <- y_seq[y_idx]
       z_val <- z_seq[z_idx]
-      data_red <- data[data[, 2] == y_val & data[, 3] == z_val, ] %>%
+      data_red <- data[data[, 2] == y_val & data[, 3] == z_val, ] |>
         matrix(ncol = ncol(data))
 
       if (nrow(data_red) == 0) {
@@ -81,10 +83,12 @@ estimate_volume <- function(data, voxel_volume) {
     missing_mins <- matrix(nrow = nrow(x_mins), ncol = ncol(x_mins))
     missing_maxs <- matrix(nrow = nrow(x_maxs), ncol = ncol(x_maxs))
     n_missing <- 0
-    for (y_idx in 1:length(y_seq)) {
-      for (z_idx in 1:length(z_seq)) {
+    for (y_idx in seq_along(y_seq)) {
+      for (z_idx in seq_along(z_seq)) {
         if (is.na(x_mins[y_idx, z_idx])) {
[…]

### 2025-02-10 code/functions/interior_identification.R

@@ -1,35 +1,66 @@
-interior_identification <- function(embedding_map, coefs, params, x, reference) {
-  orientation_c <- get_orientation(embedding_map, coefs, params, reference)[1]
+interior_identification <- function(embedding_map, coefs, params, x, ref) {
+  require(pme, quietly = TRUE, warn.conflicts = FALSE)
+  require(pracma, quietly = TRUE, warn.conflicts = FALSE)
+  require(purrr, quietly = TRUE, warn.conflicts = FALSE)
+  require(wordspace, quietly = TRUE, warn.conflicts = FALSE)
+
+  param_centers <- map(
+    seq_len(nrow(params)),
+    ~ embedding_map(params[.x, ])
+  ) |>
+    reduce(rbind)
+
+  orientation_c <- get_orientation(
+    embedding_map,
+    coefs,
+    param_centers,
+    params,
+    ref
+  )[1]
   orientation_x <- map(
-    1:nrow(x), 
-    ~ get_orientation(embedding_map, coefs, params, unlist(x[.x, ]))
-  ) %>% 
+    seq_len(nrow(x)),
+    ~ get_orientation(
+      embedding_map,
+      coefs,
+      param_centers,
+      params,
+      unlist(x[.x, ])
+    )
+  ) %>%
     unlist()
   return(orientation_x == orientation_c)
 }
 
-  get_orientation <- function(embedding_map, coefs, params, x) {
+get_orientation <- function(embedding_map, coefs, centers, params, x) {
   d <- ncol(params)
   n <- nrow(params)
 
-  augment_x <- c(x, cart2sph(x))
-  x_params <- projection_pme(augment_x, embedding_map, params[1, ])
+  augment_x <- c(x, cart2sph(x))[-6]
+  nearest_center <- map(
+    seq_len(nrow(centers)),
+    ~ dist_euclidean(augment_x, as.vector(centers[.x, ]))
+  ) |>
+    reduce(c) |>
+    which.min()
+  x_params <- projection_pme(augment_x, embedding_map, params[nearest_center, ])
   x_proj <- embedding_map(x_params)
 
-  
-  param_diff <- -1 * sweep(params, 2, x_params)
-  param_diff_norms <- wordspace::rowNorms(param_diff)
 
[…]

### 2025-02-10 code/functions/interior_identification2.R

@@ -1,0 +1,90 @@
+interior_identification <- function(embedding_map, coefs, params, x, ref) {
+  require(pme, quietly = TRUE, warn.conflicts = FALSE)
+  require(pracma, quietly = TRUE, warn.conflicts = FALSE)
+  require(purrr, quietly = TRUE, warn.conflicts = FALSE)
+  require(wordspace, quietly = TRUE, warn.conflicts = FALSE)
+
+  param_centers <- map(
+    seq_len(nrow(params)),
+    ~ embedding_map(params[.x, ])
+  ) |>
+    reduce(rbind)
+
+  augment_x <- cbind(x, cart2sph(x))[, -6]
+  augment_c <- c(ref, cart2sph(ref))[-6]
+  nearest_centers <- vector(mode = "numeric", length = nrow(augment_x))
+  for (row_idx in seq_len(nrow(augment_x))) {
+    nearest_centers[row_idx] <- map(
+      seq_len(nrow(param_centers)),
+      ~ dist_euclidean(
+        as.vector(augment_x[row_idx, ]),
+        as.vector(param_centers[.x, ])
+      )
+    ) |>
+      reduce(c) |>
+      which.min()
+  }
+
+  nearest_center_c <- map(
+    seq_len(nrow(param_centers)),
+    ~ dist_euclidean(
+      as.vector(augment_c),
+      as.vector(param_centers[.x, ])
+    )
+  ) |>
+    reduce(c) |>
+    which.min()
+
+  c_param <- projection_pme(
+    augment_c,
+    embedding_map,
+    params[nearest_center_c, ]
+  )
+  c_proj <- embedding_map(c_param)
+  c_diff <- augment_c - c_proj
+
+  x_params <- map(
+    seq_len(nrow(augment_x)),
+    ~ projection_pme(
+      as.vector(augment_x[.x, ]),
+      embedding_map,
+      params[nearest_centers[.x], ]
+    )
+  ) |>
+    reduce(rbind)
+  x_proj <- map(
+    seq_len(nrow(x_params)),
+    ~ embedding_map(as.vector(x_params[.x, ]))
+  ) |>
+    reduce(rbind)
+  x_diff <- augment_x - x_proj
+
+  angles <- map(
+    seq_len(nrow(x_diff)),
+    ~ acos(
+      (x_diff[.x, ] %*% c_diff)[1] /
+        (norm_euclidean(x_diff[.x, ]) * norm_euclidean(c_diff))
+    )
+  ) |>
[…]

### 2025-02-10 code/functions/lpme_init.R

@@ -1,125 +1,0 @@
-lpme_init <- function(df, d, alpha = 0.05, max.comp = 100, epsilon = 0.05, max.iter = 100, init = "first") {
-
-  # init argument can take three values:
-  # "first" indicates that an isomap initialization run on the first time point is used
-  # "full" indicates that an isomap initialization run on the full dataset (all time points) is used
-  # "separate" indicates that each time point will use its own isomap initialization
-
-  source("code/pme.R")
-  require(plotly)
-
-  time_points <- df[, 1] %>%
-    unique()
-
-  pme_results <- list()
-  funcs <- list()
-  x_test <- list()
-  r <- list()
-
-  msd <- list()
-
-  if (init %in% c("first", "full")) {
-    if (init == "first") {
-      init_df <- df[df[, 1] == time_points[1], -1]
-    } else if (init == "full") {
-      init_df <- df[, -1]
-    }
-
-    init_dimension.size <- dim(init_df)
-    init_D <- init_dimension.size[2]
-    init_n <- init_dimension.size[1]
-    lambda <- 4 - d
-
-    init_N0 <- 20 * init_D
-    init_est <- hdmde(init_df, init_N0, alpha, max.comp)
-    init_order <- order(init_est$mu[, 1])
-    init_theta.hat <- init_est$theta.hat[init_order]
-    init_centers <- init_est$mu[init_order, ]
-    init_sigma <- init_est$sigma
-    init_W <- diag(init_theta.hat)
-    init_X <- init_est$mu[init_order, ]
-    init_I <- length(init_theta.hat)
-
-    init_dissimilarity.matrix <- as.matrix(dist(init_X))
-    init_isomap <- isomap(init_dissimilarity.matrix, ndim = d, k = 10)
-  }
-
-  for (idx in 1:length(time_points)) {
-    df_temp <- df[df[, 1] == time_points[idx], ]
-    if (init %in% c("first", "full")) {
-      pme_results[[idx]] <- pme(
-        x.obs = df_temp[, -1],
-        d = d,
-        initialization = list(init_est, init_isomap)
-      )
-    } else {
-      pme_results[[idx]] <- pme(
[…]

### 2025-02-10 code/functions/plot_lpme_complete.R

@@ -1,9 +1,0 @@
-plot_lpme_complete <- function(df, result) {
-  opt_result <- which.min(result$MSD)
-  # t_new <- result$TNEW[[opt_result]]
-  coefs <- result$coefs[[opt_result]]
-  d_new <- ncol(coefs)
-  D_new <- ncol(df) - 1
-  time_points <- unique(df[,  1])
-  plot_lpme(df, result$embedding_map, coefs, d_new, D_new, time_points)
-}

### 2025-02-10 code/functions/pme_functions.cpp

@@ -1,195 +1,0 @@
-#include <RcppCommon.h>
-#include <dlib/optimization/optimization.h>
-
-#include <RcppArmadillo.h>
-// [[Rcpp::depends(RcppArmadillo)]]
-using namespace Rcpp;
-
-// [[Rcpp::depends(dlib)]]
-//#include <dlib/optimization.h>
-
-
-using namespace std;
-using namespace dlib;
-
-
-#include <numeric>
-#include <vector>
-#include <cmath>
-#include <complex>
-
-// [[Rcpp::export]]
-double norm_euclideanC(arma::vec x) {
-  double sq_sum = 0;
-  for (int i = 0; i < x.size(); i++) {
-    sq_sum += pow(x[i], 2);
-  }
-  return sqrt(sq_sum);
-}
-
-// [[Rcpp::export]]
-double dist_euclideanC(arma::vec x, arma::vec y) {
-  return norm_euclideanC(x - y);
-}
-
-// [[Rcpp::export]]
-double dist_euclideanF(arma::vec x, arma::vec y, Function f) {
-  NumericVector mod_x = f(x);
-  return dist_euclideanC(mod_x, y);
-}
-
-// [[Rcpp::export]]
-arma::vec dist_euclideanC_vec(arma::mat x, arma::mat y) {
-  arma::vec distances(x.n_rows);
-  for (int i = 0; i < x.n_rows; i++) {
-    distances(i) = dist_euclideanC(x.row(i).t(), y.row(i).t());
-  }
-  return distances;
-}
-
-// [[Rcpp::export]]
-double eta_kernelC(arma::vec t, int lambda) {
-
-  double lambda_num = lambda / 1.0;
-  double norm_val = norm_euclideanC(t);
-  double y;
-  if (lambda % 2 == 0) {
-    if (norm_val == 0) {
-      y = 0;
-    } else {
-      y = pow(norm_val, lambda_num) * log(norm_val);
-    }
-  } else {
-    y = pow(norm_val, lambda_num);
-  }
-  return y;
-}
-
-// [[Rcpp::export]]
-arma::mat calcE(arma::mat x, int lambda) {
-  int nrow = x.n_rows;
-  int ncol = x.n_cols;
-  arma::mat E(nrow, nrow);
-
-  for (int i = 0; i < nrow; i++) {
-    for (int j = 0; j < nrow; j++) {
[…]

### 2025-02-10 code/functions/simulations/evaluate_models.R

@@ -1,0 +1,60 @@
+evaluate_models <- function(data, models, case, d, D) {
+  require(dplyr, quietly = TRUE, warn.conflicts = FALSE)
+  require(purrr, quietly = TRUE, warn.conflicts = FALSE)
+
+  true_values <- data$df_true |>
+    select(time, contains("X")) |>
+    as.matrix()
+
+  observed_data <- data$df_observed |>
+    select(time, contains("X")) |>
+    as.matrix()
+
+  lpme_error <- map(
+    seq_len(nrow(true_values)),
+    ~ dist_euclidean(
+      true_values[.x, ],
+      models$lpme$reconstructions[.x, 1:(1 + D)]
+    )^2
+  ) |>
+    unlist() |>
+    mean()
+
+  pme_error <- map(
+    seq_len(nrow(true_values)),
+    ~ dist_euclidean(
+      true_values[.x, ], models$pme$reconstructions[.x, 1:(1 + D)]
+    )^2
+  ) |>
+    unlist() |>
+    mean()
+
+  if (!is.null(models$pc$reconstructions)) {
+    principal_curve_error <- map(
+      seq_len(nrow(true_values)),
+      ~ dist_euclidean(
+        true_values[.x, ], models$pc$reconstructions[.x, 1:(1 + D)]
+      )^2
+    ) |>
+      unlist() |>
+      mean()
+  } else {
+    principal_curve_error <- NA
+  }
+
+  data_error <- map(
+    seq_len(nrow(true_values)),
+    ~ dist_euclidean(true_values[.x, ], observed_data[.x, ])^2
+  ) |>
+    unlist() |>
+    mean()
+
+  error_list <- list(
+    lpme_error = lpme_error,
+    pme_error = pme_error,
+    principal_curve_error = principal_curve_error,
+    data_error = data_error
+  )
+
+  return(error_list)
+}

### 2025-02-10 code/functions/simulations/fit_models.R

@@ -1,0 +1,231 @@
+fit_models <- function(data, case, d, D) {
+  ##### MANIFOLD LEARNING #####
+
+  require(dplyr, quietly = TRUE, warn.conflicts = FALSE)
+  require(pme, quietly = TRUE, warn.conflicts = FALSE)
+  require(princurve, quietly = TRUE, warn.conflicts = FALSE)
+  require(purrr, quietly = TRUE, warn.conflicts = FALSE)
+
+  if (case != 8) {
+    mat <- as.matrix(data$df_observed)
+
+    lpme_results <- list()
+
+    lpme_results[[1]] <- lpme(
+      mat,
+      d,
+      verbose = FALSE,
+      print_plots = FALSE
+    )
+    lpme_reconstructions <- calculate_lpme_reconstructions(
+      lpme_results[[1]],
+      mat
+    )
+
+    # fit PME and appropriate principal curve/surface algorithms and
+    # calculate reconstructions for each
+    pme_result_list <- list()
+    pc_result_list <- list()
+
+    pme_reconstruction_list <- list()
+    pc_reconstruction_list <- list()
+
+    smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
+
+    time_values <- unique(data$df$time)
+    for (time_idx in seq_along(time_values)) {
+      temp_data <- mat[mat[, 1] == time_values[time_idx], -1]
+      pme_result_list[[time_idx]] <- pme(temp_data, d = d)
+      pme_reconstruction_list[[time_idx]] <- calculate_pme_reconstructions(
+        pme_result_list[[time_idx]],
+        temp_data
+      )
+      pme_reconstruction_list[[time_idx]] <- cbind(
+        time_values[time_idx],
+        pme_reconstruction_list[[time_idx]]
+      )
+      if (d == 1) {
+        principal_curves <- list()
+        pc_error <- vector()
+        for (smoother_idx in seq_along(smoothing_options)) {
+          principal_curves[[smoother_idx]] <- principal_curve(
+            temp_data,
+            smoother = smoothing_options[[smoother_idx]]
+          )
+          pc_error[smoother_idx] <- principal_curves[[smoother_idx]]$dist
+        }
+        opt_principal_curve <- which.min(pc_error)
+        pc_result_list[[time_idx]] <- principal_curves[[opt_principal_curve]]
+        pc_reconstruction_list[[time_idx]] <- cbind(
+          time_values[time_idx],
[…]

### 2025-02-10 code/functions/simulations/preprocess_data.R

@@ -1,0 +1,81 @@
+preprocess_data <- function(df, case, d, D) {
+  require(dplyr, quietly = TRUE, warn.conflicts = FALSE)
+  require(pracma, quietly = TRUE, warn.conflicts = FALSE)
+
+
+  ##### PREPROCESSING *****
+
+  # standardize observations to fall between -1 and 1 in each dimension
+  # use maximum observed values to standardize both observations and true values
+  time_values <- unique(df$time)
+
+  for (dim_idx in 1:D) {
+    obs_dim_idx <- 1 + d + dim_idx
+    true_dim_idx <- 1 + d + D + dim_idx
+    col_max <- max(abs(df[obs_dim_idx]))
+    df[, obs_dim_idx] <- df[, obs_dim_idx] / col_max
+    df[, true_dim_idx] <- df[, true_dim_idx] / col_max
+  }
+  df$time <- df$time / max(df$time)
+  time_values <- time_values / max(time_values)
+
+  # where, if anywhere, should case 7 (swiss roll) be included?
+  polar_cases <- c(3)
+  spherical_cases <- c(8)
+
+  if (case %in% polar_cases) {
+    observed_polar <- df |>
+      select(contains("X")) |>
+      select(-contains("true")) |>
+      as.matrix() |>
+      cart2pol()
+
+    true_polar <- df |>
+      select(contains("true")) |>
+      as.matrix() |>
+      cart2pol()
+
+    observed_polar <- observed_polar[, 1]
+    true_polar <- true_polar[, 1]
+
+    df$phi <- observed_polar
+    df$phi_true <- true_polar
+  } else if (case %in% spherical_cases) {
+    observed_spherical <- df |>
+      select(contains("X")) |>
+      select(-contains("true")) |>
+      as.matrix() |>
+      cart2sph()
+
+    true_spherical <- df |>
+      select(contains("true")) |>
+      as.matrix() |>
+      cart2sph()
+
+    observed_spherical_df <- observed_spherical[, -3] |>
+      as.data.frame()
+    names(observed_spherical_df) <- c("theta", "phi")
+
+    true_spherical_df <- true_spherical[, -3] |>
+      as.data.frame()
+    names(true_spherical_df) <- c("theta_true", "phi_true")
+
+    df <- df |>
+      bind_cols(observed_spherical_df) |>
+      bind_cols(true_spherical_df)
+  }
+
+  df_observed <- df |>
[…]

### 2025-02-10 code/functions/simulations/simulate_data.R

@@ -1,0 +1,258 @@
+#' Simulate Data
+#'
+#' Generates simulated data from given underlying manifold.
+#'
+#' @param duration The simulated study's follow-up duration
+#' @param interval The scheduled interval of study visits
+#' @param case Specifies the manifold used to generate the data
+#' @param obs_noise The standard deviation of the independent Gaussian observation-level noise
+#' @param amplitude_noise The standard deviation of Gaussian noise applied to the amplitude of the manifold at each time point
+#' @param period_noise The standard deviation of Gaussian noise applied to the period of the manifold at each time point
+#' @param time_trend Specifies systematic longitudinal changes in the manifold amplitude
+#' @param time_change Scales the longitudinal amplitude changes
+#' @param N The number of observations generated at each time point
+#'
+simulate_data <- function(
+    duration,
+    interval,
+    case,
+    obs_noise,
+    amplitude_noise,
+    period_noise,
+    time_trend,
+    time_change,
+    N = 1000) {
+  # instead of using set interval between visits consider using a parameter
+  # for the expected number of follow up visits within the study
+  # then draw the observed number of visits from a poisson distribution
+  # visit times will then be uniformly distributed between 0 and the duration
+
+  # simulate_data <- function(duration, exp_follow_up, case, obs_noise, amplitude_noise, period_noise, time_trend, time_change, N = 1000) {
+
+  require(dplyr, quietly = TRUE, warn.conflicts = FALSE)
+  require(purrr, quietly = TRUE, warn.conflicts = FALSE)
+  require(tibble, quietly = TRUE, warn.conflicts = FALSE)
+
+  # create list of functions used to define manifolds of interest
+  manifolds <- list(
+    # case 1: sine curve with limited domain
+    function(tau, amplitude, period) {
+      c(tau, amplitude[1] * sin(period[1] * tau + pi / 2))
+    },
+    # case 2: sine curve with extended domain
+    function(tau, amplitude, period) {
+      c(tau, amplitude[1] * sin(period[1] * tau))
+    },
+    # case 3: partial circle
+    function(tau, amplitude, period) {
[…]

### 2025-02-10 code/functions/weight_seq.R

@@ -1,64 +1,0 @@
-weight_seq <- function(x.obs, mu, sigma, epsilon = 0.001, max.iter = 1000) {
-  # x.obs is the dataset of interest
-  # there are n observations, and each observation is a D-dimensional point
-  # x.obs is a n-by-D matrix
-
-  # mu is a vector of the knots in a mixture density estimation
-  # sigma is the bandwidth of the density estimation
-  # epsilon is a predetermined tolerance of the Euclidean distance between thetas in two consecutive steps
-  # max.iter is a predetermined upper bound on the number of steps
-
-  n.D <- dim(x.obs)
-  n <- n.D[1]
-  D <- n.D[2]
-  N <- dim(mu)[1]
-
-  A <- matrix(NA, nrow = n, ncol = N)
-  for (j in 1:N) {
-    A.prepare <- function(x) {
-      return(ker(x, mu[j, ], sigma))
-    }
-    A[, j] <- apply(x.obs, 1, A.prepare)
-  }
-
-  theta.old <- rep(1 / N, N)
-  abs.diff <- 10 * epsilon
-  count <- 0
-  lambda.hat.old <- c(n, rep(-1, D))
-
-  while((abs.diff > epsilon) & (count <= max.iter)) {
-    W <- t(t(A) * theta.old)
-    W <- W / rowsums(W)
-    w <- colsums(W)
-
-    flambda <- function(lambda) {
-
-      denom.temp <- rowsums(t(t(cbind(rep(1, dim(mu)[1]), mu)) * lambda))
-      num.temp <- mu * w
-
-      f1 <- sum(w / denom.temp)
-      f2 <- colsums(num.temp * (as.vector(1 / denom.temp)))
-      f <- dist_euclidean(f1, 1) + dist_euclidean(f2, apply(x.obs, 2, mean))
-      return(f)
-    }
-
-    lambda.hat <- nlm(flambda, lambda.hat.old, iterlim = 1000)$estimate
-
-    theta.new <- w / rowsums(t(t(cbind(rep(1, dim(mu)[1]), mu)) * lambda.hat))
-
-    abs.diff <- dist_euclidean(theta.new, theta.old)
-    if (is.na(abs.diff)) {
-      abs.diff <- 0
-      theta.new <- theta.old
-    }
-
-    theta.old <- pmax(theta.new, 0)
-    theta.old <- pmin(theta.old, 1)
-    count <- count + 1
[…]

### 2025-02-10 code/lpme.R

@@ -1,706 +1,0 @@
-lpme <- function(df, d, tuning.para.seq = c(0, exp(-15:10)), alpha = 0.05, max.comp = 500, epsilon = 0.05, max.iter = 100, verbose = "all", print_plots = TRUE, SSD_ratio_threshold = 100, increase_threshold = 1.05, init = "full") {
-  # df is an N x (D + 1) matrix, with the first column corresponding
-  # to the time point at which each observation was collected
-  # this matrix should include the observations from all time points
-
-  source("code/pme.R")
-  source("code/functions/plot_lpme.R")
-  source("code/functions/projection_lpme.R")
-  source("code/functions/solve_eq2.R")
-  require(plotly)
-  require(svMisc)
-  require(Rcpp)
-  require(nloptr)
-  sourceCpp("code/functions/pme_functions.cpp")
-  # source("Principal_Manifold_Estimation.R")
-
-  if (verbose == "none") {
-    print.MSDs <- FALSE
-    print.SSDs <- FALSE
-  } else if (verbose == "MSD") {
-    print.MSDs <- TRUE
-    print.SSDs <- FALSE
-  } else if (verbose == "SSD") {
-    print.MSDs <- FALSE
-    print.SSDs <- TRUE
-  } else if (verbose == "all") {
-    print.MSDs <- TRUE
-    print.SSDs <- TRUE
-  }
-
-  time_points <- df[, 1] %>%
-    unique()
-
-  pme_results <- list()
-  pme_coefs <- list()
-  pme_coefs2 <- list()
-  funcs <- list()
-  clusters <- list()
-  centers <- list()
-  x_test <- list()
-  r <- list()
-  r2 <- list()
-
-  init_timevals <- list()
-  init_theta_hat <- list()
-  init_centers <- list()
-  init_sigma <- list()
-  init_clusters <- list()
-
-  if (init %in% c("first", "full")) {
-    if (init == "first") {
-      init_df <- df[df[, 1] == time_points[1], -1]
-    } else if (init == "full") {
-      for (idx in 1:length(time_points)) {
-        init_dimension_size <- dim(df[, -1])
-        init_D <- init_dimension_size[2]
[…]

### 2025-02-10 code/lpme_s3.R

@@ -1,783 +1,0 @@
-library(Rcpp)
-library(vctrs)
-
-source("code/pme.R")
-source("code/functions/plot_lpme.R")
-source("code/functions/projection_lpme.R")
-source("code/functions/solve_eq2.R")
-require(plotly)
-require(svMisc)
-require(Rcpp)
-require(nloptr)
-sourceCpp("code/functions/pme_functions.cpp")
-
-new_lpme <- function(
-    msd,
-    sol_coef,
-    times,
-    r_init,
-    r_fit,
-    d,
-    D,
-    n_knots,
-    lambda,
-    gamma,
-    sol_coef_list,
-    r_fit_list
-) {
-  lpme_list <- list(
-    d = d,
-    D = D,
-    lambda = lambda,
-    gamma = gamma,
-    times = times,
-    n_knots = n_knots,
-    r_init = r_init,
-    r_fit = r_fit,
-    sol_coef = sol_coef,
-    msd = msd,
-    sol_coef_list = sol_coef_list,
-    r_fit_list = r_fit_list
-  )
-  new_vctr(lpme_list, class = "lpme")
-}
-
-is_lpme <- function(x) {
-  inherits(x, "lpme")
-}
-
-lpme <- function(df, d, tuning.para.seq = c(0, exp(-15:10)), alpha = 0.05, max.comp = 500, epsilon = 0.05, max.iter = 100, verbose = "all", print_plots = TRUE, SSD_ratio_threshold = 100, increase_threshold = 1.05, init = "full") {
-  # df is an N x (D + 1) matrix, with the first column corresponding
-  # to the time point at which each observation was collected
-  # this matrix should include the observations from all time points
-
-  source("code/pme.R")
-  source("code/functions/plot_lpme.R")
-  source("code/functions/projection_lpme.R")
-  source("code/functions/solve_eq2.R")
-  require(plotly)
-  require(svMisc)
-  require(Rcpp)
-  require(nloptr)
-  sourceCpp("code/functions/pme_functions.cpp")
-  # source("Principal_Manifold_Estimation.R")
-
-  if (verbose == "none") {
-    print.MSDs <- FALSE
-    print.SSDs <- FALSE
-  } else if (verbose == "MSD") {
-    print.MSDs <- TRUE
-    print.SSDs <- FALSE
[…]

### 2025-02-10 code/lpme_simulations.R

@@ -1,2295 +1,0 @@
-library(tidyverse)
-library(plotly)
-library(pracma)
-library(foreach)
-library(doParallel)
-library(furrr)
-library(progress)
-library(pme)
-library(princurve)
-
-source("code/functions/sim_data.R")
-source("code/functions/calc_pme_est.R")
-source("code/functions/calc_lpme_est.R")
-source("code/prinSurf_v3.R")
-
-### Simulation Case 1
-
-sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
-  time_vals <- seq(0, max_time, interval)
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 1,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n,
-    time_change = time_change,
-    time_trend = time_trend
-  )
-
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  for (col_idx in 1:ncol(sim_df)) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
-    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
-  }
-
-  time_vals <- unique(sim_df[, 1])
-
-  lpme_result <- lpme(
-    sim_df,
-    1,
-    smoothing_method = "spline",
-    print_plots = print_plots,
-    verbose = TRUE,
-    init = "first",
-    initialization_algorithm = initialization_alg
-  )
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-
-  # lpme_result_gp <- lpme(
-  #   sim_df,
-  #   1,
-  #   smoothing_method = "gp",
-  #   print_plots = print_plots,
-  #   verbose = TRUE
-  # )
-  # lpme_vals_gp <- calc_lpme_est(lpme_result_gp, sim_df)
-
-  pme_result <- list()
[…]

### 2025-02-10 code/pme.R

@@ -1,507 +1,0 @@
-#########################################################################
-##########    Principal Manifold Estimation (PME) Algorithm   ###########
-##########                  Kun (Michael) Meng                ###########
-##########    Department of Biostatistics, Brown University   ###########
-#########################################################################
-
-
-########### Introduction ###########################################
-####################################################################
-
-# We propose an R-software function for d-dimensional estimating principle manifolds,
-# where d = 1, 2, 3, and manifolds embedded into D-dimensional Euclidean space with D
-# strictly larger than d.
-
-# The proposed function is "PME()." To construct this function, we need some functions
-# for "high dimensional mixture density estimation (HDMDE)."
-
-# This manuscript - whatever its name is - is organized as follows.
-#   (i)   In Section 1, we define some basic functions.
-#   (ii)  In Section 2, we construct a function for HDMDE. In this section, we first construct
-#         the "weight.seq" function. Then we apply this function to construct "hdmde" function.
-#   (iii) Based on this "hdmde" function, "PME" estimation function is finally given in Section 3.
-# The theoretical foundation of this R-software function is the following paper.
-#
-# K. Meng and A. Eloyan, Principal Manifolds: A framework Using Sobolev Spaces and Model
-#                        Complexity Selection Using Mixture Densities
-
-########### Section 0, Packages ####################################
-####################################################################
-
-#install.packages("MASS")
-library("MASS")
-#install.packages("Matrix")
-library("Matrix")
-#install.packages("vegan")
-library("vegan")
-#install.packages("plot3D")
-library("plot3D")
-library(Rfast)
-library(tidyverse)
-library(plotly)
-library(Rcpp)
-
-
-########### Section 1, Some Basic Functions ########################
-####################################################################
-
-## Subsection 1.1, Functions for Euclidean metrics
-source("code/functions/norm_euclidean.R")
-
[…]

### 2025-02-10 code/run_simulations.R

@@ -1,0 +1,213 @@
+library(tidyverse)
+library(plotly)
+library(pracma)
+library(furrr)
+library(progress)
+library(pme)
+library(princurve)
+library(doSNOW)
+library(foreach)
+library(doRNG)
+
+setwd("code")
+
+source("prinSurf_v3.R")
+source("functions/simulations/simulate_data.R")
+source("functions/simulations/preprocess_data.R")
+source("functions/simulations/fit_models.R")
+source("functions/simulations/evaluate_models.R")
+source("functions/simulations/display_results.R")
+source("functions/calculate_lpme_reconstructions.R")
+source("functions/calculate_pme_reconstructions.R")
+
+durations <- c(1, 2, 5)
+intervals <- c(0.25, 0.5)
+cases <- 1:8
+obs_noises <- c(0.05, 0.1, 0.25, 0.5)
+amplitude_noises <- c(0.05, 0.1, 0.25, 0.5)
+period_noises <- c(0.05, 0.1, 0.25, 0.5)
+time_trends <- c("constant", "linear", "quadratic")
+time_changes <- c(0, 0.05, 0.1, 0.25, 0.5)
+
+
+parameter_df <- expand_grid(
+  duration = durations,
+  interval = intervals,
+  case = cases,
+  obs_noise = obs_noises,
+  amplitude_noise = amplitude_noises,
+  period_noise = period_noises,
+  time_trend = time_trends,
+  time_change = time_changes
+)
+
+error_df <- tibble(
+  lpme_error = vector(mode = "numeric", length = nrow(parameter_df)),
+  pme_error = vector(mode = "numeric", length = nrow(parameter_df)),
+  pc_error = vector(mode = "numeric", length = nrow(parameter_df)),
+  data_error = vector(mode = "numeric", length = nrow(parameter_df))
+)
+
+seed_states <- list()
+
+print("Starting Simulations...")
+
+cl <- makeCluster(parallel::detectCores() - 1)
+registerDoSNOW(cl)
+
+registerDoRNG(611492)
+
+pb  <- txtProgressBar(max = nrow(parameter_df), style = 3)
+progress <- function(n) setTxtProgressBar(pb, n)
+opts <- list(progress = progress)
+error_df <- foreach(
+  row_idx = seq_len(nrow(parameter_df)),
+  .export = c(
+    "simulate_data",
+    "preprocess_data",
+    "fit_models",
+    "evaluate_models",
+    "display_results"
+  ),
[…]

### 2025-02-10 code/simulation_test.R

@@ -1,1448 +1,0 @@
-library(tidyverse)
-library(plotly)
-library(pracma)
-library(profvis)
-library(foreach)
-library(doParallel)
-library(doSNOW)
-library(furrr)
-library(progress)
-# library(multimode)
-
-source("code/functions/sim_data.R")
-source("code/functions/calc_pme_est.R")
-source("code/functions/calc_lpme_est.R")
-source("code/pme.R")
-# source("code/lpme.R")
-source("code/lpme_s3.R")
-
-### Simulation Case 1
-
-sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, n, run = 1, print_plots = FALSE) {
-  require(tidyverse)
-  source("code/functions/sim_data.R")
-  source("code/pme.R")
-  # source("code/lpme.R")
-  source("code/lpme_s3.R")
-  source("code/functions/calc_pme_est.R")
-  source("code/functions/calc_lpme_est.R")
-  time_vals <- seq(0, max_time, interval)
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 1,
-    noise = 0.15,
-    amp_noise = amp_noise,
-    period_noise = shape_noise,
-    N = n
-  )
-
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  lpme_result <- lpme(sim_df, 1, print_plots = print_plots, verbose = "MSD")
-  lpme_vals <- calc_lpme_est(lpme_result, sim_df)
-  # lpme_vals[, 1] <- sim_df[, 1]
-  pme_result <- list()
-  pme_vals <- list()
-  for (t in 1:length(time_vals)) {
-    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
-    pme_result[[t]] <- pme(temp_data, d = 1, verbose = "none")
-    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
-  }
[…]

### 2025-02-10 code/volume_estimation_simulations.R

@@ -1,384 +1,0 @@
-library(tidyverse)
-library(plotly)
-library(pracma)
-library(foreach)
-library(doParallel)
-library(furrr)
-library(progress)
-library(pme)
-library(princurve)
-library(Rfast)
-
-source("code/functions/sim_data.R")
-source("code/functions/calc_pme_est.R")
-source("code/functions/calc_lpme_est.R")
-source("code/prinSurf_v3.R")
-source("code/functions/estimate_volume.R")
-
-est_sphere_vol <- function(max_time, interval, amp_noise, r, pct_missingness, run) {
-
-  time_vals <- seq(0, max_time, interval)
-  sim_list <- lapply(
-    time_vals,
-    sim_data,
-    case = 9,
-    noise = 0.05,
-    amp_noise = amp_noise,
-    period_noise = 0,
-    N = 1000,
-    time_change = 0,
-    time_trend = "constant"
-  )
-
-  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
-  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
-  for (i in 1:length(sim_list)) {
-    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
-    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
-  }
-  sim_df <- sim_df[-1, ]
-  true_vals <- true_vals[-1, ]
-
-  for (col_idx in 2:ncol(sim_df)) {
-    sim_df[, col_idx] <- r * sim_df[, col_idx]
-    true_vals[, col_idx] <- r * true_vals[, col_idx]
-  }
-
-  true_sph <- cart2sph(true_vals[, -1])
-
-  true_r <- vector()
-  true_volume <- vector()
-  for (time_idx in 1:length(unique(true_vals[, 1]))) {
-    time_val <- unique(true_vals[, 1])[time_idx]
-    true_r[time_idx] <- mean(true_sph[true_vals[, 1] == time_val, 3])
-    true_volume[time_idx] <- (4 / 3) * pi * true_r[time_idx]^3
-  }
-
-  col_maxs <- vector()
-  for (col_idx in 1:ncol(sim_df[, -(5:7)])) {
-    col_max <- max(abs(sim_df[, col_idx]))
-    col_maxs[col_idx] <- col_max
-    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
[…]

### 2025-02-10 run_simulations.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+#SBATCH --job-name=lpme_simulations
+#SBATCH --partition defq
+#SBATCH -n 64
+#SBATCH -o lpme_simulations_%j.out
+#SBATCH -e lpme_simulations_%j.err
+#SBATCH --mail-type=ALL
+#SBATCH --mail-user=robert_zielinski1@brown.edu
+R CMD BATCH code/run_simulations.R

### 2025-02-24 manuscript/Revision 1/Reviews/main.tex

\documentclass{article}
\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}

\newcommand{\blind}{1}
\graphicspath{{figures/}}

\pdfminorversion=4

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\T}{\intercal}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\setcitestyle{aysep={}}

\begin{document}

\begin{center}
\bigskip {\Large Response to the Editor}

{\large Submission No.: 1}

{\large \bigskip Longitudinal Principal Manifold Estimation}
\end{center}

\noindent We would like to thank the Editor and Referees for their insightful comments and helpful suggestions. In this revised version, we have addressed all the comments by the AE and Referees. Point-by-point responses to these comments are presented herein.

In responding to all the comments, we have followed the following convention: each of the comments is italicized and followed by our response to the corresponding comment; any changes in the manuscript based on the corresponding comment are presented in \textcolor{blue}{blue} beneath our response.

We have formatted the manuscript using the latex template obtained from the journal website. Particularly, while implementing the revisions suggested by the AE and Referee, we maintained the manuscript length of 24 pages, as suggested by the journal website.

\vspace{0.3in}

\large{\textbf{Response to Editor}}

\vspace{0.3in}

{\it While the development of a general methodology for manifold estimation is worthwhile, you need to give more thought to how to make this general approach more suitable for the specific application. }

\begin{enumerate}

\item {\it For example, you note that your method struggles to get the shape of the hippocampus because the true shape has sharp curves in it.  Couldn't/shouldn't your methodology be modified to take account of this feature of the hippocampus?  As an analogy, when using cubic splines to fit one-dimensional curves, one can get curves that aren't differentiable by putting more than one knot at a single location.  There should be some way of doing something similar for manifold estimation.  I think it would be very interesting to try some (perhaps ad hoc) modification of your penalty function that would allow you to better fit this aspect of the hippocampus shape.}

{\bf Response:}

\item {\it Your methodological development is highly algorithmic with no actual model for the shapes of brain regions or for the measurement process.  In your simulations, it looks like your simulated data includes independent measurement error (but you need to make that clear).  Given the pre-processing steps in your data analysis, I suspect that the actual "errors" in your observation are spatially dependent.  If so, then you need to discuss this issue and do some simulations that try to more realistically mimic the measurement process in your actual data.}

{\bf Response:}

\item {\it Your criterion function (4) includes terms with derivatives in space and another with a derivative in time, but no terms that include mixed partials in space and time.  I am wondering what are the implications of this choice and how much your result would change if you included some kind of integral of a squared mixed partial derivative in (4).  I'm not sure how far you would want to go down this avenue, but I do think it is at least worth acknowledging that the form of (4) is somewhat arbitrary and that other forms might be worth considering.  I would also note that (4) puts no penalty on large linear trends in time, which may not be biologically realistic.}

{\bf Response:}

\item {\it I recommend you add the application to the title of your paper, so something like "Longitudinal Principal Manifold Estimation for Modeling Changes Over Time in Brain Regions".}

\end{enumerate}

\large{\textbf{Response to Referee 1}}

\vspace{0.25in}

The paper proposes a novel method called Longitudinal Principal Manifold Estimation (LPME) to fit the spatial Principal Manifolds smoothed over time for longitudinal data. I have the following comments:

\begin{enumerate}
\item In line 24 on page 3, it says "to develop smooth estimates of the surfaces of subcortical structures over time." Why not also consider cortical structure?

{\bf Response:}

\item In lines 30-31 on page 4, it says "the deep learning methodology used imposes costs in terms of computational requirements and interpretability". This sentence has a bias on deep learning (DL). Nowadays, GPU is almost a necessary device for brain image data analysis, due to powerful DL models in image classification and segmentation. Also, Google CoLab provides free access to GPU. DL models are very fast on GPU. For interpretability of DL, many tools have been developed. See the topic of Explainable AI.

{\bf Response:}

\item Line 15 on page 9 says "... by a study participant." Is this method used for the longitudinal data of a single subject, not simultaneously applied to a group of subjects?

\item Also line 9 on page 10 says "for each image i". But the index i is from 1 to $I_t$, so the $x_it$ is an image, not just a 3D voxel? Is the method only for subject-level data analysis, not for group-level data analysis?

\item In Eq. (3), (t,r) should be (r, t) since the domain is $R^d x R$ not $R x R^d$.

\item In figure 3, the plots for time=0.3 \& 0.6, the true manifold (red curve) doesn’t well align with the scatter plot. Is it because the noise is too strong?
And the plot for time =0.9, the LPME (blue curve) is two-mode, not as smooth as the other 2 methods and the true manifold.

\item In line 19-26 on page 25, why only consider the 463 subjects from the total 800 subjects of ADNI 1?

\item The MCI group is very heterogeneous. It's better to consider only the subgroup of MCI with later conversion to AD.

\item Line 5 on page 25 says "...meaningful between-image errors … are introduced ..." Is "meaningful" a typo?

\item For the ADNI data, you used 236 subjects from the three groups, CN, MCI, and AD,  for image segmentation, but used LPME to fit each subject individually? Where is the comparison between the 3 groups CN, AD, and MCI?

\item Timing performance of the proposed method is needed for simulation and real-data analysis.

\end{enumerate}

\vspace{0.3in}

\large{\textbf{Response to Referee 2}}

\vspace{0.25in}

{\it The authors of this manuscript propose and evaluate a longitudinal principal manifold estimation method and a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. The developed methodology is relatively computationally efficient and provides good approximation for the changes in the brain volume structures. The proposed method, LPME, is also applied to the data obtained from the longitudinal ADNI study.  The work undertaken is interesting and the new method and its application to the ADNI study data is a worthwhile undertaking. However, there are a few issues that should be addressed. First, the described methodology depends on a relatively strong assumption that the underlying manifold is minimally changing over time and the changes can be well described by a smooth time function. Second, it is not clear if the method proposed applies only to discrete time points or if it can be used in the continuous time setting, i.e. with irregular time intervals. Third, the simulation studies summarized indicate that the new method is advantageous is specific simulation settings. Are such settings relevant to the brain structure changes? Forth, is it feasible to estimate smooth changes over time with a very limited number of longitudinal time points. Please find additional comments below.}

Specific comments

\begin{enumerate}
\item Abstract:    Need to be more specific about the type of the MRI used. I believe the authors mean the structural MRI (sMRI).

\item Abstract:    The plural of “hippocampus” is “hippocampi.”

\item Introduction, p.2:  References on processing of the sMRI data are dated. There has been a lot of progress in the past decade in the sMRI processing pipelines.

\item Introduction, p.5: Are the monotonicity restrictions included in the modeling? For example, volumes of the brain structures should not increase over time.

\item Section 3, p.9:    Justification for the use of manifolds to learn about the volume changes is not fully satisfactory as volumes are scalar quantities. One does not need to estimate the surface changes in order to obtain a volume of a 3D object.

\item Section 3.1, p.11: Description of the minimization problem is unclear.

\item Section 3.2.2:    The choice of the first time point for the initialization is arbitrary and prone to error. Why the longitudinal information is not used in the initialization?

\item Table 2, p.23 and Table S1: While in many simulation settings (Cases), the LPME algorithm seems to perform well. Case 7 looks to be troublesome. Any specific reason why in this case PC/PS method performs the best on average?

\item Figure 2, p.24:    The figure is difficult to interpret as the colors for LPME and Principal Curves are similar and the symbols are hard to differentiate among the methods.

\end{enumerate}

%\bibliographystyle{plain}
%\bibliography{sample}

\end{document}

### 2025-02-24 manuscript/paper/main.tex

@@
 Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods.
-The proposed method improves the estimation of surfaces of \outdated{hippocampuses} \revision{hippocampi} and \outdated{thalamuses} \revision{thalami} using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
+The proposed method improves the estimation of surfaces of \revision{hippocampi} and  \revision{thalami} using data from participants of the Alzheimer’s Disease Neuroimaging Initiative.
 An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
